### 2024-12-23 draft_v1/draft_v1.qmd
draft

---
title: "Causal Discovery on Precarity and Depression"
#subtitle: Using the HELIUS dataset
author: 
  - name: Kyuri Park
    affil-id: 1
  - name: Leonie K. Elsenburg
    affil-id: 2
  - name: Mary Nicolao
    affil-id: 2
  - name: Karien Stronks
    affil-id: 2
  - name: Vítor V. Vasconcelos
    affil-id: 1, 3

affiliations:
  - id: 1
    name: \textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}
  - id: 2
    name: \textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}
  - id: 3
    name: \textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}   
format: 
  # docx:
  #   number-sections: true
  #   appendix: true
  pdf:
    pdf-engine: pdflatex
    fontfamily: palatino
    toc: true
    number-sections: true
    appendix: true
    colorlinks: true
    keep-tex: true
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
        \usepackage{amssymb}
        \usepackage{booktabs} % Add to your preamble for cleaner table lines
        \usepackage{makecell} % Add to your preamble for multi-line cells
date: today
execute:
  echo: false
  warning: false
bibliography: references_draft.bib
csl: apa.csl
abstract: |
  \noindent Understanding the causal mechanisms linking precarity factors and depression is critical for developing effective interventions. This study utilizes the HELIUS dataset to explore these relationships using advanced causal discovery methods. By applying algorithms such as FCI, and CCI, and combining traditional Gaussian CI tests with non-parametric approaches like RCoT, we investigate how precarity factors—including employment, social, financial, housing, and relational stress—affect depression, both as a sum score and at the individual symptom level. Our findings reveal that relational stress consistently emerges as a potential causal factor for depression, while symptoms such as sleep disturbances, guilt, and anhedonia are particularly sensitive to external stressors, acting as potential early warning signals or intervention points for prevention. Moreover, the results highlight complexities in the data, including the influence of latent confounders and the challenges of capturing cyclic relationships. Despite some limitations, such as unresolved ambiguities in causal directions and challenges with mixed data distributions, this study demonstrates the utility of causal discovery tools in disentangling the intricate interplay between social and mental health dynamics. By mapping these causal structures into computational models, future research can simulate intervention effects, providing actionable insights to mitigate the impact of precarity on mental health. This study serves as a foundational effort, offering both methodological advancements and practical implications for addressing depression at a population level.
---

```{r, include=FALSE}
library(haven)
library(dplyr)
library(skimr)
#library(bestNormalize)
library(furrr)
library(future)
library(tidyr)
library(GGally)
library(purrr)
library(gt)

library(corrplot)
library(psych)
library(FactoMineR)
library(factoextra)
library(fastICA)
library(umap)

```

```{r, include=FALSE}
# Load the dataset 
dat <- read_sav("../data/HELIUS_LEONIE.sav")

# Preprocess and scale the precariousness data
scaled_data <- dat %>%
  # Select precariousness-related variables
                # Employment Precariousness
  dplyr::select(H1_Arbeidsparticipatie, H1_WerkSit, H1_RecentErv8, 
                # Financial Precariousness
                H1_InkHhMoeite, H1_RecentErv9,                      
                # Housing Precariousness
                veilig_2012, vrz_2012, P_HUURWON,                   
                # Cultural Precariousness
                H1_Discr_sumscore, H1_SBSQ_meanscore, A_BED_RU,         
                # Social Precariousness
                H1_RecentErv5, H1_RecentErv6, H1_RecentErv7,       
                H1_SSQT, H1_SSQSa,                                  
                # Depression symptoms / sum scores
                H1_WlbvRecent1, H1_WlbvRecent2, H1_WlbvRecent3, H1_WlbvRecent4, H1_WlbvRecent5, H1_WlbvRecent6, H1_WlbvRecent7, H1_WlbvRecent_89, H1_WlbvRecent10, H1_PHQ9_sumscore,   
                # Ethnicity
                H1_etniciteit,
                # Ags
                H1_lft) %>%                                
  # Replace missing codes with NA 
  mutate(across(everything(), ~ na_if(.x, -9)), 
         across(everything(), ~ na_if(.x, -1))) %>%  
  # Remove NAs 
  na.omit() %>%
  # Convert all columns to numeric
  mutate(across(everything(), as.numeric),
        # Combine categories 5 to 10 into a single category (5-10 as one unemployed)
         H1_WerkSit = case_when(
    H1_WerkSit %in% c(5, 6, 7, 8, 9, 10) ~ 5,  
    .default = H1_WerkSit  # Keep other values as they are 
  ))  %>%
  # Reverse the values of specific variables (reverse transformation)
  mutate(across(c(H1_RecentErv8, H1_RecentErv9, veilig_2012, vrz_2012, 
                  H1_SBSQ_meanscore, A_BED_RU, H1_RecentErv5, H1_RecentErv6, 
                  H1_RecentErv7, H1_SSQT, H1_SSQSa), 
                ~ max(.x) - .x)) %>%
  # scale the data
  mutate(across(!c(H1_etniciteit, H1_lft), scale))
  # Apply Min-Max scaling to all columns
  # mutate(across(everything(), min_max_scaling))

# redefine colnames
colnames(scaled_data) <- c(
  "emp_stat",       # H1_Arbeidsparticipatie
  "work_sit",   # H1_WerkSit
  "unemp12",     # H1_RecentErv8
  "inc_dif",      # H1_InkHhMoeite
  "fincri12",# H1_RecentErv9
  "nb_safe",     # veilig_2012
  "nb_res",   # vrz_2012
  "nb_rent",         # P_HUURWON
  "discrim",      # H1_Discr_sumscore
  "hea_lit",       # H1_SBSQ_meanscore
  "cul_rec",      # A_BED_RU
  "rel_end12",   # H1_RecentErv5
  "frd_brk12", # H1_RecentErv6
  "conf12",         # H1_RecentErv7
  "soc_freq",      # H1_SSQT
  "soc_adq",      # H1_SSQSa
  "anh",       # H1_WlbvRecent1
  "dep",       # H1_WlbvRecent2
  "slp",       # H1_WlbvRecent3
  "ene",       # H1_WlbvRecent4
  "app",       # H1_WlbvRecent5
  "glt",       # H1_WlbvRecent6
  "con",       # H1_WlbvRecent7
  "mot",     # H1_WlbvRecent_89
  "sui",      # H1_WlbvRecent10
  "PHQsum",         # H1_PHQ9_sumscore
  "ethn",         # H1_etniciteit
  "age"          #"H1_lft" 
)
```

# Introduction

Mental health problems in urban areas have been reported to be on the rise. Governments have been attempting to intervene, but the complexity of mental health systems presents significant challenges in planning effective interventions, let alone understanding the underlying mechanisms driving these issues.

Recent research has aimed to identify underlying factors contributing to mental health problems, often referred to as precariousness factors. These factors encompass various aspects of life, such as employment, social connections, financial stability, housing, and cultural dimensions. This comprehensive perspective on precarity helps to highlight how different aspects of life may be interconnected. While this research has advanced our understanding of the complex interplay between precariousness factors, a key question remains unanswered: how do these factors influence mental health? Specifically, the lack of directional information --- knowing what influences what --- limits our ability to identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between precariousness factors and mental health outcomes, with a specific focus on depression, the most prevalent mental health issue in urban populations. Using causal discovery methods, we explore how different aspects of precarity influence depression and delve deeper into the dynamics at the symptom level. By examining individual depressive symptoms, we aim to identify which symptoms may act as initiators by being particularly sensitive to specific precariousness factors. Through this analysis, our goal is to uncover the causal mechanisms underlying mental health challenges and provide a foundation for developing more effective and targeted interventions.

# Methods

## Data

We use the HELIUS dataset, which captures the diverse population of Amsterdam across various ethnicities and provides comprehensive health and lifestyle data. To operationalize precariousness factors, we draw on the framework outlined in previous research (i.e., Leonie's paper) and select a set of relevant variables.

To ensure a robust representation of each precariousness factor, we conducted various exploratory analyses to identify consistent and meaningful factor structures. Based on these analyses, we identified five precariousness factors, including two related to recent stressors, each comprising multiple variables as outlined below. Detailed information on the exploratory analyses can be found in the [Appendix](#sec-appendix).

-   Employment precariousness: `emp_stat`, `work_sit`.
-   Social precariousness: `soc_freq`, `soc_adq`.
-   Housing precariousness: `nb_safe`, `nb_res`, `nb_rent`, `cul_rec`.
-   Recent relational stressors: `frd_brk12`, `conf12`.
-   Recent financial stressors: `fincri12`, `inc_diff`.

After preprocessing, the HELIUS dataset comprises 21,628 samples. In addition to the five precariousness factors, we include PHQ-9 scores — both the total sum score and individual symptom scores — to represent depression. In the subsequent causal discovery analysis, we will examine depression both as an aggregated sum score and through its individual symptom-level representations. Refer to @fig-dist for the overall distributions of the variables used in the analysis.

```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%"}
#| fig-cap: >
#|   Distributions of variables with density overlay.  
#|   
#|   *P.emp* = employment precariousness; *P.hou* = housing precariousness; *P.soc* = social precariousness; *S.fin* = recent financial stressors; *S.rel* = recent relational stressors;
#|   *PHQsum* = PHQ-9 sum score; *anh* = anhedonia; *app* = appetite; *con* = concentration; *dep* = depressed mood; *ene* = energy; *glt* = guilty; *mot* = motor; *sui* = suicidal
#| label: fig-dist

clust_data <- scaled_data |> 
  mutate(
    P.emp = rowMeans(cbind(emp_stat, work_sit)),
    P.soc = rowMeans(cbind(soc_freq, soc_adq)),
    P.hou = rowMeans(cbind(nb_safe, nb_res, nb_rent, cul_rec)),
    S.rel = rowMeans(cbind(frd_brk12, conf12)),
    S.fin = rowMeans(cbind(fincri12, inc_dif)),
    Precarity = P.emp + P.soc + P.hou + S.rel + S.fin
  ) |>
  dplyr::select(17:34) |>
  as.data.frame()

# Reshape the data to long format
clust_data_long <- clust_data |>
  dplyr::select(-age, -ethn) |>
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Plot distributions with density lines
ggplot(clust_data_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "white", bins = 15) +  # Use density for histogram
  geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "Precarity", "PHQsum", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui")), scales = "free") +
  labs(x = "", y = "") +
  theme_minimal()
```

## Causal Discovery

There are numerous causal discovery algorithms available; however, in this study, we focus on algorithms suited to the potential cyclic relationships within our system. Specifically, we use FCI (Fast Causal Inference) and CCI (Cyclic Causal Inference), both capable of accounting for such cycles under certain conditions [@mooij2020constraint; @strobl2019]. Additionally, we include the PC algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation]. For a more detailed explanation of these algorithms, please refer to @park2024discovering.

\begin{table}[ht]
\centering
\caption{Assumptions of causal discovery algorithms}
\begin{tabular}{lccccl}
\toprule
Algorithm & Acyclicity & Causal sufficiency & 
\makecell{Absence of \\ selection bias} & Output \\ 
\midrule
PC  & $\checkmark$ & $\checkmark$ & $\checkmark$ & CPDAG \\ 
FCI & $-^{a}$ & $\checkmark$ & $-^{a}$ & PAG \\ 
CCI & x & x & x & MAAG \\ 
\bottomrule
\end{tabular}
\caption*{\small{\textit{Note}. $^{a}$FCI can detect cycles in systems that lack selection bias and exhibit non-linear relationships (Mooij \& Claassen, 2020).}}
\end{table}

<!-- For word docx compiling -->
<!-- **Table 1**: Assumptions of causal discovery algorithms. -->

<!-- | Algorithm | Acyclicity | Causal sufficiency | Absence of selection bias | Output | -->
<!-- |-----------|------------|--------------------|---------------------------|--------| -->
<!-- | PC        | ✓          | ✓                  | ✓                         | CPDAG  | -->
<!-- | FCI       | -¹         | ✓                  | -¹                        | PAG    | -->
<!-- | CCI       | x          | x                  | x                         | MAAG   | -->
<!-- ------------------------------------------------------------------------------------ -->
<!-- ¹FCI can detect cycles in systems that lack selection bias and exhibit non-linear relationships (Mooij & Claassen, 2020). -->

As shown in Table 1, the resulting graphs from FCI and CCI differ slightly (*PAG*: partial ancestral graph; *MAAG*: maximal almost ancestral graph) due to their reliance on different underlying assumptions. Both encode information about causal relationships between variables, where the presence of an edge indicates causal *ancestry*.
Directed edges, $A$ \*→ $B$, specify that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$. $A$\*—$B$ represent cases where $B$ is an ancestor of $A$ in every graph in $Equiv(G)$. Circle endpoints, $A$\*-o$B$, denote ambiguity in the ancestral relationship, meaning that $B$'s ancestral status relative to $A$ varies across graphs in $Equiv(G)$.

In contrast, the graph produced by the PC algorithm is a *CPDAG* (completed partially directed acyclic graph), where directed edges ($A$ → $B$) indicate that $A$ is a direct cause (parent) of $B$. Unlike FCI and CCI, the CPDAG does not include circle symbols. Instead, when the PC algorithm cannot determine the direction, it represents uncertainty with bidirectional arrows. While PC serves as a useful reference, its strict assumptions of acyclicity and the absence of latent confounders limit its applicability in more complex settings. Therefore, our primary focus remains on the results from FCI and CCI. For completeness, all PC algorithm results are provided in the [Appendix](#sec-appendix).

One practical challenge in applying these algorithms to the HELIUS dataset is that the data does not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear. To account for this, we complement the commonly used Gaussian conditional independence test (CI test), which relies on partial correlations, with a non-parametric CI test based on kernel methods. However, kernel-based non-parametric tests are computationally demanding, particularly with large datasets like ours. To mitigate this issue, we employ Randomized Conditional Independence Test (RCIT) and the Randomized conditional Correlation Test (RCoT), which uses random Fourier features to approximate the kernel methods, thereby significantly reducing the computational cost [@strobl2019approximate]. For a more detailed explanation of RCIT and RCoT, see @sec-rcot.

## Analysis
We analyze the causal structure using two approaches: one with the PHQ sum score representing overall depression severity, and another with individual symptom scores. The PHQ sum score simplifies the analysis by reducing dimensionality, which is computationally advantageous and provides a broad, interpretable perspective on depression's relationship with precarity factors. In contrast, individual symptom scores offer a nuanced understanding, capturing the heterogeneous ways symptoms respond to precarity factors. However, analyzing individual symptoms poses methodological challenges due to their non-standard distributions and the increased complexity introduced by the higher dimensionality. By employing both approaches, we balance simplicity and granularity, ensuring robustness in our findings: the PHQ sum score captures overarching trends, while symptom-level analysis reveals detailed dynamics, essential for tailoring specific interventions.

To evaluate the sensitivity of the results to the choice of alpha levels and ensure consistency, we test two significance levels --- $\alpha = 0.01, 0.05$. To further ensure robustness, we employ bootstrapping to generate 100 bootstrap samples For each sample, we estimate causal graphs and retain only the edges and directions that appear above predefined thresholds.

The analyses are conducted under the following conditions:

-   Significance levels ($\alpha$): 0.01 and 0.05
-   Thresholds: 0.5, 0.6, 0.7, and 0.8
-   CI test: Gaussian CI test, RCoT
-   Algorithms: FCI, CCI, and PC

This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms, with each combination repeated for 100 bootstrap samples, resulting in a total of 1,600 resulting graphs. For analyses involving individual symptom variables, we streamline the setup by using thresholds of 0.6 and 0.7, reducing the number of bootstrap samples to 30. Additionally, we fix the skeleton of edges among symptom variables based on the common structure estimated across all three algorithms, thereby reducing the computational time required to estimate skeletons. This fixed structure aligns with commonly observed skeleton structures in the literature, ensuring relevance to existing findings (**cite the comp_model paper**).[^2]

To summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups. This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression. Refer to @fig-workflow for the analysis workflow.

[^2]: 
To reduce spurious edges in the symptom network, we fixed the skeleton of symptom interactions using a common structure derived from PC, FCI, and CCI algorithms with an alpha level of 0.001. This approach maintained sparsity in symptom interactions while allowing the free estimation of causal directions between symptoms and precarity factors, balancing computational efficiency with meaningful results.

```{r}
#| label: fig-workflow
#| fig-cap: "Analysis workflow applied across all three algorithms."
#| fig-align: center
#| out-width: 80%

knitr::include_graphics("img/simsetup.png")

```
<!-- \begin{figure}[htbp] -->
<!--     \centering -->
<!--         \caption{Simulation workflow} -->
<!--         \includegraphics[width=0.9\textwidth]{img/simsetup.png} -->
<!-- \end{figure} -->

# Results
## Depression as sum score
The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
@fig-sum illustrates the causal relationships between precarity factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*) under two different setups: (a) using both Gaussian CI test and RCoT, and (b) using RCoT alone. Black edges represent consistent causal relationships identified by both FCI and CCI, while gray edges denote inconsistent relationships that vary between the two methods. The edge endpoints of inconsistent gray edges are marked with circles.
In graph (a), the key pathways suggest that employment precarity (*P.emp*) and social precarity (*P.soc*) do not cause depression (*PHQsum*). Social precarity does not cause recent relational stress (*S.rel*) or financial stress (*S.fin*), and these stressors are likely related by a latent confounder. Additionally, *P.emp* and *S.fin* are identified as non-causes of housing precarity (*P.hou*). Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or if the connection is mediated by an unmeasured confounder. A similar ambiguity exists in the relationships between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*, with the methods diverging on whether these stressors influence depression or are linked through latent variables.

Graph (b), derived solely from the nonparametric RCoT test. Both graphs consistently identify that employment precarity is not a cause of housing precarity or depression, and that social precarity does not cause depression or financial stress. 
The relationship between stressors and depression remains unresolved between the two algorithms. However, RCoT provides greater confidence that depression is not the cause of financial stress and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.

Overall, both graphs together highlight the potential roles of recent stressors as being closely linked with depression, either as causes or through latent confounders. Employment and social precarity may also be influenced by depression, either directly or through latent variables. Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, either directly or through a latent confounder.

```{r}
#| layout-ncol: 2
#| label: fig-sum
#| fig-cap: "Resulting graphs of precarity factors and depression sum score using FCI and CCI"
#| fig-subcap: 
#|   - "Using both GaussianCI and RCoT"
#|   - "Using only RCoT"
#| fig-align: center
#| out-width: 100%

knitr::include_graphics("img/both_dot.png")
knitr::include_graphics("img/rcot_dot.png")

```

<!-- \begin{figure}[htbp] -->
<!--     \centering -->
<!--         \caption{Summary of all sum score graphs} -->
<!--         \includegraphics[width=1\textwidth]{img/sum_all.png} -->
<!-- \end{figure} -->
<!-- \begin{figure}[htbp] -->
<!--     \centering -->
<!--         \caption{Summary of sum score graphs given RCoT} -->
<!--         \includegraphics[width=1\textwidth]{img/sum_rcotonly.png} -->
<!-- \end{figure} -->

## Individual depression symptom
Moving from the sum score representation to the symptom-level graph provides a more granular perspective on the causal relationships between precarity factors and depression. This approach highlights the heterogeneity in how precarity factors influence individual depressive symptoms — *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood). While the sum score graph aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom graph uncovers distinct pathways for different symptoms.
In the symptom-level graph (@fig-sym), consistent relationships are represented by black solid edges, while areas of disagreement between the FCI and CCI algorithms are denoted by gray dashed edges. Endpoints marked with circles indicate differences in directional conclusions between the algorithms. Additionally, the navy dashed edges represent relationships unique to the graphs generated using Gaussian CI testing.

<!-- Overall Patterns in the Symptom Network -->
The symptom-level graph reveals a complex and interconnected structure, far more intricate than the sum score graph. Its denser network highlights the strong interdependence among symptoms and suggests the presence of latent confounding influences, as indicated by numerous bidirectional arrows.

<!-- Centrality and Connectivity -->
Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points. Among the depressive symptoms, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network. Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precarity factors. These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes.
<!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. -->

The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs. As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precarity (*P.soc*) are more likely to be influenced by depressive symptoms. The symptom-level analysis, however, provides more specificity by pinpointing the symptoms involved in these relationships. For instance, *slp* is identified as influencing *P.emp* or potentially through a latent confounder (*slp* o-> *P.emp*) , while *glt* appears to affect *P.soc* or may be linked via a confounder (*glt* o-> *P.soc*).
Additional edges are observed in the graph generated using Gaussian CI tests, such as *slp* <-> *S.rel* and *slp* o-> *P.emp*. As seen in the sum graph, Gaussian CI testing tends to produce a denser graph. In this case, causal relationships involving *slp* are particularly prominent.

Some discrepancies, however, exist between the symptom-level and sum graphs. For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precarity factors. Additionally, housing precarity (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.

<!-- However, due to computational constraints that required limiting the sample size, as well as the challenging distributions of certain symptoms, the statistical power is lower compared to the depression sum score graph. -->
```{r}
#| label: fig-sym
#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using FCI and CCI"
#| fig-align: center
#| out-width: 80%

knitr::include_graphics("img/symptom_graph_refined.png")

```

# Discussion

The present study explored the causal relationships between precarity factors and depression by employing both sum score and symptom-level analyses using causal discovery algorithms. The use of both PHQ sum scores and individual symptom scores provided a comprehensive understanding of how different aspects of precarity influence depression, revealing nuanced pathways that may be obscured when considering aggregate measures alone.

<!-- Interpretation of Findings -->
Our findings highlight the significant role of recent relational stress (*S.rel*) as a potential causal factor for depression, consistently observed across both sum score and symptom-level analyses. This consistency underscores the profound impact interpersonal relationships can have on mental health. The symptom-level analysis further identified specific symptoms, such as sleep disturbances (*slp*), guilt (*glt*), and anhedonia (*anh*), as particularly sensitive to external precarity conditions. These symptoms emerged as potential initiators or activators within the depressive symptom network, suggesting that they may serve as early warning signs or valuable targets for preventive interventions.

Employment precarity (*P.emp*) and social precarity (*P.soc*) were more likely to be influenced by depressive symptoms rather than acting as direct causes. The symptom-level graph provided additional clarity by identifying that sleep disturbances influenced employment precarity (*slp* o-> *P.emp*) and feelings of guilt affected social precarity (*glt* o-> *P.soc*). This directional insight suggests that interventions targeting specific depressive symptoms may have downstream effects on improving certain aspects of precarity.

Interestingly, financial stress (*S.fin*) did not exhibit causal relationships with individual depressive symptoms in the symptom-level analysis, contrasting with its apparent role in the sum score graphs. This divergence indicates that while financial stress may influence overall depression severity, its impact on specific symptoms may not be significant.
Also, housing precarity (*P.hou*) emerged as an isolated node in the symptom-level graph, losing all connections with other precarity factors. This isolation suggests that housing precarity may operate independently of the depressive symptom network or that its effects are not captured within the scope of the measured variables.

<!-- Limitations -->
Several limitations should be acknowledged when interpreting these findings. The prevalence of bidirectional arrows and circle-marked endpoints in the graphs reflects unresolved ambiguities in the causal relationships suggested by the data. These uncertainties underscore the need for further research, ideally incorporating datasets with a higher proportion of symptomatic individuals. The HELIUS dataset, being predominantly composed of asymptomatic samples, posed challenges in identifying clear causal directions, particularly among symptom nodes.
Future studies could address these limitations by incorporating time-series data to leverage temporal information about the relationships between precarity factors and depressive symptoms. Time-series data could provide time-specific insights and track how these relationships evolve over time. For instance, methods such as *PCMCI* [@runge2019detecting] and *tsFCI* [@entner2010causal], along with other time-series adaptations of causal discovery algorithms, could help better account for temporal dependencies and refine the analysis.

Another limitation is the lack of clear evidence for cycles within the symptom network, despite using algorithms designed to account for cyclic relationships. The CCI algorithm predominantly produced bidirectional arrows, while FCI primarily generated directional arrows, yet neither displayed patterns indicative of definitive cyclic structures. Addressing cyclic relationships is particularly challenging with observational datasets alone. Future research could benefit from refined datasets that include intervention data. Methods such as *LLC* [@hyttinen2012learning], *NODGAS-Flow* [@sethuraman2023nodags], and the recently developed *Bicycle* algorithm [@rohbeck2024] are specifically designed to utilize both observational and intervention data to uncover potential cycles.
Additionally, if time-series data becomes available, corresponding methods, as described above, could be applied to capture repetitive patterns in variable interactions that might suggest cyclic structures.

<!-- The numerous bidirectional arrows in the CCI outputs also imply the presence of latent confounders. Expanding the analysis to include additional variables, such as other mental health conditions and coping mechanisms, could provide a more comprehensive understanding of the factors influencing depressive symptoms. Additionally, integrating qualitative approaches could enrich these findings, offering contextual insights into the quantitative relationships identified. Understanding individuals' lived experiences of precarity and depression could inform the development of more tailored and effective interventions. -->

Lastly, regarding conditional independence (CI) testing, the differences between the graphs generated by Gaussian CI and RCoT underscore the methodological sensitivities in detecting causal relationships. Gaussian CI produced denser graphs, which is somewhat counterintuitive, as the Gaussian CI's strict linearity assumption would typically result in fewer detected relationships, not more.
A possible explanation for this discrepancy is that Gaussian CI's reliance on partial correlations may overestimate relationships when specific non-linear dependencies exist in the data. In contrast, RCoT, free from such assumptions, may better capture these patterns under such conditions. However, while RCoT is technically non-parametric, its performance can still be influenced by the distributional characteristics of the data. Specifically, the RBF kernel, optimized for continuous data with smooth transitions, may struggle to capture relationships in datasets with discrete or mixed distributions. In such cases, the distances between discrete points may fail to convey meaningful similarity information. As a result, RCoT might miss certain dependencies, particularly when variables in the dataset lack smooth continuity.
Future research could address these issues by exploring a broader range of CI testing approaches. One traditional approach to handle this is discretizing variables and use $G^2$ test, which may better capture dependencies in non-continuous datasets [@dojer2016learning; @neapolitan2004learning]. A more promising direction, however, lies in the development of hybrid kernels tailored for mixed datasets, effectively integrating both continuous and discrete variables into RCoT-like methods.
By systematically employing and comparing a wider variety of CI testing techniques, researchers could gain more robust insights and mitigate the limitations inherent in specific approaches. This broader exploration holds the potential to enhance the reliability of findings, particularly in datasets with complex and heterogeneous structures.

<!-- Implications -->
<!-- 1. still impressive -- symptom directions aligns much well with the other empirical studies. -->
<!-- 2. understood better which aspect of precarity is more important for depression and we can target depends on what level of intervention we are planning. -->

<!-- Implications for Research and Practice -->
Despite its limitations, this study marks a meaningful step toward understanding the mechanisms linking precarity factors and depression. By applying causal discovery methods, it moves beyond traditional association-based analyses, providing insights that can inform more precise and targeted interventions.
While the resulting graphs are preliminary and contain unresolved ambiguities, they offer a valuable starting point for leveraging causal discovery tools to investigate the causal interplay between depression and precarity factors. A promising next step would involve integrating these causal structures into computational models, such as the symptom dynamic model proposed by **our comp-model paper**. By simulating intervention effects, such models could provide more realistic insights into how targeted actions might influence symptom networks and precarity factors over time. For example, interventions focused on improving sleep hygiene or alleviating guilt could be evaluated for their cascading effects on employment and social relationships, offering actionable guidance for designing population-level mental health strategies.
<!-- Conclusion -->
As one of the early applications of causal discovery tools to the complex dynamics of depression and precarity factors, this study lays a foundation for future research. We hope it inspires further refinement of these methods and ultimately contribute to more effective solutions for alleviating depression and improving societal well-being.

# References

::: {#refs}
:::

# Appendix {#sec-appendix}

## Precariousness factors by Leonie

1.  EMPLOYMENT PRECARIOUSNESS

-   `H1_Arbeidsparticipatie`: Working status
-   `H1_WerkSit`: Which work situation most applies to you?
-   `H1_RecentErv8`: Experiences past 12 months: h. You were sacked from your job or became unemployed (*reverse*)

2.  FINANCIAL PRECARIOUSNESS

-   `H1_InkHhMoeite`: During the past year, did you have problems managing your household income?
-   `H1_RecentErv9`: Experiences past 12 months: i. You had a major financial crisis (*reverse*)

3.  HOUSING PRECARIOUSNESS

-   `veilig_2012`: Score safety (veiligheid) in 2012 (*reverse*)
-   `vrz_2012`: Score level of resources (niveau voorzieningen) in 2012 (*reverse*)
-   `P_HUURWON`: Percentage Huurwoningen

4.  CULTURAL PRECARIOUSNESS

-   `H1_Discr_sumscore`: Perceived discrimination: sum score of 9 items (range 9-45)
-   `H1_SBSQ_meanscore`: Health literacy: SBSQ meanscore (range 1-5) (*reverse*)
-   `A_BED_RU`: Aantal bedrijfsvestigingen; cultuur, recreatie, overige diensten (*reverse*)

5.  SOCIAL PRECARIOUSNESS

-   `H1_RecentErv5`: Experiences past 12 months: e. Your steady relationship ended (*reverse*)
-   `H1_RecentErv6`: Experiences past 12 months: f. A long-term friendship with a good friend or family member was broken off (*reverse*)
-   `H1_RecentErv7`: Experiences past 12 months: g. You had a serious problem with a good friend or family member, or neighbour (*reverse*)
-   `H1_SSQT`: SSQT (frequency of social contact): sum score of 5 items (range 5-20) (*reverse*)
-   `H1_SSQSa`: SSQS (adequacy of social contact): sum score of 5 items, category 3 and 4 not combined (range 5-20) (*reverse*)

```{r, echo=FALSE, fig.align='center', fig.width=15}
# check correlation
#cor(df) |> round(2)

# visualize corr
col <- colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))
corrplot(cor(scaled_data[,c(1:16, 26, 27)]), method="color", col=col(200),  
         type="upper", order="hclust", 
         addCoef.col = "black", # Add coefficient of correlation
         tl.col="black", number.cex=0.50, tl.srt=45,tl.cex=0.7, #Text label color and rotation
         # hide correlation coefficient on the principal diagonal
         diag=FALSE 
         )
```

-   **High** Correlations: `emp_stat` (employment status) and `work_sit` (work situation) have a strong positive correlation of 0.82. This suggests that individuals with higher employment status tend to have more secure or favorable work situations. `soc_freq` (social contact frequency) shows a strong positive correlation with `soc_adq` (social adequacy) at 0.61. This indicates that individuals with more frequent social contact also tend to have higher perceived adequacy of social interactions.

-   **Moderate** Correlations: `nb_safe` (neighborhood safety) and `nb_res` (resources) have a moderate positive correlation of 0.39, suggesting that areas with higher safety also have better resources. `hea_lit` (health literacy) has moderate correlations with `emp_stat` (0.26) and `work_sit` (0.25), which could mean that higher health literacy is associated with better employment situations. `frd_brk12` (friendship breakups) and `conf12` (conflicts) have a notable correlation of 0.43, indicating a relationship between having conflicts and friendship losses.

-   **Low to Moderate** Correlations in Financial Precariousness: `inc_dif` (income difficulties) has a moderate correlation with `fincri12` (financial crisis) at 0.49. This aligns with the expected relationship, where individuals who experience general income difficulties are more likely to report financial crises.

-   **Low** Correlations (0.1 - 0.2): Many variables, such as `discrim` (discrimination), `unemp12` (unemployment experience), and `rel_end12` (relationship end), have low correlations with other variables, suggesting relatively independent relationships in the context of this dataset.

## Exploratory Factor Analysis (EFA)

```{r, echo=FALSE, fig.align='center', fig.height=5, out.width="55%", message=FALSE, results='hide'}
# precarious factors 
df <- scaled_data[,1:16]

# check suggested number of factors
pl <- fa.parallel(df, fa = "fa", n.iter = 100, show.legend = FALSE)
# run EFA
efa_result <- fa(df, nfactors = 5, rotate = "promax") # oblique rotation
#print(efa_result)
```

```{r, echo=FALSE, fig.align='center', out.width="70%"}
# Create a data frame from the pattern matrix
loadings <- data.frame(
  Variable = c("emp_stat", "work_sit", "unemp12", "inc_dif", "fincri12", "nb_safe", "nb_res", "nb_rent",
               "discrim", "hea_lit", "cul_rec", "rel_end12", "frd_brk12", "conf12", "soc_freq", "soc_adq"),
  MR1 = c(0.98, 0.97, 0.17, 0.21, 0.11, -0.08, 0.02, -0.03, -0.07, 0.27, -0.03, -0.04, -0.09, -0.08, -0.02, -0.03),
  MR2 = c(-0.06, -0.08, -0.03, 0.09, 0.05, -0.08, -0.05, -0.06, 0.17, 0.08, -0.08, -0.04, -0.04, 0.00, 1.00, 0.77),
  MR3 = c(-0.01, -0.06, 0.26, 0.28, 0.47, -0.05, -0.04, -0.02, 0.20, -0.10, -0.05, 0.36, 0.63, 0.56, -0.14, -0.01),
  MR4 = c(-0.02, -0.03, -0.04, 0.06, 0.03, -0.02, 0.71, -0.09, 0.13, 0.04, 0.61, -0.03, -0.01, -0.02, -0.04, -0.12),
  MR5 = c(-0.14, -0.11, -0.03, 0.16, 0.09, 0.61, -0.21, 0.54, 0.10, 0.15, 0.04, -0.01, -0.06, -0.06, -0.10, -0.13)
)

# Convert to long format
loadings_long <- loadings %>%
  pivot_longer(cols = starts_with("MR"), names_to = "Factor", values_to = "Loading")

# Plot heatmap
ggplot(loadings_long, aes(x = Factor, y = Variable, fill = Loading)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0, limits = c(-1, 1)) +
  theme_minimal() +
  labs(title = "Factor Loadings Heatmap", x = "", y = "") 
```

### Factor Loadings (Pattern Matrix)

-   **MR1**: High loadings on `emp_stat` and `work_sit` suggest this factor captures *employment* precariousness.
-   **MR2**: Strong loadings on `soc_freq` and `soc_adq` indicate *social* precariousness.
-   **MR3**: Key items like `frd_brk12`, `conf12`, and `fincri12`, suggest recent *stressful events*.
-   **MR4**: High loadings on `nb_res` and `cul_rec` may reflect *community resources* precariousness.
-   **MR5**: Variables `nb_safe` and `nb_rent` with high loadings indicate *housing* precariousness.

### Variance Explained

The factors cumulatively explain **38%** of the variance, with MR1 being the most influential factor. Each factor contributes a smaller proportion to the total variance (MR1 at 12%, MR2 at 9%, etc.).

### Factor Intercorrelations

Factors are moderately correlated, especially between *MR1 and MR5*, and *MR2 and MR3*. This indicates that while distinct, these factors are related—reasonable in a complex socio-economic context.

### Model Fit Statistics

RMSEA (0.071) suggest an acceptable fit. Tucker Lewis Index (0.802) suggests moderate reliability for the model.

### Summary

The 5-factor model appears interpretable and captures distinct dimensions of precariousness: *employment, social, stressors, community resources, and housing precariousness*. Although the overall fit and explained variance could be stronger, these factors offer insights into the underlying structure of the data, highlighting key areas of precariousness.

## PCA

```{r, echo=FALSE, fig.align='center', out.width="70%"}
# run PCA
res.pca <- prcomp(df, scale = TRUE)
# print(res.pca)
#summary(res.pca)

# Visualize eigenvalues (scree plot)
fviz_eig(res.pca, addlabels = TRUE)
#          
# fviz_pca_var(res.pca,
#              col.var = "contrib", # Color by contributions to the PC
#              gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
#              repel = TRUE     # Avoid text overlapping
#              )
```

-   Component Retention: The scree plot shows a clear "elbow" after the first component. This steep drop suggests that most variance is explained by the first component. After Dimension 5, the percentage of explained variance decreases slightly more gradually, indicating diminishing returns for adding more components. If we need to choose multiple components, retaining the first 5 components seems reasonable, as they capture most of the variance (cumulatively explaining about 54.7% of the total variance).

```{r, echo=FALSE, fig.align='center', fig.height=4.5, fig.width = 8}
# Results for Variables
res.var <- get_pca_var(res.pca)
# res.var$coord |> round(2)         # Coordinates
# res.var$contrib |> round(2)        # Contributions to the PCs
# res.var$cos2           # Quality of representation

# corrplot(res.var$cos2, is.corr=FALSE)

res.var$contrib |> as.data.frame() |> dplyr::select(1:5) |>
  tibble::rownames_to_column(var = "var") |> # adds a column for rownames
    pivot_longer(cols = starts_with("Dim"), 
               names_to = "Dimension", 
               values_to = "Contribution") |>
# Plot with facet_wrap to show each dimension in a separate facet
ggplot(aes(x = reorder(var, -Contribution), y = Contribution, fill = Contribution)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  facet_wrap(~ Dimension, scales="free_y") +
  coord_flip() +
  labs(title = "Contribution of Variables to 5 Principal Components",
       x = "", y = "Contribution (%)") +
  scale_fill_gradient(low = "lightblue", high = "blue") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

### Explained variance (contributions) of variables

It shows the importance of variables within each component.

-   **Dim1**: High contributions are observed from `emp_stat`, `work_sit`, `inc_dif`, and `fincri12`, suggesting that this dimension captures aspects of *employment and financial* security.

-   **Dim2**: While `emp_stat` and `work_sit` overlap with Dim1, the strong contributions from `frd_brk12` and `rel_end12` indicate that this dimension captures a focus on *recent relationship stressors*.

-   **Dim3**: `cul_rec`, `nb_res` have the highest contributions, indicating this dimension likely represents *community and cultural* factors.

-   **Dim4**: `soc_freq` and `soc_adq` stand out in this dimension, suggesting an emphasis on *social* precariousness.

-   **Dim5**: `nb_safe` and `nb_rent` are the top contributors, pointing to *housing* security as key themes in this component.

### Cos² Values

Cos² (squared cosine) values, or the quality of representation, show how well each variable is represented by each dimension. where higher cos² values (closer to 1) indicate better representation of a variable by a component.

```{r, echo=FALSE, fig.align='center', out.width="70%"}
res.var$cos2 |> as.data.frame() |> dplyr::select(1:5) %>% 
  mutate(var = rownames(.)) |>
  pivot_longer(!var, names_to = "dim", values_to = "cos2") |>
  ggplot(aes(x = dim, y = var, fill = cos2)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue") +
  labs(title = "Cos2 of Variables on Each Component (1-5)", fill = "Cos2", x = "", y = "") +
  theme_minimal()
```

-   **Dim.1**: Variables `emp_stat`, `work_sit`, `inc_dif`, and `fincri12` show high cos² values, meaning that PC1 primarily captures variations in employment and financial difficulties. This component could represent *employment & finance* precariousness.

-   **Dim.2**: Variables `work_sit`, `emp_stat`, `frd_brk12`, and `conf12` are well-represented in this component, suggesting PC2 captures aspects of *recent relationship stressors*.

-   **Dim.3**: Variables `nb_res` and `cul_rec` load strongly on PC3. This may represent community or cultural resources, indicating that this component is associated with *neighborhood resources*.

-   **Dim.4**: This component has high cos² values for `nb_res`, `cul_rec`, `soc_freq`, and `soc_adq.` While `nb_res` and `cul_rec` are also prominent in PC3, PC4 uniquely captures nuanced differentiation in *social* precariousness.

-   **Dim.5**: `nb_safe` and `nb_rent` are well-represented by PC5. This component might capture *housing* precariousness.

## ICA

```{r, echo=FALSE,  fig.align='center', out.width="70%"}
# Load the library
library(fastICA)
# set the seed
set.seed(1)

# Set the number of components you want to extract
n_components <- 5

# Run ICA
ica_model <- fastICA(df,
  n.comp = n_components, alg.typ = "parallel",
  fun = "exp", method = "C", maxit = 5000
)
# glimpse(ica_model)

weight_matrix <- data.frame(t(ica_model$A))
names(weight_matrix) <- stringr::str_glue("IC{seq(weight_matrix)}")

# dist_matrix <- dist(as.matrix(weight_matrix))
# clust <- hclust(dist_matrix)
# plot(clust)

weight_matrix |>
  mutate(var = colnames(df)) |>
  pivot_longer(
    cols = starts_with("IC"),
    names_to = "IC", values_to = "loading"
  ) |>
  ggplot(aes(x = IC, y = var, fill = loading)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0, limits = c(-1, 1)) +
  labs(x="", y="") +
  theme_minimal() 
```

### Dominant Variables per Component:

For each Independent Component (IC), we can identify variables with *high absolute* values in each column. These values indicate that the IC captures a strong, independent signal associated with these variables.

-   **IC1**: `soc_freq` and `soc_adq` have strong negative loadings on this component, indicating that this component might represent *social* precariousness.

-   **IC2**: `frd_brk12`, `conf12`, `rel_end12`, `fincri12` and `unemp12` have the most substantial loadings on this component, all with negative signs. This might point to a *recent relational or social stressor* component.

-   **IC3**: `nb_res` and `cul_rec` show notable negative loadings, pointing to a focus on *community resource* precariousness.

-   **IC4**: High loadings for `nb_safe`, `nb_rent`, `nb_res`, `cul_rec`, and `discrim` suggest a theme of *housing and community-based* precariousness, reflecting both safety and social challenges within the neighborhood context.

-   **IC5**: `emp_stat` and `work_sit` both have strong negative loadings on this component, suggesting it captures *employment* precariousness.

## Hierarchical clustering

### Using Euclidean distance

-   Ward.D's method: Minimizes the variance within clusters, producing more compact and spherical clusters.
-   Single linkage: Groups clusters based on the minimum distance between points.
-   Complete linkage: Groups clusters based on the maximum distance between points.
-   Average linkage: Uses the average distance between all pairs of points in the two clusters.

```{r, warning=FALSE, message=FALSE, fig.height=7, echo=FALSE, fig.align='center'}
#| layout-ncol: 2
library(dendextend)  # For coloring the clusters
library(ggdendro)

# Compute the distance matrix between variables
distance_matrix <- dist(t(df))

# wardD
hc_ward <- hclust(distance_matrix, method = "ward.D")
# dend <- as.dendrogram(hc)
# dend_colored <- dendextend::color_branches(dend, k = 5)  # Specify the number of clusters
plot(hc_ward, main = "ward.D")
rect.hclust(hc_ward, k = 5, border = 2:6)

# Avg
hc_avg <- hclust(distance_matrix, method = "average") 
# dend <- as.dendrogram(hc)
# dend_colored <- dendextend::color_branches(dend, k = 5) 
plot(hc_avg, main = "average")
rect.hclust(hc_avg, k = 5, border = 2:6)

hc_comp <- hclust(distance_matrix, method = "complete")
# dend <- as.dendrogram(hc)
# dend_colored <- dendextend::color_branches(dend, k = 5)  
plot(hc_comp, main = "complete")
rect.hclust(hc_comp, k = 5, border = 2:6)

hc_sing <- hclust(distance_matrix, method = "single") 
# dend <- as.dendrogram(hc)
# dend_colored <- dendextend::color_branches(dend, k = 5)  # Specify the number of clusters
plot(hc_sing, main = "single")
rect.hclust(hc_sing, k = 5, border = 2:6)
```

#### Consistent Groupings (Across All or Most Methods)

-   `emp_stat` and `work_sit`: This pair consistently clusters together across all linkage methods, suggesting that they are closely related variables, likely capturing a similar aspect of the data (possibly employment status or employment-related information).

-   `nb_safe`, `nb_res`, and `nb_rent`: These variables are often grouped closely in several methods (especially Ward.D, average, and complete linkage). This suggests a similarity or common theme among them, potentially related to neighborhood or housing precariousness.

-   `soc_freq` and `soc_adq`: These two variables frequently cluster together, indicating they likely measure aspects of social frequency and adequacy in similar ways. They appear together in Ward.D, average, and complete linkage.

-   `frd_brk12` and `conf12`: These variables are often clustered closely (though they sometimes join with other variables like rel_end12), suggesting they may capture aspects of relationship or social conflict. This pair appears in close proximity, especially in average and Ward.D.

#### Inconsistent Groupings (Variability Across Methods)

-   `hea_lit`: This variable shows inconsistent clustering across methods. In Ward.D, it joins with `fincri12`, while in other methods, it’s often more isolated or grouped with variables that do not appear similar. This may suggest that `hea_lit` does not strongly correlate with other variables, or it has multidimensional aspects affecting its grouping across methods.

-   `discrim`: This variable also shows variable groupings. In Ward.D, it is grouped with `hea_lit`, while in other methods (e.g., complete and single linkage), it clusters differently, sometimes on its own. This variability may indicate that `discrim` has weaker associations with the main clusters in the data or overlaps partially with multiple clusters.

-   Social and Financial Variables (`inc_dif`, `fincri12`, `unemp12`): These variables appear together in some methods (e.g., Ward.D clusters `fincri12` and `inc_dif`), but in others, they are spread out. This inconsistency suggests that social and financial variables may not have strong or consistent ties across different methods, perhaps due to capturing different aspects of precariousness.

#### Summary

The consistent clusters are likely capturing distinct thematic dimensions of the data (e.g., employment, housing, social contact), while the inconsistent variables may reflect multifaceted or weakly correlated attributes that do not fit neatly into one cluster.

### Using Mutual Information

Using mutual information (MI) as a basis for hierarchical clustering differs from using traditional distance measures (like Euclidean distance) in a few key ways.

```{r echo=FALSE, fig.align='center', fig.height=6, out.width="60%"}
library(infotheo)  # for mutual information functions

# Calculate pairwise mutual information
mi_matrix <- matrix(NA, ncol = ncol(df), nrow = ncol(df))
colnames(mi_matrix) <- rownames(mi_matrix) <- colnames(df)

for (i in 1:ncol(df)) {
  for (j in i:ncol(df)) {
    # Calculate mutual information between pairs of variables
    mi_matrix[i, j] <- mutinformation(discretize(df[, i]), discretize(df[, j]))
    mi_matrix[j, i] <- mi_matrix[i, j]  # symmetric
  }
}

# Normalize mutual information matrix
max_mi <- max(mi_matrix, na.rm = TRUE)
mi_distance <- 1 - (mi_matrix / max_mi)

# Convert to distance object
dist_matrix_mi <- as.dist(mi_distance)

# Hierarchical clustering on MI-based distance
hc_mi <- hclust(dist_matrix_mi, method = "ward.D2")

# Plot the dendrogram
plot(hc_mi, main = "Hierarchical Clustering (Mutual Information Distance)", xlab = "", sub = "")
rect.hclust(hc_mi, k = 5, border = 2:6)

```

#### Comparison to Euclidean Distance Clustering

-   **Housing and Community** Cluster: The variables `nb_safe`, `nb_res`, `nb_rent`, and `cul_rec` cluster together, indicating a strong association among housing-related and community-based factors. This suggests a shared theme of housing or community precariousness. This grouping is also observed in the Euclidean-based clustering, but it appears more tightly connected here, potentially due to the non-linear relationships highlighted by mutual information.

-   **Employment and Social Support** Cluster: `emp_stat` and `work_sit` form a cluster, linking employment status and work situation together as they did in Euclidean-based clustering. These remain closely associated regardless of the distance metric used. `soc_freq` and `soc_adq`, related to social contact frequency and adequacy, cluster nearby, indicating they have a stronger non-linear relationship with employment variables. This is a subtle difference as Euclidean distance might not capture this association as effectively.

-   **Financial Stressor** Cluster: `inc_dif` and `fincri12`, representing income difficulties and recent financial crises, consistently cluster together in both approaches, showing a strong association, likely linear. However, mutual information-based clustering links these financial stressors with social support variables, suggesting that financial challenges may have complex dependencies with social support in this dataset.

-   **Relational Stressor** Cluster: `frd_brk12`, `conf12`, `discrim`, `hea_lit`, `unemp12`, and `rel_end12` form a *looser* cluster focused on social and relational stressors (e.g., friendship breakup, conflicts, and discrimination). Compared to Euclidean clustering, `discrim` and `hea_lit` (health literacy) appear closer to relational stressors here, indicating that non-linear relationships might play a larger role in linking these variables.

#### Summary

In conclusion, mutual information-based clustering provides an alternative perspective that can reveal more intricate associations between variables, especially for those with non-linear relationships. Compared to Euclidean clustering, it shows a similar high-level structure but emphasizes nuanced connections between variables, particularly around social support, employment, and financial stress.

## Conclusions on Precariousness factors

Based on the consistent findings across multiple analyses, we decided to exclude the variables `discrim`, `hea_lit`, `umemp12`, and `rel_end12`, as they do not clearly belong to any specific precariousness factor nor exhibit strong associations with depression (see the correlation table above). Therefore, we propose retaining the following key precariousness factors:

-   Employment Precariousness: `emp_stat`, `work_sit`
-   Social Precariousness: `soc_freq`, `soc_adq`
-   Housing Precariousness: `nb_safe`, `nb_res`, `nb_rent`, `cul_rec`
-   Recent Relational Stressors: `frd_brk12`, `conf12`
-   Recent Financial Stressors: `fincri12`, `inc_diff`

We construct each precariousness factor by calculating the mean value of the combined variables. Below, we present the updated correlation table for the newly composed factors, along with the corresponding distributions of all variables to be used in the causal discovery analysis.

```{r, echo=FALSE, fig.align='center'}
clust_data <- scaled_data |> 
  mutate(
    P.emp = rowMeans(cbind(emp_stat, work_sit)),
    P.soc = rowMeans(cbind(soc_freq, soc_adq)),
    P.hou = rowMeans(cbind(nb_safe, nb_res, nb_rent, cul_rec)),
    S.rel = rowMeans(cbind(frd_brk12, conf12)),
    S.fin = rowMeans(cbind(fincri12, inc_dif))
  ) |>
  dplyr::select(17:33) |>
  as.data.frame()

# skimr::skim(clust_data)

# visualize corr
corrplot(cor(clust_data), method="color", col=col(200),  
         type="upper", order="hclust", 
         addCoef.col = "black", # Add coefficient of correlation
         tl.col="black", number.cex=0.50, tl.srt=45,tl.cex=0.7, 
         diag=FALSE 
         )

```

```{r echo=FALSE, fig.align='center', fig.height=4.5, fig.width=8}
# Reshape the data to long format
clust_data_long <- clust_data %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")

# Plot distributions with density lines
ggplot(clust_data_long, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "white", bins = 15) +  # Use density for histogram
  geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
  facet_wrap(~ variable, scales = "free") +
  labs(title = "Distribution of All Variables with Density Overlay", x = "", y = "") +
  theme_minimal()
```

## Results from PC algorithm

```{r}
#| layout-ncol: 2
#| label: fig-pc_sum
#| fig-cap: "Resulting graphs of precarity factors and depression sum score using PC"
#| fig-subcap: 
#|   - "Using both GaussianCI and RCoT"
#|   - "Using only RCoT"
#| fig-align: center
#| out-width: 100%

knitr::include_graphics("img/sum_PC_all.png")
knitr::include_graphics("img/sum_PC_RCoTonly.png")
```

```{r}
#| layout-nrow: 2
#| label: fig-pc_sym
#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC"
#| fig-subcap: 
#|   - "Using both GaussianCI and RCoT"
#|   - "Using only RCoT"
#| fig-align: center
#| out-width: 70%

knitr::include_graphics("img/pc_sym_GaussianCIRCoT.png")
knitr::include_graphics("img/pc_sym_onlyRCoT.png")
```

\clearpage

## Randomized Conditional Independence / Correlation Test (RCIT & RCoT) {#sec-rcot}

RCIT (Randomized Conditional Independence Test) and RCoT (Randomized conditional Correlation Test) are advanced methods for scalable conditional independence (CI) testing, offering computational efficiency while maintaining the accuracy of kernel-based approaches. These methods evaluate conditional independence between two variables $X$ and $Y$ given a third variable $Z$ while addressing computational challenges inherent in kernel-based CI tests.  In this section, we provide a high-level overview of RCIT and RCoT based on [@strobl2019approximate].

### Kernel-Based Conditional Independence Testing

Traditional kernel-based CI tests, such as the Kernel Conditional Independence Test (KCIT), compute dependencies using the Hilbert-Schmidt Independence Criterion (HSIC) in reproducing kernel Hilbert spaces (RKHS) [@zhang2012kernel]. KCIT uses the following hypothesis framework:
$$
H_0: X \perp\!\!\!\perp Y \,|\, Z, \quad H_1: X \not\!\perp\!\!\!\perp Y \,|\, Z.
$$
The core quantity in KCIT is the partial cross-covariance operator:
$$
\Sigma_{XY \cdot Z} = \Sigma_{XY} - \Sigma_{XZ} \Sigma_{ZZ}^{-1} \Sigma_{ZY},
$$
where $\Sigma_{XY}$ represents the cross-covariance operator between $X$ and $Y$, and $\Sigma_{XZ} \Sigma_{ZZ}^{-1} \Sigma_{ZY}$ removes the dependence mediated by $Z$.

The squared Hilbert-Schmidt (HS) norm of $\Sigma_{XY \cdot Z}$ serves as the test statistic:
$$
\|\Sigma_{XY \cdot Z}\|^2_{HS} = 0 \quad \text{if and only if} \quad X \perp\!\!\!\perp Y \,|\, Z.
$$

KCIT estimates residual dependencies using kernel ridge regression:
$$
f^*(z) = K_Z (K_Z + \lambda I)^{-1} f(x),
$$
where $K_Z$ is the kernel matrix for $Z$, $f(x)$ is the kernel feature map for $X$, and $\lambda$ is the ridge regularization parameter. The residual function for $X$ is:
$$
f_\text{res}(x) = f(x) - f^*(z) = R_Z f(x),
$$
with:
$$
R_Z = I - K_Z (K_Z + \lambda I)^{-1}.
$$
The kernel matrix for residualized $X$ is:
$$
K_{X \cdot Z} = R_Z K_X R_Z,
$$
and similarly for $Y$, $K_{Y \cdot Z} = R_Z K_Y R_Z$.

The test statistic is computed as:
$$
T_{XY \cdot Z} = \frac{1}{n^2} \text{tr}(K_{X \cdot Z} K_{Y \cdot Z}),
$$
which estimates the Hilbert-Schmidt (HS) norm of the partial cross-covariance operator. To ensure convergence, KCIT scales the statistic by $n$:
$$
S_K = n T_{XY \cdot Z}.
$$
The null hypothesis $H_0$ is rejected if $S_K$ exceeds a threshold determined by permutation or moment-matching-based null distribution [@lindsay2000moment].

### Random Fourier Features (RFFs)

Kernel-based methods like KCIT face scalability issues, as they involve operations on $n \times n$ kernel matrices, which scale quadratically with the sample size $n$. RCIT and RCoT overcome this bottleneck using *Random Fourier Features (RFFs)* to approximate kernel operations efficiently.

#### Bochner’s Theorem
Bochner’s theorem provides the foundation for RFFs, stating that any continuous shift-invariant kernel $k(x, y)$ can be expressed as:
$$
k(x, y) = \int_{\mathbb{R}^p} e^{i \omega^\top (x - y)} \, dP_\omega,
$$
where $P_\omega$ is the spectral distribution of the kernel. For the widely used RBF kernel:
$$
k(x, y) = \exp\left(-\frac{\|x - y\|^2}{2\sigma^2}\right),
$$
$P_\omega$ follows a Gaussian distribution: $\omega \sim \mathcal{N}(0, \sigma^2 I)$.

#### RFF Approximation
Using Monte Carlo sampling, the kernel function is approximated as:
$$
k(x, y) \approx \phi(x)^\top \phi(y),
$$
where $\phi(x)$ is the random Fourier feature mapping:
$$
\phi(x) = \sqrt{\frac{2}{D}} \cos(W^\top x + b),
$$
with $W \sim \mathcal{N}(0, \sigma^2 I)$ and $b \sim \text{Uniform}(0, 2\pi)$. Here, $D$ is the number of Fourier features, which balances computational efficiency and approximation accuracy.

### Differences Between RCIT and RCoT

RCIT and RCoT differ in their test statistics, computational efficiency, and practical performance, which makes them suited for different scenarios in causal discovery. RCIT evaluates the Hilbert-Schmidt norm of the full partial cross-covariance operator, providing a general test for conditional independence but at a higher computational cost. RCoT simplifies the process by using the Frobenius norm of a finite-dimensional residualized cross-covariance matrix, significantly reducing complexity and improving scalability.

These distinctions are particularly important for large-scale datasets, where RCoT's computational efficiency makes it a practical choice for high-dimensional causal discovery tasks.

#### RCIT: Randomized Conditional Independence Test

RCIT tests full conditional independence by examining the squared Hilbert-Schmidt (HS) norm of the partial cross-covariance operator $\Sigma_{XY \cdot Z}$:
$$
S_K = n T_{XY \cdot Z} = \frac{1}{n} \text{tr}(K_{X \cdot Z} K_{Y \cdot Z}),
$$
where $T_{XY \cdot Z}$ is an empirical estimate of $\|\Sigma_{XY \cdot Z}\|^2_{HS}$. The null and alternative hypotheses are:
$$
H_0: \|\Sigma_{XY \cdot Z}\|^2_{HS} = 0, \quad H_1: \|\Sigma_{XY \cdot Z}\|^2_{HS} > 0.
$$

RCIT is a general test for conditional independence but becomes computationally demanding as the size of $Z$ increases, due to the high-dimensional kernel operations required.

#### RCoT: Randomized Conditional Correlation Test

RCoT simplifies the testing process by using a finite-dimensional partial cross-covariance matrix, avoiding full HS norm calculations. Instead, it uses the Frobenius norm of the residualized cross-covariance matrix:
$$
S' = n \|C_{AB \cdot C}\|_F^2,
$$
where $C_{AB \cdot C}$ represents the residualized cross-covariance matrix. The hypotheses are:
$$
H_0: \|C_{AB \cdot C}\|_F^2 = 0, \quad H_1: \|C_{AB \cdot C}\|_F^2 > 0.
$$

RCoT is computationally efficient and well-suited for large conditioning sets ($|Z| \geq 4$). Its simplicity enables robust calibration of the null distribution and improved scalability for high-dimensional data.

<!-- ### Advantages of RCIT and RCoT -->

<!-- - **Scalability**: Linear complexity ($O(nD)$) makes these methods efficient for large datasets. -->
<!-- - **Robustness**: Provide accurate approximations of kernel-based CI tests while significantly reducing computational costs. -->
<!-- - **Applicability**: Suitable for large-scale causal discovery applications, particularly when integrated with algorithms like PC and FCI. -->

<!-- By combining the flexibility of kernel methods with the computational efficiency of RFFs, RCIT and RCoT represent powerful tools for non-parametric CI testing, enabling their use in high-dimensional and large-scale datasets. -->

## Summarized Stable Edges Proportion

```{r}
#| label: fig-sum-mat
#| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum scores"
#| fig-align: center
#| out-width: 100%
knitr::include_graphics("img/sumgraph_mat.pdf")
```

```{r}
#| label: fig-sym-mat
#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms"
#| fig-align: center
#| out-width: 100%
knitr::include_graphics("img/symgraph_mat.pdf")
```

### 2024-12-23 draft_v1/draft_v1.tex
draft

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage[]{palatino}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{amssymb}
\usepackage{booktabs} % Add to your preamble for cleaner table lines
\usepackage{makecell} % Add to your preamble for multi-line cells
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Causal Discovery on Precarity and Depression},
  pdfauthor={Kyuri Park; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Causal Discovery on Precarity and Depression}

\author[1]{Kyuri Park}
\author[2]{Leonie K. Elsenburg}
\author[2]{Mary Nicolao}
\author[2]{Karien Stronks}
\author[1, 3]{Vítor V. Vasconcelos}

\affil[1]{\textit{Computational Science Lab, Informatics Institute, University of Amsterdam, PO Box 94323, Amsterdam, 1090GH, the Netherlands}}
\affil[2]{\textit{Department of Public and Occupational Health, Amsterdam Public Health Research Institute, Amsterdam UMC, University of Amsterdam, Amsterdam, the Netherland}}
\affil[3]{\textit{Institute for Advanced Study, University of Amsterdam, Oude Turfmarkt 147, Amsterdam, 1012GC, the Netherland}}

\date{2024-12-23}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precarity factors
and depression is critical for developing effective interventions. This
study utilizes the HELIUS dataset to explore these relationships using
advanced causal discovery methods. By applying algorithms such as FCI,
and CCI, and combining traditional Gaussian CI tests with non-parametric
approaches like RCoT, we investigate how precarity factors---including
employment, social, financial, housing, and relational stress---affect
depression, both as a sum score and at the individual symptom level. Our
findings reveal that relational stress consistently emerges as a
potential causal factor for depression, while symptoms such as sleep
disturbances, guilt, and anhedonia are particularly sensitive to
external stressors, acting as potential early warning signals or
intervention points for prevention. Moreover, the results highlight
complexities in the data, including the influence of latent confounders
and the challenges of capturing cyclic relationships. Despite some
limitations, such as unresolved ambiguities in causal directions and
challenges with mixed data distributions, this study demonstrates the
utility of causal discovery tools in disentangling the intricate
interplay between social and mental health dynamics. By mapping these
causal structures into computational models, future research can
simulate intervention effects, providing actionable insights to mitigate
the impact of precarity on mental health. This study serves as a
foundational effort, offering both methodological advancements and
practical implications for addressing depression at a population level.
\end{abstract}

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}

\section{Introduction}\label{introduction}

Mental health problems in urban areas have been reported to be on the
rise. Governments have been attempting to intervene, but the complexity
of mental health systems presents significant challenges in planning
effective interventions, let alone understanding the underlying
mechanisms driving these issues.

Recent research has aimed to identify underlying factors contributing to
mental health problems, often referred to as precariousness factors.
These factors encompass various aspects of life, such as employment,
social connections, financial stability, housing, and cultural
dimensions. This comprehensive perspective on precarity helps to
highlight how different aspects of life may be interconnected. While
this research has advanced our understanding of the complex interplay
between precariousness factors, a key question remains unanswered: how
do these factors influence mental health? Specifically, the lack of
directional information --- knowing what influences what --- limits our
ability to identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between
precariousness factors and mental health outcomes, with a specific focus
on depression, the most prevalent mental health issue in urban
populations. Using causal discovery methods, we explore how different
aspects of precarity influence depression and delve deeper into the
dynamics at the symptom level. By examining individual depressive
symptoms, we aim to identify which symptoms may act as initiators by
being particularly sensitive to specific precariousness factors. Through
this analysis, our goal is to uncover the causal mechanisms underlying
mental health challenges and provide a foundation for developing more
effective and targeted interventions.

\section{Methods}\label{methods}

\subsection{Data}\label{data}

We use the HELIUS dataset, which captures the diverse population of
Amsterdam across various ethnicities and provides comprehensive health
and lifestyle data. To operationalize precariousness factors, we draw on
the framework outlined in previous research (i.e., Leonie's paper) and
select a set of relevant variables.

To ensure a robust representation of each precariousness factor, we
conducted various exploratory analyses to identify consistent and
meaningful factor structures. Based on these analyses, we identified
five precariousness factors, including two related to recent stressors,
each comprising multiple variables as outlined below. Detailed
information on the exploratory analyses can be found in the
\hyperref[sec-appendix]{Appendix}.

\begin{itemize}
\tightlist
\item
  Employment precariousness: \texttt{emp\_stat}, \texttt{work\_sit}.
\item
  Social precariousness: \texttt{soc\_freq}, \texttt{soc\_adq}.
\item
  Housing precariousness: \texttt{nb\_safe}, \texttt{nb\_res},
  \texttt{nb\_rent}, \texttt{cul\_rec}.
\item
  Recent relational stressors: \texttt{frd\_brk12}, \texttt{conf12}.
\item
  Recent financial stressors: \texttt{fincri12}, \texttt{inc\_diff}.
\end{itemize}

After preprocessing, the HELIUS dataset comprises 21,628 samples. In
addition to the five precariousness factors, we include PHQ-9 scores ---
both the total sum score and individual symptom scores --- to represent
depression. In the subsequent causal discovery analysis, we will examine
depression both as an aggregated sum score and through its individual
symptom-level representations. Refer to Figure~\ref{fig-dist} for the
overall distributions of the variables used in the analysis.

\begin{figure}

\centering{

\includegraphics[width=0.9\textwidth,height=\textheight]{draft_v1_files/figure-pdf/fig-dist-1.pdf}

}

\caption{\label{fig-dist}Distributions of variables with density
overlay. \emph{P.emp} = employment precariousness; \emph{P.hou} =
housing precariousness; \emph{P.soc} = social precariousness;
\emph{S.fin} = recent financial stressors; \emph{S.rel} = recent
relational stressors; \emph{PHQsum} = PHQ-9 sum score; \emph{anh} =
anhedonia; \emph{app} = appetite; \emph{con} = concentration; \emph{dep}
= depressed mood; \emph{ene} = energy; \emph{glt} = guilty; \emph{mot} =
motor; \emph{sui} = suicidal}

\end{figure}%

\subsection{Causal Discovery}\label{causal-discovery}

There are numerous causal discovery algorithms available; however, in
this study, we focus on algorithms suited to the potential cyclic
relationships within our system. Specifically, we use FCI (Fast Causal
Inference) and CCI (Cyclic Causal Inference), both capable of accounting
for such cycles under certain conditions (Mooij \& Claassen, 2020;
Strobl, 2019). Additionally, we include the PC algorithm as a reference,
given its simplicity and prominence as one of the most widely known
causal discovery methods (Spirtes et al., 2001). For a more detailed
explanation of these algorithms, please refer to Park et al. (2024).

\begin{table}[ht]
\centering
\caption{Assumptions of causal discovery algorithms}
\begin{tabular}{lccccl}
\toprule
Algorithm & Acyclicity & Causal sufficiency & 
\makecell{Absence of \\ selection bias} & Output \\ 
\midrule
PC  & $\checkmark$ & $\checkmark$ & $\checkmark$ & CPDAG \\ 
FCI & $-^{a}$ & $\checkmark$ & $-^{a}$ & PAG \\ 
CCI & x & x & x & MAAG \\ 
\bottomrule
\end{tabular}
\caption*{\small{\textit{Note}. $^{a}$FCI can detect cycles in systems that lack selection bias and exhibit non-linear relationships (Mooij \& Claassen, 2020).}}
\end{table}

As shown in Table 1, the resulting graphs from FCI and CCI differ
slightly (\emph{PAG}: partial ancestral graph; \emph{MAAG}: maximal
almost ancestral graph) due to their reliance on different underlying
assumptions. Both encode information about causal relationships between
variables, where the presence of an edge indicates causal
\emph{ancestry}. Directed edges, \(A\) *→ \(B\), specify that \(B\) is
not an ancestor of \(A\) in every graph within the Markov equivalence
class, \(Equiv(G)\). \(A\)*---\(B\) represent cases where \(B\) is an
ancestor of \(A\) in every graph in \(Equiv(G)\). Circle endpoints,
\(A\)*-o\(B\), denote ambiguity in the ancestral relationship, meaning
that \(B\)'s ancestral status relative to \(A\) varies across graphs in
\(Equiv(G)\).

In contrast, the graph produced by the PC algorithm is a \emph{CPDAG}
(completed partially directed acyclic graph), where directed edges
(\(A\) → \(B\)) indicate that \(A\) is a direct cause (parent) of \(B\).
Unlike FCI and CCI, the CPDAG does not include circle symbols. Instead,
when the PC algorithm cannot determine the direction, it represents
uncertainty with bidirectional arrows. While PC serves as a useful
reference, its strict assumptions of acyclicity and the absence of
latent confounders limit its applicability in more complex settings.
Therefore, our primary focus remains on the results from FCI and CCI.
For completeness, all PC algorithm results are provided in the
\hyperref[sec-appendix]{Appendix}.

One practical challenge in applying these algorithms to the HELIUS
dataset is that the data does not follow a Gaussian distribution, and
the relationships among variables are unlikely to be strictly linear. To
account for this, we complement the commonly used Gaussian conditional
independence test (CI test), which relies on partial correlations, with
a non-parametric CI test based on kernel methods. However, kernel-based
non-parametric tests are computationally demanding, particularly with
large datasets like ours. To mitigate this issue, we employ Randomized
Conditional Independence Test (RCIT) and the Randomized conditional
Correlation Test (RCoT), which uses random Fourier features to
approximate the kernel methods, thereby significantly reducing the
computational cost (Strobl et al., 2019). For a more detailed
explanation of RCIT and RCoT, see Section~\ref{sec-rcot}.

\subsection{Analysis}\label{analysis}

We analyze the causal structure using two approaches: one with the PHQ
sum score representing overall depression severity, and another with
individual symptom scores. The PHQ sum score simplifies the analysis by
reducing dimensionality, which is computationally advantageous and
provides a broad, interpretable perspective on depression's relationship
with precarity factors. In contrast, individual symptom scores offer a
nuanced understanding, capturing the heterogeneous ways symptoms respond
to precarity factors. However, analyzing individual symptoms poses
methodological challenges due to their non-standard distributions and
the increased complexity introduced by the higher dimensionality. By
employing both approaches, we balance simplicity and granularity,
ensuring robustness in our findings: the PHQ sum score captures
overarching trends, while symptom-level analysis reveals detailed
dynamics, essential for tailoring specific interventions.

To evaluate the sensitivity of the results to the choice of alpha levels
and ensure consistency, we test two significance levels ---
\(\alpha = 0.01, 0.05\). To further ensure robustness, we employ
bootstrapping to generate 100 bootstrap samples For each sample, we
estimate causal graphs and retain only the edges and directions that
appear above predefined thresholds.

The analyses are conducted under the following conditions:

\begin{itemize}
\tightlist
\item
  Significance levels (\(\alpha\)): 0.01 and 0.05
\item
  Thresholds: 0.5, 0.6, 0.7, and 0.8
\item
  CI test: Gaussian CI test, RCoT
\item
  Algorithms: FCI, CCI, and PC
\end{itemize}

This setup yields 16 combinations (2 significance levels × 4 thresholds
× 2 CI tests), applied across three algorithms, with each combination
repeated for 100 bootstrap samples, resulting in a total of 1,600
resulting graphs. For analyses involving individual symptom variables,
we streamline the setup by using thresholds of 0.6 and 0.7, reducing the
number of bootstrap samples to 30. Additionally, we fix the skeleton of
edges among symptom variables based on the common structure estimated
across all three algorithms, thereby reducing the computational time
required to estimate skeletons. This fixed structure aligns with
commonly observed skeleton structures in the literature, ensuring
relevance to existing findings (\textbf{cite the comp\_model
paper}).\footnote{To reduce spurious edges in the symptom network, we
  fixed the skeleton of symptom interactions using a common structure
  derived from PC, FCI, and CCI algorithms with an alpha level of 0.001.
  This approach maintained sparsity in symptom interactions while
  allowing the free estimation of causal directions between symptoms and
  precarity factors, balancing computational efficiency with meaningful
  results.}

To summarize the results, we identify the most frequently occurring edge
endpoints across different experimental setups. This ensures that only
stable and consistent edges are retained, providing a clearer and more
reliable understanding of the relationships between precarity factors
and depression. Refer to Figure~\ref{fig-workflow} for the analysis
workflow.

\begin{figure}

\centering{

\includegraphics[width=0.8\textwidth,height=\textheight]{img/simsetup.png}

}

\caption{\label{fig-workflow}Analysis workflow applied across all three
algorithms.}

\end{figure}%

\section{Results}\label{results}

\subsection{Depression as sum score}\label{depression-as-sum-score}

The sum score graphs provide a high-level summary of how precarity
factors collectively influence overall depression severity, focusing on
aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
relationships between precarity factors (\emph{P.hou}, \emph{P.emp},
\emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the depression sum score
(\emph{PHQsum}) under two different setups: (a) using both Gaussian CI
test and RCoT, and (b) using RCoT alone. Black edges represent
consistent causal relationships identified by both FCI and CCI, while
gray edges denote inconsistent relationships that vary between the two
methods. The edge endpoints of inconsistent gray edges are marked with
circles. In graph (a), the key pathways suggest that employment
precarity (\emph{P.emp}) and social precarity (\emph{P.soc}) do not
cause depression (\emph{PHQsum}). Social precarity does not cause recent
relational stress (\emph{S.rel}) or financial stress (\emph{S.fin}), and
these stressors are likely related by a latent confounder. Additionally,
\emph{P.emp} and \emph{S.fin} are identified as non-causes of housing
precarity (\emph{P.hou}). Both FCI and CCI detect a dependency between
\emph{P.emp} and \emph{S.fin}, but they disagree on the direction of the
relationship, leaving it unclear whether \emph{S.fin} causes
\emph{P.emp} or if the connection is mediated by an unmeasured
confounder. A similar ambiguity exists in the relationships between
\emph{S.rel} and \emph{PHQsum} and between \emph{S.fin} and
\emph{PHQsum}, with the methods diverging on whether these stressors
influence depression or are linked through latent variables.

Graph (b), derived solely from the nonparametric RCoT test. Both graphs
consistently identify that employment precarity is not a cause of
housing precarity or depression, and that social precarity does not
cause depression or financial stress. The relationship between stressors
and depression remains unresolved between the two algorithms. However,
RCoT provides greater confidence that depression is not the cause of
financial stress and suggests a stronger likelihood that the financial
stress may contribute to depression or that their relationship is
mediated by a latent confounder. The role of relational stress has
become less pronounced, as the edge between \emph{P.soc} and
\emph{S.rel} is omitted, and the relationship between \emph{S.rel} and
\emph{S.fin} is now inconsistent between the algorithms.

Overall, both graphs together highlight the potential roles of recent
stressors as being closely linked with depression, either as causes or
through latent confounders. Employment and social precarity may also be
influenced by depression, either directly or through latent variables.
Lastly, housing precarity does not directly impact depression but
remains connected to employment precarity and financial stress, either
directly or through a latent confounder.

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{img/both_dot.png}

}

\subcaption{\label{fig-sum-1}Using both GaussianCI and RCoT}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{img/rcot_dot.png}

}

\subcaption{\label{fig-sum-2}Using only RCoT}

\end{minipage}%

\caption{\label{fig-sum}Resulting graphs of precarity factors and
depression sum score using FCI and CCI}

\end{figure}%

\subsection{Individual depression
symptom}\label{individual-depression-symptom}

Moving from the sum score representation to the symptom-level graph
provides a more granular perspective on the causal relationships between
precarity factors and depression. This approach highlights the
heterogeneity in how precarity factors influence individual depressive
symptoms --- \emph{slp} (sleep), \emph{ene} (energy), \emph{app}
(appetite), \emph{mot} (motor), \emph{sui} (suicidal), \emph{anh}
(anhedonia), \emph{glt} (guilt), and \emph{dep} (depressed mood). While
the sum score graph aggregates all symptoms into a single
measure---potentially obscuring nuanced relationships---the symptom
graph uncovers distinct pathways for different symptoms. In the
symptom-level graph (Figure~\ref{fig-sym}), consistent relationships are
represented by black solid edges, while areas of disagreement between
the FCI and CCI algorithms are denoted by gray dashed edges. Endpoints
marked with circles indicate differences in directional conclusions
between the algorithms. Additionally, the navy dashed edges represent
relationships unique to the graphs generated using Gaussian CI testing.

The symptom-level graph reveals a complex and interconnected structure,
far more intricate than the sum score graph. Its denser network
highlights the strong interdependence among symptoms and suggests the
presence of latent confounding influences, as indicated by numerous
bidirectional arrows.

Certain nodes in the network emerge as more \emph{causally} central,
highlighting their potential importance as intervention points. Among
the depressive symptoms, \emph{dep}, \emph{slp}, and \emph{glt} emerge
as particularly influential, with multiple outgoing edges
(o-\textgreater) to other symptoms, suggesting their roles as potential
key drivers within the symptom network. Especially, symptoms such as
\emph{glt}, \emph{slp}, and \emph{anh} are particularly critical,
potentially acting as initiator or activator nodes within the network
due to their apparent connections with precarity factors. These symptoms
appear to be especially sensitive to external stressors, potentially
manifesting early in response to such conditions and subsequently
activating other interconnected symptom nodes.

The causal structure involving precarity factors in the symptom-level
graph is largely consistent with the patterns observed in the sum score
graphs. As in the sum graphs, relational stress (\emph{S.rel}) emerges
as a potential causal factor for depression, while employment
(\emph{P.emp}) and social precarity (\emph{P.soc}) are more likely to be
influenced by depressive symptoms. The symptom-level analysis, however,
provides more specificity by pinpointing the symptoms involved in these
relationships. For instance, \emph{slp} is identified as influencing
\emph{P.emp} or potentially through a latent confounder (\emph{slp}
o-\textgreater{} \emph{P.emp}) , while \emph{glt} appears to affect
\emph{P.soc} or may be linked via a confounder (\emph{glt}
o-\textgreater{} \emph{P.soc}). Additional edges are observed in the
graph generated using Gaussian CI tests, such as \emph{slp}
\textless-\textgreater{} \emph{S.rel} and \emph{slp} o-\textgreater{}
\emph{P.emp}. As seen in the sum graph, Gaussian CI testing tends to
produce a denser graph. In this case, causal relationships involving
\emph{slp} are particularly prominent.

Some discrepancies, however, exist between the symptom-level and sum
graphs. For example, financial stress (\emph{S.fin}) does not have any
edges with depressive symptoms in the symptom-level graph, although it
maintains associations with other precarity factors. Additionally,
housing precarity (\emph{P.hou}) becomes entirely disconnected from the
rest of the network, appearing as an isolated node.

\begin{figure}

\centering{

\includegraphics[width=0.8\textwidth,height=\textheight]{img/symptom_graph_refined.png}

}

\caption{\label{fig-sym}Resulting graphs of precarity factors and
individual depression symptoms using FCI and CCI}

\end{figure}%

\section{Discussion}\label{discussion}

The present study explored the causal relationships between precarity
factors and depression by employing both sum score and symptom-level
analyses using causal discovery algorithms. The use of both PHQ sum
scores and individual symptom scores provided a comprehensive
understanding of how different aspects of precarity influence
depression, revealing nuanced pathways that may be obscured when
considering aggregate measures alone.

Our findings highlight the significant role of recent relational stress
(\emph{S.rel}) as a potential causal factor for depression, consistently
observed across both sum score and symptom-level analyses. This
consistency underscores the profound impact interpersonal relationships
can have on mental health. The symptom-level analysis further identified
specific symptoms, such as sleep disturbances (\emph{slp}), guilt
(\emph{glt}), and anhedonia (\emph{anh}), as particularly sensitive to
external precarity conditions. These symptoms emerged as potential
initiators or activators within the depressive symptom network,
suggesting that they may serve as early warning signs or valuable
targets for preventive interventions.

Employment precarity (\emph{P.emp}) and social precarity (\emph{P.soc})
were more likely to be influenced by depressive symptoms rather than
acting as direct causes. The symptom-level graph provided additional
clarity by identifying that sleep disturbances influenced employment
precarity (\emph{slp} o-\textgreater{} \emph{P.emp}) and feelings of
guilt affected social precarity (\emph{glt} o-\textgreater{}
\emph{P.soc}). This directional insight suggests that interventions
targeting specific depressive symptoms may have downstream effects on
improving certain aspects of precarity.

Interestingly, financial stress (\emph{S.fin}) did not exhibit causal
relationships with individual depressive symptoms in the symptom-level
analysis, contrasting with its apparent role in the sum score graphs.
This divergence indicates that while financial stress may influence
overall depression severity, its impact on specific symptoms may not be
significant. Also, housing precarity (\emph{P.hou}) emerged as an
isolated node in the symptom-level graph, losing all connections with
other precarity factors. This isolation suggests that housing precarity
may operate independently of the depressive symptom network or that its
effects are not captured within the scope of the measured variables.

Several limitations should be acknowledged when interpreting these
findings. The prevalence of bidirectional arrows and circle-marked
endpoints in the graphs reflects unresolved ambiguities in the causal
relationships suggested by the data. These uncertainties underscore the
need for further research, ideally incorporating datasets with a higher
proportion of symptomatic individuals. The HELIUS dataset, being
predominantly composed of asymptomatic samples, posed challenges in
identifying clear causal directions, particularly among symptom nodes.
Future studies could address these limitations by incorporating
time-series data to leverage temporal information about the
relationships between precarity factors and depressive symptoms.
Time-series data could provide time-specific insights and track how
these relationships evolve over time. For instance, methods such as
\emph{PCMCI} (Runge et al., 2019) and \emph{tsFCI} (Entner \& Hoyer,
2010), along with other time-series adaptations of causal discovery
algorithms, could help better account for temporal dependencies and
refine the analysis.

Another limitation is the lack of clear evidence for cycles within the
symptom network, despite using algorithms designed to account for cyclic
relationships. The CCI algorithm predominantly produced bidirectional
arrows, while FCI primarily generated directional arrows, yet neither
displayed patterns indicative of definitive cyclic structures.
Addressing cyclic relationships is particularly challenging with
observational datasets alone. Future research could benefit from refined
datasets that include intervention data. Methods such as \emph{LLC}
(Hyttinen et al., 2012), \emph{NODGAS-Flow} (Sethuraman et al., 2023),
and the recently developed \emph{Bicycle} algorithm (Rohbeck et al.,
2024) are specifically designed to utilize both observational and
intervention data to uncover potential cycles. Additionally, if
time-series data becomes available, corresponding methods, as described
above, could be applied to capture repetitive patterns in variable
interactions that might suggest cyclic structures.

Lastly, regarding conditional independence (CI) testing, the differences
between the graphs generated by Gaussian CI and RCoT underscore the
methodological sensitivities in detecting causal relationships. Gaussian
CI produced denser graphs, which is somewhat counterintuitive, as the
Gaussian CI's strict linearity assumption would typically result in
fewer detected relationships, not more. A possible explanation for this
discrepancy is that Gaussian CI's reliance on partial correlations may
overestimate relationships when specific non-linear dependencies exist
in the data. In contrast, RCoT, free from such assumptions, may better
capture these patterns under such conditions. However, while RCoT is
technically non-parametric, its performance can still be influenced by
the distributional characteristics of the data. Specifically, the RBF
kernel, optimized for continuous data with smooth transitions, may
struggle to capture relationships in datasets with discrete or mixed
distributions. In such cases, the distances between discrete points may
fail to convey meaningful similarity information. As a result, RCoT
might miss certain dependencies, particularly when variables in the
dataset lack smooth continuity. Future research could address these
issues by exploring a broader range of CI testing approaches. One
traditional approach to handle this is discretizing variables and use
\(G^2\) test, which may better capture dependencies in non-continuous
datasets (Dojer, 2016; Neapolitan et al., 2004). A more promising
direction, however, lies in the development of hybrid kernels tailored
for mixed datasets, effectively integrating both continuous and discrete
variables into RCoT-like methods. By systematically employing and
comparing a wider variety of CI testing techniques, researchers could
gain more robust insights and mitigate the limitations inherent in
specific approaches. This broader exploration holds the potential to
enhance the reliability of findings, particularly in datasets with
complex and heterogeneous structures.

Despite its limitations, this study marks a meaningful step toward
understanding the mechanisms linking precarity factors and depression.
By applying causal discovery methods, it moves beyond traditional
association-based analyses, providing insights that can inform more
precise and targeted interventions. While the resulting graphs are
preliminary and contain unresolved ambiguities, they offer a valuable
starting point for leveraging causal discovery tools to investigate the
causal interplay between depression and precarity factors. A promising
next step would involve integrating these causal structures into
computational models, such as the symptom dynamic model proposed by
\textbf{our comp-model paper}. By simulating intervention effects, such
models could provide more realistic insights into how targeted actions
might influence symptom networks and precarity factors over time. For
example, interventions focused on improving sleep hygiene or alleviating
guilt could be evaluated for their cascading effects on employment and
social relationships, offering actionable guidance for designing
population-level mental health strategies. As one of the early
applications of causal discovery tools to the complex dynamics of
depression and precarity factors, this study lays a foundation for
future research. We hope it inspires further refinement of these methods
and ultimately contribute to more effective solutions for alleviating
depression and improving societal well-being.

\section{References}\label{references}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-dojer2016learning}
Dojer, N. (2016). Learning bayesian networks from datasets joining
continuous and discrete variables. \emph{International Journal of
Approximate Reasoning}, \emph{78}, 116--124.

\bibitem[\citeproctext]{ref-entner2010causal}
Entner, D., \& Hoyer, P. O. (2010). On causal discovery from time series
data using FCI. \emph{Probabilistic Graphical Models}, \emph{16}.

\bibitem[\citeproctext]{ref-hyttinen2012learning}
Hyttinen, A., Eberhardt, F., \& Hoyer, P. O. (2012). Learning linear
cyclic causal models with latent variables. \emph{The Journal of Machine
Learning Research}, \emph{13}(1), 3387--3439.

\bibitem[\citeproctext]{ref-lindsay2000moment}
Lindsay, B. G., Pilla, R. S., \& Basak, P. (2000). Moment-based
approximations of distributions using mixtures: Theory and applications.
\emph{Annals of the Institute of Statistical Mathematics}, \emph{52},
215--230.

\bibitem[\citeproctext]{ref-mooij2020constraint}
Mooij, J. M., \& Claassen, T. (2020). Constraint-based causal discovery
using partial ancestral graphs in the presence of cycles.
\emph{Conference on Uncertainty in Artificial Intelligence}, 1159--1168.

\bibitem[\citeproctext]{ref-neapolitan2004learning}
Neapolitan, R. E. et al. (2004). \emph{Learning bayesian networks} (Vol.
38). Pearson Prentice Hall Upper Saddle River.

\bibitem[\citeproctext]{ref-park2024discovering}
Park, K., Waldorp, L. J., \& Ryan, O. (2024). Discovering cyclic causal
models in psychological research. \emph{Advances. In/Psychology},
\emph{2}, e72425.

\bibitem[\citeproctext]{ref-rohbeck2024}
Rohbeck, M., Clarke, B., Mikulik, K., Pettet, A., Stegle, O., \&
Ueltzhöffer, K. (2024). Bicycle: Intervention-based causal discovery
with cycles. In F. Locatello \& V. Didelez (Eds.), \emph{Proceedings of
the third conference on causal learning and reasoning} (Vol. 236, pp.
209--242). PMLR.
\url{https://proceedings.mlr.press/v236/rohbeck24a.html}

\bibitem[\citeproctext]{ref-runge2019detecting}
Runge, J., Nowack, P., Kretschmer, M., Flaxman, S., \& Sejdinovic, D.
(2019). Detecting and quantifying causal associations in large nonlinear
time series datasets. \emph{Science Advances}, \emph{5}(11), eaau4996.

\bibitem[\citeproctext]{ref-sethuraman2023nodags}
Sethuraman, M. G., Lopez, R., Mohan, R., Fekri, F., Biancalani, T., \&
Hütter, J.-C. (2023). NODAGS-flow: Nonlinear cyclic causal structure
learning. \emph{International Conference on Artificial Intelligence and
Statistics}, 6371--6387.

\bibitem[\citeproctext]{ref-spirtes2001causation}
Spirtes, P., Glymour, C., \& Scheines, R. (2001). \emph{Causation,
prediction, and search}. MIT press.

\bibitem[\citeproctext]{ref-strobl2019}
Strobl, E. V. (2019). A constraint-based algorithm for causal discovery
with cycles, latent variables and selection bias. \emph{International
Journal of Data Science and Analytics}, \emph{8}(1), 33--56.
\url{https://doi.org/10.1007/s41060-018-0158-2}

\bibitem[\citeproctext]{ref-strobl2019approximate}
Strobl, E. V., Zhang, K., \& Visweswaran, S. (2019). Approximate
kernel-based conditional independence tests for fast non-parametric
causal discovery. \emph{Journal of Causal Inference}, \emph{7}(1),
20180017.

\bibitem[\citeproctext]{ref-zhang2012kernel}
Zhang, K., Peters, J., Janzing, D., \& Schölkopf, B. (2012).
Kernel-based conditional independence test and application in causal
discovery. \emph{arXiv Preprint arXiv:1202.3775}.

\end{CSLReferences}

\section{Appendix}\label{sec-appendix}

\subsection{Precariousness factors by
Leonie}\label{precariousness-factors-by-leonie}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  EMPLOYMENT PRECARIOUSNESS
\end{enumerate}

\begin{itemize}
\tightlist
\item
  \texttt{H1\_Arbeidsparticipatie}: Working status
\item
  \texttt{H1\_WerkSit}: Which work situation most applies to you?
\item
  \texttt{H1\_RecentErv8}: Experiences past 12 months: h. You were
  sacked from your job or became unemployed (\emph{reverse})
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  FINANCIAL PRECARIOUSNESS
\end{enumerate}

\begin{itemize}
\tightlist
\item
  \texttt{H1\_InkHhMoeite}: During the past year, did you have problems
  managing your household income?
\item
  \texttt{H1\_RecentErv9}: Experiences past 12 months: i. You had a
  major financial crisis (\emph{reverse})
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{2}
\tightlist
\item
  HOUSING PRECARIOUSNESS
\end{enumerate}

\begin{itemize}
\tightlist
\item
  \texttt{veilig\_2012}: Score safety (veiligheid) in 2012
  (\emph{reverse})
\item
  \texttt{vrz\_2012}: Score level of resources (niveau voorzieningen) in
  2012 (\emph{reverse})
\item
  \texttt{P\_HUURWON}: Percentage Huurwoningen
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{3}
\tightlist
\item
  CULTURAL PRECARIOUSNESS
\end{enumerate}

\begin{itemize}
\tightlist
\item
  \texttt{H1\_Discr\_sumscore}: Perceived discrimination: sum score of 9
  items (range 9-45)
\item
  \texttt{H1\_SBSQ\_meanscore}: Health literacy: SBSQ meanscore (range
  1-5) (\emph{reverse})
\item
  \texttt{A\_BED\_RU}: Aantal bedrijfsvestigingen; cultuur, recreatie,
  overige diensten (\emph{reverse})
\end{itemize}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{4}
\tightlist
\item
  SOCIAL PRECARIOUSNESS
\end{enumerate}

\begin{itemize}
\tightlist
\item
  \texttt{H1\_RecentErv5}: Experiences past 12 months: e. Your steady
  relationship ended (\emph{reverse})
\item
  \texttt{H1\_RecentErv6}: Experiences past 12 months: f.~A long-term
  friendship with a good friend or family member was broken off
  (\emph{reverse})
\item
  \texttt{H1\_RecentErv7}: Experiences past 12 months: g. You had a
  serious problem with a good friend or family member, or neighbour
  (\emph{reverse})
\item
  \texttt{H1\_SSQT}: SSQT (frequency of social contact): sum score of 5
  items (range 5-20) (\emph{reverse})
\item
  \texttt{H1\_SSQSa}: SSQS (adequacy of social contact): sum score of 5
  items, category 3 and 4 not combined (range 5-20) (\emph{reverse})
\end{itemize}

\begin{center}
\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-7-1.pdf}
\end{center}

\begin{itemize}
\item
  \textbf{High} Correlations: \texttt{emp\_stat} (employment status) and
  \texttt{work\_sit} (work situation) have a strong positive correlation
  of 0.82. This suggests that individuals with higher employment status
  tend to have more secure or favorable work situations.
  \texttt{soc\_freq} (social contact frequency) shows a strong positive
  correlation with \texttt{soc\_adq} (social adequacy) at 0.61. This
  indicates that individuals with more frequent social contact also tend
  to have higher perceived adequacy of social interactions.
\item
  \textbf{Moderate} Correlations: \texttt{nb\_safe} (neighborhood
  safety) and \texttt{nb\_res} (resources) have a moderate positive
  correlation of 0.39, suggesting that areas with higher safety also
  have better resources. \texttt{hea\_lit} (health literacy) has
  moderate correlations with \texttt{emp\_stat} (0.26) and
  \texttt{work\_sit} (0.25), which could mean that higher health
  literacy is associated with better employment situations.
  \texttt{frd\_brk12} (friendship breakups) and \texttt{conf12}
  (conflicts) have a notable correlation of 0.43, indicating a
  relationship between having conflicts and friendship losses.
\item
  \textbf{Low to Moderate} Correlations in Financial Precariousness:
  \texttt{inc\_dif} (income difficulties) has a moderate correlation
  with \texttt{fincri12} (financial crisis) at 0.49. This aligns with
  the expected relationship, where individuals who experience general
  income difficulties are more likely to report financial crises.
\item
  \textbf{Low} Correlations (0.1 - 0.2): Many variables, such as
  \texttt{discrim} (discrimination), \texttt{unemp12} (unemployment
  experience), and \texttt{rel\_end12} (relationship end), have low
  correlations with other variables, suggesting relatively independent
  relationships in the context of this dataset.
\end{itemize}

\subsection{Exploratory Factor Analysis
(EFA)}\label{exploratory-factor-analysis-efa}

\begin{center}
\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-8-1.pdf}
\end{center}

\begin{center}
\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-9-1.pdf}
\end{center}

\subsubsection{Factor Loadings (Pattern
Matrix)}\label{factor-loadings-pattern-matrix}

\begin{itemize}
\tightlist
\item
  \textbf{MR1}: High loadings on \texttt{emp\_stat} and
  \texttt{work\_sit} suggest this factor captures \emph{employment}
  precariousness.
\item
  \textbf{MR2}: Strong loadings on \texttt{soc\_freq} and
  \texttt{soc\_adq} indicate \emph{social} precariousness.
\item
  \textbf{MR3}: Key items like \texttt{frd\_brk12}, \texttt{conf12}, and
  \texttt{fincri12}, suggest recent \emph{stressful events}.
\item
  \textbf{MR4}: High loadings on \texttt{nb\_res} and \texttt{cul\_rec}
  may reflect \emph{community resources} precariousness.
\item
  \textbf{MR5}: Variables \texttt{nb\_safe} and \texttt{nb\_rent} with
  high loadings indicate \emph{housing} precariousness.
\end{itemize}

\subsubsection{Variance Explained}\label{variance-explained}

The factors cumulatively explain \textbf{38\%} of the variance, with MR1
being the most influential factor. Each factor contributes a smaller
proportion to the total variance (MR1 at 12\%, MR2 at 9\%, etc.).

\subsubsection{Factor Intercorrelations}\label{factor-intercorrelations}

Factors are moderately correlated, especially between \emph{MR1 and
MR5}, and \emph{MR2 and MR3}. This indicates that while distinct, these
factors are related---reasonable in a complex socio-economic context.

\subsubsection{Model Fit Statistics}\label{model-fit-statistics}

RMSEA (0.071) suggest an acceptable fit. Tucker Lewis Index (0.802)
suggests moderate reliability for the model.

\subsubsection{Summary}\label{summary}

The 5-factor model appears interpretable and captures distinct
dimensions of precariousness: \emph{employment, social, stressors,
community resources, and housing precariousness}. Although the overall
fit and explained variance could be stronger, these factors offer
insights into the underlying structure of the data, highlighting key
areas of precariousness.

\subsection{PCA}\label{pca}

\begin{center}
\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-10-1.pdf}
\end{center}

\begin{itemize}
\tightlist
\item
  Component Retention: The scree plot shows a clear ``elbow'' after the
  first component. This steep drop suggests that most variance is
  explained by the first component. After Dimension 5, the percentage of
  explained variance decreases slightly more gradually, indicating
  diminishing returns for adding more components. If we need to choose
  multiple components, retaining the first 5 components seems
  reasonable, as they capture most of the variance (cumulatively
  explaining about 54.7\% of the total variance).
\end{itemize}

\begin{center}
\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-11-1.pdf}
\end{center}

\subsubsection{Explained variance (contributions) of
variables}\label{explained-variance-contributions-of-variables}

It shows the importance of variables within each component.

\begin{itemize}
\item
  \textbf{Dim1}: High contributions are observed from
  \texttt{emp\_stat}, \texttt{work\_sit}, \texttt{inc\_dif}, and
  \texttt{fincri12}, suggesting that this dimension captures aspects of
  \emph{employment and financial} security.
\item
  \textbf{Dim2}: While \texttt{emp\_stat} and \texttt{work\_sit} overlap
  with Dim1, the strong contributions from \texttt{frd\_brk12} and
  \texttt{rel\_end12} indicate that this dimension captures a focus on
  \emph{recent relationship stressors}.
\item
  \textbf{Dim3}: \texttt{cul\_rec}, \texttt{nb\_res} have the highest
  contributions, indicating this dimension likely represents
  \emph{community and cultural} factors.
\item
  \textbf{Dim4}: \texttt{soc\_freq} and \texttt{soc\_adq} stand out in
  this dimension, suggesting an emphasis on \emph{social}
  precariousness.
\item
  \textbf{Dim5}: \texttt{nb\_safe} and \texttt{nb\_rent} are the top
  contributors, pointing to \emph{housing} security as key themes in
  this component.
\end{itemize}

\subsubsection{Cos² Values}\label{cosuxb2-values}

Cos² (squared cosine) values, or the quality of representation, show how
well each variable is represented by each dimension. where higher cos²
values (closer to 1) indicate better representation of a variable by a
component.

\begin{center}
\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-12-1.pdf}
\end{center}

\begin{itemize}
\item
  \textbf{Dim.1}: Variables \texttt{emp\_stat}, \texttt{work\_sit},
  \texttt{inc\_dif}, and \texttt{fincri12} show high cos² values,
  meaning that PC1 primarily captures variations in employment and
  financial difficulties. This component could represent
  \emph{employment \& finance} precariousness.
\item
  \textbf{Dim.2}: Variables \texttt{work\_sit}, \texttt{emp\_stat},
  \texttt{frd\_brk12}, and \texttt{conf12} are well-represented in this
  component, suggesting PC2 captures aspects of \emph{recent
  relationship stressors}.
\item
  \textbf{Dim.3}: Variables \texttt{nb\_res} and \texttt{cul\_rec} load
  strongly on PC3. This may represent community or cultural resources,
  indicating that this component is associated with \emph{neighborhood
  resources}.
\item
  \textbf{Dim.4}: This component has high cos² values for
  \texttt{nb\_res}, \texttt{cul\_rec}, \texttt{soc\_freq}, and
  \texttt{soc\_adq.} While \texttt{nb\_res} and \texttt{cul\_rec} are
  also prominent in PC3, PC4 uniquely captures nuanced differentiation
  in \emph{social} precariousness.
\item
  \textbf{Dim.5}: \texttt{nb\_safe} and \texttt{nb\_rent} are
  well-represented by PC5. This component might capture \emph{housing}
  precariousness.
\end{itemize}

\subsection{ICA}\label{ica}

\begin{center}
\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-13-1.pdf}
\end{center}

\subsubsection{Dominant Variables per
Component:}\label{dominant-variables-per-component}

For each Independent Component (IC), we can identify variables with
\emph{high absolute} values in each column. These values indicate that
the IC captures a strong, independent signal associated with these
variables.

\begin{itemize}
\item
  \textbf{IC1}: \texttt{soc\_freq} and \texttt{soc\_adq} have strong
  negative loadings on this component, indicating that this component
  might represent \emph{social} precariousness.
\item
  \textbf{IC2}: \texttt{frd\_brk12}, \texttt{conf12},
  \texttt{rel\_end12}, \texttt{fincri12} and \texttt{unemp12} have the
  most substantial loadings on this component, all with negative signs.
  This might point to a \emph{recent relational or social stressor}
  component.
\item
  \textbf{IC3}: \texttt{nb\_res} and \texttt{cul\_rec} show notable
  negative loadings, pointing to a focus on \emph{community resource}
  precariousness.
\item
  \textbf{IC4}: High loadings for \texttt{nb\_safe}, \texttt{nb\_rent},
  \texttt{nb\_res}, \texttt{cul\_rec}, and \texttt{discrim} suggest a
  theme of \emph{housing and community-based} precariousness, reflecting
  both safety and social challenges within the neighborhood context.
\item
  \textbf{IC5}: \texttt{emp\_stat} and \texttt{work\_sit} both have
  strong negative loadings on this component, suggesting it captures
  \emph{employment} precariousness.
\end{itemize}

\subsection{Hierarchical clustering}\label{hierarchical-clustering}

\subsubsection{Using Euclidean distance}\label{using-euclidean-distance}

\begin{itemize}
\tightlist
\item
  Ward.D's method: Minimizes the variance within clusters, producing
  more compact and spherical clusters.
\item
  Single linkage: Groups clusters based on the minimum distance between
  points.
\item
  Complete linkage: Groups clusters based on the maximum distance
  between points.
\item
  Average linkage: Uses the average distance between all pairs of points
  in the two clusters.
\end{itemize}

\begin{figure}

\begin{minipage}{0.50\linewidth}
\begin{center}
\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-14-1.pdf}
\end{center}
\end{minipage}%
%
\begin{minipage}{0.50\linewidth}
\begin{center}
\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-14-2.pdf}
\end{center}
\end{minipage}%
\newline
\begin{minipage}{0.50\linewidth}
\begin{center}
\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-14-3.pdf}
\end{center}
\end{minipage}%
%
\begin{minipage}{0.50\linewidth}
\begin{center}
\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-14-4.pdf}
\end{center}
\end{minipage}%

\end{figure}%

\paragraph{Consistent Groupings (Across All or Most
Methods)}\label{consistent-groupings-across-all-or-most-methods}

\begin{itemize}
\item
  \texttt{emp\_stat} and \texttt{work\_sit}: This pair consistently
  clusters together across all linkage methods, suggesting that they are
  closely related variables, likely capturing a similar aspect of the
  data (possibly employment status or employment-related information).
\item
  \texttt{nb\_safe}, \texttt{nb\_res}, and \texttt{nb\_rent}: These
  variables are often grouped closely in several methods (especially
  Ward.D, average, and complete linkage). This suggests a similarity or
  common theme among them, potentially related to neighborhood or
  housing precariousness.
\item
  \texttt{soc\_freq} and \texttt{soc\_adq}: These two variables
  frequently cluster together, indicating they likely measure aspects of
  social frequency and adequacy in similar ways. They appear together in
  Ward.D, average, and complete linkage.
\item
  \texttt{frd\_brk12} and \texttt{conf12}: These variables are often
  clustered closely (though they sometimes join with other variables
  like rel\_end12), suggesting they may capture aspects of relationship
  or social conflict. This pair appears in close proximity, especially
  in average and Ward.D.
\end{itemize}

\paragraph{Inconsistent Groupings (Variability Across
Methods)}\label{inconsistent-groupings-variability-across-methods}

\begin{itemize}
\item
  \texttt{hea\_lit}: This variable shows inconsistent clustering across
  methods. In Ward.D, it joins with \texttt{fincri12}, while in other
  methods, it's often more isolated or grouped with variables that do
  not appear similar. This may suggest that \texttt{hea\_lit} does not
  strongly correlate with other variables, or it has multidimensional
  aspects affecting its grouping across methods.
\item
  \texttt{discrim}: This variable also shows variable groupings. In
  Ward.D, it is grouped with \texttt{hea\_lit}, while in other methods
  (e.g., complete and single linkage), it clusters differently,
  sometimes on its own. This variability may indicate that
  \texttt{discrim} has weaker associations with the main clusters in the
  data or overlaps partially with multiple clusters.
\item
  Social and Financial Variables (\texttt{inc\_dif}, \texttt{fincri12},
  \texttt{unemp12}): These variables appear together in some methods
  (e.g., Ward.D clusters \texttt{fincri12} and \texttt{inc\_dif}), but
  in others, they are spread out. This inconsistency suggests that
  social and financial variables may not have strong or consistent ties
  across different methods, perhaps due to capturing different aspects
  of precariousness.
\end{itemize}

\paragraph{Summary}\label{summary-1}

The consistent clusters are likely capturing distinct thematic
dimensions of the data (e.g., employment, housing, social contact),
while the inconsistent variables may reflect multifaceted or weakly
correlated attributes that do not fit neatly into one cluster.

\subsubsection{Using Mutual Information}\label{using-mutual-information}

Using mutual information (MI) as a basis for hierarchical clustering
differs from using traditional distance measures (like Euclidean
distance) in a few key ways.

\begin{center}
\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-15-1.pdf}
\end{center}

\paragraph{Comparison to Euclidean Distance
Clustering}\label{comparison-to-euclidean-distance-clustering}

\begin{itemize}
\item
  \textbf{Housing and Community} Cluster: The variables
  \texttt{nb\_safe}, \texttt{nb\_res}, \texttt{nb\_rent}, and
  \texttt{cul\_rec} cluster together, indicating a strong association
  among housing-related and community-based factors. This suggests a
  shared theme of housing or community precariousness. This grouping is
  also observed in the Euclidean-based clustering, but it appears more
  tightly connected here, potentially due to the non-linear
  relationships highlighted by mutual information.
\item
  \textbf{Employment and Social Support} Cluster: \texttt{emp\_stat} and
  \texttt{work\_sit} form a cluster, linking employment status and work
  situation together as they did in Euclidean-based clustering. These
  remain closely associated regardless of the distance metric used.
  \texttt{soc\_freq} and \texttt{soc\_adq}, related to social contact
  frequency and adequacy, cluster nearby, indicating they have a
  stronger non-linear relationship with employment variables. This is a
  subtle difference as Euclidean distance might not capture this
  association as effectively.
\item
  \textbf{Financial Stressor} Cluster: \texttt{inc\_dif} and
  \texttt{fincri12}, representing income difficulties and recent
  financial crises, consistently cluster together in both approaches,
  showing a strong association, likely linear. However, mutual
  information-based clustering links these financial stressors with
  social support variables, suggesting that financial challenges may
  have complex dependencies with social support in this dataset.
\item
  \textbf{Relational Stressor} Cluster: \texttt{frd\_brk12},
  \texttt{conf12}, \texttt{discrim}, \texttt{hea\_lit},
  \texttt{unemp12}, and \texttt{rel\_end12} form a \emph{looser} cluster
  focused on social and relational stressors (e.g., friendship breakup,
  conflicts, and discrimination). Compared to Euclidean clustering,
  \texttt{discrim} and \texttt{hea\_lit} (health literacy) appear closer
  to relational stressors here, indicating that non-linear relationships
  might play a larger role in linking these variables.
\end{itemize}

\paragraph{Summary}\label{summary-2}

In conclusion, mutual information-based clustering provides an
alternative perspective that can reveal more intricate associations
between variables, especially for those with non-linear relationships.
Compared to Euclidean clustering, it shows a similar high-level
structure but emphasizes nuanced connections between variables,
particularly around social support, employment, and financial stress.

\subsection{Conclusions on Precariousness
factors}\label{conclusions-on-precariousness-factors}

Based on the consistent findings across multiple analyses, we decided to
exclude the variables \texttt{discrim}, \texttt{hea\_lit},
\texttt{umemp12}, and \texttt{rel\_end12}, as they do not clearly belong
to any specific precariousness factor nor exhibit strong associations
with depression (see the correlation table above). Therefore, we propose
retaining the following key precariousness factors:

\begin{itemize}
\tightlist
\item
  Employment Precariousness: \texttt{emp\_stat}, \texttt{work\_sit}
\item
  Social Precariousness: \texttt{soc\_freq}, \texttt{soc\_adq}
\item
  Housing Precariousness: \texttt{nb\_safe}, \texttt{nb\_res},
  \texttt{nb\_rent}, \texttt{cul\_rec}
\item
  Recent Relational Stressors: \texttt{frd\_brk12}, \texttt{conf12}
\item
  Recent Financial Stressors: \texttt{fincri12}, \texttt{inc\_diff}
\end{itemize}

We construct each precariousness factor by calculating the mean value of
the combined variables. Below, we present the updated correlation table
for the newly composed factors, along with the corresponding
distributions of all variables to be used in the causal discovery
analysis.

\begin{center}
\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-1.pdf}
\end{center}

\begin{center}
\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-17-1.pdf}
\end{center}

\subsection{Results from PC algorithm}\label{results-from-pc-algorithm}

\begin{figure}

\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{img/sum_PC_all.png}

}

\subcaption{\label{fig-pc_sum-1}Using both GaussianCI and RCoT}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{img/sum_PC_RCoTonly.png}

}

\subcaption{\label{fig-pc_sum-2}Using only RCoT}

\end{minipage}%

\caption{\label{fig-pc_sum}Resulting graphs of precarity factors and
depression sum score using PC}

\end{figure}%

\begin{figure}

\begin{minipage}{\linewidth}

\centering{

\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_sym_GaussianCIRCoT.png}

}

\subcaption{\label{fig-pc_sym-1}Using both GaussianCI and RCoT}

\end{minipage}%
\newline
\begin{minipage}{\linewidth}

\centering{

\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_sym_onlyRCoT.png}

}

\subcaption{\label{fig-pc_sym-2}Using only RCoT}

\end{minipage}%

\caption{\label{fig-pc_sym}Resulting graphs of precarity factors and
individual depression symptoms using PC}

\end{figure}%

\clearpage

\subsection{Randomized Conditional Independence / Correlation Test (RCIT
\& RCoT)}\label{sec-rcot}

RCIT (Randomized Conditional Independence Test) and RCoT (Randomized
conditional Correlation Test) are advanced methods for scalable
conditional independence (CI) testing, offering computational efficiency
while maintaining the accuracy of kernel-based approaches. These methods
evaluate conditional independence between two variables \(X\) and \(Y\)
given a third variable \(Z\) while addressing computational challenges
inherent in kernel-based CI tests. In this section, we provide a
high-level overview of RCIT and RCoT based on (Strobl et al., 2019).

\subsubsection{Kernel-Based Conditional Independence
Testing}\label{kernel-based-conditional-independence-testing}

Traditional kernel-based CI tests, such as the Kernel Conditional
Independence Test (KCIT), compute dependencies using the Hilbert-Schmidt
Independence Criterion (HSIC) in reproducing kernel Hilbert spaces
(RKHS) (Zhang et al., 2012). KCIT uses the following hypothesis
framework: \[
H_0: X \perp\!\!\!\perp Y \,|\, Z, \quad H_1: X \not\!\perp\!\!\!\perp Y \,|\, Z.
\] The core quantity in KCIT is the partial cross-covariance operator:
\[
\Sigma_{XY \cdot Z} = \Sigma_{XY} - \Sigma_{XZ} \Sigma_{ZZ}^{-1} \Sigma_{ZY},
\] where \(\Sigma_{XY}\) represents the cross-covariance operator
between \(X\) and \(Y\), and
\(\Sigma_{XZ} \Sigma_{ZZ}^{-1} \Sigma_{ZY}\) removes the dependence
mediated by \(Z\).

The squared Hilbert-Schmidt (HS) norm of \(\Sigma_{XY \cdot Z}\) serves
as the test statistic: \[
\|\Sigma_{XY \cdot Z}\|^2_{HS} = 0 \quad \text{if and only if} \quad X \perp\!\!\!\perp Y \,|\, Z.
\]

KCIT estimates residual dependencies using kernel ridge regression: \[
f^*(z) = K_Z (K_Z + \lambda I)^{-1} f(x),
\] where \(K_Z\) is the kernel matrix for \(Z\), \(f(x)\) is the kernel
feature map for \(X\), and \(\lambda\) is the ridge regularization
parameter. The residual function for \(X\) is: \[
f_\text{res}(x) = f(x) - f^*(z) = R_Z f(x),
\] with: \[
R_Z = I - K_Z (K_Z + \lambda I)^{-1}.
\] The kernel matrix for residualized \(X\) is: \[
K_{X \cdot Z} = R_Z K_X R_Z,
\] and similarly for \(Y\), \(K_{Y \cdot Z} = R_Z K_Y R_Z\).

The test statistic is computed as: \[
T_{XY \cdot Z} = \frac{1}{n^2} \text{tr}(K_{X \cdot Z} K_{Y \cdot Z}),
\] which estimates the Hilbert-Schmidt (HS) norm of the partial
cross-covariance operator. To ensure convergence, KCIT scales the
statistic by \(n\): \[
S_K = n T_{XY \cdot Z}.
\] The null hypothesis \(H_0\) is rejected if \(S_K\) exceeds a
threshold determined by permutation or moment-matching-based null
distribution (Lindsay et al., 2000).

\subsubsection{Random Fourier Features
(RFFs)}\label{random-fourier-features-rffs}

Kernel-based methods like KCIT face scalability issues, as they involve
operations on \(n \times n\) kernel matrices, which scale quadratically
with the sample size \(n\). RCIT and RCoT overcome this bottleneck using
\emph{Random Fourier Features (RFFs)} to approximate kernel operations
efficiently.

\paragraph{Bochner's Theorem}\label{bochners-theorem}

Bochner's theorem provides the foundation for RFFs, stating that any
continuous shift-invariant kernel \(k(x, y)\) can be expressed as: \[
k(x, y) = \int_{\mathbb{R}^p} e^{i \omega^\top (x - y)} \, dP_\omega,
\] where \(P_\omega\) is the spectral distribution of the kernel. For
the widely used RBF kernel: \[
k(x, y) = \exp\left(-\frac{\|x - y\|^2}{2\sigma^2}\right),
\] \(P_\omega\) follows a Gaussian distribution:
\(\omega \sim \mathcal{N}(0, \sigma^2 I)\).

\paragraph{RFF Approximation}\label{rff-approximation}

Using Monte Carlo sampling, the kernel function is approximated as: \[
k(x, y) \approx \phi(x)^\top \phi(y),
\] where \(\phi(x)\) is the random Fourier feature mapping: \[
\phi(x) = \sqrt{\frac{2}{D}} \cos(W^\top x + b),
\] with \(W \sim \mathcal{N}(0, \sigma^2 I)\) and
\(b \sim \text{Uniform}(0, 2\pi)\). Here, \(D\) is the number of Fourier
features, which balances computational efficiency and approximation
accuracy.

\subsubsection{Differences Between RCIT and
RCoT}\label{differences-between-rcit-and-rcot}

RCIT and RCoT differ in their test statistics, computational efficiency,
and practical performance, which makes them suited for different
scenarios in causal discovery. RCIT evaluates the Hilbert-Schmidt norm
of the full partial cross-covariance operator, providing a general test
for conditional independence but at a higher computational cost. RCoT
simplifies the process by using the Frobenius norm of a
finite-dimensional residualized cross-covariance matrix, significantly
reducing complexity and improving scalability.

These distinctions are particularly important for large-scale datasets,
where RCoT's computational efficiency makes it a practical choice for
high-dimensional causal discovery tasks.

\paragraph{RCIT: Randomized Conditional Independence
Test}\label{rcit-randomized-conditional-independence-test}

RCIT tests full conditional independence by examining the squared
Hilbert-Schmidt (HS) norm of the partial cross-covariance operator
\(\Sigma_{XY \cdot Z}\): \[
S_K = n T_{XY \cdot Z} = \frac{1}{n} \text{tr}(K_{X \cdot Z} K_{Y \cdot Z}),
\] where \(T_{XY \cdot Z}\) is an empirical estimate of
\(\|\Sigma_{XY \cdot Z}\|^2_{HS}\). The null and alternative hypotheses
are: \[
H_0: \|\Sigma_{XY \cdot Z}\|^2_{HS} = 0, \quad H_1: \|\Sigma_{XY \cdot Z}\|^2_{HS} > 0.
\]

RCIT is a general test for conditional independence but becomes
computationally demanding as the size of \(Z\) increases, due to the
high-dimensional kernel operations required.

\paragraph{RCoT: Randomized Conditional Correlation
Test}\label{rcot-randomized-conditional-correlation-test}

RCoT simplifies the testing process by using a finite-dimensional
partial cross-covariance matrix, avoiding full HS norm calculations.
Instead, it uses the Frobenius norm of the residualized cross-covariance
matrix: \[
S' = n \|C_{AB \cdot C}\|_F^2,
\] where \(C_{AB \cdot C}\) represents the residualized cross-covariance
matrix. The hypotheses are: \[
H_0: \|C_{AB \cdot C}\|_F^2 = 0, \quad H_1: \|C_{AB \cdot C}\|_F^2 > 0.
\]

RCoT is computationally efficient and well-suited for large conditioning
sets (\(|Z| \geq 4\)). Its simplicity enables robust calibration of the
null distribution and improved scalability for high-dimensional data.

\subsection{Summarized Stable Edges
Proportion}\label{summarized-stable-edges-proportion}

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{img/sumgraph_mat.pdf}

}

\caption{\label{fig-sum-mat}Proportions of edge endpoint types for
graphs based on depression sum scores}

\end{figure}%

\begin{figure}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{img/symgraph_mat.pdf}

}

\caption{\label{fig-sym-mat}Proportions of edge endpoint types for
graphs based on individual depression symptoms}

\end{figure}%

\end{document}

### 2024-12-23 code/edge_proportion_sum.R
update script description

@@ -1,3 +1,22 @@
+## =============================================================================
+## Script for Visualizing Proportions of Edge Endpoint Types in Depression Graphs
+## =============================================================================
+#' This script computes and visualizes the proportions of edge endpoint types 
+#' (null, circle, arrowhead, arrowtail) for causal discovery results (FCI and CCI)
+#' derived from sum score and individual symptom-level data.
+#'
+#' @details
+#' - Stable edges are extracted from `.rds` files containing causal discovery results.
+#' - Proportions of edge types are computed for each edge across multiple runs.
+#' - Heatmaps are generated to visualize the dominant edge types and their proportions.
+#' 
+#' @output
+#' - Heatmaps visualizing the proportions of edge endpoint types for sum score graphs.
+#' - Heatmaps visualizing the proportions of edge endpoint types for individual symptom graphs.
+#' - Figures exported as `.pdf` files.
+## =============================================================================
+
+
 # BiocManager::install("ComplexHeatmap")
 library(ComplexHeatmap)
 library(circlize)
@@ -268,6 +287,7 @@ heatmap_CCI_plot <- ggplotify::as.ggplot(grid.grabExpr(draw(heatmap_cci))) + ggt
 )
 
 legend <- ggplotify::as.ggplot(grid.grabExpr(draw(custom_legend)))
+legend_h <- ggplotify::as.ggplot(grid.grabExpr(draw(custom_legend_h))) # horizontal version 
 
 # Combine heatmaps and legend using ggarrange
 sumgraph_mat <- ggpubr::ggarrange(
@@ -280,4 +300,20 @@ sumgraph_mat <- ggpubr::ggarrange(
 print(sumgraph_mat)
 # save plot
 #ggpubr::ggexport(filename = "sumgraph_mat.pdf", plot = sumgraph_mat,  width = 11, height = 5, units = "cm")
+
+
+
+# legend horizontal version (for symptom graph)
+symgraph_mat_h <- ggpubr::ggarrange(
+  heatmap_FCI_plot,
+  heatmap_CCI_plot,
+  legend_h,
+  nrow = 3,  # Heatmaps side by side with legend
+  heights = c(3, 3, 0.1)  # Adjust the relative width of each panel
+)
+print(symgraph_mat_h)
+# save plot
+# ggpubr::ggexport(filename = "symgraph_mat.pdf", plot = symgraph_mat_h,  width = 9, height = 20, units = "cm")
+
+
 

### 2024-12-28 draft_v1/draft_v1.qmd
add_presum_graph

@@
 # Load the dataset 
-dat <- read_sav("../data/HELIUS_LEONIE.sav")
+dat <- read_sav("../../data/HELIUS_LEONIE.sav")
 
@@
 For analyses involving individual symptom variables, we streamline the setup by using thresholds of 0.6 and 0.7, reducing the number of bootstrap samples to 30.
-Additionally, we fix the skeleton of edges among symptom variables based on the common structure estimated across all three algorithms, thereby reducing the computational time required to estimate skeletons.
-This fixed structure aligns with commonly observed skeleton structures in the literature, ensuring relevance to existing findings (**cite the comp_model paper**).[^2]
-
-To summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.
-This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression.
+<!-- Additionally, we fix the skeleton of edges among symptom variables based on the common structure estimated across all three algorithms, thereby reducing the computational time required to estimate skeletons.
+This fixed structure aligns with commonly observed skeleton structures in the literature, ensuring relevance to existing findings (**cite the comp_model paper**).[^2]  -->
+
+To summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.[^3] This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression.
 Refer to @fig-workflow for the analysis workflow.
 
-[^2]: 
-To reduce spurious edges in the symptom network, we fixed the skeleton of symptom interactions using a common structure derived from PC, FCI, and CCI algorithms with an alpha level of 0.001.
+<!-- [^2]:  -->
+<!-- To reduce spurious edges in the symptom network, we fixed the skeleton of symptom interactions using a common structure derived from PC, FCI, and CCI algorithms with an alpha level of 0.001.
 This approach maintained sparsity in symptom interactions while allowing the free estimation of causal directions between symptoms and precarity factors, balancing computational efficiency with meaningful results.
+-->
+
+[^3]:
+The detailed proportion of each edge endpoint occurrence is shown in [@sec-propmatrix].
 
@@
 Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points.
-Among the depressive symptoms, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network.
+Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network.
 Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precarity factors.
@@
 
-## Summarized Stable Edges Proportion
+## Summarized Stable Edges Proportion {#sec-propmatrix}
 

### 2024-12-28 draft_v1/draft_v1.tex
add_presum_graph

@@
 we streamline the setup by using thresholds of 0.6 and 0.7, reducing the
-number of bootstrap samples to 30. Additionally, we fix the skeleton of
-edges among symptom variables based on the common structure estimated
-across all three algorithms, thereby reducing the computational time
-required to estimate skeletons. This fixed structure aligns with
-commonly observed skeleton structures in the literature, ensuring
-relevance to existing findings (\textbf{cite the comp\_model
-paper}).\footnote{To reduce spurious edges in the symptom network, we
-  fixed the skeleton of symptom interactions using a common structure
-  derived from PC, FCI, and CCI algorithms with an alpha level of 0.001.
-  This approach maintained sparsity in symptom interactions while
-  allowing the free estimation of causal directions between symptoms and
-  precarity factors, balancing computational efficiency with meaningful
-  results.}
+number of bootstrap samples to 30.
 
 To summarize the results, we identify the most frequently occurring edge
-endpoints across different experimental setups. This ensures that only
-stable and consistent edges are retained, providing a clearer and more
-reliable understanding of the relationships between precarity factors
-and depression. Refer to Figure~\ref{fig-workflow} for the analysis
+endpoints across different experimental setups.\footnote{The detailed
+  proportion of each edge endpoint occurrence is shown in
+  Section~\ref{sec-propmatrix}.} This ensures that only stable and
+consistent edges are retained, providing a clearer and more reliable
+understanding of the relationships between precarity factors and
+depression. Refer to Figure~\ref{fig-workflow} for the analysis
 workflow.
@@
 highlighting their potential importance as intervention points. Among
-the depressive symptoms, \emph{dep}, \emph{slp}, and \emph{glt} emerge
-as particularly influential, with multiple outgoing edges
-(o-\textgreater) to other symptoms, suggesting their roles as potential
-key drivers within the symptom network. Especially, symptoms such as
-\emph{glt}, \emph{slp}, and \emph{anh} are particularly critical,
-potentially acting as initiator or activator nodes within the network
-due to their apparent connections with precarity factors. These symptoms
-appear to be especially sensitive to external stressors, potentially
-manifesting early in response to such conditions and subsequently
-activating other interconnected symptom nodes.
+the depressive symptoms, \emph{anh}, \emph{dep}, \emph{slp}, and
+\emph{glt} emerge as particularly influential, with multiple outgoing
+edges (o-\textgreater) to other symptoms, suggesting their roles as
+potential key drivers within the symptom network. Especially, symptoms
+such as \emph{glt}, \emph{slp}, and \emph{anh} are particularly
+critical, potentially acting as initiator or activator nodes within the
+network due to their apparent connections with precarity factors. These
+symptoms appear to be especially sensitive to external stressors,
+potentially manifesting early in response to such conditions and
+subsequently activating other interconnected symptom nodes.
 
@@
 
-\subsection{Summarized Stable Edges
-Proportion}\label{summarized-stable-edges-proportion}
+\subsection{Summarized Stable Edges Proportion}\label{sec-propmatrix}
 

### 2024-12-28 code/graphviz/symptom_presum_graph.py

@@ -1,0 +1,75 @@
+from graphviz import Digraph
+
+# Define fixed positions for the nodes
+fixed_positions = {
+    "anh": (4, 3),
+    "dep": (2, 3.2),
+    "slp": (2, 1),
+    "ene": (3, 4),
+    "app": (1, 4),
+    "glt": (2.5, 2),
+    "con": (0, 2.7),
+    "mot": (0.8, 2),
+    "sui": (0, 1),
+    "Precarity": (3, 0)
+}
+
+# Create the directed graph
+dot = Digraph(format='png', engine='neato')
+dot.attr(size="14,15", dpi="300", splines="true", ratio="fill")
+dot.attr(label="", labelloc="top", fontsize="20", fontname="Palatino")
+
+# Add nodes with fixed positions
+for node, pos in fixed_positions.items():
+    dot.node(node, node, shape="circle", pos=f"{pos[0]},{pos[1]}!",
+             fixedsize="true", width="1", height="1", fontsize="25")
+
+
+# Manually define edges based on the adj.matrices
+dot.edge("anh", "dep", dir="both", arrowhead="odot", arrowtail="normal", style="dashed")
+dot.edge("anh", "slp", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
+dot.edge("anh", "ene", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
+dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="normal")
+dot.edge("anh", "con", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
+dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
+
+dot.edge("dep", "slp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
+dot.edge("dep", "app", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
+dot.edge("dep", "glt", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
+dot.edge("dep", "con", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
+dot.edge("dep", "mot", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
+dot.edge("dep", "sui", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
[…]

### 2024-12-28 code/post_analysis.R

@@ -17,7 +17,7 @@
 # all conditions
 algorithm <- c("PC", "FCI", "CCI")
 alpha <- c(0.01, 0.05)
-threshold <- c(0.5, 0.6, 0.7, 0.8)
+threshold <- c(0.5, 0.6, 0.7, 0.8) # c(0.6, 0.7) # for symptom data
 citest <- c("gaussCItest", "RCoT")
 comb <- expand.grid(algorithm = algorithm, alpha = alpha, threshold = threshold, citest = citest)
 # sum score data
@@ -30,7 +30,7 @@ filenames <- glue::glue(
 # )
 
 # directory containing the files
-directory <- "data/dep_sumscore"  
+directory <- "Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_sumscore"  
 
 # Full paths to the files
 full_paths <- file.path(directory, filenames)
@@ -46,6 +46,9 @@ stable_edges_data <- lapply(rds_data, function(x) x$stable_edges)
 
 # Group matrices by algorithm
 grouped_stable_edges <- split(stable_edges_data, comb$algorithm)
+
+# Group matrices by CItest
+grouped_stable_edges_ci <- split(stable_edges_data, comb$citest)
 
 
 summarized_stable_edges <- lapply(grouped_stable_edges, function(group) {
@@ -71,13 +74,36 @@ summarized_stable_edges <- lapply(grouped_stable_edges, function(group) {
 })
 
 
+## Plot the resulting graph
+# per algorithm
+summarized_stable_edges$PC |> plotPC()
+par(mfrow=c(1,2))
+summarized_stable_edges$FCI |> plotAG()
+summarized_stable_edges$CCI |> plotAG()
 
-# Compare matrices
+# per citest
+par(mfrow=c(1,2))
+summarized_stable_edges$gaussCItest |> plotAG()
+summarized_stable_edges$RCoT |> plotAG()
+
+## Compare matrices
+# per algorithm
 fcimat <- summarized_stable_edges$FCI 
 ccimat <- summarized_stable_edges$CCI
-
 shared_edges <- which(fcimat == ccimat, arr.ind = TRUE)
 diff_edges <- which(fcimat != ccimat, arr.ind = TRUE)
+
+cat("Shared Edges:\n")
+print(shared_edges)
+
+cat("\nDiffering Edges:\n")
+print(diff_edges)
+
+# per citest
[…]

### 2024-12-28 code/snellius/precarity_sumscore.R

@@ -195,7 +195,9 @@ run_and_save <- function(alpha, threshold, citest, algorithm) {
     num_subsamples = num_subsamples,
     alpha = alpha,
     threshold = threshold,
-    citest = citest
+    citest = citest,
+    fixedEdges = fixedEdges_full,
+    fixedGaps = fixedGaps_full
   )
   
   # Create file name

### 2024-12-28 code/snellius/robust_func.R

@@ -6,7 +6,6 @@ plan(multicore)
 message("Number of parallel workers: ", nbrOfWorkers())
 
 
-
 # General function to run a causal discovery algorithm on a subsample
 run_causal_subsample <- function(data_subsample, algorithm, citest, alpha, num_f = 100, num_f2 = 10, fixedEdges = fixedEdges, fixedGaps = fixedGaps) {
   message(sprintf("Running causal discovery for algorithm: %s, CI test: %s, alpha: %f", algorithm, citest, alpha))

### 2024-12-28 draft_v1/draft_v1.qmd
draft_update_add_presum

@@
 
+## Precarity as sum score
+
+```{r}
+#| label: fig-presum
+#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using FCI and CCI"
+#| fig-align: center
+#| out-width: 80%
+
+knitr::include_graphics("img/presum_graph_refined.png")
+
+```
+
 # Discussion
@@
 #| label: fig-sum-mat
-#| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum scores"
+#| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum score"
 #| fig-align: center
@@
 ```
+
+```{r}
+#| label: fig-presum-mat
+#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precarity sum score"
+#| fig-align: center
+#| out-width: 70%
+knitr::include_graphics("img/presumgraph_mat.pdf")
+```

### 2024-12-28 draft_v1/draft_v1.tex
draft_update_add_presum

@@
 
-\date{2024-12-23}
+\date{2024-12-28}
 \begin{document}
@@
 \caption{\label{fig-sym}Resulting graphs of precarity factors and
+individual depression symptoms using FCI and CCI}
+
+\end{figure}%
+
+\subsection{Precarity as sum score}\label{precarity-as-sum-score}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth,height=\textheight]{img/presum_graph_refined.png}
+
+}
+
+\caption{\label{fig-presum}Resulting graphs of precarity sum score and
 individual depression symptoms using FCI and CCI}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-7-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-8-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-8-1.pdf}
+\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-9-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-9-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-10-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-10-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-11-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-11-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-12-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-12-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-13-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-13-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-14-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-14-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-15-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-14-2.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-15-2.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-14-3.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-15-3.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-14-4.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-15-4.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-15-1.pdf}
+\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-16-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-17-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-17-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-18-1.pdf}
 \end{center}
@@
 \caption{\label{fig-sum-mat}Proportions of edge endpoint types for
-graphs based on depression sum scores}
+graphs based on depression sum score}
 
@@
 
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/presumgraph_mat.pdf}
+
+}
+
+\caption{\label{fig-presum-mat}Proportions of edge endpoint types for
+graphs based on individual depression symptoms with precarity sum score}
+
+\end{figure}%
+
 \end{document}

### 2024-12-28 code/edge_proportion_sum.R

@@ -27,7 +27,8 @@ library(ggplotify)
 
 # Define the directory containing the .rds files
 directory <- "Precarity_HELIUS/data/dep_sumscore"  # Replace with the path you need
-#directory <- "Precarity_HELIUS/data/dep_symscore"  # individual symptom
+# directory <- "Precarity_HELIUS/data/dep_symscore"  # individual symptom
+# directory <- "Precarity_HELIUS/data/presum_symscore"  # individual symptom
 
 # List all the file names from the directory
 file_names <- list.files(directory, full.names = TRUE, pattern = "\\.rds$")
@@ -289,18 +290,31 @@ heatmap_CCI_plot <- ggplotify::as.ggplot(grid.grabExpr(draw(heatmap_cci))) + ggt
 legend <- ggplotify::as.ggplot(grid.grabExpr(draw(custom_legend)))
 legend_h <- ggplotify::as.ggplot(grid.grabExpr(draw(custom_legend_h))) # horizontal version 
 
-# Combine heatmaps and legend using ggarrange
-sumgraph_mat <- ggpubr::ggarrange(
+
+## Combine heatmaps and legend using ggarrange
+# vertical version
+sumgraph_mat_v <- ggpubr::ggarrange(
   heatmap_FCI_plot,
   heatmap_CCI_plot,
-  legend,
+  legend_v,
   ncol = 3,  # Heatmaps side by side with legend
   widths = c(3, 3, 0.9)  # Adjust the relative width of each panel
 )
-print(sumgraph_mat)
-# save plot
-#ggpubr::ggexport(filename = "sumgraph_mat.pdf", plot = sumgraph_mat,  width = 11, height = 5, units = "cm")
+print(sumgraph_mat_v)
 
+# horizontal version
+sumgraph_mat_h <- ggpubr::ggarrange(
+  heatmap_FCI_plot,
+  heatmap_CCI_plot,
+  legend_h,
+  nrow = 3,  # Heatmaps side by side with legend
+  heights = c(3, 3, 0.1)  # Adjust the relative width of each panel
+)
+print(sumgraph_mat_h)
+# save plot
+# ggpubr::ggexport(filename = "sumgraph_mat.pdf", plot = sumgraph_mat,  width = 11, height = 5, units = "cm")
+# ggpubr::ggexport(filename = "symgraph_mat.pdf", plot = sumgraph_mat,  width = 11, height = 5, units = "cm")
[…]

### 2024-12-28 code/graphviz/symptom_presum_graph.py

@@ -26,44 +26,44 @@ for node, pos in fixed_positions.items():
 
 
 # Manually define edges based on the adj.matrices
-dot.edge("anh", "dep", dir="both", arrowhead="odot", arrowtail="normal", style="dashed")
-dot.edge("anh", "slp", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
-dot.edge("anh", "ene", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
+dot.edge("anh", "dep", dir="both", arrowhead="odot", arrowtail="normal", style="dashed", color="#a9a9a9")
+dot.edge("anh", "slp", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
+dot.edge("anh", "ene", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
 dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("anh", "con", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
-dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
+dot.edge("anh", "con", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
+dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
 
-dot.edge("dep", "slp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("dep", "app", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
-dot.edge("dep", "glt", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("dep", "con", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("dep", "mot", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("dep", "sui", dir="both", arrowhead="odot", arrowtail="odot", style="dashed")
-dot.edge("dep", "Precarity", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
+dot.edge("dep", "slp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
[…]

### 2024-12-28 code/post_analysis.R

@@ -31,6 +31,8 @@ filenames <- glue::glue(
 
 # directory containing the files
 directory <- "Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_sumscore"  
+# directory <- "Precarity_HELIUS/data/dep_sumscore"  
+# directory <- "Precarity_HELIUS/data/dep_symscore"  
 
 # Full paths to the files
 full_paths <- file.path(directory, filenames)

### 2024-12-28 code/snellius/precarity_sumscore.R

@@ -129,42 +129,42 @@ sumpre_sym <- depsym |>
 # num_subsamples <- 100     # Number of subsamples
 
 
-## using fixed gaps and edges
-# Initialize a 14x14 matrix with FALSE for both fixedEdges and fixedGaps
-fixedEdges_full <- matrix(FALSE, nrow = 10, ncol = 10)
-fixedGaps_full <- matrix(FALSE, nrow = 10, ncol = 10)
-
-# Define the constraints for the first 9 variables (fixedEdges)
-fixedEdges_full[1:9, 1:9] <- matrix(
-  c(
-    FALSE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,  TRUE,
-    TRUE,  TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,
-    FALSE,  TRUE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,
-    TRUE,  TRUE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,
-    FALSE,  TRUE, FALSE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE,
-    FALSE,  TRUE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE, FALSE
-  ),
-  nrow = 9, byrow = TRUE
-)
-
-# Define the constraints for the first 9 variables (fixedGaps)
-fixedGaps_full[1:9, 1:9] <- matrix(
-  c(
-    FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
-    FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE, FALSE, FALSE, FALSE,
-    FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
[…]

### 2024-12-28 code/snellius/sym_gaussian.R

@@ -36,41 +36,41 @@ num_subsamples <- 100    # Number of subsamples
 
 ## Set up fixed gaps and edges for individual symptom analysis *speed up*
 ## using fixed gaps and edges
-# Initialize a 14x14 matrix with FALSE for both fixedEdges and fixedGaps
-fixedEdges_full <- matrix(FALSE, nrow = 14, ncol = 14)
-fixedGaps_full <- matrix(FALSE, nrow = 14, ncol = 14)
-
-# Define the constraints for the first 9 variables (fixedEdges)
-fixedEdges_full[1:9, 1:9] <- matrix(
-  c(
-    FALSE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,  TRUE,
-    TRUE,  TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,
-    FALSE,  TRUE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,
-    TRUE,  TRUE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,
-    FALSE,  TRUE, FALSE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE,
-    FALSE,  TRUE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE, FALSE
-  ),
-  nrow = 9, byrow = TRUE
-)
-
-# Define the constraints for the first 9 variables (fixedGaps)
-fixedGaps_full[1:9, 1:9] <- matrix(
-  c(
-    FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
-    FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE, FALSE, FALSE, FALSE,
[…]

### 2024-12-28 code/snellius/sym_rcot.R

@@ -43,41 +43,41 @@ num_subsamples <- 30     # Number of subsamples
 
 
 ## Set up fixed gaps and edges for individual symptom analysis (*speed up*)
-# Initialize a 14x14 matrix with FALSE for both fixedEdges and fixedGaps
-fixedEdges_full <- matrix(FALSE, nrow = 14, ncol = 14)
-fixedGaps_full <- matrix(FALSE, nrow = 14, ncol = 14)
-
-# Define the constraints for the first 9 variables (fixedEdges)
-fixedEdges_full[1:9, 1:9] <- matrix(
-  c(
-    FALSE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,  TRUE,
-    TRUE,  TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-    TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,
-    FALSE,  TRUE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,
-    TRUE,  TRUE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,
-    FALSE,  TRUE, FALSE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE,
-    FALSE,  TRUE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE, FALSE
-  ),
-  nrow = 9, byrow = TRUE
-)
-
-# Define the constraints for the first 9 variables (fixedGaps)
-fixedGaps_full[1:9, 1:9] <- matrix(
-  c(
-    FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
-    FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE, FALSE, FALSE, FALSE,
-    FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
[…]

### 2024-12-28 draft_v1/draft_v1.qmd
add_PC_presum_result

@@
 
+```{r}
+#| layout-nrow: 2
+#| label: fig-pc_presum
+#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using PC"
+#| fig-subcap: 
+#|   - "Using both GaussianCI and RCoT"
+#|   - "Using only RCoT"
+#| fig-align: center
+#| out-width: 70%
+
+knitr::include_graphics("img/pc_presum_GaussianCIRCoT.png")
+knitr::include_graphics("img/pc_presum_onlyRCoT.png")
+```
+
 \clearpage

### 2024-12-28 draft_v1/draft_v1.tex
add_PC_presum_result

@@
 
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_presum_GaussianCIRCoT.png}
+
+}
+
+\subcaption{\label{fig-pc_presum-1}Using both GaussianCI and RCoT}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_presum_onlyRCoT.png}
+
+}
+
+\subcaption{\label{fig-pc_presum-2}Using only RCoT}
+
+\end{minipage}%
+
+\caption{\label{fig-pc_presum}Resulting graphs of precarity sum score
+and individual depression symptoms using PC}
+
+\end{figure}%
+
 \clearpage

### 2024-12-28 code/graphviz/create_graph_PC.py

@@ -1,0 +1,127 @@
+from graphviz import Digraph
+
+# Define the fixed positions for the nodes
+fixed_positions = {
+    "anh": (0, 3),
+    "dep": (1, 4),
+    "slp": (2, 3),
+    "ene": (3, 4),
+    "app": (4, 3),
+    "glt": (2, 2),
+    "con": (2, 1),
+    "mot": (2, 0),
+    "sui": (1, 1),
+    "Prec": (4,1)
+    # "P.emp": (3, 0),
+    # "P.soc": (4, 1),
+    # "P.hou": (5, 0),
+    # "S.rel": (5, 2),
+    # "S.fin": (6, 1)
+}
+
+# Function to create a graph with fixed layout and bidirectional edges
+def create_curvy_graph_with_bidirectional_edges(matrix, title, positions, filename):
+    dot = Digraph(format='png', engine='neato')
+    dot.attr(size="13,13", dpi="300", splines="true", ratio="fill")
+    #dot.attr(label=title, labelloc="top", fontsize="20", fontname="Palatino")
+    # Set global node attributes for uniform size
+    dot.attr('node', fixedsize='true', width='1', height='1')
+    
+    # Add nodes with fixed positions and uniform sizes
+    for node, pos in positions.items():
+        dot.node(node, node, shape="circle", pos=f"{pos[0]},{pos[1]}!",
+                 fixedsize="true", width="1", height="1", fontsize="30")
+    
+    # Add edges based on the matrix
+    nodes = list(positions.keys())
+    for i, src in enumerate(nodes):
+        for j, dest in enumerate(nodes):
+            if i < j:  # Avoid duplicate processing
+                if matrix[i][j] == 1 and matrix[j][i] == 1:
[…]

### 2024-12-28 code/post_analysis.R

@@ -20,6 +20,10 @@ alpha <- c(0.01, 0.05)
 threshold <- c(0.5, 0.6, 0.7, 0.8) # c(0.6, 0.7) # for symptom data
 citest <- c("gaussCItest", "RCoT")
 comb <- expand.grid(algorithm = algorithm, alpha = alpha, threshold = threshold, citest = citest)
+
+# Filter combinations for "PC" and "RCoT"
+# filtered_comb <- comb[comb$algorithm == "PC" & comb$citest == "RCoT", ]
+
 # sum score data
 filenames <- glue::glue(
     "{comb$algorithm}_alpha_{sprintf('%.2f', comb$alpha)}_threshold_{sprintf('%.1f', comb$threshold)}_citest_{comb$citest}.rds"
@@ -27,6 +31,10 @@ filenames <- glue::glue(
 # # gaussian sym test data
 # filenames <- glue::glue(
 #   "gaussian{comb$algorithm}_alpha_{sprintf('%.2f', comb$alpha)}_threshold_{sprintf('%.1f', comb$threshold)}.rds"
+# )
+# Generate filenames for the filtered conditions
+# filtered_filenames <- glue::glue(
+#   "{filtered_comb$algorithm}_alpha_{sprintf('%.2f', filtered_comb$alpha)}_threshold_{sprintf('%.1f', filtered_comb$threshold)}_citest_{filtered_comb$citest}.rds"
 # )
 
 # directory containing the files
@@ -36,6 +44,7 @@ directory <- "Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_
 
 # Full paths to the files
 full_paths <- file.path(directory, filenames)
+# full_paths <- file.path(directory, filtered_filenames)
 
 # Read all .rds files into a list
 rds_data <- lapply(full_paths, readRDS)
@@ -53,6 +62,7 @@ grouped_stable_edges <- split(stable_edges_data, comb$algorithm)
 grouped_stable_edges_ci <- split(stable_edges_data, comb$citest)
 
 
+# summarize them by taking the most frequent symbol per element in matrices
 summarized_stable_edges <- lapply(grouped_stable_edges, function(group) {
   # Check dimensions from the first matrix in the group
   nrow <- nrow(group[[1]])
@@ -112,6 +122,4 @@ print(shared_edges)
 
 cat("\nDiffering Edges:\n")
 print(diff_edges)
-
-
 

### 2024-12-28 code/snellius/sumscore_pre.sh
minor_edits

@@ -7,8 +7,7 @@
 #SBATCH --exclusive 
 #SBATCH --nodes=1
 #SBATCH --tasks-per-node=4  
-#SBATCH --cpus-per-task=30
-#SBATCH --array=0-3 # Total jobs based on parameter combinations
+#SBATCH --cpus-per-task=40
 
 #Send email at start en end
 #SBATCH --mail-type=BEGIN,END
@@ -25,17 +24,17 @@ module load R-bundle-CRAN/2023.12-foss-2023a
 export R_LIBS="/gpfs/home1/kpark/rpackages:/gpfs/home1/kpark/R/x86_64-pc-linux-gnu-library/4.3:$R_LIBS"
 
 # Define parameters
-alphas=("0.01" "0.05")
-thresholds=("0.6" "0.7")
+# alphas=("0.01" "0.05")
+# thresholds=("0.6" "0.7")
  
 #Calculate indices for combinations
-alpha_index=$((SLURM_ARRAY_TASK_ID / 2))
-threshold_index=$((SLURM_ARRAY_TASK_ID % 2))
-alpha=${alphas[$alpha_index]}
-threshold=${thresholds[$threshold_index]}
+# alpha_index=$((SLURM_ARRAY_TASK_ID / 2))
+# threshold_index=$((SLURM_ARRAY_TASK_ID % 2))
+# alpha=${alphas[$alpha_index]}
+# threshold=${thresholds[$threshold_index]}
 
 # Run the R script with the specific alpha and threshold
-# Rscript $HOME/helius/code/sym_rcot.R "$alpha" "$threshold"
-Rscript $HOME/helius/code/precarity_sumscore.R "$alpha" "$threshold"
+# Rscript $HOME/helius/code/precarity_sumscore.R "$alpha" "$threshold"
+Rscript $HOME/helius/code/precarity_sumscore.R
 
 

### 2024-12-28 code/utils/rff.R

@@ -4,6 +4,7 @@ library(stats)    # For kernel functions
 library(pracma)   # For cosine function
 library(momentchi2)  # For advanced p-value calculation using Lancaster or HBE methods
 
+
 # Multinomial Kernel function for categorical variables
 multinomial_kernel <- function(x, y, theta = 1) {
   if (x == y) {

### 2025-01-01 code/graphviz/create_graph_PC.py
update_fixedEdgeGap

@@ -97,10 +97,10 @@ def create_curvy_graph_with_bidirectional_edges(matrix, title, positions, filena
 ## precarity sum score case PC
 presum_pc1 = [
  [0, 1, 1, 1, 1, 1, 1, 1, 0, 1],
- [0, 0, 1, 1, 1, 1, 1, 1, 1, 1],
+ [0, 0, 1, 1, 1, 0, 1, 1, 0, 1],
  [0, 0, 1, 0, 1, 0, 1, 0, 0, 1],
- [0, 1, 1, 0, 1, 1, 1, 1, 0, 1],
- [0, 0, 0, 0, 0, 1, 1, 1, 0, 1],
+ [0, 0, 1, 0, 1, 1, 1, 1, 0, 1],
+ [0, 0, 0, 0, 0, 0, 1, 1, 0, 1],
  [0, 1, 1, 1, 1, 0, 1, 1, 1, 1],
  [0, 0, 0, 0, 0, 0, 0, 1, 0, 1],
  [0, 0, 0, 0, 0, 0, 0, 0, 0, 1],

### 2025-01-01 code/graphviz/symptom_presum_graph.py

@@ -34,6 +34,8 @@ dot.edge("anh", "con", dir="both", arrowhead="odot", arrowtail="odot", style="da
 dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
 
 dot.edge("dep", "slp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+dot.edge("dep", "ene", dir="both", arrowhead="normal", arrowtail="normal")
+
 dot.edge("dep", "app", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
 dot.edge("dep", "glt", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
 dot.edge("dep", "con", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
@@ -48,6 +50,7 @@ dot.edge("slp", "mot", dir="both", arrowhead="odot", arrowtail="odot", style="da
 dot.edge("slp", "Precarity", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
 
 dot.edge("ene", "app", dir="both", arrowhead="normal", arrowtail="normal")
+dot.edge("ene", "con", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
 dot.edge("ene", "mot", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
 
 dot.edge("app", "glt", dir="both", arrowhead="normal", arrowtail="normal")

### 2025-01-01 code/snellius/precarity_sumscore.R

@@ -131,44 +131,39 @@ sumpre_sym <- depsym |>
 
 # ## using fixed gaps and edges
 # # Initialize a 14x14 matrix with FALSE for both fixedEdges and fixedGaps
-# fixedEdges_full <- matrix(FALSE, nrow = 10, ncol = 10)
-# fixedGaps_full <- matrix(FALSE, nrow = 10, ncol = 10)
-# 
-# # Define the constraints for the first 9 variables (fixedEdges)
-# fixedEdges_full[1:9, 1:9] <- matrix(
-#   c(
-#     FALSE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,  TRUE,
-#     TRUE,  TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,
-#     FALSE,  TRUE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,
-#     TRUE,  TRUE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,
-#     FALSE,  TRUE, FALSE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE,
-#     FALSE,  TRUE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE, FALSE
-#   ),
-#   nrow = 9, byrow = TRUE
-# )
-# 
-# # Define the constraints for the first 9 variables (fixedGaps)
-# fixedGaps_full[1:9, 1:9] <- matrix(
-#   c(
-#     FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
-#     FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE, FALSE, FALSE, FALSE,
-#     FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
-#     FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
[…]

### 2025-01-01 code/snellius/sym_gaussian.R

@@ -36,41 +36,36 @@ num_subsamples <- 100    # Number of subsamples
 
 ## Set up fixed gaps and edges for individual symptom analysis *speed up*
 ## using fixed gaps and edges
-# # Initialize a 14x14 matrix with FALSE for both fixedEdges and fixedGaps
-# fixedEdges_full <- matrix(FALSE, nrow = 14, ncol = 14)
-# fixedGaps_full <- matrix(FALSE, nrow = 14, ncol = 14)
-# 
-# # Define the constraints for the first 9 variables (fixedEdges)
-# fixedEdges_full[1:9, 1:9] <- matrix(
-#   c(
-#     FALSE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,  TRUE,
-#     TRUE,  TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,
-#     FALSE,  TRUE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,
-#     TRUE,  TRUE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,
-#     FALSE,  TRUE, FALSE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE,
-#     FALSE,  TRUE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE, FALSE
-#   ),
-#   nrow = 9, byrow = TRUE
-# )
-# 
-# # Define the constraints for the first 9 variables (fixedGaps)
-# fixedGaps_full[1:9, 1:9] <- matrix(
-#   c(
-#     FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
[…]

### 2025-01-01 code/snellius/sym_rcot.R

@@ -43,41 +43,36 @@ num_subsamples <- 30     # Number of subsamples
 
 
 ## Set up fixed gaps and edges for individual symptom analysis (*speed up*)
-# # Initialize a 14x14 matrix with FALSE for both fixedEdges and fixedGaps
-# fixedEdges_full <- matrix(FALSE, nrow = 14, ncol = 14)
-# fixedGaps_full <- matrix(FALSE, nrow = 14, ncol = 14)
-# 
-# # Define the constraints for the first 9 variables (fixedEdges)
-# fixedEdges_full[1:9, 1:9] <- matrix(
-#   c(
-#     FALSE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,  TRUE,
-#     TRUE,  TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,  TRUE, FALSE, FALSE,
-#     TRUE, FALSE,  TRUE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,
-#     FALSE,  TRUE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,  TRUE,
-#     TRUE,  TRUE,  TRUE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE, FALSE,
-#     FALSE,  TRUE, FALSE, FALSE,  TRUE,  TRUE,  TRUE, FALSE,  TRUE,
-#     FALSE,  TRUE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE, FALSE
-#   ),
-#   nrow = 9, byrow = TRUE
-# )
-# 
-# # Define the constraints for the first 9 variables (fixedGaps)
-# fixedGaps_full[1:9, 1:9] <- matrix(
-#   c(
-#     FALSE, FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE,  TRUE,  TRUE,
-#     FALSE, FALSE, FALSE, FALSE,  TRUE, FALSE, FALSE, FALSE, FALSE,
[…]

### 2025-01-01 code/utils/robust_func.R

@@ -58,7 +58,7 @@ run_causal_subsample <- function(data_subsample, algorithm, citest, alpha, num_f
     stop("Unsupported CI test specified.")
   }
   
-  # Select the CI test function
+  # Select the CI test function 
   indepTest <- switch(citest,
                       "gaussCItest" = pcalg::gaussCItest,
                       "RCoT" = RCIT::RCoT,
@@ -101,6 +101,7 @@ causal_subsampling <- function(data, algorithm, subsample_size, num_subsamples, 
     1:num_subsamples, 
     ~ {
       library(RCIT)
+      library(pcalg)
       
       data_subsample <- dplyr::sample_n(data, subsample_size, replace = TRUE)
       message(sprintf("Processing subsample: %s", .x))

### 2025-01-01 draft_v1/draft_v1.qmd
update_draftv1

@@
 After preprocessing, the HELIUS dataset comprises 21,628 samples.
-In addition to the five precariousness factors, we include PHQ-9 scores — both the total sum score and individual symptom scores — to represent depression.
-In the subsequent causal discovery analysis, we will examine depression both as an aggregated sum score and through its individual symptom-level representations.
+Along with the five precariousness factors, we also compute an overall precarity score as the combined value of these factors.
+Additionally, PHQ-9 scores are included to represent depression, both as a total sum score and as individual symptom scores.
+In the subsequent causal discovery analysis, we examine the relationship between depression and precarity using both aggregated sum scores and their individual-level representations.
 Refer to @fig-dist for the overall distributions of the variables used in the analysis.
@@
 #|   
-#|   *P.emp* = employment precariousness; *P.hou* = housing precariousness; *P.soc* = social precariousness; *S.fin* = recent financial stressors; *S.rel* = recent relational stressors;
+#|   *P.emp* = employment precariousness; *P.hou* = housing precariousness; *P.soc* = social precariousness; *S.fin* = recent financial stressors; *S.rel* = recent relational stressors; *Precarity* = overall precarity;
 #|   *PHQsum* = PHQ-9 sum score; *anh* = anhedonia; *app* = appetite; *con* = concentration; *dep* = depressed mood; *ene* = energy; *glt* = guilty; *mot* = motor; *sui* = suicidal
@@
 ## Analysis
-We analyze the causal structure using two approaches: one with the PHQ sum score representing overall depression severity, and another with individual symptom scores.
-The PHQ sum score simplifies the analysis by reducing dimensionality, which is computationally advantageous and provides a broad, interpretable perspective on depression's relationship with precarity factors.
-In contrast, individual symptom scores offer a nuanced understanding, capturing the heterogeneous ways symptoms respond to precarity factors.
-However, analyzing individual symptoms poses methodological challenges due to their non-standard distributions and the increased complexity introduced by the higher dimensionality.
-By employing both approaches, we balance simplicity and granularity, ensuring robustness in our findings: the PHQ sum score captures overarching trends, while symptom-level analysis reveals detailed dynamics, essential for tailoring specific interventions.
-
-To evaluate the sensitivity of the results to the choice of alpha levels and ensure consistency, we test two significance levels --- $\alpha = 0.01, 0.05$.
-To further ensure robustness, we employ bootstrapping to generate 100 bootstrap samples For each sample, we estimate causal graphs and retain only the edges and directions that appear above predefined thresholds.
-
+We analyze the causal structure using three approaches: examining the relationship (1) between the five individual precarity factors and the PHQ sum score, which represents overall depression severity; (2) between individual symptoms and precarity factors; and (3) between individual symptoms and overall precarity (the sum of the five precarity factors).
+Using the PHQ sum score simplifies the analysis by reducing dimensionality, offering computational efficiency and a broad, interpretable perspective on the relationship between depression and precarity factors.
+In contrast, including individual symptoms provide a more detailed understanding, capturing the diverse ways symptoms respond to different precarity factors.
+However, this approach introduces methodological challenges, such as non-standard distributions and the complexities of increased dimensionality.
+Finally, examining individual symptoms in relation to the overall precarity score complements the analysis by revealing how symptoms collectively connect to general precarity.
+By integrating these three approaches, we balance simplicity and granularity, ensuring robust findings.
+<!-- The PHQ sum score and precarity sum score capture overarching trends between overall depression and precarity, while symptom-level analysis with individual precarity factors reveals detailed dynamics, essential for tailoring specific interventions.
+-->
+
+To address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping.
+For each of the 100 bootstrap samples, we estimate causal graphs and retain only the edges and directions that exceed predefined thresholds.
 The analyses are conducted under the following conditions:
@@
 
-This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms, with each combination repeated for 100 bootstrap samples, resulting in a total of 1,600 resulting graphs.
-For analyses involving individual symptom variables, we streamline the setup by using thresholds of 0.6 and 0.7, reducing the number of bootstrap samples to 30.
+This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms.
+Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm.
+For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30 to ease computational demands.
 <!-- Additionally, we fix the skeleton of edges among symptom variables based on the common structure estimated across all three algorithms, thereby reducing the computational time required to estimate skeletons.
@@
 To summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.[^3] This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression.
-Refer to @fig-workflow for the analysis workflow.
+See @fig-workflow for an overview of the analysis workflow.
 
@@
 
-knitr::include_graphics("img/simsetup.png")
+knitr::include_graphics("img/simsetup2.png")
 
@@
 ## Precarity as sum score
+Lastly, we examine the relationships between individual symptoms and overall precarity, represented as the sum of five precariousness factors.
+This approach completes the analysis by providing a perspective on how individual symptoms relate to overall precarity.
+It complements the previous analysis, which focused on individual symptoms and individual precarity factors, by potentially capturing distributed relationships across different precarity factors through their aggregation into a single overall score.
+As before, black solid lines indicate edges agreed upon by both FCI and CCI, while gray dashed lines represent edges where the two algorithms disagree.
+Circles at the endpoints of gray dashed edges denote unresolved directionality.
+Additionally, navy dashed edges highlight relationships that appear exclusively in graphs generated using Gaussian CI tests.
+
+Compared to @fig-sym, this analysis reveals greater disagreement between FCI and CCI regarding edge directionality, as evidenced by more dashed edges.
+FCI tends to estimate more undirected edges (circles), while CCI more frequently assigns directed arrows between symptom-symptom relationships.
+For a detailed breakdown, see [@sec-propmatrix], which shows the proportion of symbols produced by each algorithm.
+When examining overall precarity, we observe that more symptoms are now connected to the precarity score (*Precarity*) compared to the individual precarity factor analysis in @fig-sym.
+In addition to *glt*, *slp*, *anh*, and *sui*, which are previously shown to be connected to individual precarity factors, *mot*, *app*, and *dep* are now identified as having causal relationships with overall precarity.
+Furthermore, when using the Gaussian CI test, con is also shown to be causally related to overall precarity.
+Notably, all these connections are directed towards precarity (o→), indicating that overall precarity is not the cause of symptoms.
+The exception is the bidirectional edge between *sui* <-> *Precarity*, which suggests a connection via latent confounders.
+FCI predominantly estimates directed arrows (→) for symptom-precarity relationships, suggesting that symptoms are driving precarity.
+With fewer variables (reduced from 14 to 9), this approach appears to have slightly more statistical power, enabling algorithms to estimate relationships with greater confidence compared to the previous analysis involving individual symptoms and individual precarity factors.
 

### 2025-01-01 draft_v1/draft_v1.tex
update_draftv1

@@
 
-\date{2024-12-28}
+\date{2024-12-30}
 \begin{document}
@@
 
-After preprocessing, the HELIUS dataset comprises 21,628 samples. In
-addition to the five precariousness factors, we include PHQ-9 scores ---
-both the total sum score and individual symptom scores --- to represent
-depression. In the subsequent causal discovery analysis, we will examine
-depression both as an aggregated sum score and through its individual
-symptom-level representations. Refer to Figure~\ref{fig-dist} for the
-overall distributions of the variables used in the analysis.
+After preprocessing, the HELIUS dataset comprises 21,628 samples. Along
+with the five precariousness factors, we also compute an overall
+precarity score as the combined value of these factors. Additionally,
+PHQ-9 scores are included to represent depression, both as a total sum
+score and as individual symptom scores. In the subsequent causal
+discovery analysis, we examine the relationship between depression and
+precarity using both aggregated sum scores and their individual-level
+representations. Refer to Figure~\ref{fig-dist} for the overall
+distributions of the variables used in the analysis.
 
@@
 \emph{S.fin} = recent financial stressors; \emph{S.rel} = recent
-relational stressors; \emph{PHQsum} = PHQ-9 sum score; \emph{anh} =
-anhedonia; \emph{app} = appetite; \emph{con} = concentration; \emph{dep}
-= depressed mood; \emph{ene} = energy; \emph{glt} = guilty; \emph{mot} =
-motor; \emph{sui} = suicidal}
+relational stressors; \emph{Precarity} = overall precarity;
+\emph{PHQsum} = PHQ-9 sum score; \emph{anh} = anhedonia; \emph{app} =
+appetite; \emph{con} = concentration; \emph{dep} = depressed mood;
+\emph{ene} = energy; \emph{glt} = guilty; \emph{mot} = motor; \emph{sui}
+= suicidal}
 
@@
 
-We analyze the causal structure using two approaches: one with the PHQ
-sum score representing overall depression severity, and another with
-individual symptom scores. The PHQ sum score simplifies the analysis by
-reducing dimensionality, which is computationally advantageous and
-provides a broad, interpretable perspective on depression's relationship
-with precarity factors. In contrast, individual symptom scores offer a
-nuanced understanding, capturing the heterogeneous ways symptoms respond
-to precarity factors. However, analyzing individual symptoms poses
-methodological challenges due to their non-standard distributions and
-the increased complexity introduced by the higher dimensionality. By
-employing both approaches, we balance simplicity and granularity,
-ensuring robustness in our findings: the PHQ sum score captures
-overarching trends, while symptom-level analysis reveals detailed
-dynamics, essential for tailoring specific interventions.
-
-To evaluate the sensitivity of the results to the choice of alpha levels
-and ensure consistency, we test two significance levels ---
-\(\alpha = 0.01, 0.05\). To further ensure robustness, we employ
-bootstrapping to generate 100 bootstrap samples For each sample, we
-estimate causal graphs and retain only the edges and directions that
-appear above predefined thresholds.
-
+We analyze the causal structure using three approaches: examining the
+relationship (1) between the five individual precarity factors and the
+PHQ sum score, which represents overall depression severity; (2) between
+individual symptoms and precarity factors; and (3) between individual
+symptoms and overall precarity (the sum of the five precarity factors).
+Using the PHQ sum score simplifies the analysis by reducing
+dimensionality, offering computational efficiency and a broad,
+interpretable perspective on the relationship between depression and
+precarity factors. In contrast, including individual symptoms provide a
+more detailed understanding, capturing the diverse ways symptoms respond
+to different precarity factors. However, this approach introduces
+methodological challenges, such as non-standard distributions and the
+complexities of increased dimensionality. Finally, examining individual
+symptoms in relation to the overall precarity score complements the
+analysis by revealing how symptoms collectively connect to general
+precarity. By integrating these three approaches, we balance simplicity
+and granularity, ensuring robust findings.
+
+To address sensitivity to parameter values and enhance the robustness of
+our results, we implement multiple settings combined with bootstrapping.
+For each of the 100 bootstrap samples, we estimate causal graphs and
+retain only the edges and directions that exceed predefined thresholds.
 The analyses are conducted under the following conditions:
@@
 This setup yields 16 combinations (2 significance levels × 4 thresholds
-× 2 CI tests), applied across three algorithms, with each combination
-repeated for 100 bootstrap samples, resulting in a total of 1,600
-resulting graphs. For analyses involving individual symptom variables,
-we streamline the setup by using thresholds of 0.6 and 0.7, reducing the
-number of bootstrap samples to 30.
+× 2 CI tests), applied across three algorithms. Each combination is
+repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per
+algorithm. For analyses involving individual symptom variables, we
+simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing
+the number of bootstrap samples to 30 to ease computational demands.
 
@@
 understanding of the relationships between precarity factors and
-depression. Refer to Figure~\ref{fig-workflow} for the analysis
-workflow.
+depression. See Figure~\ref{fig-workflow} for an overview of the
+analysis workflow.
 
@@
 
-\includegraphics[width=0.8\textwidth,height=\textheight]{img/simsetup.png}
+\includegraphics[width=0.8\textwidth,height=\textheight]{img/simsetup2.png}
 
@@
 \subsection{Precarity as sum score}\label{precarity-as-sum-score}
+
+Lastly, we examine the relationships between individual symptoms and
+overall precarity, represented as the sum of five precariousness
+factors. This approach completes the analysis by providing a perspective
+on how individual symptoms relate to overall precarity. It complements
+the previous analysis, which focused on individual symptoms and
+individual precarity factors, by potentially capturing distributed
+relationships across different precarity factors through their
+aggregation into a single overall score. As before, black solid lines
+indicate edges agreed upon by both FCI and CCI, while gray dashed lines
+represent edges where the two algorithms disagree. Circles at the
+endpoints of gray dashed edges denote unresolved directionality.
+Additionally, navy dashed edges highlight relationships that appear
+exclusively in graphs generated using Gaussian CI tests.
+
+Compared to Figure~\ref{fig-sym}, this analysis reveals greater
+disagreement between FCI and CCI regarding edge directionality, as
+evidenced by more dashed edges. FCI tends to estimate more undirected
+edges (circles), while CCI more frequently assigns directed arrows
+between symptom-symptom relationships. For a detailed breakdown, see
+Section~\ref{sec-propmatrix}, which shows the proportion of symbols
+produced by each algorithm. When examining overall precarity, we observe
+that more symptoms are now connected to the precarity score
+(\emph{Precarity}) compared to the individual precarity factor analysis
+in Figure~\ref{fig-sym}. In addition to \emph{glt}, \emph{slp},
+\emph{anh}, and \emph{sui}, which are previously shown to be connected
+to individual precarity factors, \emph{mot}, \emph{app}, and \emph{dep}
+are now identified as having causal relationships with overall
+precarity. Furthermore, when using the Gaussian CI test, con is also
+shown to be causally related to overall precarity. Notably, all these
+connections are directed towards precarity (o→), indicating that overall
+precarity is not the cause of symptoms. The exception is the
+bidirectional edge between \emph{sui} \textless-\textgreater{}
+\emph{Precarity}, which suggests a connection via latent confounders.
+FCI predominantly estimates directed arrows (→) for symptom-precarity
+relationships, suggesting that symptoms are driving precarity. With
+fewer variables (reduced from 14 to 9), this approach appears to have
+slightly more statistical power, enabling algorithms to estimate
+relationships with greater confidence compared to the previous analysis
+involving individual symptoms and individual precarity factors.
 

### 2025-01-01 code/snellius/individual_sym_pre.R
Create individual_sym_pre.R

@@ -1,0 +1,193 @@
+## =============================================================================
+## Script for Running Causal Discovery with Individual Symptoms & Precarity Factors
+## =============================================================================
+#' It runs causal discovery algorithms (PC, FCI, CCI) using multiple parameter combinations 
+#' and saves the results for further analysis.
+#'
+#' @details
+#' - Causal discovery algorithms are applied to subsampled data using combinations 
+#'   of alpha levels, and thresholds. 
+#' - Constraints on edges and gaps (fixedEdges and fixedGaps) are defined to 
+#'   refine the analysis for symptom-level data.
+#' - Results are saved in `.rds` format for both sum score and individual 
+#'   symptom analyses.
+## =============================================================================
+
+
+# load helper functions (snellius)
+source("helius/code/robust_func.R")
+
+# Load dataset (snellius)
+dat <- readRDS("helius/data/dat.rds") 
+
+# Preprocess and scale the precariousness data
+scaled_data <- dat %>%
+  # Select precariousness-related variables
+  # Employment Precariousness 
+  dplyr::select(H1_Arbeidsparticipatie, H1_WerkSit, H1_RecentErv8, 
+                # Financial Precariousness
+                H1_InkHhMoeite, H1_RecentErv9,                      
+                # Housing Precariousness
+                veilig_2012, vrz_2012, P_HUURWON,                   
+                # Cultural Precariousness
+                H1_Discr_sumscore, H1_SBSQ_meanscore, A_BED_RU,         
+                # Social Precariousness
+                H1_RecentErv5, H1_RecentErv6, H1_RecentErv7,       
+                H1_SSQT, H1_SSQSa,                                  
+                # Depression symptoms / sum scores
[…]

### 2025-01-03 code/snellius/indiv_symptom_rcot_CCI.sh
snellius_runRCoT_individualsymptom_v5

@@ -1,6 +1,6 @@
 #!/bin/bash
 #Set job requirements (note set time 1.5 to 2x longer than expected)
-#SBATCH -J sym_rcot
+#SBATCH -J sym_rcot_CCI
 
 #SBATCH -t 120:00:00 
 #SBATCH -p genoa
@@ -8,7 +8,6 @@
 #SBATCH --nodes=1
 #SBATCH --tasks-per-node=1
 #SBATCH --cpus-per-task=128
-#SBATCH --array=0-3 # Total jobs based on parameter combinations
 
 #Send email at start en end
 #SBATCH --mail-type=BEGIN,END
@@ -25,21 +24,6 @@ module load R-bundle-CRAN/2023.12-foss-2023a
 export R_LIBS="/gpfs/home1/kpark/rpackages:/gpfs/home1/kpark/R/x86_64-pc-linux-gnu-library/4.3:$R_LIBS"
 
 
-# Define parameters
-alphas=("0.01" "0.05")
-thresholds=("0.6" "0.7")
-
-# Calculate indices for parameter combinations
-alpha_index=$((SLURM_ARRAY_TASK_ID / 2)) # 2 thresholds per alpha
-threshold_index=$((SLURM_ARRAY_TASK_ID % 2)) # Modulo to wrap indices
-
-# Get specific alpha and threshold based on the indices
-alpha=${alphas[$alpha_index]}
-threshold=${thresholds[$threshold_index]}
-
-# Log the chosen parameters (for debugging purposes)
-echo "Running with alpha=$alpha and threshold=$threshold"
-
 # Run the R script with the specified alpha and threshold
-Rscript $HOME/helius/code/sym_rcot.R "$alpha" "$threshold"
+Rscript $HOME/helius/code/individual_sym_CCI.R 
 

### 2025-01-03 code/snellius/indiv_symptom_rcot_FCI.sh

@@ -1,6 +1,6 @@
 #!/bin/bash
 #Set job requirements (note set time 1.5 to 2x longer than expected)
-#SBATCH -J sym_rcot
+#SBATCH -J sym_rcot_FCI
 
 #SBATCH -t 120:00:00 
 #SBATCH -p genoa
@@ -8,7 +8,6 @@
 #SBATCH --nodes=1
 #SBATCH --tasks-per-node=1
 #SBATCH --cpus-per-task=128
-#SBATCH --array=0-3 # Total jobs based on parameter combinations
 
 #Send email at start en end
 #SBATCH --mail-type=BEGIN,END
@@ -24,22 +23,6 @@ module load R-bundle-CRAN/2023.12-foss-2023a
 # Set R_LIBS to include both directories
 export R_LIBS="/gpfs/home1/kpark/rpackages:/gpfs/home1/kpark/R/x86_64-pc-linux-gnu-library/4.3:$R_LIBS"
 
-
-# Define parameters
-alphas=("0.01" "0.05")
-thresholds=("0.6" "0.7")
-
-# Calculate indices for parameter combinations
-alpha_index=$((SLURM_ARRAY_TASK_ID / 2)) # 2 thresholds per alpha
-threshold_index=$((SLURM_ARRAY_TASK_ID % 2)) # Modulo to wrap indices
-
-# Get specific alpha and threshold based on the indices
-alpha=${alphas[$alpha_index]}
-threshold=${thresholds[$threshold_index]}
-
-# Log the chosen parameters (for debugging purposes)
-echo "Running with alpha=$alpha and threshold=$threshold"
-
 # Run the R script with the specified alpha and threshold
-Rscript $HOME/helius/code/sym_rcot.R "$alpha" "$threshold"
+Rscript $HOME/helius/code/individual_sym_FCI.R 
 

### 2025-01-03 code/snellius/individual_sym_CCI.R

@@ -148,16 +148,21 @@ fixedGaps_full[1:9, 1:9] <- matrix(c(
 ), nrow = 9, byrow = TRUE)
 
 
-# Setup parallel backend
-plan(multicore)
-message("Number of parallel workers: ", nbrOfWorkers())
+## Setup parallel backend
+# Dynamically determine available cores and set workers
+available_cores <- parallel::detectCores() - 4  # Reserve some for the system
+workers <- min(available_cores, 120)  # Use a safe limit below the connection cap
 
+# Set the future plan
+plan(multisession, workers = workers)
+# Confirm the setup
+cat("Using", workers, "parallel workers.\n")
 
 # Define parameters
 alphas = c(0.01, 0.05)
 thresholds = c(0.6, 0.7)
 citests = c("gaussCItest", "RCoT")
-algorithms = c("FCI", "CCI") #c("PC", "FCI", "CCI")
+algorithms = "CCI" #c("FCI", "CCI") #c("PC", "FCI", "CCI")
 data <- depsym
 subsample_size <- nrow(data)  # Size of each subsample
 num_subsamples <- 30     # Number of subsamples
@@ -182,7 +187,7 @@ run_and_save <- function(alpha, threshold, citest, algorithm) {
   )
   
   # Create file name
-  file_name <- glue::glue("data/individual_sym_pre/{algorithm}_alpha_{alpha}_threshold_{threshold}_citest_{citest}.rds")
+  file_name <- glue::glue("helius/data/individual_sym/{algorithm}_alpha_{alpha}_threshold_{threshold}_citest_{citest}.rds")
   saveRDS(result, file = file_name)
   
   # Print a message to track progress
@@ -190,4 +195,4 @@ run_and_save <- function(alpha, threshold, citest, algorithm) {
 }
 
 # Apply the function to each row in the parameter data frame
-pwalk(params, run_and_save, .progress = TRUE)
+purrr::pwalk(params, run_and_save, .progress = TRUE)

### 2025-01-03 code/snellius/individual_sym_FCI.R

@@ -148,16 +148,21 @@ fixedGaps_full[1:9, 1:9] <- matrix(c(
 ), nrow = 9, byrow = TRUE)
 
 
-# Setup parallel backend
-plan(multicore)
-message("Number of parallel workers: ", nbrOfWorkers())
+## Setup parallel backend
+# Dynamically determine available cores and set workers
+available_cores <- parallel::detectCores() - 4  # Reserve some for the system
+workers <- min(available_cores, 120)  # Use a safe limit below the connection cap
 
+# Set the future plan
+plan(multisession, workers = workers)
+# Confirm the setup
+cat("Using", workers, "parallel workers.\n")
 
 # Define parameters
 alphas = c(0.01, 0.05)
 thresholds = c(0.6, 0.7)
 citests = c("gaussCItest", "RCoT")
-algorithms = c("FCI", "CCI") #c("PC", "FCI", "CCI")
+algorithms = "FCI" #c("FCI", "CCI") #c("PC", "FCI", "CCI")
 data <- depsym
 subsample_size <- nrow(data)  # Size of each subsample
 num_subsamples <- 30     # Number of subsamples
@@ -182,7 +187,7 @@ run_and_save <- function(alpha, threshold, citest, algorithm) {
   )
   
   # Create file name
-  file_name <- glue::glue("data/individual_sym_pre/{algorithm}_alpha_{alpha}_threshold_{threshold}_citest_{citest}.rds")
+  file_name <- glue::glue("helius/data/individual_sym/{algorithm}_alpha_{alpha}_threshold_{threshold}_citest_{citest}.rds")
   saveRDS(result, file = file_name)
   
   # Print a message to track progress
@@ -190,4 +195,4 @@ run_and_save <- function(alpha, threshold, citest, algorithm) {
 }
 
 # Apply the function to each row in the parameter data frame
-pwalk(params, run_and_save, .progress = TRUE)
+purrr::pwalk(params, run_and_save, .progress = TRUE)

### 2025-01-03 code/snellius/robust_func.R

@@ -62,7 +62,7 @@ causal_subsampling <- function(data, algorithm, subsample_size, num_subsamples, 
     1:num_subsamples, 
     ~ {
       library(RCIT)
-      
+      library(pcalg)
       data_subsample <- dplyr::sample_n(data, subsample_size, replace = TRUE)
       message(sprintf("Processing subsample: %s", .x))
       

### 2025-01-03 draft_v1/draft_v1.qmd
draft update_till_sec2(analysis)

@@
       - title.tex
-    include-in-header:
-      text: |
-        \usepackage[noblocks]{authblk}
-        \renewcommand*{\Authsep}{, }
-        \renewcommand*{\Authand}{, }
-        \renewcommand*{\Authands}{, }
-        \renewcommand\Affilfont{\small}
-        \usepackage{amssymb}
-        \usepackage{booktabs} % Add to your preamble for cleaner table lines
-        \usepackage{makecell} % Add to your preamble for multi-line cells
+    include-in-header: header.tex
 date: today
@@
 bibliography: references_draft.bib
+nocite: |
+  @*
 csl: apa.csl
@@
 
-```{r, include=FALSE}
+```{r, include=FALSE, cache=TRUE}
 # Load the dataset 
@@
 
-```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%"}
+```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%", cache=TRUE}
 #| fig-cap: >
@@
 Additionally, we include the PC algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation].
-For a more detailed explanation of these algorithms, please refer to @park2024discovering.
 
@@
 \centering
+\small % Make the font size smaller
 \caption{Assumptions of causal discovery algorithms}
-\begin{tabular}{lccccl}
+\begin{tabular}{lccccc}
 \toprule
-Algorithm & Acyclicity & Causal sufficiency & 
-\makecell{Absence of \\ selection bias} & Output \\ 
+Algorithm & Acyclicity & \makecell{Causal \\ sufficiency} & 
+\makecell{Absence of \\ selection bias} & Linearity & Output \\ 
 \midrule
-PC  & $\checkmark$ & $\checkmark$ & $\checkmark$ & CPDAG \\ 
-FCI & $-^{a}$ & $\checkmark$ & $-^{a}$ & PAG \\ 
-CCI & x & x & x & MAAG \\ 
+PC  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & CPDAG \\ 
+FCI & $-^{a}$ & $\checkmark$ & $-^{a}$ & $-^{a}$ & PAG \\ 
+CCI & x & x & x & $\checkmark$ & \makecell{\footnotesize(partially oriented) \\ MAAG} \\ 
 \bottomrule
 \end{tabular}
-\caption*{\small{\textit{Note}. $^{a}$FCI can detect cycles in systems that lack selection bias and exhibit non-linear relationships (Mooij \& Claassen, 2020).}}
+\caption*{\footnotesize{\textit{Note}.
+$^{a}$The FCI algorithm, introduced by Spirtes (1995), is a constraint-based causal discovery method for DAGs that accounts for latent confounding and selection bias.
+Mooij \& Claassen (2020) later showed its applicability to cyclic causal discovery with latent confounding under general faithfulness and Markov conditions, assuming non-linear causal relationships.}}
 \end{table}
@@
 
-<!-- | Algorithm | Acyclicity | Causal sufficiency | Absence of selection bias | Output | -->
-<!-- |-----------|------------|--------------------|---------------------------|--------| -->
-<!-- | PC        | ✓          | ✓                  | ✓                         | CPDAG  | -->
-<!-- | FCI       | -¹         | ✓                  | -¹                        | PAG    | -->
-<!-- | CCI       | x          | x                  | x                         | MAAG   | -->
-<!-- ------------------------------------------------------------------------------------ -->
+<!-- | Algorithm | Acyclicity | Causal sufficiency | Absence of selection bias | Linearity | Output | -->
+<!-- |-----------|------------|--------------------|---------------------------|-----------|--------| -->
+<!-- | PC        | ✓          | ✓                  | ✓                         | ✓         |  CPDAG | -->
+<!-- | FCI       | -¹         | ✓                  | -¹                        | -¹        |  PAG   | -->
+<!-- | CCI       | x          | x                  | x                         | ✓         |  MAAG  | -->
+<!-- ------------------------------------------------------------------------------------------------ -->
 <!-- ¹FCI can detect cycles in systems that lack selection bias and exhibit non-linear relationships (Mooij & Claassen, 2020). -->
@@
 As shown in Table 1, the resulting graphs from FCI and CCI differ slightly (*PAG*: partial ancestral graph; *MAAG*: maximal almost ancestral graph) due to their reliance on different underlying assumptions.
-Both encode information about causal relationships between variables, where the presence of an edge indicates causal *ancestry*.
-Directed edges, $A$ \*→ $B$, specify that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$.
-$A$\*—$B$ represent cases where $B$ is an ancestor of $A$ in every graph in $Equiv(G)$.
-Circle endpoints, $A$\*-o$B$, denote ambiguity in the ancestral relationship, meaning that $B$'s ancestral status relative to $A$ varies across graphs in $Equiv(G)$.
-
-In contrast, the graph produced by the PC algorithm is a *CPDAG* (completed partially directed acyclic graph), where directed edges ($A$ → $B$) indicate that $A$ is a direct cause (parent) of $B$.
+Despite these differences, both graphs are forms of *ancestral graphs*, designed to encode causal relationships between variables, where the presence of an edge indicates causal *ancestry.*
+Directed edges, $A \stararrow B$, indicate that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$.
+The Markov equivalence class refers to the set of graphs that encode the same conditional independence relationships, such that the same *d-separation* relations hold across all graphs in the class [@spirtes2001causation].
+$A \startail B$ indicates that $B$ is an ancestor of $A$ across all graphs in $Equiv(G)$. 
+Circle endpoints, $A \starcirc B$, represent ambiguity in the ancestral relationship, meaning $B$'s ancestral status relative to $A$ varies across graphs in $Equiv(G)$.[^4] 
+When $A \arrowarrow B$, neither $A$ nor $B$ is an ancestor of the other, indicating the presence of a latent confounder between them.
+
+The algorithms also differ in their approaches and assumptions for detecting cycles.
+In CCI's MAAG, $A$ \textemdash\ $B$ indicates that $A$ is an ancestor of $B$ and $B$ is an ancestor of $A$, referring to a cycle between $A \stackedarrows B$. 
+FCI, on the other hand, identifies potential cycles more subtly: fully-connected nodes with circle endpoints ($\circirc$) in FCI's PAG may suggest a cyclic structure. 
+FCI basically provides a sufficient condition to distinguish nodes that are not part of a cycle, offering a more nuanced approach [@mooij2020constraint].
+Table 1 highlights further differences in the assumptions underlying these algorithms.
+CCI operates under the assumption of a linear system, while FCI, particularly for inferring cyclic relationships, assumes a non-linear system without selection bias and adheres to more general faithfulness and Markov conditions (i.e., *$\sigma$-separation* and *$\sigma$-faithfulness* setting) [@forre2018constraint].
+When the respective assumptions of each algorithm are satisfied, their outcomes for cyclic relationships align with those illustrated in @fig-examplegraphs.
+While an exhaustive discussion of causal discovery concepts and algorithm-specific details is beyond the scope of this paper, readers seeking a deeper understanding are encouraged to consult @park2024discovering for comprehensive insights into these methods and their applications.
+
+[^4]: $*$ serves as a \textit{meta-symbol}, representing one of the three possible edge-endpoints.
+For instance, $A \tailstar B$ can indicate any of the following edges: $A$ \textemdash\ $B$, $A\tailarrow B$, or $A \tailcirc B$ [@park2024discovering].
+
+```{r}
+#| layout-ncol: 3
+#| label: fig-examplegraphs
+#| fig-cap: Example graph $\mathcal{G}$ featuring a cycle and the corresponding MAAG from CCI and PAG from FCI.
+#| fig-subcap: 
+#|   - Example graph $\mathcal{G}$
+#|   - Corresponding MAAG 
+#|   - Corrsponding PAG
+#| fig-align: center
+#| out-width: 60%
+
+knitr::include_graphics("img/original_DCG.png")
+knitr::include_graphics("img/FCI_graph.png")
+knitr::include_graphics("img/CCI_graph.png")
+```
+
+The graph produced by the PC algorithm is a *CPDAG* (completed partially directed acyclic graph), where directed edges ($A$ → $B$) indicate that $A$ is a direct cause (parent) of $B$.
 Unlike FCI and CCI, the CPDAG does not include circle symbols.
-Instead, when the PC algorithm cannot determine the direction, it represents uncertainty with bidirectional arrows.
-While PC serves as a useful reference, its strict assumptions of acyclicity and the absence of latent confounders limit its applicability in more complex settings.
-Therefore, our primary focus remains on the results from FCI and CCI.
-For completeness, all PC algorithm results are provided in the [Appendix](#sec-appendix).
+Instead, when the PC algorithm cannot determine the direction of causality, it represents this uncertainty with bidirectional arrows.
+While the PC algorithm serves as a useful reference, its strict assumptions of acyclicity and the absence of latent confounders restrict its applicability in more complex scenarios.
+Therefore, our primary focus remains on the results from FCI and CCI, with all PC algorithm results provided in the [Appendix](#sec-appendix) for completeness.
 
 One practical challenge in applying these algorithms to the HELIUS dataset is that the data does not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear.
-To account for this, we complement the commonly used Gaussian conditional independence test (CI test), which relies on partial correlations, with a non-parametric CI test based on kernel methods.
-However, kernel-based non-parametric tests are computationally demanding, particularly with large datasets like ours.
-To mitigate this issue, we employ Randomized Conditional Independence Test (RCIT) and the Randomized conditional Correlation Test (RCoT), which uses random Fourier features to approximate the kernel methods, thereby significantly reducing the computational cost [@strobl2019approximate].
-For a more detailed explanation of RCIT and RCoT, see @sec-rcot.
+To account for this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel].
+While kernel-based conditional independence tests (KCIT) are effective, they are computationally intensive for large datasets like HELIUS, as their quadratic scaling with sample size is primarily driven by the inversion of large kernel matrices.
+To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximate kernel methods using random Fourier features.
+This approach reduces computation to linear scaling with sample size, significantly lowering computational costs [@strobl2019approximate].
+For a more detailed explanation of RCoT, refer to @sec-rcot.
 
@@
 Using the PHQ sum score simplifies the analysis by reducing dimensionality, offering computational efficiency and a broad, interpretable perspective on the relationship between depression and precarity factors.
-In contrast, including individual symptoms provide a more detailed understanding, capturing the diverse ways symptoms respond to different precarity factors.
-However, this approach introduces methodological challenges, such as non-standard distributions and the complexities of increased dimensionality.
+In contrast, analyzing individual symptoms provides a more detailed understanding by capturing the diverse ways symptoms respond to different precarity factors.
+However, this detailed approach introduces methodological challenges, including non-standard distributions of individual symptom variables and the complexities associated with increased dimensionality.
 Finally, examining individual symptoms in relation to the overall precarity score complements the analysis by revealing how symptoms collectively connect to general precarity.
-By integrating these three approaches, we balance simplicity and granularity, ensuring robust findings.
+Aggregating precarity factors may uncover relationships that could be overlooked in analyses of individual factors, particularly when a symptom has associations across multiple factors that are not strong enough to emerge in individual analyses.
+By integrating these three approaches, we achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression. 
 <!-- The PHQ sum score and precarity sum score capture overarching trends between overall depression and precarity, while symptom-level analysis with individual precarity factors reveals detailed dynamics, essential for tailoring specific interventions.
@@
 
-To address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping.
+Furthermore, to address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping.
 For each of the 100 bootstrap samples, we estimate causal graphs and retain only the edges and directions that exceed predefined thresholds.
@@
 Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm.
-For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30 to ease computational demands.
-<!-- Additionally, we fix the skeleton of edges among symptom variables based on the common structure estimated across all three algorithms, thereby reducing the computational time required to estimate skeletons.
-This fixed structure aligns with commonly observed skeleton structures in the literature, ensuring relevance to existing findings (**cite the comp_model paper**).[^2]  -->
-
-To summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.[^3] This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression.
+For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30, easing computational demands.
+To further enhance efficiency, we fix the skeleton of edges among symptom variables based on the common structure estimated across all algorithms.
+This approach minimizes spurious edges in the symptom network and significantly reduces the computational time needed for skeleton estimation.
+This fixed structure also aligns well with commonly reported skeleton structures in the literature, ensuring consistency with existing findings (**cite the comp_model paper**).[^2]
+Lastly, to summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.[^3] This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression.
 See @fig-workflow for an overview of the analysis workflow.
 
-<!-- [^2]:  -->
-<!-- To reduce spurious edges in the symptom network, we fixed the skeleton of symptom interactions using a common structure derived from PC, FCI, and CCI algorithms with an alpha level of 0.001.
-This approach maintained sparsity in symptom interactions while allowing the free estimation of causal directions between symptoms and precarity factors, balancing computational efficiency with meaningful results.
--->
+[^2]:
+We fixed the skeleton of the symptom network using a common structure derived from the PC, FCI, and CCI algorithms with an alpha level of 0.001.
+This approach preserved the complexity of symptom interactions, enabling unrestricted estimation of causal directions both among symptoms (*symptom --- symptom* interactions) and between symptoms and precarity factors.
 
 [^3]:
-The detailed proportion of each edge endpoint occurrence is shown in [@sec-propmatrix].
+The detailed proportion of each edge endpoint occurrence is provided in [@sec-propmatrix].
 
@@
 #| label: fig-sum
-#| fig-cap: "Resulting graphs of precarity factors and depression sum score using FCI and CCI"
+#| fig-cap: "Resulting graphs of precarity factors and depression sum score using FCI and CCI."
 #| fig-subcap: 
@@
 #| label: fig-sym
-#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using FCI and CCI"
+#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using FCI and CCI."
 #| fig-align: center
@@
 ## Precarity as sum score
+
 Lastly, we examine the relationships between individual symptoms and overall precarity, represented as the sum of five precariousness factors.
@@
 #| label: fig-presum
-#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using FCI and CCI"
+#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using FCI and CCI."
 #| fig-align: center
@@
 #| label: fig-pc_sum
-#| fig-cap: "Resulting graphs of precarity factors and depression sum score using PC"
+#| fig-cap: "Resulting graphs of precarity factors and depression sum score using PC."
 #| fig-subcap:
@@
 #| label: fig-pc_sym
-#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC"
+#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC."
 #| fig-subcap:
@@
 #| label: fig-pc_presum
-#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using PC"
+#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using PC."
 #| fig-subcap:
@@
 #| label: fig-sum-mat
-#| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum score"
+#| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum score."
 #| fig-align: center
@@
 #| label: fig-sym-mat
-#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms"
+#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms."
 #| fig-align: center
@@
 #| label: fig-presum-mat
-#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precarity sum score"
+#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precarity sum score."
 #| fig-align: center

### 2025-01-03 draft_v1/draft_v1.tex
draft update_till_sec2(analysis)

@@
 \usepackage{makecell} % Add to your preamble for multi-line cells
+\usepackage{tikz}
+\usetikzlibrary{arrows.meta, shapes.geometric}
+
+%% Define TikZ arrow styles
+\newcommand{\starstar}{% 
+\begin{tikzpicture}[baseline=-3pt]
+    \draw [{Rays[n=6]}-{Rays[n=6]}] (0,0) -- (0.55,0);
+\end{tikzpicture}
+}
+
+\newcommand{\circstar}{% 
+\begin{tikzpicture}
+    \draw [{Circle[open]}-{Rays[n=6]}] (0,0) -- (0.55, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\starcirc}{% 
+\begin{tikzpicture}
+    \draw [{Rays[n=6]}-{Circle[open]}] (0,0) -- (0.55, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\stararrow}{% 
+\begin{tikzpicture}
+    \draw [{Rays[n=6]}-{Straight Barb[length=2.5pt]}] (0,0) -- (0.5, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\arrowstar}{% 
+\begin{tikzpicture}
+    \draw [{Straight Barb[length=2.5pt]}-{Rays[n=6]}] (0,0) -- (0.5, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\circarrow}{% 
+\begin{tikzpicture}
+    \draw [{Circle[open]}-{Straight Barb[length=2.5pt]}] (0,0) -- (0.5, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\tailcirc}{% 
+\begin{tikzpicture}[baseline=-3pt] 
+    \draw [-{Circle[open]}] (0,0) -- (0.4, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\circtail}{% 
+\begin{tikzpicture}[baseline=-3pt] 
+    \draw [{Circle[open]}-] (0,0) -- (0.4, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\circirc}{% 
+\begin{tikzpicture}[baseline=-3pt] 
+    \draw [{Circle[open]}-{Circle[open]}] (0,0) -- (0.5, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\tailstar}{% 
+\begin{tikzpicture}[baseline=-3pt] 
+    \draw [-{Rays[n=6]}] (0,0) -- (0.5, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\startail}{% 
+\begin{tikzpicture}
+    \draw [{Rays[n=6]}-] (0,0) -- (0.5, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\tailarrow}{% 
+\begin{tikzpicture}
+    \draw [-{Straight Barb[length=2.5pt]}](0,0) -- (0.4, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\arrowtail}{% 
+\begin{tikzpicture}
+    \draw [{Straight Barb[length=2.5pt]}-](0,0) -- (0.4, 0);
+\end{tikzpicture}
+}
+
+\newcommand{\arrowarrow}{% 
+\begin{tikzpicture}
+    \draw [{Straight Barb[length=2.5pt]}-{Straight Barb[length=2.5pt]}](0,0) -- (0.4, 0);
+\end{tikzpicture}
+}
+
+\newcommand\stackedarrows{%
+        \mathrel{\vcenter{\mathsurround0pt
+                \ialign{##\crcr
+                \noalign{\nointerlineskip}$\arrowtail$\crcr
+                \noalign{\nointerlineskip}$\tailarrow$\crcr
+                }
+        }}%
+}
 \makeatletter
@@
 
-\date{2024-12-30}
+\date{2025-01-03}
 \begin{document}
@@
 given its simplicity and prominence as one of the most widely known
-causal discovery methods (Spirtes et al., 2001). For a more detailed
-explanation of these algorithms, please refer to Park et al. (2024).
+causal discovery methods (Spirtes et al., 2001).
 
@@
 almost ancestral graph) due to their reliance on different underlying
-assumptions. Both encode information about causal relationships between
+assumptions. Despite these differences, both graphs are forms of
+\emph{ancestral graphs}, designed to encode causal relationships between
 variables, where the presence of an edge indicates causal
-\emph{ancestry}. Directed edges, \(A\) *→ \(B\), specify that \(B\) is
-not an ancestor of \(A\) in every graph within the Markov equivalence
-class, \(Equiv(G)\). \(A\)*---\(B\) represent cases where \(B\) is an
-ancestor of \(A\) in every graph in \(Equiv(G)\). Circle endpoints,
-\(A\)*-o\(B\), denote ambiguity in the ancestral relationship, meaning
-that \(B\)'s ancestral status relative to \(A\) varies across graphs in
-\(Equiv(G)\).
-
-In contrast, the graph produced by the PC algorithm is a \emph{CPDAG}
-(completed partially directed acyclic graph), where directed edges
-(\(A\) → \(B\)) indicate that \(A\) is a direct cause (parent) of \(B\).
-Unlike FCI and CCI, the CPDAG does not include circle symbols. Instead,
-when the PC algorithm cannot determine the direction, it represents
-uncertainty with bidirectional arrows. While PC serves as a useful
-reference, its strict assumptions of acyclicity and the absence of
-latent confounders limit its applicability in more complex settings.
-Therefore, our primary focus remains on the results from FCI and CCI.
-For completeness, all PC algorithm results are provided in the
-\hyperref[sec-appendix]{Appendix}.
+\emph{ancestry.} Directed edges, \(A \stararrow B\), indicate that \(B\)
+is not an ancestor of \(A\) in every graph within the Markov equivalence
+class, \(Equiv(G)\). The Markov equivalence class refers to the set of
+graphs that encode the same conditional independence relationships, such
+that the same \emph{d-separation} relations hold across all graphs in
+the class (Spirtes et al., 2001). \(A \startail B\) indicates that \(B\)
+is an ancestor of \(A\) across all graphs in \(Equiv(G)\). Circle
+endpoints, \(A \starcirc B\), represent ambiguity in the ancestral
+relationship, meaning \(B\)'s ancestral status relative to \(A\) varies
+across graphs in \(Equiv(G)\).\footnote{\(*\) serves as a
+  \textit{meta-symbol}, representing one of the three possible
+  edge-endpoints. For instance, \(A \tailstar B\) can indicate any of
+  the following edges: \(A\) \textemdash~\(B\), \(A\tailarrow B\), or
+  \(A \tailcirc B\) (Park et al., 2024).} When \(A \arrowarrow B\),
+neither \(A\) nor \(B\) is an ancestor of the other, indicating the
+presence of a latent confounder between them.
+
+The algorithms also differ in their approaches and assumptions for
+detecting cycles. In CCI's MAAG, \(A\) \textemdash~\(B\) indicates that
+\(A\) is an ancestor of \(B\) and \(B\) is an ancestor of \(A\),
+referring to a cycle between \(A \stackedarrows B\). FCI, on the other
+hand, identifies potential cycles more subtly: fully-connected nodes
+with circle endpoints (\(\circirc\)) in FCI's PAG may suggest a cyclic
+structure. FCI basically provides a sufficient condition to distinguish
+nodes that are not part of a cycle, offering a more nuanced approach
+(Mooij \& Claassen, 2020). Table 1 highlights further differences in the
+assumptions underlying these algorithms. CCI operates under the
+assumption of a linear system, while FCI, particularly for inferring
+cyclic relationships, assumes a non-linear system without selection bias
+and adheres to more general faithfulness and Markov conditions (i.e.,
+\emph{\(\sigma\)-separation} and \emph{\(\sigma\)-faithfulness} setting)
+(Forré \& Mooij, 2018). When the respective assumptions of each
+algorithm are satisfied, their outcomes for cyclic relationships align
+with those illustrated in Figure~\ref{fig-examplegraphs}. While an
+exhaustive discussion of causal discovery concepts and
+algorithm-specific details is beyond the scope of this paper, readers
+seeking a deeper understanding are encouraged to consult Park et al.
+(2024) for comprehensive insights into these methods and their
+applications.
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/original_DCG.png}
+
+}
+
+\subcaption{\label{fig-examplegraphs-1}Example graph \(\mathcal{G}\)}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/FCI_graph.png}
+
+}
+
+\subcaption{\label{fig-examplegraphs-2}Corresponding MAAG}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/CCI_graph.png}
+
+}
+
+\subcaption{\label{fig-examplegraphs-3}Corrsponding PAG}
+
+\end{minipage}%
+
+\caption{\label{fig-examplegraphs}Example graph \(\mathcal{G}\)
+featuring a cycle and the corresponding MAAG from CCI and PAG from FCI.}
+
+\end{figure}%
+
+The graph produced by the PC algorithm is a \emph{CPDAG} (completed
+partially directed acyclic graph), where directed edges (\(A\) → \(B\))
+indicate that \(A\) is a direct cause (parent) of \(B\). Unlike FCI and
+CCI, the CPDAG does not include circle symbols. Instead, when the PC
+algorithm cannot determine the direction of causality, it represents
+this uncertainty with bidirectional arrows. While the PC algorithm
+serves as a useful reference, its strict assumptions of acyclicity and
+the absence of latent confounders restrict its applicability in more
+complex scenarios. Therefore, our primary focus remains on the results
+from FCI and CCI, with all PC algorithm results provided in the
+\hyperref[sec-appendix]{Appendix} for completeness.
 
@@
 the relationships among variables are unlikely to be strictly linear. To
-account for this, we complement the commonly used Gaussian conditional
-independence test (CI test), which relies on partial correlations, with
-a non-parametric CI test based on kernel methods. However, kernel-based
-non-parametric tests are computationally demanding, particularly with
-large datasets like ours. To mitigate this issue, we employ Randomized
-Conditional Independence Test (RCIT) and the Randomized conditional
-Correlation Test (RCoT), which uses random Fourier features to
-approximate the kernel methods, thereby significantly reducing the
-computational cost (Strobl et al., 2019). For a more detailed
-explanation of RCIT and RCoT, see Section~\ref{sec-rcot}.
+account for this, we supplement the commonly used Gaussian conditional
+independence (CI) test, which relies on partial correlations, with a
+non-parametric CI test based on reproducing kernel methods (Zhang et
+al., 2012). While kernel-based conditional independence tests (KCIT) are
+effective, they are computationally intensive for large datasets like
+HELIUS, as their quadratic scaling with sample size is primarily driven
+by the inversion of large kernel matrices. To mitigate this issue, we
+use the Randomized Conditional Correlation Test (RCoT), which
+approximate kernel methods using random Fourier features. This approach
+reduces computation to linear scaling with sample size, significantly
+lowering computational costs (Strobl et al., 2019). For a more detailed
+explanation of RCoT, refer to Section~\ref{sec-rcot}.
 
@@
 interpretable perspective on the relationship between depression and
-precarity factors. In contrast, including individual symptoms provide a
-more detailed understanding, capturing the diverse ways symptoms respond
-to different precarity factors. However, this approach introduces
-methodological challenges, such as non-standard distributions and the
-complexities of increased dimensionality. Finally, examining individual
+precarity factors. In contrast, analyzing individual symptoms provides a
+more detailed understanding by capturing the diverse ways symptoms
+respond to different precarity factors. However, this detailed approach
+introduces methodological challenges, including non-standard
+distributions of individual symptom variables and the complexities
+associated with increased dimensionality. Finally, examining individual
 symptoms in relation to the overall precarity score complements the
 analysis by revealing how symptoms collectively connect to general
-precarity. By integrating these three approaches, we balance simplicity
-and granularity, ensuring robust findings.
-
-To address sensitivity to parameter values and enhance the robustness of
-our results, we implement multiple settings combined with bootstrapping.
-For each of the 100 bootstrap samples, we estimate causal graphs and
-retain only the edges and directions that exceed predefined thresholds.
-The analyses are conducted under the following conditions:
+precarity. Aggregating precarity factors may uncover relationships that
+could be overlooked in analyses of individual factors, particularly when
+a symptom has associations across multiple factors that are not strong
+enough to emerge in individual analyses. By integrating these three
+approaches, we achieve a balance between capturing detailed granularity
+and gaining a broader understanding of the relationship between
+precarity and depression.
+
+Furthermore, to address sensitivity to parameter values and enhance the
+robustness of our results, we implement multiple settings combined with
+bootstrapping. For each of the 100 bootstrap samples, we estimate causal
+graphs and retain only the edges and directions that exceed predefined
+thresholds. The analyses are conducted under the following conditions:
 
@@
 simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing
-the number of bootstrap samples to 30 to ease computational demands.
-
-To summarize the results, we identify the most frequently occurring edge
-endpoints across different experimental setups.\footnote{The detailed
-  proportion of each edge endpoint occurrence is shown in
-  Section~\ref{sec-propmatrix}.} This ensures that only stable and
-consistent edges are retained, providing a clearer and more reliable
-understanding of the relationships between precarity factors and
-depression. See Figure~\ref{fig-workflow} for an overview of the
-analysis workflow.
+the number of bootstrap samples to 30, easing computational demands. To
+further enhance efficiency, we fix the skeleton of edges among symptom
+variables based on the common structure estimated across all algorithms.
+This approach minimizes spurious edges in the symptom network and
+significantly reduces the computational time needed for skeleton
+estimation. This fixed structure also aligns well with commonly reported
+skeleton structures in the literature, ensuring consistency with
+existing findings (\textbf{cite the comp\_model paper}).\footnote{We
+  fixed the skeleton of the symptom network using a common structure
+  derived from the PC, FCI, and CCI algorithms with an alpha level of
+  0.001. This approach preserved the complexity of symptom interactions,
+  enabling unrestricted estimation of causal directions both among
+  symptoms (\emph{symptom --- symptom} interactions) and between
+  symptoms and precarity factors.} Lastly, to summarize the results, we
+identify the most frequently occurring edge endpoints across different
+experimental setups.\footnote{The detailed proportion of each edge
+  endpoint occurrence is provided in Section~\ref{sec-propmatrix}.} This
+ensures that only stable and consistent edges are retained, providing a
+clearer and more reliable understanding of the relationships between
+precarity factors and depression. See Figure~\ref{fig-workflow} for an
+overview of the analysis workflow.
 
@@
 \caption{\label{fig-sum}Resulting graphs of precarity factors and
-depression sum score using FCI and CCI}
+depression sum score using FCI and CCI.}
 
@@
 \caption{\label{fig-sym}Resulting graphs of precarity factors and
-individual depression symptoms using FCI and CCI}
+individual depression symptoms using FCI and CCI.}
 
@@
 \caption{\label{fig-presum}Resulting graphs of precarity sum score and
-individual depression symptoms using FCI and CCI}
+individual depression symptoms using FCI and CCI.}
 
@@
 
+\bibitem[\citeproctext]{ref-forre2018constraint}
+Forré, P., \& Mooij, J. M. (2018). Constraint-based causal discovery for
+non-linear structural causal models with cycles and latent confounders.
+\emph{arXiv Preprint arXiv:1807.03024}.
+
 \bibitem[\citeproctext]{ref-hyttinen2012learning}
@@
 
+\bibitem[\citeproctext]{ref-spirtes_causal_1995}
+Spirtes, P., Meek, C., \& Richardson, T. (1995). Causal inference in the
+presence of latent variables and selection bias. \emph{Proceedings of
+the {Eleventh} Conference on {Uncertainty} in Artificial Intelligence},
+499--506.
+
 \bibitem[\citeproctext]{ref-strobl2019}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-8-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-9-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-9-1.pdf}
+\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-10-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-11-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-12-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-12-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-13-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-14-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-15-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-15-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-15-2.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-2.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-15-3.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-3.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-15-4.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-4.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-16-1.pdf}
+\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-17-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-18-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-1.pdf}
 \end{center}
@@
 \caption{\label{fig-pc_sum}Resulting graphs of precarity factors and
-depression sum score using PC}
+depression sum score using PC.}
 
@@
 \caption{\label{fig-pc_sym}Resulting graphs of precarity factors and
-individual depression symptoms using PC}
+individual depression symptoms using PC.}
 
@@
 \caption{\label{fig-pc_presum}Resulting graphs of precarity sum score
-and individual depression symptoms using PC}
+and individual depression symptoms using PC.}
 
@@
 \caption{\label{fig-sum-mat}Proportions of edge endpoint types for
-graphs based on depression sum score}
+graphs based on depression sum score.}
 
@@
 \caption{\label{fig-sym-mat}Proportions of edge endpoint types for
-graphs based on individual depression symptoms}
+graphs based on individual depression symptoms.}
 
@@
 \caption{\label{fig-presum-mat}Proportions of edge endpoint types for
-graphs based on individual depression symptoms with precarity sum score}
+graphs based on individual depression symptoms with precarity sum
+score.}
 

### 2025-01-04 draft_v1/draft_v1.qmd
update_discussion_limitations

@@
 @fig-sum illustrates the causal relationships between precarity factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*) under two different setups: (a) using both Gaussian CI test and RCoT, and (b) using RCoT alone.
-Black edges represent consistent causal relationships identified by both FCI and CCI, while gray edges denote inconsistent relationships that vary between the two methods.
-The edge endpoints of inconsistent gray edges are marked with circles.
+Black edges represent consistent causal relationships identified by both FCI and CCI, while gray dashed edges denote inconsistent relationships that vary between the two methods.
+The inconsistent endpoints of the gray edges are marked with circles.
+
 In graph (a), the key pathways suggest that employment precarity (*P.emp*) and social precarity (*P.soc*) do not cause depression (*PHQsum*).
-Social precarity does not cause recent relational stress (*S.rel*) or financial stress (*S.fin*), and these stressors are likely related by a latent confounder.
+*P.soc* does not cause recent relational stress (*S.rel*) or financial stress (*S.fin*), and these stressors are likely related by a latent confounder.
 Additionally, *P.emp* and *S.fin* are identified as non-causes of housing precarity (*P.hou*).
-Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or if the connection is mediated by an unmeasured confounder.
-A similar ambiguity exists in the relationships between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*, with the methods diverging on whether these stressors influence depression or are linked through latent variables.
-
-Graph (b), derived solely from the nonparametric RCoT test.
-Both graphs consistently identify that employment precarity is not a cause of housing precarity or depression, and that social precarity does not cause depression or financial stress.
+Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship.
+Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*.
+
+Graph (b), derived solely from the nonparametric RCoT method.
+Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*.
 The relationship between stressors and depression remains unresolved between the two algorithms.
-However, RCoT provides greater confidence that depression is not the cause of financial stress and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
+However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
 The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.
 
-Overall, both graphs together highlight the potential roles of recent stressors as being closely linked with depression, either as causes or through latent confounders.
+Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders.
 Employment and social precarity may also be influenced by depression, either directly or through latent variables.
-Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, either directly or through a latent confounder.
+Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways.
 
@@
 
-The present study explored the causal relationships between precarity factors and depression by employing both sum score and symptom-level analyses using causal discovery algorithms.
-The use of both PHQ sum scores and individual symptom scores provided a comprehensive understanding of how different aspects of precarity influence depression, revealing nuanced pathways that may be obscured when considering aggregate measures alone.
+The present study examined the causal relationships between precarity factors and depression using causal discovery algorithms, incorporating both sum score and symptom-level analyses.
+By analyzing PHQ sum scores alongside individual symptom scores, and comparing aggregated precarity scores with individual precarity factors, we gained a comprehensive perspective on how different aspects of precarity influence depression.
+This dual approach highlighted nuanced pathways that may be overlooked when focusing solely on aggregate measures, while also capturing broader interactions that might remain undetected in analyses limited to individual factors.
 
@@
 The prevalence of bidirectional arrows and circle-marked endpoints in the graphs reflects unresolved ambiguities in the causal relationships suggested by the data.
-These uncertainties underscore the need for further research, ideally incorporating datasets with a higher proportion of symptomatic individuals.
-The HELIUS dataset, being predominantly composed of asymptomatic samples, posed challenges in identifying clear causal directions, particularly among symptom nodes.
-Future studies could address these limitations by incorporating time-series data to leverage temporal information about the relationships between precarity factors and depressive symptoms.
-Time-series data could provide time-specific insights and track how these relationships evolve over time.
-For instance, methods such as *PCMCI* [@runge2019detecting] and *tsFCI* [@entner2010causal], along with other time-series adaptations of causal discovery algorithms, could help better account for temporal dependencies and refine the analysis.
-
-Another limitation is the lack of clear evidence for cycles within the symptom network, despite using algorithms designed to account for cyclic relationships.
-The CCI algorithm predominantly produced bidirectional arrows, while FCI primarily generated directional arrows, yet neither displayed patterns indicative of definitive cyclic structures.
-Addressing cyclic relationships is particularly challenging with observational datasets alone.
-Future research could benefit from refined datasets that include intervention data.
-Methods such as *LLC* [@hyttinen2012learning], *NODGAS-Flow* [@sethuraman2023nodags], and the recently developed *Bicycle* algorithm [@rohbeck2024] are specifically designed to utilize both observational and intervention data to uncover potential cycles.
-Additionally, if time-series data becomes available, corresponding methods, as described above, could be applied to capture repetitive patterns in variable interactions that might suggest cyclic structures.
+These uncertainties highlight the need for further research, particularly with higher-resolution datasets, such as time-series data.
+Incorporating time-series data could capture temporal dynamics between precarity factors and depressive symptoms, providing time-specific insights and revealing how these relationships evolve over time.
+Methods like *PCMCI* [@runge2019detecting], *tsFCI* [@entner2010causal], and other time-series adaptations of causal discovery algorithms could offer a more robust framework for addressing temporal dependencies and refining the analysis.
+
+Another notable limitation is the absence of clear evidence for cycles within the symptom network, despite employing algorithms designed to account for cyclic relationships.
+The CCI algorithm primarily produced bidirectional arrows, while FCI predominantly generated directional arrows, yet neither demonstrated patterns indicative of definitive cyclic structures.
+This suggests that addressing cyclic relationships is particularly challenging when relying solely on observational datasets.
+Future research could benefit from refined datasets that incorporate intervention data.
+Methods such as *LLC* [@hyttinen2012learning], *NODGAS-Flow *[@sethuraman2023nodags], and the recently developed *Bicycle* algorithm [@rohbeck2024], which utilize both observational and intervention data, may uncover potential cycles more effectively.
 
@@
 A possible explanation for this discrepancy is that Gaussian CI's reliance on partial correlations may overestimate relationships when specific non-linear dependencies exist in the data.
-In contrast, RCoT, free from such assumptions, may better capture these patterns under such conditions.
-However, while RCoT is technically non-parametric, its performance can still be influenced by the distributional characteristics of the data.
-Specifically, the RBF kernel, optimized for continuous data with smooth transitions, may struggle to capture relationships in datasets with discrete or mixed distributions.
+In contrast, RCoT, free from such assumptions, might better capture independence patterns under such conditions.
+However, RCoT is not without its limitations.
+While technically non-parametric, its performance can be influenced by the distributional characteristics of the data.
+The Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle with discrete or mixed datasets, where distances between discrete points may fail to convey meaningful similarity [@howlett2001radial].
+This limitation could explain why RCoT might miss certain dependencies, particularly in the HELIUS dataset, where some variables lack smooth continuity.
+Future research could tackle these challenges by developing hybrid kernels designed for mixed datasets, integrating both continuous and discrete variables into RCoT-like methods.
+This would have the potential to improve the reliability of findings, especially when working with datasets characterized by complex and heterogeneous structures.
+
+<!-- However, it could be also that RCoT performed somewhat suboptimal.
+while RCoT is technically non-parametric, its performance can still be influenced by the distributional characteristics of the data.
+Specifically, the Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle to capture relationships in datasets with discrete or mixed distributions.
 In such cases, the distances between discrete points may fail to convey meaningful similarity information.
-As a result, RCoT might miss certain dependencies, particularly when variables in the dataset lack smooth continuity.
-Future research could address these issues by exploring a broader range of CI testing approaches.
+As a result, RCoT might miss certain dependencies, particularly when variables in the dataset lack smooth continuity like in our HELIUS dataset.
+-->
+
+<!-- Future research could address these issues by exploring a broader range of CI testing approaches.
 One traditional approach to handle this is discretizing variables and use $G^2$ test, which may better capture dependencies in non-continuous datasets [@dojer2016learning; @neapolitan2004learning].
 A more promising direction, however, lies in the development of hybrid kernels tailored for mixed datasets, effectively integrating both continuous and discrete variables into RCoT-like methods.
-By systematically employing and comparing a wider variety of CI testing techniques, researchers could gain more robust insights and mitigate the limitations inherent in specific approaches.
+-->
+<!-- By systematically employing and comparing a wider variety of CI testing techniques, researchers could gain more robust insights and mitigate the limitations inherent in specific approaches.
 This broader exploration holds the potential to enhance the reliability of findings, particularly in datasets with complex and heterogeneous structures.
+-->
 
@@
 While the resulting graphs are preliminary and contain unresolved ambiguities, they offer a valuable starting point for leveraging causal discovery tools to investigate the causal interplay between depression and precarity factors.
-A promising next step would involve integrating these causal structures into computational models, such as the symptom dynamic model proposed by **our comp-model paper**.
+A promising next step would involve integrating these causal structures into computational models, such as the symptom dynamic model proposed in **our comp-model paper**.
 By simulating intervention effects, such models could provide more realistic insights into how targeted actions might influence symptom networks and precarity factors over time.

### 2025-01-04 draft_v1/draft_v1.tex
update_discussion_limitations

@@
 
-\date{2025-01-03}
+\date{2025-01-04}
 \begin{document}
@@
 consistent causal relationships identified by both FCI and CCI, while
-gray edges denote inconsistent relationships that vary between the two
-methods. The edge endpoints of inconsistent gray edges are marked with
-circles. In graph (a), the key pathways suggest that employment
-precarity (\emph{P.emp}) and social precarity (\emph{P.soc}) do not
-cause depression (\emph{PHQsum}). Social precarity does not cause recent
+gray dashed edges denote inconsistent relationships that vary between
+the two methods. The inconsistent endpoints of the gray edges are marked
+with circles.
+
+In graph (a), the key pathways suggest that employment precarity
+(\emph{P.emp}) and social precarity (\emph{P.soc}) do not cause
+depression (\emph{PHQsum}). \emph{P.soc} does not cause recent
 relational stress (\emph{S.rel}) or financial stress (\emph{S.fin}), and
@@
 relationship, leaving it unclear whether \emph{S.fin} causes
-\emph{P.emp} or if the connection is mediated by an unmeasured
-confounder. A similar ambiguity exists in the relationships between
-\emph{S.rel} and \emph{PHQsum} and between \emph{S.fin} and
-\emph{PHQsum}, with the methods diverging on whether these stressors
-influence depression or are linked through latent variables.
-
-Graph (b), derived solely from the nonparametric RCoT test. Both graphs
-consistently identify that employment precarity is not a cause of
-housing precarity or depression, and that social precarity does not
-cause depression or financial stress. The relationship between stressors
-and depression remains unresolved between the two algorithms. However,
-RCoT provides greater confidence that depression is not the cause of
-financial stress and suggests a stronger likelihood that the financial
-stress may contribute to depression or that their relationship is
-mediated by a latent confounder. The role of relational stress has
-become less pronounced, as the edge between \emph{P.soc} and
+\emph{P.emp} or whether a latent confounder mediates their relationship.
+Similar ambiguities are found in the connections between \emph{S.rel}
+and \emph{PHQsum} and between \emph{S.fin} and \emph{PHQsum}.
+
+Graph (b), derived solely from the nonparametric RCoT method. Both
+graphs consistently identify that \emph{P.emp} is not a cause of
+\emph{P.hou} or \emph{PHQsum}, and that \emph{P.soc} does not cause
+\emph{PHQsum} or \emph{S.fin}. The relationship between stressors and
+depression remains unresolved between the two algorithms. However, the
+graph derived from RCoT provides greater confidence that depression is
+not the cause of \emph{S.fin} and suggests a stronger likelihood that
+the financial stress may contribute to depression or that their
+relationship is mediated by a latent confounder. The role of relational
+stress has become less pronounced, as the edge between \emph{P.soc} and
 \emph{S.rel} is omitted, and the relationship between \emph{S.rel} and
@@
 
-Overall, both graphs together highlight the potential roles of recent
-stressors as being closely linked with depression, either as causes or
-through latent confounders. Employment and social precarity may also be
-influenced by depression, either directly or through latent variables.
-Lastly, housing precarity does not directly impact depression but
-remains connected to employment precarity and financial stress, either
-directly or through a latent confounder.
+Overall, both graphs together highlight the potential roles of stressors
+like \emph{S.fin} and \emph{S.rel}, which are closely tied to
+depression, either causally or via latent confounders. Employment and
+social precarity may also be influenced by depression, either directly
+or through latent variables. Lastly, housing precarity does not directly
+impact depression but remains connected to employment precarity and
+financial stress, likely mediated by direct or latent pathways.
 
@@
 
-The present study explored the causal relationships between precarity
-factors and depression by employing both sum score and symptom-level
-analyses using causal discovery algorithms. The use of both PHQ sum
-scores and individual symptom scores provided a comprehensive
-understanding of how different aspects of precarity influence
-depression, revealing nuanced pathways that may be obscured when
-considering aggregate measures alone.
+The present study examined the causal relationships between precarity
+factors and depression using causal discovery algorithms, incorporating
+both sum score and symptom-level analyses. By analyzing PHQ sum scores
+alongside individual symptom scores, and comparing aggregated precarity
+scores with individual precarity factors, we gained a comprehensive
+perspective on how different aspects of precarity influence depression.
+This dual approach highlighted nuanced pathways that may be overlooked
+when focusing solely on aggregate measures, while also capturing broader
+interactions that might remain undetected in analyses limited to
+individual factors.
 
@@
 endpoints in the graphs reflects unresolved ambiguities in the causal
-relationships suggested by the data. These uncertainties underscore the
-need for further research, ideally incorporating datasets with a higher
-proportion of symptomatic individuals. The HELIUS dataset, being
-predominantly composed of asymptomatic samples, posed challenges in
-identifying clear causal directions, particularly among symptom nodes.
-Future studies could address these limitations by incorporating
-time-series data to leverage temporal information about the
-relationships between precarity factors and depressive symptoms.
-Time-series data could provide time-specific insights and track how
-these relationships evolve over time. For instance, methods such as
-\emph{PCMCI} (Runge et al., 2019) and \emph{tsFCI} (Entner \& Hoyer,
-2010), along with other time-series adaptations of causal discovery
-algorithms, could help better account for temporal dependencies and
-refine the analysis.
-
-Another limitation is the lack of clear evidence for cycles within the
-symptom network, despite using algorithms designed to account for cyclic
-relationships. The CCI algorithm predominantly produced bidirectional
-arrows, while FCI primarily generated directional arrows, yet neither
-displayed patterns indicative of definitive cyclic structures.
-Addressing cyclic relationships is particularly challenging with
-observational datasets alone. Future research could benefit from refined
-datasets that include intervention data. Methods such as \emph{LLC}
-(Hyttinen et al., 2012), \emph{NODGAS-Flow} (Sethuraman et al., 2023),
-and the recently developed \emph{Bicycle} algorithm (Rohbeck et al.,
-2024) are specifically designed to utilize both observational and
-intervention data to uncover potential cycles. Additionally, if
-time-series data becomes available, corresponding methods, as described
-above, could be applied to capture repetitive patterns in variable
-interactions that might suggest cyclic structures.
+relationships suggested by the data. These uncertainties highlight the
+need for further research, particularly with higher-resolution datasets,
+such as time-series data. Incorporating time-series data could capture
+temporal dynamics between precarity factors and depressive symptoms,
+providing time-specific insights and revealing how these relationships
+evolve over time. Methods like \emph{PCMCI} (Runge et al., 2019),
+\emph{tsFCI} (Entner \& Hoyer, 2010), and other time-series adaptations
+of causal discovery algorithms could offer a more robust framework for
+addressing temporal dependencies and refining the analysis.
+
+Another notable limitation is the absence of clear evidence for cycles
+within the symptom network, despite employing algorithms designed to
+account for cyclic relationships. The CCI algorithm primarily produced
+bidirectional arrows, while FCI predominantly generated directional
+arrows, yet neither demonstrated patterns indicative of definitive
+cyclic structures. This suggests that addressing cyclic relationships is
+particularly challenging when relying solely on observational datasets.
+Future research could benefit from refined datasets that incorporate
+intervention data. Methods such as \emph{LLC} (Hyttinen et al., 2012),
+\emph{NODGAS-Flow }(Sethuraman et al., 2023), and the recently developed
+\emph{Bicycle} algorithm (Rohbeck et al., 2024), which utilize both
+observational and intervention data, may uncover potential cycles more
+effectively.
 
@@
 overestimate relationships when specific non-linear dependencies exist
-in the data. In contrast, RCoT, free from such assumptions, may better
-capture these patterns under such conditions. However, while RCoT is
-technically non-parametric, its performance can still be influenced by
-the distributional characteristics of the data. Specifically, the RBF
-kernel, optimized for continuous data with smooth transitions, may
-struggle to capture relationships in datasets with discrete or mixed
-distributions. In such cases, the distances between discrete points may
-fail to convey meaningful similarity information. As a result, RCoT
-might miss certain dependencies, particularly when variables in the
-dataset lack smooth continuity. Future research could address these
-issues by exploring a broader range of CI testing approaches. One
-traditional approach to handle this is discretizing variables and use
-\(G^2\) test, which may better capture dependencies in non-continuous
-datasets (Dojer, 2016; Neapolitan et al., 2004). A more promising
-direction, however, lies in the development of hybrid kernels tailored
-for mixed datasets, effectively integrating both continuous and discrete
-variables into RCoT-like methods. By systematically employing and
-comparing a wider variety of CI testing techniques, researchers could
-gain more robust insights and mitigate the limitations inherent in
-specific approaches. This broader exploration holds the potential to
-enhance the reliability of findings, particularly in datasets with
-complex and heterogeneous structures.
+in the data. In contrast, RCoT, free from such assumptions, might better
+capture independence patterns under such conditions. However, RCoT is
+not without its limitations. While technically non-parametric, its
+performance can be influenced by the distributional characteristics of
+the data. The Gaussian RBF (radial basis function) kernel, optimized for
+continuous data with smooth transitions, may struggle with discrete or
+mixed datasets, where distances between discrete points may fail to
+convey meaningful similarity (Howlett, 2001). This limitation could
+explain why RCoT might miss certain dependencies, particularly in the
+HELIUS dataset, where some variables lack smooth continuity. Future
+research could tackle these challenges by developing hybrid kernels
+designed for mixed datasets, integrating both continuous and discrete
+variables into RCoT-like methods. This would have the potential to
+improve the reliability of findings, especially when working with
+datasets characterized by complex and heterogeneous structures.
 
@@
 next step would involve integrating these causal structures into
-computational models, such as the symptom dynamic model proposed by
+computational models, such as the symptom dynamic model proposed in
 \textbf{our comp-model paper}. By simulating intervention effects, such
@@
 \emph{arXiv Preprint arXiv:1807.03024}.
+
+\bibitem[\citeproctext]{ref-howlett2001radial}
+Howlett, R. J. (2001). \emph{Radial basis function networks 1: Recent
+developments in theory and applications}.
 

### 2025-01-12 code/graphviz/presum_peralgo.py
presum_graph_update

@@ -1,0 +1,136 @@
+from graphviz import Digraph
+
+# Define fixed positions for the nodes
+fixed_positions = {
+    "anh": (4, 3),
+    "dep": (2, 3.2),
+    "slp": (2, 1),
+    "ene": (3, 4),
+    "app": (1, 4),
+    "glt": (2.5, 2),
+    "con": (0, 2.7),
+    "mot": (0.8, 2),
+    "sui": (0, 1),
+    "Precarity": (3.5, 0.5)
+}
+
+# Create the directed graph
+dot = Digraph(format='png', engine='neato')
+dot.attr(size="14,15", dpi="300", splines="true", ratio="fill")
+dot.attr(label="", labelloc="top", fontsize="20", fontname="Palatino")
+
+# Add nodes with fixed positions
+for node, pos in fixed_positions.items():
+    dot.node(node, node, shape="circle", pos=f"{pos[0]},{pos[1]}!",
+             fixedsize="true", width="1", height="1", fontsize="25")
+
+
+# Manually define edges based on the adj.matrices
+# FCI
+dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="odot")
+dot.edge("anh", "slp", dir="both", arrowhead="normal", arrowtail="odot")
+dot.edge("anh", "ene", dir="both", arrowhead="normal", arrowtail="normal")
+dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="odot")
+dot.edge("anh", "glt", dir="both", arrowhead="normal", arrowtail="normal")
+dot.edge("anh", "con", dir="both", arrowhead="normal", arrowtail="odot")
+dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot")
+
+dot.edge("dep", "slp", dir="both", arrowhead="odot", arrowtail="normal")
+dot.edge("dep", "ene", dir="both", arrowhead="odot", arrowtail="normal")
+dot.edge("dep", "app", dir="both", arrowhead="odot", arrowtail="normal")
+dot.edge("dep", "glt", dir="both", arrowhead="odot", arrowtail="normal")
+dot.edge("dep", "con", dir="both", arrowhead="odot", arrowtail="odot")
+dot.edge("dep", "mot", dir="both", arrowhead="odot", arrowtail="normal")
+dot.edge("dep", "sui", dir="both", arrowhead="normal", arrowtail="normal")
[…]

### 2025-01-12 code/graphviz/symptom_presum_graph.py

@@ -26,22 +26,22 @@ for node, pos in fixed_positions.items():
 
 
 # Manually define edges based on the adj.matrices
-dot.edge("anh", "dep", dir="both", arrowhead="odot", arrowtail="normal", style="dashed", color="#a9a9a9")
-dot.edge("anh", "slp", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
-dot.edge("anh", "ene", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
-dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("anh", "con", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
+dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+dot.edge("anh", "slp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+dot.edge("anh", "ene", dir="both", arrowhead="normal", arrowtail="normal")
+dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+dot.edge("anh", "glt", dir="both", arrowhead="normal", arrowtail="normal")
+dot.edge("anh", "con", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
 dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
 
-dot.edge("dep", "slp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
-dot.edge("dep", "ene", dir="both", arrowhead="normal", arrowtail="normal")
-
-dot.edge("dep", "app", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
-dot.edge("dep", "glt", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
-dot.edge("dep", "con", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
-dot.edge("dep", "mot", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
[…]

### 2025-01-12 code/post_analysis.R

@@ -60,7 +60,10 @@ grouped_stable_edges <- split(stable_edges_data, comb$algorithm)
 
 # Group matrices by CItest
 grouped_stable_edges_ci <- split(stable_edges_data, comb$citest)
-
+## fci gauss
+fci_gauss_matrices <- grouped_stable_edges_ci$gaussCItest[grep("FCI", names(grouped_stable_edges_ci$gaussCItest))] 
+## cci gauss
+cci_gauss_matrices <- grouped_stable_edges_ci$gaussCItest[grep("CCI", names(grouped_stable_edges_ci$gaussCItest))]
 
 # summarize them by taking the most frequent symbol per element in matrices
 summarized_stable_edges <- lapply(grouped_stable_edges, function(group) {
@@ -84,6 +87,43 @@ summarized_stable_edges <- lapply(grouped_stable_edges, function(group) {
   }
   return(summarized)  # Return the summarized matrix for this group
 })
+
+
+## for matrices
+# Function to summarize matrices by finding the most frequent element at each position
+summarize_matrices <- function(group) {
+  # Get the dimensions from the first matrix in the group (assuming all matrices have the same dimensions)
+  nrow <- nrow(group[[1]])
+  ncol <- ncol(group[[1]])
+  
+  # Create a matrix to store the most frequent element for each position
+  summarized <- matrix(NA, nrow = nrow, ncol = ncol, 
+                       dimnames = dimnames(group[[1]]))  # Preserve row and column names
+  
+  # Iterate through each cell in the matrices
+  for (i in 1:nrow) {
+    for (j in 1:ncol) {
+      # Collect all values for the same position (i, j) across all matrices in the group
+      elements <- sapply(group, function(mat) mat[i, j])
+      
+      # Find the most frequent value for this position
+      most_frequent <- as.numeric(names(sort(table(elements), decreasing = TRUE))[1])
+      
+      # Store the most frequent value at position (i, j) in the summarized matrix
+      summarized[i, j] <- most_frequent
+    }
+  }
+  
+  # Return the summarized matrix for this group
+  return(summarized)
+}
+
+## compare the gaussian only case with the ones with both RCoT and gaussianci
+summarize_matrices(fci_gauss_matrices)
+summarized_stable_edges$FCI
+
+summarize_matrices(cci_gauss_matrices)
+summarized_stable_edges$CCI 
 
 
 ## Plot the resulting graph

### 2025-01-12 code/snellius/precarity_sumscore.R

@@ -171,7 +171,7 @@ message("Number of parallel workers: ", nbrOfWorkers())
 alphas = c(0.01, 0.05)
 thresholds = c(0.6, 0.7)
 citests = c("gaussCItest", "RCoT")
-algorithms = c("FCI", "CCI") #c("PC", "FCI", "CCI")
+algorithms = c("PC", "FCI", "CCI")
 data <- sumpre_sym # sumscore 
 subsample_size <- nrow(data)  # Size of each subsample
 num_subsamples <- 30     # Number of subsamples

### 2025-01-12 code/snellius/sym_gaussian.R

@@ -24,14 +24,17 @@ source("helius/code/preprocess_dat.R")
 # Causal Discovery: Define Parameters and Run Analyses
 # ==============================================================================
 # Set parameters
-args <- commandArgs(trailingOnly = TRUE)
-
-alpha <- as.numeric(args[1]) # Significance level 
-threshold <- as.numeric(args[2]) # Threshold for stable edges 
-
-data <- depsym
-subsample_size <- nrow(data) #10000  # Size of each subsample
-num_subsamples <- 100    # Number of subsamples
+# args <- commandArgs(trailingOnly = TRUE)
+# 
+# alpha <- as.numeric(args[1]) # Significance level 
+# threshold <- as.numeric(args[2]) # Threshold for stable edges 
+alphas = c(0.01, 0.05)
+thresholds = c(0.6, 0.7)
+citests = "gaussCItest"
+algorithms = c("PC", "FCI", "CCI")
+data <- dep_sym # sumscore 
+subsample_size <- nrow(data)  # Size of each subsample
+num_subsamples <- 30     # Number of subsamples
 
 
 ## Set up fixed gaps and edges for individual symptom analysis *speed up*
@@ -68,16 +71,50 @@ fixedGaps_full[1:9, 1:9] <- matrix(c(
 ), nrow = 9, byrow = TRUE)
 
 
+# Create a data frame of all parameter combinations
+params <- tidyr::expand_grid(alpha = alphas, threshold = thresholds, citest = citests, algorithm = algorithms)
 
-# Run with GaussianCI
-pc_result <- causal_subsampling(data, algorithm = "PC", subsample_size = subsample_size, num_subsamples = num_subsamples, alpha = alpha, threshold = threshold, citest = "gaussCItest", fixedEdges = fixedEdges_full, fixedGaps = fixedGaps_full)
-saveRDS(pc_result, file = sprintf("helius/data/gaussianPC_alpha_%s_threshold_%s.rds", alpha, threshold))
 
-fci_result <- causal_subsampling(data, algorithm = "FCI", subsample_size = subsample_size, num_subsamples = num_subsamples, alpha = alpha, threshold = threshold, citest = "gaussCItest", fixedEdges = fixedEdges_full, fixedGaps = fixedGaps_full)
-saveRDS(fci_result, file = sprintf("helius/data/gaussianFCI_alpha_%s_threshold_%s.rds", alpha, threshold))
+# Setup parallel backend
+plan(multisession)
[…]

### 2025-01-12 draft_v1/draft_v1.qmd
update_draft_ presum graph

@@
 ```{r}
+#| layout-nrow: 2
 #| label: fig-presum
 #| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using FCI and CCI."
+#| fig-subcap: 
+#|   - "FCI"
+#|   - "CCI"
 #| fig-align: center
-#| out-width: 80%
-
-knitr::include_graphics("img/presum_graph_refined.png")
+#| out-width: 65%
+
+knitr::include_graphics("img/FCI_presum.png")
+knitr::include_graphics("img/CCI_presum.png")
 
@@
 #| out-width: 70%
-knitr::include_graphics("img/presumgraph_mat.pdf")
-```
+knitr::include_graphics("img/presumgraph_mat2.pdf")
+```

### 2025-01-12 draft_v1/draft_v1.tex
update_draft_ presum graph

@@
 
-\date{2025-01-04}
+\date{2025-01-12}
 \begin{document}
@@
 
+\begin{minipage}{\linewidth}
+
 \centering{
 
-\includegraphics[width=0.8\textwidth,height=\textheight]{img/presum_graph_refined.png}
-
-}
+\includegraphics[width=0.65\textwidth,height=\textheight]{img/FCI_presum.png}
+
+}
+
+\subcaption{\label{fig-presum-1}FCI}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.65\textwidth,height=\textheight]{img/CCI_presum.png}
+
+}
+
+\subcaption{\label{fig-presum-2}CCI}
+
+\end{minipage}%
 
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/presumgraph_mat.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/presumgraph_mat2.pdf}
 

### 2025-01-12 draft_v1/draft_v1.qmd
add PC_presum_graph

@@
 ```{r}
-#| layout-nrow: 2
 #| label: fig-pc_presum
 #| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using PC."
-#| fig-subcap:
-#|   - "Using both GaussianCI and RCoT"
-#|   - "Using only RCoT"
 #| fig-align: center
@@
 
-knitr::include_graphics("img/pc_presum_GaussianCIRCoT.png")
-knitr::include_graphics("img/pc_presum_onlyRCoT.png")
+knitr::include_graphics("img/PC_presum.png")
 ```

### 2025-01-12 draft_v1/draft_v1.tex
add PC_presum_graph

@@
 
-\begin{minipage}{\linewidth}
-
 \centering{
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_presum_GaussianCIRCoT.png}
-
-}
-
-\subcaption{\label{fig-pc_presum-1}Using both GaussianCI and RCoT}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_presum_onlyRCoT.png}
-
-}
-
-\subcaption{\label{fig-pc_presum-2}Using only RCoT}
-
-\end{minipage}%
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/PC_presum.png}
+
+}
 

### 2025-01-12 code/graphviz/presum_peralgo.py

@@ -26,59 +26,59 @@ for node, pos in fixed_positions.items():
 
 
 # Manually define edges based on the adj.matrices
-# FCI
-dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="odot")
-dot.edge("anh", "slp", dir="both", arrowhead="normal", arrowtail="odot")
-dot.edge("anh", "ene", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="odot")
-dot.edge("anh", "glt", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("anh", "con", dir="both", arrowhead="normal", arrowtail="odot")
-dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot")
+# # FCI
+# dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("anh", "slp", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("anh", "ene", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("anh", "glt", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("anh", "con", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot")
 
-dot.edge("dep", "slp", dir="both", arrowhead="odot", arrowtail="normal")
-dot.edge("dep", "ene", dir="both", arrowhead="odot", arrowtail="normal")
-dot.edge("dep", "app", dir="both", arrowhead="odot", arrowtail="normal")
-dot.edge("dep", "glt", dir="both", arrowhead="odot", arrowtail="normal")
-dot.edge("dep", "con", dir="both", arrowhead="odot", arrowtail="odot")
-dot.edge("dep", "mot", dir="both", arrowhead="odot", arrowtail="normal")
-dot.edge("dep", "sui", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("dep", "Precarity", dir="both", arrowhead="odot", arrowtail="none")
+# dot.edge("dep", "slp", dir="both", arrowhead="odot", arrowtail="normal")
[…]

### 2025-01-12 code/post_analysis.R

@@ -64,6 +64,8 @@ grouped_stable_edges_ci <- split(stable_edges_data, comb$citest)
 fci_gauss_matrices <- grouped_stable_edges_ci$gaussCItest[grep("FCI", names(grouped_stable_edges_ci$gaussCItest))] 
 ## cci gauss
 cci_gauss_matrices <- grouped_stable_edges_ci$gaussCItest[grep("CCI", names(grouped_stable_edges_ci$gaussCItest))]
+## pc gauss
+pc_gauss_matrices <- grouped_stable_edges_ci$gaussCItest[grep("PC", names(grouped_stable_edges_ci$gaussCItest))]
 
 # summarize them by taking the most frequent symbol per element in matrices
 summarized_stable_edges <- lapply(grouped_stable_edges, function(group) {
@@ -125,6 +127,8 @@ summarized_stable_edges$FCI
 summarize_matrices(cci_gauss_matrices)
 summarized_stable_edges$CCI 
 
+summarize_matrices(pc_gauss_matrices) |> plotPC()
+summarized_stable_edges$PC
 
 ## Plot the resulting graph
 # per algorithm

### 2025-01-16 draft_v1/draft_v1.qmd
update_pc_sym

@@
 
+<!-- ```{r} -->
+<!-- #| layout-nrow: 2 -->
+<!-- #| label: fig-pc_sym -->
+<!-- #| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC." -->
+<!-- #| fig-subcap: -->
+<!-- #|   - "Using both GaussianCI and RCoT" -->
+<!-- #|   - "Using only RCoT" -->
+<!-- #| fig-align: center -->
+<!-- #| out-width: 70% -->
+
+<!-- knitr::include_graphics("img/pc_sym_GaussianCIRCoT.png") -->
+<!-- knitr::include_graphics("img/pc_sym_onlyRCoT.png") -->
+<!-- ``` -->
+
 ```{r}
-#| layout-nrow: 2
 #| label: fig-pc_sym
 #| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC."
-#| fig-subcap:
-#|   - "Using both GaussianCI and RCoT"
-#|   - "Using only RCoT"
 #| fig-align: center
@@
 
-knitr::include_graphics("img/pc_sym_GaussianCIRCoT.png")
-knitr::include_graphics("img/pc_sym_onlyRCoT.png")
+knitr::include_graphics("img/PC_symptom.png")
 ```

### 2025-01-16 draft_v1/draft_v1.tex
update_pc_sym

@@
 
-\date{2025-01-12}
+\date{2025-01-16}
 \begin{document}
@@
 
-\begin{minipage}{\linewidth}
-
 \centering{
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_sym_GaussianCIRCoT.png}
-
-}
-
-\subcaption{\label{fig-pc_sym-1}Using both GaussianCI and RCoT}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/pc_sym_onlyRCoT.png}
-
-}
-
-\subcaption{\label{fig-pc_sym-2}Using only RCoT}
-
-\end{minipage}%
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/PC_symptom.png}
+
+}
 

### 2025-01-16 code/graphviz/create_graph_PC_sym.py

@@ -1,0 +1,98 @@
+from graphviz import Digraph
+
+# Define the fixed positions for the nodes
+fixed_positions = {
+    "anh": (0, 3),
+    "dep": (1, 4),
+    "slp": (2, 3),
+    "ene": (3, 4),
+    "app": (4, 3),
+    "glt": (2, 2),
+    "con": (2, 1),
+    "mot": (1.7, 0),
+    "sui": (1, 1),
+    # "Prec": (4,1)
+    "P.emp": (3.5, 0),
+    "P.soc": (4, 1),
+    "P.hou": (5, 0),
+    "S.rel": (5, 2),
+    "S.fin": (6, 1)
+}
+
+# Function to create a graph with fixed layout and bidirectional edges
+def create_curvy_graph_with_bidirectional_edges(matrix1, matrix2, title, positions, filename):
+    dot = Digraph(format='png', engine='neato')
+    dot.attr(size="13,13", dpi="300", splines="true", ratio="fill")
+    dot.attr('node', fixedsize='true', width='1', height='1')
+
+    # Add nodes with fixed positions and uniform sizes
+    for node, pos in positions.items():
+        dot.node(node, node, shape="circle", pos=f"{pos[0]},{pos[1]}!",
+                 fixedsize="true", width='1', height='1', fontsize="30")
+
+    # Add edges based on the matrix and highlight the differences with dashed lines
+    nodes = list(positions.keys())
+    for i, src in enumerate(nodes):
+        for j, dest in enumerate(nodes):
+            if i < j:  # Avoid duplicate processing
+                style = "solid"
+                color = "black"  # Default color
+
+                if (matrix1[i][j] == 0 and matrix1[j][i] == 0 and (matrix2[i][j] == 1 or matrix2[j][i] == 1)):
[…]

### 2025-01-16 code/snellius/sym_gaussian.R

@@ -32,7 +32,7 @@ alphas = c(0.01, 0.05)
 thresholds = c(0.6, 0.7)
 citests = "gaussCItest"
 algorithms = c("PC", "FCI", "CCI")
-data <- dep_sym # sumscore 
+data <- depsym # sumscore 
 subsample_size <- nrow(data)  # Size of each subsample
 num_subsamples <- 30     # Number of subsamples
 

### 2025-02-07 code/edge_symbol_proportion.R
fix_tie

@@ -24,6 +24,7 @@ names(rds_data) <- gsub("\\.rds$", "", basename(file_names))
 
 # Check the loaded data
 print(names(rds_data))  # View the names of the datasets
+
 # Group files by algorithm
 grouped_rds_data <- list(
   FCI = rds_data[grep("FCI", names(rds_data))],

### 2025-02-07 code/graphviz/create_graph_sum.py

@@ -15,7 +15,7 @@ dot.attr(ratio="fill")
 
 # Add title
 # dot.attr(label="RCoT", labelloc="top", fontsize="20")
-dot.attr(label="Gaussian & RCoT", labelloc="top", fontsize="20")
+dot.attr(label="", labelloc="top", fontsize="20")
 
 # Set global node attributes for uniform size
 dot.attr('node', fixedsize='true', width='0.8', height='0.8')
@@ -76,21 +76,53 @@ dot.node("S.fin", "S.fin", shape="circle", pos="1,0!")
 # dot.edge("P.soc", "S.fin", dir="both", arrowtail="normal", arrowhead="odot")  # <-o
 # dot.edge("P.hou", "S.fin", dir="both", arrowtail="odot", arrowhead="normal", color="gray", style ="dashed")  # o->
 
-# both gaussian and rcot situation
+# # both gaussian and rcot situation
+# dot.edge("PHQsum", "P.emp", dir="both", arrowtail="odot", arrowhead="normal")  # o->
+# dot.edge("PHQsum", "P.soc", dir="both", arrowtail="odot", arrowhead="normal")  # o->
+# dot.edge("PHQsum", "S.fin", dir="both", arrowtail="odot", arrowhead="odot", color="gray", style ="dashed", constraint="false")  # <-
+# dot.edge("PHQsum", "S.rel", dir="both", arrowtail="normal", arrowhead="odot", color="gray", style ="dashed")  # <->
+
+# dot.edge("P.emp", "P.hou", dir="both", arrowtail="normal", arrowhead="odot")  # <-o
+# dot.edge("P.emp", "S.fin", dir="both", arrowtail="normal", arrowhead="odot", color="gray", style ="dashed")  # <-o
+
+# dot.edge("S.rel", "S.fin", dir="both", arrowtail="normal", arrowhead="normal")  # <->
+# dot.edge("S.rel", "P.soc", dir="both", arrowtail="odot", arrowhead="normal")  # o->
+
+# dot.edge("P.soc", "S.fin", dir="both", arrowtail="normal", arrowhead="odot")  # <-o
+# dot.edge("P.hou", "S.fin", dir="both", arrowtail="odot", arrowhead="normal")  # o->
+
+
+## both gaussian and rcot situation for FCI only
+# dot.edge("PHQsum", "P.emp", dir="both", arrowtail="odot", arrowhead="normal")  # o->
[…]

### 2025-02-07 code/graphviz/symptom_manual_graph.py

@@ -71,47 +71,75 @@ for node, pos in fixed_positions.items():
 # dot.edge("S.fin", "P.emp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="gray")
 # dot.edge("S.rel", "sui", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="gray")
 
-# gewoon black instead of gray dash
-dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
+## FCI only (RCoT and Gaussian dashed)
+dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="none")
 dot.edge("anh", "slp", dir="both", arrowhead="normal", arrowtail="odot")
-#dot.edge("anh", "glt", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("anh", "ene", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("anh", "P.soc", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("app", "ene", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("app", "mot", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("app", "slp", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("app", "glt", dir="both", arrowhead="odot", arrowtail="normal", style="dashed")
-dot.edge("app", "con", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("con", "anh", dir="both", arrowhead="odot", arrowtail="normal", style="dashed")
-dot.edge("con", "mot", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("dep", "glt", dir="both", arrowhead="normal", arrowtail="odot", style="dashed")
-dot.edge("dep", "slp", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("dep", "con", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("dep", "mot", dir="both", arrowhead="normal", arrowtail="normal")
-dot.edge("dep", "sui", dir="both", arrowhead="normal", arrowtail="normal")
[…]

### 2025-02-07 code/post_analysis.R

@@ -38,7 +38,8 @@ filenames <- glue::glue(
 # )
 
 # directory containing the files
-directory <- "Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_sumscore"  
+directory <- "data/dat_sumscore" #"Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_sumscore"  
+directory <- "data/symptom_both" 
 # directory <- "Precarity_HELIUS/data/dep_sumscore"  
 # directory <- "Precarity_HELIUS/data/dep_symscore"  
 
@@ -52,14 +53,29 @@ rds_data <- lapply(full_paths, readRDS)
 # Name the datasets for easy access
 names(rds_data) <- filenames
 
+# Check the loaded data
+print(names(rds_data))  # View the names of the datasets
+
+
 # Extract 'stable_edges' matrices from each list
 stable_edges_data <- lapply(rds_data, function(x) x$stable_edges)
 
 # Group matrices by algorithm
-grouped_stable_edges <- split(stable_edges_data, comb$algorithm)
+# grouped_stable_edges <- split(stable_edges_data, comb$algorithm)
+# Group files by algorithm
+grouped_stable_edges <- list(
+  PC = stable_edges_data[grep("PC", names(stable_edges_data))],
+  FCI = stable_edges_data[grep("FCI", names(stable_edges_data))],
+  CCI = stable_edges_data[grep("CCI", names(stable_edges_data))]
+)
 
 # Group matrices by CItest
-grouped_stable_edges_ci <- split(stable_edges_data, comb$citest)
+# grouped_stable_edges_ci <- split(stable_edges_data, comb$citest)
+grouped_stable_edges_ci <- list(
+  gaussCItest = stable_edges_data[grep("gaussCItest", names(stable_edges_data))],
+  RCoT = stable_edges_data[grep("RCoT", names(stable_edges_data))]
+)
+
 ## fci gauss
 fci_gauss_matrices <- grouped_stable_edges_ci$gaussCItest[grep("FCI", names(grouped_stable_edges_ci$gaussCItest))] 
 ## cci gauss
@@ -67,7 +83,13 @@ cci_gauss_matrices <- grouped_stable_edges_ci$gaussCItest[grep("CCI", names(grou
[…]

### 2025-02-07 code/utils/plotAG.R

@@ -28,6 +28,8 @@ plotAG <- function(amat, title_text = NULL)
   amat[amat == 1] <- "odot"
   amat[amat == 2] <- "vee"
   amat[amat == 3] <- "none"
+  amat[amat == 4] <- "tee"
+  
   iE <- 0
   for (i in 1:(p-1)) {
     x <- nn[i]

### 2025-02-08 code/edge_symbol_proportion.R
update_graphs&matrices

@@ -94,12 +94,13 @@ dominant_symbol_CCI <- apply(proportion_array_CCI, c(1, 2), function(x) which.ma
 
 
 # Define blended gradient colors for all symbols
+# Modify the blend_colors function to include alpha control based on proportions
 blend_colors <- function(proportions) {
   # Normalize proportions to sum to 1
   proportions <- proportions / sum(proportions)
   
   # Define base colors for each symbol
-  base_colors <- c("darkgray", "mediumpurple4", "aquamarine4", "salmon")
+  base_colors <- c("gray91", alpha("dodgerblue4", 0.7), alpha("aquamarine4", 0.7), alpha("salmon", 0.7))
   
   # Convert base colors to RGB
   base_rgb <- t(col2rgb(base_colors) / 255)
@@ -107,8 +108,11 @@ blend_colors <- function(proportions) {
   # Compute the weighted average of the colors
   blended_rgb <- colSums(base_rgb * proportions)
   
-  # Convert back to HEX color
-  rgb(blended_rgb[1], blended_rgb[2], blended_rgb[3])
+  # Calculate alpha based on the maximum proportion (this could be modified as needed)
+  alpha_value <- max(proportions)  # Set alpha to the maximum proportion for a stronger color impact
+  
+  # Return the final blended color with the calculated alpha
+  rgb(blended_rgb[1], blended_rgb[2], blended_rgb[3], alpha = alpha_value)
 }
 
 # Generate blended colors for all cells in the matrix
@@ -140,7 +144,7 @@ custom_labels <- c(
 custom_legend_v <- Legend(
   labels = custom_labels,
   legend_gp = gpar(
-    fill = c("darkgray", "mediumpurple4", "aquamarine4", "salmon")),
+    fill = c("gray91", alpha("dodgerblue4", 0.7), alpha("aquamarine4", 0.7), alpha("salmon", 0.7))),
   labels_gp = gpar( fontsize = 12,                  # Ensure label font size increases
                     fontfamily = "Palatino"         # Set font to Palatino for labels
   ),
@@ -156,7 +160,7 @@ custom_legend_v <- Legend(
 custom_legend_h <- Legend(
   labels = custom_labels,
[…]

### 2025-02-08 code/graphviz/create_graph_sum.py

@@ -95,34 +95,34 @@ dot.node("S.fin", "S.fin", shape="circle", pos="1,0!")
 ## both gaussian and rcot situation for FCI only
 # dot.edge("PHQsum", "P.emp", dir="both", arrowtail="odot", arrowhead="normal")  # o->
 # dot.edge("PHQsum", "P.soc", dir="both", arrowtail="odot", arrowhead="normal")  # o->
-# dot.edge("PHQsum", "S.fin", dir="both", arrowtail="normal", arrowhead="odot")  # <-
-# dot.edge("PHQsum", "S.rel", dir="both", arrowtail="normal", arrowhead="odot")  # <->
+# dot.edge("PHQsum", "S.fin", dir="both", arrowtail="normal", arrowhead="none")  # <-
+# dot.edge("PHQsum", "S.rel", dir="both", arrowtail="normal", arrowhead="normal")  # <->
 
 # dot.edge("P.emp", "P.hou", dir="both", arrowtail="normal", arrowhead="odot")  # <-o
-# dot.edge("P.emp", "S.fin", dir="both", arrowtail="normal", arrowhead="normal")  # <-o
+# dot.edge("P.emp", "S.fin", dir="both", arrowtail="normal", arrowhead="odot")  # <-o
 
-# dot.edge("S.rel", "S.fin", dir="both", arrowtail="odot", arrowhead="normal")  # <->
-# dot.edge("S.rel", "P.soc", dir="both", arrowtail="odot", arrowhead="normal", color="#000080", style ="dashed")  # o->
+# dot.edge("S.rel", "S.fin", dir="both", arrowtail="normal", arrowhead="normal")  # <->
+# dot.edge("S.rel", "P.soc", dir="both", arrowtail="odot", arrowhead="normal")  # o->
 
-# dot.edge("P.soc", "S.fin", dir="both", arrowtail="normal", arrowhead="odot")  # <-o
+# dot.edge("P.soc", "S.fin", dir="both", arrowtail="normal", arrowhead="tee", style="dashed")  # <-o
 # dot.edge("P.hou", "S.fin", dir="both", arrowtail="odot", arrowhead="normal")  # o->
 # # Render and view the graph
 # dot.render("FCI_depsum", view=True)
 
 
 ## both gaussian and rcot situation for CCI only
-dot.edge("PHQsum", "P.emp", dir="both", arrowtail="odot", arrowhead="normal")  # o->
[…]

### 2025-02-08 code/graphviz/symptom_manual_graph.py

@@ -71,76 +71,150 @@ for node, pos in fixed_positions.items():
 # dot.edge("S.fin", "P.emp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="gray")
 # dot.edge("S.rel", "sui", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="gray")
 
-## FCI only (RCoT and Gaussian dashed)
+
+
+# ## FCI only (RCoT and Gaussian - tie dashed)
+# dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="none")
+# dot.edge("anh", "slp", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("anh", "ene", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("anh", "glt", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("anh", "con", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("anh", "P.soc", dir="both", arrowhead="normal", arrowtail="tee", style="dashed")
+
+
+# dot.edge("dep", "slp", dir="both", arrowhead="odot", arrowtail="normal")
+# dot.edge("dep", "ene", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("dep", "app", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("dep", "glt", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("dep", "con", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("dep", "mot", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("dep", "sui", dir="both", arrowhead="odot", arrowtail="normal")
+# dot.edge("dep", "P.soc", dir="both", arrowhead="normal", arrowtail="none")
+# dot.edge("dep", "S.fin", dir="both", arrowhead="tee", arrowtail="tee", style="dashed")
+
+
+# dot.edge("slp", "ene", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("slp", "app", dir="both", arrowhead="normal", arrowtail="odot")
+# dot.edge("slp", "glt", dir="both", arrowhead="normal", arrowtail="normal")
[…]

### 2025-02-08 code/graphviz/symptom_presum_graph.py

@@ -25,54 +25,163 @@ for node, pos in fixed_positions.items():
              fixedsize="true", width="1", height="1", fontsize="25")
 
 
-# Manually define edges based on the adj.matrices
-dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
-dot.edge("anh", "slp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+# # Manually define edges based on the adj.matrices
+# dot.edge("anh", "dep", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+# dot.edge("anh", "slp", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+# dot.edge("anh", "ene", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("anh", "app", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+# dot.edge("anh", "glt", dir="both", arrowhead="normal", arrowtail="normal")
+# dot.edge("anh", "con", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+# dot.edge("anh", "Precarity", dir="both", arrowhead="normal", arrowtail="odot", style="dashed", color="#a9a9a9")
+
+# dot.edge("dep", "slp", dir="both", arrowhead="odot", arrowtail="normal", style="dashed", color="#a9a9a9")
+# dot.edge("dep", "ene", dir="both", arrowhead="odot", arrowtail="normal", style="dashed", color="#a9a9a9")
+# dot.edge("dep", "app", dir="both", arrowhead="odot", arrowtail="normal", style="dashed", color="#a9a9a9")
+# dot.edge("dep", "glt", dir="both", arrowhead="odot", arrowtail="normal", style="dashed", color="#a9a9a9")
+# dot.edge("dep", "con", dir="both", arrowhead="odot", arrowtail="odot", style="dashed", color="#a9a9a9")
+# dot.edge("dep", "mot", dir="both", arrowhead="odot", arrowtail="normal", style="dashed", color="#a9a9a9")
+# dot.edge("dep", "sui", dir="both", arrowhead="odot", arrowtail="normal", style="dashed", color="#a9a9a9")
[…]

### 2025-02-08 code/post_analysis.R

@@ -40,6 +40,9 @@ filenames <- glue::glue(
 # directory containing the files
 directory <- "data/dat_sumscore" #"Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_sym_presum" #"Precarity_HELIUS/data/dep_sumscore"  
 directory <- "data/symptom_both" 
+
+directory <- "data/symptom_presum2" 
+
 # directory <- "Precarity_HELIUS/data/dep_sumscore"  
 # directory <- "Precarity_HELIUS/data/dep_symscore"  
 

### 2025-02-08 draft_v1/draft_v1.qmd
draft_add_newresult

--- 2025-01-12 draft_v1/draft_v1.qmd
@@
 abstract: |
-  \noindent Understanding the causal mechanisms linking precarity factors and depression is critical for developing effective interventions.
-This study utilizes the HELIUS dataset to explore these relationships using advanced causal discovery methods.
-By applying algorithms such as FCI, and CCI, and combining traditional Gaussian CI tests with non-parametric approaches like RCoT, we investigate how precarity factors—including employment, social, financial, housing, and relational stress—affect depression, both as a sum score and at the individual symptom level.
-Our findings reveal that relational stress consistently emerges as a potential causal factor for depression, while symptoms such as sleep disturbances, guilt, and anhedonia are particularly sensitive to external stressors, acting as potential early warning signals or intervention points for prevention.
+  \noindent Understanding the causal mechanisms linking precariousness and depression is critical for developing effective interventions.
+This study utilizes data from the HELIUS cohort study to explore these relationships using advanced causal discovery methods.
+By applying algorithms such as FCI, and CCI, and combining traditional Gaussian CI tests with non-parametric approaches like RCoT, we investigate how different indicators of precariousness—including factors related to employment, social relations and relational stress, financial situation, and housing—affect depressed mood, both as a sum score and at the individual symptom level.
+Our findings reveal that relational stress consistently emerges as a potential causal factor for depressed mood, while symptoms such as sleep disturbances, guilt, and anhedonia are particularly sensitive to external stressors, acting as potential early warning signals or intervention points for prevention.
 Moreover, the results highlight complexities in the data, including the influence of latent confounders and the challenges of capturing cyclic relationships.
 Despite some limitations, such as unresolved ambiguities in causal directions and challenges with mixed data distributions, this study demonstrates the utility of causal discovery tools in disentangling the intricate interplay between social and mental health dynamics.
-By mapping these causal structures into computational models, future research can simulate intervention effects, providing actionable insights to mitigate the impact of precarity on mental health.
+By mapping these causal structures into computational models, future research can simulate intervention effects, providing actionable insights to mitigate the impact of precariousness on mental health.
 This study serves as a foundational effort, offering both methodological advancements and practical implications for addressing depression at a population level.
@@
 
-Mental health problems in urban areas have been reported to be on the rise.
-Governments have been attempting to intervene, but the complexity of mental health systems presents significant challenges in planning effective interventions, let alone understanding the underlying mechanisms driving these issues.
-
-Recent research has aimed to identify underlying factors contributing to mental health problems, often referred to as precariousness factors.
-These factors encompass various aspects of life, such as employment, social connections, financial stability, housing, and cultural dimensions.
-This comprehensive perspective on precarity helps to highlight how different aspects of life may be interconnected.
-While this research has advanced our understanding of the complex interplay between precariousness factors, a key question remains unanswered: how do these factors influence mental health?
-Specifically, the lack of directional information --- knowing what influences what --- limits our ability to identify and prioritize effective intervention targets.
-
-This study aims to investigate the causal relationships between precariousness factors and mental health outcomes, with a specific focus on depression, the most prevalent mental health issue in urban populations.
-Using causal discovery methods, we explore how different aspects of precarity influence depression and delve deeper into the dynamics at the symptom level.
-By examining individual depressive symptoms, we aim to identify which symptoms may act as initiators by being particularly sensitive to specific precariousness factors.
+Mental health problems in urban areas have been reported to be on the rise the complexity of mental health systems presents significant challenges in understanding the underlying mechanisms driving these issues, let alone planning effective interventions.
+
+Research has aimed to identify underlying factors contributing to mental health problems.
+Recent research looked into the association between mental health and indicators of precariousness in various dimensions of life, indicating a high level of uncertainty and instability in people’s lives, such as factors related to employment, social relations, finances, housing, and culture (ref Leonie’s paper).
+This comprehensive perspective on precariousness helped to highlight how different aspects of life may be interconnected.
+While this research has advanced our understanding of which indicators of precariousness are related to mental health, a key question remains unanswered: how do these factors influence mental health?
+Specifically, the lack of directional information — knowing what influences what — limits our ability to identify and prioritize effective intervention targets.
+
+This study aims to investigate the causal relationships between different dimensions of precariousness and mental health outcomes, to with a specific focus on depression, the most prevalent mental health issue.
+Using causal discovery methods, we explore how different aspects of precariousness influence depression and delve deeper into the dynamics at the symptom level.
+By examining individual depressive symptoms, we aim to identify which symptoms may act as initiators by being particularly sensitive to precariousness.
 Through this analysis, our goal is to uncover the causal mechanisms underlying mental health challenges and provide a foundation for developing more effective and targeted interventions.
@@
 
-We use the HELIUS dataset, which captures the diverse population of Amsterdam across various ethnicities and provides comprehensive health and lifestyle data.
+We use data from the HELIUS study, a multi-ethnic cohort that includes participants of Dutch, Turkish, Moroccan, Surinamese and Ghanaian origin [@snijder2017cohort].
 To operationalize precariousness factors, we draw on the framework outlined in previous research (i.e., Leonie's paper) and select a set of relevant variables.
@@
 #| label: fig-sum
-#| fig-cap: "Resulting graphs of precarity factors and depression sum score using FCI and CCI."
+#| fig-cap: "Resulting graph of precarity factors and depression sum score using FCI and proportion of edge endpoint types."
 #| fig-subcap: 
-#|   - "Using both GaussianCI and RCoT"
-#|   - "Using only RCoT"
+#|   - "FCI PAG"
+#|   - "Proportion matrix"
 #| fig-align: center
@@
 
-knitr::include_graphics("img/both_dot.png")
-knitr::include_graphics("img/rcot_dot.png")
+knitr::include_graphics("img/FCI_depsum.png")
+knitr::include_graphics("img/depsum_mat_fci.pdf")
 
@@
 #| label: fig-sym
-#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using FCI and CCI."
+#| layout-nrow: 2
+#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using FCI and proportion of edge endpoint types."
+#| fig-subcap: 
+#|   - "FCI PAG"
+#|   - "Proportion matrix"
 #| fig-align: center
-#| out-width: 80%
-
-knitr::include_graphics("img/symptom_graph_refined.png")
+#| out-width: 70%
+
+knitr::include_graphics("img/symptom_graph_FCI.png")
+knitr::include_graphics("img/symptom_mat_fci.pdf")
 
@@
 #| label: fig-presum
-#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using FCI and CCI."
+#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
 #| fig-subcap: 
-#|   - "FCI"
-#|   - "CCI"
+#|   - "FCI PAG"
+#|   - "Proportion matrix"
 #| fig-align: center
-#| out-width: 65%
-
-knitr::include_graphics("img/FCI_presum.png")
-knitr::include_graphics("img/CCI_presum.png")
-
+#| out-width: 60%
+
+knitr::include_graphics("img/presum_graph_FCI.png")
+knitr::include_graphics("img/presum_mat_fci.pdf")
 ```
@@
 
+## Results from CCI algorithm
+
+```{r}
+#| layout-ncol: 2
+#| label: fig-sum-cci
+#| fig-cap: "Resulting graph of precarity factors and depression sum score using CCI and proportion of edge endpoint types."
+#| fig-subcap: 
+#|   - "CCI MAAG"
+#|   - "Proportion matrix"
+#| fig-align: center
+#| out-width: 100%
+
+knitr::include_graphics("img/CCI_depsum.png")
+knitr::include_graphics("img/depsum_mat_cci.pdf")
+
+```
+
+```{r}
+#| label: fig-sym-cci
+#| layout-nrow: 2
+#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using CCI and proportion of edge endpoint types."
+#| fig-subcap: 
+#|   - "CCI MAAG"
+#|   - "Proportion matrix"
+#| fig-align: center
+#| out-width: 60%
+
+knitr::include_graphics("img/symptom_graph_CCI.png")
+knitr::include_graphics("img/symptom_mat_cci.pdf")
+
+```
+
+```{r}
+#| layout-nrow: 2
+#| label: fig-presum-cci
+#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using CCI and proportion of edge endpoint types."
+#| fig-subcap: 
+#|   - "CCI MAAG"
+#|   - "Proportion matrix"
+#| fig-align: center
+#| out-width: 60%
+
+knitr::include_graphics("img/presum_graph_CCI.png")
+knitr::include_graphics("img/presum_mat_cci.pdf")
+```
+
+\clearpage
+
 ## Results from PC algorithm
@@
 ```{r}
-#| layout-nrow: 2
 #| label: fig-pc_sym
 #| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC."
-#| fig-subcap:
-#|   - "Using both GaussianCI and RCoT"
-#|   - "Using only RCoT"
 #| fig-align: center
-#| out-width: 70%
-
-knitr::include_graphics("img/pc_sym_GaussianCIRCoT.png")
-knitr::include_graphics("img/pc_sym_onlyRCoT.png")
+#| out-width: 60%
+
+knitr::include_graphics("img/PC_symptom.png")
 ```
@@
 #| fig-align: center
-#| out-width: 70%
+#| out-width: 60%
 
@@
 
-## Summarized Stable Edges Proportion {#sec-propmatrix}
-
-```{r}
-#| label: fig-sum-mat
-#| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum score."
-#| fig-align: center
-#| out-width: 100%
-knitr::include_graphics("img/sumgraph_mat.pdf")
-```
-
-```{r}
-#| label: fig-sym-mat
-#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms."
-#| fig-align: center
-#| out-width: 100%
-knitr::include_graphics("img/symgraph_mat.pdf")
-```
-
-```{r}
-#| label: fig-presum-mat
-#| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precarity sum score."
-#| fig-align: center
-#| out-width: 70%
-knitr::include_graphics("img/presumgraph_mat2.pdf")
-```
+<!-- ## Summarized Stable Edges Proportion {#sec-propmatrix} -->
+
+<!-- ```{r} -->
+<!-- #| label: fig-sum-mat -->
+<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum score." -->
+<!-- #| fig-align: center -->
+<!-- #| out-width: 100% -->
+<!-- knitr::include_graphics("img/sumgraph_mat.pdf") -->
+<!-- knitr::include_graphics("img/sumgraph_mat.pdf") -->
+
+<!-- ``` -->
+
+<!-- ```{r} -->
+<!-- #| label: fig-sym-mat -->
+<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms." -->
+<!-- #| fig-align: center -->
+<!-- #| out-width: 100% -->
+<!-- knitr::include_graphics("img/symgraph_mat.pdf") -->
+<!-- ``` -->
+
+<!-- ```{r} -->
+<!-- #| label: fig-presum-mat -->
+<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precarity sum score." -->
+<!-- #| fig-align: center -->
+<!-- #| out-width: 70% -->
+<!-- knitr::include_graphics("img/presumgraph_mat2.pdf") -->
+<!-- ``` -->

### 2025-02-08 draft_v1/draft_v1.tex
draft_add_newresult

@@
 
-\date{2025-01-16}
+\date{2025-02-08}
 \begin{document}
@@
 \begin{abstract}
-\noindent Understanding the causal mechanisms linking precarity factors
-and depression is critical for developing effective interventions. This
-study utilizes the HELIUS dataset to explore these relationships using
-advanced causal discovery methods. By applying algorithms such as FCI,
-and CCI, and combining traditional Gaussian CI tests with non-parametric
-approaches like RCoT, we investigate how precarity factors---including
-employment, social, financial, housing, and relational stress---affect
-depression, both as a sum score and at the individual symptom level. Our
-findings reveal that relational stress consistently emerges as a
-potential causal factor for depression, while symptoms such as sleep
-disturbances, guilt, and anhedonia are particularly sensitive to
-external stressors, acting as potential early warning signals or
-intervention points for prevention. Moreover, the results highlight
-complexities in the data, including the influence of latent confounders
-and the challenges of capturing cyclic relationships. Despite some
-limitations, such as unresolved ambiguities in causal directions and
-challenges with mixed data distributions, this study demonstrates the
-utility of causal discovery tools in disentangling the intricate
-interplay between social and mental health dynamics. By mapping these
-causal structures into computational models, future research can
-simulate intervention effects, providing actionable insights to mitigate
-the impact of precarity on mental health. This study serves as a
-foundational effort, offering both methodological advancements and
-practical implications for addressing depression at a population level.
+\noindent Understanding the causal mechanisms linking precariousness and
+depression is critical for developing effective interventions. This
+study utilizes data from the HELIUS cohort study to explore these
+relationships using advanced causal discovery methods. By applying
+algorithms such as FCI, and CCI, and combining traditional Gaussian CI
+tests with non-parametric approaches like RCoT, we investigate how
+different indicators of precariousness---including factors related to
+employment, social relations and relational stress, financial situation,
+and housing---affect depressed mood, both as a sum score and at the
+individual symptom level. Our findings reveal that relational stress
+consistently emerges as a potential causal factor for depressed mood,
+while symptoms such as sleep disturbances, guilt, and anhedonia are
+particularly sensitive to external stressors, acting as potential early
+warning signals or intervention points for prevention. Moreover, the
+results highlight complexities in the data, including the influence of
+latent confounders and the challenges of capturing cyclic relationships.
+Despite some limitations, such as unresolved ambiguities in causal
+directions and challenges with mixed data distributions, this study
+demonstrates the utility of causal discovery tools in disentangling the
+intricate interplay between social and mental health dynamics. By
+mapping these causal structures into computational models, future
+research can simulate intervention effects, providing actionable
+insights to mitigate the impact of precariousness on mental health. This
+study serves as a foundational effort, offering both methodological
+advancements and practical implications for addressing depression at a
+population level.
 \end{abstract}
@@
 Mental health problems in urban areas have been reported to be on the
-rise. Governments have been attempting to intervene, but the complexity
-of mental health systems presents significant challenges in planning
-effective interventions, let alone understanding the underlying
-mechanisms driving these issues.
-
-Recent research has aimed to identify underlying factors contributing to
-mental health problems, often referred to as precariousness factors.
-These factors encompass various aspects of life, such as employment,
-social connections, financial stability, housing, and cultural
-dimensions. This comprehensive perspective on precarity helps to
-highlight how different aspects of life may be interconnected. While
-this research has advanced our understanding of the complex interplay
-between precariousness factors, a key question remains unanswered: how
-do these factors influence mental health? Specifically, the lack of
-directional information --- knowing what influences what --- limits our
-ability to identify and prioritize effective intervention targets.
+rise the complexity of mental health systems presents significant
+challenges in understanding the underlying mechanisms driving these
+issues, let alone planning effective interventions.
+
+Research has aimed to identify underlying factors contributing to mental
+health problems. Recent research looked into the association between
+mental health and indicators of precariousness in various dimensions of
+life, indicating a high level of uncertainty and instability in people's
+lives, such as factors related to employment, social relations,
+finances, housing, and culture (ref Leonie's paper). This comprehensive
+perspective on precariousness helped to highlight how different aspects
+of life may be interconnected. While this research has advanced our
+understanding of which indicators of precariousness are related to
+mental health, a key question remains unanswered: how do these factors
+influence mental health? Specifically, the lack of directional
+information --- knowing what influences what --- limits our ability to
+identify and prioritize effective intervention targets.
 
 This study aims to investigate the causal relationships between
-precariousness factors and mental health outcomes, with a specific focus
-on depression, the most prevalent mental health issue in urban
-populations. Using causal discovery methods, we explore how different
-aspects of precarity influence depression and delve deeper into the
+different dimensions of precariousness and mental health outcomes, to
+with a specific focus on depression, the most prevalent mental health
+issue. Using causal discovery methods, we explore how different aspects
+of precariousness influence depression and delve deeper into the
 dynamics at the symptom level. By examining individual depressive
 symptoms, we aim to identify which symptoms may act as initiators by
-being particularly sensitive to specific precariousness factors. Through
-this analysis, our goal is to uncover the causal mechanisms underlying
-mental health challenges and provide a foundation for developing more
-effective and targeted interventions.
+being particularly sensitive to precariousness. Through this analysis,
+our goal is to uncover the causal mechanisms underlying mental health
+challenges and provide a foundation for developing more effective and
+targeted interventions.
 
@@
 
-We use the HELIUS dataset, which captures the diverse population of
-Amsterdam across various ethnicities and provides comprehensive health
-and lifestyle data. To operationalize precariousness factors, we draw on
-the framework outlined in previous research (i.e., Leonie's paper) and
-select a set of relevant variables.
+We use data from the HELIUS study, a multi-ethnic cohort that includes
+participants of Dutch, Turkish, Moroccan, Surinamese and Ghanaian origin
+(Snijder et al., 2017). To operationalize precariousness factors, we
+draw on the framework outlined in previous research (i.e., Leonie's
+paper) and select a set of relevant variables.
 
@@
 experimental setups.\footnote{The detailed proportion of each edge
-  endpoint occurrence is provided in Section~\ref{sec-propmatrix}.} This
+  endpoint occurrence is provided in \textbf{?@sec-propmatrix}.} This
 ensures that only stable and consistent edges are retained, providing a
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{img/both_dot.png}
-
-}
-
-\subcaption{\label{fig-sum-1}Using both GaussianCI and RCoT}
+\includegraphics[width=1\textwidth,height=\textheight]{img/FCI_depsum.png}
+
+}
+
+\subcaption{\label{fig-sum-1}FCI PAG}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{img/rcot_dot.png}
-
-}
-
-\subcaption{\label{fig-sum-2}Using only RCoT}
-
-\end{minipage}%
-
-\caption{\label{fig-sum}Resulting graphs of precarity factors and
-depression sum score using FCI and CCI.}
+\includegraphics[width=1\textwidth,height=\textheight]{img/depsum_mat_fci.pdf}
+
+}
+
+\subcaption{\label{fig-sum-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-sum}Resulting graph of precarity factors and
+depression sum score using FCI and proportion of edge endpoint types.}
 
@@
 
-\centering{
-
-\includegraphics[width=0.8\textwidth,height=\textheight]{img/symptom_graph_refined.png}
-
-}
-
-\caption{\label{fig-sym}Resulting graphs of precarity factors and
-individual depression symptoms using FCI and CCI.}
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/symptom_graph_FCI.png}
+
+}
+
+\subcaption{\label{fig-sym-1}FCI PAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/symptom_mat_fci.pdf}
+
+}
+
+\subcaption{\label{fig-sym-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-sym}Resulting graph of precarity factors and
+individual depression symptoms using FCI and proportion of edge endpoint
+types.}
 
@@
 between symptom-symptom relationships. For a detailed breakdown, see
-Section~\ref{sec-propmatrix}, which shows the proportion of symbols
+\textbf{?@sec-propmatrix}, which shows the proportion of symbols
 produced by each algorithm. When examining overall precarity, we observe
@@
 
-\includegraphics[width=0.65\textwidth,height=\textheight]{img/FCI_presum.png}
-
-}
-
-\subcaption{\label{fig-presum-1}FCI}
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_graph_FCI.png}
+
+}
+
+\subcaption{\label{fig-presum-1}FCI PAG}
 
@@
 
-\includegraphics[width=0.65\textwidth,height=\textheight]{img/CCI_presum.png}
-
-}
-
-\subcaption{\label{fig-presum-2}CCI}
-
-\end{minipage}%
-
-\caption{\label{fig-presum}Resulting graphs of precarity sum score and
-individual depression symptoms using FCI and CCI.}
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_mat_fci.pdf}
+
+}
+
+\subcaption{\label{fig-presum-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-presum}Resulting graph of precarity sum score and
+individual depression symptoms using FCI and proportion of edge endpoint
+types.}
 
@@
 
+\bibitem[\citeproctext]{ref-snijder2017cohort}
+Snijder, M. B., Galenkamp, H., Prins, M., Derks, E. M., Peters, R. J.,
+Zwinderman, A. H., \& Stronks, K. (2017). Cohort profile: The healthy
+life in an urban setting (HELIUS) study in amsterdam, the netherlands.
+\emph{BMJ Open}, \emph{7}(12), e017873.
+
 \bibitem[\citeproctext]{ref-spirtes2001causation}
@@
 
-\subsection{Results from PC algorithm}\label{results-from-pc-algorithm}
+\subsection{Results from CCI
+algorithm}\label{results-from-cci-algorithm}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{img/sum_PC_all.png}
-
-}
-
-\subcaption{\label{fig-pc_sum-1}Using both GaussianCI and RCoT}
+\includegraphics[width=1\textwidth,height=\textheight]{img/CCI_depsum.png}
+
+}
+
+\subcaption{\label{fig-sum-cci-1}CCI MAAG}
 
@@
 
+\includegraphics[width=1\textwidth,height=\textheight]{img/depsum_mat_cci.pdf}
+
+}
+
+\subcaption{\label{fig-sum-cci-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-sum-cci}Resulting graph of precarity factors and
+depression sum score using CCI and proportion of edge endpoint types.}
+
+\end{figure}%
+
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/symptom_graph_CCI.png}
+
+}
+
+\subcaption{\label{fig-sym-cci-1}CCI MAAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/symptom_mat_cci.pdf}
+
+}
+
+\subcaption{\label{fig-sym-cci-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-sym-cci}Resulting graph of precarity factors and
+individual depression symptoms using CCI and proportion of edge endpoint
+types.}
+
+\end{figure}%
+
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_graph_CCI.png}
+
+}
+
+\subcaption{\label{fig-presum-cci-1}CCI MAAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_mat_cci.pdf}
+
+}
+
+\subcaption{\label{fig-presum-cci-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-presum-cci}Resulting graph of precarity sum score
+and individual depression symptoms using CCI and proportion of edge
+endpoint types.}
+
+\end{figure}%
+
+\clearpage
+
+\subsection{Results from PC algorithm}\label{results-from-pc-algorithm}
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{img/sum_PC_all.png}
+
+}
+
+\subcaption{\label{fig-pc_sum-1}Using both GaussianCI and RCoT}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
 \includegraphics[width=1\textwidth,height=\textheight]{img/sum_PC_RCoTonly.png}
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/PC_symptom.png}
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/PC_symptom.png}
 
@@
 
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/PC_presum.png}
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/PC_presum.png}
 
@@
 
-\subsection{Summarized Stable Edges Proportion}\label{sec-propmatrix}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{img/sumgraph_mat.pdf}
-
-}
-
-\caption{\label{fig-sum-mat}Proportions of edge endpoint types for
-graphs based on depression sum score.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{img/symgraph_mat.pdf}
-
-}
-
-\caption{\label{fig-sym-mat}Proportions of edge endpoint types for
-graphs based on individual depression symptoms.}
-
-\end{figure}%
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/presumgraph_mat2.pdf}
-
-}
-
-\caption{\label{fig-presum-mat}Proportions of edge endpoint types for
-graphs based on individual depression symptoms with precarity sum
-score.}
-
-\end{figure}%
-
 \end{document}

### 2025-02-11 draft_v1/draft_v1.qmd
update_results
only FCI in main text + proportion matrix plots

@@
 
-Mental health problems in urban areas have been reported to be on the rise the complexity of mental health systems presents significant challenges in understanding the underlying mechanisms driving these issues, let alone planning effective interventions.
-
-Research has aimed to identify underlying factors contributing to mental health problems.
+Mental health problems in urban areas are on the rise, yet the complexity of mental health systems poses significant challenges in understanding the underlying mechanisms driving these issues — let alone designing effective interventions.
+Research has long sought to identify underlying factors contributing to mental health problems.
 Recent research looked into the association between mental health and indicators of precariousness in various dimensions of life, indicating a high level of uncertainty and instability in people’s lives, such as factors related to employment, social relations, finances, housing, and culture (ref Leonie’s paper).
@@
 While this research has advanced our understanding of which indicators of precariousness are related to mental health, a key question remains unanswered: how do these factors influence mental health?
-Specifically, the lack of directional information — knowing what influences what — limits our ability to identify and prioritize effective intervention targets.
+Specifically, the lack of directional insight — knowing what influences what — limits our ability to identify and prioritize effective intervention targets.
 
 This study aims to investigate the causal relationships between different dimensions of precariousness and mental health outcomes, to with a specific focus on depression, the most prevalent mental health issue.
-Using causal discovery methods, we explore how different aspects of precariousness influence depression and delve deeper into the dynamics at the symptom level.
-By examining individual depressive symptoms, we aim to identify which symptoms may act as initiators by being particularly sensitive to precariousness.
-Through this analysis, our goal is to uncover the causal mechanisms underlying mental health challenges and provide a foundation for developing more effective and targeted interventions.
+Using causal discovery methods, we explore how various aspects of precariousness contribute to depression and whether depression, in turn, exacerbates precariousness, potentially creating a feedback loop.
+Furthermore, we zoom in at the symptom level, examining individual depressive symptoms to identify those that may act as initiators — triggering the activation of the symptom network — by displaying heightened sensitivity to specific precariousness factors.
+We also investigate symptoms that may function as bridging elements, linking the dynamics of depression with precariousness by feeding back into precarious conditions, potentially reinforcing the cycle of mental health deterioration and precarious conditions.
+Through this analysis, we aim to uncover the causal mechanisms driving mental health challenges related to precariousness and establish a foundation for more targeted and effective interventions by identifying the key driving forces within this system.
 
@@
 There are numerous causal discovery algorithms available; however, in this study, we focus on algorithms suited to the potential cyclic relationships within our system.
-Specifically, we use FCI (Fast Causal Inference) and CCI (Cyclic Causal Inference), both capable of accounting for such cycles under certain conditions [@mooij2020constraint; @strobl2019].
-Additionally, we include the PC algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation].
+Specifically, we use *FCI* (Fast Causal Inference) and *CCI* (Cyclic Causal Inference), both capable of accounting for such cycles [@mooij2020constraint; @strobl2019].
+Additionally, we include the *PC* algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation].
 
@@
 As shown in Table 1, the resulting graphs from FCI and CCI differ slightly (*PAG*: partial ancestral graph; *MAAG*: maximal almost ancestral graph) due to their reliance on different underlying assumptions.
-Despite these differences, both graphs are forms of *ancestral graphs*, designed to encode causal relationships between variables, where the presence of an edge indicates causal *ancestry.*
-Directed edges, $A \stararrow B$, indicate that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$.
-The Markov equivalence class refers to the set of graphs that encode the same conditional independence relationships, such that the same *d-separation* relations hold across all graphs in the class [@spirtes2001causation].
-$A \startail B$ indicates that $B$ is an ancestor of $A$ across all graphs in $Equiv(G)$. 
+Despite these differences, both graphs belong to the class of *ancestral graphs*, which are designed to encode causal relationships between variables, where the presence of an edge indicates causal *ancestry.*
+In these graphs, directed edges, $A \stararrow B$, indicate that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$. 
+The Markov equivalence class represents a set of graphs that encode the same conditional independence relationships, ensuring that the same *d-separation* conditions hold across all graphs in the class [@spirtes2001causation].
+Conversely, an edge marked as $A \startail B$ indicates that $B$ is an ancestor of $A$ across all graphs in $Equiv(G)$. 
 Circle endpoints, $A \starcirc B$, represent ambiguity in the ancestral relationship, meaning $B$'s ancestral status relative to $A$ varies across graphs in $Equiv(G)$.[^4] 
-When $A \arrowarrow B$, neither $A$ nor $B$ is an ancestor of the other, indicating the presence of a latent confounder between them.
-
-The algorithms also differ in their approaches and assumptions for detecting cycles.
-In CCI's MAAG, $A$ \textemdash\ $B$ indicates that $A$ is an ancestor of $B$ and $B$ is an ancestor of $A$, referring to a cycle between $A \stackedarrows B$. 
-FCI, on the other hand, identifies potential cycles more subtly: fully-connected nodes with circle endpoints ($\circirc$) in FCI's PAG may suggest a cyclic structure. 
-FCI basically provides a sufficient condition to distinguish nodes that are not part of a cycle, offering a more nuanced approach [@mooij2020constraint].
+Finally, when an edge is represented as $A \arrowarrow B$, it implies that neither $A$ nor $B$ is an ancestor of the other, suggesting the presence of a latent confounder influencing both variables.
+
+The algorithms also differ in how they detect cycles and represent cyclic relationships in their resulting graphs.
+In CCI's MAAG, $A$ \textemdash\ $B$ indicates that $A$ is an ancestor of $B$, and simultaneously $B$ is an ancestor of $A$, referring to a cyclic relationship between $A \stackedarrows B$. 
+FCI, on the other hand, identifies potential cycles more subtly. In its PAG, fully-connected nodes with circle endpoints ($\circirc$) may suggest the presence of cyclic structures.
+FCI, therefore, provides a sufficient condition to distinguish variables that are not part of a cycle, offering a more nuanced approach to handling cyclic relationships [@mooij2020constraint].
+
 Table 1 highlights further differences in the assumptions underlying these algorithms.
-CCI operates under the assumption of a linear system, while FCI, particularly for inferring cyclic relationships, assumes a non-linear system without selection bias and adheres to more general faithfulness and Markov conditions (i.e., *$\sigma$-separation* and *$\sigma$-faithfulness* setting) [@forre2018constraint].
-When the respective assumptions of each algorithm are satisfied, their outcomes for cyclic relationships align with those illustrated in @fig-examplegraphs.
-While an exhaustive discussion of causal discovery concepts and algorithm-specific details is beyond the scope of this paper, readers seeking a deeper understanding are encouraged to consult @park2024discovering for comprehensive insights into these methods and their applications.
+CCI operates under the assumption of a linear system, whereas FCI, particularly when used to infer cyclic relationships, assumes a non-linear system without selection bias and adheres to more general faithfulness and Markov conditions (i.e., *$\sigma$-separation* and *$\sigma$-faithfulness* setting) [@forre2018constraint].
+When the respective assumptions of each algorithm are met, their inferred cyclic relationships align with those illustrated in @fig-examplegraphs.
+
+While CCI's MAAG provides an explicit representation of cyclic relationships, it has certain theoretical limitations, as the MAAG may not always fully preserve d-separation relations from the original graph, $\mathcal{G}$ [@strobl2019].
+Given these considerations, we examine and compare results from both algorithms, placing more weight on the FCI results, which are presented in the main results section, while discussing CCI results where relevant, with detailed findings provided in the [Appendix](#sec-appendix).
+A full discussion of causal discovery concepts and algorithmic details is beyond the scope of this paper; however, readers seeking a more in-depth understanding can refer to @park2024discovering for a comprehensive exploration of these methods and their applications.
 
@@
 Instead, when the PC algorithm cannot determine the direction of causality, it represents this uncertainty with bidirectional arrows.
-While the PC algorithm serves as a useful reference, its strict assumptions of acyclicity and the absence of latent confounders restrict its applicability in more complex scenarios.
-Therefore, our primary focus remains on the results from FCI and CCI, with all PC algorithm results provided in the [Appendix](#sec-appendix) for completeness.
-
-One practical challenge in applying these algorithms to the HELIUS dataset is that the data does not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear.
-To account for this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel].
-While kernel-based conditional independence tests (KCIT) are effective, they are computationally intensive for large datasets like HELIUS, as their quadratic scaling with sample size is primarily driven by the inversion of large kernel matrices.
-To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximate kernel methods using random Fourier features.
-This approach reduces computation to linear scaling with sample size, significantly lowering computational costs [@strobl2019approximate].
+While the PC algorithm serves as a useful reference, its strict assumptions — acyclicity and the absence of latent confounders—limit its applicability in more complex scenarios.
+For this reason, our primary focus remains on the results obtained from FCI and CCI, with all PC algorithm results provided in the [Appendix](#sec-appendix) for completeness.
+
+One practical challenge in applying these algorithms to the HELIUS dataset is that most variables do not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear.
+To address this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel].
+While kernel-based conditional independence tests (KCIT) are theoretically effective in capturing non-linear dependencies and subtle interactions in data with non-Gaussian noise, they are computationally demanding for large datasets like HELIUS.
+Their quadratic scaling with sample size is primarily due to the need to invert large kernel matrices, making them impractical for large-scale analysis [@rahimi2007random].
+To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximates kernel methods using random Fourier features.
+This approach reduces computational complexity from quadratic to linear scaling with sample size, significantly lowering computational costs while maintaining robustness [@strobl2019approximate].
 For a more detailed explanation of RCoT, refer to @sec-rcot.
 
-## Analysis
-We analyze the causal structure using three approaches: examining the relationship (1) between the five individual precarity factors and the PHQ sum score, which represents overall depression severity; (2) between individual symptoms and precarity factors; and (3) between individual symptoms and overall precarity (the sum of the five precarity factors).
-Using the PHQ sum score simplifies the analysis by reducing dimensionality, offering computational efficiency and a broad, interpretable perspective on the relationship between depression and precarity factors.
-In contrast, analyzing individual symptoms provides a more detailed understanding by capturing the diverse ways symptoms respond to different precarity factors.
-However, this detailed approach introduces methodological challenges, including non-standard distributions of individual symptom variables and the complexities associated with increased dimensionality.
-Finally, examining individual symptoms in relation to the overall precarity score complements the analysis by revealing how symptoms collectively connect to general precarity.
-Aggregating precarity factors may uncover relationships that could be overlooked in analyses of individual factors, particularly when a symptom has associations across multiple factors that are not strong enough to emerge in individual analyses.
-By integrating these three approaches, we achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression. 
+## Analysis{#sec-analysis}
+
+We analyze the causal structure using three complementary approaches: (1) Examining the relationship between the five individual precariousness factors and the PHQ sum score, which represents overall depression severity.
+(2) Investigating the relationship between individual depressive symptoms and individual precariousness factors.
+(3) Assessing the relationship between individual depressive symptoms and overall precariousness (the sum of all precariousness factors).
+Using the PHQ sum score as a measure of depression simplifies the analysis by reducing dimensionality, improving computational efficiency, and providing a broad, interpretable perspective on how precariousness factors relate to overall depression severity.
+However, this aggregated approach may overlook granular symptom-level relationships.
+
+By zooming in on individual symptoms, we capture a more detailed picture of how specific depressive symptoms interact with different precariousness factors.
+This approach allows us to map individual causal pathways, identifying which symptoms are particularly sensitive to certain precarious conditions and vice versa.
+However, this level of detail introduces methodological challenges, including non-standard distributions of symptom variables and the high-dimensional nature of the analysis, which can reduce statistical power and significantly increase computational demands—especially when using non-parametric conditional independence (CI) tests.
+
+Finally, analyzing individual symptoms in relation to overall precariousness provides an integrative perspective, revealing how depression symptoms collectively relate to precarious conditions.
+This approach complements the fully disaggregated analysis by aggregating precariousness factors, potentially uncovering relationships that may be too weak to detect when examined separately.
+This is particularly relevant when a symptom is associated with multiple precariousness factors, but these associations are not strong enough to emerge in an individual-level analysis.
+By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression.
+
 <!-- The PHQ sum score and precarity sum score capture overarching trends between overall depression and precarity, while symptom-level analysis with individual precarity factors reveals detailed dynamics, essential for tailoring specific interventions.
@@
 Furthermore, to address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping.
-For each of the 100 bootstrap samples, we estimate causal graphs and retain only the edges and directions that exceed predefined thresholds.
+For each condition, we generate 100 bootstrap samples, estimate causal graphs, and retain only the edge-endpoints that exceed predefined thresholds.
+If no edge-endpoint type surpasses the threshold, we assign a circle ($\circ$) to indicate uncertainty in the edge direction.
+
 The analyses are conducted under the following conditions:
@@
 For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30, easing computational demands.
-To further enhance efficiency, we fix the skeleton of edges among symptom variables based on the common structure estimated across all algorithms.
-This approach minimizes spurious edges in the symptom network and significantly reduces the computational time needed for skeleton estimation.
-This fixed structure also aligns well with commonly reported skeleton structures in the literature, ensuring consistency with existing findings (**cite the comp_model paper**).[^2]
-Lastly, to summarize the results, we identify the most frequently occurring edge endpoints across different experimental setups.[^3] This ensures that only stable and consistent edges are retained, providing a clearer and more reliable understanding of the relationships between precarity factors and depression.
+To further enhance efficiency, we fix the skeleton of the symptom network using a common structure derived from PC, FCI, and CCI algorithms ($\alpha$ = 0.01, RCoT CI test).
+This step constrains which edges can be present among symptoms but does not restrict the estimation of causal directions within the symptom network.
+Additionally, the entire causal structure between symptoms and precariousness factors remains fully estimable, ensuring that key relationships are still inferred.
+Beyond its computational benefits — particularly in reducing the time required for skeleton estimation — this fixed structure also aligns well with commonly reported symptom network structures in the literature [@park2025individual], ensuring consistency with existing findings.
+
+<!-- [^2] -->
+To obtain a final, stable summary of causal relationships, we identify the most frequently occurring edge-endpoints across different experimental setups.
+First, within each condition, we determine the most frequent edge-endpoint from all bootstrap samples.
+Then, across different conditions, we aggregate these selected symbols and determine the overall most common edge-endpoint for each graph entry.
+
+Since we have an even number of experimental setups, ties may occasionally occur.
+In such cases, we explicitly denote them in the resulting graph using dashed lines.
+To further help with the interpretation and understanding of the resulting graph, we provide a corresponding table for each graph, showing the proportion of each edge-endpoint type to clarify how dominant a given symbol is in each case.
+
+By double summarizing the results in this way, we generate one final graph per algorithm for each analysis approach: the partially aggregated analysis, which uses either the sum of precariousness factors or the sum of depression factors to capture overall trends, and the fully disaggregated analysis, which examines individual precariousness and depression factors separately.
+This approach ensures that only stable and consistent edges — those that persist across different parameter settings — are retained, leading to a robust and reliable understanding of the relationships between precariousness factors and depression.
 See @fig-workflow for an overview of the analysis workflow.
 
-[^2]:
-We fixed the skeleton of the symptom network using a common structure derived from the PC, FCI, and CCI algorithms with an alpha level of 0.001.
+<!-- [^2]: -->
+<!-- We fixed the skeleton of the symptom network using a common structure derived from the PC, FCI, and CCI algorithms with an alpha level of 0.01 using RCoT CI test.
 This approach preserved the complexity of symptom interactions, enabling unrestricted estimation of causal directions both among symptoms (*symptom --- symptom* interactions) and between symptoms and precarity factors.
-
-[^3]:
-The detailed proportion of each edge endpoint occurrence is provided in [@sec-propmatrix].
+-->
+
+<!-- [^3]: -->
+<!-- The detailed proportion of each edge endpoint occurrence is provided in [@sec-propmatrix]. -->
 
@@
 The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
-@fig-sum illustrates the causal relationships between precarity factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*) under two different setups: (a) using both Gaussian CI test and RCoT, and (b) using RCoT alone.
-Black edges represent consistent causal relationships identified by both FCI and CCI, while gray dashed edges denote inconsistent relationships that vary between the two methods.
-The inconsistent endpoints of the gray edges are marked with circles.
-
-In graph (a), the key pathways suggest that employment precarity (*P.emp*) and social precarity (*P.soc*) do not cause depression (*PHQsum*).
-*P.soc* does not cause recent relational stress (*S.rel*) or financial stress (*S.fin*), and these stressors are likely related by a latent confounder.
-Additionally, *P.emp* and *S.fin* are identified as non-causes of housing precarity (*P.hou*).
-Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship.
+@fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
+As described in [@sec-analysis], dashed edges indicate ties, with the exact edge endpoint where the tie occurs is represented by a bold horizontal bar.
+
+The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (PHQsum).
+Employment precarity (P.emp) and social precarity (P.soc) are not identified as causes of depression, whereas financial stress (S.fin) appears to play a causal role.
+While P.emp and P.soc are generally not recognized as causes of other precariousness factors, S.fin emerges as a potential cause, as indicated by its circle edge endpoint.
+Supporting this, the proportion matrix plot in panel (b) shows that S.fin has some probability of causally influencing either P.emp or P.soc.
+
+The matrix plot can be interpreted such that the symbol in matrix[$i$, $j$] represents the relationship $i -~[symbol] \  j$.
+For example, if matrix[$i$, $j$] is > and matrix[j, i] is $\circ$, then the inferred relationship is $i \circarrow j$.
+In the table, different edge types are represented by distinct colors: light gray for the absence of an edge, blue for circles, green for arrowheads, and coral for arrowtails.
+The colors are blended based on the proportion of each edge type, with higher proportions increasing the opacity of the corresponding color, making dominant symbols more visually prominent.
+
+Relational stress (S.rel) plays a more nuanced role, interacting with depression and S.fin  through a latent confounder.
+Its relationship with P.soc remains less certain, as the proportion table suggests a fair probability that this connection is absent, uncertain, or bidirectional.
+Meanwhile, housing precarity (P.hou) is not directly causally related to depression but is linked to P.emp and S.fin.
+While P.emp and S.fin are identified as non-causes of P.hou, it remains unclear whether P.hou causally influences P.emp or if their relationship is mediated by an unobserved confounder.
+
+The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify S.fin as a cause of depression.
+Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI.
+Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.
+
+<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precarity  (P.emp) and social precarity (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.
+Additionally, P.emp and S.fin are identified as non-causes of housing precarity (P.hou).
+Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship.
+Similar ambiguities are found in the connections between S.rel and PHQsum and between S.fin and PHQsum -->
+
+<!-- Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship.
 Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*.
-
-Graph (b), derived solely from the nonparametric RCoT method.
+-->
+
+<!-- Graph (b), derived solely from the nonparametric RCoT method.
 Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*.
-The relationship between stressors and depression remains unresolved between the two algorithms.
+-->
+<!-- The relationship between stressors and depression remains unresolved between the two algorithms.
 However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
-The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.
-
-Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders.
+-->
+<!-- The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.
+-->
+
+<!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders.
 Employment and social precarity may also be influenced by depression, either directly or through latent variables.
 Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways.
+-->
 
@@
 ## Individual depression symptom
-Moving from the sum score representation to the symptom-level graph provides a more granular perspective on the causal relationships between precarity factors and depression.
-This approach highlights the heterogeneity in how precarity factors influence individual depressive symptoms — *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood).
-While the sum score graph aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom graph uncovers distinct pathways for different symptoms.
-In the symptom-level graph (@fig-sym), consistent relationships are represented by black solid edges, while areas of disagreement between the FCI and CCI algorithms are denoted by gray dashed edges.
-Endpoints marked with circles indicate differences in directional conclusions between the algorithms.
-Additionally, the navy dashed edges represent relationships unique to the graphs generated using Gaussian CI testing.
-
-<!-- Overall Patterns in the Symptom Network -->
-The symptom-level graph reveals a complex and interconnected structure, far more intricate than the sum score graph.
-Its denser network highlights the strong interdependence among symptoms and suggests the presence of latent confounding influences, as indicated by numerous bidirectional arrows.
-
-<!-- Centrality and Connectivity -->
-Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points.
+
+Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms.
+Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph highlights heterogeneity in how precariousness factors influence individual depressive symptoms: con (concentration), slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal), anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.
+
+The symptom-level graph in @fig-sym panel (a) reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges.
+As before, dashed edges denote ties, with specific ties marked by bold horizontal bars.
+In @fig-sym panel (b), the proportion matrix provides more insight into these directionalities.
+Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type.
+However, blue regions, which correspond to circle endpoints, are particularly common for anh, slp, ene, and sui, reflecting uncertainty in the causal direction among these symptoms.
+Notably, the only arrowtail connection appears between dep and anh, suggesting that anhedonia (anh) is a potential cause of depressed mood (dep).
+
+Looking at the symptom-precariousness connections, one of the key patterns is that ties are most frequently found in the relationships between individual symptoms and precariousness factors.
+This suggests that these causal links may be less stable across different conditions.
+Upon closer examination, most ties arise due to discrepancies between the Gaussian CI test and the RCoT test—with Gaussian CI favoring arrowheads and RCoT more often predicting the absence of an edge.
+
+This discrepancy likely arises from a combination of factors. 
+First, as the analysis shifts from aggregated to disaggregated variables, the number of conditional independence (CI) tests increases, leading to a loss of statistical power in high-dimensional causal discovery.
+With more variables being conditioned on, the data becomes sparser, making it harder to detect weaker dependencies.
+Additionally, over-conditioning—controlling for too many variables—can artificially remove statistical associations, leading to false negatives.
+Beyond this general statistical power issue, the fundamental differences in how these CI tests operate contribute to the inconsistency.
+RCoT, as a nonparametric test, does not assume linearity and can capture both linear and nonlinear dependencies.
+However, it is more conservative and requires larger sample sizes, making it more prone to false negatives in high-dimensional settings.
+In contrast, partial correlation (Gaussian CI test) assumes linear Gaussian relationships, making it more permissive and sometimes detecting weak dependencies that may not be statistically meaningful.
+As a result, when Gaussian CI detects an edge while RCoT does not, it could indicate that the relationship is weak, strictly linear, or requires more data for reliable nonparametric detection.
+This explains why many ties in the proportion matrix show a 50-50 split between symbols—where half of the conditions with RCoT suggest the absence of an edge, while the other half using Gaussian CI test favor an arrowhead.
+This systematic divergence highlights the challenges of causal discovery in high-dimensional settings, where varying statistical assumptions can lead to differing conclusions about the presence and direction of causal relationships.
+
+Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of financial stress (S.fin), which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality.
+One exception is the stronger evidence suggesting that S.fin may cause changes in app (appetite), while its relationships with other symptoms, such as ene, mot, and dep, remain ambiguous.
+Similarly, social precarity (P.soc) exhibits numerous connections with depressive symptoms, with most edges pointing toward P.soc rather than outward from it.
+This suggests that depressive symptoms, particularly dep and glt, may contribute to worsening social precarity rather than the other way around.
+Additionally, anh also shows some probability of causally influencing P.soc, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.
+
+Other precariousness factors exhibit more uncertain but still notable relationships.
+Relational stress (S.rel) shows weak but existing connections with slp, glt, and app, though directionality remains unclear in many cases.
+Employment precarity (P.emp) also connects to symptoms like slp, mot, and con, but, like S.rel, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
+In line with the aggregated graph in @fig-sum, housing precarity (P.hou) does not appear to have any direct relationship with depression symptoms but consistently shows associations with employment precarity (P.emp) and financial stress (S.fin).
+The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with financial stress (S.fin) exhibiting the strongest tendency to influence other precariousness factors.
+
+Among depressive symptoms, dep, glt, and slp appear to be the most connected to precariousness factors, while P.soc has the most connections with symptoms, predominantly as a recipient rather than a driver of influence.
+Within the symptom network, slp, sui, and anh emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms.
+On the other hand, con, mot, dep, and glt, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
+Considering both symptom-precariousness connections and symptom-level dynamics, sleep disturbance (slp) emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network.
+This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precarity.
+On the other hand, social precarity (P.soc) serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.
+
+Finally, the results from CCI are largely consistent with those from FCI, but with some key differences.
+CCI tends to favor arrowheads more frequently, resulting in a greater number of bidirectional edges, which suggests a higher involvement of latent confounders.
+Additionally, CCI exhibits more tie situations, though, similar to FCI, most ties occur between absence and arrowhead edges, reflecting the discrepancy between the Gaussian CI test and RCoT—where the Gaussian test favors arrowheads, while RCoT more often suggests the absence of an edge.
+For a detailed visualization of the CCI-derived graph and proportion matrix, see @fig-sym-cci.
+
+<!-- <!-- Centrality and Connectivity --> 
+<!-- Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points.
 Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network.
@@
 These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes.
-<!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. -->
-
-The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs.
+-->
+<!-- <!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. --> 
+<!-- The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs.
 As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precarity (*P.soc*) are more likely to be influenced by depressive symptoms.
-The symptom-level analysis, however, provides more specificity by pinpointing the symptoms involved in these relationships.
+-->
+
+<!-- The symptom-level analysis, however, provides more specificity by pinpointing the symptoms involved in these relationships.
 For instance, *slp* is identified as influencing *P.emp* or potentially through a latent confounder (*slp* o-> *P.emp*) , while *glt* appears to affect *P.soc* or may be linked via a confounder (*glt* o-> *P.soc*).
-Additional edges are observed in the graph generated using Gaussian CI tests, such as *slp* <-> *S.rel* and *slp* o-> *P.emp*.
+-->
+<!-- Additional edges are observed in the graph generated using Gaussian CI tests, such as *slp* <-> *S.rel* and *slp* o-> *P.emp*.
 As seen in the sum graph, Gaussian CI testing tends to produce a denser graph.
 In this case, causal relationships involving *slp* are particularly prominent.
-
-Some discrepancies, however, exist between the symptom-level and sum graphs.
+-->
+
+<!-- Some discrepancies, however, exist between the symptom-level and sum graphs.
 For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precarity factors.
 Additionally, housing precarity (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.
+-->
 
@@
 
-Lastly, we examine the relationships between individual symptoms and overall precarity, represented as the sum of five precariousness factors.
-This approach completes the analysis by providing a perspective on how individual symptoms relate to overall precarity.
-It complements the previous analysis, which focused on individual symptoms and individual precarity factors, by potentially capturing distributed relationships across different precarity factors through their aggregation into a single overall score.
-As before, black solid lines indicate edges agreed upon by both FCI and CCI, while gray dashed lines represent edges where the two algorithms disagree.
+Lastly, we examine the relationships between individual symptoms and overall precarity, an aggregated measure represented by the sum of five precariousness factors.
+This analysis provides a complementary perspective, capturing broader patterns that may not be evident in the disaggregated symptom-precariousness analysis.
+By aggregating precariousness factors into a single score, this approach may capture distributed relationships across different precarity factors that were previously overlooked in the disaggregated analysis.
+
+Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum panel (a)) shows more certainty in causal directions, with only a few edges resulting in ties, all of which involve the overall precarity factor.
+This is even more evident in the proportion matrix (@fig-presum panel (b)), where symptom-to-symptom interactions almost fully converge to a single edge type.
+Additionally, all symptom interactions are connected either through bidirectional edges or a combination of an arrowhead and a circle, suggesting a strong presence of latent confounders.
+
+Examining the symptom-precarity connections, we observe a stronger overall trend of depressive symptoms influencing precarity, rather than the other way around.
+Specifically, anh, dep, and slp show a high probability of causing precarity, whereas app, glt, and mot exhibit more uncertainty in directionality, often resulting in circle endpoints.
+This pattern suggests that when precariousness factors are aggregated, the dominant causal flow is from depressive symptoms to precarity, rather than precarity driving depression.
+Particularly, dep emerges as the strongest predictor of precarity, exhibiting the highest proportion of arrowtails, consistent with findings from the disaggregated analysis.
+
+Comparing this with the CCI-derived graph (@fig-presum-cci), several key differences stand out.
+The CCI results reveal that most symptoms are interconnected through bidirectional edges, suggesting a strong presence of latent confounders.
+However, dep appears to be a distinct exception, as it is predominantly caused by nearly all other symptoms, with one exception sui, which does not contribute to dep.
+Similarly, con is influenced by multiple symptoms, albeit with weaker support compared to dep.
+Most interestingly, the CCI results highlight a distinct feedback loop between dep and overall precarity, represented by a tail-tail edge (\textemdash \).
+This suggests a possible reinforcing cycle between depression and overall precariousness, where depression not only arises from precarious conditions but also contributes to their persistence, creating a self-sustaining dynamic.
+
+Overall, the aggregated precariousness graph supports the key findings from both the disaggregated analysis and the depression sum score analysis, providing greater certainty in several important patterns.
+The results highlight the significant role of latent confounders in symptom-to-symptom interactions.
+While certain stressor-related precariousness factors, particularly S.fin and S.rel, appear to causally influence depressive symptoms, other precariousness factors are more likely driven by depression rather than acting as primary causes, suggesting that precariousness may function more as a consequence of depression than its source.
+Some symptoms, particularly glt, slp, and dep, show greater sensitivity to precarious conditions, while also playing a central role in symptom dynamics.
+Among them, slp and glt emerge as influential symptoms within the depression network, whereas con and dep, despite their high connectivity, behave more as outcomes rather than primary drivers.
+Together, these findings suggest that depression—especially symptoms like sleep disturbances and guilt—may contribute to sustaining precarious conditions, possibly forming a reciprocal cycle between depression and precariousness.
+
+<!-- As before, black solid lines indicate edges agreed upon by both FCI and CCI, while gray dashed lines represent edges where the two algorithms disagree.
 Circles at the endpoints of gray dashed edges denote unresolved directionality.
 Additionally, navy dashed edges highlight relationships that appear exclusively in graphs generated using Gaussian CI tests.
-
-Compared to @fig-sym, this analysis reveals greater disagreement between FCI and CCI regarding edge directionality, as evidenced by more dashed edges.
+-->
+<!-- Compared to @fig-sym, this analysis reveals greater disagreement between FCI and CCI regarding edge directionality, as evidenced by more dashed edges.
 FCI tends to estimate more undirected edges (circles), while CCI more frequently assigns directed arrows between symptom-symptom relationships.
 For a detailed breakdown, see [@sec-propmatrix], which shows the proportion of symbols produced by each algorithm.
-When examining overall precarity, we observe that more symptoms are now connected to the precarity score (*Precarity*) compared to the individual precarity factor analysis in @fig-sym.
+-->
+<!-- When examining overall precarity, we observe that more symptoms are now connected to the precarity score (*Precarity*) compared to the individual precarity factor analysis in @fig-sym.
 In addition to *glt*, *slp*, *anh*, and *sui*, which are previously shown to be connected to individual precarity factors, *mot*, *app*, and *dep* are now identified as having causal relationships with overall precarity.
@@
 The exception is the bidirectional edge between *sui* <-> *Precarity*, which suggests a connection via latent confounders.
-FCI predominantly estimates directed arrows (→) for symptom-precarity relationships, suggesting that symptoms are driving precarity.
+-->
+<!-- FCI predominantly estimates directed arrows (→) for symptom-precarity relationships, suggesting that symptoms are driving precarity.
 With fewer variables (reduced from 14 to 9), this approach appears to have slightly more statistical power, enabling algorithms to estimate relationships with greater confidence compared to the previous analysis involving individual symptoms and individual precarity factors.
+-->
 
@@
 
-The present study examined the causal relationships between precarity factors and depression using causal discovery algorithms, incorporating both sum score and symptom-level analyses.
-By analyzing PHQ sum scores alongside individual symptom scores, and comparing aggregated precarity scores with individual precarity factors, we gained a comprehensive perspective on how different aspects of precarity influence depression.
-This dual approach highlighted nuanced pathways that may be overlooked when focusing solely on aggregate measures, while also capturing broader interactions that might remain undetected in analyses limited to individual factors.
+This study examined the causal relationships between precariousness factors and depression using causal discovery algorithms, integrating both sum score and symptom-level analyses.
+By analyzing PHQ sum scores alongside individual symptom scores, and comparing aggregated precarity scores with individual precariousness factors, we gained a comprehensive perspective on how different aspects of precariousness influence depression.
+This dual approach revealed nuanced pathways that may be overlooked when relying solely on aggregate measures, while also capturing broader interactions that might remain undetected in analyses focused on individual factors.
 
 <!-- Interpretation of Findings -->
-Our findings highlight the significant role of recent relational stress (*S.rel*) as a potential causal factor for depression, consistently observed across both sum score and symptom-level analyses.
-This consistency underscores the profound impact interpersonal relationships can have on mental health.
-The symptom-level analysis further identified specific symptoms, such as sleep disturbances (*slp*), guilt (*glt*), and anhedonia (*anh*), as particularly sensitive to external precarity conditions.
-These symptoms emerged as potential initiators or activators within the depressive symptom network, suggesting that they may serve as early warning signs or valuable targets for preventive interventions.
-
-Employment precarity (*P.emp*) and social precarity (*P.soc*) were more likely to be influenced by depressive symptoms rather than acting as direct causes.
+Our findings highlight the significant role of recent financial stress (S.fin) as a potential causal factor for depression, consistently observed across both sum score and symptom-level analyses.
+This consistency underscores the profound impact of financial stressors on mental health.
+The symptom-level analysis further identified specific symptoms—particularly sleep disturbances (slp), depressed mood (dep), guilt (glt), and anhedonia (anh)—as being especially sensitive to external precarious conditions.
+
+Among these symptoms, dep stands out as a distinct causal factor for precarity in both aggregated and disaggregated analyses, with CCI even suggesting a possible cyclic relationship between them.
+This implies that depression and precariousness may reinforce one another, potentially leading to a worsening cycle over time.
+Within the symptom network, dep also exhibits a high degree of connectivity, but with mostly incoming arrows, suggesting that as symptom dynamics intensify, dep is quickly impacted by multiple factors—particularly P.soc (social precarity), further fueling the depression-precariousness cycle.
+
+Beyond dep, slp emerges as a critical symptom in this dynamic.
+It connects with both social and employment precariousness, while also playing a central role in symptom interactions, affecting other symptoms more than most others.
+This suggests that slp may act as an initiator within the depressive symptom system, potentially triggering or amplifying symptom cascades.
+
+Among the precariousness factors, P.soc appears to be one of the most pivotal—not as a driver of symptoms, but rather as a consequence of depression.
+It maintains the strongest connectivity with symptoms, yet its directionality indicates that it is largely shaped by depressive symptoms rather than acting as a primary cause.
+Together, these patterns suggest a broader directional dynamic in which financial stress (S.fin) exacerbates depressive symptoms, which in turn feed into social precarity, perpetuating the cycle.
+
+This directional insight suggests that interventions targeting specific depressive symptoms, particularly dep, may help prevent downstream effects on worsening precarious conditions.
+Alternatively, early intervention focusing on slp may be a promising strategy, as slp appears to act as a potential initiator or activator within the depressive symptom network.
+This positioning suggests that slp could serve as an early warning sign, offering a valuable target for preventive interventions aimed at breaking the depression-precariousness cycle before it intensifies.
+
+<!-- emerged as potential initiators or activators within the depressive symptom network, suggesting that they may serve as early warning signs or valuable targets for preventive interventions. -->
+
+<!-- Employment precarity (*P.emp*) and social precarity (*P.soc*) were more likely to be influenced by depressive symptoms rather than acting as direct causes.
 The symptom-level graph provided additional clarity by identifying that sleep disturbances influenced employment precarity (*slp* o-> *P.emp*) and feelings of guilt affected social precarity (*glt* o-> *P.soc*).
 This directional insight suggests that interventions targeting specific depressive symptoms may have downstream effects on improving certain aspects of precarity.
-
-Interestingly, financial stress (*S.fin*) did not exhibit causal relationships with individual depressive symptoms in the symptom-level analysis, contrasting with its apparent role in the sum score graphs.
+-->
+
+<!-- Interestingly, financial stress (*S.fin*) did not exhibit causal relationships with individual depressive symptoms in the symptom-level analysis, contrasting with its apparent role in the sum score graphs.
 This divergence indicates that while financial stress may influence overall depression severity, its impact on specific symptoms may not be significant.
-Also, housing precarity (*P.hou*) emerged as an isolated node in the symptom-level graph, losing all connections with other precarity factors.
+-->
+<!-- Also, housing precarity (*P.hou*) emerged as an isolated node in the symptom-level graph, losing all connections with other precarity factors.
 This isolation suggests that housing precarity may operate independently of the depressive symptom network or that its effects are not captured within the scope of the measured variables.
+-->
 
@@
 While the resulting graphs are preliminary and contain unresolved ambiguities, they offer a valuable starting point for leveraging causal discovery tools to investigate the causal interplay between depression and precarity factors.
-A promising next step would involve integrating these causal structures into computational models, such as the symptom dynamic model proposed in **our comp-model paper**.
+A promising next step would involve integrating these causal structures into computational models, such as the symptom dynamic model proposed in @park2025individual.
 By simulating intervention effects, such models could provide more realistic insights into how targeted actions might influence symptom networks and precarity factors over time.
@@
 
-## Results from CCI algorithm
+## Results from CCI algorithm {#sec-cci}
 

### 2025-02-11 draft_v1/draft_v1.tex
update_results
only FCI in main text + proportion matrix plots

@@
 
-\date{2025-02-08}
+\date{2025-02-11}
 \begin{document}
@@
 
-Mental health problems in urban areas have been reported to be on the
-rise the complexity of mental health systems presents significant
-challenges in understanding the underlying mechanisms driving these
-issues, let alone planning effective interventions.
-
-Research has aimed to identify underlying factors contributing to mental
-health problems. Recent research looked into the association between
-mental health and indicators of precariousness in various dimensions of
-life, indicating a high level of uncertainty and instability in people's
-lives, such as factors related to employment, social relations,
-finances, housing, and culture (ref Leonie's paper). This comprehensive
-perspective on precariousness helped to highlight how different aspects
-of life may be interconnected. While this research has advanced our
-understanding of which indicators of precariousness are related to
-mental health, a key question remains unanswered: how do these factors
-influence mental health? Specifically, the lack of directional
-information --- knowing what influences what --- limits our ability to
-identify and prioritize effective intervention targets.
+Mental health problems in urban areas are on the rise, yet the
+complexity of mental health systems poses significant challenges in
+understanding the underlying mechanisms driving these issues --- let
+alone designing effective interventions. Research has long sought to
+identify underlying factors contributing to mental health problems.
+Recent research looked into the association between mental health and
+indicators of precariousness in various dimensions of life, indicating a
+high level of uncertainty and instability in people's lives, such as
+factors related to employment, social relations, finances, housing, and
+culture (ref Leonie's paper). This comprehensive perspective on
+precariousness helped to highlight how different aspects of life may be
+interconnected. While this research has advanced our understanding of
+which indicators of precariousness are related to mental health, a key
+question remains unanswered: how do these factors influence mental
+health? Specifically, the lack of directional insight --- knowing what
+influences what --- limits our ability to identify and prioritize
+effective intervention targets.
 
@@
 with a specific focus on depression, the most prevalent mental health
-issue. Using causal discovery methods, we explore how different aspects
-of precariousness influence depression and delve deeper into the
-dynamics at the symptom level. By examining individual depressive
-symptoms, we aim to identify which symptoms may act as initiators by
-being particularly sensitive to precariousness. Through this analysis,
-our goal is to uncover the causal mechanisms underlying mental health
-challenges and provide a foundation for developing more effective and
-targeted interventions.
+issue. Using causal discovery methods, we explore how various aspects of
+precariousness contribute to depression and whether depression, in turn,
+exacerbates precariousness, potentially creating a feedback loop.
+Furthermore, we zoom in at the symptom level, examining individual
+depressive symptoms to identify those that may act as initiators ---
+triggering the activation of the symptom network --- by displaying
+heightened sensitivity to specific precariousness factors. We also
+investigate symptoms that may function as bridging elements, linking the
+dynamics of depression with precariousness by feeding back into
+precarious conditions, potentially reinforcing the cycle of mental
+health deterioration and precarious conditions. Through this analysis,
+we aim to uncover the causal mechanisms driving mental health challenges
+related to precariousness and establish a foundation for more targeted
+and effective interventions by identifying the key driving forces within
+this system.
 
@@
 this study, we focus on algorithms suited to the potential cyclic
-relationships within our system. Specifically, we use FCI (Fast Causal
-Inference) and CCI (Cyclic Causal Inference), both capable of accounting
-for such cycles under certain conditions (Mooij \& Claassen, 2020;
-Strobl, 2019). Additionally, we include the PC algorithm as a reference,
-given its simplicity and prominence as one of the most widely known
-causal discovery methods (Spirtes et al., 2001).
+relationships within our system. Specifically, we use \emph{FCI} (Fast
+Causal Inference) and \emph{CCI} (Cyclic Causal Inference), both capable
+of accounting for such cycles (Mooij \& Claassen, 2020; Strobl, 2019).
+Additionally, we include the \emph{PC} algorithm as a reference, given
+its simplicity and prominence as one of the most widely known causal
+discovery methods (Spirtes et al., 2001).
 
@@
 almost ancestral graph) due to their reliance on different underlying
-assumptions. Despite these differences, both graphs are forms of
-\emph{ancestral graphs}, designed to encode causal relationships between
-variables, where the presence of an edge indicates causal
-\emph{ancestry.} Directed edges, \(A \stararrow B\), indicate that \(B\)
-is not an ancestor of \(A\) in every graph within the Markov equivalence
-class, \(Equiv(G)\). The Markov equivalence class refers to the set of
-graphs that encode the same conditional independence relationships, such
-that the same \emph{d-separation} relations hold across all graphs in
-the class (Spirtes et al., 2001). \(A \startail B\) indicates that \(B\)
-is an ancestor of \(A\) across all graphs in \(Equiv(G)\). Circle
-endpoints, \(A \starcirc B\), represent ambiguity in the ancestral
-relationship, meaning \(B\)'s ancestral status relative to \(A\) varies
-across graphs in \(Equiv(G)\).\footnote{\(*\) serves as a
-  \textit{meta-symbol}, representing one of the three possible
+assumptions. Despite these differences, both graphs belong to the class
+of \emph{ancestral graphs}, which are designed to encode causal
+relationships between variables, where the presence of an edge indicates
+causal \emph{ancestry.} In these graphs, directed edges,
+\(A \stararrow B\), indicate that \(B\) is not an ancestor of \(A\) in
+every graph within the Markov equivalence class, \(Equiv(G)\). The
+Markov equivalence class represents a set of graphs that encode the same
+conditional independence relationships, ensuring that the same
+\emph{d-separation} conditions hold across all graphs in the class
+(Spirtes et al., 2001). Conversely, an edge marked as \(A \startail B\)
+indicates that \(B\) is an ancestor of \(A\) across all graphs in
+\(Equiv(G)\). Circle endpoints, \(A \starcirc B\), represent ambiguity
+in the ancestral relationship, meaning \(B\)'s ancestral status relative
+to \(A\) varies across graphs in \(Equiv(G)\).\footnote{\(*\) serves as
+  a \textit{meta-symbol}, representing one of the three possible
   edge-endpoints. For instance, \(A \tailstar B\) can indicate any of
   the following edges: \(A\) \textemdash~\(B\), \(A\tailarrow B\), or
-  \(A \tailcirc B\) (Park et al., 2024).} When \(A \arrowarrow B\),
-neither \(A\) nor \(B\) is an ancestor of the other, indicating the
-presence of a latent confounder between them.
-
-The algorithms also differ in their approaches and assumptions for
-detecting cycles. In CCI's MAAG, \(A\) \textemdash~\(B\) indicates that
-\(A\) is an ancestor of \(B\) and \(B\) is an ancestor of \(A\),
-referring to a cycle between \(A \stackedarrows B\). FCI, on the other
-hand, identifies potential cycles more subtly: fully-connected nodes
-with circle endpoints (\(\circirc\)) in FCI's PAG may suggest a cyclic
-structure. FCI basically provides a sufficient condition to distinguish
-nodes that are not part of a cycle, offering a more nuanced approach
-(Mooij \& Claassen, 2020). Table 1 highlights further differences in the
-assumptions underlying these algorithms. CCI operates under the
-assumption of a linear system, while FCI, particularly for inferring
-cyclic relationships, assumes a non-linear system without selection bias
-and adheres to more general faithfulness and Markov conditions (i.e.,
+  \(A \tailcirc B\) (Park et al., 2024).} Finally, when an edge is
+represented as \(A \arrowarrow B\), it implies that neither \(A\) nor
+\(B\) is an ancestor of the other, suggesting the presence of a latent
+confounder influencing both variables.
+
+The algorithms also differ in how they detect cycles and represent
+cyclic relationships in their resulting graphs. In CCI's MAAG, \(A\)
+\textemdash~\(B\) indicates that \(A\) is an ancestor of \(B\), and
+simultaneously \(B\) is an ancestor of \(A\), referring to a cyclic
+relationship between \(A \stackedarrows B\). FCI, on the other hand,
+identifies potential cycles more subtly. In its PAG, fully-connected
+nodes with circle endpoints (\(\circirc\)) may suggest the presence of
+cyclic structures. FCI, therefore, provides a sufficient condition to
+distinguish variables that are not part of a cycle, offering a more
+nuanced approach to handling cyclic relationships (Mooij \& Claassen,
+2020).
+
+Table 1 highlights further differences in the assumptions underlying
+these algorithms. CCI operates under the assumption of a linear system,
+whereas FCI, particularly when used to infer cyclic relationships,
+assumes a non-linear system without selection bias and adheres to more
+general faithfulness and Markov conditions (i.e.,
 \emph{\(\sigma\)-separation} and \emph{\(\sigma\)-faithfulness} setting)
 (Forré \& Mooij, 2018). When the respective assumptions of each
-algorithm are satisfied, their outcomes for cyclic relationships align
-with those illustrated in Figure~\ref{fig-examplegraphs}. While an
-exhaustive discussion of causal discovery concepts and
-algorithm-specific details is beyond the scope of this paper, readers
-seeking a deeper understanding are encouraged to consult Park et al.
-(2024) for comprehensive insights into these methods and their
-applications.
+algorithm are met, their inferred cyclic relationships align with those
+illustrated in Figure~\ref{fig-examplegraphs}.
+
+While CCI's MAAG provides an explicit representation of cyclic
+relationships, it has certain theoretical limitations, as the MAAG may
+not always fully preserve d-separation relations from the original
+graph, \(\mathcal{G}\) (Strobl, 2019). Given these considerations, we
+examine and compare results from both algorithms, placing more weight on
+the FCI results, which are presented in the main results section, while
+discussing CCI results where relevant, with detailed findings provided
+in the \hyperref[sec-appendix]{Appendix}. A full discussion of causal
+discovery concepts and algorithmic details is beyond the scope of this
+paper; however, readers seeking a more in-depth understanding can refer
+to Park et al. (2024) for a comprehensive exploration of these methods
+and their applications.
 
@@
 this uncertainty with bidirectional arrows. While the PC algorithm
-serves as a useful reference, its strict assumptions of acyclicity and
-the absence of latent confounders restrict its applicability in more
-complex scenarios. Therefore, our primary focus remains on the results
-from FCI and CCI, with all PC algorithm results provided in the
-\hyperref[sec-appendix]{Appendix} for completeness.
+serves as a useful reference, its strict assumptions --- acyclicity and
+the absence of latent confounders---limit its applicability in more
+complex scenarios. For this reason, our primary focus remains on the
+results obtained from FCI and CCI, with all PC algorithm results
+provided in the \hyperref[sec-appendix]{Appendix} for completeness.
 
 One practical challenge in applying these algorithms to the HELIUS
-dataset is that the data does not follow a Gaussian distribution, and
-the relationships among variables are unlikely to be strictly linear. To
-account for this, we supplement the commonly used Gaussian conditional
-independence (CI) test, which relies on partial correlations, with a
-non-parametric CI test based on reproducing kernel methods (Zhang et
-al., 2012). While kernel-based conditional independence tests (KCIT) are
-effective, they are computationally intensive for large datasets like
-HELIUS, as their quadratic scaling with sample size is primarily driven
-by the inversion of large kernel matrices. To mitigate this issue, we
-use the Randomized Conditional Correlation Test (RCoT), which
-approximate kernel methods using random Fourier features. This approach
-reduces computation to linear scaling with sample size, significantly
-lowering computational costs (Strobl et al., 2019). For a more detailed
+dataset is that most variables do not follow a Gaussian distribution,
+and the relationships among variables are unlikely to be strictly
+linear. To address this, we supplement the commonly used Gaussian
+conditional independence (CI) test, which relies on partial
+correlations, with a non-parametric CI test based on reproducing kernel
+methods (Zhang et al., 2012). While kernel-based conditional
+independence tests (KCIT) are theoretically effective in capturing
+non-linear dependencies and subtle interactions in data with
+non-Gaussian noise, they are computationally demanding for large
+datasets like HELIUS. Their quadratic scaling with sample size is
+primarily due to the need to invert large kernel matrices, making them
+impractical for large-scale analysis (Rahimi \& Recht, 2007). To
+mitigate this issue, we use the Randomized Conditional Correlation Test
+(RCoT), which approximates kernel methods using random Fourier features.
+This approach reduces computational complexity from quadratic to linear
+scaling with sample size, significantly lowering computational costs
+while maintaining robustness (Strobl et al., 2019). For a more detailed
 explanation of RCoT, refer to Section~\ref{sec-rcot}.
 
-\subsection{Analysis}\label{analysis}
-
-We analyze the causal structure using three approaches: examining the
-relationship (1) between the five individual precarity factors and the
-PHQ sum score, which represents overall depression severity; (2) between
-individual symptoms and precarity factors; and (3) between individual
-symptoms and overall precarity (the sum of the five precarity factors).
-Using the PHQ sum score simplifies the analysis by reducing
-dimensionality, offering computational efficiency and a broad,
-interpretable perspective on the relationship between depression and
-precarity factors. In contrast, analyzing individual symptoms provides a
-more detailed understanding by capturing the diverse ways symptoms
-respond to different precarity factors. However, this detailed approach
-introduces methodological challenges, including non-standard
-distributions of individual symptom variables and the complexities
-associated with increased dimensionality. Finally, examining individual
-symptoms in relation to the overall precarity score complements the
-analysis by revealing how symptoms collectively connect to general
-precarity. Aggregating precarity factors may uncover relationships that
-could be overlooked in analyses of individual factors, particularly when
-a symptom has associations across multiple factors that are not strong
-enough to emerge in individual analyses. By integrating these three
-approaches, we achieve a balance between capturing detailed granularity
-and gaining a broader understanding of the relationship between
-precarity and depression.
+\subsection{Analysis}\label{sec-analysis}
+
+We analyze the causal structure using three complementary approaches:
+(1) Examining the relationship between the five individual
+precariousness factors and the PHQ sum score, which represents overall
+depression severity. (2) Investigating the relationship between
+individual depressive symptoms and individual precariousness factors.
+(3) Assessing the relationship between individual depressive symptoms
+and overall precariousness (the sum of all precariousness factors).
+Using the PHQ sum score as a measure of depression simplifies the
+analysis by reducing dimensionality, improving computational efficiency,
+and providing a broad, interpretable perspective on how precariousness
+factors relate to overall depression severity. However, this aggregated
+approach may overlook granular symptom-level relationships.
+
+By zooming in on individual symptoms, we capture a more detailed picture
+of how specific depressive symptoms interact with different
+precariousness factors. This approach allows us to map individual causal
+pathways, identifying which symptoms are particularly sensitive to
+certain precarious conditions and vice versa. However, this level of
+detail introduces methodological challenges, including non-standard
+distributions of symptom variables and the high-dimensional nature of
+the analysis, which can reduce statistical power and significantly
+increase computational demands---especially when using non-parametric
+conditional independence (CI) tests.
+
+Finally, analyzing individual symptoms in relation to overall
+precariousness provides an integrative perspective, revealing how
+depression symptoms collectively relate to precarious conditions. This
+approach complements the fully disaggregated analysis by aggregating
+precariousness factors, potentially uncovering relationships that may be
+too weak to detect when examined separately. This is particularly
+relevant when a symptom is associated with multiple precariousness
+factors, but these associations are not strong enough to emerge in an
+individual-level analysis. By integrating these three approaches ---
+ranging from partially aggregated to fully disaggregated --- we aim to
+achieve a balance between capturing detailed granularity and gaining a
+broader understanding of the relationship between precarity and
+depression.
 
@@
 robustness of our results, we implement multiple settings combined with
-bootstrapping. For each of the 100 bootstrap samples, we estimate causal
-graphs and retain only the edges and directions that exceed predefined
-thresholds. The analyses are conducted under the following conditions:
+bootstrapping. For each condition, we generate 100 bootstrap samples,
+estimate causal graphs, and retain only the edge-endpoints that exceed
+predefined thresholds. If no edge-endpoint type surpasses the threshold,
+we assign a circle (\(\circ\)) to indicate uncertainty in the edge
+direction.
+
+The analyses are conducted under the following conditions:
 
@@
 the number of bootstrap samples to 30, easing computational demands. To
-further enhance efficiency, we fix the skeleton of edges among symptom
-variables based on the common structure estimated across all algorithms.
-This approach minimizes spurious edges in the symptom network and
-significantly reduces the computational time needed for skeleton
-estimation. This fixed structure also aligns well with commonly reported
-skeleton structures in the literature, ensuring consistency with
-existing findings (\textbf{cite the comp\_model paper}).\footnote{We
-  fixed the skeleton of the symptom network using a common structure
-  derived from the PC, FCI, and CCI algorithms with an alpha level of
-  0.001. This approach preserved the complexity of symptom interactions,
-  enabling unrestricted estimation of causal directions both among
-  symptoms (\emph{symptom --- symptom} interactions) and between
-  symptoms and precarity factors.} Lastly, to summarize the results, we
-identify the most frequently occurring edge endpoints across different
-experimental setups.\footnote{The detailed proportion of each edge
-  endpoint occurrence is provided in \textbf{?@sec-propmatrix}.} This
-ensures that only stable and consistent edges are retained, providing a
-clearer and more reliable understanding of the relationships between
-precarity factors and depression. See Figure~\ref{fig-workflow} for an
-overview of the analysis workflow.
+further enhance efficiency, we fix the skeleton of the symptom network
+using a common structure derived from PC, FCI, and CCI algorithms
+(\(\alpha\) = 0.01, RCoT CI test). This step constrains which edges can
+be present among symptoms but does not restrict the estimation of causal
+directions within the symptom network. Additionally, the entire causal
+structure between symptoms and precariousness factors remains fully
+estimable, ensuring that key relationships are still inferred. Beyond
+its computational benefits --- particularly in reducing the time
+required for skeleton estimation --- this fixed structure also aligns
+well with commonly reported symptom network structures in the literature
+(Park et al., 2025), ensuring consistency with existing findings.
+
+To obtain a final, stable summary of causal relationships, we identify
+the most frequently occurring edge-endpoints across different
+experimental setups. First, within each condition, we determine the most
+frequent edge-endpoint from all bootstrap samples. Then, across
+different conditions, we aggregate these selected symbols and determine
+the overall most common edge-endpoint for each graph entry.
+
+Since we have an even number of experimental setups, ties may
+occasionally occur. In such cases, we explicitly denote them in the
+resulting graph using dashed lines. To further help with the
+interpretation and understanding of the resulting graph, we provide a
+corresponding table for each graph, showing the proportion of each
+edge-endpoint type to clarify how dominant a given symbol is in each
+case.
+
+By double summarizing the results in this way, we generate one final
+graph per algorithm for each analysis approach: the partially aggregated
+analysis, which uses either the sum of precariousness factors or the sum
+of depression factors to capture overall trends, and the fully
+disaggregated analysis, which examines individual precariousness and
+depression factors separately. This approach ensures that only stable
+and consistent edges --- those that persist across different parameter
+settings --- are retained, leading to a robust and reliable
+understanding of the relationships between precariousness factors and
+depression. See Figure~\ref{fig-workflow} for an overview of the
+analysis workflow.
 
@@
 aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
-relationships between precarity factors (\emph{P.hou}, \emph{P.emp},
-\emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the depression sum score
-(\emph{PHQsum}) under two different setups: (a) using both Gaussian CI
-test and RCoT, and (b) using RCoT alone. Black edges represent
-consistent causal relationships identified by both FCI and CCI, while
-gray dashed edges denote inconsistent relationships that vary between
-the two methods. The inconsistent endpoints of the gray edges are marked
-with circles.
-
-In graph (a), the key pathways suggest that employment precarity
-(\emph{P.emp}) and social precarity (\emph{P.soc}) do not cause
-depression (\emph{PHQsum}). \emph{P.soc} does not cause recent
-relational stress (\emph{S.rel}) or financial stress (\emph{S.fin}), and
-these stressors are likely related by a latent confounder. Additionally,
-\emph{P.emp} and \emph{S.fin} are identified as non-causes of housing
-precarity (\emph{P.hou}). Both FCI and CCI detect a dependency between
-\emph{P.emp} and \emph{S.fin}, but they disagree on the direction of the
-relationship, leaving it unclear whether \emph{S.fin} causes
-\emph{P.emp} or whether a latent confounder mediates their relationship.
-Similar ambiguities are found in the connections between \emph{S.rel}
-and \emph{PHQsum} and between \emph{S.fin} and \emph{PHQsum}.
-
-Graph (b), derived solely from the nonparametric RCoT method. Both
-graphs consistently identify that \emph{P.emp} is not a cause of
-\emph{P.hou} or \emph{PHQsum}, and that \emph{P.soc} does not cause
-\emph{PHQsum} or \emph{S.fin}. The relationship between stressors and
-depression remains unresolved between the two algorithms. However, the
-graph derived from RCoT provides greater confidence that depression is
-not the cause of \emph{S.fin} and suggests a stronger likelihood that
-the financial stress may contribute to depression or that their
-relationship is mediated by a latent confounder. The role of relational
-stress has become less pronounced, as the edge between \emph{P.soc} and
-\emph{S.rel} is omitted, and the relationship between \emph{S.rel} and
-\emph{S.fin} is now inconsistent between the algorithms.
-
-Overall, both graphs together highlight the potential roles of stressors
-like \emph{S.fin} and \emph{S.rel}, which are closely tied to
-depression, either causally or via latent confounders. Employment and
-social precarity may also be influenced by depression, either directly
-or through latent variables. Lastly, housing precarity does not directly
-impact depression but remains connected to employment precarity and
-financial stress, likely mediated by direct or latent pathways.
+relationships between precariousness factors (\emph{P.hou},
+\emph{P.emp}, \emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the
+depression sum score (\emph{PHQsum}), as identified by the FCI
+algorithm. As described in Section~\ref{sec-analysis}, dashed edges
+indicate ties, with the exact edge endpoint where the tie occurs is
+represented by a bold horizontal bar.
+
+The causal graph in panel (a) highlights key pathways in the
+relationships between precarity factors and depression (PHQsum).
+Employment precarity (P.emp) and social precarity (P.soc) are not
+identified as causes of depression, whereas financial stress (S.fin)
+appears to play a causal role. While P.emp and P.soc are generally not
+recognized as causes of other precariousness factors, S.fin emerges as a
+potential cause, as indicated by its circle edge endpoint. Supporting
+this, the proportion matrix plot in panel (b) shows that S.fin has some
+probability of causally influencing either P.emp or P.soc.
+
+The matrix plot can be interpreted such that the symbol in
+matrix{[}\(i\), \(j\){]} represents the relationship
+\(i -~[symbol] \  j\). For example, if matrix{[}\(i\), \(j\){]} is
+\textgreater{} and matrix{[}j, i{]} is \(\circ\), then the inferred
+relationship is \(i \circarrow j\). In the table, different edge types
+are represented by distinct colors: light gray for the absence of an
+edge, blue for circles, green for arrowheads, and coral for arrowtails.
+The colors are blended based on the proportion of each edge type, with
+higher proportions increasing the opacity of the corresponding color,
+making dominant symbols more visually prominent.
+
+Relational stress (S.rel) plays a more nuanced role, interacting with
+depression and S.fin through a latent confounder. Its relationship with
+P.soc remains less certain, as the proportion table suggests a fair
+probability that this connection is absent, uncertain, or bidirectional.
+Meanwhile, housing precarity (P.hou) is not directly causally related to
+depression but is linked to P.emp and S.fin. While P.emp and S.fin are
+identified as non-causes of P.hou, it remains unclear whether P.hou
+causally influences P.emp or if their relationship is mediated by an
+unobserved confounder.
+
+The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
+largely consistent with those from the FCI algorithm, with one
+exception: CCI does not identify S.fin as a cause of depression.
+Additionally, CCI introduces greater uncertainty in edge directions,
+with more ties mainly between circle and arrowhead endpoints, and tends
+to favor arrowheads more frequently than FCI. Despite these differences,
+the skeleton structure and overall causal directions derived from CCI
+align well with the results of the FCI algorithm, supporting the key
+pathways.
 
@@
 
-Moving from the sum score representation to the symptom-level graph
-provides a more granular perspective on the causal relationships between
-precarity factors and depression. This approach highlights the
-heterogeneity in how precarity factors influence individual depressive
-symptoms --- \emph{slp} (sleep), \emph{ene} (energy), \emph{app}
-(appetite), \emph{mot} (motor), \emph{sui} (suicidal), \emph{anh}
-(anhedonia), \emph{glt} (guilt), and \emph{dep} (depressed mood). While
-the sum score graph aggregates all symptoms into a single
-measure---potentially obscuring nuanced relationships---the symptom
-graph uncovers distinct pathways for different symptoms. In the
-symptom-level graph (Figure~\ref{fig-sym}), consistent relationships are
-represented by black solid edges, while areas of disagreement between
-the FCI and CCI algorithms are denoted by gray dashed edges. Endpoints
-marked with circles indicate differences in directional conclusions
-between the algorithms. Additionally, the navy dashed edges represent
-relationships unique to the graphs generated using Gaussian CI testing.
-
-The symptom-level graph reveals a complex and interconnected structure,
-far more intricate than the sum score graph. Its denser network
-highlights the strong interdependence among symptoms and suggests the
-presence of latent confounding influences, as indicated by numerous
-bidirectional arrows.
-
-Certain nodes in the network emerge as more \emph{causally} central,
-highlighting their potential importance as intervention points. Among
-the depressive symptoms, \emph{anh}, \emph{dep}, \emph{slp}, and
-\emph{glt} emerge as particularly influential, with multiple outgoing
-edges (o-\textgreater) to other symptoms, suggesting their roles as
-potential key drivers within the symptom network. Especially, symptoms
-such as \emph{glt}, \emph{slp}, and \emph{anh} are particularly
-critical, potentially acting as initiator or activator nodes within the
-network due to their apparent connections with precarity factors. These
-symptoms appear to be especially sensitive to external stressors,
-potentially manifesting early in response to such conditions and
-subsequently activating other interconnected symptom nodes.
-
-The causal structure involving precarity factors in the symptom-level
-graph is largely consistent with the patterns observed in the sum score
-graphs. As in the sum graphs, relational stress (\emph{S.rel}) emerges
-as a potential causal factor for depression, while employment
-(\emph{P.emp}) and social precarity (\emph{P.soc}) are more likely to be
-influenced by depressive symptoms. The symptom-level analysis, however,
-provides more specificity by pinpointing the symptoms involved in these
-relationships. For instance, \emph{slp} is identified as influencing
-\emph{P.emp} or potentially through a latent confounder (\emph{slp}
-o-\textgreater{} \emph{P.emp}) , while \emph{glt} appears to affect
-\emph{P.soc} or may be linked via a confounder (\emph{glt}
-o-\textgreater{} \emph{P.soc}). Additional edges are observed in the
-graph generated using Gaussian CI tests, such as \emph{slp}
-\textless-\textgreater{} \emph{S.rel} and \emph{slp} o-\textgreater{}
-\emph{P.emp}. As seen in the sum graph, Gaussian CI testing tends to
-produce a denser graph. In this case, causal relationships involving
-\emph{slp} are particularly prominent.
-
-Some discrepancies, however, exist between the symptom-level and sum
-graphs. For example, financial stress (\emph{S.fin}) does not have any
-edges with depressive symptoms in the symptom-level graph, although it
-maintains associations with other precarity factors. Additionally,
-housing precarity (\emph{P.hou}) becomes entirely disconnected from the
-rest of the network, appearing as an isolated node.
+Moving from the sum score representation to the disaggregated
+symptom-level graph provides a more granular perspective on the causal
+relationships between precariousness factors and depression symptoms.
+Unlike the sum score graph, which aggregates all symptoms into a single
+measure---potentially obscuring nuanced relationships---the
+symptom-level graph highlights heterogeneity in how precariousness
+factors influence individual depressive symptoms: con (concentration),
+slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal),
+anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.
+
+The symptom-level graph in Figure~\ref{fig-sym} panel (a) reveals a more
+complex and interconnected structure than the sum score graph,
+reflecting the strong interdependence among symptoms and suggesting much
+presence of latent confounders, as indicated by numerous bidirectional
+edges. As before, dashed edges denote ties, with specific ties marked by
+bold horizontal bars. In Figure~\ref{fig-sym} panel (b), the proportion
+matrix provides more insight into these directionalities. Here, opaque
+green dominates most symptom-to-symptom connections, indicating that
+arrowheads are the most frequently inferred edge type. However, blue
+regions, which correspond to circle endpoints, are particularly common
+for anh, slp, ene, and sui, reflecting uncertainty in the causal
+direction among these symptoms. Notably, the only arrowtail connection
+appears between dep and anh, suggesting that anhedonia (anh) is a
+potential cause of depressed mood (dep).
+
+Looking at the symptom-precariousness connections, one of the key
+patterns is that ties are most frequently found in the relationships
+between individual symptoms and precariousness factors. This suggests
+that these causal links may be less stable across different conditions.
+Upon closer examination, most ties arise due to discrepancies between
+the Gaussian CI test and the RCoT test---with Gaussian CI favoring
+arrowheads and RCoT more often predicting the absence of an edge.
+
+This discrepancy likely arises from a combination of factors. First, as
+the analysis shifts from aggregated to disaggregated variables, the
+number of conditional independence (CI) tests increases, leading to a
+loss of statistical power in high-dimensional causal discovery. With
+more variables being conditioned on, the data becomes sparser, making it
+harder to detect weaker dependencies. Additionally,
+over-conditioning---controlling for too many variables---can
+artificially remove statistical associations, leading to false
+negatives. Beyond this general statistical power issue, the fundamental
+differences in how these CI tests operate contribute to the
+inconsistency. RCoT, as a nonparametric test, does not assume linearity
+and can capture both linear and nonlinear dependencies. However, it is
+more conservative and requires larger sample sizes, making it more prone
+to false negatives in high-dimensional settings. In contrast, partial
+correlation (Gaussian CI test) assumes linear Gaussian relationships,
+making it more permissive and sometimes detecting weak dependencies that
+may not be statistically meaningful. As a result, when Gaussian CI
+detects an edge while RCoT does not, it could indicate that the
+relationship is weak, strictly linear, or requires more data for
+reliable nonparametric detection. This explains why many ties in the
+proportion matrix show a 50-50 split between symbols---where half of the
+conditions with RCoT suggest the absence of an edge, while the other
+half using Gaussian CI test favor an arrowhead. This systematic
+divergence highlights the challenges of causal discovery in
+high-dimensional settings, where varying statistical assumptions can
+lead to differing conclusions about the presence and direction of causal
+relationships.
+
+Despite these differences, some consistent patterns emerge across both
+CI tests, particularly in the case of financial stress (S.fin), which
+appears to be connected to nearly all other variables---though many of
+these connections are marked by ties, reflecting uncertainty in
+directionality. One exception is the stronger evidence suggesting that
+S.fin may cause changes in app (appetite), while its relationships with
+other symptoms, such as ene, mot, and dep, remain ambiguous. Similarly,
+social precarity (P.soc) exhibits numerous connections with depressive
+symptoms, with most edges pointing toward P.soc rather than outward from
+it. This suggests that depressive symptoms, particularly dep and glt,
+may contribute to worsening social precarity rather than the other way
+around. Additionally, anh also shows some probability of causally
+influencing P.soc, reinforcing the idea that social precarity is more
+often a consequence rather than a driver of depressive symptoms.
+
+Other precariousness factors exhibit more uncertain but still notable
+relationships. Relational stress (S.rel) shows weak but existing
+connections with slp, glt, and app, though directionality remains
+unclear in many cases. Employment precarity (P.emp) also connects to
+symptoms like slp, mot, and con, but, like S.rel, these relationships
+exhibit a 50/50 split in directionality, reflecting uncertainty in the
+inferred causal paths. In line with the aggregated graph in
+Figure~\ref{fig-sum}, housing precarity (P.hou) does not appear to have
+any direct relationship with depression symptoms but consistently shows
+associations with employment precarity (P.emp) and financial stress
+(S.fin). The causal relationships among precariousness factors remain
+largely unchanged from the aggregated analysis, with financial stress
+(S.fin) exhibiting the strongest tendency to influence other
+precariousness factors.
+
+Among depressive symptoms, dep, glt, and slp appear to be the most
+connected to precariousness factors, while P.soc has the most
+connections with symptoms, predominantly as a recipient rather than a
+driver of influence. Within the symptom network, slp, sui, and anh
+emerge as causally influential symptoms, as they exhibit more outgoing
+arrows compared to other symptoms. On the other hand, con, mot, dep, and
+glt, despite having high connectivity, predominantly receive incoming
+arrows, indicating they are more likely effects rather than causes.
+Considering both symptom-precariousness connections and symptom-level
+dynamics, sleep disturbance (slp) emerges as a central symptom, given
+its strong ties to precariousness factors and its influential role
+within the symptom network. This suggests that sleep issues could be an
+initiating symptom, particularly sensitive to relational stress and
+employment precarity. On the other hand, social precarity (P.soc) serves
+as a key bridge between the depression and precariousness subsystems, as
+depressive symptoms appear to feed back into social precarity,
+reinforcing a self-sustaining dynamic between depression and precarious
+conditions.
+
+Finally, the results from CCI are largely consistent with those from
+FCI, but with some key differences. CCI tends to favor arrowheads more
+frequently, resulting in a greater number of bidirectional edges, which
+suggests a higher involvement of latent confounders. Additionally, CCI
+exhibits more tie situations, though, similar to FCI, most ties occur
+between absence and arrowhead edges, reflecting the discrepancy between
+the Gaussian CI test and RCoT---where the Gaussian test favors
+arrowheads, while RCoT more often suggests the absence of an edge. For a
+detailed visualization of the CCI-derived graph and proportion matrix,
+see Figure~\ref{fig-sym-cci}.
 
@@
 Lastly, we examine the relationships between individual symptoms and
-overall precarity, represented as the sum of five precariousness
-factors. This approach completes the analysis by providing a perspective
-on how individual symptoms relate to overall precarity. It complements
-the previous analysis, which focused on individual symptoms and
-individual precarity factors, by potentially capturing distributed
-relationships across different precarity factors through their
-aggregation into a single overall score. As before, black solid lines
-indicate edges agreed upon by both FCI and CCI, while gray dashed lines
-represent edges where the two algorithms disagree. Circles at the
-endpoints of gray dashed edges denote unresolved directionality.
-Additionally, navy dashed edges highlight relationships that appear
-exclusively in graphs generated using Gaussian CI tests.
-
-Compared to Figure~\ref{fig-sym}, this analysis reveals greater
-disagreement between FCI and CCI regarding edge directionality, as
-evidenced by more dashed edges. FCI tends to estimate more undirected
-edges (circles), while CCI more frequently assigns directed arrows
-between symptom-symptom relationships. For a detailed breakdown, see
-\textbf{?@sec-propmatrix}, which shows the proportion of symbols
-produced by each algorithm. When examining overall precarity, we observe
-that more symptoms are now connected to the precarity score
-(\emph{Precarity}) compared to the individual precarity factor analysis
-in Figure~\ref{fig-sym}. In addition to \emph{glt}, \emph{slp},
-\emph{anh}, and \emph{sui}, which are previously shown to be connected
-to individual precarity factors, \emph{mot}, \emph{app}, and \emph{dep}
-are now identified as having causal relationships with overall
-precarity. Furthermore, when using the Gaussian CI test, con is also
-shown to be causally related to overall precarity. Notably, all these
-connections are directed towards precarity (o→), indicating that overall
-precarity is not the cause of symptoms. The exception is the
-bidirectional edge between \emph{sui} \textless-\textgreater{}
-\emph{Precarity}, which suggests a connection via latent confounders.
-FCI predominantly estimates directed arrows (→) for symptom-precarity
-relationships, suggesting that symptoms are driving precarity. With
-fewer variables (reduced from 14 to 9), this approach appears to have
-slightly more statistical power, enabling algorithms to estimate
-relationships with greater confidence compared to the previous analysis
-involving individual symptoms and individual precarity factors.
+overall precarity, an aggregated measure represented by the sum of five
+precariousness factors. This analysis provides a complementary
+perspective, capturing broader patterns that may not be evident in the
+disaggregated symptom-precariousness analysis. By aggregating
+precariousness factors into a single score, this approach may capture
+distributed relationships across different precarity factors that were
+previously overlooked in the disaggregated analysis.
+
+Unlike the disaggregated graph in Figure~\ref{fig-sym}, the aggregated
+symptom-precariousness graph (Figure~\ref{fig-presum} panel (a)) shows
+more certainty in causal directions, with only a few edges resulting in
+ties, all of which involve the overall precarity factor. This is even
+more evident in the proportion matrix (Figure~\ref{fig-presum} panel
+(b)), where symptom-to-symptom interactions almost fully converge to a
+single edge type. Additionally, all symptom interactions are connected
+either through bidirectional edges or a combination of an arrowhead and
+a circle, suggesting a strong presence of latent confounders.
+
+Examining the symptom-precarity connections, we observe a stronger
+overall trend of depressive symptoms influencing precarity, rather than
+the other way around. Specifically, anh, dep, and slp show a high
+probability of causing precarity, whereas app, glt, and mot exhibit more
+uncertainty in directionality, often resulting in circle endpoints. This
+pattern suggests that when precariousness factors are aggregated, the
+dominant causal flow is from depressive symptoms to precarity, rather
+than precarity driving depression. Particularly, dep emerges as the
+strongest predictor of precarity, exhibiting the highest proportion of
+arrowtails, consistent with findings from the disaggregated analysis.
+
+Comparing this with the CCI-derived graph (Figure~\ref{fig-presum-cci}),
+several key differences stand out. The CCI results reveal that most
+symptoms are interconnected through bidirectional edges, suggesting a
+strong presence of latent confounders. However, dep appears to be a
+distinct exception, as it is predominantly caused by nearly all other
+symptoms, with one exception sui, which does not contribute to dep.
+Similarly, con is influenced by multiple symptoms, albeit with weaker
+support compared to dep. Most interestingly, the CCI results highlight a
+distinct feedback loop between dep and overall precarity, represented by
+a tail-tail edge (\textemdash ). This suggests a possible reinforcing
+cycle between depression and overall precariousness, where depression
+not only arises from precarious conditions but also contributes to their
+persistence, creating a self-sustaining dynamic.
+
+Overall, the aggregated precariousness graph supports the key findings
+from both the disaggregated analysis and the depression sum score
+analysis, providing greater certainty in several important patterns. The
+results highlight the significant role of latent confounders in
+symptom-to-symptom interactions. While certain stressor-related
+precariousness factors, particularly S.fin and S.rel, appear to causally
+influence depressive symptoms, other precariousness factors are more
+likely driven by depression rather than acting as primary causes,
+suggesting that precariousness may function more as a consequence of
+depression than its source. Some symptoms, particularly glt, slp, and
+dep, show greater sensitivity to precarious conditions, while also
+playing a central role in symptom dynamics. Among them, slp and glt
+emerge as influential symptoms within the depression network, whereas
+con and dep, despite their high connectivity, behave more as outcomes
+rather than primary drivers. Together, these findings suggest that
+depression---especially symptoms like sleep disturbances and guilt---may
+contribute to sustaining precarious conditions, possibly forming a
+reciprocal cycle between depression and precariousness.
 
@@
 
-The present study examined the causal relationships between precarity
-factors and depression using causal discovery algorithms, incorporating
+This study examined the causal relationships between precariousness
+factors and depression using causal discovery algorithms, integrating
 both sum score and symptom-level analyses. By analyzing PHQ sum scores
 alongside individual symptom scores, and comparing aggregated precarity
-scores with individual precarity factors, we gained a comprehensive
-perspective on how different aspects of precarity influence depression.
-This dual approach highlighted nuanced pathways that may be overlooked
-when focusing solely on aggregate measures, while also capturing broader
-interactions that might remain undetected in analyses limited to
-individual factors.
-
-Our findings highlight the significant role of recent relational stress
-(\emph{S.rel}) as a potential causal factor for depression, consistently
+scores with individual precariousness factors, we gained a comprehensive
+perspective on how different aspects of precariousness influence
+depression. This dual approach revealed nuanced pathways that may be
+overlooked when relying solely on aggregate measures, while also
+capturing broader interactions that might remain undetected in analyses
+focused on individual factors.
+
+Our findings highlight the significant role of recent financial stress
+(S.fin) as a potential causal factor for depression, consistently
 observed across both sum score and symptom-level analyses. This
-consistency underscores the profound impact interpersonal relationships
-can have on mental health. The symptom-level analysis further identified
-specific symptoms, such as sleep disturbances (\emph{slp}), guilt
-(\emph{glt}), and anhedonia (\emph{anh}), as particularly sensitive to
-external precarity conditions. These symptoms emerged as potential
-initiators or activators within the depressive symptom network,
-suggesting that they may serve as early warning signs or valuable
-targets for preventive interventions.
-
-Employment precarity (\emph{P.emp}) and social precarity (\emph{P.soc})
-were more likely to be influenced by depressive symptoms rather than
-acting as direct causes. The symptom-level graph provided additional
-clarity by identifying that sleep disturbances influenced employment
-precarity (\emph{slp} o-\textgreater{} \emph{P.emp}) and feelings of
-guilt affected social precarity (\emph{glt} o-\textgreater{}
-\emph{P.soc}). This directional insight suggests that interventions
-targeting specific depressive symptoms may have downstream effects on
-improving certain aspects of precarity.
-
-Interestingly, financial stress (\emph{S.fin}) did not exhibit causal
-relationships with individual depressive symptoms in the symptom-level
-analysis, contrasting with its apparent role in the sum score graphs.
-This divergence indicates that while financial stress may influence
-overall depression severity, its impact on specific symptoms may not be
-significant. Also, housing precarity (\emph{P.hou}) emerged as an
-isolated node in the symptom-level graph, losing all connections with
-other precarity factors. This isolation suggests that housing precarity
-may operate independently of the depressive symptom network or that its
-effects are not captured within the scope of the measured variables.
+consistency underscores the profound impact of financial stressors on
+mental health. The symptom-level analysis further identified specific
+symptoms---particularly sleep disturbances (slp), depressed mood (dep),
+guilt (glt), and anhedonia (anh)---as being especially sensitive to
+external precarious conditions.
+
+Among these symptoms, dep stands out as a distinct causal factor for
+precarity in both aggregated and disaggregated analyses, with CCI even
+suggesting a possible cyclic relationship between them. This implies
+that depression and precariousness may reinforce one another,
+potentially leading to a worsening cycle over time. Within the symptom
+network, dep also exhibits a high degree of connectivity, but with
+mostly incoming arrows, suggesting that as symptom dynamics intensify,
+dep is quickly impacted by multiple factors---particularly P.soc (social
+precarity), further fueling the depression-precariousness cycle.
+
+Beyond dep, slp emerges as a critical symptom in this dynamic. It
+connects with both social and employment precariousness, while also
+playing a central role in symptom interactions, affecting other symptoms
+more than most others. This suggests that slp may act as an initiator
+within the depressive symptom system, potentially triggering or
+amplifying symptom cascades.
+
+Among the precariousness factors, P.soc appears to be one of the most
+pivotal---not as a driver of symptoms, but rather as a consequence of
+depression. It maintains the strongest connectivity with symptoms, yet
+its directionality indicates that it is largely shaped by depressive
+symptoms rather than acting as a primary cause. Together, these patterns
+suggest a broader directional dynamic in which financial stress (S.fin)
+exacerbates depressive symptoms, which in turn feed into social
+precarity, perpetuating the cycle.
+
+This directional insight suggests that interventions targeting specific
+depressive symptoms, particularly dep, may help prevent downstream
+effects on worsening precarious conditions. Alternatively, early
+intervention focusing on slp may be a promising strategy, as slp appears
+to act as a potential initiator or activator within the depressive
+symptom network. This positioning suggests that slp could serve as an
+early warning sign, offering a valuable target for preventive
+interventions aimed at breaking the depression-precariousness cycle
+before it intensifies.
 
@@
 next step would involve integrating these causal structures into
-computational models, such as the symptom dynamic model proposed in
-\textbf{our comp-model paper}. By simulating intervention effects, such
-models could provide more realistic insights into how targeted actions
-might influence symptom networks and precarity factors over time. For
-example, interventions focused on improving sleep hygiene or alleviating
-guilt could be evaluated for their cascading effects on employment and
-social relationships, offering actionable guidance for designing
+computational models, such as the symptom dynamic model proposed in Park
+et al. (2025). By simulating intervention effects, such models could
+provide more realistic insights into how targeted actions might
+influence symptom networks and precarity factors over time. For example,
+interventions focused on improving sleep hygiene or alleviating guilt
+could be evaluated for their cascading effects on employment and social
+relationships, offering actionable guidance for designing
 population-level mental health strategies. As one of the early
@@
 
+\bibitem[\citeproctext]{ref-park2025individual}
+Park, K., Waldorp, L., \& Vasconcelos, V. V. (2025). \emph{The
+individual-and population-level mechanistic implications of statistical
+networks of symptoms}.
+
+\bibitem[\citeproctext]{ref-rahimi2007random}
+Rahimi, A., \& Recht, B. (2007). Random features for large-scale kernel
+machines. \emph{Advances in Neural Information Processing Systems},
+\emph{20}.
+
 \bibitem[\citeproctext]{ref-rohbeck2024}
@@
 209--242). PMLR.
-\url{https://proceedings.mlr.press/v236/rohbeck24a.html}
 
@@
 
-\subsection{Results from CCI
-algorithm}\label{results-from-cci-algorithm}
+\subsection{Results from CCI algorithm}\label{sec-cci}
 

### 2025-05-16 draft_v1/draft_v1.qmd
draft_update

@@
 ---
-title: "Causal Discovery on Precarity and Depression"
+title: "From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study"
 #subtitle: Using the HELIUS dataset
@@
 abstract: |
-  \noindent Understanding the causal mechanisms linking precariousness and depression is critical for developing effective interventions.
+  \noindent This study investigates the dynamic relationship between precarious life conditions and depression by combining constraint-based causal discovery with computational modeling.
+Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between financial stress, domain-specific precariousness indicators, and depressive symptoms.
+Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
+  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precarity under varying levels of financial stress reduction.
+Simulation results revealed that the system's responsiveness depends not only on the intensity of external intervention but also on internal structural features such as feedback strength and noise.
+In particular, some configurations exhibited bistability, requiring interventions to surpass a critical threshold to trigger lasting improvement, while others responded more proportionally.
+  Together, these findings emphasize that internal feedback dynamics—not just external adversity—can shape the impact of mental health interventions.
+More broadly, our study offers a scalable framework for integrating causal discovery and dynamical simulation to better understand and inform policy responses to complex psychosocial phenomena.
+
+  <!-- Understanding the causal mechanisms linking precariousness and depression is critical for developing effective interventions.
 This study utilizes data from the HELIUS cohort study to explore these relationships using advanced causal discovery methods.
@@
 This study serves as a foundational effort, offering both methodological advancements and practical implications for addressing depression at a population level.
+-->
 ---
@@
 
-Mental health problems in urban areas are on the rise, yet the complexity of mental health systems poses significant challenges in understanding the underlying mechanisms driving these issues — let alone designing effective interventions.
-Research has long sought to identify underlying factors contributing to mental health problems.
-Recent research looked into the association between mental health and indicators of precariousness in various dimensions of life, indicating a high level of uncertainty and instability in people’s lives, such as factors related to employment, social relations, finances, housing, and culture (ref Leonie’s paper).
-This comprehensive perspective on precariousness helped to highlight how different aspects of life may be interconnected.
-While this research has advanced our understanding of which indicators of precariousness are related to mental health, a key question remains unanswered: how do these factors influence mental health?
-Specifically, the lack of directional insight — knowing what influences what — limits our ability to identify and prioritize effective intervention targets.
-
-This study aims to investigate the causal relationships between different dimensions of precariousness and mental health outcomes, to with a specific focus on depression, the most prevalent mental health issue.
-Using causal discovery methods, we explore how various aspects of precariousness contribute to depression and whether depression, in turn, exacerbates precariousness, potentially creating a feedback loop.
-Furthermore, we zoom in at the symptom level, examining individual depressive symptoms to identify those that may act as initiators — triggering the activation of the symptom network — by displaying heightened sensitivity to specific precariousness factors.
-We also investigate symptoms that may function as bridging elements, linking the dynamics of depression with precariousness by feeding back into precarious conditions, potentially reinforcing the cycle of mental health deterioration and precarious conditions.
-Through this analysis, we aim to uncover the causal mechanisms driving mental health challenges related to precariousness and establish a foundation for more targeted and effective interventions by identifying the key driving forces within this system.
+Mental health disorders represent a growing global health concern, especially in urbanized regions [@WHO_2022; @gruebner2017cities].
+In 2019, one in eight people worldwide were living with a mental health condition, with the highest disability-adjusted life years (DALYs) due to mental and addictive disorders concentrated in high-income countries such as those in Northern Europe, North America, and Australia [@rehm2019global; @WHO_2022].
+Despite growing awareness, governments have struggled to design effective responses.
+Mental health outcomes arise from complex, multi-level dynamics—intertwined with the social, economic, and spatial structures of urban environments—making both diagnosis and intervention design deeply challenging [@van2021advancing].
+
+Recent work has shifted focus from individual-level vulnerability to broader social and structural determinants.
+While high-income nations show elevated overall DALYs, within-country disparities reveal that income inequality is a significant predictor of mental health burden [@rehm2019global].
+Moreover, precarious working conditions, housing instability, and neighborhood disadvantage have all been linked to increased risk of depression and anxiety [@pevalin2017impact; @fone2014effect; @rugulies2023work; @ronnblad2019precarious].
+This growing literature highlights that poor mental health is not merely a personal issue, but one rooted in sustained socioeconomic stress and structural uncertainty.
+
+Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains—including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering].
+This broader view reveals how different forms of insecurity can co-occur and compound, forming an ecosystem of risk.
+However, despite growing evidence of association between precariousness and mental health, a fundamental question remains unresolved: *How do these factors causally influence one another?* Most existing studies rely on cross-sectional correlations, leaving open questions about directionality and feedback.
+Without stronger causal insight, it remains difficult to identify priority targets for intervention or to anticipate unintended effects.
+
+This study aims to investigate the causal mechanisms linking precariousness and depression—and how those mechanisms shape system responses to external intervention.
+Specifically, we ask: *How does the internal configuration of the precarity–depression relationship affect the outcomes of stress-reducing interventions?*
+To answer this, we combine two approaches.
+First, we use cycle-capable causal discovery algorithms to infer directional relationships among financial stress, domain-specific precariousness factors, and depressive symptoms.
+By analyzing both aggregate scores (e.g., PHQ-9 total, composite precarity index) and disaggregated variables (individual symptoms and life-domain indicators), we identify both broad associations and fine-grained causal pathways.
+Second, we translate these insights into a computational dynamical model that simulates how depression and precarity co-evolve under varying levels of external stress.
+This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention.
+In some configurations, the system exhibits bistability, remaining stuck in a high-risk state unless external support exceeds a critical threshold.
+In others, it shows unistable behavior, responding more gradually to incremental changes.
+By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precarity system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.
+
+<!-- Mental health problems are rising globally, and populations facing chronic stress and socioeconomic instability bear a disproportionate share of this burden.
+Despite growing recognition of these challenges, designing effective policies remains difficult—largely because mental health outcomes emerge from complex, interacting factors that do not operate in isolation.
+To inform better interventions, we must move beyond identifying which factors are associated with mental health problems and instead uncover how those factors influence one another, the underlying causal mechanisms.
+-->
+
+<!-- Precarious life conditions—such as unstable housing, insecure employment, and financial hardship—are now well established as key correlates of depression and other mental health outcomes.
+However, the causal dynamics linking these domains remain poorly understood.
+In particular, it is unclear how different forms of precariousness and depressive symptoms influence one another, and whether their interplay creates self-reinforcing loops that stabilize poor mental health and social adversity over time.
+-->
+
+<!-- This study aims to investigate the causal mechanisms linking precariousness and depression, and to examine how those mechanisms shape the system’s response to external intervention.
+Specifically, we ask: How do different configurations of the precarity–depression relationship affect the outcomes of interventions aimed at reducing external stress?
+-->
+
+<!-- To address this, we combine causal discovery and mechanistic modeling.
+We apply cycle-capable causal discovery algorithms to infer directional relationships among financial stress, precariousness indicators, and depressive symptoms.
+Our analysis includes both aggregate scores (e.g., PHQ-9 total, precarity composite index) and disaggregated variables (individual symptoms and domain-specific precariousness), allowing us to capture both broad associations and fine-grained pathways.
+-->
+
+<!-- Building on the discovered causal patterns, we simulate a set of scenarios using a computational model that captures the co-evolution of depression and precarity under varying levels of financial stress.
+This framework allows us to examine how internal system properties—particularly feedback strength and stochastic noise—influence responsiveness to intervention.
+In some regimes, the system exhibits bistability, resisting change unless pushed past a critical threshold.
+In others, it transitions gradually, reflecting unistable dynamics.
+Through this integrative approach, we aim to show how the internal architecture of the depression–precarity system fundamentally shapes its susceptibility to change—highlighting the importance of structural understanding in designing effective, equitable mental health interventions.
+-->
 
@@
 
-We use data from the HELIUS study, a multi-ethnic cohort that includes participants of Dutch, Turkish, Moroccan, Surinamese and Ghanaian origin [@snijder2017cohort].
-To operationalize precariousness factors, we draw on the framework outlined in previous research (i.e., Leonie's paper) and select a set of relevant variables.
-
-To ensure a robust representation of each precariousness factor, we conducted various exploratory analyses to identify consistent and meaningful factor structures.
+We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study, which has been described in detail elsewhere [@stronks2013unravelling; @snijder2017cohort].
+It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and which provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement].
+To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
+To ensure a robust representation of precariousness in our causal discovery models, we conducted various exploratory analyses to identify consistent and meaningful data-driven factor structures.
 Based on these analyses, we identified five precariousness factors, including two related to recent stressors, each comprising multiple variables as outlined below.
-Detailed information on the exploratory analyses can be found in the [Appendix](#sec-appendix).
+DFurther details on the HELIUS study and factor extraction are provided in the [Appendix](#sec-appendix).
 
@@
 
-## Causal Discovery
-
-There are numerous causal discovery algorithms available; however, in this study, we focus on algorithms suited to the potential cyclic relationships within our system.
-Specifically, we use *FCI* (Fast Causal Inference) and *CCI* (Cyclic Causal Inference), both capable of accounting for such cycles [@mooij2020constraint; @strobl2019].
+## Causal Discovery{#sec-analysis}
+
+To uncover directional relationships between precariousness and depression, we used constraint-based causal discovery algorithms that allow for feedback loops and latent confounding.
+Specifically, we applied the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019].
+For reference, we also included the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].
+
+A key challenge in applying these algorithms to the HELIUS dataset is that many variables are non-Gaussian and their relationships are likely nonlinear.
+To address this, we supplemented the commonly used Gaussian conditional independence (CI) test—based on partial correlations—with a non-parametric alternative: the *Randomized Conditional Correlation Test* (RCoT).
+RCoT approximates kernel-based CI testing using random Fourier features, significantly reducing computational demands while maintaining sensitivity to nonlinear dependencies [@strobl2019approximate; @zhang2012kernel].
+
+We examined the causal structure using three complementary approaches:
+
+1. Aggregate-level analysis: relationships between five domain-specific precariousness factors and the PHQ-9 sum score.
+2. Fully disaggregated analysis: relationships between individual precariousness items and individual depressive symptoms.
+3. Mixed analysis: relationships between individual symptoms and a composite index of overall precariousness.
+
+The aggregate analysis reduces dimensionality and yields interpretable summaries of how domains of precarity relate to overall depression severity.
+However, it may obscure finer-grained effects.
+The disaggregated analysis allows for precise mapping of which symptoms are influenced by (or influence) specific precariousness conditions, but it introduces complexity due to the high dimensionality and distributional properties of the data.
+The mixed approach, in turn, integrates these perspectives by examining how individual symptoms respond to cumulative precarity across domains.
+
+To ensure robust estimation and guard against sensitivity to parameter settings, we implemented a systematic grid of analysis conditions:
+
+* Significance levels: $\alpha$ = 0.01 and 0.05
+* Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8
+* CI tests: Gaussian and RCoT
+* Algorithms: FCI, CCI, and PC
+
+Each combination was bootstrapped 100 times for sum-score level analyses and 30 times for symptom-level analyses, yielding a total of 1,600 graphs per algorithm for the full analysis.
+For symptom networks, we fixed the skeleton structure using a consensus graph estimated across PC, FCI, and CCI ($\alpha$ = 0.01, RCoT).
+This constraint improves computational efficiency without limiting the estimation of directional edges between symptoms and precariousness variables.
+
+To summarize the results, we adopted a two-level aggregation strategy.
+First, within each analysis condition, we recorded the most frequently observed edge-endpoint type (e.g., directed, undirected, uncertain) from the bootstraps.
+Then, across all conditions, we tallied the most frequent endpoint type per edge.
+In cases of ties, we assigned dashed edges to reflect uncertainty.
+For transparency, we accompany each graph with a table showing the relative frequencies of each edge symbol.
+
+This procedure yields a final graph for each algorithm and analysis type—offering both robustness and interpretability. See @fig-workflow for an overview of the causal discovery pipeline.
+
+For detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, see the [Appendix](#sec-appendix).
+
+<!-- To uncover directional relationships between precariousness and depression, we used constraint-based causal discovery algorithms capable of detecting feedback loops and latent confounding.
+Specifically, we applied the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which support cyclic and non-acyclic structures [@mooij2020constraint; @strobl2019].
 Additionally, we include the *PC* algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation].
+-->
+
+<!-- One practical challenge in applying these algorithms to the HELIUS dataset is that most variables do not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear.
+To address this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel].
+-->
+<!-- While kernel-based conditional independence tests (KCIT) are theoretically effective in capturing non-linear dependencies and subtle interactions in data with non-Gaussian noise, they are computationally demanding for large datasets like HELIUS.
+Their quadratic scaling with sample size is primarily due to the need to invert large kernel matrices, making them impractical for large-scale analysis [@rahimi2007random].
+-->
+<!-- To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximates kernel methods using random Fourier features.
+This approach reduces computational complexity from quadratic to linear scaling with sample size, significantly lowering computational costs while maintaining robustness [@strobl2019approximate].
+For a more detailed explanation of RCoT, refer to @sec-rcot.
+-->
+
+<!-- We analyze the causal structure using three complementary approaches: (1) Examining the relationship between the five individual precariousness factors and the PHQ sum score, which represents overall depression severity.
+-->
+<!-- (2) Investigating the relationship between individual depressive symptoms and individual precariousness factors.
+(3) Assessing the relationship between individual depressive symptoms and overall precariousness (the sum of all precariousness factors).
+-->
+<!-- Using the PHQ sum score as a measure of depression simplifies the analysis by reducing dimensionality, improving computational efficiency, and providing a broad, interpretable perspective on how precariousness factors relate to overall depression severity.
+However, this aggregated approach may overlook granular symptom-level relationships.
+-->
+
+<!-- By zooming in on individual symptoms, we capture a more detailed picture of how specific depressive symptoms interact with different precariousness factors.
+This approach allows us to map individual causal pathways, identifying which symptoms are particularly sensitive to certain precarious conditions and vice versa.
+However, this level of detail introduces methodological challenges, including non-standard distributions of symptom variables and the high-dimensional nature of the analysis, which can reduce statistical power and significantly increase computational demands—especially when using non-parametric conditional independence (CI) tests.
+-->
+
+<!-- Finally, analyzing individual symptoms in relation to overall precariousness provides an integrative perspective, revealing how depression symptoms collectively relate to precarious conditions.
+This approach complements the fully disaggregated analysis by aggregating precariousness factors, potentially uncovering relationships that may be too weak to detect when examined separately.
+This is particularly relevant when a symptom is associated with multiple precariousness factors, but these associations are not strong enough to emerge in an individual-level analysis.
+-->
+<!-- By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression.
+-->
+
+<!-- Furthermore, to address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping.
+For each condition, we generate 100 bootstrap samples, estimate causal graphs, and retain only the edge-endpoints that exceed predefined thresholds.
+If no edge-endpoint type surpasses the threshold, we assign a circle ($\circ$) to indicate uncertainty in the edge direction.
+-->
+
+<!-- The analyses are conducted under the following conditions: -->
+
+<!-- -   Significance levels ($\alpha$): 0.01 and 0.05 -->
+<!-- -   Thresholds: 0.5, 0.6, 0.7, and 0.8 -->
+<!-- -   CI test: Gaussian CI test, RCoT -->
+<!-- -   Algorithms: FCI, CCI, and PC -->
+
+<!-- This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms.
+Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm.
+-->
+<!-- For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30, easing computational demands.
+-->
+<!-- To further enhance efficiency, we fix the skeleton of the symptom network using a common structure derived from PC, FCI, and CCI algorithms ($\alpha$ = 0.01, RCoT CI test).
+This step constrains which edges can be present among symptoms but does not restrict the estimation of causal directions within the symptom network.
+Additionally, the entire causal structure between symptoms and precariousness factors remains fully estimable, ensuring that key relationships are still inferred.
+-->
+<!-- Beyond its computational benefits — particularly in reducing the time required for skeleton estimation — this fixed structure also aligns well with commonly reported symptom network structures in the literature [@park2025individual], ensuring consistency with existing findings.
+-->
+
+<!-- To obtain a final, stable summary of causal relationships, we identify the most frequently occurring edge-endpoints across different experimental setups.
+First, within each condition, we determine the most frequent edge-endpoint from all bootstrap samples.
+Then, across different conditions, we aggregate these selected symbols and determine the overall most common edge-endpoint for each graph entry.
+-->
+
+<!-- Since we have an even number of experimental setups, ties may occasionally occur.
+In such cases, we explicitly denote them in the resulting graph using dashed lines.
+To further help with the interpretation and understanding of the resulting graph, we provide a corresponding table for each graph, showing the proportion of each edge-endpoint type to clarify how dominant a given symbol is in each case.
+-->
+
+<!-- By double summarizing the results in this way, we generate one final graph per algorithm for each analysis approach: the partially aggregated analysis, which uses either the sum of precariousness factors or the sum of depression factors to capture overall trends, and the fully disaggregated analysis, which examines individual precariousness and depression factors separately.
+-->
+<!-- This approach ensures that only stable and consistent edges — those that persist across different parameter settings — are retained, leading to a robust and reliable understanding of the relationships between precariousness factors and depression.
+See @fig-workflow for an overview of the analysis workflow.
+-->
+
+```{r}
+#| label: fig-workflow
+#| fig-cap: "Analysis workflow applied across all three algorithms."
+#| fig-align: center
+#| out-width: 80%
+
+knitr::include_graphics("img/simsetup2.png")
+
+```
+
+## Modeling Dynamics between Precarity and Depression
+
+To explore how internal system structure shapes responses to external support, we implement a high-level computational model that formalizes key mechanisms suggested by the causal discovery analysis.
+Rather than attempting to replicate the full estimated causal graph, this model abstracts the essential structure into a compact, tractable system focused on dynamic interactions between depression and social precariousness, both influenced by financial stress.
+
+As illustrated in @fig-concept, the model includes three components: depression ($D$), represented by the PHQ-9 sum score, social precariousness ($P$), and external financial stress ($S$).
+Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S$ acting as a shared input.
+These dynamics are formalized as a pair of nonlinear stochastic differential equations (SDEs):
+
+$$
+\begin{aligned}
+dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\
+dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2 
+\end{aligned}
+$$
+
+Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress.
+The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which variable adjusts toward its input-driven target state.
+The coefficients $\alpha_1, \alpha_2$ and $\beta_1, \beta_2$ represent the strengths of the causal influences between variables.
+Stochastic fluctuations are introduced through the terms $\sigma_1 dW_1$ and $\sigma_2 dW_2$, where $dW_1$ and $dW_2$ are independent Wiener processes.
+The system uses the hyperbolic tangent function ($tanh$) to bound the influence of inputs, ensuring smooth, saturating dynamics and preventing runaway feedback loops.
+Importantly, the nonlinearity introduced by the $tanh$ function also allows the system to exhibit bistability, supporting the existence of two stable states: one corresponding to low depression and precariousness, and the other to high depression and precariousness.
+
+<!-- Building on the causal structure derived from the result of causal discovery, we implement a high-level simulation model to explore the potential impact of targeted interventions on precariousness.
+This serves as a proof-of-concept for translating causal relationships into dynamic systems that can generate actionable insights.
+-->
+
+<!-- The structure of the simulation model, illustrated in @fig-concept, captures the dynamic interplay between depression ($PHQsum$) and social precariousness ($P.soc$), both of which are influenced by an external financial stressor ($S.fin$).
+Importantly, the model incorporates cyclic dependencies between depression and precariousness, reflecting their reciprocal influence over time.
+-->
+<!-- This system is formalized as a set of non-linear stochastic differential equations (SDEs): -->
+
+<!-- $$ -->
+<!-- \begin{aligned} -->
+<!-- dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\ -->
+<!-- dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2  -->
+<!-- \end{aligned} -->
+<!-- $$ -->
+
+<!-- Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress.
+The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which variable adjusts toward its input-driven target state.
+The coefficients $\alpha_1, \alpha_2$ and $\beta_1, \beta_2$ represent the strengths of the causal influences between variables.
+-->
+<!-- Stochastic fluctuations are introduced through the terms $\sigma_1 dW_1$ and $\sigma_2 dW_2$, where $dW_1$ and $dW_2$ are independent Wiener processes.
+The system uses the hyperbolic tangent function ($tanh$) to bound the influence of inputs, ensuring smooth, saturating dynamics and preventing runaway feedback loops.
+Importantly, the nonlinearity introduced by the $tanh$ function also allows the system to exhibit bistability, supporting the existence of two stable states: one corresponding to low depression and precariousness, and the other to high depression and precariousness.
+-->
+
+```{r}
+#| label: fig-concept
+#| fig-cap: "Conceptual high-level causal structure of precariousness and depression system."
+#| fig-align: center
+#| out-width: 35%
+
+knitr::include_graphics("img/conceptmodel.png")
+```
+
+### Model Calibration
+
+To identify realistic parameter settings for the simulation model, we calibrated six free parameters—$\alpha_1$, $\alpha_2$, $\beta_1$, $\beta_2$, and noise amplitudes $\sigma_1$, $\sigma_2$—using the Non-dominated Sorting Genetic Algorithm II (NSGA-II) [@deb2002fast], a multi-objective evolutionary optimization method well-suited for noisy and high-dimensional search spaces.
+The goal was to match simulated system behavior to empirical summary statistics derived from the HELIUS dataset, including the means and variances of depression ($D$) and social precariousness ($P$), as well as their partial correlations after adjusting for financial stress ($S$), specifically $\rho_{DP.S}$, $\rho_{DS.P}$, and $\rho_{PS.D}$.
+Given the stochastic nature of the model, each candidate parameter set was evaluated by simulating 1,000 independent trajectories, each representing an individual with a fixed $S$ value sampled from the observed distribution.
+Final-state values were extracted and aggregated to compute population-level summary statistics.
+Fitness was assessed as the absolute deviation from the empirical targets, aggregated across all objectives.
+NSGA-II was implemented via the nsga2R package [@nsga2R], with a population of 300 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2).
+This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives.
+From the resulting Pareto front (@fig-pareto), we selected the top 30 parameter sets with the lowest combined error to use in the subsequent simulations assessing intervention effects.
+
+```{r}
+#| label: fig-pareto
+#| fig-cap: >
+#|   Pareto front of simulated parameter sets.
+#|   *Each point represents a solution evaluated by its total variance error (x-axis) and total partial correlation error (y-axis), with color indicating total mean error.
+#|   Red crosses highlight the top 30 solutions used for simulation.*
+#| fig-align: center
+#| out-width: 80%
+knitr::include_graphics("img/paretoplot.png")
+```
+
+### Intervention Simulation
+
+To evaluate the effects of hypothetical support, we simulate scenarios in which each individual’s financial stress level ($S$) is proportionally reduced toward the minimum observed value.
+This models a population-wide intervention applied at varying strengths --- from no reduction in stress (0%) to complete alignment with the minimum observed stress level (100%).
+
+For each of the top 30 calibrated parameter sets, we simulate the system’s evolution across 300 individuals under a continuum of intervention levels.
+Each simulation run generates final values for $D$ and $P$, which are then used to assess how internal model parameters influence the system’s behavior under changing external conditions.
+
+# Results
+
+## Causal Structure Linking Precarity and Depression
+
+### Depression as sum score
+The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
+@fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
+As described in [@sec-analysis], dashed edges indicate ties, with the exact edge endpoint where the tie occurs is represented by a bold horizontal bar.
+
+The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (PHQsum).
+Employment precarity (P.emp) and social precarity (P.soc) are not identified as causes of depression, whereas financial stress (S.fin) appears to play a causal role.
+While P.emp and P.soc are generally not recognized as causes of other precariousness factors, S.fin emerges as a potential cause, as indicated by its circle edge endpoint.
+Supporting this, the proportion matrix plot in panel (b) shows that S.fin has some probability of causally influencing either P.emp or P.soc.
+
+The matrix plot can be interpreted such that the symbol in matrix[$i$, $j$] represents the relationship $i -~[symbol] \  j$.
+For example, if matrix[$i$, $j$] is > and matrix[j, i] is $\circ$, then the inferred relationship is $i \circarrow j$.
+In the table, different edge types are represented by distinct colors: light gray for the absence of an edge, blue for circles, green for arrowheads, and coral for arrowtails.
+The colors are blended based on the proportion of each edge type, with higher proportions increasing the opacity of the corresponding color, making dominant symbols more visually prominent.
+
+Relational stress (S.rel) plays a more nuanced role, interacting with depression and S.fin  through a latent confounder.
+Its relationship with P.soc remains less certain, as the proportion table suggests a fair probability that this connection is absent, uncertain, or bidirectional.
+Meanwhile, housing precarity (P.hou) is not directly causally related to depression but is linked to P.emp and S.fin.
+While P.emp and S.fin are identified as non-causes of P.hou, it remains unclear whether P.hou causally influences P.emp or if their relationship is mediated by an unobserved confounder.
+
+The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify S.fin as a cause of depression.
+Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI.
+Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.
+
+<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precarity  (P.emp) and social precarity (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.
+Additionally, P.emp and S.fin are identified as non-causes of housing precarity (P.hou).
+Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship.
+Similar ambiguities are found in the connections between S.rel and PHQsum and between S.fin and PHQsum -->
+
+<!-- Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship.
+Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*.
+-->
+
+<!-- Graph (b), derived solely from the nonparametric RCoT method.
+Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*.
+-->
+<!-- The relationship between stressors and depression remains unresolved between the two algorithms.
+However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
+-->
+<!-- The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.
+-->
+
+<!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders.
+Employment and social precarity may also be influenced by depression, either directly or through latent variables.
+Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways.
+-->
+
+```{r}
+#| layout-ncol: 2
+#| label: fig-sum
+#| fig-cap: "Resulting graph of precarity factors and depression sum score using FCI and proportion of edge endpoint types."
+#| fig-subcap: 
+#|   - "FCI PAG"
+#|   - "Proportion matrix"
+#| fig-align: center
+#| out-width: 100%
+
+knitr::include_graphics("img/FCI_depsum.png")
+knitr::include_graphics("img/depsum_mat_fci.pdf")
+
+```
+
+<!-- \begin{figure}[htbp] -->
+<!--     \centering -->
+<!--         \caption{Summary of all sum score graphs} -->
+<!--         \includegraphics[width=1\textwidth]{img/sum_all.png} -->
+<!-- \end{figure} -->
+<!-- \begin{figure}[htbp] -->
+<!--     \centering -->
+<!--         \caption{Summary of sum score graphs given RCoT} -->
+<!--         \includegraphics[width=1\textwidth]{img/sum_rcotonly.png} -->
+<!-- \end{figure} -->
+
+### Individual depression symptom
+
+Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms.
+Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph highlights heterogeneity in how precariousness factors influence individual depressive symptoms: con (concentration), slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal), anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.
+
+The symptom-level graph in @fig-sym panel (a) reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges.
+As before, dashed edges denote ties, with specific ties marked by bold horizontal bars.
+In @fig-sym panel (b), the proportion matrix provides more insight into these directionalities.
+Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type.
+However, blue regions, which correspond to circle endpoints, are particularly common for anh, slp, ene, and sui, reflecting uncertainty in the causal direction among these symptoms.
+Notably, the only arrowtail connection appears between dep and anh, suggesting that anhedonia (anh) is a potential cause of depressed mood (dep).
+
+Looking at the symptom-precariousness connections, one of the key patterns is that ties are most frequently found in the relationships between individual symptoms and precariousness factors.
+This suggests that these causal links may be less stable across different conditions.
+Upon closer examination, most ties arise due to discrepancies between the Gaussian CI test and the RCoT test—with Gaussian CI favoring arrowheads and RCoT more often predicting the absence of an edge.
+
+This discrepancy likely arises from a combination of factors. 
+First, as the analysis shifts from aggregated to disaggregated variables, the number of conditional independence (CI) tests increases, leading to a loss of statistical power in high-dimensional causal discovery.
+With more variables being conditioned on, the data becomes sparser, making it harder to detect weaker dependencies.
+Additionally, over-conditioning—controlling for too many variables—can artificially remove statistical associations, leading to false negatives.
+Beyond this general statistical power issue, the fundamental differences in how these CI tests operate contribute to the inconsistency.
+RCoT, as a nonparametric test, does not assume linearity and can capture both linear and nonlinear dependencies.
+However, it is more conservative and requires larger sample sizes, making it more prone to false negatives in high-dimensional settings.
+In contrast, partial correlation (Gaussian CI test) assumes linear Gaussian relationships, making it more permissive and sometimes detecting weak dependencies that may not be statistically meaningful.
+As a result, when Gaussian CI detects an edge while RCoT does not, it could indicate that the relationship is weak, strictly linear, or requires more data for reliable nonparametric detection.
+This explains why many ties in the proportion matrix show a 50-50 split between symbols—where half of the conditions with RCoT suggest the absence of an edge, while the other half using Gaussian CI test favor an arrowhead.
+This systematic divergence highlights the challenges of causal discovery in high-dimensional settings, where varying statistical assumptions can lead to differing conclusions about the presence and direction of causal relationships.
+
+Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of financial stress (S.fin), which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality.
+One exception is the stronger evidence suggesting that S.fin may cause changes in app (appetite), while its relationships with other symptoms, such as ene, mot, and dep, remain ambiguous.
+Similarly, social precarity (P.soc) exhibits numerous connections with depressive symptoms, with most edges pointing toward P.soc rather than outward from it.
+This suggests that depressive symptoms, particularly dep and glt, may contribute to worsening social precarity rather than the other way around.
+Additionally, anh also shows some probability of causally influencing P.soc, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.
+
+Other precariousness factors exhibit more uncertain but still notable relationships.
+Relational stress (S.rel) shows weak but existing connections with slp, glt, and app, though directionality remains unclear in many cases.
+Employment precarity (P.emp) also connects to symptoms like slp, mot, and con, but, like S.rel, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
+In line with the aggregated graph in @fig-sum, housing precarity (P.hou) does not appear to have any direct relationship with depression symptoms but consistently shows associations with employment precarity (P.emp) and financial stress (S.fin).
+The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with financial stress (S.fin) exhibiting the strongest tendency to influence other precariousness factors.
+
+Among depressive symptoms, dep, glt, and slp appear to be the most connected to precariousness factors, while P.soc has the most connections with symptoms, predominantly as a recipient rather than a driver of influence.
+Within the symptom network, slp, sui, and anh emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms.
+On the other hand, con, mot, dep, and glt, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
+Considering both symptom-precariousness connections and symptom-level dynamics, sleep disturbance (slp) emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network.
+This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precarity.
+On the other hand, social precarity (P.soc) serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.
+
+Finally, the results from CCI are largely consistent with those from FCI, but with some key differences.
+CCI tends to favor arrowheads more frequently, resulting in a greater number of bidirectional edges, which suggests a higher involvement of latent confounders.
+Additionally, CCI exhibits more tie situations, though, similar to FCI, most ties occur between absence and arrowhead edges, reflecting the discrepancy between the Gaussian CI test and RCoT—where the Gaussian test favors arrowheads, while RCoT more often suggests the absence of an edge.
+For a detailed visualization of the CCI-derived graph and proportion matrix, see @fig-sym-cci.
+
+<!-- <!-- Centrality and Connectivity --> 
+<!-- Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points.
+Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network.
+Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precarity factors.
+These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes.
+-->
+<!-- <!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. --> 
+<!-- The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs.
+As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precarity (*P.soc*) are more likely to be influenced by depressive symptoms.
+-->
+
+<!-- The symptom-level analysis, however, provides more specificity by pinpointing the symptoms involved in these relationships.
+For instance, *slp* is identified as influencing *P.emp* or potentially through a latent confounder (*slp* o-> *P.emp*) , while *glt* appears to affect *P.soc* or may be linked via a confounder (*glt* o-> *P.soc*).
+-->
+<!-- Additional edges are observed in the graph generated using Gaussian CI tests, such as *slp* <-> *S.rel* and *slp* o-> *P.emp*.
+As seen in the sum graph, Gaussian CI testing tends to produce a denser graph.
+In this case, causal relationships involving *slp* are particularly prominent.
+-->
+
+<!-- Some discrepancies, however, exist between the symptom-level and sum graphs.
+For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precarity factors.
+Additionally, housing precarity (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.
+-->
+
+<!-- However, due to computational constraints that required limiting the sample size, as well as the challenging distributions of certain symptoms, the statistical power is lower compared to the depression sum score graph.
+-->
+```{r}
+#| label: fig-sym
+#| layout-nrow: 2
+#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using FCI and proportion of edge endpoint types."
+#| fig-subcap: 
+#|   - "FCI PAG"
+#|   - "Proportion matrix"
+#| fig-align: center
+#| out-width: 70%
+
+knitr::include_graphics("img/symptom_graph_FCI.png")
+knitr::include_graphics("img/symptom_mat_fci.pdf")
+
+```
+
+### Precarity as sum score
+
+Lastly, we examine the relationships between individual symptoms and overall precarity, an aggregated measure represented by the sum of five precariousness factors.
+This analysis provides a complementary perspective, capturing broader patterns that may not be evident in the disaggregated symptom-precariousness analysis.
+By aggregating precariousness factors into a single score, this approach may capture distributed relationships across different precarity factors that were previously overlooked in the disaggregated analysis.
+
+Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum panel (a)) shows more certainty in causal directions, with only a few edges resulting in ties, all of which involve the overall precarity factor.
+This is even more evident in the proportion matrix (@fig-presum panel (b)), where symptom-to-symptom interactions almost fully converge to a single edge type.
+Additionally, all symptom interactions are connected either through bidirectional edges or a combination of an arrowhead and a circle, suggesting a strong presence of latent confounders.
+
+Examining the symptom-precarity connections, we observe a stronger overall trend of depressive symptoms influencing precarity, rather than the other way around.
+Specifically, anh, dep, and slp show a high probability of causing precarity, whereas app, glt, and mot exhibit more uncertainty in directionality, often resulting in circle endpoints.
+This pattern suggests that when precariousness factors are aggregated, the dominant causal flow is from depressive symptoms to precarity, rather than precarity driving depression.
+Particularly, dep emerges as the strongest predictor of precarity, exhibiting the highest proportion of arrowtails, consistent with findings from the disaggregated analysis.
+
+Comparing this with the CCI-derived graph (@fig-presum-cci), several key differences stand out.
+The CCI results reveal that most symptoms are interconnected through bidirectional edges, suggesting a strong presence of latent confounders.
+However, dep appears to be a distinct exception, as it is predominantly caused by nearly all other symptoms, with one exception sui, which does not contribute to dep.
+Similarly, con is influenced by multiple symptoms, albeit with weaker support compared to dep.
+Most interestingly, the CCI results highlight a distinct feedback loop between dep and overall precarity, represented by a tail-tail edge (\textemdash \).
+This suggests a possible reinforcing cycle between depression and overall precariousness, where depression not only arises from precarious conditions but also contributes to their persistence, creating a self-sustaining dynamic.
+
+Overall, the aggregated precariousness graph supports the key findings from both the disaggregated analysis and the depression sum score analysis, providing greater certainty in several important patterns.
+The results highlight the significant role of latent confounders in symptom-to-symptom interactions.
+While certain stressor-related precariousness factors, particularly S.fin and S.rel, appear to causally influence depressive symptoms, other precariousness factors are more likely driven by depression rather than acting as primary causes, suggesting that precariousness may function more as a consequence of depression than its source.
+Some symptoms, particularly glt, slp, and dep, show greater sensitivity to precarious conditions, while also playing a central role in symptom dynamics.
+Among them, slp and glt emerge as influential symptoms within the depression network, whereas con and dep, despite their high connectivity, behave more as outcomes rather than primary drivers.
+Together, these findings suggest that depression—especially symptoms like sleep disturbances and guilt—may contribute to sustaining precarious conditions, possibly forming a reciprocal cycle between depression and precariousness.
+
+<!-- As before, black solid lines indicate edges agreed upon by both FCI and CCI, while gray dashed lines represent edges where the two algorithms disagree.
+Circles at the endpoints of gray dashed edges denote unresolved directionality.
+Additionally, navy dashed edges highlight relationships that appear exclusively in graphs generated using Gaussian CI tests.
+-->
+<!-- Compared to @fig-sym, this analysis reveals greater disagreement between FCI and CCI regarding edge directionality, as evidenced by more dashed edges.
+FCI tends to estimate more undirected edges (circles), while CCI more frequently assigns directed arrows between symptom-symptom relationships.
+For a detailed breakdown, see [@sec-propmatrix], which shows the proportion of symbols produced by each algorithm.
+-->
+<!-- When examining overall precarity, we observe that more symptoms are now connected to the precarity score (*Precarity*) compared to the individual precarity factor analysis in @fig-sym.
+In addition to *glt*, *slp*, *anh*, and *sui*, which are previously shown to be connected to individual precarity factors, *mot*, *app*, and *dep* are now identified as having causal relationships with overall precarity.
+Furthermore, when using the Gaussian CI test, con is also shown to be causally related to overall precarity.
+Notably, all these connections are directed towards precarity (o→), indicating that overall precarity is not the cause of symptoms.
+The exception is the bidirectional edge between *sui* <-> *Precarity*, which suggests a connection via latent confounders.
+-->
+<!-- FCI predominantly estimates directed arrows (→) for symptom-precarity relationships, suggesting that symptoms are driving precarity.
+With fewer variables (reduced from 14 to 9), this approach appears to have slightly more statistical power, enabling algorithms to estimate relationships with greater confidence compared to the previous analysis involving individual symptoms and individual precarity factors.
+-->
+
+```{r}
+#| layout-nrow: 2
+#| label: fig-presum
+#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
+#| fig-subcap: 
+#|   - "FCI PAG"
+#|   - "Proportion matrix"
+#| fig-align: center
+#| out-width: 60%
+
+knitr::include_graphics("img/presum_graph_FCI.png")
+knitr::include_graphics("img/presum_mat_fci.pdf")
+```
+
+## Simulating the Dynamics Between Precarity and Depression Under Intervention
+
+To evaluate how internal system structure influences the effect of external support, we simulated outcomes under a range of intervention strengths across the top 30 parameter sets from the calibrated model ensemble.
+The top panel of @fig-intervention shows mean levels of depression ($D$) and precarity ($P$) after simulation, with each line representing one parameter set.
+Lines are color-coded by combined feedback and input strength, calculated as the product $\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2$.
+
+Parameter sets with stronger interactions exhibit more pronounced declines in both depression and precarity in response to intervention.
+At full intervention—modeled as a complete shift of external stress ($S$) toward its observed minimum—the mean depression level decreases from -0.113 to -0.963, and precarity from -0.100 to -0.958.
+These changes correspond to reductions of approximately 90.8\% relative to their initial magnitudes, indicating that targeted intervention can shift the system toward a substantially lower-risk equilibrium.
+
+To better understand these dynamics, we focused on two illustrative parameter sets: one with strong feedback/input coupling and low noise (a "bistable" candidate), and one with weaker coupling and higher noise (a "unistable" candidate).
+In nonlinear systems, bistability refers to the existence of two distinct stable equilibria—meaning the system’s long-term outcome depends on its starting point or perturbations.
+
+In our model, bistability arises from the interaction between strong internal coupling—particularly the product $\alpha_2 \times \beta_2$—and the system’s nonlinearity.
+The $\tanh(\cdot)$ activation function saturates at high input values, meaning sufficient gain is required to push the system into nonlinear regimes where multiple stable points emerge.
+If the feedback gain is too low, or the system is overwhelmed by noise, it behaves more linearly and settles into a single equilibrium—yielding a unistable configuration.[^1]
+
+[^1]: This resembles a classic *pitchfork bifurcation* in discrete dynamical systems: when the gain $k$ in a scalar map like $x_{t+1} = \tanh(k x_t)$ exceeds a critical value, the system shifts from a single fixed point to two stable attractors and one unstable one.
+In our model, high values of $\alpha_2 \times \beta_2$ similarly enable the emergence of multiple attractor states.
+
+The bottom panel of @fig-intervention illustrates these contrasting behaviors.
+In the bistable model, a clear bifurcation emerges: small interventions have little effect, but once a threshold is crossed, the system abruptly transitions to a low-risk state.
+In the unistable model, the transition is gradual and continuous, with outcomes improving incrementally as stress is reduced.
+
+These findings underscore that the impact of intervention depends not only on its intensity, but on the system's internal configuration.
+In bistable regimes, small or moderate interventions may be insufficient unless they push the system past a tipping point.
+Unistable systems, by contrast, respond more linearly, offering greater predictability and flexibility in designing incremental interventions.
+
+This distinction is especially relevant for mental health interventions in populations exposed to persistent stressors like housing instability, financial strain, or job insecurity.
+In such contexts, external adversity may be reinforced through internal feedback loops between psychological distress and precarious conditions.
+Recognizing when a system is bistable may help identify when strong, early, or multifaceted interventions are needed to produce lasting change—while unistable dynamics may justify more gradual or targeted efforts.
+
+Tailoring intervention strategies to the system’s underlying structure may enhance their effectiveness and sustainability, particularly in addressing the compounding effects of socioeconomic precarity on mental health.
+
+<!-- The final outcomes, specifically the mean levels of depression ($D$) and precarity ($P$), are presented in the top panel of @fig-intervention.
+Each trajectory corresponds to a parameter set and is color-coded by its combined feedback and input strength, defined as the product $\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2$.
+-->
+
+<!-- Systems with stronger parameter interactions exhibit more pronounced shifts in behavior, particularly under intensive interventions.
+At full intervention (i.e., maximum stress reduction), the mean depression level decreased from -0.113 to -0.963, and precarity from -0.100 to -0.958.
+These changes represent reductions of approximately 90.8\% relative to their respective baseline magnitudes, indicating that effective intervention can push the system close to its lowest-risk equilibrium.
+-->
+
+<!-- To explore these dynamics in more detail, we focused on two representative parameter sets: one with high feedback/input strength and low noise (a bistable candidate), and another with weak parameter interactions and high noise (a unistable candidate).
+In nonlinear dynamical systems, bistability refers to the existence of two distinct stable equilibria—meaning the long-term outcome depends on initial conditions or perturbations.
+-->
+
+<!-- In our model, bistability emerges from an interplay between the strength of parameter-driven feedback/input loops and the magnitude of stochastic noise.
+The internal loop between $D$ and $P$ — governed by the product $\alpha_2 \times \beta_2$ — allows changes in one variable to reinforce the other.
+At the same time, the $\tanh(\cdot)$ nonlinearity in the system saturates for large inputs, flattening the system’s responsiveness.
+For multiple stable equilibria to emerge, the internal gain must be large enough to push the system into these nonlinear (saturated) regions.
+If this gain is too small, or if noise is too high, the system behaves more linearly and converges to a single stable point — a unistable regime.[^1]  -->
+
+<!-- [^1]: This condition resembles the classical *pitchfork* bifurcation in discrete dynamical systems: when the gain $k$ in a scalar system such as $x_{t+1} = \tanh(k x_t)$ exceeds a critical threshold (e.g., $k>1$), the system admits three fixed points—two stable and one unstable—thus producing bistability.
+In our two-variable model, sufficiently high values of $\alpha_2 \times \beta_2$ similarly give rise to a self-reinforcing loop that sustains two attractor states.
+-->
+
+<!-- This behavior is illustrated in the bottom panel of @fig-intervention.
+In the bistable model, we observe a bifurcation in final outcomes: as stress is reduced, the system undergoes a sharp transition from a high-$D$/$P$ regime to a low one—indicating a tipping point.
+In contrast, the unistable model shows a smooth, continuous response to increasing intervention, consistent with a single stable equilibrium.
+-->
+
+<!-- These findings highlight how a system’s internal structure — including both nonlinear feedback gain and noise levels — fundamentally shapes its response to intervention.
+In bistable regimes, small or moderate interventions may have little effect unless they exceed a critical threshold sufficient to tip the system out of a high-risk state.
+In unistable systems, by contrast, outcomes improve more gradually and predictably with increasing support.
+-->
+
+<!-- This distinction is particularly relevant when addressing persistent stressors associated with precariousness, such as unstable housing, employment insecurity, or chronic financial strain.
+In these contexts, the effects of external adversity may be reinforced internally through feedback loops between stress exposure and mental health.
+Such dynamics can stabilize harmful states and limit responsiveness to minor interventions—emphasizing the need for targeted, high-leverage support.
+-->
+
+<!-- Recognizing whether a population’s mental health dynamics follow a bistable or unistable structure can guide policy design.
+For bistable systems, early or forceful interventions may be required to trigger lasting improvement.
+For unistable systems, more moderate or incremental measures may suffice.
+By tailoring strategies to the system’s underlying structure, interventions can be made more effective and context-sensitive—ultimately enhancing support for those facing precarity-driven mental health risks.
+-->
+
+```{r}
+#| label: fig-intervention
+#| fig-cap: "**Top**: Final Mean D and P Across Intervention Sizes.
+*Mean trajectories of depression ($D$) and precarity ($P$) across varying intervention strengths, simulated over the top 30 parameter sets.
+Each line represents one parameter set and is colored by its combined feedback and input strength.* **Bottom**: Smoothed Density Heatmap of Final $D$ and $P$ for Two Parameter Sets.
+*One exhibits unistability (top two panels) and the other exhibits bistability (bottom two panels).
+Bistable dynamics reveal a sharp transition between distinct outcome regimes, while unistable dynamics show gradual, continuous change across the intervention gradient.*"
+#| fig-align: center
+#| out-width: 100%
+
+knitr::include_graphics("img/intervention_plot2.pdf")
+```
+
+# Discussion
+
+<!-- Summmary & Interpretation of Findings -->
+This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precarity interact—and how their internal structure modulates responsiveness to intervention.
+Using cross-sectional data from the HELIUS study, we applied cycle-capable discovery algorithms to estimate plausible directional relationships among financial stress, multiple domains of precariousness, and depressive symptoms.
+By analyzing both aggregate constructs and individual variables, we aimed to capture patterns that may be masked in either extreme of abstraction (i.e., purely aggregate or disaggregated analyses).
+
+The causal analysis consistently identified financial stress (*S.fin*) as a key plausible driver of depression, appearing across both aggregated and disaggregated models.
+At the symptom level, we found that sleep disturbance (*slp*), depressed mood (*dep*), guilt (*glt*), and anhedonia (*anh*) were especially sensitive to external stressors.
+Among these, *dep* emerged as a central hub: not only was it heavily influenced by other symptoms, it also exerted directional influence on social precarity (*P.soc*), suggesting a feedback loop in which worsening mood contributes to social disconnection, which in turn may exacerbate depressive symptoms.
+The CCI algorithm even suggested a possible cyclic relationship between dep and overall precarity, hinting at a structurally self-reinforcing mechanism.
+*Slp* also played a key initiating role.
+It showed directed connections to both social and employment precarity and exhibited many outgoing edges within the symptom network.
+This suggests *slp* may function as an early warning signal or entry point into the depression–precarity cycle, making it a promising target for preventive intervention.
+In contrast, *P.soc* more often appeared as an effect rather than a cause—receiving arrows from symptoms like *dep* and *glt*, but rarely sending them.
+This asymmetry implies that social precarity may often reflect downstream consequences of worsening mental health rather than being its root cause.
+These patterns offer actionable insights for intervention.
+Addressing symptoms like *slp* early may prevent escalation into broader depressive states or worsening precarity.
+Intervening on *dep*, as a central node with potential feedback effects, could also disrupt the self-reinforcing dynamics between mental health and social isolation.
+
+To examine how such structures shape responsiveness to support, we developed a simplified two-variable nonlinear dynamical model capturing the feedback between depression and precarity, modulated by financial stress.
+Calibrated against empirical summary statistics from the HELIUS dataset, the model allowed us to simulate how these dynamics evolve under varying levels of stress reduction.
+The simulations revealed that intervention effectiveness depends not just on the magnitude of external stressors, but on the internal configuration of the system itself.
+In bistable regimes—characterized by strong feedback and low noise—the system could remain stuck in a high-risk state unless an intervention crosses a critical threshold, tipping it into a healthier equilibrium.
+By contrast, unistable systems—driven by weaker coupling or greater noise—responded more gradually and proportionally to increasing support.
+These dynamics carry practical policy implications.
+In bistable systems, even moderate interventions may have little effect unless they are forceful enough to tip the system toward a lower-risk state.
+This underscores the importance of early or intensive support when targeting self-reinforcing cycles of depression and precarity.
+Unistable systems, by contrast, are more responsive to incremental support, suggesting that sustained, moderate interventions may be sufficient in such contexts.
+
+<!-- <!-- Interpretation of Findings --> 
+<!-- Our findings highlight the significant role of recent financial stress (S.fin) as a potential causal factor for depression, consistently observed across both sum score and symptom-level analyses.
+This consistency underscores the profound impact of financial stressors on mental health.
+The symptom-level analysis further identified specific symptoms—particularly sleep disturbances (slp), depressed mood (dep), guilt (glt), and anhedonia (anh)—as being especially sensitive to external precarious conditions.
+-->
+
+<!-- Among these symptoms, dep stands out as a distinct causal factor for precarity in both aggregated and disaggregated analyses, with CCI even suggesting a possible cyclic relationship between them.
+This implies that depression and precariousness may reinforce one another, potentially leading to a worsening cycle over time.
+Within the symptom network, dep also exhibits a high degree of connectivity, but with mostly incoming arrows, suggesting that as symptom dynamics intensify, dep is quickly impacted by multiple factors—particularly P.soc (social precarity), further fueling the depression-precariousness cycle.
+-->
+
+<!-- Beyond dep, slp emerges as a critical symptom in this dynamic.
+It connects with both social and employment precariousness, while also playing a central role in symptom interactions, affecting other symptoms more than most others.
+This suggests that slp may act as an initiator within the depressive symptom system, potentially triggering or amplifying symptom cascades.
+-->
+
+<!-- Among the precariousness factors, P.soc appears to be one of the most pivotal—not as a driver of symptoms, but rather as a consequence of depression.
+It maintains the strongest connectivity with symptoms, yet its directionality indicates that it is largely shaped by depressive symptoms rather than acting as a primary cause.
+Together, these patterns suggest a broader directional dynamic in which financial stress (S.fin) exacerbates depressive symptoms, which in turn feed into social precarity, perpetuating the cycle.
+-->
+
+<!-- This directional insight suggests that interventions targeting specific depressive symptoms, particularly dep, may help prevent downstream effects on worsening precarious conditions.
+Alternatively, early intervention focusing on slp may be a promising strategy, as slp appears to act as a potential initiator or activator within the depressive symptom network.
+This positioning suggests that slp could serve as an early warning sign, offering a valuable target for preventive interventions aimed at breaking the depression-precariousness cycle before it intensifies.
+-->
+
+<!-- Building on these insights, we developed a simplified computational model capturing the core mechanisms identified in the data: bidirectional feedback between depression and precarity, driven by external stress.
+Calibrated against empirical summary statistics, the model allowed us to explore how different parameter configurations—particularly feedback strength and stochastic noise—modulate the system’s response to simulated interventions.
+Some systems exhibited bistability, resisting change until pushed past a critical threshold, while others responded more gradually and predictably.
+These patterns have direct policy implications: in bistable regimes, intensive or early interventions may be required to shift individuals out of high-risk states, whereas in unistable systems, more moderate, sustained support may suffice.
+-->
+
+<!-- A key contribution of this work is to show that the structure of the depression–precarity system matters: not all forms of adversity respond equally to intervention, and this heterogeneity may stem from internal feedback dynamics as much as from external conditions.
+Recognizing whether a given population operates within a bistable or unistable regime could inform the design and timing of mental health interventions—particularly in contexts of chronic socioeconomic stress.
+-->
+
+<!-- Building on these findings, we developed a simplified two-variable nonlinear dynamical system to simulate the co-evolution of depression and social precarity under varying levels of financial stress.
+This model served as a conceptual abstraction of the inferred causal structure, enabling us to assess how the system’s internal configuration—particularly feedback strength, input coupling, and stochastic noise—modulates its behavior in response to intervention.
+-->
+
+<!-- Simulation results revealed that system structure critically influences responsiveness to intervention.
+In regimes with strong internal feedback loop and low noise, the model exhibits bistability: it can remain in a high-risk state unless a sufficiently large intervention tips it into a lower-risk configuration.
+By contrast, systems with weaker interactions or higher noise display unistable dynamics, where outcomes improve more gradually and proportionally with increasing intervention strength.
+-->
+
+<!-- Limitations -->
+That said, our study has several limitations.
+While the FCI and CCI algorithms are designed to recover causal structure from cross-sectional data—including feedback and latent confounding—many edge directions in our results remain ambiguous, as indicated by the prevalence of circle endpoints and bidirectional arrows, especially in weakly connected regions of the graph.
+These uncertainties were most evident in the links between individual symptoms and precariousness factors, where edge estimates varied across algorithms and conditional independence (CI) tests.
+This likely reflects limited statistical signal in high-dimensional settings, where large conditioning sets reduce power and small effect sizes are harder to detect.
+Although the CCI algorithm is explicitly designed to detect cycles, we found little clear evidence of feedback: most candidate loops appeared only as bidirectional edges, which may reflect latent confounding or statistical imprecision rather than true cyclic structure.
+This could point to either an absence of strong feedback at the symptom level or limitations in the sensitivity of full-graph discovery approaches applied to densely interconnected variables.
+More focused strategies, such as local structure tests or theory-guided subgraph modeling, may help resolve these ambiguities.
+In addition, while our analyses relied solely on observational data, some structural uncertainties—particularly those involving latent confounding or bidirectional influence—may ultimately require interventional or semi-interventional designs.
+Extensions such as LLC [@hyttinen2012learning], NODAGS-Flow [@sethuraman2023nodags], or Bicycle [@rohbeck2024], which integrate observational and interventional data, may prove valuable where such data are available.
+
+Second, our findings showed sensitivity to the choice of conditional independence (CI) test.
+The Gaussian CI test, which assumes linear relationships and Gaussian noise, often produced denser graphs—likely due to its tendency to overestimate weak or noisy linear associations.
+In contrast, RCoT, a nonparametric alternative, avoids these assumptions but relies on a Gaussian radial basis function (RBF) kernel to assess dependence.
+While flexible in capturing nonlinear relationships, this kernel is optimized for smooth, continuous variables and may underperform with discrete or ordinal data.
+Its similarity estimates can be unreliable when applied to variables with discontinuities, potentially leading to missed dependencies—particularly relevant in the HELIUS dataset, where some variables are ordinal or mixed-type [@howlett2001radial].
+To address this, future work should consider adapting kernel-based CI tests to better handle mixed data structures—for example, by employing hybrid kernels that explicitly combine continuous and discrete similarity measures.
+This could improve the reliability of nonparametric causal discovery in settings with heterogeneous variable types.
+
+Finally, the structure of the dynamical model introduces its own limitations.
+While the model was designed to reflect key features of the discovered causal structure, it represents a highly simplified system: only two variables—depression and social precarity—modulated by external financial stress.
+This abstraction allowed for tractable analysis and visual intuition, particularly around nonlinear behaviors such as bistability, but omits other potentially influential constructs highlighted in the causal graphs, including relational stress, employment conditions and housing conditions.
+Moreover, the model’s nonlinearity—introduced via the hyperbolic tangent function—was chosen for convenience rather than empirical justification.
+While this function supports saturation and the emergence of tipping points, the presence of bistability is a product of simulation, not empirical verification.
+It remains possible that simpler, linear dynamics could account for the observed data just as well.
+Future work should further investigate whether bistability is a necessary feature of depression–precarity dynamics, and explore alternative functional forms—both linear and nonlinear—using richer data and stronger empirical constraints.
+
+<!-- Several limitations should be acknowledged when interpreting these findings.
+The prevalence of bidirectional arrows and circle-marked endpoints in the graphs reflects unresolved ambiguities in the causal relationships suggested by the data.
+These uncertainties highlight the need for further research, particularly with higher-resolution datasets, such as time-series data.
+Incorporating time-series data could capture temporal dynamics between precarity factors and depressive symptoms, providing time-specific insights and revealing how these relationships evolve over time.
+Methods like *PCMCI* [@runge2019detecting], *tsFCI* [@entner2010causal], and other time-series adaptations of causal discovery algorithms could offer a more robust framework for addressing temporal dependencies and refining the analysis.
+-->
+<!-- Another notable limitation is the absence of clear evidence for cycles within the symptom network, despite employing algorithms designed to account for cyclic relationships.
+The CCI algorithm primarily produced bidirectional arrows, while FCI predominantly generated directional arrows, yet neither demonstrated patterns indicative of definitive cyclic structures.
+This suggests that addressing cyclic relationships is particularly challenging when relying solely on observational datasets.
+Future research could benefit from refined datasets that incorporate intervention data.
+Methods such as *LLC* [@hyttinen2012learning], *NODGAS-Flow *[@sethuraman2023nodags], and the recently developed *Bicycle* algorithm [@rohbeck2024], which utilize both observational and intervention data, may uncover potential cycles more effectively.
+-->
+<!-- Lastly, regarding conditional independence (CI) testing, the differences between the graphs generated by Gaussian CI and RCoT underscore the methodological sensitivities in detecting causal relationships.
+Gaussian CI produced denser graphs, which is somewhat counterintuitive, as the Gaussian CI's strict linearity assumption would typically result in fewer detected relationships, not more.
+-->
+<!-- A possible explanation for this discrepancy is that Gaussian CI's reliance on partial correlations may overestimate relationships when specific non-linear dependencies exist in the data.
+In contrast, RCoT, free from such assumptions, might better capture independence patterns under such conditions.
+-->
+<!-- However, RCoT is not without its limitations.
+While technically non-parametric, its performance can be influenced by the distributional characteristics of the data.
+The Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle with discrete or mixed datasets, where distances between discrete points may fail to convey meaningful similarity [@howlett2001radial].
+This limitation could explain why RCoT might miss certain dependencies, particularly in the HELIUS dataset, where some variables lack smooth continuity.
+-->
+<!-- Future research could tackle these challenges by developing hybrid kernels designed for mixed datasets, integrating both continuous and discrete variables into RCoT-like methods.
+This would have the potential to improve the reliability of findings, especially when working with datasets characterized by complex and heterogeneous structures.
+-->
+
+<!-- A key limitation of our study lies in the structure and assumptions of the dynamical model.
+While the model is grounded in causal patterns identified from the data, it is intentionally simplified, focusing solely on two interacting variables—depression and social precarity—modulated by financial stress.
+This abstraction facilitates tractable simulation and exploration of systemic behaviors such as bistability.
+However, it excludes several factors identified in the causal graph as potentially influential, including relational stress, housing conditions, and other contextual or protective variables like coping strategies or social support.
+These omissions constrain the model’s ability to represent the multifaceted nature of real-world dynamics.
+-->
+<!-- Additionally, the model’s nonlinearity—implemented via a hyperbolic tangent function—was chosen for its analytical properties, such as bounded outputs and saturating responses, rather than empirical grounding.
+While such nonlinearities are necessary to produce bistable dynamics, the presence of bistability in our model is a product of simulation rather than empirical verification.
+We do not currently have time-series data to confirm the existence of tipping points or regime shifts in real-world trajectories of depression and precarity.
+It is also possible that a linear model would suffice to capture much of the observed behavior, and that the current structure overcomplicates model formulation.
+Future work should seek for securing stronger evidence for bistability in the system and empirically test for the nonlinearity and compare model variants—linear and nonlinear—against richer longitudinal data to validate which formulations best capture the underlying dynamics.
+-->
+
+<!-- The numerous bidirectional arrows in the CCI outputs also imply the presence of latent confounders.
+Expanding the analysis to include additional variables, such as other mental health conditions and coping mechanisms, could provide a more comprehensive understanding of the factors influencing depressive symptoms.
+Additionally, integrating qualitative approaches could enrich these findings, offering contextual insights into the quantitative relationships identified.
+Understanding individuals' lived experiences of precarity and depression could inform the development of more tailored and effective interventions.
+-->
+
+<!-- However, it could be also that RCoT performed somewhat suboptimal.
+while RCoT is technically non-parametric, its performance can still be influenced by the distributional characteristics of the data.
+Specifically, the Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle to capture relationships in datasets with discrete or mixed distributions.
+In such cases, the distances between discrete points may fail to convey meaningful similarity information.
+As a result, RCoT might miss certain dependencies, particularly when variables in the dataset lack smooth continuity like in our HELIUS dataset.
+-->
+
+<!-- Future research could address these issues by exploring a broader range of CI testing approaches.
+One traditional approach to handle this is discretizing variables and use $G^2$ test, which may better capture dependencies in non-continuous datasets [@dojer2016learning; @neapolitan2004learning].
+A more promising direction, however, lies in the development of hybrid kernels tailored for mixed datasets, effectively integrating both continuous and discrete variables into RCoT-like methods.
+-->
+<!-- By systematically employing and comparing a wider variety of CI testing techniques, researchers could gain more robust insights and mitigate the limitations inherent in specific approaches.
+This broader exploration holds the potential to enhance the reliability of findings, particularly in datasets with complex and heterogeneous structures.
+-->
+
+<!-- Implications -->
+<!-- 1. still impressive -- symptom directions aligns much well with the other empirical studies. -->
+<!-- 2. understood better which aspect of precarity is more important for depression and we can target depends on what level of intervention we are planning. -->
+
+<!-- Implications -->
+Our results reveal that the effectiveness of interventions targeting depression and precarity is shaped not only by the strength of external stressors but also by the internal structure of the system.
+The interplay between feedback strength and noise determines whether a system responds gradually or only after crossing a tipping point—highlighting the need for intervention strategies attuned to these underlying dynamics.
+
+These findings underscore the importance of considering internal feedback—not just external adversity—when designing interventions.
+Mental health systems are not always passively responsive to support; their structure can constrain how much impact any given policy has.
+By diagnosing these structural features—through tools like causal discovery and computational modeling—researchers and policymakers can better tailor interventions to the underlying mechanics of vulnerability.
+
+Beyond these specific findings, this study offers a generalizable framework for studying complex mental health phenomena.
+Causal discovery enables principled hypothesis generation from observational data, while dynamical modeling helps test how those structures behave under intervention.
+Together, these methods move us from descriptive patterns to mechanistic insight, with practical relevance for real-world mental health policy.
+
+Future work could build richer dynamical models grounded in higher-resolution empirical data—potentially incorporating time, interventions, or more granular psychosocial constructs.
+Such extensions would help validate whether mechanisms proposed here, such as bistability, manifest in real-world trajectories.
+Ultimately, we hope this study encourages further integration of causal structure learning and system-level modeling to improve mental health outcomes under conditions of socioeconomic precarity.
+
+# References
+
+::: {#refs}
+:::
+
+# Appendix {#sec-appendix}
+
+## HELIUS study
+
+The HELIUS (HEalthy LIfe in an Urban Setting) study is a large-scale, multiethnic cohort study conducted in Amsterdam.
+Participants were randomly selected from the municipality register and stratified by ethnic origin to ensure balanced representation across six major groups.
+Invitation letters were sent by mail, followed by a reminder after two weeks.
+Non-respondents were contacted via home visits where applicable.
+
+Of those invited, approximately 55% responded (Dutch: 55%, Surinamese: 62%, Ghanaian: 57%, Turkish: 46%, Moroccan: 48%).
+Among those contacted, 50% agreed to participate (Dutch: 60%, Surinamese: 51%, Ghanaian: 61%, Turkish: 41%, Moroccan: 43%), resulting in an overall participation rate of 28%.
+
+Participants completed either a digital or paper version of the questionnaire, with assistance offered when needed, and received a confirmation letter for a physical examination appointment.
+
+At baseline (2011–2015), 24,780 individuals were enrolled.
+Of these, 23,936 participants completed the questionnaire that included items relevant to precariousness.
+After excluding individuals with missing data on key variables, the final analytical sample consisted of 21,628 participants.
+
+The HELIUS study received approval from the Medical Ethics Committee of the Academic Medical Center (AMC), and written informed consent was obtained from all participants prior to enrollment.
+
+## Causal Discovery Primer
+
+As shown in Table 1, the resulting graphs from FCI and CCI differ slightly (*PAG*: partial ancestral graph; *MAAG*: maximal almost ancestral graph) due to their reliance on different underlying assumptions.
+Despite these differences, both graphs belong to the class of *ancestral graphs*, which are designed to encode causal relationships between variables, where the presence of an edge indicates causal *ancestry.*
+In these graphs, directed edges, $A \stararrow B$, indicate that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$. 
+The Markov equivalence class represents a set of graphs that encode the same conditional independence relationships, ensuring that the same *d-separation* conditions hold across all graphs in the class [@spirtes2001causation].
+Conversely, an edge marked as $A \startail B$ indicates that $B$ is an ancestor of $A$ across all graphs in $Equiv(G)$. 
+Circle endpoints, $A \starcirc B$, represent ambiguity in the ancestral relationship, meaning $B$'s ancestral status relative to $A$ varies across graphs in $Equiv(G)$.[^4] 
+Finally, when an edge is represented as $A \arrowarrow B$, it implies that neither $A$ nor $B$ is an ancestor of the other, suggesting the presence of a latent confounder influencing both variables.
 
@@
 
-As shown in Table 1, the resulting graphs from FCI and CCI differ slightly (*PAG*: partial ancestral graph; *MAAG*: maximal almost ancestral graph) due to their reliance on different underlying assumptions.
-Despite these differences, both graphs belong to the class of *ancestral graphs*, which are designed to encode causal relationships between variables, where the presence of an edge indicates causal *ancestry.*
-In these graphs, directed edges, $A \stararrow B$, indicate that $B$ is not an ancestor of $A$ in every graph within the Markov equivalence class, $Equiv(G)$. 
-The Markov equivalence class represents a set of graphs that encode the same conditional independence relationships, ensuring that the same *d-separation* conditions hold across all graphs in the class [@spirtes2001causation].
-Conversely, an edge marked as $A \startail B$ indicates that $B$ is an ancestor of $A$ across all graphs in $Equiv(G)$. 
-Circle endpoints, $A \starcirc B$, represent ambiguity in the ancestral relationship, meaning $B$'s ancestral status relative to $A$ varies across graphs in $Equiv(G)$.[^4] 
-Finally, when an edge is represented as $A \arrowarrow B$, it implies that neither $A$ nor $B$ is an ancestor of the other, suggesting the presence of a latent confounder influencing both variables.
-
 The algorithms also differ in how they detect cycles and represent cyclic relationships in their resulting graphs.
@@
 For this reason, our primary focus remains on the results obtained from FCI and CCI, with all PC algorithm results provided in the [Appendix](#sec-appendix) for completeness.
-
-One practical challenge in applying these algorithms to the HELIUS dataset is that most variables do not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear.
-To address this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel].
-While kernel-based conditional independence tests (KCIT) are theoretically effective in capturing non-linear dependencies and subtle interactions in data with non-Gaussian noise, they are computationally demanding for large datasets like HELIUS.
-Their quadratic scaling with sample size is primarily due to the need to invert large kernel matrices, making them impractical for large-scale analysis [@rahimi2007random].
-To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximates kernel methods using random Fourier features.
-This approach reduces computational complexity from quadratic to linear scaling with sample size, significantly lowering computational costs while maintaining robustness [@strobl2019approximate].
-For a more detailed explanation of RCoT, refer to @sec-rcot.
-
-## Analysis{#sec-analysis}
-
-We analyze the causal structure using three complementary approaches: (1) Examining the relationship between the five individual precariousness factors and the PHQ sum score, which represents overall depression severity.
-(2) Investigating the relationship between individual depressive symptoms and individual precariousness factors.
-(3) Assessing the relationship between individual depressive symptoms and overall precariousness (the sum of all precariousness factors).
-Using the PHQ sum score as a measure of depression simplifies the analysis by reducing dimensionality, improving computational efficiency, and providing a broad, interpretable perspective on how precariousness factors relate to overall depression severity.
-However, this aggregated approach may overlook granular symptom-level relationships.
-
-By zooming in on individual symptoms, we capture a more detailed picture of how specific depressive symptoms interact with different precariousness factors.
-This approach allows us to map individual causal pathways, identifying which symptoms are particularly sensitive to certain precarious conditions and vice versa.
-However, this level of detail introduces methodological challenges, including non-standard distributions of symptom variables and the high-dimensional nature of the analysis, which can reduce statistical power and significantly increase computational demands—especially when using non-parametric conditional independence (CI) tests.
-
-Finally, analyzing individual symptoms in relation to overall precariousness provides an integrative perspective, revealing how depression symptoms collectively relate to precarious conditions.
-This approach complements the fully disaggregated analysis by aggregating precariousness factors, potentially uncovering relationships that may be too weak to detect when examined separately.
-This is particularly relevant when a symptom is associated with multiple precariousness factors, but these associations are not strong enough to emerge in an individual-level analysis.
-By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression.
-
-<!-- The PHQ sum score and precarity sum score capture overarching trends between overall depression and precarity, while symptom-level analysis with individual precarity factors reveals detailed dynamics, essential for tailoring specific interventions.
--->
-
-Furthermore, to address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping.
-For each condition, we generate 100 bootstrap samples, estimate causal graphs, and retain only the edge-endpoints that exceed predefined thresholds.
-If no edge-endpoint type surpasses the threshold, we assign a circle ($\circ$) to indicate uncertainty in the edge direction.
-
-The analyses are conducted under the following conditions:
-
--   Significance levels ($\alpha$): 0.01 and 0.05
--   Thresholds: 0.5, 0.6, 0.7, and 0.8
--   CI test: Gaussian CI test, RCoT
--   Algorithms: FCI, CCI, and PC
-
-This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms.
-Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm.
-For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30, easing computational demands.
-To further enhance efficiency, we fix the skeleton of the symptom network using a common structure derived from PC, FCI, and CCI algorithms ($\alpha$ = 0.01, RCoT CI test).
-This step constrains which edges can be present among symptoms but does not restrict the estimation of causal directions within the symptom network.
-Additionally, the entire causal structure between symptoms and precariousness factors remains fully estimable, ensuring that key relationships are still inferred.
-Beyond its computational benefits — particularly in reducing the time required for skeleton estimation — this fixed structure also aligns well with commonly reported symptom network structures in the literature [@park2025individual], ensuring consistency with existing findings.
-
-<!-- [^2] -->
-To obtain a final, stable summary of causal relationships, we identify the most frequently occurring edge-endpoints across different experimental setups.
-First, within each condition, we determine the most frequent edge-endpoint from all bootstrap samples.
-Then, across different conditions, we aggregate these selected symbols and determine the overall most common edge-endpoint for each graph entry.
-
-Since we have an even number of experimental setups, ties may occasionally occur.
-In such cases, we explicitly denote them in the resulting graph using dashed lines.
-To further help with the interpretation and understanding of the resulting graph, we provide a corresponding table for each graph, showing the proportion of each edge-endpoint type to clarify how dominant a given symbol is in each case.
-
-By double summarizing the results in this way, we generate one final graph per algorithm for each analysis approach: the partially aggregated analysis, which uses either the sum of precariousness factors or the sum of depression factors to capture overall trends, and the fully disaggregated analysis, which examines individual precariousness and depression factors separately.
-This approach ensures that only stable and consistent edges — those that persist across different parameter settings — are retained, leading to a robust and reliable understanding of the relationships between precariousness factors and depression.
-See @fig-workflow for an overview of the analysis workflow.
-
-<!-- [^2]: -->
-<!-- We fixed the skeleton of the symptom network using a common structure derived from the PC, FCI, and CCI algorithms with an alpha level of 0.01 using RCoT CI test.
-This approach preserved the complexity of symptom interactions, enabling unrestricted estimation of causal directions both among symptoms (*symptom --- symptom* interactions) and between symptoms and precarity factors.
--->
-
-<!-- [^3]: -->
-<!-- The detailed proportion of each edge endpoint occurrence is provided in [@sec-propmatrix]. -->
-
-```{r}
-#| label: fig-workflow
-#| fig-cap: "Analysis workflow applied across all three algorithms."
-#| fig-align: center
-#| out-width: 80%
-
-knitr::include_graphics("img/simsetup2.png")
-
-```
-<!-- \begin{figure}[htbp] -->
-<!--     \centering -->
-<!--         \caption{Simulation workflow} -->
-<!--         \includegraphics[width=0.9\textwidth]{img/simsetup.png} -->
-<!-- \end{figure} -->
-
-# Results
-## Depression as sum score
-The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
-@fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
-As described in [@sec-analysis], dashed edges indicate ties, with the exact edge endpoint where the tie occurs is represented by a bold horizontal bar.
-
-The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (PHQsum).
-Employment precarity (P.emp) and social precarity (P.soc) are not identified as causes of depression, whereas financial stress (S.fin) appears to play a causal role.
-While P.emp and P.soc are generally not recognized as causes of other precariousness factors, S.fin emerges as a potential cause, as indicated by its circle edge endpoint.
-Supporting this, the proportion matrix plot in panel (b) shows that S.fin has some probability of causally influencing either P.emp or P.soc.
-
-The matrix plot can be interpreted such that the symbol in matrix[$i$, $j$] represents the relationship $i -~[symbol] \  j$.
-For example, if matrix[$i$, $j$] is > and matrix[j, i] is $\circ$, then the inferred relationship is $i \circarrow j$.
-In the table, different edge types are represented by distinct colors: light gray for the absence of an edge, blue for circles, green for arrowheads, and coral for arrowtails.
-The colors are blended based on the proportion of each edge type, with higher proportions increasing the opacity of the corresponding color, making dominant symbols more visually prominent.
-
-Relational stress (S.rel) plays a more nuanced role, interacting with depression and S.fin  through a latent confounder.
-Its relationship with P.soc remains less certain, as the proportion table suggests a fair probability that this connection is absent, uncertain, or bidirectional.
-Meanwhile, housing precarity (P.hou) is not directly causally related to depression but is linked to P.emp and S.fin.
-While P.emp and S.fin are identified as non-causes of P.hou, it remains unclear whether P.hou causally influences P.emp or if their relationship is mediated by an unobserved confounder.
-
-The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify S.fin as a cause of depression.
-Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI.
-Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.
-
-<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precarity  (P.emp) and social precarity (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.
-Additionally, P.emp and S.fin are identified as non-causes of housing precarity (P.hou).
-Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship.
-Similar ambiguities are found in the connections between S.rel and PHQsum and between S.fin and PHQsum -->
-
-<!-- Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship.
-Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*.
--->
-
-<!-- Graph (b), derived solely from the nonparametric RCoT method.
-Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*.
--->
-<!-- The relationship between stressors and depression remains unresolved between the two algorithms.
-However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
--->
-<!-- The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.
--->
-
-<!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders.
-Employment and social precarity may also be influenced by depression, either directly or through latent variables.
-Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways.
--->
-
-```{r}
-#| layout-ncol: 2
-#| label: fig-sum
-#| fig-cap: "Resulting graph of precarity factors and depression sum score using FCI and proportion of edge endpoint types."
-#| fig-subcap: 
-#|   - "FCI PAG"
-#|   - "Proportion matrix"
-#| fig-align: center
-#| out-width: 100%
-
-knitr::include_graphics("img/FCI_depsum.png")
-knitr::include_graphics("img/depsum_mat_fci.pdf")
-
-```
-
-<!-- \begin{figure}[htbp] -->
-<!--     \centering -->
-<!--         \caption{Summary of all sum score graphs} -->
-<!--         \includegraphics[width=1\textwidth]{img/sum_all.png} -->
-<!-- \end{figure} -->
-<!-- \begin{figure}[htbp] -->
-<!--     \centering -->
-<!--         \caption{Summary of sum score graphs given RCoT} -->
-<!--         \includegraphics[width=1\textwidth]{img/sum_rcotonly.png} -->
-<!-- \end{figure} -->
-
-## Individual depression symptom
-
-Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms.
-Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph highlights heterogeneity in how precariousness factors influence individual depressive symptoms: con (concentration), slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal), anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.
-
-The symptom-level graph in @fig-sym panel (a) reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges.
-As before, dashed edges denote ties, with specific ties marked by bold horizontal bars.
-In @fig-sym panel (b), the proportion matrix provides more insight into these directionalities.
-Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type.
-However, blue regions, which correspond to circle endpoints, are particularly common for anh, slp, ene, and sui, reflecting uncertainty in the causal direction among these symptoms.
-Notably, the only arrowtail connection appears between dep and anh, suggesting that anhedonia (anh) is a potential cause of depressed mood (dep).
-
-Looking at the symptom-precariousness connections, one of the key patterns is that ties are most frequently found in the relationships between individual symptoms and precariousness factors.
-This suggests that these causal links may be less stable across different conditions.
-Upon closer examination, most ties arise due to discrepancies between the Gaussian CI test and the RCoT test—with Gaussian CI favoring arrowheads and RCoT more often predicting the absence of an edge.
-
-This discrepancy likely arises from a combination of factors. 
-First, as the analysis shifts from aggregated to disaggregated variables, the number of conditional independence (CI) tests increases, leading to a loss of statistical power in high-dimensional causal discovery.
-With more variables being conditioned on, the data becomes sparser, making it harder to detect weaker dependencies.
-Additionally, over-conditioning—controlling for too many variables—can artificially remove statistical associations, leading to false negatives.
-Beyond this general statistical power issue, the fundamental differences in how these CI tests operate contribute to the inconsistency.
-RCoT, as a nonparametric test, does not assume linearity and can capture both linear and nonlinear dependencies.
-However, it is more conservative and requires larger sample sizes, making it more prone to false negatives in high-dimensional settings.
-In contrast, partial correlation (Gaussian CI test) assumes linear Gaussian relationships, making it more permissive and sometimes detecting weak dependencies that may not be statistically meaningful.
-As a result, when Gaussian CI detects an edge while RCoT does not, it could indicate that the relationship is weak, strictly linear, or requires more data for reliable nonparametric detection.
-This explains why many ties in the proportion matrix show a 50-50 split between symbols—where half of the conditions with RCoT suggest the absence of an edge, while the other half using Gaussian CI test favor an arrowhead.
-This systematic divergence highlights the challenges of causal discovery in high-dimensional settings, where varying statistical assumptions can lead to differing conclusions about the presence and direction of causal relationships.
-
-Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of financial stress (S.fin), which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality.
-One exception is the stronger evidence suggesting that S.fin may cause changes in app (appetite), while its relationships with other symptoms, such as ene, mot, and dep, remain ambiguous.
-Similarly, social precarity (P.soc) exhibits numerous connections with depressive symptoms, with most edges pointing toward P.soc rather than outward from it.
-This suggests that depressive symptoms, particularly dep and glt, may contribute to worsening social precarity rather than the other way around.
-Additionally, anh also shows some probability of causally influencing P.soc, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.
-
-Other precariousness factors exhibit more uncertain but still notable relationships.
-Relational stress (S.rel) shows weak but existing connections with slp, glt, and app, though directionality remains unclear in many cases.
-Employment precarity (P.emp) also connects to symptoms like slp, mot, and con, but, like S.rel, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
-In line with the aggregated graph in @fig-sum, housing precarity (P.hou) does not appear to have any direct relationship with depression symptoms but consistently shows associations with employment precarity (P.emp) and financial stress (S.fin).
-The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with financial stress (S.fin) exhibiting the strongest tendency to influence other precariousness factors.
-
-Among depressive symptoms, dep, glt, and slp appear to be the most connected to precariousness factors, while P.soc has the most connections with symptoms, predominantly as a recipient rather than a driver of influence.
-Within the symptom network, slp, sui, and anh emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms.
-On the other hand, con, mot, dep, and glt, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
-Considering both symptom-precariousness connections and symptom-level dynamics, sleep disturbance (slp) emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network.
-This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precarity.
-On the other hand, social precarity (P.soc) serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.
-
-Finally, the results from CCI are largely consistent with those from FCI, but with some key differences.
-CCI tends to favor arrowheads more frequently, resulting in a greater number of bidirectional edges, which suggests a higher involvement of latent confounders.
-Additionally, CCI exhibits more tie situations, though, similar to FCI, most ties occur between absence and arrowhead edges, reflecting the discrepancy between the Gaussian CI test and RCoT—where the Gaussian test favors arrowheads, while RCoT more often suggests the absence of an edge.
-For a detailed visualization of the CCI-derived graph and proportion matrix, see @fig-sym-cci.
-
-<!-- <!-- Centrality and Connectivity --> 
-<!-- Certain nodes in the network emerge as more *causally* central, highlighting their potential importance as intervention points.
-Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network.
-Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precarity factors.
-These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes.
--->
-<!-- <!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. --> 
-<!-- The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs.
-As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precarity (*P.soc*) are more likely to be influenced by depressive symptoms.
--->
-
-<!-- The symptom-level analysis, however, provides more specificity by pinpointing the symptoms involved in these relationships.
-For instance, *slp* is identified as influencing *P.emp* or potentially through a latent confounder (*slp* o-> *P.emp*) , while *glt* appears to affect *P.soc* or may be linked via a confounder (*glt* o-> *P.soc*).
--->
-<!-- Additional edges are observed in the graph generated using Gaussian CI tests, such as *slp* <-> *S.rel* and *slp* o-> *P.emp*.
-As seen in the sum graph, Gaussian CI testing tends to produce a denser graph.
-In this case, causal relationships involving *slp* are particularly prominent.
--->
-
-<!-- Some discrepancies, however, exist between the symptom-level and sum graphs.
-For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precarity factors.
-Additionally, housing precarity (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.
--->
-
-<!-- However, due to computational constraints that required limiting the sample size, as well as the challenging distributions of certain symptoms, the statistical power is lower compared to the depression sum score graph.
--->
-```{r}
-#| label: fig-sym
-#| layout-nrow: 2
-#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using FCI and proportion of edge endpoint types."
-#| fig-subcap: 
-#|   - "FCI PAG"
-#|   - "Proportion matrix"
-#| fig-align: center
-#| out-width: 70%
-
-knitr::include_graphics("img/symptom_graph_FCI.png")
-knitr::include_graphics("img/symptom_mat_fci.pdf")
-
-```
-
-## Precarity as sum score
-
-Lastly, we examine the relationships between individual symptoms and overall precarity, an aggregated measure represented by the sum of five precariousness factors.
-This analysis provides a complementary perspective, capturing broader patterns that may not be evident in the disaggregated symptom-precariousness analysis.
-By aggregating precariousness factors into a single score, this approach may capture distributed relationships across different precarity factors that were previously overlooked in the disaggregated analysis.
-
-Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum panel (a)) shows more certainty in causal directions, with only a few edges resulting in ties, all of which involve the overall precarity factor.
-This is even more evident in the proportion matrix (@fig-presum panel (b)), where symptom-to-symptom interactions almost fully converge to a single edge type.
-Additionally, all symptom interactions are connected either through bidirectional edges or a combination of an arrowhead and a circle, suggesting a strong presence of latent confounders.
-
-Examining the symptom-precarity connections, we observe a stronger overall trend of depressive symptoms influencing precarity, rather than the other way around.
-Specifically, anh, dep, and slp show a high probability of causing precarity, whereas app, glt, and mot exhibit more uncertainty in directionality, often resulting in circle endpoints.
-This pattern suggests that when precariousness factors are aggregated, the dominant causal flow is from depressive symptoms to precarity, rather than precarity driving depression.
-Particularly, dep emerges as the strongest predictor of precarity, exhibiting the highest proportion of arrowtails, consistent with findings from the disaggregated analysis.
-
-Comparing this with the CCI-derived graph (@fig-presum-cci), several key differences stand out.
-The CCI results reveal that most symptoms are interconnected through bidirectional edges, suggesting a strong presence of latent confounders.
-However, dep appears to be a distinct exception, as it is predominantly caused by nearly all other symptoms, with one exception sui, which does not contribute to dep.
-Similarly, con is influenced by multiple symptoms, albeit with weaker support compared to dep.
-Most interestingly, the CCI results highlight a distinct feedback loop between dep and overall precarity, represented by a tail-tail edge (\textemdash \).
-This suggests a possible reinforcing cycle between depression and overall precariousness, where depression not only arises from precarious conditions but also contributes to their persistence, creating a self-sustaining dynamic.
-
-Overall, the aggregated precariousness graph supports the key findings from both the disaggregated analysis and the depression sum score analysis, providing greater certainty in several important patterns.
-The results highlight the significant role of latent confounders in symptom-to-symptom interactions.
-While certain stressor-related precariousness factors, particularly S.fin and S.rel, appear to causally influence depressive symptoms, other precariousness factors are more likely driven by depression rather than acting as primary causes, suggesting that precariousness may function more as a consequence of depression than its source.
-Some symptoms, particularly glt, slp, and dep, show greater sensitivity to precarious conditions, while also playing a central role in symptom dynamics.
-Among them, slp and glt emerge as influential symptoms within the depression network, whereas con and dep, despite their high connectivity, behave more as outcomes rather than primary drivers.
-Together, these findings suggest that depression—especially symptoms like sleep disturbances and guilt—may contribute to sustaining precarious conditions, possibly forming a reciprocal cycle between depression and precariousness.
-
-<!-- As before, black solid lines indicate edges agreed upon by both FCI and CCI, while gray dashed lines represent edges where the two algorithms disagree.
-Circles at the endpoints of gray dashed edges denote unresolved directionality.
-Additionally, navy dashed edges highlight relationships that appear exclusively in graphs generated using Gaussian CI tests.
--->
-<!-- Compared to @fig-sym, this analysis reveals greater disagreement between FCI and CCI regarding edge directionality, as evidenced by more dashed edges.
-FCI tends to estimate more undirected edges (circles), while CCI more frequently assigns directed arrows between symptom-symptom relationships.
-For a detailed breakdown, see [@sec-propmatrix], which shows the proportion of symbols produced by each algorithm.
--->
-<!-- When examining overall precarity, we observe that more symptoms are now connected to the precarity score (*Precarity*) compared to the individual precarity factor analysis in @fig-sym.
-In addition to *glt*, *slp*, *anh*, and *sui*, which are previously shown to be connected to individual precarity factors, *mot*, *app*, and *dep* are now identified as having causal relationships with overall precarity.
-Furthermore, when using the Gaussian CI test, con is also shown to be causally related to overall precarity.
-Notably, all these connections are directed towards precarity (o→), indicating that overall precarity is not the cause of symptoms.
-The exception is the bidirectional edge between *sui* <-> *Precarity*, which suggests a connection via latent confounders.
--->
-<!-- FCI predominantly estimates directed arrows (→) for symptom-precarity relationships, suggesting that symptoms are driving precarity.
-With fewer variables (reduced from 14 to 9), this approach appears to have slightly more statistical power, enabling algorithms to estimate relationships with greater confidence compared to the previous analysis involving individual symptoms and individual precarity factors.
--->
-
-```{r}
-#| layout-nrow: 2
-#| label: fig-presum
-#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
-#| fig-subcap: 
-#|   - "FCI PAG"
-#|   - "Proportion matrix"
-#| fig-align: center
-#| out-width: 60%
-
-knitr::include_graphics("img/presum_graph_FCI.png")
-knitr::include_graphics("img/presum_mat_fci.pdf")
-```
-
-# Discussion
-
-This study examined the causal relationships between precariousness factors and depression using causal discovery algorithms, integrating both sum score and symptom-level analyses.
-By analyzing PHQ sum scores alongside individual symptom scores, and comparing aggregated precarity scores with individual precariousness factors, we gained a comprehensive perspective on how different aspects of precariousness influence depression.
-This dual approach revealed nuanced pathways that may be overlooked when relying solely on aggregate measures, while also capturing broader interactions that might remain undetected in analyses focused on individual factors.
-
-<!-- Interpretation of Findings -->
-Our findings highlight the significant role of recent financial stress (S.fin) as a potential causal factor for depression, consistently observed across both sum score and symptom-level analyses.
-This consistency underscores the profound impact of financial stressors on mental health.
-The symptom-level analysis further identified specific symptoms—particularly sleep disturbances (slp), depressed mood (dep), guilt (glt), and anhedonia (anh)—as being especially sensitive to external precarious conditions.
-
-Among these symptoms, dep stands out as a distinct causal factor for precarity in both aggregated and disaggregated analyses, with CCI even suggesting a possible cyclic relationship between them.
-This implies that depression and precariousness may reinforce one another, potentially leading to a worsening cycle over time.
-Within the symptom network, dep also exhibits a high degree of connectivity, but with mostly incoming arrows, suggesting that as symptom dynamics intensify, dep is quickly impacted by multiple factors—particularly P.soc (social precarity), further fueling the depression-precariousness cycle.
-
-Beyond dep, slp emerges as a critical symptom in this dynamic.
-It connects with both social and employment precariousness, while also playing a central role in symptom interactions, affecting other symptoms more than most others.
-This suggests that slp may act as an initiator within the depressive symptom system, potentially triggering or amplifying symptom cascades.
-
-Among the precariousness factors, P.soc appears to be one of the most pivotal—not as a driver of symptoms, but rather as a consequence of depression.
-It maintains the strongest connectivity with symptoms, yet its directionality indicates that it is largely shaped by depressive symptoms rather than acting as a primary cause.
-Together, these patterns suggest a broader directional dynamic in which financial stress (S.fin) exacerbates depressive symptoms, which in turn feed into social precarity, perpetuating the cycle.
-
-This directional insight suggests that interventions targeting specific depressive symptoms, particularly dep, may help prevent downstream effects on worsening precarious conditions.
-Alternatively, early intervention focusing on slp may be a promising strategy, as slp appears to act as a potential initiator or activator within the depressive symptom network.
-This positioning suggests that slp could serve as an early warning sign, offering a valuable target for preventive interventions aimed at breaking the depression-precariousness cycle before it intensifies.
-
-<!-- emerged as potential initiators or activators within the depressive symptom network, suggesting that they may serve as early warning signs or valuable targets for preventive interventions. -->
-
-<!-- Employment precarity (*P.emp*) and social precarity (*P.soc*) were more likely to be influenced by depressive symptoms rather than acting as direct causes.
-The symptom-level graph provided additional clarity by identifying that sleep disturbances influenced employment precarity (*slp* o-> *P.emp*) and feelings of guilt affected social precarity (*glt* o-> *P.soc*).
-This directional insight suggests that interventions targeting specific depressive symptoms may have downstream effects on improving certain aspects of precarity.
--->
-
-<!-- Interestingly, financial stress (*S.fin*) did not exhibit causal relationships with individual depressive symptoms in the symptom-level analysis, contrasting with its apparent role in the sum score graphs.
-This divergence indicates that while financial stress may influence overall depression severity, its impact on specific symptoms may not be significant.
--->
-<!-- Also, housing precarity (*P.hou*) emerged as an isolated node in the symptom-level graph, losing all connections with other precarity factors.
-This isolation suggests that housing precarity may operate independently of the depressive symptom network or that its effects are not captured within the scope of the measured variables.
--->
-
-<!-- Limitations -->
-Several limitations should be acknowledged when interpreting these findings.
-The prevalence of bidirectional arrows and circle-marked endpoints in the graphs reflects unresolved ambiguities in the causal relationships suggested by the data.
-These uncertainties highlight the need for further research, particularly with higher-resolution datasets, such as time-series data.
-Incorporating time-series data could capture temporal dynamics between precarity factors and depressive symptoms, providing time-specific insights and revealing how these relationships evolve over time.
-Methods like *PCMCI* [@runge2019detecting], *tsFCI* [@entner2010causal], and other time-series adaptations of causal discovery algorithms could offer a more robust framework for addressing temporal dependencies and refining the analysis.
-
-Another notable limitation is the absence of clear evidence for cycles within the symptom network, despite employing algorithms designed to account for cyclic relationships.
-The CCI algorithm primarily produced bidirectional arrows, while FCI predominantly generated directional arrows, yet neither demonstrated patterns indicative of definitive cyclic structures.
-This suggests that addressing cyclic relationships is particularly challenging when relying solely on observational datasets.
-Future research could benefit from refined datasets that incorporate intervention data.
-Methods such as *LLC* [@hyttinen2012learning], *NODGAS-Flow *[@sethuraman2023nodags], and the recently developed *Bicycle* algorithm [@rohbeck2024], which utilize both observational and intervention data, may uncover potential cycles more effectively.
-
-<!-- The numerous bidirectional arrows in the CCI outputs also imply the presence of latent confounders.
-Expanding the analysis to include additional variables, such as other mental health conditions and coping mechanisms, could provide a more comprehensive understanding of the factors influencing depressive symptoms.
-Additionally, integrating qualitative approaches could enrich these findings, offering contextual insights into the quantitative relationships identified.
-Understanding individuals' lived experiences of precarity and depression could inform the development of more tailored and effective interventions.
--->
-
-Lastly, regarding conditional independence (CI) testing, the differences between the graphs generated by Gaussian CI and RCoT underscore the methodological sensitivities in detecting causal relationships.
-Gaussian CI produced denser graphs, which is somewhat counterintuitive, as the Gaussian CI's strict linearity assumption would typically result in fewer detected relationships, not more.
-A possible explanation for this discrepancy is that Gaussian CI's reliance on partial correlations may overestimate relationships when specific non-linear dependencies exist in the data.
-In contrast, RCoT, free from such assumptions, might better capture independence patterns under such conditions.
-However, RCoT is not without its limitations.
-While technically non-parametric, its performance can be influenced by the distributional characteristics of the data.
-The Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle with discrete or mixed datasets, where distances between discrete points may fail to convey meaningful similarity [@howlett2001radial].
-This limitation could explain why RCoT might miss certain dependencies, particularly in the HELIUS dataset, where some variables lack smooth continuity.
-Future research could tackle these challenges by developing hybrid kernels designed for mixed datasets, integrating both continuous and discrete variables into RCoT-like methods.
-This would have the potential to improve the reliability of findings, especially when working with datasets characterized by complex and heterogeneous structures.
-
-<!-- However, it could be also that RCoT performed somewhat suboptimal.
-while RCoT is technically non-parametric, its performance can still be influenced by the distributional characteristics of the data.
-Specifically, the Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle to capture relationships in datasets with discrete or mixed distributions.
-In such cases, the distances between discrete points may fail to convey meaningful similarity information.
-As a result, RCoT might miss certain dependencies, particularly when variables in the dataset lack smooth continuity like in our HELIUS dataset.
--->
-
-<!-- Future research could address these issues by exploring a broader range of CI testing approaches.
-One traditional approach to handle this is discretizing variables and use $G^2$ test, which may better capture dependencies in non-continuous datasets [@dojer2016learning; @neapolitan2004learning].
-A more promising direction, however, lies in the development of hybrid kernels tailored for mixed datasets, effectively integrating both continuous and discrete variables into RCoT-like methods.
--->
-<!-- By systematically employing and comparing a wider variety of CI testing techniques, researchers could gain more robust insights and mitigate the limitations inherent in specific approaches.
-This broader exploration holds the potential to enhance the reliability of findings, particularly in datasets with complex and heterogeneous structures.
--->
-
-<!-- Implications -->
-<!-- 1. still impressive -- symptom directions aligns much well with the other empirical studies. -->
-<!-- 2. understood better which aspect of precarity is more important for depression and we can target depends on what level of intervention we are planning. -->
-
-<!-- Implications for Research and Practice -->
-Despite its limitations, this study marks a meaningful step toward understanding the mechanisms linking precarity factors and depression.
-By applying causal discovery methods, it moves beyond traditional association-based analyses, providing insights that can inform more precise and targeted interventions.
-While the resulting graphs are preliminary and contain unresolved ambiguities, they offer a valuable starting point for leveraging causal discovery tools to investigate the causal interplay between depression and precarity factors.
-A promising next step would involve integrating these causal structures into computational models, such as the symptom dynamic model proposed in @park2025individual.
-By simulating intervention effects, such models could provide more realistic insights into how targeted actions might influence symptom networks and precarity factors over time.
-For example, interventions focused on improving sleep hygiene or alleviating guilt could be evaluated for their cascading effects on employment and social relationships, offering actionable guidance for designing population-level mental health strategies.
-<!-- Conclusion -->
-As one of the early applications of causal discovery tools to the complex dynamics of depression and precarity factors, this study lays a foundation for future research.
-We hope it inspires further refinement of these methods and ultimately contribute to more effective solutions for alleviating depression and improving societal well-being.
-
-# References
-
-::: {#refs}
-:::
-
-# Appendix {#sec-appendix}
 
@@
 
+# Acknowledgements {.unnumbered}
+
+The HELIUS study is conducted by the Amsterdam University Medical Centers, location AMC and the Public Health Service of Amsterdam.
+Both organisations provided core support for HELIUS.
+The HELIUS study is also funded by the Dutch Heart Foundation, the Netherlands Organization for Health Research and Development (ZonMw), the European Union (FP-7), and the European Fund for the Integration of non-EU immigrants (EIF).
+We are most grateful to the participants of the HELIUS study and the management team, research nurses, interviewers, research assistants and other staff who have taken part in gathering the data of this study.
+
 <!-- ### Advantages of RCIT and RCoT -->

### 2025-05-16 draft_v1/draft_v1.tex
draft_update

@@
 \usepackage{tikz}
+\usepackage{amsmath}
 \usetikzlibrary{arrows.meta, shapes.geometric}
@@
 \hypersetup{
-  pdftitle={Causal Discovery on Precarity and Depression},
+  pdftitle={From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study},
   pdfauthor={Kyuri Park; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
@@
 
-\title{Causal Discovery on Precarity and Depression}
+\title{From Causal Discovery to Intervention Simulation: Modeling
+Precarity and Depression in the HELIUS Study}
 
@@
 
-\date{2025-02-11}
+\date{2025-05-14}
 \begin{document}
@@
 \begin{abstract}
-\noindent Understanding the causal mechanisms linking precariousness and
-depression is critical for developing effective interventions. This
-study utilizes data from the HELIUS cohort study to explore these
-relationships using advanced causal discovery methods. By applying
-algorithms such as FCI, and CCI, and combining traditional Gaussian CI
-tests with non-parametric approaches like RCoT, we investigate how
-different indicators of precariousness---including factors related to
-employment, social relations and relational stress, financial situation,
-and housing---affect depressed mood, both as a sum score and at the
-individual symptom level. Our findings reveal that relational stress
-consistently emerges as a potential causal factor for depressed mood,
-while symptoms such as sleep disturbances, guilt, and anhedonia are
-particularly sensitive to external stressors, acting as potential early
-warning signals or intervention points for prevention. Moreover, the
-results highlight complexities in the data, including the influence of
-latent confounders and the challenges of capturing cyclic relationships.
-Despite some limitations, such as unresolved ambiguities in causal
-directions and challenges with mixed data distributions, this study
-demonstrates the utility of causal discovery tools in disentangling the
-intricate interplay between social and mental health dynamics. By
-mapping these causal structures into computational models, future
-research can simulate intervention effects, providing actionable
-insights to mitigate the impact of precariousness on mental health. This
-study serves as a foundational effort, offering both methodological
-advancements and practical implications for addressing depression at a
-population level.
+\noindent This study investigates the dynamic relationship between
+precarious life conditions and depression by combining constraint-based
+causal discovery with computational modeling. Using cross-sectional data
+from the HELIUS study, we applied cycle-capable causal discovery
+algorithms to uncover directional relationships between financial
+stress, domain-specific precariousness indicators, and depressive
+symptoms. Our multi-resolution analysis---spanning both aggregate and
+symptom-level constructs---identified financial stress as a robust
+driver of depression, and highlighted key symptoms, such as sleep
+disturbance and depressed mood, as likely initiators or reinforcers of
+downstream precariousness. To explore how these structural insights
+shape intervention effectiveness, we developed a simplified nonlinear
+dynamical model simulating the co-evolution of depression and social
+precarity under varying levels of financial stress reduction. Simulation
+results revealed that the system's responsiveness depends not only on
+the intensity of external intervention but also on internal structural
+features such as feedback strength and noise. In particular, some
+configurations exhibited bistability, requiring interventions to surpass
+a critical threshold to trigger lasting improvement, while others
+responded more proportionally. Together, these findings emphasize that
+internal feedback dynamics---not just external adversity---can shape the
+impact of mental health interventions. More broadly, our study offers a
+scalable framework for integrating causal discovery and dynamical
+simulation to better understand and inform policy responses to complex
+psychosocial phenomena.
 \end{abstract}
@@
 
-Mental health problems in urban areas are on the rise, yet the
-complexity of mental health systems poses significant challenges in
-understanding the underlying mechanisms driving these issues --- let
-alone designing effective interventions. Research has long sought to
-identify underlying factors contributing to mental health problems.
-Recent research looked into the association between mental health and
-indicators of precariousness in various dimensions of life, indicating a
-high level of uncertainty and instability in people's lives, such as
-factors related to employment, social relations, finances, housing, and
-culture (ref Leonie's paper). This comprehensive perspective on
-precariousness helped to highlight how different aspects of life may be
-interconnected. While this research has advanced our understanding of
-which indicators of precariousness are related to mental health, a key
-question remains unanswered: how do these factors influence mental
-health? Specifically, the lack of directional insight --- knowing what
-influences what --- limits our ability to identify and prioritize
-effective intervention targets.
-
-This study aims to investigate the causal relationships between
-different dimensions of precariousness and mental health outcomes, to
-with a specific focus on depression, the most prevalent mental health
-issue. Using causal discovery methods, we explore how various aspects of
-precariousness contribute to depression and whether depression, in turn,
-exacerbates precariousness, potentially creating a feedback loop.
-Furthermore, we zoom in at the symptom level, examining individual
-depressive symptoms to identify those that may act as initiators ---
-triggering the activation of the symptom network --- by displaying
-heightened sensitivity to specific precariousness factors. We also
-investigate symptoms that may function as bridging elements, linking the
-dynamics of depression with precariousness by feeding back into
-precarious conditions, potentially reinforcing the cycle of mental
-health deterioration and precarious conditions. Through this analysis,
-we aim to uncover the causal mechanisms driving mental health challenges
-related to precariousness and establish a foundation for more targeted
-and effective interventions by identifying the key driving forces within
-this system.
+Mental health disorders represent a growing global health concern,
+especially in urbanized regions (Gruebner et al., 2017; World Health
+Organization, 2022). In 2019, one in eight people worldwide were living
+with a mental health condition, with the highest disability-adjusted
+life years (DALYs) due to mental and addictive disorders concentrated in
+high-income countries such as those in Northern Europe, North America,
+and Australia (Rehm \& Shield, 2019; World Health Organization, 2022).
+Despite growing awareness, governments have struggled to design
+effective responses. Mental health outcomes arise from complex,
+multi-level dynamics---intertwined with the social, economic, and
+spatial structures of urban environments---making both diagnosis and
+intervention design deeply challenging (Van Der Wal et al., 2021).
+
+Recent work has shifted focus from individual-level vulnerability to
+broader social and structural determinants. While high-income nations
+show elevated overall DALYs, within-country disparities reveal that
+income inequality is a significant predictor of mental health burden
+(Rehm \& Shield, 2019). Moreover, precarious working conditions, housing
+instability, and neighborhood disadvantage have all been linked to
+increased risk of depression and anxiety (Fone et al., 2014; Pevalin et
+al., 2017; Rönnblad et al., 2019; Rugulies et al., 2023). This growing
+literature highlights that poor mental health is not merely a personal
+issue, but one rooted in sustained socioeconomic stress and structural
+uncertainty.
+
+Building on this, recent studies have introduced the concept of
+precariousness as a multidimensional condition characterized by
+instability and lack of control across several life domains---including
+employment, finances, housing, social relations, and cultural belonging
+(Elsenburg et al., 2025). This broader view reveals how different forms
+of insecurity can co-occur and compound, forming an ecosystem of risk.
+However, despite growing evidence of association between precariousness
+and mental health, a fundamental question remains unresolved: \emph{How
+do these factors causally influence one another?} Most existing studies
+rely on cross-sectional correlations, leaving open questions about
+directionality and feedback. Without stronger causal insight, it remains
+difficult to identify priority targets for intervention or to anticipate
+unintended effects.
+
+This study aims to investigate the causal mechanisms linking
+precariousness and depression---and how those mechanisms shape system
+responses to external intervention. Specifically, we ask: \emph{How does
+the internal configuration of the precarity--depression relationship
+affect the outcomes of stress-reducing interventions?} To answer this,
+we combine two approaches. First, we use cycle-capable causal discovery
+algorithms to infer directional relationships among financial stress,
+domain-specific precariousness factors, and depressive symptoms. By
+analyzing both aggregate scores (e.g., PHQ-9 total, composite precarity
+index) and disaggregated variables (individual symptoms and life-domain
+indicators), we identify both broad associations and fine-grained causal
+pathways. Second, we translate these insights into a computational
+dynamical model that simulates how depression and precarity co-evolve
+under varying levels of external stress. This model enables us to probe
+how system properties---such as feedback strength and stochastic
+noise---govern responsiveness to intervention. In some configurations,
+the system exhibits bistability, remaining stuck in a high-risk state
+unless external support exceeds a critical threshold. In others, it
+shows unistable behavior, responding more gradually to incremental
+changes. By integrating causal discovery with dynamic simulation, we aim
+to demonstrate how the internal structure of the depression--precarity
+system shapes its susceptibility to change---and, by extension, how
+intervention strategies might be tailored to the architecture of
+underlying vulnerability.
 
@@
 
-We use data from the HELIUS study, a multi-ethnic cohort that includes
-participants of Dutch, Turkish, Moroccan, Surinamese and Ghanaian origin
-(Snijder et al., 2017). To operationalize precariousness factors, we
-draw on the framework outlined in previous research (i.e., Leonie's
-paper) and select a set of relevant variables.
-
-To ensure a robust representation of each precariousness factor, we
-conducted various exploratory analyses to identify consistent and
-meaningful factor structures. Based on these analyses, we identified
-five precariousness factors, including two related to recent stressors,
-each comprising multiple variables as outlined below. Detailed
-information on the exploratory analyses can be found in the
-\hyperref[sec-appendix]{Appendix}.
+We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study,
+which has been described in detail elsewhere (Snijder et al., 2017;
+Stronks et al., 2013). It captures the diverse population of the city of
+Amsterdam by including the six main ethnic groups and which provides
+comprehensive health and lifestyle data, including depressive symptoms
+as measured by the PHQ-9 (Galenkamp et al., 2017). To operationalize
+indicators of precariousness, we draw on the framework outlined in
+previous research (Elsenburg et al., 2025) and select a set of relevant
+variables. To ensure a robust representation of precariousness in our
+causal discovery models, we conducted various exploratory analyses to
+identify consistent and meaningful data-driven factor structures. Based
+on these analyses, we identified five precariousness factors, including
+two related to recent stressors, each comprising multiple variables as
+outlined below. DFurther details on the HELIUS study and factor
+extraction are provided in the \hyperref[sec-appendix]{Appendix}.
 
@@
 \caption{\label{fig-dist}Distributions of variables with density
-overlay. \emph{P.emp} = employment precariousness; \emph{P.hou} =
-housing precariousness; \emph{P.soc} = social precariousness;
-\emph{S.fin} = recent financial stressors; \emph{S.rel} = recent
-relational stressors; \emph{Precarity} = overall precarity;
-\emph{PHQsum} = PHQ-9 sum score; \emph{anh} = anhedonia; \emph{app} =
-appetite; \emph{con} = concentration; \emph{dep} = depressed mood;
-\emph{ene} = energy; \emph{glt} = guilty; \emph{mot} = motor; \emph{sui}
-= suicidal}
+overlay.\\
+\emph{P.emp} = employment precariousness; \emph{P.hou} = housing
+precariousness; \emph{P.soc} = social precariousness; \emph{S.fin} =
+recent financial stressors; \emph{S.rel} = recent relational stressors;
+\emph{Precarity} = overall precarity; \emph{PHQsum} = PHQ-9 sum score;
+\emph{anh} = anhedonia; \emph{app} = appetite; \emph{con} =
+concentration; \emph{dep} = depressed mood; \emph{ene} = energy;
+\emph{glt} = guilty; \emph{mot} = motor; \emph{sui} = suicidal}
 
@@
 
-\subsection{Causal Discovery}\label{causal-discovery}
-
-There are numerous causal discovery algorithms available; however, in
-this study, we focus on algorithms suited to the potential cyclic
-relationships within our system. Specifically, we use \emph{FCI} (Fast
-Causal Inference) and \emph{CCI} (Cyclic Causal Inference), both capable
-of accounting for such cycles (Mooij \& Claassen, 2020; Strobl, 2019).
-Additionally, we include the \emph{PC} algorithm as a reference, given
-its simplicity and prominence as one of the most widely known causal
-discovery methods (Spirtes et al., 2001).
-
-\begin{table}[ht]
-\centering
-\small % Make the font size smaller
-\caption{Assumptions of causal discovery algorithms}
-\begin{tabular}{lccccc}
-\toprule
-Algorithm & Acyclicity & \makecell{Causal \\ sufficiency} & 
-\makecell{Absence of \\ selection bias} & Linearity & Output \\ 
-\midrule
-PC  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & CPDAG \\ 
-FCI & $-^{a}$ & $\checkmark$ & $-^{a}$ & $-^{a}$ & PAG \\ 
-CCI & x & x & x & $\checkmark$ & \makecell{\footnotesize(partially oriented) \\ MAAG} \\ 
-\bottomrule
-\end{tabular}
-\caption*{\footnotesize{\textit{Note}.
-$^{a}$The FCI algorithm, introduced by Spirtes (1995), is a constraint-based causal discovery method for DAGs that accounts for latent confounding and selection bias.
-Mooij \& Claassen (2020) later showed its applicability to cyclic causal discovery with latent confounding under general faithfulness and Markov conditions, assuming non-linear causal relationships.}}
-\end{table}
-
-As shown in Table 1, the resulting graphs from FCI and CCI differ
-slightly (\emph{PAG}: partial ancestral graph; \emph{MAAG}: maximal
-almost ancestral graph) due to their reliance on different underlying
-assumptions. Despite these differences, both graphs belong to the class
-of \emph{ancestral graphs}, which are designed to encode causal
-relationships between variables, where the presence of an edge indicates
-causal \emph{ancestry.} In these graphs, directed edges,
-\(A \stararrow B\), indicate that \(B\) is not an ancestor of \(A\) in
-every graph within the Markov equivalence class, \(Equiv(G)\). The
-Markov equivalence class represents a set of graphs that encode the same
-conditional independence relationships, ensuring that the same
-\emph{d-separation} conditions hold across all graphs in the class
-(Spirtes et al., 2001). Conversely, an edge marked as \(A \startail B\)
-indicates that \(B\) is an ancestor of \(A\) across all graphs in
-\(Equiv(G)\). Circle endpoints, \(A \starcirc B\), represent ambiguity
-in the ancestral relationship, meaning \(B\)'s ancestral status relative
-to \(A\) varies across graphs in \(Equiv(G)\).\footnote{\(*\) serves as
-  a \textit{meta-symbol}, representing one of the three possible
-  edge-endpoints. For instance, \(A \tailstar B\) can indicate any of
-  the following edges: \(A\) \textemdash~\(B\), \(A\tailarrow B\), or
-  \(A \tailcirc B\) (Park et al., 2024).} Finally, when an edge is
-represented as \(A \arrowarrow B\), it implies that neither \(A\) nor
-\(B\) is an ancestor of the other, suggesting the presence of a latent
-confounder influencing both variables.
-
-The algorithms also differ in how they detect cycles and represent
-cyclic relationships in their resulting graphs. In CCI's MAAG, \(A\)
-\textemdash~\(B\) indicates that \(A\) is an ancestor of \(B\), and
-simultaneously \(B\) is an ancestor of \(A\), referring to a cyclic
-relationship between \(A \stackedarrows B\). FCI, on the other hand,
-identifies potential cycles more subtly. In its PAG, fully-connected
-nodes with circle endpoints (\(\circirc\)) may suggest the presence of
-cyclic structures. FCI, therefore, provides a sufficient condition to
-distinguish variables that are not part of a cycle, offering a more
-nuanced approach to handling cyclic relationships (Mooij \& Claassen,
-2020).
-
-Table 1 highlights further differences in the assumptions underlying
-these algorithms. CCI operates under the assumption of a linear system,
-whereas FCI, particularly when used to infer cyclic relationships,
-assumes a non-linear system without selection bias and adheres to more
-general faithfulness and Markov conditions (i.e.,
-\emph{\(\sigma\)-separation} and \emph{\(\sigma\)-faithfulness} setting)
-(Forré \& Mooij, 2018). When the respective assumptions of each
-algorithm are met, their inferred cyclic relationships align with those
-illustrated in Figure~\ref{fig-examplegraphs}.
-
-While CCI's MAAG provides an explicit representation of cyclic
-relationships, it has certain theoretical limitations, as the MAAG may
-not always fully preserve d-separation relations from the original
-graph, \(\mathcal{G}\) (Strobl, 2019). Given these considerations, we
-examine and compare results from both algorithms, placing more weight on
-the FCI results, which are presented in the main results section, while
-discussing CCI results where relevant, with detailed findings provided
-in the \hyperref[sec-appendix]{Appendix}. A full discussion of causal
-discovery concepts and algorithmic details is beyond the scope of this
-paper; however, readers seeking a more in-depth understanding can refer
-to Park et al. (2024) for a comprehensive exploration of these methods
-and their applications.
-
-\begin{figure}
-
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.6\textwidth,height=\textheight]{img/original_DCG.png}
-
-}
-
-\subcaption{\label{fig-examplegraphs-1}Example graph \(\mathcal{G}\)}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.6\textwidth,height=\textheight]{img/FCI_graph.png}
-
-}
-
-\subcaption{\label{fig-examplegraphs-2}Corresponding MAAG}
-
-\end{minipage}%
-%
-\begin{minipage}{0.33\linewidth}
-
-\centering{
-
-\includegraphics[width=0.6\textwidth,height=\textheight]{img/CCI_graph.png}
-
-}
-
-\subcaption{\label{fig-examplegraphs-3}Corrsponding PAG}
-
-\end{minipage}%
-
-\caption{\label{fig-examplegraphs}Example graph \(\mathcal{G}\)
-featuring a cycle and the corresponding MAAG from CCI and PAG from FCI.}
-
-\end{figure}%
-
-The graph produced by the PC algorithm is a \emph{CPDAG} (completed
-partially directed acyclic graph), where directed edges (\(A\) → \(B\))
-indicate that \(A\) is a direct cause (parent) of \(B\). Unlike FCI and
-CCI, the CPDAG does not include circle symbols. Instead, when the PC
-algorithm cannot determine the direction of causality, it represents
-this uncertainty with bidirectional arrows. While the PC algorithm
-serves as a useful reference, its strict assumptions --- acyclicity and
-the absence of latent confounders---limit its applicability in more
-complex scenarios. For this reason, our primary focus remains on the
-results obtained from FCI and CCI, with all PC algorithm results
-provided in the \hyperref[sec-appendix]{Appendix} for completeness.
-
-One practical challenge in applying these algorithms to the HELIUS
-dataset is that most variables do not follow a Gaussian distribution,
-and the relationships among variables are unlikely to be strictly
-linear. To address this, we supplement the commonly used Gaussian
-conditional independence (CI) test, which relies on partial
-correlations, with a non-parametric CI test based on reproducing kernel
-methods (Zhang et al., 2012). While kernel-based conditional
-independence tests (KCIT) are theoretically effective in capturing
-non-linear dependencies and subtle interactions in data with
-non-Gaussian noise, they are computationally demanding for large
-datasets like HELIUS. Their quadratic scaling with sample size is
-primarily due to the need to invert large kernel matrices, making them
-impractical for large-scale analysis (Rahimi \& Recht, 2007). To
-mitigate this issue, we use the Randomized Conditional Correlation Test
-(RCoT), which approximates kernel methods using random Fourier features.
-This approach reduces computational complexity from quadratic to linear
-scaling with sample size, significantly lowering computational costs
-while maintaining robustness (Strobl et al., 2019). For a more detailed
-explanation of RCoT, refer to Section~\ref{sec-rcot}.
-
-\subsection{Analysis}\label{sec-analysis}
-
-We analyze the causal structure using three complementary approaches:
-(1) Examining the relationship between the five individual
-precariousness factors and the PHQ sum score, which represents overall
-depression severity. (2) Investigating the relationship between
-individual depressive symptoms and individual precariousness factors.
-(3) Assessing the relationship between individual depressive symptoms
-and overall precariousness (the sum of all precariousness factors).
-Using the PHQ sum score as a measure of depression simplifies the
-analysis by reducing dimensionality, improving computational efficiency,
-and providing a broad, interpretable perspective on how precariousness
-factors relate to overall depression severity. However, this aggregated
-approach may overlook granular symptom-level relationships.
-
-By zooming in on individual symptoms, we capture a more detailed picture
-of how specific depressive symptoms interact with different
-precariousness factors. This approach allows us to map individual causal
-pathways, identifying which symptoms are particularly sensitive to
-certain precarious conditions and vice versa. However, this level of
-detail introduces methodological challenges, including non-standard
-distributions of symptom variables and the high-dimensional nature of
-the analysis, which can reduce statistical power and significantly
-increase computational demands---especially when using non-parametric
-conditional independence (CI) tests.
-
-Finally, analyzing individual symptoms in relation to overall
-precariousness provides an integrative perspective, revealing how
-depression symptoms collectively relate to precarious conditions. This
-approach complements the fully disaggregated analysis by aggregating
-precariousness factors, potentially uncovering relationships that may be
-too weak to detect when examined separately. This is particularly
-relevant when a symptom is associated with multiple precariousness
-factors, but these associations are not strong enough to emerge in an
-individual-level analysis. By integrating these three approaches ---
-ranging from partially aggregated to fully disaggregated --- we aim to
-achieve a balance between capturing detailed granularity and gaining a
-broader understanding of the relationship between precarity and
-depression.
-
-Furthermore, to address sensitivity to parameter values and enhance the
-robustness of our results, we implement multiple settings combined with
-bootstrapping. For each condition, we generate 100 bootstrap samples,
-estimate causal graphs, and retain only the edge-endpoints that exceed
-predefined thresholds. If no edge-endpoint type surpasses the threshold,
-we assign a circle (\(\circ\)) to indicate uncertainty in the edge
-direction.
-
-The analyses are conducted under the following conditions:
+\subsection{Causal Discovery}\label{sec-analysis}
+
+To uncover directional relationships between precariousness and
+depression, we used constraint-based causal discovery algorithms that
+allow for feedback loops and latent confounding. Specifically, we
+applied the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
+Inference} (CCI) algorithms, both of which can detect cycles and
+non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
+reference, we also included the \emph{PC} algorithm, one of the most
+widely used methods for acyclic causal discovery (Spirtes et al., 2001).
+
+A key challenge in applying these algorithms to the HELIUS dataset is
+that many variables are non-Gaussian and their relationships are likely
+nonlinear. To address this, we supplemented the commonly used Gaussian
+conditional independence (CI) test---based on partial
+correlations---with a non-parametric alternative: the \emph{Randomized
+Conditional Correlation Test} (RCoT). RCoT approximates kernel-based CI
+testing using random Fourier features, significantly reducing
+computational demands while maintaining sensitivity to nonlinear
+dependencies (Strobl et al., 2019; Zhang et al., 2012).
+
+We examined the causal structure using three complementary approaches:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  Aggregate-level analysis: relationships between five domain-specific
+  precariousness factors and the PHQ-9 sum score.
+\item
+  Fully disaggregated analysis: relationships between individual
+  precariousness items and individual depressive symptoms.
+\item
+  Mixed analysis: relationships between individual symptoms and a
+  composite index of overall precariousness.
+\end{enumerate}
+
+The aggregate analysis reduces dimensionality and yields interpretable
+summaries of how domains of precarity relate to overall depression
+severity. However, it may obscure finer-grained effects. The
+disaggregated analysis allows for precise mapping of which symptoms are
+influenced by (or influence) specific precariousness conditions, but it
+introduces complexity due to the high dimensionality and distributional
+properties of the data. The mixed approach, in turn, integrates these
+perspectives by examining how individual symptoms respond to cumulative
+precarity across domains.
+
+To ensure robust estimation and guard against sensitivity to parameter
+settings, we implemented a systematic grid of analysis conditions:
 
@@
 \item
-  Significance levels (\(\alpha\)): 0.01 and 0.05
-\item
-  Thresholds: 0.5, 0.6, 0.7, and 0.8
-\item
-  CI test: Gaussian CI test, RCoT
+  Significance levels: \(\alpha\) = 0.01 and 0.05
+\item
+  Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8
+\item
+  CI tests: Gaussian and RCoT
 \item
@@
 
-This setup yields 16 combinations (2 significance levels × 4 thresholds
-× 2 CI tests), applied across three algorithms. Each combination is
-repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per
-algorithm. For analyses involving individual symptom variables, we
-simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing
-the number of bootstrap samples to 30, easing computational demands. To
-further enhance efficiency, we fix the skeleton of the symptom network
-using a common structure derived from PC, FCI, and CCI algorithms
-(\(\alpha\) = 0.01, RCoT CI test). This step constrains which edges can
-be present among symptoms but does not restrict the estimation of causal
-directions within the symptom network. Additionally, the entire causal
-structure between symptoms and precariousness factors remains fully
-estimable, ensuring that key relationships are still inferred. Beyond
-its computational benefits --- particularly in reducing the time
-required for skeleton estimation --- this fixed structure also aligns
-well with commonly reported symptom network structures in the literature
-(Park et al., 2025), ensuring consistency with existing findings.
-
-To obtain a final, stable summary of causal relationships, we identify
-the most frequently occurring edge-endpoints across different
-experimental setups. First, within each condition, we determine the most
-frequent edge-endpoint from all bootstrap samples. Then, across
-different conditions, we aggregate these selected symbols and determine
-the overall most common edge-endpoint for each graph entry.
-
-Since we have an even number of experimental setups, ties may
-occasionally occur. In such cases, we explicitly denote them in the
-resulting graph using dashed lines. To further help with the
-interpretation and understanding of the resulting graph, we provide a
-corresponding table for each graph, showing the proportion of each
-edge-endpoint type to clarify how dominant a given symbol is in each
-case.
-
-By double summarizing the results in this way, we generate one final
-graph per algorithm for each analysis approach: the partially aggregated
-analysis, which uses either the sum of precariousness factors or the sum
-of depression factors to capture overall trends, and the fully
-disaggregated analysis, which examines individual precariousness and
-depression factors separately. This approach ensures that only stable
-and consistent edges --- those that persist across different parameter
-settings --- are retained, leading to a robust and reliable
-understanding of the relationships between precariousness factors and
-depression. See Figure~\ref{fig-workflow} for an overview of the
-analysis workflow.
+Each combination was bootstrapped 100 times for sum-score level analyses
+and 30 times for symptom-level analyses, yielding a total of 1,600
+graphs per algorithm for the full analysis. For symptom networks, we
+fixed the skeleton structure using a consensus graph estimated across
+PC, FCI, and CCI (\(\alpha\) = 0.01, RCoT). This constraint improves
+computational efficiency without limiting the estimation of directional
+edges between symptoms and precariousness variables.
+
+To summarize the results, we adopted a two-level aggregation strategy.
+First, within each analysis condition, we recorded the most frequently
+observed edge-endpoint type (e.g., directed, undirected, uncertain) from
+the bootstraps. Then, across all conditions, we tallied the most
+frequent endpoint type per edge. In cases of ties, we assigned dashed
+edges to reflect uncertainty. For transparency, we accompany each graph
+with a table showing the relative frequencies of each edge symbol.
+
+This procedure yields a final graph for each algorithm and analysis
+type---offering both robustness and interpretability. See
+Figure~\ref{fig-workflow} for an overview of the causal discovery
+pipeline.
+
+For detailed explanations of the algorithms, edge interpretations, and
+graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions
+and limitations, see the \hyperref[sec-appendix]{Appendix}.
 
@@
 
+\subsection{Modeling Dynamics between Precarity and
+Depression}\label{modeling-dynamics-between-precarity-and-depression}
+
+To explore how internal system structure shapes responses to external
+support, we implement a high-level computational model that formalizes
+key mechanisms suggested by the causal discovery analysis. Rather than
+attempting to replicate the full estimated causal graph, this model
+abstracts the essential structure into a compact, tractable system
+focused on dynamic interactions between depression and social
+precariousness, both influenced by financial stress.
+
+As illustrated in Figure~\ref{fig-concept}, the model includes three
+components: depression (\(D\)), represented by the PHQ-9 sum score,
+social precariousness (\(P\)), and external financial stress (\(S\)).
+Depression and precariousness are modeled as coupled state variables
+that influence each other over time, with \(S\) acting as a shared
+input. These dynamics are formalized as a pair of nonlinear stochastic
+differential equations (SDEs):
+
+\[
+\begin{aligned}
+dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\
+dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2 
+\end{aligned}
+\]
+
+Here, \(D\) denotes depression, \(P\) denotes social precariousness, and
+\(S\) is the external stressor associated with financial stress. The
+parameters \(\lambda_D\) and \(\lambda_P\) control the timescales over
+which variable adjusts toward its input-driven target state. The
+coefficients \(\alpha_1, \alpha_2\) and \(\beta_1, \beta_2\) represent
+the strengths of the causal influences between variables. Stochastic
+fluctuations are introduced through the terms \(\sigma_1 dW_1\) and
+\(\sigma_2 dW_2\), where \(dW_1\) and \(dW_2\) are independent Wiener
+processes. The system uses the hyperbolic tangent function (\(tanh\)) to
+bound the influence of inputs, ensuring smooth, saturating dynamics and
+preventing runaway feedback loops. Importantly, the nonlinearity
+introduced by the \(tanh\) function also allows the system to exhibit
+bistability, supporting the existence of two stable states: one
+corresponding to low depression and precariousness, and the other to
+high depression and precariousness.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.35\textwidth,height=\textheight]{img/conceptmodel.png}
+
+}
+
+\caption{\label{fig-concept}Conceptual high-level causal structure of
+precariousness and depression system.}
+
+\end{figure}%
+
+\subsubsection{Model Calibration}\label{model-calibration}
+
+To identify realistic parameter settings for the simulation model, we
+calibrated six free parameters---\(\alpha_1\), \(\alpha_2\),
+\(\beta_1\), \(\beta_2\), and noise amplitudes \(\sigma_1\),
+\(\sigma_2\)---using the Non-dominated Sorting Genetic Algorithm II
+(NSGA-II) (Deb et al., 2002), a multi-objective evolutionary
+optimization method well-suited for noisy and high-dimensional search
+spaces. The goal was to match simulated system behavior to empirical
+summary statistics derived from the HELIUS dataset, including the means
+and variances of depression (\(D\)) and social precariousness (\(P\)),
+as well as their partial correlations after adjusting for financial
+stress (\(S\)), specifically \(\rho_{DP.S}\), \(\rho_{DS.P}\), and
+\(\rho_{PS.D}\). Given the stochastic nature of the model, each
+candidate parameter set was evaluated by simulating 1,000 independent
+trajectories, each representing an individual with a fixed \(S\) value
+sampled from the observed distribution. Final-state values were
+extracted and aggregated to compute population-level summary statistics.
+Fitness was assessed as the absolute deviation from the empirical
+targets, aggregated across all objectives. NSGA-II was implemented via
+the nsga2R package (Franz \& Nakamura, 2015), with a population of 300
+evolved over 300 generations using standard hyperparameters (crossover
+probability = 0.9, mutation probability = 0.25, and tournament size =
+2). This process yielded a diverse ensemble of Pareto-efficient
+solutions balancing trade-offs between objectives. From the resulting
+Pareto front (Figure~\ref{fig-pareto}), we selected the top 30 parameter
+sets with the lowest combined error to use in the subsequent simulations
+assessing intervention effects.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth,height=\textheight]{img/paretoplot.png}
+
+}
+
+\caption{\label{fig-pareto}Pareto front of simulated parameter sets.
+\emph{Each point represents a solution evaluated by its total variance
+error (x-axis) and total partial correlation error (y-axis), with color
+indicating total mean error. Red crosses highlight the top 30 solutions
+used for simulation.}}
+
+\end{figure}%
+
+\subsubsection{Intervention Simulation}\label{intervention-simulation}
+
+To evaluate the effects of hypothetical support, we simulate scenarios
+in which each individual's financial stress level (\(S\)) is
+proportionally reduced toward the minimum observed value. This models a
+population-wide intervention applied at varying strengths --- from no
+reduction in stress (0\%) to complete alignment with the minimum
+observed stress level (100\%).
+
+For each of the top 30 calibrated parameter sets, we simulate the
+system's evolution across 300 individuals under a continuum of
+intervention levels. Each simulation run generates final values for
+\(D\) and \(P\), which are then used to assess how internal model
+parameters influence the system's behavior under changing external
+conditions.
+
 \section{Results}\label{results}
 
-\subsection{Depression as sum score}\label{depression-as-sum-score}
+\subsection{Causal Structure Linking Precarity and
+Depression}\label{causal-structure-linking-precarity-and-depression}
+
+\subsubsection{Depression as sum score}\label{depression-as-sum-score}
 
@@
 
-\subsection{Individual depression
+\subsubsection{Individual depression
 symptom}\label{individual-depression-symptom}
@@
 
-\subsection{Precarity as sum score}\label{precarity-as-sum-score}
+\subsubsection{Precarity as sum score}\label{precarity-as-sum-score}
 
@@
 
+\subsection{Simulating the Dynamics Between Precarity and Depression
+Under
+Intervention}\label{simulating-the-dynamics-between-precarity-and-depression-under-intervention}
+
+To evaluate how internal system structure influences the effect of
+external support, we simulated outcomes under a range of intervention
+strengths across the top 30 parameter sets from the calibrated model
+ensemble. The top panel of Figure~\ref{fig-intervention} shows mean
+levels of depression (\(D\)) and precarity (\(P\)) after simulation,
+with each line representing one parameter set. Lines are color-coded by
+combined feedback and input strength, calculated as the product
+\(\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2\).
+
+Parameter sets with stronger interactions exhibit more pronounced
+declines in both depression and precarity in response to intervention.
+At full intervention---modeled as a complete shift of external stress
+(\(S\)) toward its observed minimum---the mean depression level
+decreases from -0.113 to -0.963, and precarity from -0.100 to -0.958.
+These changes correspond to reductions of approximately 90.8\% relative
+to their initial magnitudes, indicating that targeted intervention can
+shift the system toward a substantially lower-risk equilibrium.
+
+To better understand these dynamics, we focused on two illustrative
+parameter sets: one with strong feedback/input coupling and low noise (a
+``bistable'' candidate), and one with weaker coupling and higher noise
+(a ``unistable'' candidate). In nonlinear systems, bistability refers to
+the existence of two distinct stable equilibria---meaning the system's
+long-term outcome depends on its starting point or perturbations.
+
+In our model, bistability arises from the interaction between strong
+internal coupling---particularly the product
+\(\alpha_2 \times \beta_2\)---and the system's nonlinearity. The
+\(\tanh(\cdot)\) activation function saturates at high input values,
+meaning sufficient gain is required to push the system into nonlinear
+regimes where multiple stable points emerge. If the feedback gain is too
+low, or the system is overwhelmed by noise, it behaves more linearly and
+settles into a single equilibrium---yielding a unistable
+configuration.\footnote{This resembles a classic \emph{pitchfork
+  bifurcation} in discrete dynamical systems: when the gain \(k\) in a
+  scalar map like \(x_{t+1} = \tanh(k x_t)\) exceeds a critical value,
+  the system shifts from a single fixed point to two stable attractors
+  and one unstable one. In our model, high values of
+  \(\alpha_2 \times \beta_2\) similarly enable the emergence of multiple
+  attractor states.}
+
+The bottom panel of Figure~\ref{fig-intervention} illustrates these
+contrasting behaviors. In the bistable model, a clear bifurcation
+emerges: small interventions have little effect, but once a threshold is
+crossed, the system abruptly transitions to a low-risk state. In the
+unistable model, the transition is gradual and continuous, with outcomes
+improving incrementally as stress is reduced.
+
+These findings underscore that the impact of intervention depends not
+only on its intensity, but on the system's internal configuration. In
+bistable regimes, small or moderate interventions may be insufficient
+unless they push the system past a tipping point. Unistable systems, by
+contrast, respond more linearly, offering greater predictability and
+flexibility in designing incremental interventions.
+
+This distinction is especially relevant for mental health interventions
+in populations exposed to persistent stressors like housing instability,
+financial strain, or job insecurity. In such contexts, external
+adversity may be reinforced through internal feedback loops between
+psychological distress and precarious conditions. Recognizing when a
+system is bistable may help identify when strong, early, or multifaceted
+interventions are needed to produce lasting change---while unistable
+dynamics may justify more gradual or targeted efforts.
+
+Tailoring intervention strategies to the system's underlying structure
+may enhance their effectiveness and sustainability, particularly in
+addressing the compounding effects of socioeconomic precarity on mental
+health.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{img/intervention_plot2.pdf}
+
+}
+
+\caption{\label{fig-intervention}\textbf{Top}: Final Mean D and P Across
+Intervention Sizes. \emph{Mean trajectories of depression (\(D\)) and
+precarity (\(P\)) across varying intervention strengths, simulated over
+the top 30 parameter sets. Each line represents one parameter set and is
+colored by its combined feedback and input strength.} \textbf{Bottom}:
+Smoothed Density Heatmap of Final \(D\) and \(P\) for Two Parameter
+Sets. \emph{One exhibits unistability (top two panels) and the other
+exhibits bistability (bottom two panels). Bistable dynamics reveal a
+sharp transition between distinct outcome regimes, while unistable
+dynamics show gradual, continuous change across the intervention
+gradient.}}
+
+\end{figure}%
+
 \section{Discussion}\label{discussion}
 
-This study examined the causal relationships between precariousness
-factors and depression using causal discovery algorithms, integrating
-both sum score and symptom-level analyses. By analyzing PHQ sum scores
-alongside individual symptom scores, and comparing aggregated precarity
-scores with individual precariousness factors, we gained a comprehensive
-perspective on how different aspects of precariousness influence
-depression. This dual approach revealed nuanced pathways that may be
-overlooked when relying solely on aggregate measures, while also
-capturing broader interactions that might remain undetected in analyses
-focused on individual factors.
-
-Our findings highlight the significant role of recent financial stress
-(S.fin) as a potential causal factor for depression, consistently
-observed across both sum score and symptom-level analyses. This
-consistency underscores the profound impact of financial stressors on
-mental health. The symptom-level analysis further identified specific
-symptoms---particularly sleep disturbances (slp), depressed mood (dep),
-guilt (glt), and anhedonia (anh)---as being especially sensitive to
-external precarious conditions.
-
-Among these symptoms, dep stands out as a distinct causal factor for
-precarity in both aggregated and disaggregated analyses, with CCI even
-suggesting a possible cyclic relationship between them. This implies
-that depression and precariousness may reinforce one another,
-potentially leading to a worsening cycle over time. Within the symptom
-network, dep also exhibits a high degree of connectivity, but with
-mostly incoming arrows, suggesting that as symptom dynamics intensify,
-dep is quickly impacted by multiple factors---particularly P.soc (social
-precarity), further fueling the depression-precariousness cycle.
-
-Beyond dep, slp emerges as a critical symptom in this dynamic. It
-connects with both social and employment precariousness, while also
-playing a central role in symptom interactions, affecting other symptoms
-more than most others. This suggests that slp may act as an initiator
-within the depressive symptom system, potentially triggering or
-amplifying symptom cascades.
-
-Among the precariousness factors, P.soc appears to be one of the most
-pivotal---not as a driver of symptoms, but rather as a consequence of
-depression. It maintains the strongest connectivity with symptoms, yet
-its directionality indicates that it is largely shaped by depressive
-symptoms rather than acting as a primary cause. Together, these patterns
-suggest a broader directional dynamic in which financial stress (S.fin)
-exacerbates depressive symptoms, which in turn feed into social
-precarity, perpetuating the cycle.
-
-This directional insight suggests that interventions targeting specific
-depressive symptoms, particularly dep, may help prevent downstream
-effects on worsening precarious conditions. Alternatively, early
-intervention focusing on slp may be a promising strategy, as slp appears
-to act as a potential initiator or activator within the depressive
-symptom network. This positioning suggests that slp could serve as an
-early warning sign, offering a valuable target for preventive
-interventions aimed at breaking the depression-precariousness cycle
-before it intensifies.
-
-Several limitations should be acknowledged when interpreting these
-findings. The prevalence of bidirectional arrows and circle-marked
-endpoints in the graphs reflects unresolved ambiguities in the causal
-relationships suggested by the data. These uncertainties highlight the
-need for further research, particularly with higher-resolution datasets,
-such as time-series data. Incorporating time-series data could capture
-temporal dynamics between precarity factors and depressive symptoms,
-providing time-specific insights and revealing how these relationships
-evolve over time. Methods like \emph{PCMCI} (Runge et al., 2019),
-\emph{tsFCI} (Entner \& Hoyer, 2010), and other time-series adaptations
-of causal discovery algorithms could offer a more robust framework for
-addressing temporal dependencies and refining the analysis.
-
-Another notable limitation is the absence of clear evidence for cycles
-within the symptom network, despite employing algorithms designed to
-account for cyclic relationships. The CCI algorithm primarily produced
-bidirectional arrows, while FCI predominantly generated directional
-arrows, yet neither demonstrated patterns indicative of definitive
-cyclic structures. This suggests that addressing cyclic relationships is
-particularly challenging when relying solely on observational datasets.
-Future research could benefit from refined datasets that incorporate
-intervention data. Methods such as \emph{LLC} (Hyttinen et al., 2012),
-\emph{NODGAS-Flow }(Sethuraman et al., 2023), and the recently developed
-\emph{Bicycle} algorithm (Rohbeck et al., 2024), which utilize both
-observational and intervention data, may uncover potential cycles more
-effectively.
-
-Lastly, regarding conditional independence (CI) testing, the differences
-between the graphs generated by Gaussian CI and RCoT underscore the
-methodological sensitivities in detecting causal relationships. Gaussian
-CI produced denser graphs, which is somewhat counterintuitive, as the
-Gaussian CI's strict linearity assumption would typically result in
-fewer detected relationships, not more. A possible explanation for this
-discrepancy is that Gaussian CI's reliance on partial correlations may
-overestimate relationships when specific non-linear dependencies exist
-in the data. In contrast, RCoT, free from such assumptions, might better
-capture independence patterns under such conditions. However, RCoT is
-not without its limitations. While technically non-parametric, its
-performance can be influenced by the distributional characteristics of
-the data. The Gaussian RBF (radial basis function) kernel, optimized for
-continuous data with smooth transitions, may struggle with discrete or
-mixed datasets, where distances between discrete points may fail to
-convey meaningful similarity (Howlett, 2001). This limitation could
-explain why RCoT might miss certain dependencies, particularly in the
-HELIUS dataset, where some variables lack smooth continuity. Future
-research could tackle these challenges by developing hybrid kernels
-designed for mixed datasets, integrating both continuous and discrete
-variables into RCoT-like methods. This would have the potential to
-improve the reliability of findings, especially when working with
-datasets characterized by complex and heterogeneous structures.
-
-Despite its limitations, this study marks a meaningful step toward
-understanding the mechanisms linking precarity factors and depression.
-By applying causal discovery methods, it moves beyond traditional
-association-based analyses, providing insights that can inform more
-precise and targeted interventions. While the resulting graphs are
-preliminary and contain unresolved ambiguities, they offer a valuable
-starting point for leveraging causal discovery tools to investigate the
-causal interplay between depression and precarity factors. A promising
-next step would involve integrating these causal structures into
-computational models, such as the symptom dynamic model proposed in Park
-et al. (2025). By simulating intervention effects, such models could
-provide more realistic insights into how targeted actions might
-influence symptom networks and precarity factors over time. For example,
-interventions focused on improving sleep hygiene or alleviating guilt
-could be evaluated for their cascading effects on employment and social
-relationships, offering actionable guidance for designing
-population-level mental health strategies. As one of the early
-applications of causal discovery tools to the complex dynamics of
-depression and precarity factors, this study lays a foundation for
-future research. We hope it inspires further refinement of these methods
-and ultimately contribute to more effective solutions for alleviating
-depression and improving societal well-being.
+This study combined constraint-based causal discovery and dynamical
+modeling to investigate how depression and socioeconomic precarity
+interact---and how their internal structure modulates responsiveness to
+intervention. Using cross-sectional data from the HELIUS study, we
+applied cycle-capable discovery algorithms to estimate plausible
+directional relationships among financial stress, multiple domains of
+precariousness, and depressive symptoms. By analyzing both aggregate
+constructs and individual variables, we aimed to capture patterns that
+may be masked in either extreme of abstraction (i.e., purely aggregate
+or disaggregated analyses).
+
+The causal analysis consistently identified financial stress
+(\emph{S.fin}) as a key plausible driver of depression, appearing across
+both aggregated and disaggregated models. At the symptom level, we found
+that sleep disturbance (\emph{slp}), depressed mood (\emph{dep}), guilt
+(\emph{glt}), and anhedonia (\emph{anh}) were especially sensitive to
+external stressors. Among these, \emph{dep} emerged as a central hub:
+not only was it heavily influenced by other symptoms, it also exerted
+directional influence on social precarity (\emph{P.soc}), suggesting a
+feedback loop in which worsening mood contributes to social
+disconnection, which in turn may exacerbate depressive symptoms. The CCI
+algorithm even suggested a possible cyclic relationship between dep and
+overall precarity, hinting at a structurally self-reinforcing mechanism.
+\emph{Slp} also played a key initiating role. It showed directed
+connections to both social and employment precarity and exhibited many
+outgoing edges within the symptom network. This suggests \emph{slp} may
+function as an early warning signal or entry point into the
+depression--precarity cycle, making it a promising target for preventive
+intervention. In contrast, \emph{P.soc} more often appeared as an effect
+rather than a cause---receiving arrows from symptoms like \emph{dep} and
+\emph{glt}, but rarely sending them. This asymmetry implies that social
+precarity may often reflect downstream consequences of worsening mental
+health rather than being its root cause. These patterns offer actionable
+insights for intervention. Addressing symptoms like \emph{slp} early may
+prevent escalation into broader depressive states or worsening
+precarity. Intervening on \emph{dep}, as a central node with potential
+feedback effects, could also disrupt the self-reinforcing dynamics
+between mental health and social isolation.
+
+To examine how such structures shape responsiveness to support, we
+developed a simplified two-variable nonlinear dynamical model capturing
+the feedback between depression and precarity, modulated by financial
+stress. Calibrated against empirical summary statistics from the HELIUS
+dataset, the model allowed us to simulate how these dynamics evolve
+under varying levels of stress reduction. The simulations revealed that
+intervention effectiveness depends not just on the magnitude of external
+stressors, but on the internal configuration of the system itself. In
+bistable regimes---characterized by strong feedback and low noise---the
+system could remain stuck in a high-risk state unless an intervention
+crosses a critical threshold, tipping it into a healthier equilibrium.
+By contrast, unistable systems---driven by weaker coupling or greater
+noise---responded more gradually and proportionally to increasing
+support. These dynamics carry practical policy implications. In bistable
+systems, even moderate interventions may have little effect unless they
+are forceful enough to tip the system toward a lower-risk state. This
+underscores the importance of early or intensive support when targeting
+self-reinforcing cycles of depression and precarity. Unistable systems,
+by contrast, are more responsive to incremental support, suggesting that
+sustained, moderate interventions may be sufficient in such contexts.
+
+That said, our study has several limitations. While the FCI and CCI
+algorithms are designed to recover causal structure from cross-sectional
+data---including feedback and latent confounding---many edge directions
+in our results remain ambiguous, as indicated by the prevalence of
+circle endpoints and bidirectional arrows, especially in weakly
+connected regions of the graph. These uncertainties were most evident in
+the links between individual symptoms and precariousness factors, where
+edge estimates varied across algorithms and conditional independence
+(CI) tests. This likely reflects limited statistical signal in
+high-dimensional settings, where large conditioning sets reduce power
+and small effect sizes are harder to detect. Although the CCI algorithm
+is explicitly designed to detect cycles, we found little clear evidence
+of feedback: most candidate loops appeared only as bidirectional edges,
+which may reflect latent confounding or statistical imprecision rather
+than true cyclic structure. This could point to either an absence of
+strong feedback at the symptom level or limitations in the sensitivity
+of full-graph discovery approaches applied to densely interconnected
+variables. More focused strategies, such as local structure tests or
+theory-guided subgraph modeling, may help resolve these ambiguities. In
+addition, while our analyses relied solely on observational data, some
+structural uncertainties---particularly those involving latent
+confounding or bidirectional influence---may ultimately require
+interventional or semi-interventional designs. Extensions such as LLC
+(Hyttinen et al., 2012), NODAGS-Flow (Sethuraman et al., 2023), or
+Bicycle (Rohbeck et al., 2024), which integrate observational and
+interventional data, may prove valuable where such data are available.
+
+Second, our findings showed sensitivity to the choice of conditional
+independence (CI) test. The Gaussian CI test, which assumes linear
+relationships and Gaussian noise, often produced denser graphs---likely
+due to its tendency to overestimate weak or noisy linear associations.
+In contrast, RCoT, a nonparametric alternative, avoids these assumptions
+but relies on a Gaussian radial basis function (RBF) kernel to assess
+dependence. While flexible in capturing nonlinear relationships, this
+kernel is optimized for smooth, continuous variables and may
+underperform with discrete or ordinal data. Its similarity estimates can
+be unreliable when applied to variables with discontinuities,
+potentially leading to missed dependencies---particularly relevant in
+the HELIUS dataset, where some variables are ordinal or mixed-type
+(Howlett, 2001). To address this, future work should consider adapting
+kernel-based CI tests to better handle mixed data structures---for
+example, by employing hybrid kernels that explicitly combine continuous
+and discrete similarity measures. This could improve the reliability of
+nonparametric causal discovery in settings with heterogeneous variable
+types.
+
+Finally, the structure of the dynamical model introduces its own
+limitations. While the model was designed to reflect key features of the
+discovered causal structure, it represents a highly simplified system:
+only two variables---depression and social precarity---modulated by
+external financial stress. This abstraction allowed for tractable
+analysis and visual intuition, particularly around nonlinear behaviors
+such as bistability, but omits other potentially influential constructs
+highlighted in the causal graphs, including relational stress,
+employment conditions and housing conditions. Moreover, the model's
+nonlinearity---introduced via the hyperbolic tangent function---was
+chosen for convenience rather than empirical justification. While this
+function supports saturation and the emergence of tipping points, the
+presence of bistability is a product of simulation, not empirical
+verification. It remains possible that simpler, linear dynamics could
+account for the observed data just as well. Future work should further
+investigate whether bistability is a necessary feature of
+depression--precarity dynamics, and explore alternative functional
+forms---both linear and nonlinear---using richer data and stronger
+empirical constraints.
+
+Our results reveal that the effectiveness of interventions targeting
+depression and precarity is shaped not only by the strength of external
+stressors but also by the internal structure of the system. The
+interplay between feedback strength and noise determines whether a
+system responds gradually or only after crossing a tipping
+point---highlighting the need for intervention strategies attuned to
+these underlying dynamics.
+
+These findings underscore the importance of considering internal
+feedback---not just external adversity---when designing interventions.
+Mental health systems are not always passively responsive to support;
+their structure can constrain how much impact any given policy has. By
+diagnosing these structural features---through tools like causal
+discovery and computational modeling---researchers and policymakers can
+better tailor interventions to the underlying mechanics of
+vulnerability.
+
+Beyond these specific findings, this study offers a generalizable
+framework for studying complex mental health phenomena. Causal discovery
+enables principled hypothesis generation from observational data, while
+dynamical modeling helps test how those structures behave under
+intervention. Together, these methods move us from descriptive patterns
+to mechanistic insight, with practical relevance for real-world mental
+health policy.
+
+Future work could build richer dynamical models grounded in
+higher-resolution empirical data---potentially incorporating time,
+interventions, or more granular psychosocial constructs. Such extensions
+would help validate whether mechanisms proposed here, such as
+bistability, manifest in real-world trajectories. Ultimately, we hope
+this study encourages further integration of causal structure learning
+and system-level modeling to improve mental health outcomes under
+conditions of socioeconomic precarity.
 
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-deb2002fast}
+Deb, K., Pratap, A., Agarwal, S., \& Meyarivan, T. (2002). A fast and
+elitist multiobjective genetic algorithm: NSGA-II. \emph{IEEE
+Transactions on Evolutionary Computation}, \emph{6}(2), 182--197.
+
 \bibitem[\citeproctext]{ref-dojer2016learning}
@@
 
+\bibitem[\citeproctext]{ref-elsenburg2025clustering}
+Elsenburg, L. K., Nicolaou, M., Galenkamp, H., Lakerveld, J., \&
+Stronks, K. (2025). The clustering of disadvantage in different life
+dimensions across ethnic groups: A network analysis of indicators of
+precariousness in the HELIUS study. \emph{Social Science \& Medicine},
+117970.
+
 \bibitem[\citeproctext]{ref-entner2010causal}
@@
 data using FCI. \emph{Probabilistic Graphical Models}, \emph{16}.
+
+\bibitem[\citeproctext]{ref-fone2014effect}
+Fone, D., White, J., Farewell, D., Kelly, M., John, G., Lloyd, K.,
+Williams, G., \& Dunstan, F. (2014). Effect of neighbourhood deprivation
+and social cohesion on mental health inequality: A multilevel
+population-based longitudinal study. \emph{Psychological Medicine},
+\emph{44}(11), 2449--2460.
 
@@
 \emph{arXiv Preprint arXiv:1807.03024}.
+
+\bibitem[\citeproctext]{ref-nsga2R}
+Franz, R., \& Nakamura, R. (2015). \emph{nsga2R: Elitist non-dominated
+sorting genetic algorithm for multi-objective optimization}.
+\url{https://CRAN.R-project.org/package=nsga2R}
+
+\bibitem[\citeproctext]{ref-galenkamp2017measurement}
+Galenkamp, H., Stronks, K., Snijder, M. B., \& Derks, E. M. (2017).
+Measurement invariance testing of the PHQ-9 in a multi-ethnic population
+in europe: The HELIUS study. \emph{BMC Psychiatry}, \emph{17}, 1--14.
+
+\bibitem[\citeproctext]{ref-gruebner2017cities}
+Gruebner, O., Rapp, M. A., Adli, M., Kluge, U., Galea, S., \& Heinz, A.
+(2017). Cities and mental health. \emph{Deutsches {Ä}rzteblatt
+International}, \emph{114}(8), 121.
 
@@
 
+\bibitem[\citeproctext]{ref-pevalin2017impact}
+Pevalin, D. J., Reeves, A., Baker, E., \& Bentley, R. (2017). The impact
+of persistent poor housing conditions on mental health: A longitudinal
+population-based study. \emph{Preventive Medicine}, \emph{105},
+304--310.
+
 \bibitem[\citeproctext]{ref-rahimi2007random}
@@
 \emph{20}.
+
+\bibitem[\citeproctext]{ref-rehm2019global}
+Rehm, J., \& Shield, K. D. (2019). Global burden of disease and the
+impact of mental and addictive disorders. \emph{Current Psychiatry
+Reports}, \emph{21}, 1--7.
 
@@
 
+\bibitem[\citeproctext]{ref-ronnblad2019precarious}
+Rönnblad, T., Grönholm, E., Jonsson, J., Koranyi, I., Orellana, C.,
+Kreshpaj, B., Chen, L., Stockfelt, L., \& Bodin, T. (2019). Precarious
+employment and mental health. \emph{Scandinavian Journal of Work,
+Environment \& Health}, \emph{45}(5), 429--443.
+
+\bibitem[\citeproctext]{ref-rugulies2023work}
+Rugulies, R., Aust, B., Greiner, B. A., Arensman, E., Kawakami, N.,
+LaMontagne, A. D., \& Madsen, I. E. (2023). Work-related causes of
+mental health conditions and interventions for their improvement in
+workplaces. \emph{The Lancet}, \emph{402}(10410), 1368--1381.
+
 \bibitem[\citeproctext]{ref-runge2019detecting}
@@
 
+\bibitem[\citeproctext]{ref-storn1997differential}
+Storn, R., \& Price, K. (1997). Differential evolution--a simple and
+efficient heuristic for global optimization over continuous spaces.
+\emph{Journal of Global Optimization}, \emph{11}, 341--359.
+
 \bibitem[\citeproctext]{ref-strobl2019}
@@
 
+\bibitem[\citeproctext]{ref-stronks2013unravelling}
+Stronks, K., Snijder, M. B., Peters, R. J., Prins, M., Schene, A. H., \&
+Zwinderman, A. H. (2013). Unravelling the impact of ethnicity on health
+in europe: The HELIUS study. \emph{BMC Public Health}, \emph{13}, 1--10.
+
+\bibitem[\citeproctext]{ref-van2021advancing}
+Van Der Wal, J. M., Van Borkulo, C. D., Deserno, M. K., Breedvelt, J.
+J., Lees, M., Lokman, J. C., Borsboom, D., Denys, D., Holst, R. J. van,
+Smidt, M. P., et al. (2021). Advancing urban mental health research:
+From complexity science to actionable targets for intervention.
+\emph{The Lancet Psychiatry}, \emph{8}(11), 991--1000.
+
+\bibitem[\citeproctext]{ref-WHO_2022}
+World Health Organization. (2022). \emph{Mental disorders}. World Health
+Organization.
+\url{https://www.who.int/news-room/fact-sheets/detail/mental-disorders}
+
 \bibitem[\citeproctext]{ref-zhang2012kernel}
@@
 
+\subsection{HELIUS study}\label{helius-study}
+
+The HELIUS (HEalthy LIfe in an Urban Setting) study is a large-scale,
+multiethnic cohort study conducted in Amsterdam. Participants were
+randomly selected from the municipality register and stratified by
+ethnic origin to ensure balanced representation across six major groups.
+Invitation letters were sent by mail, followed by a reminder after two
+weeks. Non-respondents were contacted via home visits where applicable.
+
+Of those invited, approximately 55\% responded (Dutch: 55\%, Surinamese:
+62\%, Ghanaian: 57\%, Turkish: 46\%, Moroccan: 48\%). Among those
+contacted, 50\% agreed to participate (Dutch: 60\%, Surinamese: 51\%,
+Ghanaian: 61\%, Turkish: 41\%, Moroccan: 43\%), resulting in an overall
+participation rate of 28\%.
+
+Participants completed either a digital or paper version of the
+questionnaire, with assistance offered when needed, and received a
+confirmation letter for a physical examination appointment.
+
+At baseline (2011--2015), 24,780 individuals were enrolled. Of these,
+23,936 participants completed the questionnaire that included items
+relevant to precariousness. After excluding individuals with missing
+data on key variables, the final analytical sample consisted of 21,628
+participants.
+
+The HELIUS study received approval from the Medical Ethics Committee of
+the Academic Medical Center (AMC), and written informed consent was
+obtained from all participants prior to enrollment.
+
+\subsection{Causal Discovery Primer}\label{causal-discovery-primer}
+
+As shown in Table 1, the resulting graphs from FCI and CCI differ
+slightly (\emph{PAG}: partial ancestral graph; \emph{MAAG}: maximal
+almost ancestral graph) due to their reliance on different underlying
+assumptions. Despite these differences, both graphs belong to the class
+of \emph{ancestral graphs}, which are designed to encode causal
+relationships between variables, where the presence of an edge indicates
+causal \emph{ancestry.} In these graphs, directed edges,
+\(A \stararrow B\), indicate that \(B\) is not an ancestor of \(A\) in
+every graph within the Markov equivalence class, \(Equiv(G)\). The
+Markov equivalence class represents a set of graphs that encode the same
+conditional independence relationships, ensuring that the same
+\emph{d-separation} conditions hold across all graphs in the class
+(Spirtes et al., 2001). Conversely, an edge marked as \(A \startail B\)
+indicates that \(B\) is an ancestor of \(A\) across all graphs in
+\(Equiv(G)\). Circle endpoints, \(A \starcirc B\), represent ambiguity
+in the ancestral relationship, meaning \(B\)'s ancestral status relative
+to \(A\) varies across graphs in \(Equiv(G)\).\footnote{\(*\) serves as
+  a \textit{meta-symbol}, representing one of the three possible
+  edge-endpoints. For instance, \(A \tailstar B\) can indicate any of
+  the following edges: \(A\) \textemdash~\(B\), \(A\tailarrow B\), or
+  \(A \tailcirc B\) (Park et al., 2024).} Finally, when an edge is
+represented as \(A \arrowarrow B\), it implies that neither \(A\) nor
+\(B\) is an ancestor of the other, suggesting the presence of a latent
+confounder influencing both variables.
+
+\begin{table}[ht]
+\centering
+\small % Make the font size smaller
+\caption{Assumptions of causal discovery algorithms}
+\begin{tabular}{lccccc}
+\toprule
+Algorithm & Acyclicity & \makecell{Causal \\ sufficiency} & 
+\makecell{Absence of \\ selection bias} & Linearity & Output \\ 
+\midrule
+PC  & $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & CPDAG \\ 
+FCI & $-^{a}$ & $\checkmark$ & $-^{a}$ & $-^{a}$ & PAG \\ 
+CCI & x & x & x & $\checkmark$ & \makecell{\footnotesize(partially oriented) \\ MAAG} \\ 
+\bottomrule
+\end{tabular}
+\caption*{\footnotesize{\textit{Note}.
+$^{a}$The FCI algorithm, introduced by Spirtes (1995), is a constraint-based causal discovery method for DAGs that accounts for latent confounding and selection bias.
+Mooij \& Claassen (2020) later showed its applicability to cyclic causal discovery with latent confounding under general faithfulness and Markov conditions, assuming non-linear causal relationships.}}
+\end{table}
+
+The algorithms also differ in how they detect cycles and represent
+cyclic relationships in their resulting graphs. In CCI's MAAG, \(A\)
+\textemdash~\(B\) indicates that \(A\) is an ancestor of \(B\), and
+simultaneously \(B\) is an ancestor of \(A\), referring to a cyclic
+relationship between \(A \stackedarrows B\). FCI, on the other hand,
+identifies potential cycles more subtly. In its PAG, fully-connected
+nodes with circle endpoints (\(\circirc\)) may suggest the presence of
+cyclic structures. FCI, therefore, provides a sufficient condition to
+distinguish variables that are not part of a cycle, offering a more
+nuanced approach to handling cyclic relationships (Mooij \& Claassen,
+2020).
+
+Table 1 highlights further differences in the assumptions underlying
+these algorithms. CCI operates under the assumption of a linear system,
+whereas FCI, particularly when used to infer cyclic relationships,
+assumes a non-linear system without selection bias and adheres to more
+general faithfulness and Markov conditions (i.e.,
+\emph{\(\sigma\)-separation} and \emph{\(\sigma\)-faithfulness} setting)
+(Forré \& Mooij, 2018). When the respective assumptions of each
+algorithm are met, their inferred cyclic relationships align with those
+illustrated in Figure~\ref{fig-examplegraphs}.
+
+While CCI's MAAG provides an explicit representation of cyclic
+relationships, it has certain theoretical limitations, as the MAAG may
+not always fully preserve d-separation relations from the original
+graph, \(\mathcal{G}\) (Strobl, 2019). Given these considerations, we
+examine and compare results from both algorithms, placing more weight on
+the FCI results, which are presented in the main results section, while
+discussing CCI results where relevant, with detailed findings provided
+in the \hyperref[sec-appendix]{Appendix}. A full discussion of causal
+discovery concepts and algorithmic details is beyond the scope of this
+paper; however, readers seeking a more in-depth understanding can refer
+to Park et al. (2024) for a comprehensive exploration of these methods
+and their applications.
+
+\begin{figure}
+
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/original_DCG.png}
+
+}
+
+\subcaption{\label{fig-examplegraphs-1}Example graph \(\mathcal{G}\)}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/FCI_graph.png}
+
+}
+
+\subcaption{\label{fig-examplegraphs-2}Corresponding MAAG}
+
+\end{minipage}%
+%
+\begin{minipage}{0.33\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/CCI_graph.png}
+
+}
+
+\subcaption{\label{fig-examplegraphs-3}Corrsponding PAG}
+
+\end{minipage}%
+
+\caption{\label{fig-examplegraphs}Example graph \(\mathcal{G}\)
+featuring a cycle and the corresponding MAAG from CCI and PAG from FCI.}
+
+\end{figure}%
+
+The graph produced by the PC algorithm is a \emph{CPDAG} (completed
+partially directed acyclic graph), where directed edges (\(A\) → \(B\))
+indicate that \(A\) is a direct cause (parent) of \(B\). Unlike FCI and
+CCI, the CPDAG does not include circle symbols. Instead, when the PC
+algorithm cannot determine the direction of causality, it represents
+this uncertainty with bidirectional arrows. While the PC algorithm
+serves as a useful reference, its strict assumptions --- acyclicity and
+the absence of latent confounders---limit its applicability in more
+complex scenarios. For this reason, our primary focus remains on the
+results obtained from FCI and CCI, with all PC algorithm results
+provided in the \hyperref[sec-appendix]{Appendix} for completeness.
+
 \subsection{Precariousness factors by
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-9-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-12-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-10-1.pdf}
+\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-13-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-11-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-14-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-12-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-15-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-13-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-14-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-17-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-15-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-18-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-2.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-2.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-3.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-3.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-4.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-4.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-17-1.pdf}
+\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-20-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-18-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-21-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-1.pdf}
+\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-22-1.pdf}
 \end{center}
@@
 
+\section*{Acknowledgements}\label{acknowledgements}
+\addcontentsline{toc}{section}{Acknowledgements}
+
+The HELIUS study is conducted by the Amsterdam University Medical
+Centers, location AMC and the Public Health Service of Amsterdam. Both
+organisations provided core support for HELIUS. The HELIUS study is also
+funded by the Dutch Heart Foundation, the Netherlands Organization for
+Health Research and Development (ZonMw), the European Union (FP-7), and
+the European Fund for the Integration of non-EU immigrants (EIF). We are
+most grateful to the participants of the HELIUS study and the management
+team, research nurses, interviewers, research assistants and other staff
+who have taken part in gathering the data of this study.
+
 \end{document}

### 2025-05-16 code/causal_analysis/main_analysis.R
restructure_causal_analysis

@@ -133,7 +133,7 @@ agg_data <- scaled_data |>
   dplyr::select(S_combined, P_combined, PHQsum) |>
   # Convert to a data frame
   as.data.frame()
-
+c
 
 ## =============================================================================
 ## Causal discovery: run analysis under each setup

### 2025-05-16 code/utils/libraries.R

@@ -9,20 +9,29 @@ library(purrr)       # Functional programming tools
 library(magrittr)    # Pipe operators and functional utilities
 library(stringr)     # String manipulation
 library(glue)        # Dynamic string interpolation
+
 # Data input/output
 library(haven)       # Reading SPSS .sav files
+
 # Visualization
 library(ggplot2)     # Data visualization
 library(qgraph)      # Graph-based visualizations
+
 # Parallel processing
 library(furrr)       # Parallelized mapping functions
 library(future)      # Setting up parallel execution plans
+
 # Causal discovery
 library(pcalg)       # Algorithms for causal discovery
 #devtools::install_github("kyurip/RCIT", force = TRUE)
 library(RCIT)        # Kernel-based conditional independence tests
 #devtools::install_github("kyurip/CCI_KP", force = TRUE)
 library(CCI.KP)      # CCI for causal discovery
+
 # Graph handling
 library(graph)       # Graph structure utilities
+
+# SDE handling
+library(sde)
+library(deSolve)
 

### 2025-05-16 code/causal_analysis/main_analysis.R
separate preprocessing_dat

@@ -1,139 +1,20 @@
 ## =============================================================================
 ## Script for data preprocessing and running causal discovery
 ## =============================================================================
-#' This script preprocesses a dataset, scales variables, and clusters data 
-#' related to precariousness and depression symptoms. It runs causal discovery 
+#' This script loads the pre-processed dataset, and runs causal discovery 
 #' algorithms (PC, FCI, CCI) under multiple parameter combinations and saves the 
 #' results for further analysis.
 #'
 #' @details
-#' - The dataset is scaled and processed to derive both depression sum scores 
-#'   and symptom-level data for analysis.
-#' - Causal discovery algorithms are applied on subsampled data using 
+#' - Causal discovery algorithms are applied on bootstrapped datasets using 
 #'   combinations of alpha levels, thresholds, and CI tests.
 #' - Results are saved in `.rds` format for both sum score and individual 
 #'   symptom analyses.
 ## =============================================================================
 
-# ==============================================================================
-# Load required libraries and functions
-# ==============================================================================
-source("Precarity_HELIUS/code/utils/libraries.R")   # Load libraries
-source("Precarity_HELIUS/code/utils/robust_func.R") # Load helper functions
+# load preprocessed data
+source("Precarity_HELIUS/code/utils/preprocess_dat.R") 
 
-# Load the dataset 
-dat <- read_sav("data/HELIUS_LEONIE.sav")
-
-## =============================================================================
-## Preprocess and scale the precariousness data
-## =============================================================================
-scaled_data <- dat %>%
-  # Select precariousness-related variables
-  # Employment Precariousness
-  dplyr::select(H1_Arbeidsparticipatie, H1_WerkSit, H1_RecentErv8, 
-                # Financial Precariousness
[…]

### 2025-05-16 code/intervention_simulation/param_estimation.R

@@ -1,0 +1,78 @@
+
+# parallel processing
+plan(multisession)  
+
+# NSGA2 - multioptimization
+multi_obj_fun_parallel <- function(params, N_runs = 1000) {
+  # Simulate N_runs "individuals", each with one trajectory
+  sims <- future_map(1:N_runs, ~ simulate_sde_fixed_S(params, T = 1000), .options = furrr_options(seed = TRUE))
+  
+  # Extract final time point (e.g., as if it's one observation per person)
+  final_D_values <- map_dbl(sims, ~ tail(.x$D, 1))
+  final_P_values <- map_dbl(sims, ~ tail(.x$P, 1))
+  final_S_values <- map_dbl(sims, ~ tail(.x$S, 1))
+  
+  # Summary statistics across the simulated individuals
+  mean_D <- mean(final_D_values)
+  var_D  <- var(final_D_values)
+  mean_P <- mean(final_P_values)
+  var_P  <- var(final_P_values)
+  
+  # Partial correlations across simulated individuals
+  pcor_DP <- partial_correlation(final_D_values, final_P_values, final_S_values)
+  pcor_DS <- partial_correlation(final_D_values, final_S_values, final_P_values)
+  pcor_PS <- partial_correlation(final_P_values, final_S_values, final_D_values)
+  
+  # Compare to real data (depsum)
+  return(c(
+    abs(mean_D - mean(depsum$PHQsum)),
+    abs(var_D  - var(depsum$PHQsum)),
+    abs(mean_P - mean(depsum$P.soc)),
+    abs(var_P  - var(depsum$P.soc)),
+    abs(pcor_DP - partial_correlation(depsum$PHQsum, depsum$P.soc, depsum$S.fin)),
+    abs(pcor_DS - partial_correlation(depsum$PHQsum, depsum$S.fin, depsum$P.soc)),
+    abs(pcor_PS - partial_correlation(depsum$P.soc, depsum$S.fin, depsum$PHQsum))
+  ))
+}
+
+# ======= 1. Setup =======
+library(tidyverse)
+library(furrr)
+library(reshape2)
+library(fmsb)
+
+library(nsga2R)
+
+
+result2 <- nsga2R::nsga2R(
+  fn = multi_obj_fun,
+  varNo = 6,                   # Number of parameters
+  objDim = 5,                  # Number of objectives
+  lowerBounds = rep(0, 6),
+  upperBounds = rep(3, 6),
+  popSize = 500,
+  generations = 1000,
[…]

### 2025-05-16 code/utils/preprocess_dat.R

@@ -1,0 +1,136 @@
+## =============================================================================
+## Script for data preprocessing 
+## =============================================================================
+#' This script loads, cleans, and preprocesses the HELIUS dataset.
+#' It prepares standardized variables related to precariousness and depression,
+#' constructs derived indicators and creates multiple structured datasets 
+#' used in downstream modeling.
+#'
+#' @details
+#' - Constructs derived variables capturing multiple dimensions of precariousness:
+#'   employment, social, financial, housing, and cultural.
+#' - Creates symptom-level depression variables and a PHQ sum score.
+#' - Outputs structured datasets: symptom-level (`depsym`), sum-score-level (`depsum`), 
+#'   and aggregated-level (`agg_data`) for modeling and simulation.
+## =============================================================================
+
+
+
+# ==============================================================================
+# Load required libraries and functions
+# ==============================================================================
+source("Precarity_HELIUS/code/utils/libraries.R")   # Load libraries
+source("Precarity_HELIUS/code/utils/robust_func.R") # Load helper functions
+
+# Load the dataset 
+dat <- read_sav("data/helius_dat/HELIUS_LEONIE.sav")
+
+## =============================================================================
+## Preprocess and scale the precariousness data
+## =============================================================================
+scaled_data <- dat %>%
+  # Select precariousness-related variables
+  # Employment Precariousness
+  dplyr::select(H1_Arbeidsparticipatie, H1_WerkSit, H1_RecentErv8, 
+                # Financial Precariousness
+                H1_InkHhMoeite, H1_RecentErv9,                      
+                # Housing Precariousness
+                veilig_2012, vrz_2012, P_HUURWON,                   
+                # Cultural Precariousness
[…]

### 2025-05-16 code/utils/robust_func.R

@@ -37,7 +37,7 @@
 
 
 # Load necessary packages
-source("Precarity_HELIUS/code/utils/libraries.R")
+# source("Precarity_HELIUS/code/utils/libraries.R")
 
 # Setup parallel backend
 plan(multicore); 

### 2025-05-16 code/intervention_simulation/optim_sdemodel.R
add optimization script

@@ -1,0 +1,113 @@
+## =============================================================================
+## Simulation-Based Multi-Objective Optimization for SDE Model
+## =============================================================================
+#' This script performs multi-objective calibration of a SDE model of 
+#' depression (D) and precariousness (P), influenced by 
+#' a fixed stressor (S). The model parameters are optimized using NSGA-II to 
+#' match empirical summary statistics (means, variances, partial correlations).
+## =============================================================================
+
+
+# Function to compute partial correlation manually
+partial_correlation <- function(x, y, z) {
+  r_xy <- cor(x, y)  # Pearson correlation between x and y
+  r_xz <- cor(x, z)  # Pearson correlation between x and z
+  r_yz <- cor(y, z)  # Pearson correlation between y and z
+  
+  # Compute partial correlation using the formula
+  r_xy_z <- (r_xy - r_xz * r_yz) / sqrt((1 - r_xz^2) * (1 - r_yz^2))
+  
+  return(r_xy_z)
+}
+
+
+# Function to simulate time trajectories given fixed S per individual
+simulate_sde_fixed_S <- function(params, T = 100, dt = 0.01) {
+  lambda_D <- lambda_P <- 1
+  
+  beta1  <- params[1]   # Effect of S on D
+  beta2  <- params[2]   # Effect of P on D
+  alpha1 <- params[3]   # Effect of S on P
+  alpha2 <- params[4]   # Effect of D on P
+  sigma1 <- params[5]   # Noise on P
+  sigma2 <- params[6]   # Noise on D
+  
+  D <- numeric(T)
+  P <- numeric(T)
+  
+  # Sample one S value from empirical dist for the whole trajectory
+  S_val <- sample(depsum$S.fin, 1)
+  S <- rep(S_val, T)
+  
+  for (t in 2:T) {
+    dW1 <- rnorm(1, 0, sqrt(dt))
+    dW2 <- rnorm(1, 0, sqrt(dt))
+    
+    P[t] <- P[t - 1] + lambda_P * (tanh(alpha1 * S[t] + alpha2 * D[t - 1]) - P[t - 1]) * dt + sigma1 * dW1    
+    D[t] <- D[t - 1] + lambda_D * (tanh(beta1 * S[t] + beta2 * P[t - 1]) - D[t - 1]) * dt + sigma2 * dW2
+    
+  }
+  
[…]

### 2025-05-16 code/intervention_simulation/param_estimation.R

@@ -1,78 +1,0 @@
-
-# parallel processing
-plan(multisession)  
-
-# NSGA2 - multioptimization
-multi_obj_fun_parallel <- function(params, N_runs = 1000) {
-  # Simulate N_runs "individuals", each with one trajectory
-  sims <- future_map(1:N_runs, ~ simulate_sde_fixed_S(params, T = 1000), .options = furrr_options(seed = TRUE))
-  
-  # Extract final time point (e.g., as if it's one observation per person)
-  final_D_values <- map_dbl(sims, ~ tail(.x$D, 1))
-  final_P_values <- map_dbl(sims, ~ tail(.x$P, 1))
-  final_S_values <- map_dbl(sims, ~ tail(.x$S, 1))
-  
-  # Summary statistics across the simulated individuals
-  mean_D <- mean(final_D_values)
-  var_D  <- var(final_D_values)
-  mean_P <- mean(final_P_values)
-  var_P  <- var(final_P_values)
-  
-  # Partial correlations across simulated individuals
-  pcor_DP <- partial_correlation(final_D_values, final_P_values, final_S_values)
-  pcor_DS <- partial_correlation(final_D_values, final_S_values, final_P_values)
-  pcor_PS <- partial_correlation(final_P_values, final_S_values, final_D_values)
-  
-  # Compare to real data (depsum)
-  return(c(
-    abs(mean_D - mean(depsum$PHQsum)),
-    abs(var_D  - var(depsum$PHQsum)),
-    abs(mean_P - mean(depsum$P.soc)),
-    abs(var_P  - var(depsum$P.soc)),
-    abs(pcor_DP - partial_correlation(depsum$PHQsum, depsum$P.soc, depsum$S.fin)),
-    abs(pcor_DS - partial_correlation(depsum$PHQsum, depsum$S.fin, depsum$P.soc)),
-    abs(pcor_PS - partial_correlation(depsum$P.soc, depsum$S.fin, depsum$PHQsum))
-  ))
-}
-
-# ======= 1. Setup =======
-library(tidyverse)
-library(furrr)
-library(reshape2)
-library(fmsb)
-
-library(nsga2R)
-
-
-result2 <- nsga2R::nsga2R(
-  fn = multi_obj_fun,
-  varNo = 6,                   # Number of parameters
-  objDim = 5,                  # Number of objectives
-  lowerBounds = rep(0, 6),
-  upperBounds = rep(3, 6),
-  popSize = 500,
-  generations = 1000,
[…]

### 2025-05-16 code/utils/preprocess_dat.R

@@ -23,7 +23,7 @@ source("Precarity_HELIUS/code/utils/libraries.R")   # Load libraries
 source("Precarity_HELIUS/code/utils/robust_func.R") # Load helper functions
 
 # Load the dataset 
-dat <- read_sav("data/helius_dat/HELIUS_LEONIE.sav")
+dat <- read_sav("Precarity_HELIUS/data/helius_dat/HELIUS_LEONIE.sav")
 
 ## =============================================================================
 ## Preprocess and scale the precariousness data

### 2025-05-17 code/intervention_simulation/pareto.R
pareto_plot

@@ -1,0 +1,97 @@
+## =============================================================================
+## Analyze NSGA-II Results and Plot Pareto Frontier
+## =============================================================================
+# Load libraries
+source("Precarity_HELIUS/code/utils/libraries.R")   
+
+# ---- 1. Load NSGA-II Results ----
+result <- readRDS("Precarity_HELIUS/data/intervention/multioptim_result_6var7optim_500pop_fixedS.rds") 
+
+param_df <- as.data.frame(result$parameters) %>%
+  setNames(c("beta1", "beta2", "alpha1", "alpha2", "sigma1", "sigma2"))
+
+obj_df <- as.data.frame(result$objectives) %>%
+  setNames(c("mean_D_err", "var_D_err", "mean_P_err", "var_P_err", 
+             "pcor_DP_err", "pcor_DS_err", "pcor_PS_err"))
+
+df <- bind_cols(param_df, obj_df) %>%
+  mutate(total_error = rowSums(across(ends_with("_err"))))
+
+# ---- 2. Visualize Parameter Distributions ----
+param_long <- param_df %>%
+  pivot_longer(everything(), names_to = "Parameter", values_to = "Value")
+
+ggplot(param_long, aes(x = Value, fill = Parameter)) +
+  geom_histogram(bins = 30, alpha = 0.6, color = "black") +
+  facet_wrap(~Parameter, scales = "free") +
+  theme_minimal() +
+  labs(title = "Distribution of Parameter Values (Final Population)")
+
+# ---- 3. Identify Top 5% Solutions ----
+threshold <- quantile(df$total_error, 0.05)
+df <- df %>% mutate(Top = total_error <= threshold)
+
+top_df <- df %>% filter(Top) %>% arrange(total_error)
+
+# Visualize Top Parameter Distributions
+top_df %>%
+  pivot_longer(cols = starts_with(c("beta", "alpha", "sigma")), 
+               names_to = "Parameter", values_to = "Value") %>%
+  ggplot(aes(x = Value, fill = Parameter)) +
+  geom_histogram(bins = 30, alpha = 0.7, color = "black") +
+  facet_wrap(~Parameter, scales = "free") +
+  theme_minimal() +
+  labs(title = "Distribution of Parameters (Top Solutions)")
+
+# ---- 4. Pareto Frontier Plot ----
+# Filter reasonable range for clarity
+df_filtered <- df %>% filter(
[…]

### 2025-05-18 code/intervention_simulation/01_analyze_nsga2.R
update_intervention_code

@@ -1,0 +1,66 @@
+## =============================================================================
+## Script: 01_analyze_nsga2.R
+## Analyze NSGA-II Results and Visualize Pareto Frontier
+## =============================================================================
+
+# Load libraries
+source("Precarity_HELIUS/code/utils/libraries.R")
+
+# Load NSGA-II results
+result <- readRDS("Precarity_HELIUS/data/intervention/multioptim_result_6var7optim_500pop_fixedS.rds") 
+
+# Prepare parameter and objective data
+df <- bind_cols(
+  as.data.frame(result$parameters) %>% setNames(c("beta1", "beta2", "alpha1", "alpha2", "sigma1", "sigma2")),
+  as.data.frame(result$objectives) %>% setNames(c("mean_D_err", "var_D_err", "mean_P_err", "var_P_err", "pcor_DP_err", "pcor_DS_err", "pcor_PS_err"))
+) %>%
+  mutate(total_error = rowSums(across(ends_with("_err"))))
+
+# Visualize parameter distributions
+param_long <- df %>% select(beta1:sigma2) %>%
+  pivot_longer(everything(), names_to = "Parameter", values_to = "Value")
+
+ggplot(param_long, aes(x = Value, fill = Parameter)) +
+  geom_histogram(bins = 30, alpha = 0.6, color = "black") +
+  facet_wrap(~Parameter, scales = "free") +
+  theme_minimal() +
+  labs(title = "Distribution of Parameter Values (Final Population)")
+
+# Filter and score Pareto-efficient solutions
+df_filtered <- df %>%
+  filter(
+    var_D_err + var_P_err < 2,
+    mean_D_err + mean_P_err < 2,
+    pcor_DP_err + pcor_DS_err + pcor_PS_err < 2
+  ) %>%
+  mutate(
+    total_variance_error = var_D_err + var_P_err,
+    total_partial_error = pcor_DP_err + pcor_DS_err + pcor_PS_err,
+    total_mean_error = mean_D_err + mean_P_err,
+    total_error_sum = total_variance_error + total_partial_error + total_mean_error,
+    Solution = paste0("S", row_number())
+  )
+
+# Select top solutions
+top_solutions <- df_filtered %>% slice_min(total_error_sum, n = 30)
+
+# Plot Pareto front
+paretoplot <- ggplot(df_filtered, aes(x = total_variance_error, y = total_partial_error)) +
[…]

### 2025-05-18 code/intervention_simulation/02_simulate_intervention.R

@@ -1,0 +1,118 @@
+## =============================================================================
+## Script: 02_simulate_intervention.R
+## Simulate D–P System Dynamics Under External Stress Interventions
+## =============================================================================
+# Description:
+# This script runs simulations of the depression (D) and precariousness (P) system
+# under varying levels of stress reduction (interventions on S). It uses top 
+# parameter sets previously identified via multi-objective optimization, simulates
+# post-intervention trajectories, and saves both aggregate trends and detailed 
+# outcomes for selected candidates (Sol_1 and Sol_21).
+## =============================================================================
+
+# parallel processing
+plan(multisession)
+
+# ------------------------------------------------------------------------------
+# 1. Load and Annotate Top Parameter Sets
+# ------------------------------------------------------------------------------
+
+top_solutions <- readRDS("Precarity_HELIUS/data/intervention/top_solutions_nsga2.rds")
+
+# Add derived features for parameter diagnostics
+param_sets_top <- top_solutions %>%
+  mutate(
+    ParamID           = paste0("Sol_", row_number()),
+    feedback_strength = beta2 * alpha2,
+    external_strength = beta1 * alpha1,
+    param_strength    = beta1 * beta2 * alpha1 * alpha2,
+    feedback_ratio    = feedback_strength / external_strength,
+    sigmasize         = sigma1 + sigma2,
+    bistability_index = feedback_strength / sigmasize
+  )
+
+# Identify candidates for case-study plots
+bistable_candidate  <- param_sets_top %>% slice_max(bistability_index, n = 2)
+unistable_candidate <- param_sets_top %>% slice_min(bistability_index, n = 2)
+
+# Extract full parameter matrix and specific candidates
+param_sets_all <- param_sets_top %>% select(ParamID, beta1:sigma2)
+param_sets_candidates <- list(
+  "Sol_1"  = unlist(param_sets_top %>% filter(ParamID == "Sol_1")  %>% select(beta1:sigma2)),
+  "Sol_21" = unlist(param_sets_top %>% filter(ParamID == "Sol_21") %>% select(beta1:sigma2))
+)
+
+# ------------------------------------------------------------------------------
+# 2. Define Simulation Settings and Function
[…]

### 2025-05-18 code/intervention_simulation/03_plot_results.R

@@ -1,0 +1,132 @@
+## =============================================================================
+## Script: 03_plot_results.R
+## Plot Intervention Effects: Line Summaries and Density Heatmaps
+## =============================================================================
+
+# -----------------------------------------------------------------------------
+# 1. Load Simulation Outputs
+# -----------------------------------------------------------------------------
+
+# Load line summary results for top 30 parameter sets
+results_DP <- readRDS("Precarity_HELIUS/data/intervention/simulated_intervention_top30.rds")
+
+# Load final-state simulations for Sol_1 and Sol_21
+final_sol1_sol21 <- readRDS("Precarity_HELIUS/data/intervention/final_sol1_sol21.rds")
+
+# -----------------------------------------------------------------------------
+# 2. Plot Line Summary: Mean Final D and P Across Intervention Levels
+# -----------------------------------------------------------------------------
+
+# Prepare data in long format
+results_long <- results_DP %>%
+  pivot_longer(cols = c("Mean_D", "Mean_P"),
+               names_to = "Variable", values_to = "Value")
+
+# Line plot of intervention curves
+plot_lines <- ggplot(results_long, aes(x = Intervention, y = Value,
+                                       group = interaction(ParamID, Variable),
+                                       color = param_strength)) +
+  geom_line(linewidth = 0.5) +
+  facet_grid(. ~ Variable, scales = "free_y") +
+  scale_color_viridis_c(option = "D", name = "Feedback and\nInput Strength") +
+  labs(
+    title = "Final Mean D and P Across Intervention Sizes",
+    subtitle = expression(paste("Top 30 Parameter Sets | Color = ", alpha[1], "×", beta[1], "×", alpha[2], "×", beta[2])),
+    x = "Intervention Size", y = "Final Value"
+  ) +
+  theme_minimal(base_size = 14, base_family = "Palatino") +
+  theme(
+    strip.text       = element_text(face = "bold", size = 14),
+    axis.line        = element_line(color = "black", linewidth = 0.3),
+    legend.position  = "bottom",
+    legend.key.width = unit(1, "null"),
+    panel.grid       = element_blank()
+  )
+
+# -----------------------------------------------------------------------------
[…]

### 2025-05-18 code/utils/libraries.R

@@ -9,6 +9,8 @@ library(purrr)       # Functional programming tools
 library(magrittr)    # Pipe operators and functional utilities
 library(stringr)     # String manipulation
 library(glue)        # Dynamic string interpolation
+library(tidyverse)
+library(MASS)
 
 # Data input/output
 library(haven)       # Reading SPSS .sav files
@@ -16,6 +18,10 @@ library(haven)       # Reading SPSS .sav files
 # Visualization
 library(ggplot2)     # Data visualization
 library(qgraph)      # Graph-based visualizations
+library(plotly)
+library(GGally)
+library(viridis)
+library(patchwork)
 
 # Parallel processing
 library(furrr)       # Parallelized mapping functions
@@ -34,4 +40,7 @@ library(graph)       # Graph structure utilities
 # SDE handling
 library(sde)
 library(deSolve)
+
+# Optimization
+library(nsga2R)
 

### 2025-05-18 code/intervention_simulation/00_optimize_sdemodel.R
edit_minor

@@ -90,8 +90,6 @@ multi_obj_fun_parallel <- function(params, N_runs = 1000) {
 # Run NSGA-II
 plan(multisession) # parallel processing
 
-
-plan(multisession)
 result7 <- nsga2R::nsga2R(
   fn = function(params) multi_obj_fun_parallel(params),  # Fix N_runs
   varNo = 6,                   # Number of parameters
@@ -108,6 +106,6 @@ result7 <- nsga2R::nsga2R(
 )
 
 # save result
-# saveRDS(result7, "multioptim_result_6var7optim_500pop_fixedS.rds")
+# saveRDS(result7, "Precarity_HELIUS/data/intervention/multioptim_result_6var7optim_500pop_fixedS.rds")
 
 

### 2025-05-18 code/intervention_simulation/01_analyze_nsga2.R

@@ -62,5 +62,4 @@ paretoplot <- ggplot(df_filtered, aes(x = total_variance_error, y = total_partia
 # ggsave("Precarity_HELIUS/figures/pareto_frontier.pdf", paretoplot, width = 8, height = 6)
 
 # Save top solutions
-top_solutions_out <- top_solutions %>% mutate(ParamID = paste0("Sol_", row_number()))
-# saveRDS(top_solutions_out, "Precarity_HELIUS/data/intervention/top_solutions_nsga2.rds")
+# saveRDS(top_solutions, "Precarity_HELIUS/data/intervention/top_solutions_nsga2.rds")

### 2025-05-18 code/intervention_simulation/02_simulate_intervention.R

@@ -47,7 +47,7 @@ param_sets_candidates <- list(
 # ------------------------------------------------------------------------------
 
 intervention_sizes <- seq(0, 1, by = 0.01)
-set.seed(42)
+set.seed(42) # set random seed
 S_vals <- sample(depsum$S.fin, 300, replace = TRUE)
 
 # Simulates the final D and P values for each S value and intervention strength

### 2025-05-18 code/intervention_simulation/pareto.R

@@ -1,97 +1,0 @@
-## =============================================================================
-## Analyze NSGA-II Results and Plot Pareto Frontier
-## =============================================================================
-# Load libraries
-source("Precarity_HELIUS/code/utils/libraries.R")   
-
-# ---- 1. Load NSGA-II Results ----
-result <- readRDS("Precarity_HELIUS/data/intervention/multioptim_result_6var7optim_500pop_fixedS.rds") 
-
-param_df <- as.data.frame(result$parameters) %>%
-  setNames(c("beta1", "beta2", "alpha1", "alpha2", "sigma1", "sigma2"))
-
-obj_df <- as.data.frame(result$objectives) %>%
-  setNames(c("mean_D_err", "var_D_err", "mean_P_err", "var_P_err", 
-             "pcor_DP_err", "pcor_DS_err", "pcor_PS_err"))
-
-df <- bind_cols(param_df, obj_df) %>%
-  mutate(total_error = rowSums(across(ends_with("_err"))))
-
-# ---- 2. Visualize Parameter Distributions ----
-param_long <- param_df %>%
-  pivot_longer(everything(), names_to = "Parameter", values_to = "Value")
-
-ggplot(param_long, aes(x = Value, fill = Parameter)) +
-  geom_histogram(bins = 30, alpha = 0.6, color = "black") +
-  facet_wrap(~Parameter, scales = "free") +
-  theme_minimal() +
-  labs(title = "Distribution of Parameter Values (Final Population)")
-
-# ---- 3. Identify Top 5% Solutions ----
-threshold <- quantile(df$total_error, 0.05)
-df <- df %>% mutate(Top = total_error <= threshold)
-
-top_df <- df %>% filter(Top) %>% arrange(total_error)
-
-# Visualize Top Parameter Distributions
-top_df %>%
-  pivot_longer(cols = starts_with(c("beta", "alpha", "sigma")), 
-               names_to = "Parameter", values_to = "Value") %>%
-  ggplot(aes(x = Value, fill = Parameter)) +
-  geom_histogram(bins = 30, alpha = 0.7, color = "black") +
-  facet_wrap(~Parameter, scales = "free") +
-  theme_minimal() +
-  labs(title = "Distribution of Parameters (Top Solutions)")
-
-# ---- 4. Pareto Frontier Plot ----
-# Filter reasonable range for clarity
-df_filtered <- df %>% filter(
[…]

### 2025-05-18 draft_v1/draft_v1.qmd
update_draft

@@
 
-Parameter sets with stronger interactions exhibit more pronounced declines in both depression and precarity in response to intervention.
-At full intervention—modeled as a complete shift of external stress ($S$) toward its observed minimum—the mean depression level decreases from -0.113 to -0.963, and precarity from -0.100 to -0.958.
-These changes correspond to reductions of approximately 90.8\% relative to their initial magnitudes, indicating that targeted intervention can shift the system toward a substantially lower-risk equilibrium.
+Parameter sets with stronger interaction dynamics exhibited more pronounced reductions in both depression and precariousness following intervention.
+At full intervention—modeled as shifting the external stressor ($S$) to its observed minimum—the reduction in depression ranged from [0.68*SD* – 0.97*SD*], and for social precariousness from [0.77*SD* – 0.94*SD*], corresponding to the lowest to highest values of parameter strength.[^3] These results suggest that systems with stronger internal dynamics are more responsive to external change, enabling targeted interventions to more effectively shift the system toward a lower-risk equilibrium.
+
+[^3]: Standard deviations (*SD*) are based on the original scale of each variable: $\textit{SD}_D = 1.0$ for depression ($D$) and $\textit{SD}_P = 0.9$ for precariousness ($P$).
+Parameter strength is defined as the product $\beta_1 \alpha_1 \beta_2 \alpha_2$, reflecting the combined influence of internal feedback and external input sensitivity.
 
@@
 
-[^1]: This resembles a classic *pitchfork bifurcation* in discrete dynamical systems: when the gain $k$ in a scalar map like $x_{t+1} = \tanh(k x_t)$ exceeds a critical value, the system shifts from a single fixed point to two stable attractors and one unstable one.
-In our model, high values of $\alpha_2 \times \beta_2$ similarly enable the emergence of multiple attractor states.
+[^1]: This is similar to a well-known pattern in dynamical systems called a *pitchfork bifurcation*.
+In simple terms, when the system’s internal feedback becomes strong enough—for example, when $k$ in the equation $x_{t+1} = \tanh(k x_t)$ passes a certain threshold—the system changes from having one stable state to having two possible stable states (and one unstable one in between).
+In our model, this kind of shift happens when the feedback loop between depression and precariousness—captured by the product $\alpha_2 \times \beta_2$—becomes strong.
+It allows the system to settle into different long-term states, depending on initial conditions.
 

### 2025-05-18 draft_v1/draft_v1.tex
update_draft

@@
 
-\date{2025-05-14}
+\date{2025-05-18}
 \begin{document}
@@
 
-Parameter sets with stronger interactions exhibit more pronounced
-declines in both depression and precarity in response to intervention.
-At full intervention---modeled as a complete shift of external stress
-(\(S\)) toward its observed minimum---the mean depression level
-decreases from -0.113 to -0.963, and precarity from -0.100 to -0.958.
-These changes correspond to reductions of approximately 90.8\% relative
-to their initial magnitudes, indicating that targeted intervention can
-shift the system toward a substantially lower-risk equilibrium.
+Parameter sets with stronger interaction dynamics exhibited more
+pronounced reductions in both depression and precariousness following
+intervention. At full intervention---modeled as shifting the external
+stressor (\(S\)) to its observed minimum---the reduction in depression
+ranged from {[}0.68\emph{SD} -- 0.97\emph{SD}{]}, and for social
+precariousness from {[}0.77\emph{SD} -- 0.94\emph{SD}{]}, corresponding
+to the lowest to highest values of parameter strength.\footnote{Standard
+  deviations (\emph{SD}) are based on the original scale of each
+  variable: \(\textit{SD}_D = 1.0\) for depression (\(D\)) and
+  \(\textit{SD}_P = 0.9\) for precariousness (\(P\)). Parameter strength
+  is defined as the product \(\beta_1 \alpha_1 \beta_2 \alpha_2\),
+  reflecting the combined influence of internal feedback and external
+  input sensitivity.} These results suggest that systems with stronger
+internal dynamics are more responsive to external change, enabling
+targeted interventions to more effectively shift the system toward a
+lower-risk equilibrium.
 
@@
 settles into a single equilibrium---yielding a unistable
-configuration.\footnote{This resembles a classic \emph{pitchfork
-  bifurcation} in discrete dynamical systems: when the gain \(k\) in a
-  scalar map like \(x_{t+1} = \tanh(k x_t)\) exceeds a critical value,
-  the system shifts from a single fixed point to two stable attractors
-  and one unstable one. In our model, high values of
-  \(\alpha_2 \times \beta_2\) similarly enable the emergence of multiple
-  attractor states.}
+configuration.\footnote{This is similar to a well-known pattern in
+  dynamical systems called a \emph{pitchfork bifurcation}. In simple
+  terms, when the system's internal feedback becomes strong enough---for
+  example, when \(k\) in the equation \(x_{t+1} = \tanh(k x_t)\) passes
+  a certain threshold---the system changes from having one stable state
+  to having two possible stable states (and one unstable one in
+  between). In our model, this kind of shift happens when the feedback
+  loop between depression and precariousness---captured by the product
+  \(\alpha_2 \times \beta_2\)---becomes strong. It allows the system to
+  settle into different long-term states, depending on initial
+  conditions.}
 

### 2025-05-20 draft_v1/draft_v1.qmd
draft update

@@
     include-in-header: header.tex
+    crossref:
+      appendix-title: "App."
+      appendix-delim: ":"
 date: today
@@
   \noindent This study investigates the dynamic relationship between precarious life conditions and depression by combining constraint-based causal discovery with computational modeling.
-Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between financial stress, domain-specific precariousness indicators, and depressive symptoms.
+Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between domain-specific precariousness indicators, and depressive symptoms.
 Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
@@
 To answer this, we combine two approaches.
-First, we use cycle-capable causal discovery algorithms to infer directional relationships among financial stress, domain-specific precariousness factors, and depressive symptoms.
-By analyzing both aggregate scores (e.g., PHQ-9 total, composite precarity index) and disaggregated variables (individual symptoms and life-domain indicators), we identify both broad associations and fine-grained causal pathways.
+First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors, and depressive symptoms.
+By examining both aggregated measures (depression sum score and a composite precarity index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
 Second, we translate these insights into a computational dynamical model that simulates how depression and precarity co-evolve under varying levels of external stress.
@@
 To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
-To ensure a robust representation of precariousness in our causal discovery models, we conducted various exploratory analyses to identify consistent and meaningful data-driven factor structures.
-Based on these analyses, we identified five precariousness factors, including two related to recent stressors, each comprising multiple variables as outlined below.
-DFurther details on the HELIUS study and factor extraction are provided in the [Appendix](#sec-appendix).
-
--   Employment precariousness: `emp_stat`, `work_sit`.
--   Social precariousness: `soc_freq`, `soc_adq`.
--   Housing precariousness: `nb_safe`, `nb_res`, `nb_rent`, `cul_rec`.
--   Recent relational stressors: `frd_brk12`, `conf12`.
--   Recent financial stressors: `fincri12`, `inc_diff`.
-
-After preprocessing, the HELIUS dataset comprises 21,628 samples.
-Along with the five precariousness factors, we also compute an overall precarity score as the combined value of these factors.
-Additionally, PHQ-9 scores are included to represent depression, both as a total sum score and as individual symptom scores.
-In the subsequent causal discovery analysis, we examine the relationship between depression and precarity using both aggregated sum scores and their individual-level representations.
-Refer to @fig-dist for the overall distributions of the variables used in the analysis.
+To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures.
+These analyses led to the identification of five distinct precariousness factors—three domain-based and two reflecting recent stressors—each composed of multiple observed variables, as outlined below.
+Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).
+
+-   Employment precariousness: `emp_stat` (current employment status), `work_sit` (nature of work situation).
+-   Social precariousness: `soc_freq` (frequency of social contact), `soc_adq` (perceived adequacy of social support).
+-   Housing precariousness: `nb_safe` (neighborhood safety), `nb_res` (access to neighborhood resources), `nb_rent` ( proportion of rental housing), `cul_rec` (availability of cultural resources).
+-   Recent relational stressors: `frd_brk12` (experience of a friendship breakup), `conf12` (recent interpersonal or relational conflict).
+-   Recent financial stressors: `fincri12` (experience of a financial crisis), `inc_diff` ( difficulty managing household income).
+
+After preprocessing, the HELIUS dataset includes 21,628 samples.
+In addition to the five identified precariousness factors, we computed an overall precarity score by aggregating their values.
+Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores.
+In the subsequent causal discovery analysis, we explore the relationship between depression and precarity using both aggregated scores and their disaggregated components.
+See @fig-dist for the overall distributions of the variables used in the analysis.
 
@@
 
-To uncover directional relationships between precariousness and depression, we used constraint-based causal discovery algorithms that allow for feedback loops and latent confounding.
-Specifically, we applied the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019].
-For reference, we also included the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].
-
-A key challenge in applying these algorithms to the HELIUS dataset is that many variables are non-Gaussian and their relationships are likely nonlinear.
-To address this, we supplemented the commonly used Gaussian conditional independence (CI) test—based on partial correlations—with a non-parametric alternative: the *Randomized Conditional Correlation Test* (RCoT).
-RCoT approximates kernel-based CI testing using random Fourier features, significantly reducing computational demands while maintaining sensitivity to nonlinear dependencies [@strobl2019approximate; @zhang2012kernel].
-
-We examined the causal structure using three complementary approaches:
+To uncover directional relationships between precariousness and depression, we use constraint-based causal discovery algorithms that allow for feedback loops and latent confounding.
+Specifically, we apply the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019].
+For reference, we also include the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].
+While our full analysis includes results from three causal discovery algorithms—FCI, CCI, and PC—we focus on FCI in the main text, as it offers a strong balance between flexibility and theoretical soundness.
+FCI accommodates latent confounding and potential feedback, making it well-suited for the complexity of our data [@mooij2020constraint].
+The PC algorithm is included for reference but relies on more restrictive assumptions, such as acyclicity and no unmeasured confounding.
+CCI, while capable of modeling cycles explicitly, comes with certain theoretical limitations.
+Results from both PC and CCI are provided in the [Appendix](#sec-appendix) ([@sec-cci; @sec-pc]) for completeness and comparison.
+For readers interested in detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, please refer to the [Appendix](#sec-appendix) ([@sec-causalprimer]).
+
+A key challenge in applying these algorithms to the HELIUS dataset is that many variables are non-Gaussian and likely exhibit nonlinear relationships.
+While kernel-based conditional independence tests (such as KCIT) are well-suited for capturing such complex dependencies, they are computationally intensive—scaling quadratically with sample size due to the need to invert large kernel matrices—making them impractical for large datasets like HELIUS [@rahimi2007random].
+To address this, we use the Randomized Conditional Correlation Test (RCoT), a nonparametric alternative that approximates kernel methods using random Fourier features.
+This reduces computational complexity from quadratic to linear in sample size, substantially lowering runtime while preserving sensitivity to nonlinear relationships [@strobl2019approximate; @zhang2012kernel].
+For further details, see @sec-rcot.
+
+We examine the causal structure using three complementary approaches:
 
@@
 
-To ensure robust estimation and guard against sensitivity to parameter settings, we implemented a systematic grid of analysis conditions:
-
-* Significance levels: $\alpha$ = 0.01 and 0.05
-* Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8
-* CI tests: Gaussian and RCoT
-* Algorithms: FCI, CCI, and PC
-
-Each combination was bootstrapped 100 times for sum-score level analyses and 30 times for symptom-level analyses, yielding a total of 1,600 graphs per algorithm for the full analysis.
+To robustly estimate the causal structure between depression and precariousness variables, we implement a comprehensive bootstrapped causal discovery procedure.
+This procedure systematically varies key analysis parameters across a grid of configurations, enabling us to assess the stability of inferred relationships under different settings.
+Specifically, we vary four main components:
+
+* Significance levels for conditional independence tests: $\alpha$ = 0.01 and 0.05.
+* Stability thresholds: 0.5, 0.6, 0.7, and 0.8, which determine how frequently an edge must appear across bootstraps to be retained.
+* Conditional independence (CI) tests: Gaussian CI test (partial correlation) and RCoT (a nonparametric test).
+* Causal discovery algorithms: FCI, CCI, and PC.
+
+For each unique parameter combination (2 significance levels × 4 thresholds × 2 CI tests = 16 settings), we run 100 bootstrap samples per algorithm for the aggregated analyses (see Step 1 in @fig-workflow).
+For symptom-level analyses, which are more computationally intensive, we use 30 bootstrap samples per setting.[^5]
+
+[^5]: For symptom-level analyses, we further improve computational efficiency by fixing the skeleton structure—the undirected network of potential connections—using a consensus graph derived from all three algorithms (with $\alpha = 0.01$ and the RCoT test).
+This allows subsequent analyses to focus exclusively on estimating edge directions, avoiding the repeated and costly computation of skeletons, and thus substantially reducing runtime.
+
+To summarize these results, we apply a two-level aggregation strategy:
+
+1.
+**Within each condition** (e.g., fixed $\alpha$, threshold, CI test, and algorithm), we identify the most frequently occurring edge type (e.g., arrowhead, circle, none) across bootstraps for each pair of variables (Step 2 in @fig-workflow).
+2.
+**Across conditions**, we then determine the most dominant edge type based on majority voting.
+In cases where no edge type is clearly dominant (i.e., a tie), we mark the edge as uncertain, represented with a dashed line in the final graph (Step 3 in @fig-workflow).
+
+This summarization process is repeated separately for each causal discovery algorithm (FCI, CCI, PC), yielding a single robust graph per algorithm.
+These final graphs represent only the most stable and consistent relationships across a wide range of plausible parameterizations, enhancing both reliability and interpretability.
+To support transparency, each final graph is accompanied by a matrix summarizing the relative frequency of each edge type for every variable pair—enabling readers to assess the degree of uncertainty or agreement underlying each connection.
+
+<!-- To ensure robust estimation and guard against sensitivity to parameter settings, we implemented a systematic grid of analysis conditions: -->
+
+<!-- * Significance levels: $\alpha$ = 0.01 and 0.05 -->
+<!-- * Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8 -->
+<!-- * CI tests: Gaussian and RCoT -->
+<!-- * Algorithms: FCI, CCI, and PC -->
+
+<!-- Each combination was bootstrapped 100 times for sum-score level analyses and 30 times for symptom-level analyses, yielding a total of 1,600 graphs per algorithm for the full analysis.
 For symptom networks, we fixed the skeleton structure using a consensus graph estimated across PC, FCI, and CCI ($\alpha$ = 0.01, RCoT).
 This constraint improves computational efficiency without limiting the estimation of directional edges between symptoms and precariousness variables.
-
-To summarize the results, we adopted a two-level aggregation strategy.
+-->
+
+<!-- To summarize the results, we adopted a two-level aggregation strategy.
 First, within each analysis condition, we recorded the most frequently observed edge-endpoint type (e.g., directed, undirected, uncertain) from the bootstraps.
@@
 For transparency, we accompany each graph with a table showing the relative frequencies of each edge symbol.
-
-This procedure yields a final graph for each algorithm and analysis type—offering both robustness and interpretability. See @fig-workflow for an overview of the causal discovery pipeline.
-
-For detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, see the [Appendix](#sec-appendix).
+-->
+
+<!-- This procedure yields a final graph for each algorithm and analysis type—offering both robustness and interpretability. See @fig-workflow for an overview of the causal discovery pipeline. -->
+<!-- For detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, see the [Appendix](#sec-appendix).
+-->
 
@@
 #| label: fig-workflow
-#| fig-cap: "Analysis workflow applied across all three algorithms."
+#| fig-cap: "Causal discovery workflow applied across all three algorithms (FCI, CCI, and PC).
+*The process involves three main steps: (1) bootstrapping the dataset to create multiple resampled datasets; (2) estimating causal graphs under varying parameter configurations; and (3) summarizing results by identifying the most frequent edge type across bootstraps and settings to construct a final stable graph.*"
 #| fig-align: center
-#| out-width: 80%
-
-knitr::include_graphics("img/simsetup2.png")
+#| out-width: 90%
+
+knitr::include_graphics("img/simsetup.pdf")
 
@@
 #| label: fig-concept
-#| fig-cap: "Conceptual high-level causal structure of precariousness and depression system."
+#| fig-cap: "High-level conceptual model of causal structure linking social precariousness (P.soc) and depression (PHQsum), influenced by external financial stress (S.fin)."
 #| fig-align: center
-#| out-width: 35%
+#| out-width: 40%
 
@@
 
-To identify realistic parameter settings for the simulation model, we calibrated six free parameters—$\alpha_1$, $\alpha_2$, $\beta_1$, $\beta_2$, and noise amplitudes $\sigma_1$, $\sigma_2$—using the Non-dominated Sorting Genetic Algorithm II (NSGA-II) [@deb2002fast], a multi-objective evolutionary optimization method well-suited for noisy and high-dimensional search spaces.
-The goal was to match simulated system behavior to empirical summary statistics derived from the HELIUS dataset, including the means and variances of depression ($D$) and social precariousness ($P$), as well as their partial correlations after adjusting for financial stress ($S$), specifically $\rho_{DP.S}$, $\rho_{DS.P}$, and $\rho_{PS.D}$.
-Given the stochastic nature of the model, each candidate parameter set was evaluated by simulating 1,000 independent trajectories, each representing an individual with a fixed $S$ value sampled from the observed distribution.
-Final-state values were extracted and aggregated to compute population-level summary statistics.
-Fitness was assessed as the absolute deviation from the empirical targets, aggregated across all objectives.
-NSGA-II was implemented via the nsga2R package [@nsga2R], with a population of 300 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2).
+To identify realistic parameter settings for the simulation model, we calibrate six free parameters—$\alpha_1$, $\alpha_2$, $\beta_1$, $\beta_2$, and noise amplitudes $\sigma_1$, $\sigma_2$—using the *Non-dominated Sorting Genetic Algorithm II* (NSGA-II) [@deb2002fast], a multi-objective evolutionary optimization method well-suited for noisy and high-dimensional search spaces.
+The goal is to match simulated system behavior to empirical summary statistics derived from the HELIUS dataset.
+Specifically, we aim to match the observed means and variances of depression ($D$) and social precariousness ($P$), along with all three partial correlations among $D$, $P$, and financial stress ($S$): $\rho_{DP.S}$, $\rho_{DS.P}$, and $\rho_{PS.D}$.
+
+Given the stochastic nature of the model, each candidate parameter set is evaluated by simulating 1,000 virtual individuals, each assigned a fixed $S$ value sampled from the empirical distribution.
+From these simulations, we compute population-level summary statistics based on the final states.
+The fitness of each parameter set is measured by how closely its summary statistics matched the empirical data, using the total absolute deviation across all objectives.
+
+NSGA-II is implemented via the `nsga2R` package [@nsga2R], with a population of 500 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2).
 This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives.
@@
 
-For each of the top 30 calibrated parameter sets, we simulate the system’s evolution across 300 individuals under a continuum of intervention levels.
-Each simulation run generates final values for $D$ and $P$, which are then used to assess how internal model parameters influence the system’s behavior under changing external conditions.
+For each of the top 30 parameter sets identified through calibration, we run simulations on 300 virtual individuals across this full range of intervention levels.
+In each simulation, we observe how depression ($D$) and precariousness ($P$) evolve over time, capturing their final values after the system stabilizes.
+This setup allows us to evaluate how different internal model dynamics (e.g., feedback strength) influence the system’s response to external change.
 
@@
 
-The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (PHQsum).
-Employment precarity (P.emp) and social precarity (P.soc) are not identified as causes of depression, whereas financial stress (S.fin) appears to play a causal role.
-While P.emp and P.soc are generally not recognized as causes of other precariousness factors, S.fin emerges as a potential cause, as indicated by its circle edge endpoint.
-Supporting this, the proportion matrix plot in panel (b) shows that S.fin has some probability of causally influencing either P.emp or P.soc.
+The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (*PHQsum*).
+Employment precarity (*P.emp*) and social precarity (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role.
+While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint.
+Supporting this, the proportion matrix plot in panel (b) shows that *S.fin* has some probability of causally influencing either *P.emp* or *P.soc.*
 
@@
 
-Relational stress (S.rel) plays a more nuanced role, interacting with depression and S.fin  through a latent confounder.
-Its relationship with P.soc remains less certain, as the proportion table suggests a fair probability that this connection is absent, uncertain, or bidirectional.
-Meanwhile, housing precarity (P.hou) is not directly causally related to depression but is linked to P.emp and S.fin.
-While P.emp and S.fin are identified as non-causes of P.hou, it remains unclear whether P.hou causally influences P.emp or if their relationship is mediated by an unobserved confounder.
-
-The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify S.fin as a cause of depression.
+Relational stress (*S.rel*) plays a more nuanced role, interacting with depression and *S.fin*  through a latent confounder.
+Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 indicates a considerable probability that this relationship is either absent, ambiguous, or potentially bidirectional.
+Meanwhile, housing precarity (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*.
+While *P.emp* and *S.fin* are identified as non-causes of *P.hou*, it remains unclear whether *P.hou* causally influences *P.emp* or if their relationship is mediated by an unobserved confounder.
+
+The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify *S.fin* as a cause of depression.
 Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI.
@@
 Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms.
-Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph highlights heterogeneity in how precariousness factors influence individual depressive symptoms: con (concentration), slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal), anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.
-
-The symptom-level graph in @fig-sym panel (a) reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges.
+Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.
+
+The symptom-level graph in @fig-sym-1 reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges.
 As before, dashed edges denote ties, with specific ties marked by bold horizontal bars.
-In @fig-sym panel (b), the proportion matrix provides more insight into these directionalities.
+In @fig-sym-2, the proportion matrix provides more insight into these directionalities.
 Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type.
-However, blue regions, which correspond to circle endpoints, are particularly common for anh, slp, ene, and sui, reflecting uncertainty in the causal direction among these symptoms.
-Notably, the only arrowtail connection appears between dep and anh, suggesting that anhedonia (anh) is a potential cause of depressed mood (dep).
+However, blue regions, which correspond to circle endpoints, are particularly common for *anh*, *slp*, *ene*, and *sui*, reflecting uncertainty in the causal direction among these symptoms.
+Notably, the only arrowtail connection appears between *dep* and *anh*, suggesting that anhedonia is a potential cause of depressed mood.
 
@@
 This suggests that these causal links may be less stable across different conditions.
-Upon closer examination, most ties arise due to discrepancies between the Gaussian CI test and the RCoT test—with Gaussian CI favoring arrowheads and RCoT more often predicting the absence of an edge.
+Closer examination reveals that most ties arise due to discrepancies between the Gaussian CI test and the RCoT test—with Gaussian CI favoring arrowheads and RCoT more often predicting the absence of an edge.
 
 This discrepancy likely arises from a combination of factors. 
-First, as the analysis shifts from aggregated to disaggregated variables, the number of conditional independence (CI) tests increases, leading to a loss of statistical power in high-dimensional causal discovery.
-With more variables being conditioned on, the data becomes sparser, making it harder to detect weaker dependencies.
-Additionally, over-conditioning—controlling for too many variables—can artificially remove statistical associations, leading to false negatives.
-Beyond this general statistical power issue, the fundamental differences in how these CI tests operate contribute to the inconsistency.
-RCoT, as a nonparametric test, does not assume linearity and can capture both linear and nonlinear dependencies.
-However, it is more conservative and requires larger sample sizes, making it more prone to false negatives in high-dimensional settings.
-In contrast, partial correlation (Gaussian CI test) assumes linear Gaussian relationships, making it more permissive and sometimes detecting weak dependencies that may not be statistically meaningful.
-As a result, when Gaussian CI detects an edge while RCoT does not, it could indicate that the relationship is weak, strictly linear, or requires more data for reliable nonparametric detection.
-This explains why many ties in the proportion matrix show a 50-50 split between symbols—where half of the conditions with RCoT suggest the absence of an edge, while the other half using Gaussian CI test favor an arrowhead.
-This systematic divergence highlights the challenges of causal discovery in high-dimensional settings, where varying statistical assumptions can lead to differing conclusions about the presence and direction of causal relationships.
-
-Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of financial stress (S.fin), which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality.
-One exception is the stronger evidence suggesting that S.fin may cause changes in app (appetite), while its relationships with other symptoms, such as ene, mot, and dep, remain ambiguous.
-Similarly, social precarity (P.soc) exhibits numerous connections with depressive symptoms, with most edges pointing toward P.soc rather than outward from it.
-This suggests that depressive symptoms, particularly dep and glt, may contribute to worsening social precarity rather than the other way around.
-Additionally, anh also shows some probability of causally influencing P.soc, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.
+As the analysis moves from aggregated to disaggregated variables, the number of CI tests increases substantially, leading to a loss of statistical power.
+With more variables being conditioned on, the effective sample size per test diminishes, making it harder to detect subtle dependencies.
+Over-conditioning—controlling for too many variables—can also obscure real relationships, resulting in false negatives.
+Beyond these statistical power issues, the underlying assumptions of the two CI tests contribute to their divergence.
+RCoT is a nonparametric test that captures both linear and nonlinear dependencies without assuming a specific distributional form.
+However, it is more conservative and requires larger sample sizes, especially in high-dimensional settings.
+In contrast, the Gaussian CI test is based on partial correlations and assumes linear Gaussian relationships, making it more permissive and more likely to detect weak or spurious associations.
+As a result, it is common to see edges where Gaussian CI suggests a directional relationship, while RCoT indicates no edge.
+In the proportion matrix, this appears as a 50–50 tie between absence and arrowhead symbols.
+This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related.
+
+Despite these differences, some consistent patterns emerge across both CI tests, particularly in the case of *S.fin*, which appears to be connected to nearly all other variables—though many of these connections are marked by ties, reflecting uncertainty in directionality.
+One exception is the stronger evidence suggesting that *S.fin* may cause changes in *app*, while its relationships with other symptoms, such as *ene*, *mot*, and *dep*, remain ambiguous.
+Similarly, *P.soc* exhibits numerous connections with depressive symptoms, with most edges pointing toward *P.soc* rather than outward from it.
+This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precarity rather than the other way around.
+Additionally, *anh* also shows some probability of causally influencing *P.soc*, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.
 
 Other precariousness factors exhibit more uncertain but still notable relationships.
-Relational stress (S.rel) shows weak but existing connections with slp, glt, and app, though directionality remains unclear in many cases.
-Employment precarity (P.emp) also connects to symptoms like slp, mot, and con, but, like S.rel, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
-In line with the aggregated graph in @fig-sum, housing precarity (P.hou) does not appear to have any direct relationship with depression symptoms but consistently shows associations with employment precarity (P.emp) and financial stress (S.fin).
-The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with financial stress (S.fin) exhibiting the strongest tendency to influence other precariousness factors.
-
-Among depressive symptoms, dep, glt, and slp appear to be the most connected to precariousness factors, while P.soc has the most connections with symptoms, predominantly as a recipient rather than a driver of influence.
-Within the symptom network, slp, sui, and anh emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms.
-On the other hand, con, mot, dep, and glt, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
-Considering both symptom-precariousness connections and symptom-level dynamics, sleep disturbance (slp) emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network.
+*S.rel* shows weak but existing connections with *slp*, *glt*, and *app*, though directionality remains unclear in many cases.
+*P.emp* also connects to symptoms like *slp*, *mot*, and *con*, but, like *S.rel*, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
+In line with the aggregated graph in @fig-sum, *P.hou* does not appear to have any direct relationship with depression symptoms but consistently shows associations with *P.emp* and *S.fin*.
+The causal relationships among precariousness factors remain largely unchanged from the aggregated analysis, with *S.fin* exhibiting the strongest tendency to influence other precariousness factors.
+
+Among depressive symptoms, *dep*, *glt*, and *slp* appear to be the most connected to precariousness factors, while *P.soc* has the most connections with symptoms, predominantly as a recipient rather than a driver of influence.
+Within the symptom network, *slp*, *sui*, and *anh* emerge as causally influential symptoms, as they exhibit more outgoing arrows compared to other symptoms.
+On the other hand, *con*, *mot*, *dep*, and *glt*, despite having high connectivity, predominantly receive incoming arrows, indicating they are more likely effects rather than causes.
+Considering both symptom-precariousness connections and symptom-level dynamics, *slp* emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network.
 This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precarity.
-On the other hand, social precarity (P.soc) serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.
+On the other hand, *P.soc* serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.
 
@@
 
-Lastly, we examine the relationships between individual symptoms and overall precarity, an aggregated measure represented by the sum of five precariousness factors.
+Lastly, we examine the relationships between individual symptoms and overall precariousness, an aggregated measure represented by the sum of five precariousness factors.
 This analysis provides a complementary perspective, capturing broader patterns that may not be evident in the disaggregated symptom-precariousness analysis.
-By aggregating precariousness factors into a single score, this approach may capture distributed relationships across different precarity factors that were previously overlooked in the disaggregated analysis.
-
-Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum panel (a)) shows more certainty in causal directions, with only a few edges resulting in ties, all of which involve the overall precarity factor.
-This is even more evident in the proportion matrix (@fig-presum panel (b)), where symptom-to-symptom interactions almost fully converge to a single edge type.
+By aggregating precariousness factors into a single score, this approach may capture distributed relationships across different precariousness factors that were previously overlooked in the disaggregated analysis.
+
+Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum-1) shows more certainty in causal directions, with only a few edges resulting in ties, all of which involve the overall precariousness factor.
+This is even more evident in the proportion matrix (@fig-presum-2), where symptom-to-symptom interactions almost fully converge to a single edge type.
 Additionally, all symptom interactions are connected either through bidirectional edges or a combination of an arrowhead and a circle, suggesting a strong presence of latent confounders.
 
-Examining the symptom-precarity connections, we observe a stronger overall trend of depressive symptoms influencing precarity, rather than the other way around.
-Specifically, anh, dep, and slp show a high probability of causing precarity, whereas app, glt, and mot exhibit more uncertainty in directionality, often resulting in circle endpoints.
-This pattern suggests that when precariousness factors are aggregated, the dominant causal flow is from depressive symptoms to precarity, rather than precarity driving depression.
-Particularly, dep emerges as the strongest predictor of precarity, exhibiting the highest proportion of arrowtails, consistent with findings from the disaggregated analysis.
+Examining the symptom-precariousness connections, we observe a stronger overall trend of depressive symptoms influencing precariousness, rather than the other way around.
+Specifically, *anh*, *dep*, and *slp* show a high probability of causing precariousness, whereas *app*, *glt*, and *mot* exhibit more uncertainty in directionality, often resulting in circle endpoints.
+This pattern suggests that when precariousness factors are aggregated, the dominant causal flow is from depressive symptoms to precariousness, rather than precariousness driving depression.
+Particularly, *dep* emerges as the strongest predictor of precariousness, exhibiting the highest proportion of arrowtails, consistent with findings from the disaggregated analysis.
 
@@
 The CCI results reveal that most symptoms are interconnected through bidirectional edges, suggesting a strong presence of latent confounders.
-However, dep appears to be a distinct exception, as it is predominantly caused by nearly all other symptoms, with one exception sui, which does not contribute to dep.
-Similarly, con is influenced by multiple symptoms, albeit with weaker support compared to dep.
-Most interestingly, the CCI results highlight a distinct feedback loop between dep and overall precarity, represented by a tail-tail edge (\textemdash \).
+However, *dep* appears to be a distinct exception, as it is predominantly caused by nearly all other symptoms, with one exception *sui*, which does not contribute to *dep*.
+Similarly, *con* is influenced by multiple symptoms, albeit with weaker support compared to *dep*.
+Most interestingly, the CCI results highlight a distinct feedback loop between *dep* and overall precariousness, represented by a tail-tail edge (\textemdash \).
 This suggests a possible reinforcing cycle between depression and overall precariousness, where depression not only arises from precarious conditions but also contributes to their persistence, creating a self-sustaining dynamic.
@@
 The results highlight the significant role of latent confounders in symptom-to-symptom interactions.
-While certain stressor-related precariousness factors, particularly S.fin and S.rel, appear to causally influence depressive symptoms, other precariousness factors are more likely driven by depression rather than acting as primary causes, suggesting that precariousness may function more as a consequence of depression than its source.
-Some symptoms, particularly glt, slp, and dep, show greater sensitivity to precarious conditions, while also playing a central role in symptom dynamics.
-Among them, slp and glt emerge as influential symptoms within the depression network, whereas con and dep, despite their high connectivity, behave more as outcomes rather than primary drivers.
+While certain stressor-related precariousness factors, particularly *S.fin* and *S.rel*, appear to causally influence depressive symptoms, other precariousness factors are more likely driven by depression rather than acting as primary causes, suggesting that precariousness may function more as a consequence of depression than its source.
+Some symptoms, particularly *glt*, *slp*, and *dep*, show greater sensitivity to precarious conditions, while also playing a central role in symptom dynamics.
+Among them, *slp* and *glt* emerge as influential symptoms within the depression network, whereas *con* and *dep*, despite their high connectivity, behave more as outcomes rather than primary drivers.
 Together, these findings suggest that depression—especially symptoms like sleep disturbances and guilt—may contribute to sustaining precarious conditions, possibly forming a reciprocal cycle between depression and precariousness.
@@
 [^3]: Standard deviations (*SD*) are based on the original scale of each variable: $\textit{SD}_D = 1.0$ for depression ($D$) and $\textit{SD}_P = 0.9$ for precariousness ($P$).
-Parameter strength is defined as the product $\beta_1 \alpha_1 \beta_2 \alpha_2$, reflecting the combined influence of internal feedback and external input sensitivity.
-
-To better understand these dynamics, we focused on two illustrative parameter sets: one with strong feedback/input coupling and low noise (a "bistable" candidate), and one with weaker coupling and higher noise (a "unistable" candidate).
+Parameter strength is defined as the product $\alpha_1 \times \beta_1 \times \alpha_2 \times \beta_2$, reflecting the combined influence of internal feedback and external input sensitivity.
+
+To better understand these dynamics, we focused on two illustrative parameter sets: one with strong feedback/input coupling and low noise (a *bistable* candidate), and one with weaker coupling and higher noise (a *unistable* candidate).
 In nonlinear systems, bistability refers to the existence of two distinct stable equilibria—meaning the system’s long-term outcome depends on its starting point or perturbations.
@@
 This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precarity interact—and how their internal structure modulates responsiveness to intervention.
-Using cross-sectional data from the HELIUS study, we applied cycle-capable discovery algorithms to estimate plausible directional relationships among financial stress, multiple domains of precariousness, and depressive symptoms.
+Using cross-sectional data from the HELIUS study, we applied cycle-capable discovery algorithms to estimate plausible directional relationships among multiple domains of precariousness factors, and depressive symptoms.
 By analyzing both aggregate constructs and individual variables, we aimed to capture patterns that may be masked in either extreme of abstraction (i.e., purely aggregate or disaggregated analyses).
 
-The causal analysis consistently identified financial stress (*S.fin*) as a key plausible driver of depression, appearing across both aggregated and disaggregated models.
-At the symptom level, we found that sleep disturbance (*slp*), depressed mood (*dep*), guilt (*glt*), and anhedonia (*anh*) were especially sensitive to external stressors.
-Among these, *dep* emerged as a central hub: not only was it heavily influenced by other symptoms, it also exerted directional influence on social precarity (*P.soc*), suggesting a feedback loop in which worsening mood contributes to social disconnection, which in turn may exacerbate depressive symptoms.
-The CCI algorithm even suggested a possible cyclic relationship between dep and overall precarity, hinting at a structurally self-reinforcing mechanism.
-*Slp* also played a key initiating role.
-It showed directed connections to both social and employment precarity and exhibited many outgoing edges within the symptom network.
-This suggests *slp* may function as an early warning signal or entry point into the depression–precarity cycle, making it a promising target for preventive intervention.
-In contrast, *P.soc* more often appeared as an effect rather than a cause—receiving arrows from symptoms like *dep* and *glt*, but rarely sending them.
-This asymmetry implies that social precarity may often reflect downstream consequences of worsening mental health rather than being its root cause.
+The causal analysis consistently identified financial stress as a key plausible driver of depression, appearing across both aggregated and disaggregated models.
+At the symptom level, we found that sleep disturbance, depressed mood, guilt, and anhedonia were especially sensitive to external stressors.
+Among these, the symptom depressed mood emerged as a central hub: not only was it heavily influenced by other symptoms, it also exerted directional influence on social precariousness, suggesting a feedback loop in which worsening mood contributes to social disconnection, which in turn may exacerbate depressive symptoms.
+The CCI algorithm even suggested a possible cyclic relationship between depressed mood and overall precariousness, hinting at a structurally self-reinforcing mechanism.
+Symptom sleep disturbance also played a key initiating role.
+It showed directed connections to both social and employment precariousness and exhibited many outgoing edges within the symptom network.
+This suggests the symptom sleep disturbance may function as an early warning signal or entry point into the depression–precariousness cycle, making it a promising target for preventive intervention.
+In contrast, Social precariousness more often appeared as an effect rather than a cause—receiving arrows from symptoms like depressed mood and guilt, but rarely sending them.
+This asymmetry implies that social precariousness may often reflect downstream consequences of worsening mental health rather than being its root cause.
 These patterns offer actionable insights for intervention.
-Addressing symptoms like *slp* early may prevent escalation into broader depressive states or worsening precarity.
-Intervening on *dep*, as a central node with potential feedback effects, could also disrupt the self-reinforcing dynamics between mental health and social isolation.
+Addressing symptoms like sleep disturbance early may prevent escalation into broader depressive states or worsening precarity.
+Intervening on depressed mood, as a central node with potential feedback effects, could also disrupt the self-reinforcing dynamics between mental health and social isolation.
 
@@
 While the FCI and CCI algorithms are designed to recover causal structure from cross-sectional data—including feedback and latent confounding—many edge directions in our results remain ambiguous, as indicated by the prevalence of circle endpoints and bidirectional arrows, especially in weakly connected regions of the graph.
-These uncertainties were most evident in the links between individual symptoms and precariousness factors, where edge estimates varied across algorithms and conditional independence (CI) tests.
+These uncertainties were most evident in the links between individual symptoms and precariousness factors, where edge estimates varied across algorithms and CI tests.
 This likely reflects limited statistical signal in high-dimensional settings, where large conditioning sets reduce power and small effect sizes are harder to detect.
@@
 
-Second, our findings showed sensitivity to the choice of conditional independence (CI) test.
+Second, our findings showed sensitivity to the choice of CI test.
 The Gaussian CI test, which assumes linear relationships and Gaussian noise, often produced denser graphs—likely due to its tendency to overestimate weak or noisy linear associations.
@@
 Finally, the structure of the dynamical model introduces its own limitations.
-While the model was designed to reflect key features of the discovered causal structure, it represents a highly simplified system: only two variables—depression and social precarity—modulated by external financial stress.
+While the model was designed to reflect key features of the discovered causal structure, it represents a highly simplified system: only two variables—depression and social precariousness—modulated by external financial stress.
 This abstraction allowed for tractable analysis and visual intuition, particularly around nonlinear behaviors such as bistability, but omits other potentially influential constructs highlighted in the causal graphs, including relational stress, employment conditions and housing conditions.
@@
 
-## HELIUS study
+## HELIUS study {#sec-helius}
 
@@
 
-## Causal Discovery Primer
+## Causal Discovery Primer {#sec-causalprimer}
 
@@
 While CCI's MAAG provides an explicit representation of cyclic relationships, it has certain theoretical limitations, as the MAAG may not always fully preserve d-separation relations from the original graph, $\mathcal{G}$ [@strobl2019].
-Given these considerations, we examine and compare results from both algorithms, placing more weight on the FCI results, which are presented in the main results section, while discussing CCI results where relevant, with detailed findings provided in the [Appendix](#sec-appendix).
+Given these considerations, we examine and compare results from both algorithms, placing more weight on the FCI results, which are presented in the main results section, while discussing CCI results where relevant, with detailed findings provided in @sec-cci.
 A full discussion of causal discovery concepts and algorithmic details is beyond the scope of this paper; however, readers seeking a more in-depth understanding can refer to @park2024discovering for a comprehensive exploration of these methods and their applications.
@@
 While the PC algorithm serves as a useful reference, its strict assumptions — acyclicity and the absence of latent confounders—limit its applicability in more complex scenarios.
-For this reason, our primary focus remains on the results obtained from FCI and CCI, with all PC algorithm results provided in the [Appendix](#sec-appendix) for completeness.
-
-## Precariousness factors by Leonie
+For this reason, our primary focus remains on the results obtained from FCI and CCI, with all PC algorithm results provided in @sec-pc for completeness.
+
+## Precariousness Domains Based on Elsenburg et al. (2025)
 
@@
 
-## Exploratory Factor Analysis (EFA)
+### Exploratory Factor Analysis (EFA)
 
@@
 
-### Factor Loadings (Pattern Matrix)
+#### Factor Loadings (Pattern Matrix)
 
@@
 
-### Variance Explained
+#### Variance Explained
 
@@
 
-### Factor Intercorrelations
+#### Factor Intercorrelations
 
@@
 
-### Model Fit Statistics
+#### Model Fit Statistics
 
@@
 
-### Summary
+#### Summary
 
@@
 
-## PCA
+### PCA
 
@@
 
-### Explained variance (contributions) of variables
+#### Explained variance (contributions) of variables
 
@@
 
-### Cos² Values
+#### Cos² Values
 
@@
 
-## ICA
+### ICA
 
@@
 
-### Dominant Variables per Component:
+#### Dominant Variables per Component:
 
@@
 
-## Hierarchical clustering
-
-### Using Euclidean distance
+### Hierarchical clustering
+
+#### Using Euclidean distance
 
@@
 
-#### Consistent Groupings (Across All or Most Methods)
+##### Consistent Groupings (Across All or Most Methods)
 
@@
 
-#### Inconsistent Groupings (Variability Across Methods)
+##### Inconsistent Groupings (Variability Across Methods)
 
@@
 
-#### Summary
+##### Summary
 
@@
 
-### Using Mutual Information
+#### Using Mutual Information
 
@@
 
-#### Comparison to Euclidean Distance Clustering
+##### Comparison to Euclidean Distance Clustering
 
@@
 
-## Conclusions on Precariousness factors
+### Conclusions on Precariousness factors
 
@@
   ) |>
-  dplyr::select(17:33) |>
+  dplyr::select(17:33, -c(age, ethn)) |>
   as.data.frame()
@@
 
-## Results from PC algorithm
+## Results from PC algorithm {#sec-pc}
 

### 2025-05-20 draft_v1/draft_v1.tex
draft update

@@
 
-\date{2025-05-18}
+\date{2025-05-20}
 \begin{document}
@@
 from the HELIUS study, we applied cycle-capable causal discovery
-algorithms to uncover directional relationships between financial
-stress, domain-specific precariousness indicators, and depressive
-symptoms. Our multi-resolution analysis---spanning both aggregate and
-symptom-level constructs---identified financial stress as a robust
-driver of depression, and highlighted key symptoms, such as sleep
-disturbance and depressed mood, as likely initiators or reinforcers of
-downstream precariousness. To explore how these structural insights
-shape intervention effectiveness, we developed a simplified nonlinear
+algorithms to uncover directional relationships between domain-specific
+precariousness indicators, and depressive symptoms. Our multi-resolution
+analysis---spanning both aggregate and symptom-level
+constructs---identified financial stress as a robust driver of
+depression, and highlighted key symptoms, such as sleep disturbance and
+depressed mood, as likely initiators or reinforcers of downstream
+precariousness. To explore how these structural insights shape
+intervention effectiveness, we developed a simplified nonlinear
 dynamical model simulating the co-evolution of depression and social
@@
 we combine two approaches. First, we use cycle-capable causal discovery
-algorithms to infer directional relationships among financial stress,
-domain-specific precariousness factors, and depressive symptoms. By
-analyzing both aggregate scores (e.g., PHQ-9 total, composite precarity
-index) and disaggregated variables (individual symptoms and life-domain
-indicators), we identify both broad associations and fine-grained causal
-pathways. Second, we translate these insights into a computational
-dynamical model that simulates how depression and precarity co-evolve
-under varying levels of external stress. This model enables us to probe
-how system properties---such as feedback strength and stochastic
-noise---govern responsiveness to intervention. In some configurations,
-the system exhibits bistability, remaining stuck in a high-risk state
-unless external support exceeds a critical threshold. In others, it
-shows unistable behavior, responding more gradually to incremental
-changes. By integrating causal discovery with dynamic simulation, we aim
-to demonstrate how the internal structure of the depression--precarity
+algorithms to infer directional relationships among domain-specific
+precariousness factors, and depressive symptoms. By examining both
+aggregated measures (depression sum score and a composite precarity
+index) and disaggregated variables (individual depressive symptoms and
+specific life-domain indicators), we capture broad patterns of
+relationship as well as fine-grained causal pathways. Second, we
+translate these insights into a computational dynamical model that
+simulates how depression and precarity co-evolve under varying levels of
+external stress. This model enables us to probe how system
+properties---such as feedback strength and stochastic noise---govern
+responsiveness to intervention. In some configurations, the system
+exhibits bistability, remaining stuck in a high-risk state unless
+external support exceeds a critical threshold. In others, it shows
+unistable behavior, responding more gradually to incremental changes. By
+integrating causal discovery with dynamic simulation, we aim to
+demonstrate how the internal structure of the depression--precarity
 system shapes its susceptibility to change---and, by extension, how
@@
 variables. To ensure a robust representation of precariousness in our
-causal discovery models, we conducted various exploratory analyses to
-identify consistent and meaningful data-driven factor structures. Based
-on these analyses, we identified five precariousness factors, including
-two related to recent stressors, each comprising multiple variables as
-outlined below. DFurther details on the HELIUS study and factor
-extraction are provided in the \hyperref[sec-appendix]{Appendix}.
+causal discovery models, we conducted exploratory analyses to identify
+consistent and meaningful data-driven factor structures. These analyses
+led to the identification of five distinct precariousness
+factors---three domain-based and two reflecting recent stressors---each
+composed of multiple observed variables, as outlined below. Additional
+details on the HELIUS study and the factor extraction process are
+provided in the \hyperref[sec-appendix]{Appendix}.
 
@@
 \item
-  Employment precariousness: \texttt{emp\_stat}, \texttt{work\_sit}.
-\item
-  Social precariousness: \texttt{soc\_freq}, \texttt{soc\_adq}.
-\item
-  Housing precariousness: \texttt{nb\_safe}, \texttt{nb\_res},
-  \texttt{nb\_rent}, \texttt{cul\_rec}.
-\item
-  Recent relational stressors: \texttt{frd\_brk12}, \texttt{conf12}.
-\item
-  Recent financial stressors: \texttt{fincri12}, \texttt{inc\_diff}.
+  Employment precariousness: \texttt{emp\_stat} (current employment
+  status), \texttt{work\_sit} (nature of work situation).
+\item
+  Social precariousness: \texttt{soc\_freq} (frequency of social
+  contact), \texttt{soc\_adq} (perceived adequacy of social support).
+\item
+  Housing precariousness: \texttt{nb\_safe} (neighborhood safety),
+  \texttt{nb\_res} (access to neighborhood resources), \texttt{nb\_rent}
+  ( proportion of rental housing), \texttt{cul\_rec} (availability of
+  cultural resources).
+\item
+  Recent relational stressors: \texttt{frd\_brk12} (experience of a
+  friendship breakup), \texttt{conf12} (recent interpersonal or
+  relational conflict).
+\item
+  Recent financial stressors: \texttt{fincri12} (experience of a
+  financial crisis), \texttt{inc\_diff} ( difficulty managing household
+  income).
 \end{itemize}
 
-After preprocessing, the HELIUS dataset comprises 21,628 samples. Along
-with the five precariousness factors, we also compute an overall
-precarity score as the combined value of these factors. Additionally,
-PHQ-9 scores are included to represent depression, both as a total sum
+After preprocessing, the HELIUS dataset includes 21,628 samples. In
+addition to the five identified precariousness factors, we computed an
+overall precarity score by aggregating their values. Depression is
+represented using the PHQ-9 (Kroenke et al., 2001), both as a total sum
 score and as individual symptom scores. In the subsequent causal
-discovery analysis, we examine the relationship between depression and
-precarity using both aggregated sum scores and their individual-level
-representations. Refer to Figure~\ref{fig-dist} for the overall
-distributions of the variables used in the analysis.
+discovery analysis, we explore the relationship between depression and
+precarity using both aggregated scores and their disaggregated
+components. See Figure~\ref{fig-dist} for the overall distributions of
+the variables used in the analysis.
 
@@
 To uncover directional relationships between precariousness and
-depression, we used constraint-based causal discovery algorithms that
-allow for feedback loops and latent confounding. Specifically, we
-applied the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
+depression, we use constraint-based causal discovery algorithms that
+allow for feedback loops and latent confounding. Specifically, we apply
+the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
 Inference} (CCI) algorithms, both of which can detect cycles and
 non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
-reference, we also included the \emph{PC} algorithm, one of the most
+reference, we also include the \emph{PC} algorithm, one of the most
 widely used methods for acyclic causal discovery (Spirtes et al., 2001).
+While our full analysis includes results from three causal discovery
+algorithms---FCI, CCI, and PC---we focus on FCI in the main text, as it
+offers a strong balance between flexibility and theoretical soundness.
+FCI accommodates latent confounding and potential feedback, making it
+well-suited for the complexity of our data (Mooij \& Claassen, 2020).
+The PC algorithm is included for reference but relies on more
+restrictive assumptions, such as acyclicity and no unmeasured
+confounding. CCI, while capable of modeling cycles explicitly, comes
+with certain theoretical limitations. Results from both PC and CCI are
+provided in the \hyperref[sec-appendix]{Appendix}
+(Section~\ref{sec-cci}, Section~\ref{sec-pc}) for completeness and
+comparison. For readers interested in detailed explanations of the
+algorithms, edge interpretations, and graph types (e.g., PAG, MAAG,
+CPDAG), as well as algorithmic assumptions and limitations, please refer
+to the \hyperref[sec-appendix]{Appendix}
+(Section~\ref{sec-causalprimer}).
 
 A key challenge in applying these algorithms to the HELIUS dataset is
-that many variables are non-Gaussian and their relationships are likely
-nonlinear. To address this, we supplemented the commonly used Gaussian
-conditional independence (CI) test---based on partial
-correlations---with a non-parametric alternative: the \emph{Randomized
-Conditional Correlation Test} (RCoT). RCoT approximates kernel-based CI
-testing using random Fourier features, significantly reducing
-computational demands while maintaining sensitivity to nonlinear
-dependencies (Strobl et al., 2019; Zhang et al., 2012).
-
-We examined the causal structure using three complementary approaches:
+that many variables are non-Gaussian and likely exhibit nonlinear
+relationships. While kernel-based conditional independence tests (such
+as KCIT) are well-suited for capturing such complex dependencies, they
+are computationally intensive---scaling quadratically with sample size
+due to the need to invert large kernel matrices---making them
+impractical for large datasets like HELIUS (Rahimi \& Recht, 2007). To
+address this, we use the Randomized Conditional Correlation Test (RCoT),
+a nonparametric alternative that approximates kernel methods using
+random Fourier features. This reduces computational complexity from
+quadratic to linear in sample size, substantially lowering runtime while
+preserving sensitivity to nonlinear relationships (Strobl et al., 2019;
+Zhang et al., 2012). For further details, see Section~\ref{sec-rcot}.
+
+We examine the causal structure using three complementary approaches:
 
@@
 
-To ensure robust estimation and guard against sensitivity to parameter
-settings, we implemented a systematic grid of analysis conditions:
+To robustly estimate the causal structure between depression and
+precariousness variables, we implement a comprehensive bootstrapped
+causal discovery procedure. This procedure systematically varies key
+analysis parameters across a grid of configurations, enabling us to
+assess the stability of inferred relationships under different settings.
+Specifically, we vary four main components:
 
@@
 \item
-  Significance levels: \(\alpha\) = 0.01 and 0.05
-\item
-  Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8
-\item
-  CI tests: Gaussian and RCoT
-\item
-  Algorithms: FCI, CCI, and PC
+  Significance levels for conditional independence tests: \(\alpha\) =
+  0.01 and 0.05.
+\item
+  Stability thresholds: 0.5, 0.6, 0.7, and 0.8, which determine how
+  frequently an edge must appear across bootstraps to be retained.
+\item
+  Conditional independence (CI) tests: Gaussian CI test (partial
+  correlation) and RCoT (a nonparametric test).
+\item
+  Causal discovery algorithms: FCI, CCI, and PC.
 \end{itemize}
 
-Each combination was bootstrapped 100 times for sum-score level analyses
-and 30 times for symptom-level analyses, yielding a total of 1,600
-graphs per algorithm for the full analysis. For symptom networks, we
-fixed the skeleton structure using a consensus graph estimated across
-PC, FCI, and CCI (\(\alpha\) = 0.01, RCoT). This constraint improves
-computational efficiency without limiting the estimation of directional
-edges between symptoms and precariousness variables.
-
-To summarize the results, we adopted a two-level aggregation strategy.
-First, within each analysis condition, we recorded the most frequently
-observed edge-endpoint type (e.g., directed, undirected, uncertain) from
-the bootstraps. Then, across all conditions, we tallied the most
-frequent endpoint type per edge. In cases of ties, we assigned dashed
-edges to reflect uncertainty. For transparency, we accompany each graph
-with a table showing the relative frequencies of each edge symbol.
-
-This procedure yields a final graph for each algorithm and analysis
-type---offering both robustness and interpretability. See
-Figure~\ref{fig-workflow} for an overview of the causal discovery
-pipeline.
-
-For detailed explanations of the algorithms, edge interpretations, and
-graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions
-and limitations, see the \hyperref[sec-appendix]{Appendix}.
+For each unique parameter combination (2 significance levels × 4
+thresholds × 2 CI tests = 16 settings), we run 100 bootstrap samples per
+algorithm for the aggregated analyses (see Step 1 in
+Figure~\ref{fig-workflow}). For symptom-level analyses, which are more
+computationally intensive, we use 30 bootstrap samples per
+setting.\footnote{For symptom-level analyses, we further improve
+  computational efficiency by fixing the skeleton structure---the
+  undirected network of potential connections---using a consensus graph
+  derived from all three algorithms (with \(\alpha = 0.01\) and the RCoT
+  test). This allows subsequent analyses to focus exclusively on
+  estimating edge directions, avoiding the repeated and costly
+  computation of skeletons, and thus substantially reducing runtime.}
+
+To summarize these results, we apply a two-level aggregation strategy:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \textbf{Within each condition} (e.g., fixed \(\alpha\), threshold, CI
+  test, and algorithm), we identify the most frequently occurring edge
+  type (e.g., arrowhead, circle, none) across bootstraps for each pair
+  of variables (Step 2 in Figure~\ref{fig-workflow}).
+\item
+  \textbf{Across conditions}, we then determine the most dominant edge
+  type based on majority voting. In cases where no edge type is clearly
+  dominant (i.e., a tie), we mark the edge as uncertain, represented
+  with a dashed line in the final graph (Step 3 in
+  Figure~\ref{fig-workflow}).
+\end{enumerate}
+
+This summarization process is repeated separately for each causal
+discovery algorithm (FCI, CCI, PC), yielding a single robust graph per
+algorithm. These final graphs represent only the most stable and
+consistent relationships across a wide range of plausible
+parameterizations, enhancing both reliability and interpretability. To
+support transparency, each final graph is accompanied by a matrix
+summarizing the relative frequency of each edge type for every variable
+pair---enabling readers to assess the degree of uncertainty or agreement
+underlying each connection.
 
@@
 
-\includegraphics[width=0.8\textwidth,height=\textheight]{img/simsetup2.png}
-
-}
-
-\caption{\label{fig-workflow}Analysis workflow applied across all three
-algorithms.}
+\includegraphics[width=0.9\textwidth,height=\textheight]{img/simsetup.pdf}
+
+}
+
+\caption{\label{fig-workflow}Causal discovery workflow applied across
+all three algorithms (FCI, CCI, and PC). \emph{The process involves
+three main steps: (1) bootstrapping the dataset to create multiple
+resampled datasets; (2) estimating causal graphs under varying parameter
+configurations; and (3) summarizing results by identifying the most
+frequent edge type across bootstraps and settings to construct a final
+stable graph.}}
 
@@
 
-\includegraphics[width=0.35\textwidth,height=\textheight]{img/conceptmodel.png}
-
-}
-
-\caption{\label{fig-concept}Conceptual high-level causal structure of
-precariousness and depression system.}
+\includegraphics[width=0.4\textwidth,height=\textheight]{img/conceptmodel.png}
+
+}
+
+\caption{\label{fig-concept}High-level conceptual model of causal
+structure linking social precariousness (P.soc) and depression (PHQsum),
+influenced by external financial stress (S.fin).}
 
@@
 To identify realistic parameter settings for the simulation model, we
-calibrated six free parameters---\(\alpha_1\), \(\alpha_2\),
-\(\beta_1\), \(\beta_2\), and noise amplitudes \(\sigma_1\),
-\(\sigma_2\)---using the Non-dominated Sorting Genetic Algorithm II
-(NSGA-II) (Deb et al., 2002), a multi-objective evolutionary
-optimization method well-suited for noisy and high-dimensional search
-spaces. The goal was to match simulated system behavior to empirical
-summary statistics derived from the HELIUS dataset, including the means
-and variances of depression (\(D\)) and social precariousness (\(P\)),
-as well as their partial correlations after adjusting for financial
-stress (\(S\)), specifically \(\rho_{DP.S}\), \(\rho_{DS.P}\), and
-\(\rho_{PS.D}\). Given the stochastic nature of the model, each
-candidate parameter set was evaluated by simulating 1,000 independent
-trajectories, each representing an individual with a fixed \(S\) value
-sampled from the observed distribution. Final-state values were
-extracted and aggregated to compute population-level summary statistics.
-Fitness was assessed as the absolute deviation from the empirical
-targets, aggregated across all objectives. NSGA-II was implemented via
-the nsga2R package (Franz \& Nakamura, 2015), with a population of 300
-evolved over 300 generations using standard hyperparameters (crossover
-probability = 0.9, mutation probability = 0.25, and tournament size =
-2). This process yielded a diverse ensemble of Pareto-efficient
-solutions balancing trade-offs between objectives. From the resulting
-Pareto front (Figure~\ref{fig-pareto}), we selected the top 30 parameter
-sets with the lowest combined error to use in the subsequent simulations
-assessing intervention effects.
+calibrate six free parameters---\(\alpha_1\), \(\alpha_2\), \(\beta_1\),
+\(\beta_2\), and noise amplitudes \(\sigma_1\), \(\sigma_2\)---using the
+\emph{Non-dominated Sorting Genetic Algorithm II} (NSGA-II) (Deb et al.,
+2002), a multi-objective evolutionary optimization method well-suited
+for noisy and high-dimensional search spaces. The goal is to match
+simulated system behavior to empirical summary statistics derived from
+the HELIUS dataset. Specifically, we aim to match the observed means and
+variances of depression (\(D\)) and social precariousness (\(P\)), along
+with all three partial correlations among \(D\), \(P\), and financial
+stress (\(S\)): \(\rho_{DP.S}\), \(\rho_{DS.P}\), and \(\rho_{PS.D}\).
+
+Given the stochastic nature of the model, each candidate parameter set
+is evaluated by simulating 1,000 virtual individuals, each assigned a
+fixed \(S\) value sampled from the empirical distribution. From these
+simulations, we compute population-level summary statistics based on the
+final states. The fitness of each parameter set is measured by how
+closely its summary statistics matched the empirical data, using the
+total absolute deviation across all objectives.
+
+NSGA-II is implemented via the \texttt{nsga2R} package (Franz \&
+Nakamura, 2015), with a population of 500 evolved over 300 generations
+using standard hyperparameters (crossover probability = 0.9, mutation
+probability = 0.25, and tournament size = 2). This process yielded a
+diverse ensemble of Pareto-efficient solutions balancing trade-offs
+between objectives. From the resulting Pareto front
+(Figure~\ref{fig-pareto}), we selected the top 30 parameter sets with
+the lowest combined error to use in the subsequent simulations assessing
+intervention effects.
 
@@
 
-For each of the top 30 calibrated parameter sets, we simulate the
-system's evolution across 300 individuals under a continuum of
-intervention levels. Each simulation run generates final values for
-\(D\) and \(P\), which are then used to assess how internal model
-parameters influence the system's behavior under changing external
-conditions.
+For each of the top 30 parameter sets identified through calibration, we
+run simulations on 300 virtual individuals across this full range of
+intervention levels. In each simulation, we observe how depression
+(\(D\)) and precariousness (\(P\)) evolve over time, capturing their
+final values after the system stabilizes. This setup allows us to
+evaluate how different internal model dynamics (e.g., feedback strength)
+influence the system's response to external change.
 
@@
 The causal graph in panel (a) highlights key pathways in the
-relationships between precarity factors and depression (PHQsum).
-Employment precarity (P.emp) and social precarity (P.soc) are not
-identified as causes of depression, whereas financial stress (S.fin)
-appears to play a causal role. While P.emp and P.soc are generally not
-recognized as causes of other precariousness factors, S.fin emerges as a
-potential cause, as indicated by its circle edge endpoint. Supporting
-this, the proportion matrix plot in panel (b) shows that S.fin has some
-probability of causally influencing either P.emp or P.soc.
+relationships between precarity factors and depression (\emph{PHQsum}).
+Employment precarity (\emph{P.emp}) and social precarity (\emph{P.soc})
+are not identified as causes of depression, whereas financial stress
+(\emph{S.fin}) appears to play a causal role. While \emph{P.emp} and
+\emph{P.soc} are generally not recognized as causes of other
+precariousness factors, \emph{S.fin} emerges as a potential cause, as
+indicated by its circle edge endpoint. Supporting this, the proportion
+matrix plot in panel (b) shows that \emph{S.fin} has some probability of
+causally influencing either \emph{P.emp} or \emph{P.soc.}
 
@@
 
-Relational stress (S.rel) plays a more nuanced role, interacting with
-depression and S.fin through a latent confounder. Its relationship with
-P.soc remains less certain, as the proportion table suggests a fair
-probability that this connection is absent, uncertain, or bidirectional.
-Meanwhile, housing precarity (P.hou) is not directly causally related to
-depression but is linked to P.emp and S.fin. While P.emp and S.fin are
-identified as non-causes of P.hou, it remains unclear whether P.hou
-causally influences P.emp or if their relationship is mediated by an
+Relational stress (\emph{S.rel}) plays a more nuanced role, interacting
+with depression and \emph{S.fin} through a latent confounder. Its
+connection to \emph{P.soc} is less definitive---the proportion matrix in
+Figure~\ref{fig-sum-2} indicates a considerable probability that this
+relationship is either absent, ambiguous, or potentially bidirectional.
+Meanwhile, housing precarity (\emph{P.hou}) is not directly causally
+related to depression but is linked to \emph{P.emp} and \emph{S.fin}.
+While \emph{P.emp} and \emph{S.fin} are identified as non-causes of
+\emph{P.hou}, it remains unclear whether \emph{P.hou} causally
+influences \emph{P.emp} or if their relationship is mediated by an
 unobserved confounder.
@@
 largely consistent with those from the FCI algorithm, with one
-exception: CCI does not identify S.fin as a cause of depression.
+exception: CCI does not identify \emph{S.fin} as a cause of depression.
 Additionally, CCI introduces greater uncertainty in edge directions,
@@
 measure---potentially obscuring nuanced relationships---the
-symptom-level graph highlights heterogeneity in how precariousness
-factors influence individual depressive symptoms: con (concentration),
-slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal),
-anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.
-
-The symptom-level graph in Figure~\ref{fig-sym} panel (a) reveals a more
-complex and interconnected structure than the sum score graph,
-reflecting the strong interdependence among symptoms and suggesting much
-presence of latent confounders, as indicated by numerous bidirectional
-edges. As before, dashed edges denote ties, with specific ties marked by
-bold horizontal bars. In Figure~\ref{fig-sym} panel (b), the proportion
-matrix provides more insight into these directionalities. Here, opaque
-green dominates most symptom-to-symptom connections, indicating that
+symptom-level graph reveals the heterogeneity in how specific symptoms
+(\emph{con} (concentration), \emph{slp} (sleep), \emph{ene} (energy),
+\emph{app} (appetite), \emph{mot} (motor), \emph{sui} (suicidal),
+\emph{anh} (anhedonia), \emph{glt} (guilt), and \emph{dep} (depressed
+mood)) are influenced by, and in turn influence, different forms of
+precariousness.
+
+The symptom-level graph in Figure~\ref{fig-sym-1} reveals a more complex
+and interconnected structure than the sum score graph, reflecting the
+strong interdependence among symptoms and suggesting much presence of
+latent confounders, as indicated by numerous bidirectional edges. As
+before, dashed edges denote ties, with specific ties marked by bold
+horizontal bars. In Figure~\ref{fig-sym-2}, the proportion matrix
+provides more insight into these directionalities. Here, opaque green
+dominates most symptom-to-symptom connections, indicating that
 arrowheads are the most frequently inferred edge type. However, blue
 regions, which correspond to circle endpoints, are particularly common
-for anh, slp, ene, and sui, reflecting uncertainty in the causal
-direction among these symptoms. Notably, the only arrowtail connection
-appears between dep and anh, suggesting that anhedonia (anh) is a
-potential cause of depressed mood (dep).
+for \emph{anh}, \emph{slp}, \emph{ene}, and \emph{sui}, reflecting
+uncertainty in the causal direction among these symptoms. Notably, the
+only arrowtail connection appears between \emph{dep} and \emph{anh},
+suggesting that anhedonia is a potential cause of depressed mood.
 
@@
 that these causal links may be less stable across different conditions.
-Upon closer examination, most ties arise due to discrepancies between
-the Gaussian CI test and the RCoT test---with Gaussian CI favoring
-arrowheads and RCoT more often predicting the absence of an edge.
-
-This discrepancy likely arises from a combination of factors. First, as
-the analysis shifts from aggregated to disaggregated variables, the
-number of conditional independence (CI) tests increases, leading to a
-loss of statistical power in high-dimensional causal discovery. With
-more variables being conditioned on, the data becomes sparser, making it
-harder to detect weaker dependencies. Additionally,
-over-conditioning---controlling for too many variables---can
-artificially remove statistical associations, leading to false
-negatives. Beyond this general statistical power issue, the fundamental
-differences in how these CI tests operate contribute to the
-inconsistency. RCoT, as a nonparametric test, does not assume linearity
-and can capture both linear and nonlinear dependencies. However, it is
-more conservative and requires larger sample sizes, making it more prone
-to false negatives in high-dimensional settings. In contrast, partial
-correlation (Gaussian CI test) assumes linear Gaussian relationships,
-making it more permissive and sometimes detecting weak dependencies that
-may not be statistically meaningful. As a result, when Gaussian CI
-detects an edge while RCoT does not, it could indicate that the
-relationship is weak, strictly linear, or requires more data for
-reliable nonparametric detection. This explains why many ties in the
-proportion matrix show a 50-50 split between symbols---where half of the
-conditions with RCoT suggest the absence of an edge, while the other
-half using Gaussian CI test favor an arrowhead. This systematic
-divergence highlights the challenges of causal discovery in
-high-dimensional settings, where varying statistical assumptions can
-lead to differing conclusions about the presence and direction of causal
-relationships.
+Closer examination reveals that most ties arise due to discrepancies
+between the Gaussian CI test and the RCoT test---with Gaussian CI
+favoring arrowheads and RCoT more often predicting the absence of an
+edge.
+
+This discrepancy likely arises from a combination of factors. As the
+analysis moves from aggregated to disaggregated variables, the number of
+CI tests increases substantially, leading to a loss of statistical
+power. With more variables being conditioned on, the effective sample
+size per test diminishes, making it harder to detect subtle
+dependencies. Over-conditioning---controlling for too many
+variables---can also obscure real relationships, resulting in false
+negatives. Beyond these statistical power issues, the underlying
+assumptions of the two CI tests contribute to their divergence. RCoT is
+a nonparametric test that captures both linear and nonlinear
+dependencies without assuming a specific distributional form. However,
+it is more conservative and requires larger sample sizes, especially in
+high-dimensional settings. In contrast, the Gaussian CI test is based on
+partial correlations and assumes linear Gaussian relationships, making
+it more permissive and more likely to detect weak or spurious
+associations. As a result, it is common to see edges where Gaussian CI
+suggests a directional relationship, while RCoT indicates no edge. In
+the proportion matrix, this appears as a 50--50 tie between absence and
+arrowhead symbols. This systematic divergence illustrates the broader
+challenge of causal discovery in high-dimensional data: different
+statistical assumptions can lead to different conclusions about
+whether---and in which direction---variables are causally related.
 
 Despite these differences, some consistent patterns emerge across both
-CI tests, particularly in the case of financial stress (S.fin), which
-appears to be connected to nearly all other variables---though many of
-these connections are marked by ties, reflecting uncertainty in
+CI tests, particularly in the case of \emph{S.fin}, which appears to be
+connected to nearly all other variables---though many of these
+connections are marked by ties, reflecting uncertainty in
 directionality. One exception is the stronger evidence suggesting that
-S.fin may cause changes in app (appetite), while its relationships with
-other symptoms, such as ene, mot, and dep, remain ambiguous. Similarly,
-social precarity (P.soc) exhibits numerous connections with depressive
-symptoms, with most edges pointing toward P.soc rather than outward from
-it. This suggests that depressive symptoms, particularly dep and glt,
-may contribute to worsening social precarity rather than the other way
-around. Additionally, anh also shows some probability of causally
-influencing P.soc, reinforcing the idea that social precarity is more
-often a consequence rather than a driver of depressive symptoms.
+\emph{S.fin} may cause changes in \emph{app}, while its relationships
+with other symptoms, such as \emph{ene}, \emph{mot}, and \emph{dep},
+remain ambiguous. Similarly, \emph{P.soc} exhibits numerous connections
+with depressive symptoms, with most edges pointing toward \emph{P.soc}
+rather than outward from it. This suggests that depressive symptoms,
+particularly \emph{dep} and \emph{glt}, may contribute to worsening
+social precarity rather than the other way around. Additionally,
+\emph{anh} also shows some probability of causally influencing
+\emph{P.soc}, reinforcing the idea that social precarity is more often a
+consequence rather than a driver of depressive symptoms.
 
 Other precariousness factors exhibit more uncertain but still notable
-relationships. Relational stress (S.rel) shows weak but existing
-connections with slp, glt, and app, though directionality remains
-unclear in many cases. Employment precarity (P.emp) also connects to
-symptoms like slp, mot, and con, but, like S.rel, these relationships
-exhibit a 50/50 split in directionality, reflecting uncertainty in the
-inferred causal paths. In line with the aggregated graph in
-Figure~\ref{fig-sum}, housing precarity (P.hou) does not appear to have
-any direct relationship with depression symptoms but consistently shows
-associations with employment precarity (P.emp) and financial stress
-(S.fin). The causal relationships among precariousness factors remain
-largely unchanged from the aggregated analysis, with financial stress
-(S.fin) exhibiting the strongest tendency to influence other
-precariousness factors.
-
-Among depressive symptoms, dep, glt, and slp appear to be the most
-connected to precariousness factors, while P.soc has the most
-connections with symptoms, predominantly as a recipient rather than a
-driver of influence. Within the symptom network, slp, sui, and anh
-emerge as causally influential symptoms, as they exhibit more outgoing
-arrows compared to other symptoms. On the other hand, con, mot, dep, and
-glt, despite having high connectivity, predominantly receive incoming
-arrows, indicating they are more likely effects rather than causes.
-Considering both symptom-precariousness connections and symptom-level
-dynamics, sleep disturbance (slp) emerges as a central symptom, given
+relationships. \emph{S.rel} shows weak but existing connections with
+\emph{slp}, \emph{glt}, and \emph{app}, though directionality remains
+unclear in many cases. \emph{P.emp} also connects to symptoms like
+\emph{slp}, \emph{mot}, and \emph{con}, but, like \emph{S.rel}, these
+relationships exhibit a 50/50 split in directionality, reflecting
+uncertainty in the inferred causal paths. In line with the aggregated
+graph in Figure~\ref{fig-sum}, \emph{P.hou} does not appear to have any
+direct relationship with depression symptoms but consistently shows
+associations with \emph{P.emp} and \emph{S.fin}. The causal
+relationships among precariousness factors remain largely unchanged from
+the aggregated analysis, with \emph{S.fin} exhibiting the strongest
+tendency to influence other precariousness factors.
+
+Among depressive symptoms, \emph{dep}, \emph{glt}, and \emph{slp} appear
+to be the most connected to precariousness factors, while \emph{P.soc}
+has the most connections with symptoms, predominantly as a recipient
+rather than a driver of influence. Within the symptom network,
+\emph{slp}, \emph{sui}, and \emph{anh} emerge as causally influential
+symptoms, as they exhibit more outgoing arrows compared to other
+symptoms. On the other hand, \emph{con}, \emph{mot}, \emph{dep}, and
+\emph{glt}, despite having high connectivity, predominantly receive
+incoming arrows, indicating they are more likely effects rather than
+causes. Considering both symptom-precariousness connections and
+symptom-level dynamics, \emph{slp} emerges as a central symptom, given
 its strong ties to precariousness factors and its influential role
@@
 initiating symptom, particularly sensitive to relational stress and
-employment precarity. On the other hand, social precarity (P.soc) serves
-as a key bridge between the depression and precariousness subsystems, as
+employment precarity. On the other hand, \emph{P.soc} serves as a key
+bridge between the depression and precariousness subsystems, as
 depressive symptoms appear to feed back into social precarity,
@@
 Lastly, we examine the relationships between individual symptoms and
-overall precarity, an aggregated measure represented by the sum of five
-precariousness factors. This analysis provides a complementary
+overall precariousness, an aggregated measure represented by the sum of
+five precariousness factors. This analysis provides a complementary
 perspective, capturing broader patterns that may not be evident in the
@@
 precariousness factors into a single score, this approach may capture
-distributed relationships across different precarity factors that were
-previously overlooked in the disaggregated analysis.
+distributed relationships across different precariousness factors that
+were previously overlooked in the disaggregated analysis.
 
 Unlike the disaggregated graph in Figure~\ref{fig-sym}, the aggregated
-symptom-precariousness graph (Figure~\ref{fig-presum} panel (a)) shows
-more certainty in causal directions, with only a few edges resulting in
-ties, all of which involve the overall precarity factor. This is even
-more evident in the proportion matrix (Figure~\ref{fig-presum} panel
-(b)), where symptom-to-symptom interactions almost fully converge to a
-single edge type. Additionally, all symptom interactions are connected
-either through bidirectional edges or a combination of an arrowhead and
-a circle, suggesting a strong presence of latent confounders.
-
-Examining the symptom-precarity connections, we observe a stronger
-overall trend of depressive symptoms influencing precarity, rather than
-the other way around. Specifically, anh, dep, and slp show a high
-probability of causing precarity, whereas app, glt, and mot exhibit more
-uncertainty in directionality, often resulting in circle endpoints. This
-pattern suggests that when precariousness factors are aggregated, the
-dominant causal flow is from depressive symptoms to precarity, rather
-than precarity driving depression. Particularly, dep emerges as the
-strongest predictor of precarity, exhibiting the highest proportion of
-arrowtails, consistent with findings from the disaggregated analysis.
+symptom-precariousness graph (Figure~\ref{fig-presum-1}) shows more
+certainty in causal directions, with only a few edges resulting in ties,
+all of which involve the overall precariousness factor. This is even
+more evident in the proportion matrix (Figure~\ref{fig-presum-2}), where
+symptom-to-symptom interactions almost fully converge to a single edge
+type. Additionally, all symptom interactions are connected either
+through bidirectional edges or a combination of an arrowhead and a
+circle, suggesting a strong presence of latent confounders.
+
+Examining the symptom-precariousness connections, we observe a stronger
+overall trend of depressive symptoms influencing precariousness, rather
+than the other way around. Specifically, \emph{anh}, \emph{dep}, and
+\emph{slp} show a high probability of causing precariousness, whereas
+\emph{app}, \emph{glt}, and \emph{mot} exhibit more uncertainty in
+directionality, often resulting in circle endpoints. This pattern
+suggests that when precariousness factors are aggregated, the dominant
+causal flow is from depressive symptoms to precariousness, rather than
+precariousness driving depression. Particularly, \emph{dep} emerges as
+the strongest predictor of precariousness, exhibiting the highest
+proportion of arrowtails, consistent with findings from the
+disaggregated analysis.
 
@@
 symptoms are interconnected through bidirectional edges, suggesting a
-strong presence of latent confounders. However, dep appears to be a
-distinct exception, as it is predominantly caused by nearly all other
-symptoms, with one exception sui, which does not contribute to dep.
-Similarly, con is influenced by multiple symptoms, albeit with weaker
-support compared to dep. Most interestingly, the CCI results highlight a
-distinct feedback loop between dep and overall precarity, represented by
-a tail-tail edge (\textemdash ). This suggests a possible reinforcing
-cycle between depression and overall precariousness, where depression
-not only arises from precarious conditions but also contributes to their
-persistence, creating a self-sustaining dynamic.
+strong presence of latent confounders. However, \emph{dep} appears to be
+a distinct exception, as it is predominantly caused by nearly all other
+symptoms, with one exception \emph{sui}, which does not contribute to
+\emph{dep}. Similarly, \emph{con} is influenced by multiple symptoms,
+albeit with weaker support compared to \emph{dep}. Most interestingly,
+the CCI results highlight a distinct feedback loop between \emph{dep}
+and overall precariousness, represented by a tail-tail edge
+(\textemdash ). This suggests a possible reinforcing cycle between
+depression and overall precariousness, where depression not only arises
+from precarious conditions but also contributes to their persistence,
+creating a self-sustaining dynamic.
 
@@
 symptom-to-symptom interactions. While certain stressor-related
-precariousness factors, particularly S.fin and S.rel, appear to causally
-influence depressive symptoms, other precariousness factors are more
-likely driven by depression rather than acting as primary causes,
-suggesting that precariousness may function more as a consequence of
-depression than its source. Some symptoms, particularly glt, slp, and
-dep, show greater sensitivity to precarious conditions, while also
-playing a central role in symptom dynamics. Among them, slp and glt
-emerge as influential symptoms within the depression network, whereas
-con and dep, despite their high connectivity, behave more as outcomes
+precariousness factors, particularly \emph{S.fin} and \emph{S.rel},
+appear to causally influence depressive symptoms, other precariousness
+factors are more likely driven by depression rather than acting as
+primary causes, suggesting that precariousness may function more as a
+consequence of depression than its source. Some symptoms, particularly
+\emph{glt}, \emph{slp}, and \emph{dep}, show greater sensitivity to
+precarious conditions, while also playing a central role in symptom
+dynamics. Among them, \emph{slp} and \emph{glt} emerge as influential
+symptoms within the depression network, whereas \emph{con} and
+\emph{dep}, despite their high connectivity, behave more as outcomes
 rather than primary drivers. Together, these findings suggest that
@@
   \(\textit{SD}_P = 0.9\) for precariousness (\(P\)). Parameter strength
-  is defined as the product \(\beta_1 \alpha_1 \beta_2 \alpha_2\),
-  reflecting the combined influence of internal feedback and external
-  input sensitivity.} These results suggest that systems with stronger
+  is defined as the product
+  \(\alpha_1 \times \beta_1 \times \alpha_2 \times \beta_2\), reflecting
+  the combined influence of internal feedback and external input
+  sensitivity.} These results suggest that systems with stronger
 internal dynamics are more responsive to external change, enabling
@@
 parameter sets: one with strong feedback/input coupling and low noise (a
-``bistable'' candidate), and one with weaker coupling and higher noise
-(a ``unistable'' candidate). In nonlinear systems, bistability refers to
-the existence of two distinct stable equilibria---meaning the system's
-long-term outcome depends on its starting point or perturbations.
+\emph{bistable} candidate), and one with weaker coupling and higher
+noise (a \emph{unistable} candidate). In nonlinear systems, bistability
+refers to the existence of two distinct stable equilibria---meaning the
+system's long-term outcome depends on its starting point or
+perturbations.
 
@@
 applied cycle-capable discovery algorithms to estimate plausible
-directional relationships among financial stress, multiple domains of
-precariousness, and depressive symptoms. By analyzing both aggregate
-constructs and individual variables, we aimed to capture patterns that
-may be masked in either extreme of abstraction (i.e., purely aggregate
-or disaggregated analyses).
-
-The causal analysis consistently identified financial stress
-(\emph{S.fin}) as a key plausible driver of depression, appearing across
-both aggregated and disaggregated models. At the symptom level, we found
-that sleep disturbance (\emph{slp}), depressed mood (\emph{dep}), guilt
-(\emph{glt}), and anhedonia (\emph{anh}) were especially sensitive to
-external stressors. Among these, \emph{dep} emerged as a central hub:
-not only was it heavily influenced by other symptoms, it also exerted
-directional influence on social precarity (\emph{P.soc}), suggesting a
-feedback loop in which worsening mood contributes to social
-disconnection, which in turn may exacerbate depressive symptoms. The CCI
-algorithm even suggested a possible cyclic relationship between dep and
-overall precarity, hinting at a structurally self-reinforcing mechanism.
-\emph{Slp} also played a key initiating role. It showed directed
-connections to both social and employment precarity and exhibited many
-outgoing edges within the symptom network. This suggests \emph{slp} may
-function as an early warning signal or entry point into the
-depression--precarity cycle, making it a promising target for preventive
-intervention. In contrast, \emph{P.soc} more often appeared as an effect
-rather than a cause---receiving arrows from symptoms like \emph{dep} and
-\emph{glt}, but rarely sending them. This asymmetry implies that social
-precarity may often reflect downstream consequences of worsening mental
-health rather than being its root cause. These patterns offer actionable
-insights for intervention. Addressing symptoms like \emph{slp} early may
-prevent escalation into broader depressive states or worsening
-precarity. Intervening on \emph{dep}, as a central node with potential
-feedback effects, could also disrupt the self-reinforcing dynamics
-between mental health and social isolation.
+directional relationships among multiple domains of precariousness
+factors, and depressive symptoms. By analyzing both aggregate constructs
+and individual variables, we aimed to capture patterns that may be
+masked in either extreme of abstraction (i.e., purely aggregate or
+disaggregated analyses).
+
+The causal analysis consistently identified financial stress as a key
+plausible driver of depression, appearing across both aggregated and
+disaggregated models. At the symptom level, we found that sleep
+disturbance, depressed mood, guilt, and anhedonia were especially
+sensitive to external stressors. Among these, the symptom depressed mood
+emerged as a central hub: not only was it heavily influenced by other
+symptoms, it also exerted directional influence on social
+precariousness, suggesting a feedback loop in which worsening mood
+contributes to social disconnection, which in turn may exacerbate
+depressive symptoms. The CCI algorithm even suggested a possible cyclic
+relationship between depressed mood and overall precariousness, hinting
+at a structurally self-reinforcing mechanism. Symptom sleep disturbance
+also played a key initiating role. It showed directed connections to
+both social and employment precariousness and exhibited many outgoing
+edges within the symptom network. This suggests the symptom sleep
+disturbance may function as an early warning signal or entry point into
+the depression--precariousness cycle, making it a promising target for
+preventive intervention. In contrast, Social precariousness more often
+appeared as an effect rather than a cause---receiving arrows from
+symptoms like depressed mood and guilt, but rarely sending them. This
+asymmetry implies that social precariousness may often reflect
+downstream consequences of worsening mental health rather than being its
+root cause. These patterns offer actionable insights for intervention.
+Addressing symptoms like sleep disturbance early may prevent escalation
+into broader depressive states or worsening precarity. Intervening on
+depressed mood, as a central node with potential feedback effects, could
+also disrupt the self-reinforcing dynamics between mental health and
+social isolation.
 
@@
 the links between individual symptoms and precariousness factors, where
-edge estimates varied across algorithms and conditional independence
-(CI) tests. This likely reflects limited statistical signal in
-high-dimensional settings, where large conditioning sets reduce power
-and small effect sizes are harder to detect. Although the CCI algorithm
-is explicitly designed to detect cycles, we found little clear evidence
-of feedback: most candidate loops appeared only as bidirectional edges,
-which may reflect latent confounding or statistical imprecision rather
-than true cyclic structure. This could point to either an absence of
-strong feedback at the symptom level or limitations in the sensitivity
-of full-graph discovery approaches applied to densely interconnected
-variables. More focused strategies, such as local structure tests or
-theory-guided subgraph modeling, may help resolve these ambiguities. In
-addition, while our analyses relied solely on observational data, some
-structural uncertainties---particularly those involving latent
-confounding or bidirectional influence---may ultimately require
-interventional or semi-interventional designs. Extensions such as LLC
-(Hyttinen et al., 2012), NODAGS-Flow (Sethuraman et al., 2023), or
-Bicycle (Rohbeck et al., 2024), which integrate observational and
-interventional data, may prove valuable where such data are available.
-
-Second, our findings showed sensitivity to the choice of conditional
-independence (CI) test. The Gaussian CI test, which assumes linear
-relationships and Gaussian noise, often produced denser graphs---likely
-due to its tendency to overestimate weak or noisy linear associations.
-In contrast, RCoT, a nonparametric alternative, avoids these assumptions
-but relies on a Gaussian radial basis function (RBF) kernel to assess
-dependence. While flexible in capturing nonlinear relationships, this
-kernel is optimized for smooth, continuous variables and may
-underperform with discrete or ordinal data. Its similarity estimates can
-be unreliable when applied to variables with discontinuities,
-potentially leading to missed dependencies---particularly relevant in
-the HELIUS dataset, where some variables are ordinal or mixed-type
-(Howlett, 2001). To address this, future work should consider adapting
-kernel-based CI tests to better handle mixed data structures---for
-example, by employing hybrid kernels that explicitly combine continuous
-and discrete similarity measures. This could improve the reliability of
-nonparametric causal discovery in settings with heterogeneous variable
-types.
+edge estimates varied across algorithms and CI tests. This likely
+reflects limited statistical signal in high-dimensional settings, where
+large conditioning sets reduce power and small effect sizes are harder
+to detect. Although the CCI algorithm is explicitly designed to detect
+cycles, we found little clear evidence of feedback: most candidate loops
+appeared only as bidirectional edges, which may reflect latent
+confounding or statistical imprecision rather than true cyclic
+structure. This could point to either an absence of strong feedback at
+the symptom level or limitations in the sensitivity of full-graph
+discovery approaches applied to densely interconnected variables. More
+focused strategies, such as local structure tests or theory-guided
+subgraph modeling, may help resolve these ambiguities. In addition,
+while our analyses relied solely on observational data, some structural
+uncertainties---particularly those involving latent confounding or
+bidirectional influence---may ultimately require interventional or
+semi-interventional designs. Extensions such as LLC (Hyttinen et al.,
+2012), NODAGS-Flow (Sethuraman et al., 2023), or Bicycle (Rohbeck et
+al., 2024), which integrate observational and interventional data, may
+prove valuable where such data are available.
+
+Second, our findings showed sensitivity to the choice of CI test. The
+Gaussian CI test, which assumes linear relationships and Gaussian noise,
+often produced denser graphs---likely due to its tendency to
+overestimate weak or noisy linear associations. In contrast, RCoT, a
+nonparametric alternative, avoids these assumptions but relies on a
+Gaussian radial basis function (RBF) kernel to assess dependence. While
+flexible in capturing nonlinear relationships, this kernel is optimized
+for smooth, continuous variables and may underperform with discrete or
+ordinal data. Its similarity estimates can be unreliable when applied to
+variables with discontinuities, potentially leading to missed
+dependencies---particularly relevant in the HELIUS dataset, where some
+variables are ordinal or mixed-type (Howlett, 2001). To address this,
+future work should consider adapting kernel-based CI tests to better
+handle mixed data structures---for example, by employing hybrid kernels
+that explicitly combine continuous and discrete similarity measures.
+This could improve the reliability of nonparametric causal discovery in
+settings with heterogeneous variable types.
 
@@
 discovered causal structure, it represents a highly simplified system:
-only two variables---depression and social precarity---modulated by
+only two variables---depression and social precariousness---modulated by
 external financial stress. This abstraction allowed for tractable
@@
 
+\bibitem[\citeproctext]{ref-kroenke2001phq}
+Kroenke, K., Spitzer, R. L., \& Williams, J. B. (2001). The PHQ-9:
+Validity of a brief depression severity measure. \emph{Journal of
+General Internal Medicine}, \emph{16}(9), 606--613.
+
 \bibitem[\citeproctext]{ref-lindsay2000moment}
@@
 
-\subsection{HELIUS study}\label{helius-study}
+\subsection{HELIUS study}\label{sec-helius}
 
@@
 
-\subsection{Causal Discovery Primer}\label{causal-discovery-primer}
+\subsection{Causal Discovery Primer}\label{sec-causalprimer}
 
@@
 discussing CCI results where relevant, with detailed findings provided
-in the \hyperref[sec-appendix]{Appendix}. A full discussion of causal
-discovery concepts and algorithmic details is beyond the scope of this
-paper; however, readers seeking a more in-depth understanding can refer
-to Park et al. (2024) for a comprehensive exploration of these methods
-and their applications.
+in Section~\ref{sec-cci}. A full discussion of causal discovery concepts
+and algorithmic details is beyond the scope of this paper; however,
+readers seeking a more in-depth understanding can refer to Park et al.
+(2024) for a comprehensive exploration of these methods and their
+applications.
 
@@
 results obtained from FCI and CCI, with all PC algorithm results
-provided in the \hyperref[sec-appendix]{Appendix} for completeness.
-
-\subsection{Precariousness factors by
-Leonie}\label{precariousness-factors-by-leonie}
+provided in Section~\ref{sec-pc} for completeness.
+
+\subsection{Precariousness Domains Based on Elsenburg et
+al.~(2025)}\label{precariousness-domains-based-on-elsenburg-et-al.-2025}
 
@@
 
-\subsection{Exploratory Factor Analysis
+\subsubsection{Exploratory Factor Analysis
 (EFA)}\label{exploratory-factor-analysis-efa}
@@
 
-\subsubsection{Factor Loadings (Pattern
+\paragraph{Factor Loadings (Pattern
 Matrix)}\label{factor-loadings-pattern-matrix}
@@
 
-\subsubsection{Variance Explained}\label{variance-explained}
+\paragraph{Variance Explained}\label{variance-explained}
 
@@
 
-\subsubsection{Factor Intercorrelations}\label{factor-intercorrelations}
+\paragraph{Factor Intercorrelations}\label{factor-intercorrelations}
 
@@
 
-\subsubsection{Model Fit Statistics}\label{model-fit-statistics}
+\paragraph{Model Fit Statistics}\label{model-fit-statistics}
 
@@
 
-\subsubsection{Summary}\label{summary}
+\paragraph{Summary}\label{summary}
 
@@
 
-\subsection{PCA}\label{pca}
+\subsubsection{PCA}\label{pca}
 
@@
 
-\subsubsection{Explained variance (contributions) of
+\paragraph{Explained variance (contributions) of
 variables}\label{explained-variance-contributions-of-variables}
@@
 
-\subsubsection{Cos² Values}\label{cosuxb2-values}
+\paragraph{Cos² Values}\label{cosuxb2-values}
 
@@
 
-\subsection{ICA}\label{ica}
+\subsubsection{ICA}\label{ica}
 
@@
 
-\subsubsection{Dominant Variables per
+\paragraph{Dominant Variables per
 Component:}\label{dominant-variables-per-component}
@@
 
-\subsection{Hierarchical clustering}\label{hierarchical-clustering}
-
-\subsubsection{Using Euclidean distance}\label{using-euclidean-distance}
+\subsubsection{Hierarchical clustering}\label{hierarchical-clustering}
+
+\paragraph{Using Euclidean distance}\label{using-euclidean-distance}
 
@@
 
-\paragraph{Consistent Groupings (Across All or Most
+\subparagraph{Consistent Groupings (Across All or Most
 Methods)}\label{consistent-groupings-across-all-or-most-methods}
@@
 
-\paragraph{Inconsistent Groupings (Variability Across
+\subparagraph{Inconsistent Groupings (Variability Across
 Methods)}\label{inconsistent-groupings-variability-across-methods}
@@
 
-\paragraph{Summary}\label{summary-1}
+\subparagraph{Summary}\label{summary-1}
 
@@
 
-\subsubsection{Using Mutual Information}\label{using-mutual-information}
+\paragraph{Using Mutual Information}\label{using-mutual-information}
 
@@
 
-\paragraph{Comparison to Euclidean Distance
+\subparagraph{Comparison to Euclidean Distance
 Clustering}\label{comparison-to-euclidean-distance-clustering}
@@
 
-\paragraph{Summary}\label{summary-2}
+\subparagraph{Summary}\label{summary-2}
 
@@
 
-\subsection{Conclusions on Precariousness
+\subsubsection{Conclusions on Precariousness
 factors}\label{conclusions-on-precariousness-factors}
@@
 
-\subsection{Results from PC algorithm}\label{results-from-pc-algorithm}
+\subsection{Results from PC algorithm}\label{sec-pc}
 

### 2025-05-20 code/intervention_simulation/00_optimize_sdemodel.R

@@ -97,7 +97,7 @@ result7 <- nsga2R::nsga2R(
   lowerBounds = rep(0, 6),
   upperBounds = rep(2, 6),
   popSize = 500,
-  generations = 100,
+  generations = 300,
   cprob = 0.9,                 # Crossover probability
   mprob = 1 / 4,               # Mutation probability
   tourSize = 2,               # Tournament size

### 2025-06-10 draft_v1/draft_v1.qmd
update_draft_Leoniefeedback

@@
 ---
-title: "From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study"
+title: "From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study"
 #subtitle: Using the HELIUS dataset
@@
 Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
-  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precarity under varying levels of financial stress reduction.
+  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precariousness under varying levels of financial stress reduction.
 Simulation results revealed that the system's responsiveness depends not only on the intensity of external intervention but also on internal structural features such as feedback strength and noise.
@@
 
-Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains—including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering].
+Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains—including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering; @mckee2017living].
 This broader view reveals how different forms of insecurity can co-occur and compound, forming an ecosystem of risk.
@@
 This study aims to investigate the causal mechanisms linking precariousness and depression—and how those mechanisms shape system responses to external intervention.
-Specifically, we ask: *How does the internal configuration of the precarity–depression relationship affect the outcomes of stress-reducing interventions?*
+Specifically, we ask: *How does the internal configuration of the precariousness–depression relationship affect the outcomes of stress-reducing interventions?*
 To answer this, we combine two approaches.
 First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors, and depressive symptoms.
-By examining both aggregated measures (depression sum score and a composite precarity index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
-Second, we translate these insights into a computational dynamical model that simulates how depression and precarity co-evolve under varying levels of external stress.
+By examining both aggregated measures (depression sum score and a composite precariousness index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
+Second, we translate these insights into a computational dynamical model that simulates how depression and precariousness co-evolve under varying levels of external stress.
 This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention.
@@
 In others, it shows unistable behavior, responding more gradually to incremental changes.
-By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precarity system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.
+By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precariousness system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.
 
@@
 <!-- This study aims to investigate the causal mechanisms linking precariousness and depression, and to examine how those mechanisms shape the system’s response to external intervention.
-Specifically, we ask: How do different configurations of the precarity–depression relationship affect the outcomes of interventions aimed at reducing external stress?
+Specifically, we ask: How do different configurations of the precariousness–depression relationship affect the outcomes of interventions aimed at reducing external stress?
 -->
@@
 We apply cycle-capable causal discovery algorithms to infer directional relationships among financial stress, precariousness indicators, and depressive symptoms.
-Our analysis includes both aggregate scores (e.g., PHQ-9 total, precarity composite index) and disaggregated variables (individual symptoms and domain-specific precariousness), allowing us to capture both broad associations and fine-grained pathways.
--->
-
-<!-- Building on the discovered causal patterns, we simulate a set of scenarios using a computational model that captures the co-evolution of depression and precarity under varying levels of financial stress.
+Our analysis includes both aggregate scores (e.g., PHQ-9 total, precariousness composite index) and disaggregated variables (individual symptoms and domain-specific precariousness), allowing us to capture both broad associations and fine-grained pathways.
+-->
+
+<!-- Building on the discovered causal patterns, we simulate a set of scenarios using a computational model that captures the co-evolution of depression and precariousness under varying levels of financial stress.
 This framework allows us to examine how internal system properties—particularly feedback strength and stochastic noise—influence responsiveness to intervention.
@@
 In others, it transitions gradually, reflecting unistable dynamics.
-Through this integrative approach, we aim to show how the internal architecture of the depression–precarity system fundamentally shapes its susceptibility to change—highlighting the importance of structural understanding in designing effective, equitable mental health interventions.
+Through this integrative approach, we aim to show how the internal architecture of the depression–precariousness system fundamentally shapes its susceptibility to change—highlighting the importance of structural understanding in designing effective, equitable mental health interventions.
 -->
@@
 We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study, which has been described in detail elsewhere [@stronks2013unravelling; @snijder2017cohort].
-It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and which provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement].
+It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement].
 To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
@@
 
-After preprocessing, the HELIUS dataset includes 21,628 samples.
-In addition to the five identified precariousness factors, we computed an overall precarity score by aggregating their values.
+After preprocessing, the HELIUS dataset includes 21,628 individuals.
+In addition to the five identified precariousness factors, we computed an overall precariousness score by by summing these five factor scores, thereby capturing a composite measure of accumulated precariousness across domains.
 Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores.
-In the subsequent causal discovery analysis, we explore the relationship between depression and precarity using both aggregated scores and their disaggregated components.
+In the subsequent causal discovery analysis, we explore the relationship between depression and precariousness using both aggregated scores and their disaggregated components.
 See @fig-dist for the overall distributions of the variables used in the analysis.
@@
 #|   Distributions of variables with density overlay.  
+#|   *This shows the distributions of the variables used in the causal analysis, plotted as histograms with overlaid density estimates.
+The x-axis represents the value of each variable after preprocessing.
+The y-axis shows the corresponding estimated probability density.*
+#|   ***P.emp*** = *employment precariousness*;  ***P.hou*** = *housing precariousness*;  ***P.soc*** = *social precariousness*;  ***S.fin*** = *recent financial stressors*;  ***S.rel*** = *recent relational stressors*;  ***precariousness*** = *overall precariousness*;  ***PHQsum*** = *PHQ-9 sum score*;  ***anh*** = *anhedonia*;  ***app*** = *appetite*;  ***con*** = *concentration*;  ***dep*** = *depressed mood*;  ***ene*** = *energy*;  ***glt*** = *guilty*;  ***mot*** = *motor*;  ***slp*** = *sleep*;  ***sui*** = *suicidal*.
 #|   
-#|   *P.emp* = employment precariousness; *P.hou* = housing precariousness; *P.soc* = social precariousness; *S.fin* = recent financial stressors; *S.rel* = recent relational stressors; *Precarity* = overall precarity;
-#|   *PHQsum* = PHQ-9 sum score; *anh* = anhedonia; *app* = appetite; *con* = concentration; *dep* = depressed mood; *ene* = energy; *glt* = guilty; *mot* = motor; *sui* = suicidal
 #| label: fig-dist
@@
     S.fin = rowMeans(cbind(fincri12, inc_dif)),
-    Precarity = P.emp + P.soc + P.hou + S.rel + S.fin
+    precariousness = P.emp + P.soc + P.hou + S.rel + S.fin
   ) |>
@@
   geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
-  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "Precarity", "PHQsum", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui")), scales = "free") +
+  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "precariousness", "PHQsum", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui")), scales = "free") +
   labs(x = "", y = "") +
@@
 
-The aggregate analysis reduces dimensionality and yields interpretable summaries of how domains of precarity relate to overall depression severity.
+The aggregate analysis reduces dimensionality and yields interpretable summaries of how domains of precariousness relate to overall depression severity.
 However, it may obscure finer-grained effects.
 The disaggregated analysis allows for precise mapping of which symptoms are influenced by (or influence) specific precariousness conditions, but it introduces complexity due to the high dimensionality and distributional properties of the data.
-The mixed approach, in turn, integrates these perspectives by examining how individual symptoms respond to cumulative precarity across domains.
+The mixed approach, in turn, integrates these perspectives by examining how individual symptoms respond to cumulative precariousness across domains.
 
@@
 -->
-<!-- By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precarity and depression.
+<!-- By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precariousness and depression.
 -->
@@
 
-## Modeling Dynamics between Precarity and Depression
+## Modeling Dynamics between precariousness and Depression
 
@@
 
-NSGA-II is implemented via the `nsga2R` package [@nsga2R], with a population of 500 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2).
+<!-- NSGA-II is implemented via the `nsga2R` package [@nsga2R], with a population of 500 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2).
 This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives.
 From the resulting Pareto front (@fig-pareto), we selected the top 30 parameter sets with the lowest combined error to use in the subsequent simulations assessing intervention effects.
+-->
+
+To explore optimal parameter configurations, we used the NSGA-II algorithm as implemented in the `nsga2R` package [@nsga2R].
+This algorithm is inspired by natural selection and evolves a population of candidate solutions over multiple generations to balance competing objectives.
+We began with a population of 500 randomly generated parameter sets and allowed them to evolve over 300 generations.
+During this process, solutions were selected, combined, and mutated using standard settings: a crossover probability of 0.9 (which controls how often solutions are combined), a mutation probability of 0.25 (introducing random variation), and a tournament size of 2 (used to pick the best candidates for reproduction).
+The result is a Pareto front—a set of solutions that are considered optimal in the sense that no single objective can be improved without worsening another.
+From this diverse set of trade-off solutions (see @fig-pareto), we selected the 30 parameter sets with the lowest overall error across objectives.
+These top-performing configurations were then used in the simulation phase to evaluate the potential effects of different interventions.
 
@@
 
-## Causal Structure Linking Precarity and Depression
+## Causal Structure Linking Precariousness and Depression
 
 ### Depression as sum score
-The sum score graphs provide a high-level summary of how precarity factors collectively influence overall depression severity, focusing on aggregated relationships.
+The sum score graphs provide a high-level summary of how precariousness factors collectively influence overall depression severity, focusing on aggregated relationships.
 @fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
-As described in [@sec-analysis], dashed edges indicate ties, with the exact edge endpoint where the tie occurs is represented by a bold horizontal bar.
-
-The causal graph in panel (a) highlights key pathways in the relationships between precarity factors and depression (*PHQsum*).
-Employment precarity (*P.emp*) and social precarity (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role.
+As described in [@sec-analysis], dashed edges indicate ties, and the edge endpoint at which the tie occurs is marked by a bold horizontal bar.
+
+The causal graph in panel (a) highlights key pathways in the relationships between precariousness factors and depression (*PHQsum*).
+Employment precariousness (*P.emp*) and social precariousness (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role.
 While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint.
@@
 Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 indicates a considerable probability that this relationship is either absent, ambiguous, or potentially bidirectional.
-Meanwhile, housing precarity (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*.
+Meanwhile, housing precariousness (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*.
 While *P.emp* and *S.fin* are identified as non-causes of *P.hou*, it remains unclear whether *P.hou* causally influences *P.emp* or if their relationship is mediated by an unobserved confounder.
@@
 
-<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precarity  (P.emp) and social precarity (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.
-Additionally, P.emp and S.fin are identified as non-causes of housing precarity (P.hou).
+<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precariousness  (P.emp) and social precariousness (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.
+Additionally, P.emp and S.fin are identified as non-causes of housing precariousness (P.hou).
 Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship.
@@
 <!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders.
-Employment and social precarity may also be influenced by depression, either directly or through latent variables.
-Lastly, housing precarity does not directly impact depression but remains connected to employment precarity and financial stress, likely mediated by direct or latent pathways.
+Employment and social precariousness may also be influenced by depression, either directly or through latent variables.
+Lastly, housing precariousness does not directly impact depression but remains connected to employment precariousness and financial stress, likely mediated by direct or latent pathways.
 -->
@@
 #| label: fig-sum
-#| fig-cap: "Resulting graph of precarity factors and depression sum score using FCI and proportion of edge endpoint types."
+#| fig-cap: "Resulting graph of precariousness factors and depression sum score using FCI and proportion of edge endpoint types."
 #| fig-subcap: 
@@
 Similarly, *P.soc* exhibits numerous connections with depressive symptoms, with most edges pointing toward *P.soc* rather than outward from it.
-This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precarity rather than the other way around.
-Additionally, *anh* also shows some probability of causally influencing *P.soc*, reinforcing the idea that social precarity is more often a consequence rather than a driver of depressive symptoms.
+This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precariousness rather than the other way around.
+Additionally, *anh* also shows some probability of causally influencing *P.soc*, reinforcing the idea that social precariousness is more often a consequence rather than a driver of depressive symptoms.
 
@@
 Considering both symptom-precariousness connections and symptom-level dynamics, *slp* emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network.
-This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precarity.
-On the other hand, *P.soc* serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precarity, reinforcing a self-sustaining dynamic between depression and precarious conditions.
+This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precariousness.
+On the other hand, *P.soc* serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precariousness, reinforcing a self-sustaining dynamic between depression and precarious conditions.
 
@@
 Among the depressive symptoms, *anh*, *dep*, *slp*, and *glt* emerge as particularly influential, with multiple outgoing edges (o->) to other symptoms,  suggesting their roles as potential key drivers within the symptom network.
-Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precarity factors.
+Especially, symptoms such as *glt*, *slp*, and *anh* are particularly critical, potentially acting as initiator or activator nodes within the network due to their apparent connections with precariousness factors.
 These symptoms appear to be especially sensitive to external stressors, potentially manifesting early in response to such conditions and subsequently activating other interconnected symptom nodes.
@@
 <!-- <!-- This sensitivity indicates their potential as early warning signs or key targets for interventions aimed at preventing the progression to severe depression. --> 
-<!-- The causal structure involving precarity factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs.
-As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precarity (*P.soc*) are more likely to be influenced by depressive symptoms.
+<!-- The causal structure involving precariousness factors in the symptom-level graph is largely consistent with the patterns observed in the sum score graphs.
+As in the sum graphs, relational stress (*S.rel*) emerges as a potential causal factor for depression, while employment (*P.emp*) and social precariousness (*P.soc*) are more likely to be influenced by depressive symptoms.
 -->
@@
 <!-- Some discrepancies, however, exist between the symptom-level and sum graphs.
-For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precarity factors.
-Additionally, housing precarity (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.
+For example, financial stress (*S.fin*) does not have any edges with depressive symptoms in the symptom-level graph, although it maintains associations with other precariousness factors.
+Additionally, housing precariousness (*P.hou*) becomes entirely disconnected from the rest of the network, appearing as an isolated node.
 -->
@@
 #| layout-nrow: 2
-#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using FCI and proportion of edge endpoint types."
+#| fig-cap: "Resulting graph of precariousness factors and individual depression symptoms using FCI and proportion of edge endpoint types."
 #| fig-subcap: 
@@
 
-### Precarity as sum score
+### Precariousness as sum score
 
@@
 While certain stressor-related precariousness factors, particularly *S.fin* and *S.rel*, appear to causally influence depressive symptoms, other precariousness factors are more likely driven by depression rather than acting as primary causes, suggesting that precariousness may function more as a consequence of depression than its source.
-Some symptoms, particularly *glt*, *slp*, and *dep*, show greater sensitivity to precarious conditions, while also playing a central role in symptom dynamics.
+Some symptoms, particularly *glt*, *slp*, and *dep*, appear to influence precarious conditions while also occupying central roles within the symptom network.
 Among them, *slp* and *glt* emerge as influential symptoms within the depression network, whereas *con* and *dep*, despite their high connectivity, behave more as outcomes rather than primary drivers.
@@
 #| label: fig-presum
-#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
+#| fig-cap: "Resulting graph of precariousness sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
 #| fig-subcap: 
@@
 
-## Simulating the Dynamics Between Precarity and Depression Under Intervention
+## Simulating the Dynamics Between Precariousness and Depression Under Intervention
 
 To evaluate how internal system structure influences the effect of external support, we simulated outcomes under a range of intervention strengths across the top 30 parameter sets from the calibrated model ensemble.
-The top panel of @fig-intervention shows mean levels of depression ($D$) and precarity ($P$) after simulation, with each line representing one parameter set.
+The top panel of @fig-intervention shows mean levels of depression ($D$) and precariousness ($P$) after simulation, with each line representing one parameter set.
 Lines are color-coded by combined feedback and input strength, calculated as the product $\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2$.
@@
 
-Tailoring intervention strategies to the system’s underlying structure may enhance their effectiveness and sustainability, particularly in addressing the compounding effects of socioeconomic precarity on mental health.
-
-<!-- The final outcomes, specifically the mean levels of depression ($D$) and precarity ($P$), are presented in the top panel of @fig-intervention.
+Tailoring intervention strategies to the system’s underlying structure may enhance their effectiveness and sustainability, particularly in addressing the compounding effects of socioeconomic precariousness on mental health.
+
+<!-- The final outcomes, specifically the mean levels of depression ($D$) and precariousness ($P$), are presented in the top panel of @fig-intervention.
 Each trajectory corresponds to a parameter set and is color-coded by its combined feedback and input strength, defined as the product $\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2$.
@@
 <!-- Systems with stronger parameter interactions exhibit more pronounced shifts in behavior, particularly under intensive interventions.
-At full intervention (i.e., maximum stress reduction), the mean depression level decreased from -0.113 to -0.963, and precarity from -0.100 to -0.958.
+At full intervention (i.e., maximum stress reduction), the mean depression level decreased from -0.113 to -0.963, and precariousness from -0.100 to -0.958.
 These changes represent reductions of approximately 90.8\% relative to their respective baseline magnitudes, indicating that effective intervention can push the system close to its lowest-risk equilibrium.
@@
 For unistable systems, more moderate or incremental measures may suffice.
-By tailoring strategies to the system’s underlying structure, interventions can be made more effective and context-sensitive—ultimately enhancing support for those facing precarity-driven mental health risks.
+By tailoring strategies to the system’s underlying structure, interventions can be made more effective and context-sensitive—ultimately enhancing support for those facing precariousness-driven mental health risks.
 -->
@@
 #| fig-cap: "**Top**: Final Mean D and P Across Intervention Sizes.
-*Mean trajectories of depression ($D$) and precarity ($P$) across varying intervention strengths, simulated over the top 30 parameter sets.
+*Mean trajectories of depression ($D$) and precariousness ($P$) across varying intervention strengths, simulated over the top 30 parameter sets.
 Each line represents one parameter set and is colored by its combined feedback and input strength.* **Bottom**: Smoothed Density Heatmap of Final $D$ and $P$ for Two Parameter Sets.
@@
 <!-- Summmary & Interpretation of Findings -->
-This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precarity interact—and how their internal structure modulates responsiveness to intervention.
+This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact—and how their internal structure modulates responsiveness to intervention.
 Using cross-sectional data from the HELIUS study, we applied cycle-capable discovery algorithms to estimate plausible directional relationships among multiple domains of precariousness factors, and depressive symptoms.
@@
 These patterns offer actionable insights for intervention.
-Addressing symptoms like sleep disturbance early may prevent escalation into broader depressive states or worsening precarity.
+Addressing symptoms like sleep disturbance early may prevent escalation into broader depressive states or worsening precariousness.
 Intervening on depressed mood, as a central node with potential feedback effects, could also disrupt the self-reinforcing dynamics between mental health and social isolation.
 
-To examine how such structures shape responsiveness to support, we developed a simplified two-variable nonlinear dynamical model capturing the feedback between depression and precarity, modulated by financial stress.
+To examine how such structures shape responsiveness to support, we developed a simplified two-variable nonlinear dynamical model capturing the feedback between depression and precariousness, modulated by financial stress.
 Calibrated against empirical summary statistics from the HELIUS dataset, the model allowed us to simulate how these dynamics evolve under varying levels of stress reduction.
@@
 In bistable systems, even moderate interventions may have little effect unless they are forceful enough to tip the system toward a lower-risk state.
-This underscores the importance of early or intensive support when targeting self-reinforcing cycles of depression and precarity.
+This underscores the importance of early or intensive support when targeting self-reinforcing cycles of depression and precariousness.
 Unistable systems, by contrast, are more responsive to incremental support, suggesting that sustained, moderate interventions may be sufficient in such contexts.
@@
 
-<!-- Among these symptoms, dep stands out as a distinct causal factor for precarity in both aggregated and disaggregated analyses, with CCI even suggesting a possible cyclic relationship between them.
+<!-- Among these symptoms, dep stands out as a distinct causal factor for precariousness in both aggregated and disaggregated analyses, with CCI even suggesting a possible cyclic relationship between them.
 This implies that depression and precariousness may reinforce one another, potentially leading to a worsening cycle over time.
-Within the symptom network, dep also exhibits a high degree of connectivity, but with mostly incoming arrows, suggesting that as symptom dynamics intensify, dep is quickly impacted by multiple factors—particularly P.soc (social precarity), further fueling the depression-precariousness cycle.
+Within the symptom network, dep also exhibits a high degree of connectivity, but with mostly incoming arrows, suggesting that as symptom dynamics intensify, dep is quickly impacted by multiple factors—particularly P.soc (social precariousness), further fueling the depression-precariousness cycle.
 -->
@@
 It maintains the strongest connectivity with symptoms, yet its directionality indicates that it is largely shaped by depressive symptoms rather than acting as a primary cause.
-Together, these patterns suggest a broader directional dynamic in which financial stress (S.fin) exacerbates depressive symptoms, which in turn feed into social precarity, perpetuating the cycle.
+Together, these patterns suggest a broader directional dynamic in which financial stress (S.fin) exacerbates depressive symptoms, which in turn feed into social precariousness, perpetuating the cycle.
 -->
@@
 
-<!-- Building on these insights, we developed a simplified computational model capturing the core mechanisms identified in the data: bidirectional feedback between depression and precarity, driven by external stress.
+<!-- Building on these insights, we developed a simplified computational model capturing the core mechanisms identified in the data: bidirectional feedback between depression and precariousness, driven by external stress.
 Calibrated against empirical summary statistics, the model allowed us to explore how different parameter configurations—particularly feedback strength and stochastic noise—modulate the system’s response to simulated interventions.
@@
 
-<!-- A key contribution of this work is to show that the structure of the depression–precarity system matters: not all forms of adversity respond equally to intervention, and this heterogeneity may stem from internal feedback dynamics as much as from external conditions.
+<!-- A key contribution of this work is to show that the structure of the depression–precariousness system matters: not all forms of adversity respond equally to intervention, and this heterogeneity may stem from internal feedback dynamics as much as from external conditions.
 Recognizing whether a given population operates within a bistable or unistable regime could inform the design and timing of mental health interventions—particularly in contexts of chronic socioeconomic stress.
@@
 
-<!-- Building on these findings, we developed a simplified two-variable nonlinear dynamical system to simulate the co-evolution of depression and social precarity under varying levels of financial stress.
+<!-- Building on these findings, we developed a simplified two-variable nonlinear dynamical system to simulate the co-evolution of depression and social precariousness under varying levels of financial stress.
 This model served as a conceptual abstraction of the inferred causal structure, enabling us to assess how the system’s internal configuration—particularly feedback strength, input coupling, and stochastic noise—modulates its behavior in response to intervention.
@@
 It remains possible that simpler, linear dynamics could account for the observed data just as well.
-Future work should further investigate whether bistability is a necessary feature of depression–precarity dynamics, and explore alternative functional forms—both linear and nonlinear—using richer data and stronger empirical constraints.
+Future work should further investigate whether bistability is a necessary feature of depression–precariousness dynamics, and explore alternative functional forms—both linear and nonlinear—using richer data and stronger empirical constraints.
 
@@
 These uncertainties highlight the need for further research, particularly with higher-resolution datasets, such as time-series data.
-Incorporating time-series data could capture temporal dynamics between precarity factors and depressive symptoms, providing time-specific insights and revealing how these relationships evolve over time.
+Incorporating time-series data could capture temporal dynamics between precariousness factors and depressive symptoms, providing time-specific insights and revealing how these relationships evolve over time.
 Methods like *PCMCI* [@runge2019detecting], *tsFCI* [@entner2010causal], and other time-series adaptations of causal discovery algorithms could offer a more robust framework for addressing temporal dependencies and refining the analysis.
@@
 <!-- A key limitation of our study lies in the structure and assumptions of the dynamical model.
-While the model is grounded in causal patterns identified from the data, it is intentionally simplified, focusing solely on two interacting variables—depression and social precarity—modulated by financial stress.
+While the model is grounded in causal patterns identified from the data, it is intentionally simplified, focusing solely on two interacting variables—depression and social precariousness—modulated by financial stress.
 This abstraction facilitates tractable simulation and exploration of systemic behaviors such as bistability.
@@
 While such nonlinearities are necessary to produce bistable dynamics, the presence of bistability in our model is a product of simulation rather than empirical verification.
-We do not currently have time-series data to confirm the existence of tipping points or regime shifts in real-world trajectories of depression and precarity.
+We do not currently have time-series data to confirm the existence of tipping points or regime shifts in real-world trajectories of depression and precariousness.
 It is also possible that a linear model would suffice to capture much of the observed behavior, and that the current structure overcomplicates model formulation.
@@
 Additionally, integrating qualitative approaches could enrich these findings, offering contextual insights into the quantitative relationships identified.
-Understanding individuals' lived experiences of precarity and depression could inform the development of more tailored and effective interventions.
+Understanding individuals' lived experiences of precariousness and depression could inform the development of more tailored and effective interventions.
 -->
@@
 <!-- 1. still impressive -- symptom directions aligns much well with the other empirical studies. -->
-<!-- 2. understood better which aspect of precarity is more important for depression and we can target depends on what level of intervention we are planning. -->
+<!-- 2. understood better which aspect of precariousness is more important for depression and we can target depends on what level of intervention we are planning. -->
 
 <!-- Implications -->
-Our results reveal that the effectiveness of interventions targeting depression and precarity is shaped not only by the strength of external stressors but also by the internal structure of the system.
+Our results reveal that the effectiveness of interventions targeting depression and precariousness is shaped not only by the strength of external stressors but also by the internal structure of the system.
 The interplay between feedback strength and noise determines whether a system responds gradually or only after crossing a tipping point—highlighting the need for intervention strategies attuned to these underlying dynamics.
@@
 These findings underscore the importance of considering internal feedback—not just external adversity—when designing interventions.
-Mental health systems are not always passively responsive to support; their structure can constrain how much impact any given policy has.
+Mental health systems are not always passively responsive to support; their structure can constrain how much impact any given policy has [@wagenaar2013public; @desavigny2009systems].
 By diagnosing these structural features—through tools like causal discovery and computational modeling—researchers and policymakers can better tailor interventions to the underlying mechanics of vulnerability.
@@
 Such extensions would help validate whether mechanisms proposed here, such as bistability, manifest in real-world trajectories.
-Ultimately, we hope this study encourages further integration of causal structure learning and system-level modeling to improve mental health outcomes under conditions of socioeconomic precarity.
+Ultimately, we hope this study encourages further integration of causal structure learning and system-level modeling to improve mental health outcomes under conditions of socioeconomic precariousness.
 
@@
 #| label: fig-sum-cci
-#| fig-cap: "Resulting graph of precarity factors and depression sum score using CCI and proportion of edge endpoint types."
+#| fig-cap: "Resulting graph of precariousness factors and depression sum score using CCI and proportion of edge endpoint types."
 #| fig-subcap: 
@@
 #| layout-nrow: 2
-#| fig-cap: "Resulting graph of precarity factors and individual depression symptoms using CCI and proportion of edge endpoint types."
+#| fig-cap: "Resulting graph of precariousness factors and individual depression symptoms using CCI and proportion of edge endpoint types."
 #| fig-subcap: 
@@
 #| label: fig-presum-cci
-#| fig-cap: "Resulting graph of precarity sum score and individual depression symptoms using CCI and proportion of edge endpoint types."
+#| fig-cap: "Resulting graph of precariousness sum score and individual depression symptoms using CCI and proportion of edge endpoint types."
 #| fig-subcap: 
@@
 #| label: fig-pc_sum
-#| fig-cap: "Resulting graphs of precarity factors and depression sum score using PC."
+#| fig-cap: "Resulting graphs of precariousness factors and depression sum score using PC."
 #| fig-subcap:
@@
 #| label: fig-pc_sym
-#| fig-cap: "Resulting graphs of precarity factors and individual depression symptoms using PC."
+#| fig-cap: "Resulting graphs of precariousness factors and individual depression symptoms using PC."
 #| fig-align: center
@@
 #| label: fig-pc_presum
-#| fig-cap: "Resulting graphs of precarity sum score and individual depression symptoms using PC."
+#| fig-cap: "Resulting graphs of precariousness sum score and individual depression symptoms using PC."
 #| fig-align: center
@@
 <!-- #| label: fig-presum-mat -->
-<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precarity sum score." -->
+<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precariousness sum score." -->
 <!-- #| fig-align: center -->

### 2025-06-10 draft_v1/draft_v1.tex
update_draft_Leoniefeedback

@@
 \hypersetup{
-  pdftitle={From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study},
+  pdftitle={From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study},
   pdfauthor={Kyuri Park; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
@@
 \title{From Causal Discovery to Intervention Simulation: Modeling
-Precarity and Depression in the HELIUS Study}
+Precariousness and Depression in the HELIUS Study}
 
@@
 
-\date{2025-05-20}
+\date{2025-06-10}
 \begin{document}
@@
 dynamical model simulating the co-evolution of depression and social
-precarity under varying levels of financial stress reduction. Simulation
-results revealed that the system's responsiveness depends not only on
-the intensity of external intervention but also on internal structural
-features such as feedback strength and noise. In particular, some
-configurations exhibited bistability, requiring interventions to surpass
-a critical threshold to trigger lasting improvement, while others
-responded more proportionally. Together, these findings emphasize that
-internal feedback dynamics---not just external adversity---can shape the
-impact of mental health interventions. More broadly, our study offers a
-scalable framework for integrating causal discovery and dynamical
-simulation to better understand and inform policy responses to complex
-psychosocial phenomena.
+precariousness under varying levels of financial stress reduction.
+Simulation results revealed that the system's responsiveness depends not
+only on the intensity of external intervention but also on internal
+structural features such as feedback strength and noise. In particular,
+some configurations exhibited bistability, requiring interventions to
+surpass a critical threshold to trigger lasting improvement, while
+others responded more proportionally. Together, these findings emphasize
+that internal feedback dynamics---not just external adversity---can
+shape the impact of mental health interventions. More broadly, our study
+offers a scalable framework for integrating causal discovery and
+dynamical simulation to better understand and inform policy responses to
+complex psychosocial phenomena.
 \end{abstract}
@@
 employment, finances, housing, social relations, and cultural belonging
-(Elsenburg et al., 2025). This broader view reveals how different forms
-of insecurity can co-occur and compound, forming an ecosystem of risk.
-However, despite growing evidence of association between precariousness
-and mental health, a fundamental question remains unresolved: \emph{How
-do these factors causally influence one another?} Most existing studies
-rely on cross-sectional correlations, leaving open questions about
-directionality and feedback. Without stronger causal insight, it remains
-difficult to identify priority targets for intervention or to anticipate
-unintended effects.
+(Elsenburg et al., 2025; McKee et al., 2017). This broader view reveals
+how different forms of insecurity can co-occur and compound, forming an
+ecosystem of risk. However, despite growing evidence of association
+between precariousness and mental health, a fundamental question remains
+unresolved: \emph{How do these factors causally influence one another?}
+Most existing studies rely on cross-sectional correlations, leaving open
+questions about directionality and feedback. Without stronger causal
+insight, it remains difficult to identify priority targets for
+intervention or to anticipate unintended effects.
 
@@
 responses to external intervention. Specifically, we ask: \emph{How does
-the internal configuration of the precarity--depression relationship
-affect the outcomes of stress-reducing interventions?} To answer this,
-we combine two approaches. First, we use cycle-capable causal discovery
-algorithms to infer directional relationships among domain-specific
-precariousness factors, and depressive symptoms. By examining both
-aggregated measures (depression sum score and a composite precarity
-index) and disaggregated variables (individual depressive symptoms and
-specific life-domain indicators), we capture broad patterns of
-relationship as well as fine-grained causal pathways. Second, we
+the internal configuration of the precariousness--depression
+relationship affect the outcomes of stress-reducing interventions?} To
+answer this, we combine two approaches. First, we use cycle-capable
+causal discovery algorithms to infer directional relationships among
+domain-specific precariousness factors, and depressive symptoms. By
+examining both aggregated measures (depression sum score and a composite
+precariousness index) and disaggregated variables (individual depressive
+symptoms and specific life-domain indicators), we capture broad patterns
+of relationship as well as fine-grained causal pathways. Second, we
 translate these insights into a computational dynamical model that
-simulates how depression and precarity co-evolve under varying levels of
-external stress. This model enables us to probe how system
+simulates how depression and precariousness co-evolve under varying
+levels of external stress. This model enables us to probe how system
 properties---such as feedback strength and stochastic noise---govern
@@
 integrating causal discovery with dynamic simulation, we aim to
-demonstrate how the internal structure of the depression--precarity
+demonstrate how the internal structure of the depression--precariousness
 system shapes its susceptibility to change---and, by extension, how
@@
 Stronks et al., 2013). It captures the diverse population of the city of
-Amsterdam by including the six main ethnic groups and which provides
+Amsterdam by including the six main ethnic groups and provides
 comprehensive health and lifestyle data, including depressive symptoms
@@
 
-After preprocessing, the HELIUS dataset includes 21,628 samples. In
+After preprocessing, the HELIUS dataset includes 21,628 individuals. In
 addition to the five identified precariousness factors, we computed an
-overall precarity score by aggregating their values. Depression is
-represented using the PHQ-9 (Kroenke et al., 2001), both as a total sum
-score and as individual symptom scores. In the subsequent causal
-discovery analysis, we explore the relationship between depression and
-precarity using both aggregated scores and their disaggregated
-components. See Figure~\ref{fig-dist} for the overall distributions of
-the variables used in the analysis.
+overall precariousness score by by summing these five factor scores,
+thereby capturing a composite measure of accumulated precariousness
+across domains. Depression is represented using the PHQ-9 (Kroenke et
+al., 2001), both as a total sum score and as individual symptom scores.
+In the subsequent causal discovery analysis, we explore the relationship
+between depression and precariousness using both aggregated scores and
+their disaggregated components. See Figure~\ref{fig-dist} for the
+overall distributions of the variables used in the analysis.
 
@@
 \caption{\label{fig-dist}Distributions of variables with density
-overlay.\\
-\emph{P.emp} = employment precariousness; \emph{P.hou} = housing
-precariousness; \emph{P.soc} = social precariousness; \emph{S.fin} =
-recent financial stressors; \emph{S.rel} = recent relational stressors;
-\emph{Precarity} = overall precarity; \emph{PHQsum} = PHQ-9 sum score;
-\emph{anh} = anhedonia; \emph{app} = appetite; \emph{con} =
-concentration; \emph{dep} = depressed mood; \emph{ene} = energy;
-\emph{glt} = guilty; \emph{mot} = motor; \emph{sui} = suicidal}
+overlay. \emph{This shows the distributions of the variables used in the
+causal analysis, plotted as histograms with overlaid density estimates.
+The x-axis represents the value of each variable after preprocessing.
+The y-axis shows the corresponding estimated probability density.}
+\textbf{\emph{P.emp}} = \emph{employment precariousness};
+\textbf{\emph{P.hou}} = \emph{housing precariousness};
+\textbf{\emph{P.soc}} = \emph{social precariousness};
+\textbf{\emph{S.fin}} = \emph{recent financial stressors};
+\textbf{\emph{S.rel}} = \emph{recent relational stressors};
+\textbf{\emph{precariousness}} = \emph{overall precariousness};
+\textbf{\emph{PHQsum}} = \emph{PHQ-9 sum score}; \textbf{\emph{anh}} =
+\emph{anhedonia}; \textbf{\emph{app}} = \emph{appetite};
+\textbf{\emph{con}} = \emph{concentration}; \textbf{\emph{dep}} =
+\emph{depressed mood}; \textbf{\emph{ene}} = \emph{energy};
+\textbf{\emph{glt}} = \emph{guilty}; \textbf{\emph{mot}} = \emph{motor};
+\textbf{\emph{slp}} = \emph{sleep}; \textbf{\emph{sui}} =
+\emph{suicidal}.}
 
@@
 The aggregate analysis reduces dimensionality and yields interpretable
-summaries of how domains of precarity relate to overall depression
+summaries of how domains of precariousness relate to overall depression
 severity. However, it may obscure finer-grained effects. The
@@
 perspectives by examining how individual symptoms respond to cumulative
-precarity across domains.
+precariousness across domains.
 
@@
 
-\subsection{Modeling Dynamics between Precarity and
-Depression}\label{modeling-dynamics-between-precarity-and-depression}
+\subsection{Modeling Dynamics between precariousness and
+Depression}\label{modeling-dynamics-between-precariousness-and-depression}
 
@@
 
-NSGA-II is implemented via the \texttt{nsga2R} package (Franz \&
-Nakamura, 2015), with a population of 500 evolved over 300 generations
-using standard hyperparameters (crossover probability = 0.9, mutation
-probability = 0.25, and tournament size = 2). This process yielded a
-diverse ensemble of Pareto-efficient solutions balancing trade-offs
-between objectives. From the resulting Pareto front
-(Figure~\ref{fig-pareto}), we selected the top 30 parameter sets with
-the lowest combined error to use in the subsequent simulations assessing
-intervention effects.
+To explore optimal parameter configurations, we used the NSGA-II
+algorithm as implemented in the \texttt{nsga2R} package (Franz \&
+Nakamura, 2015). This algorithm is inspired by natural selection and
+evolves a population of candidate solutions over multiple generations to
+balance competing objectives. We began with a population of 500 randomly
+generated parameter sets and allowed them to evolve over 300
+generations. During this process, solutions were selected, combined, and
+mutated using standard settings: a crossover probability of 0.9 (which
+controls how often solutions are combined), a mutation probability of
+0.25 (introducing random variation), and a tournament size of 2 (used to
+pick the best candidates for reproduction). The result is a Pareto
+front---a set of solutions that are considered optimal in the sense that
+no single objective can be improved without worsening another. From this
+diverse set of trade-off solutions (see Figure~\ref{fig-pareto}), we
+selected the 30 parameter sets with the lowest overall error across
+objectives. These top-performing configurations were then used in the
+simulation phase to evaluate the potential effects of different
+interventions.
 
@@
 
-\subsection{Causal Structure Linking Precarity and
-Depression}\label{causal-structure-linking-precarity-and-depression}
+\subsection{Causal Structure Linking Precariousness and
+Depression}\label{causal-structure-linking-precariousness-and-depression}
 
@@
 
-The sum score graphs provide a high-level summary of how precarity
+The sum score graphs provide a high-level summary of how precariousness
 factors collectively influence overall depression severity, focusing on
@@
 algorithm. As described in Section~\ref{sec-analysis}, dashed edges
-indicate ties, with the exact edge endpoint where the tie occurs is
-represented by a bold horizontal bar.
+indicate ties, and the edge endpoint at which the tie occurs is marked
+by a bold horizontal bar.
 
 The causal graph in panel (a) highlights key pathways in the
-relationships between precarity factors and depression (\emph{PHQsum}).
-Employment precarity (\emph{P.emp}) and social precarity (\emph{P.soc})
-are not identified as causes of depression, whereas financial stress
-(\emph{S.fin}) appears to play a causal role. While \emph{P.emp} and
-\emph{P.soc} are generally not recognized as causes of other
-precariousness factors, \emph{S.fin} emerges as a potential cause, as
-indicated by its circle edge endpoint. Supporting this, the proportion
-matrix plot in panel (b) shows that \emph{S.fin} has some probability of
-causally influencing either \emph{P.emp} or \emph{P.soc.}
+relationships between precariousness factors and depression
+(\emph{PHQsum}). Employment precariousness (\emph{P.emp}) and social
+precariousness (\emph{P.soc}) are not identified as causes of
+depression, whereas financial stress (\emph{S.fin}) appears to play a
+causal role. While \emph{P.emp} and \emph{P.soc} are generally not
+recognized as causes of other precariousness factors, \emph{S.fin}
+emerges as a potential cause, as indicated by its circle edge endpoint.
+Supporting this, the proportion matrix plot in panel (b) shows that
+\emph{S.fin} has some probability of causally influencing either
+\emph{P.emp} or \emph{P.soc.}
 
@@
 relationship is either absent, ambiguous, or potentially bidirectional.
-Meanwhile, housing precarity (\emph{P.hou}) is not directly causally
-related to depression but is linked to \emph{P.emp} and \emph{S.fin}.
-While \emph{P.emp} and \emph{S.fin} are identified as non-causes of
-\emph{P.hou}, it remains unclear whether \emph{P.hou} causally
-influences \emph{P.emp} or if their relationship is mediated by an
-unobserved confounder.
+Meanwhile, housing precariousness (\emph{P.hou}) is not directly
+causally related to depression but is linked to \emph{P.emp} and
+\emph{S.fin}. While \emph{P.emp} and \emph{S.fin} are identified as
+non-causes of \emph{P.hou}, it remains unclear whether \emph{P.hou}
+causally influences \emph{P.emp} or if their relationship is mediated by
+an unobserved confounder.
 
@@
 
-\caption{\label{fig-sum}Resulting graph of precarity factors and
+\caption{\label{fig-sum}Resulting graph of precariousness factors and
 depression sum score using FCI and proportion of edge endpoint types.}
@@
 particularly \emph{dep} and \emph{glt}, may contribute to worsening
-social precarity rather than the other way around. Additionally,
+social precariousness rather than the other way around. Additionally,
 \emph{anh} also shows some probability of causally influencing
-\emph{P.soc}, reinforcing the idea that social precarity is more often a
-consequence rather than a driver of depressive symptoms.
+\emph{P.soc}, reinforcing the idea that social precariousness is more
+often a consequence rather than a driver of depressive symptoms.
 
@@
 initiating symptom, particularly sensitive to relational stress and
-employment precarity. On the other hand, \emph{P.soc} serves as a key
-bridge between the depression and precariousness subsystems, as
-depressive symptoms appear to feed back into social precarity,
+employment precariousness. On the other hand, \emph{P.soc} serves as a
+key bridge between the depression and precariousness subsystems, as
+depressive symptoms appear to feed back into social precariousness,
 reinforcing a self-sustaining dynamic between depression and precarious
@@
 
-\caption{\label{fig-sym}Resulting graph of precarity factors and
+\caption{\label{fig-sym}Resulting graph of precariousness factors and
 individual depression symptoms using FCI and proportion of edge endpoint
@@
 
-\subsubsection{Precarity as sum score}\label{precarity-as-sum-score}
+\subsubsection{Precariousness as sum
+score}\label{precariousness-as-sum-score}
 
@@
 consequence of depression than its source. Some symptoms, particularly
-\emph{glt}, \emph{slp}, and \emph{dep}, show greater sensitivity to
-precarious conditions, while also playing a central role in symptom
-dynamics. Among them, \emph{slp} and \emph{glt} emerge as influential
+\emph{glt}, \emph{slp}, and \emph{dep}, appear to influence precarious
+conditions while also occupying central roles within the symptom
+network. Among them, \emph{slp} and \emph{glt} emerge as influential
 symptoms within the depression network, whereas \emph{con} and
@@
 
-\caption{\label{fig-presum}Resulting graph of precarity sum score and
-individual depression symptoms using FCI and proportion of edge endpoint
-types.}
+\caption{\label{fig-presum}Resulting graph of precariousness sum score
+and individual depression symptoms using FCI and proportion of edge
+endpoint types.}
 
@@
 
-\subsection{Simulating the Dynamics Between Precarity and Depression
-Under
-Intervention}\label{simulating-the-dynamics-between-precarity-and-depression-under-intervention}
+\subsection{Simulating the Dynamics Between Precariousness and
+Depression Under
+Intervention}\label{simulating-the-dynamics-between-precariousness-and-depression-under-intervention}
 
@@
 ensemble. The top panel of Figure~\ref{fig-intervention} shows mean
-levels of depression (\(D\)) and precarity (\(P\)) after simulation,
-with each line representing one parameter set. Lines are color-coded by
-combined feedback and input strength, calculated as the product
-\(\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2\).
+levels of depression (\(D\)) and precariousness (\(P\)) after
+simulation, with each line representing one parameter set. Lines are
+color-coded by combined feedback and input strength, calculated as the
+product \(\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2\).
 
@@
 may enhance their effectiveness and sustainability, particularly in
-addressing the compounding effects of socioeconomic precarity on mental
-health.
+addressing the compounding effects of socioeconomic precariousness on
+mental health.
 
@@
 Intervention Sizes. \emph{Mean trajectories of depression (\(D\)) and
-precarity (\(P\)) across varying intervention strengths, simulated over
-the top 30 parameter sets. Each line represents one parameter set and is
-colored by its combined feedback and input strength.} \textbf{Bottom}:
-Smoothed Density Heatmap of Final \(D\) and \(P\) for Two Parameter
-Sets. \emph{One exhibits unistability (top two panels) and the other
-exhibits bistability (bottom two panels). Bistable dynamics reveal a
-sharp transition between distinct outcome regimes, while unistable
-dynamics show gradual, continuous change across the intervention
-gradient.}}
+precariousness (\(P\)) across varying intervention strengths, simulated
+over the top 30 parameter sets. Each line represents one parameter set
+and is colored by its combined feedback and input strength.}
+\textbf{Bottom}: Smoothed Density Heatmap of Final \(D\) and \(P\) for
+Two Parameter Sets. \emph{One exhibits unistability (top two panels) and
+the other exhibits bistability (bottom two panels). Bistable dynamics
+reveal a sharp transition between distinct outcome regimes, while
+unistable dynamics show gradual, continuous change across the
+intervention gradient.}}
 
@@
 This study combined constraint-based causal discovery and dynamical
-modeling to investigate how depression and socioeconomic precarity
+modeling to investigate how depression and socioeconomic precariousness
 interact---and how their internal structure modulates responsiveness to
@@
 Addressing symptoms like sleep disturbance early may prevent escalation
-into broader depressive states or worsening precarity. Intervening on
-depressed mood, as a central node with potential feedback effects, could
-also disrupt the self-reinforcing dynamics between mental health and
-social isolation.
+into broader depressive states or worsening precariousness. Intervening
+on depressed mood, as a central node with potential feedback effects,
+could also disrupt the self-reinforcing dynamics between mental health
+and social isolation.
 
@@
 developed a simplified two-variable nonlinear dynamical model capturing
-the feedback between depression and precarity, modulated by financial
-stress. Calibrated against empirical summary statistics from the HELIUS
-dataset, the model allowed us to simulate how these dynamics evolve
-under varying levels of stress reduction. The simulations revealed that
-intervention effectiveness depends not just on the magnitude of external
-stressors, but on the internal configuration of the system itself. In
-bistable regimes---characterized by strong feedback and low noise---the
-system could remain stuck in a high-risk state unless an intervention
-crosses a critical threshold, tipping it into a healthier equilibrium.
-By contrast, unistable systems---driven by weaker coupling or greater
-noise---responded more gradually and proportionally to increasing
-support. These dynamics carry practical policy implications. In bistable
-systems, even moderate interventions may have little effect unless they
-are forceful enough to tip the system toward a lower-risk state. This
-underscores the importance of early or intensive support when targeting
-self-reinforcing cycles of depression and precarity. Unistable systems,
-by contrast, are more responsive to incremental support, suggesting that
-sustained, moderate interventions may be sufficient in such contexts.
+the feedback between depression and precariousness, modulated by
+financial stress. Calibrated against empirical summary statistics from
+the HELIUS dataset, the model allowed us to simulate how these dynamics
+evolve under varying levels of stress reduction. The simulations
+revealed that intervention effectiveness depends not just on the
+magnitude of external stressors, but on the internal configuration of
+the system itself. In bistable regimes---characterized by strong
+feedback and low noise---the system could remain stuck in a high-risk
+state unless an intervention crosses a critical threshold, tipping it
+into a healthier equilibrium. By contrast, unistable systems---driven by
+weaker coupling or greater noise---responded more gradually and
+proportionally to increasing support. These dynamics carry practical
+policy implications. In bistable systems, even moderate interventions
+may have little effect unless they are forceful enough to tip the system
+toward a lower-risk state. This underscores the importance of early or
+intensive support when targeting self-reinforcing cycles of depression
+and precariousness. Unistable systems, by contrast, are more responsive
+to incremental support, suggesting that sustained, moderate
+interventions may be sufficient in such contexts.
 
@@
 investigate whether bistability is a necessary feature of
-depression--precarity dynamics, and explore alternative functional
+depression--precariousness dynamics, and explore alternative functional
 forms---both linear and nonlinear---using richer data and stronger
@@
 Our results reveal that the effectiveness of interventions targeting
-depression and precarity is shaped not only by the strength of external
-stressors but also by the internal structure of the system. The
+depression and precariousness is shaped not only by the strength of
+external stressors but also by the internal structure of the system. The
 interplay between feedback strength and noise determines whether a
@@
 Mental health systems are not always passively responsive to support;
-their structure can constrain how much impact any given policy has. By
-diagnosing these structural features---through tools like causal
-discovery and computational modeling---researchers and policymakers can
-better tailor interventions to the underlying mechanics of
-vulnerability.
+their structure can constrain how much impact any given policy has
+(Savigny \& Adam, 2009; Wagenaar \& Burris, 2013). By diagnosing these
+structural features---through tools like causal discovery and
+computational modeling---researchers and policymakers can better tailor
+interventions to the underlying mechanics of vulnerability.
 
@@
 and system-level modeling to improve mental health outcomes under
-conditions of socioeconomic precarity.
+conditions of socioeconomic precariousness.
 
@@
 
+\bibitem[\citeproctext]{ref-mckee2017living}
+McKee, M., Reeves, A., Clair, A., \& Stuckler, D. (2017). Living on the
+edge: Precariousness and why it matters for health. \emph{Archives of
+Public Health}, \emph{75}, 1--10.
+
 \bibitem[\citeproctext]{ref-mooij2020constraint}
@@
 
+\bibitem[\citeproctext]{ref-desavigny2009systems}
+Savigny, D. de, \& Adam, T. (Eds.). (2009). \emph{Systems thinking for
+health systems strengthening}. Alliance for Health Policy; Systems
+Research, World Health Organization.
+
 \bibitem[\citeproctext]{ref-sethuraman2023nodags}
@@
 \emph{The Lancet Psychiatry}, \emph{8}(11), 991--1000.
+
+\bibitem[\citeproctext]{ref-wagenaar2013public}
+Wagenaar, A. C., \& Burris, S. (2013). \emph{Public health law research:
+Theory and methods}. John Wiley \& Sons.
 
@@
 
-\caption{\label{fig-sum-cci}Resulting graph of precarity factors and
-depression sum score using CCI and proportion of edge endpoint types.}
+\caption{\label{fig-sum-cci}Resulting graph of precariousness factors
+and depression sum score using CCI and proportion of edge endpoint
+types.}
 
@@
 
-\caption{\label{fig-sym-cci}Resulting graph of precarity factors and
-individual depression symptoms using CCI and proportion of edge endpoint
-types.}
+\caption{\label{fig-sym-cci}Resulting graph of precariousness factors
+and individual depression symptoms using CCI and proportion of edge
+endpoint types.}
 
@@
 
-\caption{\label{fig-presum-cci}Resulting graph of precarity sum score
-and individual depression symptoms using CCI and proportion of edge
-endpoint types.}
+\caption{\label{fig-presum-cci}Resulting graph of precariousness sum
+score and individual depression symptoms using CCI and proportion of
+edge endpoint types.}
 
@@
 
-\caption{\label{fig-pc_sum}Resulting graphs of precarity factors and
-depression sum score using PC.}
+\caption{\label{fig-pc_sum}Resulting graphs of precariousness factors
+and depression sum score using PC.}
 
@@
 
-\caption{\label{fig-pc_sym}Resulting graphs of precarity factors and
-individual depression symptoms using PC.}
+\caption{\label{fig-pc_sym}Resulting graphs of precariousness factors
+and individual depression symptoms using PC.}
 
@@
 
-\caption{\label{fig-pc_presum}Resulting graphs of precarity sum score
-and individual depression symptoms using PC.}
+\caption{\label{fig-pc_presum}Resulting graphs of precariousness sum
+score and individual depression symptoms using PC.}
 

### 2025-06-24 draft_v1/draft_v2.tex
add linear model

--- 2025-06-10 draft_v1/draft_v1.tex
@@
 
-\date{2025-06-10}
+\date{2025-06-24}
 \begin{document}
@@
 
-\includegraphics[width=0.9\textwidth,height=\textheight]{draft_v1_files/figure-pdf/fig-dist-1.pdf}
+\includegraphics[width=0.9\textwidth,height=\textheight]{draft_v2_files/figure-pdf/fig-dist-1.pdf}
 
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-12-1.pdf}
+\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-12-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-13-1.pdf}
+\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-13-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-14-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-14-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-15-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-15-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-16-1.pdf}
+\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-16-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-17-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-17-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-18-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-18-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-1.pdf}
+\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-19-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-2.pdf}
+\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-19-2.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-3.pdf}
+\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-19-3.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-19-4.pdf}
+\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-19-4.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v1_files/figure-pdf/unnamed-chunk-20-1.pdf}
+\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-20-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-21-1.pdf}
+\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-21-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v1_files/figure-pdf/unnamed-chunk-22-1.pdf}
+\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-22-1.pdf}
 \end{center}

### 2025-06-24 code/intervention_simulation/04_linear_model.R

@@ -1,0 +1,188 @@
+# Load data & libraries
+source("Precarity_HELIUS/code/utils/preprocess_dat.R") 
+
+set.seed(123) # set seed
+
+# ========== 0. Data & Covariance Setup ==========
+init_covars <- function(clust_data) {
+  list(
+    varD  = 1,
+    varP  = var(clust_data$P.soc),
+    varS  = var(clust_data$S.fin),
+    covDP = cov(clust_data$PHQsum, clust_data$P.soc),
+    covDS = cov(clust_data$PHQsum, clust_data$S.fin),
+    covPS = cov(clust_data$P.soc,   clust_data$S.fin)
+  )
+}
+covs <- init_covars(clust_data)
+
+# ========== 1. Covariance → α₍dp₎ & Parameter Estimation ==========
+get_model_params <- function(alpha_dp, covs) {
+  with(covs, {
+    αds <- (-alpha_dp * covPS + covDS) / varS
+    αps <- (-covDS^2 * covPS + covDS * (alpha_dp * covPS^2 - alpha_dp * varP * varS + 2*covDP*varS) - covPS * varD * varS) /
+      (varS*(covDS^2 - varD*varS))
+    αpd <- (-alpha_dp * covPS^2 + alpha_dp * varP * varS - 2*covDP*varS + 2*covDS*covPS) /
+      (covDS^2 - varD*varS)
+    σd2 <- ((alpha_dp*covPS - covDS)^2) / varS
+    σp2 <- ((alpha_dp*varS   - covDP)^2) / varS
+    list(alpha_ds = αds, alpha_ps = αps, alpha_pd = αpd,
+         sigma_d = sqrt(σd2), sigma_p = sqrt(σp2))
+  })
+}
+
+estimate_alpha_dp <- function(covs) {
+  loss <- function(adp) {
+    p <- get_model_params(adp, covs)
+    ests <- c(p$alpha_ds * covs$varS + adp * covs$covPS,
+              p$alpha_ps * covs$varS + p$alpha_pd * covs$covDS,
+              adp * covs$varS)
+    sum((ests - c(covs$covDS, covs$covPS, covs$covDP))^2)
+  }
+  optimize(loss, interval = c(0, 1))$minimum
+}
+αdp_est <- estimate_alpha_dp(covs)
+
+# ========== 2. Parameter Tradeoff Plot ==========
+αdp_grid <- seq(0, 0.698379, by = 0.05)
+
+param_grid <- map_dfr(αdp_grid, ~ {
[…]

### 2025-06-24 code/intervention_simulation/04_linear_model.R
Update 04_linear_model.R

@@ -1,3 +1,18 @@
+## =============================================================================
+## Linear Feedback SDE Model Analysis: Covariance-Based Calibration & Trajectory Dynamics
+## =============================================================================
+#' This script implements a stochastic linear feedback model of depression (D) and
+#' precarity (P), influenced by stressor (S), using empirical covariances for calibration.
+#'
+#' 0. Load and calculate empirical covariance structure from ‘clust_data’.
+#' 1. Estimate α₍dp₎ by minimizing discrepancy between modeled and observed covariances.
+#' 2. Analyze parameter trade-offs across a grid of α₍dp₎ values with annotated plots.
+#' 3. Run snapshot simulations at key time points and visualize mean responses to interventions.
+#' 4. Simulate full trajectories under “intervention ON/OFF” and examine recovery dynamics
+#'     with both individual paths and averaged curves.
+#' 5. Assemble all visuals into a combined multi-panel figure.
+## =============================================================================
+
 # Load data & libraries
 source("Precarity_HELIUS/code/utils/preprocess_dat.R") 
 

### 2025-06-27 code/intervention_simulation/04_linear_model.R
clean up linearmod_intervention

@@ -1,203 +1,450 @@
 ## =============================================================================
-## Linear Feedback SDE Model Analysis: Covariance-Based Calibration & Trajectory Dynamics
+## Linear Feedback SDE Model Analysis
 ## =============================================================================
-#' This script implements a stochastic linear feedback model of depression (D) and
-#' precarity (P), influenced by stressor (S), using empirical covariances for calibration.
+#' This script implements a stochastic linear feedback model of Depression (D)
+#' and Precarity (P), influenced by Stressor (S), calibrated using empirical covariances.
 #'
-#' 0. Load and calculate empirical covariance structure from ‘clust_data’.
-#' 1. Estimate α₍dp₎ by minimizing discrepancy between modeled and observed covariances.
-#' 2. Analyze parameter trade-offs across a grid of α₍dp₎ values with annotated plots.
-#' 3. Run snapshot simulations at key time points and visualize mean responses to interventions.
-#' 4. Simulate full trajectories under “intervention ON/OFF” and examine recovery dynamics
-#'     with both individual paths and averaged curves.
-#' 5. Assemble all visuals into a combined multi-panel figure.
+#' **Workflow:**
+#'  0. Load data & compute covariance structure  
+#'  1. Calibrate model by estimating α_dp (effect of P → D) from covariances  
+#'  2. Visualize parameter trade-offs across α_dp grid  
+#'  3. Run snapshot simulations at selected times → intervention effects  
+#'  4. Simulate full trajectories with intervention ON/OFF → recovery dynamics  
+#'  5. Combine plots using patchwork for publication-ready output
 ## =============================================================================
 
-# Load data & libraries
-source("Precarity_HELIUS/code/utils/preprocess_dat.R") 
+# Load libraries & data
+library(dplyr)
+library(tidyr)
+library(purrr)
+library(ggplot2)
+library(furrr)
+library(scales)
[…]

### 2025-06-27 draft_v1/draft_v2.qmd
update draft_v2

--- 2025-06-10 draft_v1/draft_v1.qmd
@@
 abstract: |
-  \noindent This study investigates the dynamic relationship between precarious life conditions and depression by combining constraint-based causal discovery with computational modeling.
-Using cross-sectional data from the HELIUS study, we applied cycle-capable causal discovery algorithms to uncover directional relationships between domain-specific precariousness indicators, and depressive symptoms.
-Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—identified financial stress as a robust driver of depression, and highlighted key symptoms, such as sleep disturbance and depressed mood, as likely initiators or reinforcers of downstream precariousness.
-  To explore how these structural insights shape intervention effectiveness, we developed a simplified nonlinear dynamical model simulating the co-evolution of depression and social precariousness under varying levels of financial stress reduction.
-Simulation results revealed that the system's responsiveness depends not only on the intensity of external intervention but also on internal structural features such as feedback strength and noise.
-In particular, some configurations exhibited bistability, requiring interventions to surpass a critical threshold to trigger lasting improvement, while others responded more proportionally.
-  Together, these findings emphasize that internal feedback dynamics—not just external adversity—can shape the impact of mental health interventions.
-More broadly, our study offers a scalable framework for integrating causal discovery and dynamical simulation to better understand and inform policy responses to complex psychosocial phenomena.
+  \noindent This study investigates how the internal structure of depression–precariousness dynamics shapes responsiveness to external support.
+Using cross-sectional data from the HELIUS study, we first apply cycle-capable causal discovery algorithms to estimate directional dependencies between financial stress, domain-specific precariousness, and depressive symptoms.
+Our multi-resolution analysis—spanning both aggregate and symptom-level constructs—highlights financial stress as a plausible upstream driver of depression, with symptoms such as sleep disturbance and depressed mood emerging as initiators of broader social vulnerability.
+  To translate these structural insights into dynamic predictions, we construct a family of analytically calibrated linear stochastic models.
+Simulations reveal that variations in internal structure can substantially alter system responses to financial stress reduction.
+Some models show rapid but transient improvements in depression, while others adjust more slowly yet sustain gains over time.
+Precariousness tends to respond more immediately across all configurations, though its trajectory also shifts subtly with structural variation.
+  Together, these findings demonstrate that intervention outcomes depend not only on the magnitude of external input but also on latent properties of the system.
+More broadly, this study illustrates how combining causal discovery with dynamical modeling supports principled, simulation-based exploration of intervention scenarios—grounded in data, yet able to probe counterfactual structure.
+This approach offers a scalable framework for diagnosing hidden system constraints and informing targeted public health strategies in complex psychosocial domains.
 
@@
 Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S$ acting as a shared input.
-These dynamics are formalized as a pair of nonlinear stochastic differential equations (SDEs):
+These dynamics are formalized as a pair of linear stochastic differential equations (SDEs):
 
@@
 \begin{aligned}
-dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\
-dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2 
+dP &= \lambda_P \left(\alpha_{PS} S + \alpha_{PD} D - P \right) dt + \sigma_{P} dW_{P} \\
+dD &= \lambda_D \left(\alpha_{DS} S + \alpha_{DP} P - D \right) dt + \sigma_{D} dW_{D} 
 \end{aligned}
@@
 Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress.
-The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which variable adjusts toward its input-driven target state.
-The coefficients $\alpha_1, \alpha_2$ and $\beta_1, \beta_2$ represent the strengths of the causal influences between variables.
-Stochastic fluctuations are introduced through the terms $\sigma_1 dW_1$ and $\sigma_2 dW_2$, where $dW_1$ and $dW_2$ are independent Wiener processes.
-The system uses the hyperbolic tangent function ($tanh$) to bound the influence of inputs, ensuring smooth, saturating dynamics and preventing runaway feedback loops.
-Importantly, the nonlinearity introduced by the $tanh$ function also allows the system to exhibit bistability, supporting the existence of two stable states: one corresponding to low depression and precariousness, and the other to high depression and precariousness.
-
-<!-- Building on the causal structure derived from the result of causal discovery, we implement a high-level simulation model to explore the potential impact of targeted interventions on precariousness.
-This serves as a proof-of-concept for translating causal relationships into dynamic systems that can generate actionable insights.
--->
-
-<!-- The structure of the simulation model, illustrated in @fig-concept, captures the dynamic interplay between depression ($PHQsum$) and social precariousness ($P.soc$), both of which are influenced by an external financial stressor ($S.fin$).
-Importantly, the model incorporates cyclic dependencies between depression and precariousness, reflecting their reciprocal influence over time.
--->
-<!-- This system is formalized as a set of non-linear stochastic differential equations (SDEs): -->
-
-<!-- $$ -->
-<!-- \begin{aligned} -->
-<!-- dP &= \lambda_P \left( \tanh(\alpha_1 S + \alpha_2 D) - P \right) dt + \sigma_1 dW_1 \\ -->
-<!-- dD &= \lambda_D \left( \tanh(\beta_1 S + \beta_2 P) - D \right) dt + \sigma_2 dW_2  -->
-<!-- \end{aligned} -->
-<!-- $$ -->
-
-<!-- Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress.
-The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which variable adjusts toward its input-driven target state.
-The coefficients $\alpha_1, \alpha_2$ and $\beta_1, \beta_2$ represent the strengths of the causal influences between variables.
--->
-<!-- Stochastic fluctuations are introduced through the terms $\sigma_1 dW_1$ and $\sigma_2 dW_2$, where $dW_1$ and $dW_2$ are independent Wiener processes.
-The system uses the hyperbolic tangent function ($tanh$) to bound the influence of inputs, ensuring smooth, saturating dynamics and preventing runaway feedback loops.
-Importantly, the nonlinearity introduced by the $tanh$ function also allows the system to exhibit bistability, supporting the existence of two stable states: one corresponding to low depression and precariousness, and the other to high depression and precariousness.
--->
+The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which each variable adjusts toward its input-driven target state.
+The coefficients $\alpha_{PS}$ and $\alpha_{DS}$ quantify the direct effects of the stressor on $P$ and $D$, while $\alpha_{DP}$ and $\alpha_{PD}$ represent the cross-influences between depression and precariousness.
+Stochastic fluctuations are introduced through the noise terms $\sigma_P, dW_P$ and $\sigma_D, dW_D$, with $dW_P$ and $dW_D$ representing independent Wiener processes.
+This linear structure offers a transparent and analytically tractable framework for examining how variations in external financial stress influence the population-level dynamics of depression and social precariousness.
+Crucially, the model's simplicity enables direct calibration from empirical data: all parameters can be derived analytically from observed variances and covariances in the HELIUS cohort.
+The following  [@sec-calibration] details the procedure used to perform this calibration.
 
@@
 
-### Model Calibration
-
-To identify realistic parameter settings for the simulation model, we calibrate six free parameters—$\alpha_1$, $\alpha_2$, $\beta_1$, $\beta_2$, and noise amplitudes $\sigma_1$, $\sigma_2$—using the *Non-dominated Sorting Genetic Algorithm II* (NSGA-II) [@deb2002fast], a multi-objective evolutionary optimization method well-suited for noisy and high-dimensional search spaces.
-The goal is to match simulated system behavior to empirical summary statistics derived from the HELIUS dataset.
-Specifically, we aim to match the observed means and variances of depression ($D$) and social precariousness ($P$), along with all three partial correlations among $D$, $P$, and financial stress ($S$): $\rho_{DP.S}$, $\rho_{DS.P}$, and $\rho_{PS.D}$.
-
-Given the stochastic nature of the model, each candidate parameter set is evaluated by simulating 1,000 virtual individuals, each assigned a fixed $S$ value sampled from the empirical distribution.
-From these simulations, we compute population-level summary statistics based on the final states.
-The fitness of each parameter set is measured by how closely its summary statistics matched the empirical data, using the total absolute deviation across all objectives.
-
-<!-- NSGA-II is implemented via the `nsga2R` package [@nsga2R], with a population of 500 evolved over 300 generations using standard hyperparameters (crossover probability = 0.9, mutation probability = 0.25, and tournament size = 2).
-This process yielded a diverse ensemble of Pareto-efficient solutions balancing trade-offs between objectives.
-From the resulting Pareto front (@fig-pareto), we selected the top 30 parameter sets with the lowest combined error to use in the subsequent simulations assessing intervention effects.
--->
-
-To explore optimal parameter configurations, we used the NSGA-II algorithm as implemented in the `nsga2R` package [@nsga2R].
-This algorithm is inspired by natural selection and evolves a population of candidate solutions over multiple generations to balance competing objectives.
-We began with a population of 500 randomly generated parameter sets and allowed them to evolve over 300 generations.
-During this process, solutions were selected, combined, and mutated using standard settings: a crossover probability of 0.9 (which controls how often solutions are combined), a mutation probability of 0.25 (introducing random variation), and a tournament size of 2 (used to pick the best candidates for reproduction).
-The result is a Pareto front—a set of solutions that are considered optimal in the sense that no single objective can be improved without worsening another.
-From this diverse set of trade-off solutions (see @fig-pareto), we selected the 30 parameter sets with the lowest overall error across objectives.
-These top-performing configurations were then used in the simulation phase to evaluate the potential effects of different interventions.
+### Model Calibration {#sec-calibration}
+
+In our linear model, all parameters are derived analytically to reproduce the empirical second-order moments observed in the HELIUS dataset.
+The system treats depression ($D$) and social precariousness ($P$) as jointly evolving under the influence of a shared external stressor ($S$), forming a coupled system of Ornstein–Uhlenbeck processes.
+Under the assumption of stationarity, the steady-state covariances between these variables admit closed-form solutions for the system’s structural parameters.
+
+We fix the adjustment rates $\lambda_D = \lambda_P = 1$, which normalizes the time scale given the cross-sectional nature of the HELIUS data.
+To resolve the remaining degrees of freedom, we treat the feedback from precariousness to depression ($\alpha_{DP}$) as a free parameter, and solve algebraically for all other coefficients — including the direct effects of stress ($\alpha_{DS}, \alpha_{PS}$), the reverse feedback from depression to precariousness ($\alpha_{PD}$), and the noise terms ($\sigma_D, \sigma_P$).
+These solutions are expressed as closed-form functions of the empirical covariances: $\mathrm{Var}(D)$, $\mathrm{Var}(P)$, $\mathrm{Var}(S)$, $\mathrm{Cov}(D, P)$, $\mathrm{Cov}(D, S)$, and $\mathrm{Cov}(P, S)$.
+
+This structure creates a family of analytically calibrated models, each consistent with the same observed covariances but differing in their internal configuration.
+However, we constrain the space of valid solutions by imposing two empirical and theoretical conditions: (1) all structural parameters must be positive, reflecting the assumption that stress and precariousness contribute adversely to depression; and (2) the system must be dynamically stable, ensuring convergence to a stationary distribution.
+These constraints restrict the allowable range of $\alpha_{DP}$ to approximately $[0, 0.698]$ (see Appendix [@sec-analcalibration]).
+
+@fig-tradeoff illustrates how the remaining parameters adjust as $\alpha_{DP}$ increases to maintain consistency with the empirical covariances.
+As $\alpha_{DP}$ increases, $\alpha_{PD}$ must decline to preserve stability, since the product $\alpha_{DP} \cdot \alpha_{PD}$ determines the strength of the feedback loop and must remain below 1.
+This creates a compensatory trade-off: as influence from $P$ to $D$ strengthens, the influence from $D$ to $P$ weakens.
+Meanwhile, the external coupling terms — $\alpha_{DS}$ and $\alpha_{PS}$ — also shift to absorb variance in a way that maintains alignment with the empirical data.
+Notably, the noise amplitudes $\sigma_D$ and $\sigma_P$ decline slightly as $\alpha_{DP}$ increases.
+This indicates that more of the observed variability is captured by deterministic interactions, with less left to be explained by stochastic fluctuations.
+
+Rather than being a limitation, this flexibility enables exploration of a spectrum of equally data-consistent feedback regimes.
+Each represents a different plausible causal architecture, offering insight into how internal dynamics modulate the system's responsiveness to external stress — even when the observed covariances remain unchanged.
 
 ```{r}
-#| label: fig-pareto
+#| label: fig-tradeoff
 #| fig-cap: >
-#|   Pareto front of simulated parameter sets.
-#|   *Each point represents a solution evaluated by its total variance error (x-axis) and total partial correlation error (y-axis), with color indicating total mean error.
-#|   Red crosses highlight the top 30 solutions used for simulation.*
+#|   Trade-off curves showing how calibrated parameters vary with $\alpha_{DP}$.
+#|   *Each line shows how a different parameter adjusts in response to changes in $\alpha_{DP}$ to maintain consistency with empirical covariances from the HELIUS data. 
+#|   All configurations shown are equally consistent with the data, illustrating how internal dynamics adjust as feedback strength changes.*
 #| fig-align: center
-#| out-width: 80%
-knitr::include_graphics("img/paretoplot.png")
+#| out-width: 100%
+knitr::include_graphics("img/param_tradeoff.pdf")
 ```
@@
 
-To evaluate the effects of hypothetical support, we simulate scenarios in which each individual’s financial stress level ($S$) is proportionally reduced toward the minimum observed value.
-This models a population-wide intervention applied at varying strengths --- from no reduction in stress (0%) to complete alignment with the minimum observed stress level (100%).
-
-For each of the top 30 parameter sets identified through calibration, we run simulations on 300 virtual individuals across this full range of intervention levels.
-In each simulation, we observe how depression ($D$) and precariousness ($P$) evolve over time, capturing their final values after the system stabilizes.
-This setup allows us to evaluate how different internal model dynamics (e.g., feedback strength) influence the system’s response to external change.
+To evaluate how internal structure affects system responsiveness, we simulate a family of linear models subjected to external intervention.
+Each model corresponds to a different value of $\alpha_{DP}$, representing the strength of influence from social precariousness to depression.
+The range spans from 0 (no coupling) to 0.65 (maximal admissible value), with all other parameters derived analytically to preserve the empirical covariances observed in the HELIUS data (see @fig-tradeoff).
+
+In each simulation, financial stress ($S$) is proportionally reduced toward the lowest observed value, modeling a population-wide intervention applied at varying intensities (0–100%).
+For each model, we simulate 1000 virtual individuals whose baseline stress levels are sampled from the empirical distribution and scaled according to the intervention level.
+
+We assess the system’s behavior using two complementary approaches:
+
+- Snapshot analysis captures the average levels of depression ($D$) and precariousness ($P$) at selected time points ($t = 100$, $300$, $500$) following intervention onset.
+Since all linear models converge to the same equilibrium, this highlights how internal feedback affects the speed and pathway of adjustment.
+
+- Full trajectory simulation tracks system dynamics over time for two representative models: one with minimal ($\alpha_{DP} = 0$) and one with maximal ($\alpha_{DP} = 0.65$) feedback.
+Each simulation includes an intervention period followed by its withdrawal, revealing differences in reactivity, delay, and recovery stability.
+
+Together, these simulations demonstrate how internal structure governs not just whether change occurs, but how it unfolds — shaping the pace, persistence, and variability of response to external support.
 
@@
 
-To evaluate how internal system structure influences the effect of external support, we simulated outcomes under a range of intervention strengths across the top 30 parameter sets from the calibrated model ensemble.
-The top panel of @fig-intervention shows mean levels of depression ($D$) and precariousness ($P$) after simulation, with each line representing one parameter set.
-Lines are color-coded by combined feedback and input strength, calculated as the product $\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2$.
-
-Parameter sets with stronger interaction dynamics exhibited more pronounced reductions in both depression and precariousness following intervention.
-At full intervention—modeled as shifting the external stressor ($S$) to its observed minimum—the reduction in depression ranged from [0.68*SD* – 0.97*SD*], and for social precariousness from [0.77*SD* – 0.94*SD*], corresponding to the lowest to highest values of parameter strength.[^3] These results suggest that systems with stronger internal dynamics are more responsive to external change, enabling targeted interventions to more effectively shift the system toward a lower-risk equilibrium.
-
-[^3]: Standard deviations (*SD*) are based on the original scale of each variable: $\textit{SD}_D = 1.0$ for depression ($D$) and $\textit{SD}_P = 0.9$ for precariousness ($P$).
-Parameter strength is defined as the product $\alpha_1 \times \beta_1 \times \alpha_2 \times \beta_2$, reflecting the combined influence of internal feedback and external input sensitivity.
-
-To better understand these dynamics, we focused on two illustrative parameter sets: one with strong feedback/input coupling and low noise (a *bistable* candidate), and one with weaker coupling and higher noise (a *unistable* candidate).
-In nonlinear systems, bistability refers to the existence of two distinct stable equilibria—meaning the system’s long-term outcome depends on its starting point or perturbations.
-
-In our model, bistability arises from the interaction between strong internal coupling—particularly the product $\alpha_2 \times \beta_2$—and the system’s nonlinearity.
-The $\tanh(\cdot)$ activation function saturates at high input values, meaning sufficient gain is required to push the system into nonlinear regimes where multiple stable points emerge.
-If the feedback gain is too low, or the system is overwhelmed by noise, it behaves more linearly and settles into a single equilibrium—yielding a unistable configuration.[^1]
-
-[^1]: This is similar to a well-known pattern in dynamical systems called a *pitchfork bifurcation*.
-In simple terms, when the system’s internal feedback becomes strong enough—for example, when $k$ in the equation $x_{t+1} = \tanh(k x_t)$ passes a certain threshold—the system changes from having one stable state to having two possible stable states (and one unstable one in between).
-In our model, this kind of shift happens when the feedback loop between depression and precariousness—captured by the product $\alpha_2 \times \beta_2$—becomes strong.
-It allows the system to settle into different long-term states, depending on initial conditions.
-
-The bottom panel of @fig-intervention illustrates these contrasting behaviors.
-In the bistable model, a clear bifurcation emerges: small interventions have little effect, but once a threshold is crossed, the system abruptly transitions to a low-risk state.
-In the unistable model, the transition is gradual and continuous, with outcomes improving incrementally as stress is reduced.
-
-These findings underscore that the impact of intervention depends not only on its intensity, but on the system's internal configuration.
-In bistable regimes, small or moderate interventions may be insufficient unless they push the system past a tipping point.
-Unistable systems, by contrast, respond more linearly, offering greater predictability and flexibility in designing incremental interventions.
-
-This distinction is especially relevant for mental health interventions in populations exposed to persistent stressors like housing instability, financial strain, or job insecurity.
-In such contexts, external adversity may be reinforced through internal feedback loops between psychological distress and precarious conditions.
-Recognizing when a system is bistable may help identify when strong, early, or multifaceted interventions are needed to produce lasting change—while unistable dynamics may justify more gradual or targeted efforts.
-
-Tailoring intervention strategies to the system’s underlying structure may enhance their effectiveness and sustainability, particularly in addressing the compounding effects of socioeconomic precariousness on mental health.
-
-<!-- The final outcomes, specifically the mean levels of depression ($D$) and precariousness ($P$), are presented in the top panel of @fig-intervention.
-Each trajectory corresponds to a parameter set and is color-coded by its combined feedback and input strength, defined as the product $\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2$.
--->
-
-<!-- Systems with stronger parameter interactions exhibit more pronounced shifts in behavior, particularly under intensive interventions.
-At full intervention (i.e., maximum stress reduction), the mean depression level decreased from -0.113 to -0.963, and precariousness from -0.100 to -0.958.
-These changes represent reductions of approximately 90.8\% relative to their respective baseline magnitudes, indicating that effective intervention can push the system close to its lowest-risk equilibrium.
--->
-
-<!-- To explore these dynamics in more detail, we focused on two representative parameter sets: one with high feedback/input strength and low noise (a bistable candidate), and another with weak parameter interactions and high noise (a unistable candidate).
-In nonlinear dynamical systems, bistability refers to the existence of two distinct stable equilibria—meaning the long-term outcome depends on initial conditions or perturbations.
--->
-
-<!-- In our model, bistability emerges from an interplay between the strength of parameter-driven feedback/input loops and the magnitude of stochastic noise.
-The internal loop between $D$ and $P$ — governed by the product $\alpha_2 \times \beta_2$ — allows changes in one variable to reinforce the other.
-At the same time, the $\tanh(\cdot)$ nonlinearity in the system saturates for large inputs, flattening the system’s responsiveness.
-For multiple stable equilibria to emerge, the internal gain must be large enough to push the system into these nonlinear (saturated) regions.
-If this gain is too small, or if noise is too high, the system behaves more linearly and converges to a single stable point — a unistable regime.[^1]  -->
-
-<!-- [^1]: This condition resembles the classical *pitchfork* bifurcation in discrete dynamical systems: when the gain $k$ in a scalar system such as $x_{t+1} = \tanh(k x_t)$ exceeds a critical threshold (e.g., $k>1$), the system admits three fixed points—two stable and one unstable—thus producing bistability.
-In our two-variable model, sufficiently high values of $\alpha_2 \times \beta_2$ similarly give rise to a self-reinforcing loop that sustains two attractor states.
--->
-
-<!-- This behavior is illustrated in the bottom panel of @fig-intervention.
-In the bistable model, we observe a bifurcation in final outcomes: as stress is reduced, the system undergoes a sharp transition from a high-$D$/$P$ regime to a low one—indicating a tipping point.
-In contrast, the unistable model shows a smooth, continuous response to increasing intervention, consistent with a single stable equilibrium.
--->
-
-<!-- These findings highlight how a system’s internal structure — including both nonlinear feedback gain and noise levels — fundamentally shapes its response to intervention.
-In bistable regimes, small or moderate interventions may have little effect unless they exceed a critical threshold sufficient to tip the system out of a high-risk state.
-In unistable systems, by contrast, outcomes improve more gradually and predictably with increasing support.
--->
-
-<!-- This distinction is particularly relevant when addressing persistent stressors associated with precariousness, such as unstable housing, employment insecurity, or chronic financial strain.
-In these contexts, the effects of external adversity may be reinforced internally through feedback loops between stress exposure and mental health.
-Such dynamics can stabilize harmful states and limit responsiveness to minor interventions—emphasizing the need for targeted, high-leverage support.
--->
-
-<!-- Recognizing whether a population’s mental health dynamics follow a bistable or unistable structure can guide policy design.
-For bistable systems, early or forceful interventions may be required to trigger lasting improvement.
-For unistable systems, more moderate or incremental measures may suffice.
-By tailoring strategies to the system’s underlying structure, interventions can be made more effective and context-sensitive—ultimately enhancing support for those facing precariousness-driven mental health risks.
--->
+To assess how internal structure influences the system’s responsiveness to external support, we simulate a set of linear models that differ in the assumed strength of influence from precariousness to depression, $\alpha_{DP}$.
+For each model, we generate a virtual population of 1,000 individuals and examine system behavior in two complementary ways: by analyzing average outcomes at selected time points following intervention onset, and by simulating full temporal trajectories for two representative cases at the extremes of the $\alpha_{DP}$ range.
+
+@fig-intervention summarizes how the internal structure of the model—specifically, variation in $\alpha_{DP}$—shapes the system’s short-term and long-term responses to external support.
+Panel A shows the mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500 after intervention onset, across a range of $\alpha_{DP}$ values.
+Each line represents a different model, color-coded from low (purple) to high (yellow) values of $\alpha_{DP}$.
+
+In the early phase of intervention ($t = 100$), depression and precariousness exhibit contrasting patterns.
+Depression declines more rapidly in models with lower $\alpha_{DP}$, where its dynamics are driven more directly by the external stressor $S$.
+As $\alpha_{DP}$ increases, $D$ becomes more dependent on precariousness ($P$), which itself adjusts more slowly—delaying the depressive response.
+Precariousness, by contrast, shows slightly faster early reductions in models with higher $\alpha_{DP}$.
+This reflects a shift in influence: as $\alpha_{DP}$ increases, $P$ becomes more directly responsive to $S$ (via increasing $\alpha_{PS}$) and less influenced by $D$, allowing it to adjust more promptly to changes in the stressor.
+
+Despite these short-term differences, all models ultimately converge to similar steady-state levels under sustained intervention.
+By time step 300, convergence is already substantial—though some variation remains, particularly in $P$, where gradients between models are still visible.
+By step 500, trajectories have nearly fully aligned across all configurations, consistent with the shared equilibrium structure of the linear system.
+This convergence highlights that the key distinctions lie not in whether intervention is effective, but in how quickly and through which pathways change unfolds.
+
+Panel B shows full time trajectories for two representative models: one with no coupling ($\alpha_{DP} = 0$) and one at the upper admissible limit ($\alpha_{DP} = 0.65$).
+In both, intervention begins at step 300 and ends at step 1500.
+In the uncoupled model, depression responds rapidly and rebounds shortly after support is removed.
+In the high-$\alpha_{DP}$ model, $D$ responds more gradually and recovers more slowly—reflecting its greater dependence on the slower-changing $P$.
+Precariousness declines in both models during intervention and returns to baseline afterward, but with a slightly sharper drop in the higher $\alpha_{DP}$ condition.
+This is because $P$ becomes more directly sensitive to $S$ as $\alpha_{PS}$ increases with $\alpha_{DP}$.
+
+These asymmetries further highlight how internal structure—not just external inputs—shapes the system’s trajectory.
+Specifically, changes in coupling parameters reconfigure how stress reduction propagates through the system, altering the speed and pathway of response even when long-run outcomes remain unchanged.
 
@@
 #| label: fig-intervention
-#| fig-cap: "**Top**: Final Mean D and P Across Intervention Sizes.
-*Mean trajectories of depression ($D$) and precariousness ($P$) across varying intervention strengths, simulated over the top 30 parameter sets.
-Each line represents one parameter set and is colored by its combined feedback and input strength.* **Bottom**: Smoothed Density Heatmap of Final $D$ and $P$ for Two Parameter Sets.
-*One exhibits unistability (top two panels) and the other exhibits bistability (bottom two panels).
-Bistable dynamics reveal a sharp transition between distinct outcome regimes, while unistable dynamics show gradual, continuous change across the intervention gradient.*"
+#| fig-cap: >
+#|   System response to external intervention across variations in internal structure.
+#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across different values of $\alpha_{DP}$ and intervention strengths.
+#|   Models with lower $\alpha_{DP}$ show faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
+#|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
+#|   Depression responds more gradually and persistently in the high $\alpha_{DP}$ model, whereas precariousness adjusts more quickly.*
+
 #| fig-align: center
@@
 
-knitr::include_graphics("img/intervention_plot2.pdf")
+knitr::include_graphics("img/intervention_linear.pdf")
 ```
@@
 
-To examine how such structures shape responsiveness to support, we developed a simplified two-variable nonlinear dynamical model capturing the feedback between depression and precariousness, modulated by financial stress.
+To explore how internal structure shapes response to external support, we developed a linear stochastic model of depression and precariousness, influenced by financial stress.
+Crucially, we analytically calibrated this model to match empirical covariances from the HELIUS data, ensuring that all parameter sets were observationally equivalent but differed in their internal configuration.
+We treated the parameter $\alpha_{DP}$—the effect of precariousness on depression—as a free variable, and derived all other parameters accordingly, yielding a family of valid models consistent with the data.
+Simulation results showed that even with identical steady-state covariances, differences in internal structure produced systematic variation in the timing and profile of system responses to intervention.
+In models with lower $\alpha_{DP}$, depression responded more quickly to reduced stress because its dynamics were more directly coupled to the external input.
+In contrast, as $\alpha_{DP}$ increased and depression became more indirectly influenced—mediated by the slower-changing precariousness variable—its response was more gradual and persistent.
+Precariousness, while generally more directly responsive across all models, also showed subtle differences, with slightly faster early declines in configurations where it was more sensitive to external stress.
+These results demonstrate that intervention impact depends not only on the external conditions (e.g., stress reduction), but also on the internal structure of the system—how variables are coupled and through which pathways change flows.
+Tailoring interventions to system configuration—using more intensive or earlier support in systems with indirect or delayed pathways, and sustained support in systems with rapid but unstable responses—may help optimize mental health outcomes under conditions of social and economic adversity.
+
+<!-- To examine how such structures shape responsiveness to support, we developed a simplified two-variable nonlinear dynamical model capturing the feedback between depression and precariousness, modulated by financial stress.
 Calibrated against empirical summary statistics from the HELIUS dataset, the model allowed us to simulate how these dynamics evolve under varying levels of stress reduction.
-The simulations revealed that intervention effectiveness depends not just on the magnitude of external stressors, but on the internal configuration of the system itself.
+-->
+<!-- The simulations revealed that intervention effectiveness depends not just on the magnitude of external stressors, but on the internal configuration of the system itself.
 In bistable regimes—characterized by strong feedback and low noise—the system could remain stuck in a high-risk state unless an intervention crosses a critical threshold, tipping it into a healthier equilibrium.
 By contrast, unistable systems—driven by weaker coupling or greater noise—responded more gradually and proportionally to increasing support.
-These dynamics carry practical policy implications.
+-->
+<!-- These dynamics carry practical policy implications.
 In bistable systems, even moderate interventions may have little effect unless they are forceful enough to tip the system toward a lower-risk state.
@@
 Unistable systems, by contrast, are more responsive to incremental support, suggesting that sustained, moderate interventions may be sufficient in such contexts.
-
-<!-- <!-- Interpretation of Findings --> 
-<!-- Our findings highlight the significant role of recent financial stress (S.fin) as a potential causal factor for depression, consistently observed across both sum score and symptom-level analyses.
-This consistency underscores the profound impact of financial stressors on mental health.
-The symptom-level analysis further identified specific symptoms—particularly sleep disturbances (slp), depressed mood (dep), guilt (glt), and anhedonia (anh)—as being especially sensitive to external precarious conditions.
--->
-
-<!-- Among these symptoms, dep stands out as a distinct causal factor for precariousness in both aggregated and disaggregated analyses, with CCI even suggesting a possible cyclic relationship between them.
-This implies that depression and precariousness may reinforce one another, potentially leading to a worsening cycle over time.
-Within the symptom network, dep also exhibits a high degree of connectivity, but with mostly incoming arrows, suggesting that as symptom dynamics intensify, dep is quickly impacted by multiple factors—particularly P.soc (social precariousness), further fueling the depression-precariousness cycle.
--->
-
-<!-- Beyond dep, slp emerges as a critical symptom in this dynamic.
-It connects with both social and employment precariousness, while also playing a central role in symptom interactions, affecting other symptoms more than most others.
-This suggests that slp may act as an initiator within the depressive symptom system, potentially triggering or amplifying symptom cascades.
--->
-
-<!-- Among the precariousness factors, P.soc appears to be one of the most pivotal—not as a driver of symptoms, but rather as a consequence of depression.
-It maintains the strongest connectivity with symptoms, yet its directionality indicates that it is largely shaped by depressive symptoms rather than acting as a primary cause.
-Together, these patterns suggest a broader directional dynamic in which financial stress (S.fin) exacerbates depressive symptoms, which in turn feed into social precariousness, perpetuating the cycle.
--->
-
-<!-- This directional insight suggests that interventions targeting specific depressive symptoms, particularly dep, may help prevent downstream effects on worsening precarious conditions.
-Alternatively, early intervention focusing on slp may be a promising strategy, as slp appears to act as a potential initiator or activator within the depressive symptom network.
-This positioning suggests that slp could serve as an early warning sign, offering a valuable target for preventive interventions aimed at breaking the depression-precariousness cycle before it intensifies.
--->
-
-<!-- Building on these insights, we developed a simplified computational model capturing the core mechanisms identified in the data: bidirectional feedback between depression and precariousness, driven by external stress.
-Calibrated against empirical summary statistics, the model allowed us to explore how different parameter configurations—particularly feedback strength and stochastic noise—modulate the system’s response to simulated interventions.
-Some systems exhibited bistability, resisting change until pushed past a critical threshold, while others responded more gradually and predictably.
-These patterns have direct policy implications: in bistable regimes, intensive or early interventions may be required to shift individuals out of high-risk states, whereas in unistable systems, more moderate, sustained support may suffice.
--->
-
-<!-- A key contribution of this work is to show that the structure of the depression–precariousness system matters: not all forms of adversity respond equally to intervention, and this heterogeneity may stem from internal feedback dynamics as much as from external conditions.
-Recognizing whether a given population operates within a bistable or unistable regime could inform the design and timing of mental health interventions—particularly in contexts of chronic socioeconomic stress.
--->
-
-<!-- Building on these findings, we developed a simplified two-variable nonlinear dynamical system to simulate the co-evolution of depression and social precariousness under varying levels of financial stress.
-This model served as a conceptual abstraction of the inferred causal structure, enabling us to assess how the system’s internal configuration—particularly feedback strength, input coupling, and stochastic noise—modulates its behavior in response to intervention.
--->
-
-<!-- Simulation results revealed that system structure critically influences responsiveness to intervention.
-In regimes with strong internal feedback loop and low noise, the model exhibits bistability: it can remain in a high-risk state unless a sufficiently large intervention tips it into a lower-risk configuration.
-By contrast, systems with weaker interactions or higher noise display unistable dynamics, where outcomes improve more gradually and proportionally with increasing intervention strength.
 -->
@@
 
-Finally, the structure of the dynamical model introduces its own limitations.
+Finally, the structure of our dynamical model introduces important limitations.
+While the linear formulation offered analytical tractability and transparent calibration, it likely oversimplifies the true dynamics of depression and socioeconomic precariousness.
+A growing body of evidence suggests that psychosocial systems are fundamentally nonlinear.
+Empirical and theoretical work points to phenomena such as threshold effects, where improvements only occur after a tipping point is crossed [@van2014critical; @park2025individual]; saturation, where the impact of support diminishes as basic needs are met [@cramer2010comorbidity]; and path dependence, where past adversity shapes sensitivity to future stressors [@wichers2016critical].
+These behaviors are fundamentally incompatible with linear dynamics, yet they may be essential for understanding why some individuals remain trapped in high-risk states while others recover more easily.
+Future modeling efforts should therefore incorporate mechanisms such as saturating input-response functions (e.g., sigmoidal or tanh forms) and state-dependent feedback (where coupling strength varies by system state).
+Incorporating such features would allow models to capture not just average effects, but also the critical conditions under which meaningful change becomes possible—clarifying whether some subgroups require forceful, threshold-crossing interventions, while others may benefit from sustained, moderate support.
+Ultimately, embracing nonlinear dynamics may be key to understanding whether different populations face genuinely distinct stability regimes—and whether effective intervention requires shifting the system across a critical boundary, rather than simply nudging it along its current trajectory.
+<!-- Finally, the structure of the dynamical model introduces its own limitations.
 While the model was designed to reflect key features of the discovered causal structure, it represents a highly simplified system: only two variables—depression and social precariousness—modulated by external financial stress.
@@
 Future work should further investigate whether bistability is a necessary feature of depression–precariousness dynamics, and explore alternative functional forms—both linear and nonlinear—using richer data and stronger empirical constraints.
-
-<!-- Several limitations should be acknowledged when interpreting these findings.
-The prevalence of bidirectional arrows and circle-marked endpoints in the graphs reflects unresolved ambiguities in the causal relationships suggested by the data.
-These uncertainties highlight the need for further research, particularly with higher-resolution datasets, such as time-series data.
-Incorporating time-series data could capture temporal dynamics between precariousness factors and depressive symptoms, providing time-specific insights and revealing how these relationships evolve over time.
-Methods like *PCMCI* [@runge2019detecting], *tsFCI* [@entner2010causal], and other time-series adaptations of causal discovery algorithms could offer a more robust framework for addressing temporal dependencies and refining the analysis.
--->
-<!-- Another notable limitation is the absence of clear evidence for cycles within the symptom network, despite employing algorithms designed to account for cyclic relationships.
-The CCI algorithm primarily produced bidirectional arrows, while FCI predominantly generated directional arrows, yet neither demonstrated patterns indicative of definitive cyclic structures.
-This suggests that addressing cyclic relationships is particularly challenging when relying solely on observational datasets.
-Future research could benefit from refined datasets that incorporate intervention data.
-Methods such as *LLC* [@hyttinen2012learning], *NODGAS-Flow *[@sethuraman2023nodags], and the recently developed *Bicycle* algorithm [@rohbeck2024], which utilize both observational and intervention data, may uncover potential cycles more effectively.
--->
-<!-- Lastly, regarding conditional independence (CI) testing, the differences between the graphs generated by Gaussian CI and RCoT underscore the methodological sensitivities in detecting causal relationships.
-Gaussian CI produced denser graphs, which is somewhat counterintuitive, as the Gaussian CI's strict linearity assumption would typically result in fewer detected relationships, not more.
--->
-<!-- A possible explanation for this discrepancy is that Gaussian CI's reliance on partial correlations may overestimate relationships when specific non-linear dependencies exist in the data.
-In contrast, RCoT, free from such assumptions, might better capture independence patterns under such conditions.
--->
-<!-- However, RCoT is not without its limitations.
-While technically non-parametric, its performance can be influenced by the distributional characteristics of the data.
-The Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle with discrete or mixed datasets, where distances between discrete points may fail to convey meaningful similarity [@howlett2001radial].
-This limitation could explain why RCoT might miss certain dependencies, particularly in the HELIUS dataset, where some variables lack smooth continuity.
--->
-<!-- Future research could tackle these challenges by developing hybrid kernels designed for mixed datasets, integrating both continuous and discrete variables into RCoT-like methods.
-This would have the potential to improve the reliability of findings, especially when working with datasets characterized by complex and heterogeneous structures.
--->
-
-<!-- A key limitation of our study lies in the structure and assumptions of the dynamical model.
-While the model is grounded in causal patterns identified from the data, it is intentionally simplified, focusing solely on two interacting variables—depression and social precariousness—modulated by financial stress.
-This abstraction facilitates tractable simulation and exploration of systemic behaviors such as bistability.
-However, it excludes several factors identified in the causal graph as potentially influential, including relational stress, housing conditions, and other contextual or protective variables like coping strategies or social support.
-These omissions constrain the model’s ability to represent the multifaceted nature of real-world dynamics.
--->
-<!-- Additionally, the model’s nonlinearity—implemented via a hyperbolic tangent function—was chosen for its analytical properties, such as bounded outputs and saturating responses, rather than empirical grounding.
-While such nonlinearities are necessary to produce bistable dynamics, the presence of bistability in our model is a product of simulation rather than empirical verification.
-We do not currently have time-series data to confirm the existence of tipping points or regime shifts in real-world trajectories of depression and precariousness.
-It is also possible that a linear model would suffice to capture much of the observed behavior, and that the current structure overcomplicates model formulation.
-Future work should seek for securing stronger evidence for bistability in the system and empirically test for the nonlinearity and compare model variants—linear and nonlinear—against richer longitudinal data to validate which formulations best capture the underlying dynamics.
--->
-
-<!-- The numerous bidirectional arrows in the CCI outputs also imply the presence of latent confounders.
-Expanding the analysis to include additional variables, such as other mental health conditions and coping mechanisms, could provide a more comprehensive understanding of the factors influencing depressive symptoms.
-Additionally, integrating qualitative approaches could enrich these findings, offering contextual insights into the quantitative relationships identified.
-Understanding individuals' lived experiences of precariousness and depression could inform the development of more tailored and effective interventions.
--->
-
-<!-- However, it could be also that RCoT performed somewhat suboptimal.
-while RCoT is technically non-parametric, its performance can still be influenced by the distributional characteristics of the data.
-Specifically, the Gaussian RBF (radial basis function) kernel, optimized for continuous data with smooth transitions, may struggle to capture relationships in datasets with discrete or mixed distributions.
-In such cases, the distances between discrete points may fail to convey meaningful similarity information.
-As a result, RCoT might miss certain dependencies, particularly when variables in the dataset lack smooth continuity like in our HELIUS dataset.
--->
-
-<!-- Future research could address these issues by exploring a broader range of CI testing approaches.
-One traditional approach to handle this is discretizing variables and use $G^2$ test, which may better capture dependencies in non-continuous datasets [@dojer2016learning; @neapolitan2004learning].
-A more promising direction, however, lies in the development of hybrid kernels tailored for mixed datasets, effectively integrating both continuous and discrete variables into RCoT-like methods.
--->
-<!-- By systematically employing and comparing a wider variety of CI testing techniques, researchers could gain more robust insights and mitigate the limitations inherent in specific approaches.
-This broader exploration holds the potential to enhance the reliability of findings, particularly in datasets with complex and heterogeneous structures.
 -->
@@
 <!-- Implications -->
-<!-- 1. still impressive -- symptom directions aligns much well with the other empirical studies. -->
-<!-- 2. understood better which aspect of precariousness is more important for depression and we can target depends on what level of intervention we are planning. -->
-
-<!-- Implications -->
-Our results reveal that the effectiveness of interventions targeting depression and precariousness is shaped not only by the strength of external stressors but also by the internal structure of the system.
-The interplay between feedback strength and noise determines whether a system responds gradually or only after crossing a tipping point—highlighting the need for intervention strategies attuned to these underlying dynamics.
-
-These findings underscore the importance of considering internal feedback—not just external adversity—when designing interventions.
-Mental health systems are not always passively responsive to support; their structure can constrain how much impact any given policy has [@wagenaar2013public; @desavigny2009systems].
-By diagnosing these structural features—through tools like causal discovery and computational modeling—researchers and policymakers can better tailor interventions to the underlying mechanics of vulnerability.
-
-Beyond these specific findings, this study offers a generalizable framework for studying complex mental health phenomena.
-Causal discovery enables principled hypothesis generation from observational data, while dynamical modeling helps test how those structures behave under intervention.
-Together, these methods move us from descriptive patterns to mechanistic insight, with practical relevance for real-world mental health policy.
-
-Future work could build richer dynamical models grounded in higher-resolution empirical data—potentially incorporating time, interventions, or more granular psychosocial constructs.
-Such extensions would help validate whether mechanisms proposed here, such as bistability, manifest in real-world trajectories.
-Ultimately, we hope this study encourages further integration of causal structure learning and system-level modeling to improve mental health outcomes under conditions of socioeconomic precariousness.
+Our findings demonstrate that intervention effectiveness depends not only on external stressors, but also on the internal structure of the system.
+While all models converge to the same long-term outcome, internal coupling shapes how quickly and through which pathways change unfolds.
+This highlights the importance of designing interventions that are sensitive to the system’s internal architecture.
+
+As public systems research has long shown, social and psychological systems do not always respond passively to external inputs; internal constraints can limit the impact of even well-intentioned policies [@wagenaar2013public; @desavigny2009systems].
+By using causal discovery and computational modeling, researchers and policymakers can better identify these structural constraints and tailor interventions accordingly.
+
+More broadly, this study offers a generalizable framework for investigating complex mental health dynamics.
+Causal discovery supports principled hypothesis generation from observational data, while dynamical modeling enables exploration of how hypothesized structures behave under external change.
+Together, these tools move us beyond static descriptions toward mechanistic insight—with direct relevance for real-world mental health policy and intervention design.
+
+Future research should build on this foundation by developing richer, more nuanced dynamical models grounded in longitudinal and interventional data.
+Incorporating time, multiple domains of precarity, and adaptive feedback processes would help assess whether the mechanisms proposed here—such as delayed response, saturation, or multi-stability—manifest in actual mental health trajectories.
+Ultimately, we hope this study encourages further integration of causal structure learning and systems modeling in efforts to understand and  improve mental health outcomes under conditions of socioeconomic precariousness.
 
@@
 
+<!-- ## Analytical Calibration of the Linear Stochastic Model {#sec-analcalibration} -->
+
+<!-- This appendix describes the derivation and calibration of parameters for a linear stochastic model linking depression ($D$), social precariousness ($P$), and an external stressor ($S$).
+The model is based on coupled Ornstein–Uhlenbeck (OU) processes and is analytically calibrated using empirical covariances from the HELIUS dataset.
+-->
+
+<!-- ### Model Structure -->
+
+<!-- The model assumes the following system of linear stochastic differential equations: -->
+
+<!-- $$ -->
+<!-- dD = \lambda_D \left( \alpha_{DS} S + \alpha_{DP} P - D \right) dt + \sigma_D\, dW_D  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- dP = \lambda_P \left( \alpha_{PS} S + \alpha_{PD} D - P \right) dt + \sigma_P\, dW_P -->
+<!-- $$ -->
+
+<!-- where: -->
+
+<!--  - $D$ = depression -->
+<!--  - $P$ = social precariousness -->
+<!--  - $S$ = exogenous financial stressor (fixed or sampled from empirical distribution) -->
+<!--  - $\lambda_D$, $\lambda_P$ = adjustment rates (set to 1) -->
+<!--  - $\alpha_{DS}, \alpha_{DP}, \alpha_{PS}, \alpha_{PD}$ = coupling coefficients -->
+<!--  - $\sigma_D, \sigma_P$ = noise amplitudes -->
+<!--  - $dW_D$, $dW_P$ = independent Wiener processes -->
+
+<!-- ### Stationary Covariance Solution -->
+
+<!-- At stationarity, the joint distribution of $D$ and $P$ is Gaussian. The system can be written in matrix form: -->
+
+<!-- $$ -->
+<!-- \frac{d\mathbf{x}}{dt} = A \mathbf{x} + B \boldsymbol{\xi}(t) -->
+<!-- \quad \text{where } \mathbf{x} = \begin{bmatrix} D \\ P \end{bmatrix} -->
+<!-- $$ -->
+
+<!-- The drift matrix $A$ and diffusion matrix $BB^\top$ are: -->
+
+<!-- $$ -->
+<!-- A = -->
+<!-- \begin{bmatrix} -->
+<!-- -\lambda_D & \lambda_D \alpha_{DP} \\ -->
+<!-- \lambda_P \alpha_{PD} & -\lambda_P -->
+<!-- \end{bmatrix}, \quad -->
+<!-- BB^\top = -->
+<!-- \begin{bmatrix} -->
+<!-- \sigma_D^2 & 0 \\ -->
+<!-- 0 & \sigma_P^2 -->
+<!-- \end{bmatrix} -->
+<!-- $$ -->
+
+<!-- Let $\Sigma_{DP}$ be the stationary $2 \times 2$ covariance matrix of $(D, P)$: -->
+
+<!-- $$ -->
+<!-- \Sigma_{DP} = -->
+<!-- \begin{bmatrix} -->
+<!-- v_D & c_{DP} \\ -->
+<!-- c_{DP} & v_P -->
+<!-- \end{bmatrix} -->
+<!-- $$ -->
+
+<!-- Then, the Lyapunov equation is: -->
+
+<!-- $$A \Sigma_{DP} + \Sigma_{DP} A^\top + BB^\top = 0$$ -->
+
+<!-- This provides a system of three equations in three unknowns: $v_D$, $v_P$, and $c_{DP}$. -->
+
+<!-- ### Covariances with the Stressor -->
+
+<!-- To complete the full $3 \times 3$ covariance matrix of $(D, P, S)$, we define: -->
+
+<!-- $$ -->
+<!-- \Delta = 1 - \alpha_{PD} \alpha_{DP} \\ -->
+<!-- k_D = \frac{\alpha_{DP} \alpha_{PS} + \alpha_{DS}}{\Delta} \\ -->
+<!-- k_P = \frac{\alpha_{PD} \alpha_{DS} + \alpha_{PS}}{\Delta} -->
+<!-- $$ -->
+
+<!-- Assuming $\mathrm{Var}(S) = v_S$, the unconditional covariances are: -->
+
+<!-- $$ -->
+<!-- \mathrm{Cov}(D, S) = k_D v_S  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Cov}(P, S) = k_P v_S  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Cov}(D, P) = k_D k_P v_S + c_{DP}  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Var}(D) = k_D^2 v_S + v_D  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Var}(P) = k_P^2 v_S + v_P -->
+<!-- $$ -->
+
+<!-- ### Empirical Calibration Procedure -->
+
+<!-- Empirical estimates from HELIUS provide: -->
+
+<!-- - $\mathrm{Var}(D)$, $\mathrm{Var}(P)$, $\mathrm{Var}(S)$ -->
+<!-- - $\mathrm{Cov}(D, S)$, $\mathrm{Cov}(P, S)$, $\mathrm{Cov}(D, P)$ -->
+
+<!-- Fix $\lambda_D = \lambda_P = 1$ and treat $\alpha_{DP}$ as a free parameter. Then solve for the five unknowns: -->
+<!-- $\alpha_{DS}, \alpha_{PS}, \alpha_{PD}, \sigma_D^2, \sigma_P^2$ -->
+<!-- symbolically using the observed moments. -->
+
+<!-- For example: -->
+<!-- $\alpha_{DS} = \frac{\mathrm{Cov}(D, S) - \alpha_{PS} \cdot \mathrm{Cov}(P, S)}{v_S}$ -->
+
+<!-- The solutions are expressed as closed-form functions of empirical covariances and the chosen $\alpha_{DP}$ value. -->
+
+<!-- ### Parameter Functions -->
+
+<!-- The following functional forms are derived in the implementation: -->
+
+<!-- $$ -->
+<!-- \sigma_D = f_{\sigma_D}(\mathrm{Cov}(D, S), \mathrm{Cov}(D, P), \mathrm{Cov}(P, S), \alpha_{DP}) -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \sigma_P = f_{\sigma_P}(\mathrm{Cov}(D, S), \mathrm{Cov}(D, P), \mathrm{Cov}(P, S), \alpha_{DP})  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \alpha_{DS} = f_{\alpha_{DS}}(\mathrm{Cov}(D, S), \mathrm{Cov}(D, P), \mathrm{Cov}(P, S), \alpha_{DP})  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \alpha_{PS} = f_{\alpha_{PS}}(\mathrm{Cov}(D, S), \mathrm{Cov}(D, P), \mathrm{Cov}(P, S), \alpha_{DP})  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \alpha_{PD} = f_{\alpha_{PD}}(\mathrm{Cov}(D, S), \mathrm{Cov}(D, P), \mathrm{Cov}(P, S), \alpha_{DP}) -->
+<!-- $$ -->
+
+<!-- The empirical values used in our study are: -->
+
+<!-- $$ -->
+<!-- \mathrm{Var}(D) = 1.000  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Var}(P) = 0.807  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Var}(S) = 0.743 -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Cov}(D, P) = 0.374 -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Cov}(D, S) = 0.357  -->
+<!-- $$ -->
+<!-- $$ -->
+<!-- \mathrm{Cov}(P, S) = 0.256 -->
+<!-- $$ -->
+
+<!-- ### Summary -->
+
+<!-- This calibration process enables exact, closed-form derivation of all model parameters directly from empirical covariance data.
+It avoids the need for iterative optimization and allows immediate simulation of the linear system under empirical constraints.
+-->
+
+## Analytical Calibration of the Linear Feedback Model{#sec-analcalibration}
+
+We consider a continuous-time, linear stochastic differential equation (SDE) model describing the coupled dynamics between depression ($D$) and social precariousness ($P$), each influenced by an external stressor ($S$).
+The system is specified as:
+
+$$
+\begin{aligned}
+\frac{dD}{dt} &= \alpha_{DS} S + \alpha_{DP} P - D + \sigma_D \cdot \eta_D(t), \\
+\frac{dP}{dt} &= \alpha_{PS} S + \alpha_{PD} D - P + \sigma_P \cdot \eta_P(t),
+\end{aligned}
+$$
+
+where:
+
+* $S$ is a fixed or exogenous input (external financial stress),
+* $\alpha_{DP}$ represents the feedback from precariousness to depression,
+* $\alpha_{PD}$ represents the feedback from depression to precariousness,
+* $\alpha_{DS}$ and $alpha_{PS}$ capture the direct effects of $S$,
+* $\sigma_D, \sigma_P$ are stochastic noise amplitudes,
+* $\eta_D(t), \eta_P(t)$ are independent standard Wiener processes.
+
+We fix the decay rates $\lambda_D = \lambda_P = 1$, normalizing the time scale due to lack of temporal resolution in the HELIUS dataset.
+
+### Stationarity Assumption
+
+We assume the joint distribution of $(D, P)$ reaches stationarity, leading to a multivariate Ornstein–Uhlenbeck process.
+Under this assumption, the system has a unique stationary distribution characterized by its covariance structure, which we match to the empirical data.
+
+Let:
+
+$$
+\Sigma_{DD} = \mathrm{Var}(D), \quad \Sigma_{PP} = \mathrm{Var}(P), \quad \Sigma_{SS} = \mathrm{Var}(S),
+$$
+
+$$
+\Sigma_{DP} = \mathrm{Cov}(D, P), \quad \Sigma_{DS} = \mathrm{Cov}(D, S), \quad \Sigma_{PS} = \mathrm{Cov}(P, S).
+$$
+
+From the HELIUS data, we use the following observed values:
+
+$\Sigma_{DD} = 1.000, \quad \Sigma_{PP} = 0.807, \quad \Sigma_{SS} = 0.743, \quad \Sigma_{DP} = 0.374, \quad \Sigma_{DS} = 0.357, \quad \Sigma_{PS} = 0.256.$
+
+### Analytical Parameter Derivation
+
+We consider a linear stochastic dynamical system in which depression ($D$) and social precariousness ($P$) evolve under mutual influence and are jointly affected by an exogenous stressor ($S$).
+We assume stationary dynamics governed by a two-dimensional Ornstein–Uhlenbeck process:
+
+$$
+\begin{aligned}
+dD_t &= \left(-D_t + \alpha_{DP} P_t + \alpha_{DS} S_t\right)dt + \sigma_D\, dW_{1t}, \\
+dP_t &= \left(-P_t + \alpha_{PD} D_t + \alpha_{PS} S_t\right)dt + \sigma_P\, dW_{2t}.
+\end{aligned}
+$$
+
+Here, we fix the autoregressive decay rates to 1 for identifiability and treat $\alpha_{DP}$ — the influence of precariousness on depression — as a free parameter.
+All other parameters are then derived to ensure the model reproduces the empirical second-order moments: variances $(\Sigma_{DD}, \Sigma_{PP}, \Sigma_{SS})$ and covariances $(\Sigma_{DP}, \Sigma_{DS}, \Sigma_{PS})$.
+
+Under the stationary solution of the linear SDE system, the following algebraic identities hold:
+
+$$
+\begin{aligned}
+\Sigma_{DS} &= \alpha_{DS} \Sigma_{SS} + \alpha_{DP} \Sigma_{PS}, \\
+\Sigma_{DP} &= \alpha_{DS} \Sigma_{PS} + \alpha_{DP} \Sigma_{PP}, \\
+\Sigma_{PS} &= \alpha_{PS} \Sigma_{SS} + \alpha_{PD} \Sigma_{DS}.
+\end{aligned}
+$$
+
+Solving these simultaneously yields the following closed-form expressions as functions of $\alpha_{DP}$:
+
+$$
+\alpha_{DS} = \frac{\Sigma_{DS} - \alpha_{DP} \cdot \Sigma_{PS}}{\Sigma_{SS}},
+$$
+
+$$
+\alpha_{PD} = \frac{2 \Sigma_{DS} \Sigma_{PS} - 2 \Sigma_{DP} \Sigma_{SS} - \alpha_{DP} \Sigma_{PS}^2 + \alpha_{DP} \Sigma_{PP} \Sigma_{SS}}{\Sigma_{DS}^2 - \Sigma_{DD} \Sigma_{SS}},
+$$
+
+$$
+\alpha_{PS} = \frac{
+\Sigma_{DS}^2 \Sigma_{PS}
++ \Sigma_{PS} \Sigma_{DD} \Sigma_{SS}
++ \Sigma_{DS}(-2 \Sigma_{DP} \Sigma_{SS} - \alpha_{DP} \Sigma_{PS}^2 + \alpha_{DP} \Sigma_{PP} \Sigma_{SS})
+}{
+\Sigma_{SS}(\Sigma_{DS}^2 - \Sigma_{DD} \Sigma_{SS})
+}.
+$$
+
+The noise amplitudes $\sigma_D^2$ and $\sigma_P^2$ are derived from the Lyapunov equation associated with the stationary covariance matrix. These yield:
+
+$$
+\sigma_D^2 = \frac{-2 \left(\Sigma_{DS}^2 - \Sigma_{DD} \Sigma_{SS} - \Sigma_{DS} \Sigma_{PS} \alpha_{DP} + \Sigma_{DP} \Sigma_{SS} \alpha_{DP} \right)}{\Sigma_{SS}},
+$$
+
+$$
+\sigma_P^2 = \frac{K}{\Sigma_{SS} (\Sigma_{DD} \Sigma_{SS} - \Sigma_{DS}^2)},
+$$
+
+where
+
+$$
+\begin{aligned}
+K &= 8 \Sigma_{DP} \Sigma_{DS} \Sigma_{PS} \Sigma_{SS}
+- 4 \Sigma_{DP}^2 \Sigma_{SS}^2
+- 2 \Sigma_{DS}^2 (\Sigma_{PS}^2 + \Sigma_{PP} \Sigma_{SS}) \\
+&\quad + 2 \Sigma_{DS} \Sigma_{PS} (\Sigma_{PS}^2 - \Sigma_{PP} \Sigma_{SS}) \alpha_{DP}
++ 2 \Sigma_{SS} (\Sigma_{PP} \Sigma_{SS} - \Sigma_{PS}^2)(\Sigma_{DD} + \Sigma_{DP} \alpha_{DP}).
+\end{aligned}
+$$
+
+These expressions define a one-parameter family of structurally distinct yet observationally equivalent models, each exactly matching the empirical second-order statistics.
+
+### Constraints on $\alpha_{DP}$
+
+Although all expressions are algebraically valid for arbitrary real $\alpha_{DP}$, admissibility requires that the resulting model remains meaningful and stable.
+
+#### Positivity Constraints
+
+To ensure interpretability, we require all effects and noise terms to be strictly positive:
+
+$$
+\alpha_{DS} > 0, \quad \alpha_{PD} > 0, \quad \alpha_{PS} > 0, \quad \sigma_D^2 > 0, \quad \sigma_P^2 > 0.
+$$
+
+Each of these is a function of $\alpha_{DP}$, and thus these conditions restrict its allowable range. In particular:
+
+* $\alpha_{PD} = 0$ when $\alpha_{DP} \approx 0.698$,
+* $\alpha_{DS} = 0$ when $\alpha_{DP} \approx 1.39$,
+* $\sigma_D^2 = 0$ when $\alpha_{DP} \approx 3.30$,
+* $\sigma_P^2 = 0$ at a value given exactly by:
+
+$$
+\alpha_{DP}^{(\sigma_P = 0)} =
+\frac{
+- \Sigma_{DD} \Sigma_{PP} \Sigma_{SS}^2 + \Sigma_{DD} \Sigma_{PS}^2 \Sigma_{SS}
++ 2 \Sigma_{DP}^2 \Sigma_{SS}^2 - 4 \Sigma_{DP} \Sigma_{DS} \Sigma_{PS} \Sigma_{SS}
++ \Sigma_{DS}^2 \Sigma_{PP} \Sigma_{SS} + \Sigma_{DS}^2 \Sigma_{PS}^2
+}{
+\Sigma_{DP} \Sigma_{PP} \Sigma_{SS}^2 - \Sigma_{DP} \Sigma_{PS}^2 \Sigma_{SS}
+- \Sigma_{DS} \Sigma_{PP} \Sigma_{PS} \Sigma_{SS} + \Sigma_{DS} \Sigma_{PS}^3
+}.
+$$
+
+This value falls above $\alpha_{DP} = 0.698$, and thus does not tighten the admissible upper bound.
+
+#### Stability Constraint
+
+The system is stable if the deterministic dynamics converge, which requires the drift matrix
+
+$$
+A = \begin{bmatrix}
+-1 & \alpha_{DP} \\
+\alpha_{PD} & -1
+\end{bmatrix}
+$$
+
+to have all eigenvalues with negative real parts. This is equivalent to:
+
+$$
+\text{Tr}(A) < 0 \quad \text{and} \quad \det(A) = 1 - \alpha_{DP} \alpha_{PD} > 0.
+$$
+
+The trace condition is always satisfied ($\text{Tr}(A) = -2$), so stability reduces to:
+
+$$
+\alpha_{DP} \cdot \alpha_{PD} < 1.
+$$
+
+This condition fails exactly when $\alpha_{PD} \to 0^+$, which coincides with the tightest positivity constraint.
+
+### Admissible Range
+
+Numerical evaluation using empirical covariances from the HELIUS dataset shows that:
+
+* $\alpha_{PD} \to 0$ at $\alpha_{DP} \approx 0.698$,
+* All other positivity and stability constraints are looser.
+
+Hence, the maximum valid value is:
+
+$$
+\boxed{\alpha_{DP}^{\max} = 0.698}
+$$
+
+All simulations and model variants are confined to the admissible domain:
+
+$$
+\alpha_{DP} \in [0,\, 0.698]
+$$
+
+This guarantees that resulting systems are dynamically stable, fully positive, and consistent with empirical second-order moments.
+
 # Acknowledgements {.unnumbered}
@@
 We are most grateful to the participants of the HELIUS study and the management team, research nurses, interviewers, research assistants and other staff who have taken part in gathering the data of this study.
-
-<!-- ### Advantages of RCIT and RCoT -->
-
-<!-- - **Scalability**: Linear complexity ($O(nD)$) makes these methods efficient for large datasets. -->
-<!-- - **Robustness**: Provide accurate approximations of kernel-based CI tests while significantly reducing computational costs. -->
-<!-- - **Applicability**: Suitable for large-scale causal discovery applications, particularly when integrated with algorithms like PC and FCI. -->
-
-<!-- By combining the flexibility of kernel methods with the computational efficiency of RFFs, RCIT and RCoT represent powerful tools for non-parametric CI testing, enabling their use in high-dimensional and large-scale datasets.
--->
-
-<!-- ## Summarized Stable Edges Proportion {#sec-propmatrix} -->
-
-<!-- ```{r} -->
-<!-- #| label: fig-sum-mat -->
-<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on depression sum score." -->
-<!-- #| fig-align: center -->
-<!-- #| out-width: 100% -->
-<!-- knitr::include_graphics("img/sumgraph_mat.pdf") -->
-<!-- knitr::include_graphics("img/sumgraph_mat.pdf") -->
-
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- #| label: fig-sym-mat -->
-<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms." -->
-<!-- #| fig-align: center -->
-<!-- #| out-width: 100% -->
-<!-- knitr::include_graphics("img/symgraph_mat.pdf") -->
-<!-- ``` -->
-
-<!-- ```{r} -->
-<!-- #| label: fig-presum-mat -->
-<!-- #| fig-cap: "Proportions of edge endpoint types for graphs based on individual depression symptoms with precariousness sum score." -->
-<!-- #| fig-align: center -->
-<!-- #| out-width: 70% -->
-<!-- knitr::include_graphics("img/presumgraph_mat2.pdf") -->
-<!-- ``` -->

### 2025-06-27 draft_v1/draft_v2.tex
update draft_v2

@@
 
-\date{2025-06-24}
+\date{2025-06-27}
 \begin{document}
@@
 \begin{abstract}
-\noindent This study investigates the dynamic relationship between
-precarious life conditions and depression by combining constraint-based
-causal discovery with computational modeling. Using cross-sectional data
-from the HELIUS study, we applied cycle-capable causal discovery
-algorithms to uncover directional relationships between domain-specific
-precariousness indicators, and depressive symptoms. Our multi-resolution
-analysis---spanning both aggregate and symptom-level
-constructs---identified financial stress as a robust driver of
-depression, and highlighted key symptoms, such as sleep disturbance and
-depressed mood, as likely initiators or reinforcers of downstream
-precariousness. To explore how these structural insights shape
-intervention effectiveness, we developed a simplified nonlinear
-dynamical model simulating the co-evolution of depression and social
-precariousness under varying levels of financial stress reduction.
-Simulation results revealed that the system's responsiveness depends not
-only on the intensity of external intervention but also on internal
-structural features such as feedback strength and noise. In particular,
-some configurations exhibited bistability, requiring interventions to
-surpass a critical threshold to trigger lasting improvement, while
-others responded more proportionally. Together, these findings emphasize
-that internal feedback dynamics---not just external adversity---can
-shape the impact of mental health interventions. More broadly, our study
-offers a scalable framework for integrating causal discovery and
-dynamical simulation to better understand and inform policy responses to
-complex psychosocial phenomena.
+\noindent This study investigates how the internal structure of
+depression--precariousness dynamics shapes responsiveness to external
+support. Using cross-sectional data from the HELIUS study, we first
+apply cycle-capable causal discovery algorithms to estimate directional
+dependencies between financial stress, domain-specific precariousness,
+and depressive symptoms. Our multi-resolution analysis---spanning both
+aggregate and symptom-level constructs---highlights financial stress as
+a plausible upstream driver of depression, with symptoms such as sleep
+disturbance and depressed mood emerging as initiators of broader social
+vulnerability. To translate these structural insights into dynamic
+predictions, we construct a family of analytically calibrated linear
+stochastic models. Simulations reveal that variations in internal
+structure can substantially alter system responses to financial stress
+reduction. Some models show rapid but transient improvements in
+depression, while others adjust more slowly yet sustain gains over time.
+Precariousness tends to respond more immediately across all
+configurations, though its trajectory also shifts subtly with structural
+variation. Together, these findings demonstrate that intervention
+outcomes depend not only on the magnitude of external input but also on
+latent properties of the system. More broadly, this study illustrates
+how combining causal discovery with dynamical modeling supports
+principled, simulation-based exploration of intervention
+scenarios---grounded in data, yet able to probe counterfactual
+structure. This approach offers a scalable framework for diagnosing
+hidden system constraints and informing targeted public health
+strategies in complex psychosocial domains.
 \end{abstract}
@@
 that influence each other over time, with \(S\) acting as a shared
-input. These dynamics are formalized as a pair of nonlinear stochastic
+input. These dynamics are formalized as a pair of linear stochastic
 differential equations (SDEs):
@@
 parameters \(\lambda_D\) and \(\lambda_P\) control the timescales over
-which variable adjusts toward its input-driven target state. The
-coefficients \(\alpha_1, \alpha_2\) and \(\beta_1, \beta_2\) represent
-the strengths of the causal influences between variables. Stochastic
-fluctuations are introduced through the terms \(\sigma_1 dW_1\) and
-\(\sigma_2 dW_2\), where \(dW_1\) and \(dW_2\) are independent Wiener
-processes. The system uses the hyperbolic tangent function (\(tanh\)) to
-bound the influence of inputs, ensuring smooth, saturating dynamics and
-preventing runaway feedback loops. Importantly, the nonlinearity
-introduced by the \(tanh\) function also allows the system to exhibit
-bistability, supporting the existence of two stable states: one
-corresponding to low depression and precariousness, and the other to
-high depression and precariousness.
+which each variable adjusts toward its input-driven target state. The
+coefficients \(\alpha_{PS}\) and \(\alpha_{DS}\) quantify the direct
+effects of the stressor on \(P\) and \(D\), while \(\alpha_{DP}\) and
+\(\alpha_{PD}\) represent the cross-influences between depression and
+precariousness. Stochastic fluctuations are introduced through the noise
+terms \(\sigma_P, dW_P\) and \(\sigma_D, dW_D\), with \(dW_P\) and
+\(dW_D\) representing independent Wiener processes. This linear
+structure offers a transparent and analytically tractable framework for
+examining how variations in external financial stress influence the
+population-level dynamics of depression and social precariousness.
+Crucially, the model's simplicity enables direct calibration from
+empirical data: all parameters can be derived analytically from observed
+variances and covariances in the HELIUS cohort. The following
+Section~\ref{sec-calibration} details the procedure used to perform this
+calibration.
 
@@
 
-\subsubsection{Model Calibration}\label{model-calibration}
-
-To identify realistic parameter settings for the simulation model, we
-calibrate six free parameters---\(\alpha_1\), \(\alpha_2\), \(\beta_1\),
-\(\beta_2\), and noise amplitudes \(\sigma_1\), \(\sigma_2\)---using the
-\emph{Non-dominated Sorting Genetic Algorithm II} (NSGA-II) (Deb et al.,
-2002), a multi-objective evolutionary optimization method well-suited
-for noisy and high-dimensional search spaces. The goal is to match
-simulated system behavior to empirical summary statistics derived from
-the HELIUS dataset. Specifically, we aim to match the observed means and
-variances of depression (\(D\)) and social precariousness (\(P\)), along
-with all three partial correlations among \(D\), \(P\), and financial
-stress (\(S\)): \(\rho_{DP.S}\), \(\rho_{DS.P}\), and \(\rho_{PS.D}\).
-
-Given the stochastic nature of the model, each candidate parameter set
-is evaluated by simulating 1,000 virtual individuals, each assigned a
-fixed \(S\) value sampled from the empirical distribution. From these
-simulations, we compute population-level summary statistics based on the
-final states. The fitness of each parameter set is measured by how
-closely its summary statistics matched the empirical data, using the
-total absolute deviation across all objectives.
-
-To explore optimal parameter configurations, we used the NSGA-II
-algorithm as implemented in the \texttt{nsga2R} package (Franz \&
-Nakamura, 2015). This algorithm is inspired by natural selection and
-evolves a population of candidate solutions over multiple generations to
-balance competing objectives. We began with a population of 500 randomly
-generated parameter sets and allowed them to evolve over 300
-generations. During this process, solutions were selected, combined, and
-mutated using standard settings: a crossover probability of 0.9 (which
-controls how often solutions are combined), a mutation probability of
-0.25 (introducing random variation), and a tournament size of 2 (used to
-pick the best candidates for reproduction). The result is a Pareto
-front---a set of solutions that are considered optimal in the sense that
-no single objective can be improved without worsening another. From this
-diverse set of trade-off solutions (see Figure~\ref{fig-pareto}), we
-selected the 30 parameter sets with the lowest overall error across
-objectives. These top-performing configurations were then used in the
-simulation phase to evaluate the potential effects of different
-interventions.
+\subsubsection{Model Calibration}\label{sec-calibration}
+
+In our linear model, all parameters are derived analytically to
+reproduce the empirical second-order moments observed in the HELIUS
+dataset. The system treats depression (\(D\)) and social precariousness
+(\(P\)) as jointly evolving under the influence of a shared external
+stressor (\(S\)), forming a coupled system of Ornstein--Uhlenbeck
+processes. Under the assumption of stationarity, the steady-state
+covariances between these variables admit closed-form solutions for the
+system's structural parameters.
+
+We fix the adjustment rates \(\lambda_D = \lambda_P = 1\), which
+normalizes the time scale given the cross-sectional nature of the HELIUS
+data. To resolve the remaining degrees of freedom, we treat the feedback
+from precariousness to depression (\(\alpha_{DP}\)) as a free parameter,
+and solve algebraically for all other coefficients --- including the
+direct effects of stress (\(\alpha_{DS}, \alpha_{PS}\)), the reverse
+feedback from depression to precariousness (\(\alpha_{PD}\)), and the
+noise terms (\(\sigma_D, \sigma_P\)). These solutions are expressed as
+closed-form functions of the empirical covariances: \(\mathrm{Var}(D)\),
+\(\mathrm{Var}(P)\), \(\mathrm{Var}(S)\), \(\mathrm{Cov}(D, P)\),
+\(\mathrm{Cov}(D, S)\), and \(\mathrm{Cov}(P, S)\).
+
+This structure creates a family of analytically calibrated models, each
+consistent with the same observed covariances but differing in their
+internal configuration. However, we constrain the space of valid
+solutions by imposing two empirical and theoretical conditions: (1) all
+structural parameters must be positive, reflecting the assumption that
+stress and precariousness contribute adversely to depression; and (2)
+the system must be dynamically stable, ensuring convergence to a
+stationary distribution. These constraints restrict the allowable range
+of \(\alpha_{DP}\) to approximately \([0, 0.698]\) (see Appendix
+Section~\ref{sec-analcalibration}).
+
+Figure~\ref{fig-tradeoff} illustrates how the remaining parameters
+adjust as \(\alpha_{DP}\) increases to maintain consistency with the
+empirical covariances. As \(\alpha_{DP}\) increases, \(\alpha_{PD}\)
+must decline to preserve stability, since the product
+\(\alpha_{DP} \cdot \alpha_{PD}\) determines the strength of the
+feedback loop and must remain below 1. This creates a compensatory
+trade-off: as influence from \(P\) to \(D\) strengthens, the influence
+from \(D\) to \(P\) weakens. Meanwhile, the external coupling terms ---
+\(\alpha_{DS}\) and \(\alpha_{PS}\) --- also shift to absorb variance in
+a way that maintains alignment with the empirical data. Notably, the
+noise amplitudes \(\sigma_D\) and \(\sigma_P\) decline slightly as
+\(\alpha_{DP}\) increases. This indicates that more of the observed
+variability is captured by deterministic interactions, with less left to
+be explained by stochastic fluctuations.
+
+Rather than being a limitation, this flexibility enables exploration of
+a spectrum of equally data-consistent feedback regimes. Each represents
+a different plausible causal architecture, offering insight into how
+internal dynamics modulate the system's responsiveness to external
+stress --- even when the observed covariances remain unchanged.
 
@@
 
-\includegraphics[width=0.8\textwidth,height=\textheight]{img/paretoplot.png}
-
-}
-
-\caption{\label{fig-pareto}Pareto front of simulated parameter sets.
-\emph{Each point represents a solution evaluated by its total variance
-error (x-axis) and total partial correlation error (y-axis), with color
-indicating total mean error. Red crosses highlight the top 30 solutions
-used for simulation.}}
+\includegraphics[width=1\textwidth,height=\textheight]{img/param_tradeoff.pdf}
+
+}
+
+\caption{\label{fig-tradeoff}Trade-off curves showing how calibrated
+parameters vary with \(\alpha_{DP}\). \emph{Each line shows how a
+different parameter adjusts in response to changes in \(\alpha_{DP}\) to
+maintain consistency with empirical covariances from the HELIUS data.
+All configurations shown are equally consistent with the data,
+illustrating how internal dynamics adjust as feedback strength
+changes.}}
 
@@
 
-To evaluate the effects of hypothetical support, we simulate scenarios
-in which each individual's financial stress level (\(S\)) is
-proportionally reduced toward the minimum observed value. This models a
-population-wide intervention applied at varying strengths --- from no
-reduction in stress (0\%) to complete alignment with the minimum
-observed stress level (100\%).
-
-For each of the top 30 parameter sets identified through calibration, we
-run simulations on 300 virtual individuals across this full range of
-intervention levels. In each simulation, we observe how depression
-(\(D\)) and precariousness (\(P\)) evolve over time, capturing their
-final values after the system stabilizes. This setup allows us to
-evaluate how different internal model dynamics (e.g., feedback strength)
-influence the system's response to external change.
+To evaluate how internal structure affects system responsiveness, we
+simulate a family of linear models subjected to external intervention.
+Each model corresponds to a different value of \(\alpha_{DP}\),
+representing the strength of influence from social precariousness to
+depression. The range spans from 0 (no coupling) to 0.65 (maximal
+admissible value), with all other parameters derived analytically to
+preserve the empirical covariances observed in the HELIUS data (see
+Figure~\ref{fig-tradeoff}).
+
+In each simulation, financial stress (\(S\)) is proportionally reduced
+toward the lowest observed value, modeling a population-wide
+intervention applied at varying intensities (0--100\%). For each model,
+we simulate 1000 virtual individuals whose baseline stress levels are
+sampled from the empirical distribution and scaled according to the
+intervention level.
+
+We assess the system's behavior using two complementary approaches:
+
+\begin{itemize}
+\item
+  Snapshot analysis captures the average levels of depression (\(D\))
+  and precariousness (\(P\)) at selected time points (\(t = 100\),
+  \(300\), \(500\)) following intervention onset. Since all linear
+  models converge to the same equilibrium, this highlights how internal
+  feedback affects the speed and pathway of adjustment.
+\item
+  Full trajectory simulation tracks system dynamics over time for two
+  representative models: one with minimal (\(\alpha_{DP} = 0\)) and one
+  with maximal (\(\alpha_{DP} = 0.65\)) feedback. Each simulation
+  includes an intervention period followed by its withdrawal, revealing
+  differences in reactivity, delay, and recovery stability.
+\end{itemize}
+
+Together, these simulations demonstrate how internal structure governs
+not just whether change occurs, but how it unfolds --- shaping the pace,
+persistence, and variability of response to external support.
 
@@
 
-To evaluate how internal system structure influences the effect of
-external support, we simulated outcomes under a range of intervention
-strengths across the top 30 parameter sets from the calibrated model
-ensemble. The top panel of Figure~\ref{fig-intervention} shows mean
-levels of depression (\(D\)) and precariousness (\(P\)) after
-simulation, with each line representing one parameter set. Lines are
-color-coded by combined feedback and input strength, calculated as the
-product \(\alpha_1 \times \alpha_2 \times \beta_1 \times \beta_2\).
-
-Parameter sets with stronger interaction dynamics exhibited more
-pronounced reductions in both depression and precariousness following
-intervention. At full intervention---modeled as shifting the external
-stressor (\(S\)) to its observed minimum---the reduction in depression
-ranged from {[}0.68\emph{SD} -- 0.97\emph{SD}{]}, and for social
-precariousness from {[}0.77\emph{SD} -- 0.94\emph{SD}{]}, corresponding
-to the lowest to highest values of parameter strength.\footnote{Standard
-  deviations (\emph{SD}) are based on the original scale of each
-  variable: \(\textit{SD}_D = 1.0\) for depression (\(D\)) and
-  \(\textit{SD}_P = 0.9\) for precariousness (\(P\)). Parameter strength
-  is defined as the product
-  \(\alpha_1 \times \beta_1 \times \alpha_2 \times \beta_2\), reflecting
-  the combined influence of internal feedback and external input
-  sensitivity.} These results suggest that systems with stronger
-internal dynamics are more responsive to external change, enabling
-targeted interventions to more effectively shift the system toward a
-lower-risk equilibrium.
-
-To better understand these dynamics, we focused on two illustrative
-parameter sets: one with strong feedback/input coupling and low noise (a
-\emph{bistable} candidate), and one with weaker coupling and higher
-noise (a \emph{unistable} candidate). In nonlinear systems, bistability
-refers to the existence of two distinct stable equilibria---meaning the
-system's long-term outcome depends on its starting point or
-perturbations.
-
-In our model, bistability arises from the interaction between strong
-internal coupling---particularly the product
-\(\alpha_2 \times \beta_2\)---and the system's nonlinearity. The
-\(\tanh(\cdot)\) activation function saturates at high input values,
-meaning sufficient gain is required to push the system into nonlinear
-regimes where multiple stable points emerge. If the feedback gain is too
-low, or the system is overwhelmed by noise, it behaves more linearly and
-settles into a single equilibrium---yielding a unistable
-configuration.\footnote{This is similar to a well-known pattern in
-  dynamical systems called a \emph{pitchfork bifurcation}. In simple
-  terms, when the system's internal feedback becomes strong enough---for
-  example, when \(k\) in the equation \(x_{t+1} = \tanh(k x_t)\) passes
-  a certain threshold---the system changes from having one stable state
-  to having two possible stable states (and one unstable one in
-  between). In our model, this kind of shift happens when the feedback
-  loop between depression and precariousness---captured by the product
-  \(\alpha_2 \times \beta_2\)---becomes strong. It allows the system to
-  settle into different long-term states, depending on initial
-  conditions.}
-
-The bottom panel of Figure~\ref{fig-intervention} illustrates these
-contrasting behaviors. In the bistable model, a clear bifurcation
-emerges: small interventions have little effect, but once a threshold is
-crossed, the system abruptly transitions to a low-risk state. In the
-unistable model, the transition is gradual and continuous, with outcomes
-improving incrementally as stress is reduced.
-
-These findings underscore that the impact of intervention depends not
-only on its intensity, but on the system's internal configuration. In
-bistable regimes, small or moderate interventions may be insufficient
-unless they push the system past a tipping point. Unistable systems, by
-contrast, respond more linearly, offering greater predictability and
-flexibility in designing incremental interventions.
-
-This distinction is especially relevant for mental health interventions
-in populations exposed to persistent stressors like housing instability,
-financial strain, or job insecurity. In such contexts, external
-adversity may be reinforced through internal feedback loops between
-psychological distress and precarious conditions. Recognizing when a
-system is bistable may help identify when strong, early, or multifaceted
-interventions are needed to produce lasting change---while unistable
-dynamics may justify more gradual or targeted efforts.
-
-Tailoring intervention strategies to the system's underlying structure
-may enhance their effectiveness and sustainability, particularly in
-addressing the compounding effects of socioeconomic precariousness on
-mental health.
+To assess how internal structure influences the system's responsiveness
+to external support, we simulate a set of linear models that differ in
+the assumed strength of influence from precariousness to depression,
+\(\alpha_{DP}\). For each model, we generate a virtual population of
+1,000 individuals and examine system behavior in two complementary ways:
+by analyzing average outcomes at selected time points following
+intervention onset, and by simulating full temporal trajectories for two
+representative cases at the extremes of the \(\alpha_{DP}\) range.
+
+Figure~\ref{fig-intervention} summarizes how the internal structure of
+the model---specifically, variation in \(\alpha_{DP}\)---shapes the
+system's short-term and long-term responses to external support. Panel A
+shows the mean levels of depression (\(D\)) and precariousness (\(P\))
+at time steps 100, 300, and 500 after intervention onset, across a range
+of \(\alpha_{DP}\) values. Each line represents a different model,
+color-coded from low (purple) to high (yellow) values of
+\(\alpha_{DP}\).
+
+In the early phase of intervention (\(t = 100\)), depression and
+precariousness exhibit contrasting patterns. Depression declines more
+rapidly in models with lower \(\alpha_{DP}\), where its dynamics are
+driven more directly by the external stressor \(S\). As \(\alpha_{DP}\)
+increases, \(D\) becomes more dependent on precariousness (\(P\)), which
+itself adjusts more slowly---delaying the depressive response.
+Precariousness, by contrast, shows slightly faster early reductions in
+models with higher \(\alpha_{DP}\). This reflects a shift in influence:
+as \(\alpha_{DP}\) increases, \(P\) becomes more directly responsive to
+\(S\) (via increasing \(\alpha_{PS}\)) and less influenced by \(D\),
+allowing it to adjust more promptly to changes in the stressor.
+
+Despite these short-term differences, all models ultimately converge to
+similar steady-state levels under sustained intervention. By time step
+300, convergence is already substantial---though some variation remains,
+particularly in \(P\), where gradients between models are still visible.
+By step 500, trajectories have nearly fully aligned across all
+configurations, consistent with the shared equilibrium structure of the
+linear system. This convergence highlights that the key distinctions lie
+not in whether intervention is effective, but in how quickly and through
+which pathways change unfolds.
+
+Panel B shows full time trajectories for two representative models: one
+with no coupling (\(\alpha_{DP} = 0\)) and one at the upper admissible
+limit (\(\alpha_{DP} = 0.65\)). In both, intervention begins at step 300
+and ends at step 1500. In the uncoupled model, depression responds
+rapidly and rebounds shortly after support is removed. In the
+high-\(\alpha_{DP}\) model, \(D\) responds more gradually and recovers
+more slowly---reflecting its greater dependence on the slower-changing
+\(P\). Precariousness declines in both models during intervention and
+returns to baseline afterward, but with a slightly sharper drop in the
+higher \(\alpha_{DP}\) condition. This is because \(P\) becomes more
+directly sensitive to \(S\) as \(\alpha_{PS}\) increases with
+\(\alpha_{DP}\).
+
+These asymmetries further highlight how internal structure---not just
+external inputs---shapes the system's trajectory. Specifically, changes
+in coupling parameters reconfigure how stress reduction propagates
+through the system, altering the speed and pathway of response even when
+long-run outcomes remain unchanged.
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{img/intervention_plot2.pdf}
-
-}
-
-\caption{\label{fig-intervention}\textbf{Top}: Final Mean D and P Across
-Intervention Sizes. \emph{Mean trajectories of depression (\(D\)) and
-precariousness (\(P\)) across varying intervention strengths, simulated
-over the top 30 parameter sets. Each line represents one parameter set
-and is colored by its combined feedback and input strength.}
-\textbf{Bottom}: Smoothed Density Heatmap of Final \(D\) and \(P\) for
-Two Parameter Sets. \emph{One exhibits unistability (top two panels) and
-the other exhibits bistability (bottom two panels). Bistable dynamics
-reveal a sharp transition between distinct outcome regimes, while
-unistable dynamics show gradual, continuous change across the
-intervention gradient.}}
+\includegraphics{img/intervention_linear.pdf}
+
+}
+
+\caption{\label{fig-intervention}System response to external
+intervention across variations in internal structure.
+\emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
+precariousness (\(P\)) at time steps 100, 300, and 500, across different
+values of \(\alpha_{DP}\) and intervention strengths. Models with lower
+\(\alpha_{DP}\) show faster early reductions in \(D\), while higher
+\(\alpha_{DP}\) leads to more immediate declines in \(P\). \textbf{Panel
+B}: Full time trajectories for two models---one with no coupling
+(\(\alpha_{DP} = 0\)) and one with strong coupling
+(\(\alpha_{DP} = 0.65\))---illustrating system response and recovery
+following intervention onset (step 300) and withdrawal (step 1500).
+Depression responds more gradually and persistently in the high
+\(\alpha_{DP}\) model, whereas precariousness adjusts more quickly.}}
 
@@
 
-To examine how such structures shape responsiveness to support, we
-developed a simplified two-variable nonlinear dynamical model capturing
-the feedback between depression and precariousness, modulated by
-financial stress. Calibrated against empirical summary statistics from
-the HELIUS dataset, the model allowed us to simulate how these dynamics
-evolve under varying levels of stress reduction. The simulations
-revealed that intervention effectiveness depends not just on the
-magnitude of external stressors, but on the internal configuration of
-the system itself. In bistable regimes---characterized by strong
-feedback and low noise---the system could remain stuck in a high-risk
-state unless an intervention crosses a critical threshold, tipping it
-into a healthier equilibrium. By contrast, unistable systems---driven by
-weaker coupling or greater noise---responded more gradually and
-proportionally to increasing support. These dynamics carry practical
-policy implications. In bistable systems, even moderate interventions
-may have little effect unless they are forceful enough to tip the system
-toward a lower-risk state. This underscores the importance of early or
-intensive support when targeting self-reinforcing cycles of depression
-and precariousness. Unistable systems, by contrast, are more responsive
-to incremental support, suggesting that sustained, moderate
-interventions may be sufficient in such contexts.
+To explore how internal structure shapes response to external support,
+we developed a linear stochastic model of depression and precariousness,
+influenced by financial stress. Crucially, we analytically calibrated
+this model to match empirical covariances from the HELIUS data, ensuring
+that all parameter sets were observationally equivalent but differed in
+their internal configuration. We treated the parameter
+\(\alpha_{DP}\)---the effect of precariousness on depression---as a free
+variable, and derived all other parameters accordingly, yielding a
+family of valid models consistent with the data. Simulation results
+showed that even with identical steady-state covariances, differences in
+internal structure produced systematic variation in the timing and
+profile of system responses to intervention. In models with lower
+\(\alpha_{DP}\), depression responded more quickly to reduced stress
+because its dynamics were more directly coupled to the external input.
+In contrast, as \(\alpha_{DP}\) increased and depression became more
+indirectly influenced---mediated by the slower-changing precariousness
+variable---its response was more gradual and persistent. Precariousness,
+while generally more directly responsive across all models, also showed
+subtle differences, with slightly faster early declines in
+configurations where it was more sensitive to external stress. These
+results demonstrate that intervention impact depends not only on the
+external conditions (e.g., stress reduction), but also on the internal
+structure of the system---how variables are coupled and through which
+pathways change flows. Tailoring interventions to system
+configuration---using more intensive or earlier support in systems with
+indirect or delayed pathways, and sustained support in systems with
+rapid but unstable responses---may help optimize mental health outcomes
+under conditions of social and economic adversity.
 
@@
 
-Finally, the structure of the dynamical model introduces its own
-limitations. While the model was designed to reflect key features of the
-discovered causal structure, it represents a highly simplified system:
-only two variables---depression and social precariousness---modulated by
-external financial stress. This abstraction allowed for tractable
-analysis and visual intuition, particularly around nonlinear behaviors
-such as bistability, but omits other potentially influential constructs
-highlighted in the causal graphs, including relational stress,
-employment conditions and housing conditions. Moreover, the model's
-nonlinearity---introduced via the hyperbolic tangent function---was
-chosen for convenience rather than empirical justification. While this
-function supports saturation and the emergence of tipping points, the
-presence of bistability is a product of simulation, not empirical
-verification. It remains possible that simpler, linear dynamics could
-account for the observed data just as well. Future work should further
-investigate whether bistability is a necessary feature of
-depression--precariousness dynamics, and explore alternative functional
-forms---both linear and nonlinear---using richer data and stronger
-empirical constraints.
-
-Our results reveal that the effectiveness of interventions targeting
-depression and precariousness is shaped not only by the strength of
-external stressors but also by the internal structure of the system. The
-interplay between feedback strength and noise determines whether a
-system responds gradually or only after crossing a tipping
-point---highlighting the need for intervention strategies attuned to
-these underlying dynamics.
-
-These findings underscore the importance of considering internal
-feedback---not just external adversity---when designing interventions.
-Mental health systems are not always passively responsive to support;
-their structure can constrain how much impact any given policy has
-(Savigny \& Adam, 2009; Wagenaar \& Burris, 2013). By diagnosing these
-structural features---through tools like causal discovery and
-computational modeling---researchers and policymakers can better tailor
-interventions to the underlying mechanics of vulnerability.
-
-Beyond these specific findings, this study offers a generalizable
-framework for studying complex mental health phenomena. Causal discovery
-enables principled hypothesis generation from observational data, while
-dynamical modeling helps test how those structures behave under
-intervention. Together, these methods move us from descriptive patterns
-to mechanistic insight, with practical relevance for real-world mental
-health policy.
-
-Future work could build richer dynamical models grounded in
-higher-resolution empirical data---potentially incorporating time,
-interventions, or more granular psychosocial constructs. Such extensions
-would help validate whether mechanisms proposed here, such as
-bistability, manifest in real-world trajectories. Ultimately, we hope
-this study encourages further integration of causal structure learning
-and system-level modeling to improve mental health outcomes under
-conditions of socioeconomic precariousness.
+Finally, the structure of our dynamical model introduces important
+limitations. While the linear formulation offered analytical
+tractability and transparent calibration, it likely oversimplifies the
+true dynamics of depression and socioeconomic precariousness. A growing
+body of evidence suggests that psychosocial systems are fundamentally
+nonlinear. Empirical and theoretical work points to phenomena such as
+threshold effects, where improvements only occur after a tipping point
+is crossed (Leemput et al., 2014; Park et al., 2025); saturation, where
+the impact of support diminishes as basic needs are met (Cramer et al.,
+2010); and path dependence, where past adversity shapes sensitivity to
+future stressors (Wichers et al., 2016). These behaviors are
+fundamentally incompatible with linear dynamics, yet they may be
+essential for understanding why some individuals remain trapped in
+high-risk states while others recover more easily. Future modeling
+efforts should therefore incorporate mechanisms such as saturating
+input-response functions (e.g., sigmoidal or tanh forms) and
+state-dependent feedback (where coupling strength varies by system
+state). Incorporating such features would allow models to capture not
+just average effects, but also the critical conditions under which
+meaningful change becomes possible---clarifying whether some subgroups
+require forceful, threshold-crossing interventions, while others may
+benefit from sustained, moderate support. Ultimately, embracing
+nonlinear dynamics may be key to understanding whether different
+populations face genuinely distinct stability regimes---and whether
+effective intervention requires shifting the system across a critical
+boundary, rather than simply nudging it along its current trajectory.
+
+Our findings demonstrate that intervention effectiveness depends not
+only on external stressors, but also on the internal structure of the
+system. While all models converge to the same long-term outcome,
+internal coupling shapes how quickly and through which pathways change
+unfolds. This highlights the importance of designing interventions that
+are sensitive to the system's internal architecture.
+
+As public systems research has long shown, social and psychological
+systems do not always respond passively to external inputs; internal
+constraints can limit the impact of even well-intentioned policies
+(Savigny \& Adam, 2009; Wagenaar \& Burris, 2013). By using causal
+discovery and computational modeling, researchers and policymakers can
+better identify these structural constraints and tailor interventions
+accordingly.
+
+More broadly, this study offers a generalizable framework for
+investigating complex mental health dynamics. Causal discovery supports
+principled hypothesis generation from observational data, while
+dynamical modeling enables exploration of how hypothesized structures
+behave under external change. Together, these tools move us beyond
+static descriptions toward mechanistic insight---with direct relevance
+for real-world mental health policy and intervention design.
+
+Future research should build on this foundation by developing richer,
+more nuanced dynamical models grounded in longitudinal and
+interventional data. Incorporating time, multiple domains of precarity,
+and adaptive feedback processes would help assess whether the mechanisms
+proposed here---such as delayed response, saturation, or
+multi-stability---manifest in actual mental health trajectories.
+Ultimately, we hope this study encourages further integration of causal
+structure learning and systems modeling in efforts to understand and
+improve mental health outcomes under conditions of socioeconomic
+precariousness.
 
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-butterworth2012role}
+Butterworth, P., Olesen, S. C., \& Leach, L. S. (2012). The role of
+hardship in the association between socio-economic position and
+depression. \emph{Australian \& New Zealand Journal of Psychiatry},
+\emph{46}(4), 364--373.
+
+\bibitem[\citeproctext]{ref-cramer2010comorbidity}
+Cramer, A. O., Waldorp, L. J., Van Der Maas, H. L., \& Borsboom, D.
+(2010). Comorbidity: A network perspective. \emph{Behavioral and Brain
+Sciences}, \emph{33}(2-3), 137--150.
+
 \bibitem[\citeproctext]{ref-deb2002fast}
@@
 
+\bibitem[\citeproctext]{ref-van2014critical}
+Leemput, I. A. van de, Wichers, M., Cramer, A. O., Borsboom, D.,
+Tuerlinckx, F., Kuppens, P., Van Nes, E. H., Viechtbauer, W., Giltay, E.
+J., Aggen, S. H., et al. (2014). Critical slowing down as early warning
+for the onset and termination of depression. \emph{Proceedings of the
+National Academy of Sciences}, \emph{111}(1), 87--92.
+
 \bibitem[\citeproctext]{ref-lindsay2000moment}
@@
 
+\bibitem[\citeproctext]{ref-lorant2003socioeconomic}
+Lorant, V., Deliège, D., Eaton, W., Robert, A., Philippot, P., \&
+Ansseau, M. (2003). Socioeconomic inequalities in depression: A
+meta-analysis. \emph{American Journal of Epidemiology}, \emph{157}(2),
+98--112.
+
 \bibitem[\citeproctext]{ref-mckee2017living}
@@
 Reports}, \emph{21}, 1--7.
+
+\bibitem[\citeproctext]{ref-ridley2020poverty}
+Ridley, M., Rao, G., Schilbach, F., \& Patel, V. (2020). Poverty,
+depression, and anxiety: Causal evidence and mechanisms. \emph{Science},
+\emph{370}(6522), eaay0214.
 
@@
 Theory and methods}. John Wiley \& Sons.
+
+\bibitem[\citeproctext]{ref-wichers2016critical}
+Wichers, M., Groot, P. C., Psychosystems, E., Group, E., et al. (2016).
+Critical slowing down as a personalized early warning signal for
+depression. \emph{Psychotherapy and Psychosomatics}, \emph{85}(2),
+114--116.
 
@@
 
+\subsection{Analytical Calibration of the Linear Feedback
+Model}\label{sec-analcalibration}
+
+We consider a continuous-time, linear stochastic differential equation
+(SDE) model describing the coupled dynamics between depression (\(D\))
+and social precariousness (\(P\)), each influenced by an external
+stressor (\(S\)). The system is specified as:
+
+\[
+\begin{aligned}
+\frac{dD}{dt} &= \alpha_{DS} S + \alpha_{DP} P - D + \sigma_D \cdot \eta_D(t), \\
+\frac{dP}{dt} &= \alpha_{PS} S + \alpha_{PD} D - P + \sigma_P \cdot \eta_P(t),
+\end{aligned}
+\]
+
+where:
+
+\begin{itemize}
+\tightlist
+\item
+  \(S\) is a fixed or exogenous input (external financial stress),
+\item
+  \(\alpha_{DP}\) represents the feedback from precariousness to
+  depression,
+\item
+  \(\alpha_{PD}\) represents the feedback from depression to
+  precariousness,
+\item
+  \(\alpha_{DS}\) and \(alpha_{PS}\) capture the direct effects of
+  \(S\),
+\item
+  \(\sigma_D, \sigma_P\) are stochastic noise amplitudes,
+\item
+  \(\eta_D(t), \eta_P(t)\) are independent standard Wiener processes.
+\end{itemize}
+
+We fix the decay rates \(\lambda_D = \lambda_P = 1\), normalizing the
+time scale due to lack of temporal resolution in the HELIUS dataset.
+
+\subsubsection{Stationarity Assumption}\label{stationarity-assumption}
+
+We assume the joint distribution of \((D, P)\) reaches stationarity,
+leading to a multivariate Ornstein--Uhlenbeck process. Under this
+assumption, the system has a unique stationary distribution
+characterized by its covariance structure, which we match to the
+empirical data.
+
+Let:
+
+\[
+\Sigma_{DD} = \mathrm{Var}(D), \quad \Sigma_{PP} = \mathrm{Var}(P), \quad \Sigma_{SS} = \mathrm{Var}(S),
+\]
+
+\[
+\Sigma_{DP} = \mathrm{Cov}(D, P), \quad \Sigma_{DS} = \mathrm{Cov}(D, S), \quad \Sigma_{PS} = \mathrm{Cov}(P, S).
+\]
+
+From the HELIUS data, we use the following observed values:
+
+\(\Sigma_{DD} = 1.000, \quad \Sigma_{PP} = 0.807, \quad \Sigma_{SS} = 0.743, \quad \Sigma_{DP} = 0.374, \quad \Sigma_{DS} = 0.357, \quad \Sigma_{PS} = 0.256.\)
+
+\subsubsection{Analytical Parameter
+Derivation}\label{analytical-parameter-derivation}
+
+We consider a linear stochastic dynamical system in which depression
+(\(D\)) and social precariousness (\(P\)) evolve under mutual influence
+and are jointly affected by an exogenous stressor (\(S\)). We assume
+stationary dynamics governed by a two-dimensional Ornstein--Uhlenbeck
+process:
+
+\[
+\begin{aligned}
+dD_t &= \left(-D_t + \alpha_{DP} P_t + \alpha_{DS} S_t\right)dt + \sigma_D\, dW_{1t}, \\
+dP_t &= \left(-P_t + \alpha_{PD} D_t + \alpha_{PS} S_t\right)dt + \sigma_P\, dW_{2t}.
+\end{aligned}
+\]
+
+Here, we fix the autoregressive decay rates to 1 for identifiability and
+treat \(\alpha_{DP}\) --- the influence of precariousness on depression
+--- as a free parameter. All other parameters are then derived to ensure
+the model reproduces the empirical second-order moments: variances
+\((\Sigma_{DD}, \Sigma_{PP}, \Sigma_{SS})\) and covariances
+\((\Sigma_{DP}, \Sigma_{DS}, \Sigma_{PS})\).
+
+Under the stationary solution of the linear SDE system, the following
+algebraic identities hold:
+
+\[
+\begin{aligned}
+\Sigma_{DS} &= \alpha_{DS} \Sigma_{SS} + \alpha_{DP} \Sigma_{PS}, \\
+\Sigma_{DP} &= \alpha_{DS} \Sigma_{PS} + \alpha_{DP} \Sigma_{PP}, \\
+\Sigma_{PS} &= \alpha_{PS} \Sigma_{SS} + \alpha_{PD} \Sigma_{DS}.
+\end{aligned}
+\]
+
+Solving these simultaneously yields the following closed-form
+expressions as functions of \(\alpha_{DP}\):
+
+\[
+\alpha_{DS} = \frac{\Sigma_{DS} - \alpha_{DP} \cdot \Sigma_{PS}}{\Sigma_{SS}},
+\]
+
+\[
+\alpha_{PD} = \frac{2 \Sigma_{DS} \Sigma_{PS} - 2 \Sigma_{DP} \Sigma_{SS} - \alpha_{DP} \Sigma_{PS}^2 + \alpha_{DP} \Sigma_{PP} \Sigma_{SS}}{\Sigma_{DS}^2 - \Sigma_{DD} \Sigma_{SS}},
+\]
+
+\[
+\alpha_{PS} = \frac{
+\Sigma_{DS}^2 \Sigma_{PS}
++ \Sigma_{PS} \Sigma_{DD} \Sigma_{SS}
++ \Sigma_{DS}(-2 \Sigma_{DP} \Sigma_{SS} - \alpha_{DP} \Sigma_{PS}^2 + \alpha_{DP} \Sigma_{PP} \Sigma_{SS})
+}{
+\Sigma_{SS}(\Sigma_{DS}^2 - \Sigma_{DD} \Sigma_{SS})
+}.
+\]
+
+The noise amplitudes \(\sigma_D^2\) and \(\sigma_P^2\) are derived from
+the Lyapunov equation associated with the stationary covariance matrix.
+These yield:
+
+\[
+\sigma_D^2 = \frac{-2 \left(\Sigma_{DS}^2 - \Sigma_{DD} \Sigma_{SS} - \Sigma_{DS} \Sigma_{PS} \alpha_{DP} + \Sigma_{DP} \Sigma_{SS} \alpha_{DP} \right)}{\Sigma_{SS}},
+\]
+
+\[
+\sigma_P^2 = \frac{K}{\Sigma_{SS} (\Sigma_{DD} \Sigma_{SS} - \Sigma_{DS}^2)},
+\]
+
+where
+
+\[
+\begin{aligned}
+K &= 8 \Sigma_{DP} \Sigma_{DS} \Sigma_{PS} \Sigma_{SS}
+- 4 \Sigma_{DP}^2 \Sigma_{SS}^2
+- 2 \Sigma_{DS}^2 (\Sigma_{PS}^2 + \Sigma_{PP} \Sigma_{SS}) \\
+&\quad + 2 \Sigma_{DS} \Sigma_{PS} (\Sigma_{PS}^2 - \Sigma_{PP} \Sigma_{SS}) \alpha_{DP}
++ 2 \Sigma_{SS} (\Sigma_{PP} \Sigma_{SS} - \Sigma_{PS}^2)(\Sigma_{DD} + \Sigma_{DP} \alpha_{DP}).
+\end{aligned}
+\]
+
+These expressions define a one-parameter family of structurally distinct
+yet observationally equivalent models, each exactly matching the
+empirical second-order statistics.
+
+\subsubsection{\texorpdfstring{Constraints on
+\(\alpha_{DP}\)}{Constraints on \textbackslash alpha\_\{DP\}}}\label{constraints-on-alpha_dp}
+
+Although all expressions are algebraically valid for arbitrary real
+\(\alpha_{DP}\), admissibility requires that the resulting model remains
+meaningful and stable.
+
+\paragraph{Positivity Constraints}\label{positivity-constraints}
+
+To ensure interpretability, we require all effects and noise terms to be
+strictly positive:
+
+\[
+\alpha_{DS} > 0, \quad \alpha_{PD} > 0, \quad \alpha_{PS} > 0, \quad \sigma_D^2 > 0, \quad \sigma_P^2 > 0.
+\]
+
+Each of these is a function of \(\alpha_{DP}\), and thus these
+conditions restrict its allowable range. In particular:
+
+\begin{itemize}
+\tightlist
+\item
+  \(\alpha_{PD} = 0\) when \(\alpha_{DP} \approx 0.698\),
+\item
+  \(\alpha_{DS} = 0\) when \(\alpha_{DP} \approx 1.39\),
+\item
+  \(\sigma_D^2 = 0\) when \(\alpha_{DP} \approx 3.30\),
+\item
+  \(\sigma_P^2 = 0\) at a value given exactly by:
+\end{itemize}
+
+\[
+\alpha_{DP}^{(\sigma_P = 0)} =
+\frac{
+- \Sigma_{DD} \Sigma_{PP} \Sigma_{SS}^2 + \Sigma_{DD} \Sigma_{PS}^2 \Sigma_{SS}
++ 2 \Sigma_{DP}^2 \Sigma_{SS}^2 - 4 \Sigma_{DP} \Sigma_{DS} \Sigma_{PS} \Sigma_{SS}
++ \Sigma_{DS}^2 \Sigma_{PP} \Sigma_{SS} + \Sigma_{DS}^2 \Sigma_{PS}^2
+}{
+\Sigma_{DP} \Sigma_{PP} \Sigma_{SS}^2 - \Sigma_{DP} \Sigma_{PS}^2 \Sigma_{SS}
+- \Sigma_{DS} \Sigma_{PP} \Sigma_{PS} \Sigma_{SS} + \Sigma_{DS} \Sigma_{PS}^3
+}.
+\]
+
+This value falls above \(\alpha_{DP} = 0.698\), and thus does not
+tighten the admissible upper bound.
+
+\paragraph{Stability Constraint}\label{stability-constraint}
+
+The system is stable if the deterministic dynamics converge, which
+requires the drift matrix
+
+\[
+A = \begin{bmatrix}
+-1 & \alpha_{DP} \\
+\alpha_{PD} & -1
+\end{bmatrix}
+\]
+
+to have all eigenvalues with negative real parts. This is equivalent to:
+
+\[
+\text{Tr}(A) < 0 \quad \text{and} \quad \det(A) = 1 - \alpha_{DP} \alpha_{PD} > 0.
+\]
+
+The trace condition is always satisfied (\(\text{Tr}(A) = -2\)), so
+stability reduces to:
+
+\[
+\alpha_{DP} \cdot \alpha_{PD} < 1.
+\]
+
+This condition fails exactly when \(\alpha_{PD} \to 0^+\), which
+coincides with the tightest positivity constraint.
+
+\subsubsection{Admissible Range}\label{admissible-range}
+
+Numerical evaluation using empirical covariances from the HELIUS dataset
+shows that:
+
+\begin{itemize}
+\tightlist
+\item
+  \(\alpha_{PD} \to 0\) at \(\alpha_{DP} \approx 0.698\),
+\item
+  All other positivity and stability constraints are looser.
+\end{itemize}
+
+Hence, the maximum valid value is:
+
+\[
+\boxed{\alpha_{DP}^{\max} = 0.698}
+\]
+
+All simulations and model variants are confined to the admissible
+domain:
+
+\[
+\alpha_{DP} \in [0,\, 0.698]
+\]
+
+This guarantees that resulting systems are dynamically stable, fully
+positive, and consistent with empirical second-order moments.
+
 \section*{Acknowledgements}\label{acknowledgements}

### 2025-06-28 draft_v1/draft_v2.qmd
update draft

@@
 * $\alpha_{PD}$ represents the feedback from depression to precariousness,
-* $\alpha_{DS}$ and $alpha_{PS}$ capture the direct effects of $S$,
+* $\alpha_{DS}$ and $\alpha_{PS}$ capture the direct effects of $S$,
 * $\sigma_D, \sigma_P$ are stochastic noise amplitudes,

### 2025-07-27 draft_v1/draft_v2.qmd
draft edit_Vitorfeedback

@@
 Mental health disorders represent a growing global health concern, especially in urbanized regions [@WHO_2022; @gruebner2017cities].
-In 2019, one in eight people worldwide were living with a mental health condition, with the highest disability-adjusted life years (DALYs) due to mental and addictive disorders concentrated in high-income countries such as those in Northern Europe, North America, and Australia [@rehm2019global; @WHO_2022].
+In 2019, one in eight people worldwide were living with a mental health condition, with the highest disability-adjusted life years (DALYs) due to mental and addictive disorders present in high-income countries such as those in Northern Europe, North America, and Australia [@rehm2019global; @WHO_2022].
 Despite growing awareness, governments have struggled to design effective responses.
-Mental health outcomes arise from complex, multi-level dynamics—intertwined with the social, economic, and spatial structures of urban environments—making both diagnosis and intervention design deeply challenging [@van2021advancing].
+Mental health outcomes arise from complex, multi-level dynamics between the social, economic, and spatial structures of urban environments—making both diagnosis and intervention design deeply challenging [@van2021advancing].
 
@@
 Moreover, precarious working conditions, housing instability, and neighborhood disadvantage have all been linked to increased risk of depression and anxiety [@pevalin2017impact; @fone2014effect; @rugulies2023work; @ronnblad2019precarious].
-This growing literature highlights that poor mental health is not merely a personal issue, but one rooted in sustained socioeconomic stress and structural uncertainty.
-
-Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains—including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering; @mckee2017living].
+This growing literature highlights that poor mental health is not merely a personal issue but one rooted in sustained socioeconomic stress and structural uncertainty.
+
+Building on this, recent studies have introduced the concept of precariousness as a multidimensional condition characterized by instability and lack of control across several life domains, including employment, finances, housing, social relations, and cultural belonging [@elsenburg2025clustering; @mckee2017living].
 This broader view reveals how different forms of insecurity can co-occur and compound, forming an ecosystem of risk.
 However, despite growing evidence of association between precariousness and mental health, a fundamental question remains unresolved: *How do these factors causally influence one another?* Most existing studies rely on cross-sectional correlations, leaving open questions about directionality and feedback.
-Without stronger causal insight, it remains difficult to identify priority targets for intervention or to anticipate unintended effects.
-
-This study aims to investigate the causal mechanisms linking precariousness and depression—and how those mechanisms shape system responses to external intervention.
+Without stronger causal insight, it remains difficult to identify priority targets for intervention or to anticipate and quantify intended and unintended effects.
+
+This study aims to investigate the causal mechanisms linking precariousness and depression and how those mechanisms shape system responses to external intervention.
 Specifically, we ask: *How does the internal configuration of the precariousness–depression relationship affect the outcomes of stress-reducing interventions?*
 To answer this, we combine two approaches.
-First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors, and depressive symptoms.
+First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors and depressive symptoms.
 By examining both aggregated measures (depression sum score and a composite precariousness index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
@@
 This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. 
-In some configurations, the system exhibits bistability, remaining stuck in a high-risk state unless external support exceeds a critical threshold.
-In others, it shows unistable behavior, responding more gradually to incremental changes.
-By integrating causal discovery with dynamic simulation, we aim to demonstrate how the internal structure of the depression–precariousness system shapes its susceptibility to change—and, by extension, how intervention strategies might be tailored to the architecture of underlying vulnerability.
+When the effects of depression symptoms onto the social precarity subsystem are weaker than the reverse path, there is higher sensitivity of depression to intervention, reflecting both a faster reduction post financial intervention but also quicker relapse once financial stresses are reestablished.
+By integrating causal discovery with dynamic simulation, we demonstrate that intervention strategies can be tailored to the underlying architecture of vulnerability.
 
@@
 
-We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study, which has been described in detail elsewhere [@stronks2013unravelling; @snijder2017cohort].
+We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study [@stronks2013unravelling; @snijder2017cohort].
 It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement].
@@
 To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures.
-These analyses led to the identification of five distinct precariousness factors—three domain-based and two reflecting recent stressors—each composed of multiple observed variables, as outlined below.
+These analyses led to the identification of five distinct precariousness factors: three reflecting structural domains—employment, housing, and social relationships—and two capturing recent life stressors—financial and relational.
+Each factor is composed of multiple observed variables, as outlined below.
 Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).
@@
 
-After preprocessing, the HELIUS dataset includes 21,628 individuals.
-In addition to the five identified precariousness factors, we computed an overall precariousness score by by summing these five factor scores, thereby capturing a composite measure of accumulated precariousness across domains.
+Following data cleaning and normalization procedures, the analytical sample comprises 21,628 individuals.
+In addition to the five identified precariousness factors, we computed an overall precariousness score by summing these five factor scores, thereby capturing a composite measure of accumulated precariousness across domains.
 Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores.
@@
   geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
-  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "precariousness", "PHQsum", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui")), scales = "free") +
+  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "precariousness", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui", "PHQsum")), scales = "free") +
   labs(x = "", y = "") +
   theme_minimal()
+
 ```
@@
 The PC algorithm is included for reference but relies on more restrictive assumptions, such as acyclicity and no unmeasured confounding.
-CCI, while capable of modeling cycles explicitly, comes with certain theoretical limitations.
+CCI, while capable of modeling cycles explicitly, has its own theoretical limitations; for example, its output may not fully preserve the d-separation properties of the true causal graph [@strobl2019].
 Results from both PC and CCI are provided in the [Appendix](#sec-appendix) ([@sec-cci; @sec-pc]) for completeness and comparison.
@@
 1.
-**Within each condition** (e.g., fixed $\alpha$, threshold, CI test, and algorithm), we identify the most frequently occurring edge type (e.g., arrowhead, circle, none) across bootstraps for each pair of variables (Step 2 in @fig-workflow).
+**Within each condition** (fixed $\alpha$, threshold, CI test, and algorithm), we identify the most frequently occurring edge type (e.g., arrowhead, circle, none) across bootstraps for each pair of variables (Step 2 in @fig-workflow).
 2.
-**Across conditions**, we then determine the most dominant edge type based on majority voting.
+**Across conditions**, we then determine the most dominant edge type based on its frequency.
 In cases where no edge type is clearly dominant (i.e., a tie), we mark the edge as uncertain, represented with a dashed line in the final graph (Step 3 in @fig-workflow).
@@
 These final graphs represent only the most stable and consistent relationships across a wide range of plausible parameterizations, enhancing both reliability and interpretability.
-To support transparency, each final graph is accompanied by a matrix summarizing the relative frequency of each edge type for every variable pair—enabling readers to assess the degree of uncertainty or agreement underlying each connection.
+To support transparency, each final graph is accompanied by a matrix summarizing the relative frequency of each edge type for every variable pair, enabling readers to assess the degree of uncertainty or agreement underlying each connection.
 
@@
 
-As illustrated in @fig-concept, the model includes three components: depression ($D$), represented by the PHQ-9 sum score, social precariousness ($P$), and external financial stress ($S$).
-Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S$ acting as a shared input.
+As illustrated in [@fig-concept], the model includes three components: social precariousness $P_i(t)$, depressive symptoms $D_i(t)$, and a static individual-specific external stressor $S_i$, representing financial stress.
+Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S_i$ acting as a shared input.
 These dynamics are formalized as a pair of linear stochastic differential equations (SDEs):
@@
 \begin{aligned}
-dP &= \lambda_P \left(\alpha_{PS} S + \alpha_{PD} D - P \right) dt + \sigma_{P} dW_{P} \\
-dD &= \lambda_D \left(\alpha_{DS} S + \alpha_{DP} P - D \right) dt + \sigma_{D} dW_{D} 
+dP_i(t) &= \lambda_P \big( \alpha_{PS} S_i + \alpha_{PD} D_i(t) - P_i(t) \big) \, dt + \sigma_P \, dW_{P,i}(t) \\
+dD_i(t) &= \lambda_D \big( \alpha_{DS} S_i + \alpha_{DP} P_i(t) - D_i(t) \big) \, dt + \sigma_D \, dW_{D,i}(t)
 \end{aligned}
@@
 
-Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress.
+Here, $D_i(t)$ denotes the depressive state and $P_i(t)$ denotes the level of social precariousness for individual $i$ at time $t$, with $S_i$ representing a static, individual-specific external stressor (financial stress).
+The parameters $\lambda_D$, $\lambda_P$, as well as the coefficients $\alpha$ and noise terms $\sigma$, are assumed to be constant across individuals and time.
+The parameters $\lambda_D$ and $\lambda_P$ govern the timescales over which $D_i(t)$ and $P_i(t)$ adjust toward their respective input-driven target states.
+The coefficients $\alpha_{DS}$ and $\alpha_{PS}$ quantify the direct effects of the external stressor $S_i$ on depression and precariousness, respectively, while $\alpha_{DP}$ and $\alpha_{PD}$ capture the mutual influence between the two dynamic processes.
+Stochastic fluctuations are introduced via the independent Wiener processes $dW_{D,i}(t)$ and $dW_{P,i}(t)$, scaled by their respective volatility parameters $\sigma_D$ and $\sigma_P$.
+This linear structure offers a transparent and analytically tractable framework for modeling how exogenous variation in financial stress may propagate through individual dynamics of depression and precariousness.
+Crucially, all parameters can be calibrated directly from empirical covariances and variances observed in the HELIUS cohort.
+The resulting steady-state distribution of the model aligns with empirical measures: PHQ-9 total score (PHQsum) for depression, and the derived social precariousness factor (P.soc) for social precarity.
+The full calibration procedure is described in [@sec-calibration].
+
+<!-- Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress.
 The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which each variable adjusts toward its input-driven target state.
@@
 Stochastic fluctuations are introduced through the noise terms $\sigma_P, dW_P$ and $\sigma_D, dW_D$, with $dW_P$ and $dW_D$ representing independent Wiener processes.
-This linear structure offers a transparent and analytically tractable framework for examining how variations in external financial stress influence the population-level dynamics of depression and social precariousness.
+-->
+<!-- This linear structure offers a transparent and analytically tractable framework for examining how variations in external financial stress influence the population-level dynamics of depression and social precariousness.
 Crucially, the model's simplicity enables direct calibration from empirical data: all parameters can be derived analytically from observed variances and covariances in the HELIUS cohort.
 The following  [@sec-calibration] details the procedure used to perform this calibration.
+-->
 
@@
 #| label: fig-concept
-#| fig-cap: "High-level conceptual model of causal structure linking social precariousness (P.soc) and depression (PHQsum), influenced by external financial stress (S.fin)."
+#| fig-cap: "High-level schematic of the dynamic model structure used in simulation.
+Arrows represent influence coefficients $\\alpha$, matching the terms in the system of stochastic differential equations.
+$D_i(t)$ and $P_i(t)$ denote depressive symptoms and social precariousness for individual $i$, with $S_i$ representing static financial stress."
 #| fig-align: center
@@
 
-knitr::include_graphics("img/conceptmodel.png")
+knitr::include_graphics("img/conceptmodel.pdf")
 ```
@@
 We fix the adjustment rates $\lambda_D = \lambda_P = 1$, which normalizes the time scale given the cross-sectional nature of the HELIUS data.
-To resolve the remaining degrees of freedom, we treat the feedback from precariousness to depression ($\alpha_{DP}$) as a free parameter, and solve algebraically for all other coefficients — including the direct effects of stress ($\alpha_{DS}, \alpha_{PS}$), the reverse feedback from depression to precariousness ($\alpha_{PD}$), and the noise terms ($\sigma_D, \sigma_P$).
+To resolve the remaining degree of freedom, we treat the feedback from precariousness to depression ($\alpha_{DP}$) as a free parameter, and we solve algebraically for all other coefficients — specifically the direct effects of stress ($\alpha_{DS}, \alpha_{PS}$), the reverse feedback from depression to precariousness ($\alpha_{PD}$), and the noise terms ($\sigma_D, \sigma_P$).
 These solutions are expressed as closed-form functions of the empirical covariances: $\mathrm{Var}(D)$, $\mathrm{Var}(P)$, $\mathrm{Var}(S)$, $\mathrm{Cov}(D, P)$, $\mathrm{Cov}(D, S)$, and $\mathrm{Cov}(P, S)$.
@@
 This structure creates a family of analytically calibrated models, each consistent with the same observed covariances but differing in their internal configuration.
-However, we constrain the space of valid solutions by imposing two empirical and theoretical conditions: (1) all structural parameters must be positive, reflecting the assumption that stress and precariousness contribute adversely to depression; and (2) the system must be dynamically stable, ensuring convergence to a stationary distribution.
-These constraints restrict the allowable range of $\alpha_{DP}$ to approximately $[0, 0.698]$ (see Appendix [@sec-analcalibration]).
+We constrain the space of valid solutions by imposing that all structural parameters must be positive, reflecting the assumption that stress and precariousness contribute adversely to depression.
+This constraint restricts the allowable range of $\alpha_{DP}$ to approximately $[0, 0.698]$ (see Appendix [@sec-analcalibration]).
 
 @fig-tradeoff illustrates how the remaining parameters adjust as $\alpha_{DP}$ increases to maintain consistency with the empirical covariances.
-As $\alpha_{DP}$ increases, $\alpha_{PD}$ must decline to preserve stability, since the product $\alpha_{DP} \cdot \alpha_{PD}$ determines the strength of the feedback loop and must remain below 1.
-This creates a compensatory trade-off: as influence from $P$ to $D$ strengthens, the influence from $D$ to $P$ weakens.
+As $\alpha_{DP}$ increases, $\alpha_{PD}$ tends to decrease, reflecting a compensatory trade-off between the mutual influences of depression and precariousness.
+This interplay helps preserve the model’s match with observed covariances, without requiring major shifts in external input strength.
 Meanwhile, the external coupling terms — $\alpha_{DS}$ and $\alpha_{PS}$ — also shift to absorb variance in a way that maintains alignment with the empirical data.
@@
 
-In each simulation, financial stress ($S$) is proportionally reduced toward the lowest observed value, modeling a population-wide intervention applied at varying intensities (0–100%).
+In each simulation, financial stress of individual $i$ ($S_i$) is proportionally reduced toward the lowest observed value, modeling a population-wide intervention applied at varying intensities (0–100%).
 For each model, we simulate 1000 virtual individuals whose baseline stress levels are sampled from the empirical distribution and scaled according to the intervention level.
@@
 While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint.
-Supporting this, the proportion matrix plot in panel (b) shows that *S.fin* has some probability of causally influencing either *P.emp* or *P.soc.*
+This pattern is further supported by the edge-type proportion matrix shown in panel (b), which aggregates edge inference outcomes across bootstrapped datasets.
 
@@
 Relational stress (*S.rel*) plays a more nuanced role, interacting with depression and *S.fin*  through a latent confounder.
-Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 indicates a considerable probability that this relationship is either absent, ambiguous, or potentially bidirectional.
+Its connection to *P.soc* is less definitive—the proportion matrix in @fig-sum-2 suggests that this relationship is frequently absent, ambiguous, or potentially bidirectional across bootstrap samples.
 Meanwhile, housing precariousness (*P.hou*) is not directly causally related to depression but is linked to *P.emp* and *S.fin*.
@@
 Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.
+
+The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI analysis, with one key distinction: CCI shows greater ambiguity in the directional relationship between *S.fin* and *PHQsum*: CCI assigns an arrowhead from *S.fin* to *PHQsum* in roughly 50% of bootstraps, while the remaining cases display an undirected (circle) endpoint.
+Overall, CCI introduces greater uncertainty in edge orientations relative to FCI, yielding more undirected endpoints and more frequent bidirectional (confounded) edges, which in turn contribute to a slightly higher overall count of arrowheads when directions are inferred.
+Despite this, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm.
+For example, both algorithms identify directional pathways from the stressors (*S.rel*, *S.fin*) to *P.soc*, and suggest that *P.soc* is not a cause of *PHQsum*, but more likely affected by it.
+*P.emp* is not inferred to cause any of its neighboring variables (*P.hou*, *PHQsum*, *S.fin*), and instead appears to be affected by them.
+*P.hou*, in turn, is connected to *P.emp* and *S.fin* with circle endpoints, indicating that it may be a potential source of influence rather than an outcome.
+These consistent pathways underscore the robustness of the core relational architecture across algorithms amid variation in directional confidence.
 
@@
 #| label: fig-sum
-#| fig-cap: "Resulting graph of precariousness factors and depression sum score using FCI and proportion of edge endpoint types."
+#| fig-cap: "Causal relationships between five precariousness factors and depression sum score (*PHQsum*), as estimated using the FCI algorithm across bootstrapped samples.
+**(a)** Aggregated causal graph showing the most frequent edge types across bootstraps.
+Dashed edges indicate ties, with a bold horizontal bar marking where the tie occurs.
+**(b)** Proportion matrix summarizing edge types across all bootstraps: each matrix cell $[i, j]$ indicates how often a given edge from variable $i$ to variable $j$ exhibited a particular edge mark (e.g., arrowhead, tail, or circle) adjacent to $j$.
+Green indicates arrowheads, coral arrowtails, blue circle marks, and gray absence of an edge.
+Colors are blended and shaded by frequency, with dominant symbols appearing more opaque.
+For example, matrix[$\\mathit{S.fin}$, $\\mathit{PHQsum}$] shows a green arrowhead ($>$) and matrix[$\\mathit{PHQsum}$, $\\mathit{S.fin}$] shows a coral arrowtail (\\textemdash), together indicating that $\\mathit{S.fin} \\rightarrow \\mathit{PHQsum}$ in the majority of bootstrapped graphs."
 #| fig-subcap: 
@@
 Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms.
-Unlike the sum score graph, which aggregates all symptoms into a single measure—potentially obscuring nuanced relationships—the symptom-level graph reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.
+Unlike the sum score graph, which aggregates all symptoms into a single measure, the symptom-level graph reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.
 
@@
 In the proportion matrix, this appears as a 50–50 tie between absence and arrowhead symbols.
-This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related.
+<!-- This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related.
+-->
 
@@
 This suggests that depressive symptoms, particularly *dep* and *glt*, may contribute to worsening social precariousness rather than the other way around.
-Additionally, *anh* also shows some probability of causally influencing *P.soc*, reinforcing the idea that social precariousness is more often a consequence rather than a driver of depressive symptoms.
-
-Other precariousness factors exhibit more uncertain but still notable relationships.
+Additionally, *anh* also shows signs of a potential causal influence on *P.soc*, reinforcing the idea that social precariousness is more often a consequence rather than a driver of depressive symptoms.
+
+Other precariousness factors exhibit more uncertainty but still notable relationships.
 *S.rel* shows weak but existing connections with *slp*, *glt*, and *app*, though directionality remains unclear in many cases.
@@
 
-To assess how internal structure influences the system’s responsiveness to external support, we simulate a set of linear models that differ in the assumed strength of influence from precariousness to depression, $\alpha_{DP}$.
-For each model, we generate a virtual population of 1,000 individuals and examine system behavior in two complementary ways: by analyzing average outcomes at selected time points following intervention onset, and by simulating full temporal trajectories for two representative cases at the extremes of the $\alpha_{DP}$ range.
-
-@fig-intervention summarizes how the internal structure of the model—specifically, variation in $\alpha_{DP}$—shapes the system’s short-term and long-term responses to external support.
-Panel A shows the mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500 after intervention onset, across a range of $\alpha_{DP}$ values.
-Each line represents a different model, color-coded from low (purple) to high (yellow) values of $\alpha_{DP}$.
+To assess how internal structure influences the system’s responsiveness to external support, we simulate a set of linear models that are consistent with the data but vary in the assumed strength of influence from precariousness to depression ($\alpha_{DP}$).
+For each model, we generate a virtual population of 1,000 individuals and simulate an external intervention that reduces financial stress ($S$), which acts as a shared input across individuals.
+This intervention defines the control input, while different values of $\alpha_{DP}$ represent alternative internal feedback scenarios.
+We examine system behavior in two complementary ways: (1) by analyzing average levels of depression and precariousness at selected time points following intervention onset, and (2) by simulating full temporal trajectories under sustained intervention followed by recovery, using two representative cases at the extremes of the $\alpha_{DP}$ range.
+
+@fig-intervention summarizes how variation in $\alpha_{DP}$ shapes system-level responses.
+Panel A shows the mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, with the x-axis indicating the strength of external support, defined as the proportional reduction of $S$ toward its minimum observed value.
+Each line represents a different scenario with a different $\alpha_{DP}$, color-coded from low (purple) to high (yellow).
 
 In the early phase of intervention ($t = 100$), depression and precariousness exhibit contrasting patterns.
-Depression declines more rapidly in models with lower $\alpha_{DP}$, where its dynamics are driven more directly by the external stressor $S$.
+Depression declines more rapidly in scenarios with lower $\alpha_{DP}$, where its dynamics are driven more directly by the external stressor $S$.
 As $\alpha_{DP}$ increases, $D$ becomes more dependent on precariousness ($P$), which itself adjusts more slowly—delaying the depressive response.
-Precariousness, by contrast, shows slightly faster early reductions in models with higher $\alpha_{DP}$.
+Precariousness, by contrast, shows slightly faster early reductions in scenarios with higher $\alpha_{DP}$.
 This reflects a shift in influence: as $\alpha_{DP}$ increases, $P$ becomes more directly responsive to $S$ (via increasing $\alpha_{PS}$) and less influenced by $D$, allowing it to adjust more promptly to changes in the stressor.
 
-Despite these short-term differences, all models ultimately converge to similar steady-state levels under sustained intervention.
+Despite these short-term differences, all scenarios ultimately converge to similar steady-state levels under sustained intervention.
 By time step 300, convergence is already substantial—though some variation remains, particularly in $P$, where gradients between models are still visible.
@@
 
-Panel B shows full time trajectories for two representative models: one with no coupling ($\alpha_{DP} = 0$) and one at the upper admissible limit ($\alpha_{DP} = 0.65$).
+Panel B shows full time trajectories for two representative cases: one with minimal internal feedback ($\alpha_{DP} = 0$) and one at the upper admissible limit ($\alpha_{DP} = 0.65$).
 In both, intervention begins at step 300 and ends at step 1500.
-In the uncoupled model, depression responds rapidly and rebounds shortly after support is removed.
-In the high-$\alpha_{DP}$ model, $D$ responds more gradually and recovers more slowly—reflecting its greater dependence on the slower-changing $P$.
-Precariousness declines in both models during intervention and returns to baseline afterward, but with a slightly sharper drop in the higher $\alpha_{DP}$ condition.
-This is because $P$ becomes more directly sensitive to $S$ as $\alpha_{PS}$ increases with $\alpha_{DP}$.
+When $\alpha_{DP} = 0$, depression responds rapidly to the reduction in stress and rebounds shortly after support is withdrawn.
+In contrast, when $\alpha_{DP}$ is high, $D$ responds more gradually and recovers more slowly—reflecting its increased dependence on the slower-changing $P$.
+Precariousness declines in both scenarios during intervention and returns to baseline afterward, but with a slightly sharper drop in the higher $\alpha_{DP}$ condition.
+This occurs because $P$ becomes more directly responsive to $S$, as $\alpha_{PS}$ increases with $\alpha_{DP}$.
 
@@
 This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact—and how their internal structure modulates responsiveness to intervention.
-Using cross-sectional data from the HELIUS study, we applied cycle-capable discovery algorithms to estimate plausible directional relationships among multiple domains of precariousness factors, and depressive symptoms.
+Using cross-sectional data from the HELIUS study, we applied cycle-capable discovery algorithms to estimate plausible directional relationships among multiple domains of precariousness factors and depressive symptoms.
 By analyzing both aggregate constructs and individual variables, we aimed to capture patterns that may be masked in either extreme of abstraction (i.e., purely aggregate or disaggregated analyses).
@@
 The causal analysis consistently identified financial stress as a key plausible driver of depression, appearing across both aggregated and disaggregated models. 
-At the symptom level, we found that sleep disturbance, depressed mood, guilt, and anhedonia were especially sensitive to external stressors.
-Among these, the symptom depressed mood emerged as a central hub: not only was it heavily influenced by other symptoms, it also exerted directional influence on social precariousness, suggesting a feedback loop in which worsening mood contributes to social disconnection, which in turn may exacerbate depressive symptoms.
-The CCI algorithm even suggested a possible cyclic relationship between depressed mood and overall precariousness, hinting at a structurally self-reinforcing mechanism.
+At the symptom level, several nodes—sleep disturbance, depressed mood, guilt, and anhedonia—emerged as especially sensitive to external influences.
+Among these, depressed mood played a particularly central role: it was influenced by other symptoms and, in turn, showed directional influence on social precariousness, suggesting a feedback mechanism wherein worsening mood fosters social withdrawal, which may subsequently exacerbate depressive symptoms.
+The CCI algorithm further suggested a potential cyclic link between depressed mood and overall precariousness, hinting that symptom and precariousness factors may co-evolve in mutually reinforcing ways.
 Symptom sleep disturbance also played a key initiating role.
@@
 This suggests the symptom sleep disturbance may function as an early warning signal or entry point into the depression–precariousness cycle, making it a promising target for preventive intervention.
-In contrast, Social precariousness more often appeared as an effect rather than a cause—receiving arrows from symptoms like depressed mood and guilt, but rarely sending them.
-This asymmetry implies that social precariousness may often reflect downstream consequences of worsening mental health rather than being its root cause.
-These patterns offer actionable insights for intervention.
-Addressing symptoms like sleep disturbance early may prevent escalation into broader depressive states or worsening precariousness.
-Intervening on depressed mood, as a central node with potential feedback effects, could also disrupt the self-reinforcing dynamics between mental health and social isolation.
+In contrast, social precariousness more often appeared as an effect rather than a cause—receiving arrows from symptoms like depressed mood and guilt, but rarely sending them.
+This asymmetry implies that social precariousness may often reflect downstream consequences of worsening mental health rather than being its root cause [@lund2010poverty; @compton2015social].
+While such relationships are complex and likely reciprocal over time, the directional patterns observed provide a useful foundation for hypothesis generation and targeted intervention.
+
+Taken together, these patterns offer actionable insights for intervention.
+Symptoms such as sleep disturbance, which show early and wide-reaching influence within the symptom network, may serve as effective targets for prevention—limiting escalation into broader depressive states or worsening precariousness.
+Similarly, depressed mood, which functions as a central hub influencing both symptom and precariousness variables, may offer a leverage point for disrupting reinforcing cycles of mental health decline and social isolation.
+At the same time, the broader causal structure points to interconnected feedback loops that span both psychological symptoms and external conditions such as financial and social precariousness—factors that often reflect the influence of structural inequalities (e.g., income instability, housing insecurity, or social marginalization).
+These complex dynamics suggest that single-point interventions may be insufficient to produce lasting change.
+Supporting this, recent simulation work by @park2025role demonstrates that feedback loops—even when modest in strength—can substantially reduce the long-term impact of isolated interventions, as system dynamics tend to re-amplify vulnerability through alternate pathways.
+In line with emerging perspectives in systems-oriented mental health research [@borsboom2017network], our findings underscore the need for multi-pronged strategies that address not only internal symptom dynamics but also the broader socioeconomic conditions that sustain vulnerability.
+Intervening simultaneously at multiple points of reinforcement may be essential to shift the system out of a maladaptive attractor and toward more stable trajectories of well-being.
 
@@
 Crucially, we analytically calibrated this model to match empirical covariances from the HELIUS data, ensuring that all parameter sets were observationally equivalent but differed in their internal configuration.
-We treated the parameter $\alpha_{DP}$—the effect of precariousness on depression—as a free variable, and derived all other parameters accordingly, yielding a family of valid models consistent with the data.
+We treated the parameter $\alpha_{DP}$—the effect of precariousness on depression—as a free variable and derived all other parameters accordingly, yielding a family of valid models consistent with the data.
 Simulation results showed that even with identical steady-state covariances, differences in internal structure produced systematic variation in the timing and profile of system responses to intervention.
-In models with lower $\alpha_{DP}$, depression responded more quickly to reduced stress because its dynamics were more directly coupled to the external input.
-In contrast, as $\alpha_{DP}$ increased and depression became more indirectly influenced—mediated by the slower-changing precariousness variable—its response was more gradual and persistent.
+When the effect of precariousness on depression was weaker, depression responded more rapidly to reductions in financial stress, reflecting a direct coupling to the external input.
+In contrast, as the effect of precariousness on depression increased and depression became more indirectly influenced—mediated by the slower-changing precariousness variable—its response was more gradual and persistent.
 Precariousness, while generally more directly responsive across all models, also showed subtle differences, with slightly faster early declines in configurations where it was more sensitive to external stress.
-These results demonstrate that intervention impact depends not only on the external conditions (e.g., stress reduction), but also on the internal structure of the system—how variables are coupled and through which pathways change flows.
-Tailoring interventions to system configuration—using more intensive or earlier support in systems with indirect or delayed pathways, and sustained support in systems with rapid but unstable responses—may help optimize mental health outcomes under conditions of social and economic adversity.
+These results highlight that intervention effectiveness is shaped not just by external input (e.g., stress reduction) but also by internal system architecture—how tightly different elements are coupled and through what paths change propagates.
+One-size-fits-all approaches may therefore fall short.
+From a policy or clinical perspective, more tailored interventions—targeting early-acting symptoms in some cases, or sustaining support over time in others—may be needed to achieve durable change, particularly in populations facing layered psychosocial vulnerabilities.
 
@@
 Finally, the structure of our dynamical model introduces important limitations.
-While the linear formulation offered analytical tractability and transparent calibration, it likely oversimplifies the true dynamics of depression and socioeconomic precariousness.
+The model focused on a restricted set of variables (depression, social precariousness, and financial stress).
+This design was intentionally simplified to isolate core mechanisms of interest, but it necessarily omits other relevant domains of precariousness, additional stressors, and potential unintended consequences of intervention.
+In addition, while the linear formulation offered analytical tractability and transparent calibration, it likely oversimplifies the true dynamics of depression and socioeconomic precariousness.
 A growing body of evidence suggests that psychosocial systems are fundamentally nonlinear.
@@
 These behaviors are fundamentally incompatible with linear dynamics, yet they may be essential for understanding why some individuals remain trapped in high-risk states while others recover more easily.
-Future modeling efforts should therefore incorporate mechanisms such as saturating input-response functions (e.g., sigmoidal or tanh forms) and state-dependent feedback (where coupling strength varies by system state).
-Incorporating such features would allow models to capture not just average effects, but also the critical conditions under which meaningful change becomes possible—clarifying whether some subgroups require forceful, threshold-crossing interventions, while others may benefit from sustained, moderate support.
-Ultimately, embracing nonlinear dynamics may be key to understanding whether different populations face genuinely distinct stability regimes—and whether effective intervention requires shifting the system across a critical boundary, rather than simply nudging it along its current trajectory.
+Future modeling efforts should therefore incorporate mechanisms such as saturating input-response functions (e.g., sigmoidal form) and state-dependent feedback (where coupling strength varies by system state).
+Incorporating such features would allow models to capture not just average effects but also the critical conditions under which meaningful change becomes possible—clarifying whether some subgroups require forceful, threshold-crossing interventions, while others may benefit from sustained, moderate support.
+
 <!-- Finally, the structure of the dynamical model introduces its own limitations.
@@
 <!-- Implications -->
-Our findings demonstrate that intervention effectiveness depends not only on external stressors, but also on the internal structure of the system.
+Our findings demonstrate that intervention effectiveness depends not only on external stressors but also on the internal structure of the system.
 While all models converge to the same long-term outcome, internal coupling shapes how quickly and through which pathways change unfolds.
-This highlights the importance of designing interventions that are sensitive to the system’s internal architecture.
-
-As public systems research has long shown, social and psychological systems do not always respond passively to external inputs; internal constraints can limit the impact of even well-intentioned policies [@wagenaar2013public; @desavigny2009systems].
+This highlights the importance of designing interventions that are sensitive to the system’s internal architecture, particularly when timely or rapid impact is a key concern.
+
+As public health research has long shown, social and psychological systems do not always respond passively to external inputs; internal constraints can limit the impact of even well-intentioned policies [@wagenaar2013public; @desavigny2009systems].
 By using causal discovery and computational modeling, researchers and policymakers can better identify these structural constraints and tailor interventions accordingly.
@@
 Causal discovery supports principled hypothesis generation from observational data, while dynamical modeling enables exploration of how hypothesized structures behave under external change.
-Together, these tools move us beyond static descriptions toward mechanistic insight—with direct relevance for real-world mental health policy and intervention design.
+Together, these tools move us beyond static descriptions toward mechanistic insight with direct relevance for real-world mental health policy and intervention design.
 

### 2025-07-27 draft_v1/draft_v2.tex
update draft
draft edit_Vitorfeedback

@@
 
-\date{2025-06-27}
+\date{2025-07-26}
 \begin{document}
@@
 with a mental health condition, with the highest disability-adjusted
-life years (DALYs) due to mental and addictive disorders concentrated in
+life years (DALYs) due to mental and addictive disorders present in
 high-income countries such as those in Northern Europe, North America,
@@
 effective responses. Mental health outcomes arise from complex,
-multi-level dynamics---intertwined with the social, economic, and
-spatial structures of urban environments---making both diagnosis and
+multi-level dynamics between the social, economic, and spatial
+structures of urban environments---making both diagnosis and
 intervention design deeply challenging (Van Der Wal et al., 2021).
@@
 literature highlights that poor mental health is not merely a personal
-issue, but one rooted in sustained socioeconomic stress and structural
+issue but one rooted in sustained socioeconomic stress and structural
 uncertainty.
@@
 precariousness as a multidimensional condition characterized by
-instability and lack of control across several life domains---including
+instability and lack of control across several life domains, including
 employment, finances, housing, social relations, and cultural belonging
@@
 insight, it remains difficult to identify priority targets for
-intervention or to anticipate unintended effects.
+intervention or to anticipate and quantify intended and unintended
+effects.
 
 This study aims to investigate the causal mechanisms linking
-precariousness and depression---and how those mechanisms shape system
+precariousness and depression and how those mechanisms shape system
 responses to external intervention. Specifically, we ask: \emph{How does
@@
 causal discovery algorithms to infer directional relationships among
-domain-specific precariousness factors, and depressive symptoms. By
+domain-specific precariousness factors and depressive symptoms. By
 examining both aggregated measures (depression sum score and a composite
@@
 properties---such as feedback strength and stochastic noise---govern
-responsiveness to intervention. In some configurations, the system
-exhibits bistability, remaining stuck in a high-risk state unless
-external support exceeds a critical threshold. In others, it shows
-unistable behavior, responding more gradually to incremental changes. By
-integrating causal discovery with dynamic simulation, we aim to
-demonstrate how the internal structure of the depression--precariousness
-system shapes its susceptibility to change---and, by extension, how
-intervention strategies might be tailored to the architecture of
-underlying vulnerability.
+responsiveness to intervention. When the effects of depression symptoms
+onto the social precarity subsystem are weaker than the reverse path,
+there is higher sensitivity of depression to intervention, reflecting
+both a faster reduction post financial intervention but also quicker
+relapse once financial stresses are reestablished. By integrating causal
+discovery with dynamic simulation, we demonstrate that intervention
+strategies can be tailored to the underlying architecture of
+vulnerability.
 
@@
 
-We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study,
-which has been described in detail elsewhere (Snijder et al., 2017;
-Stronks et al., 2013). It captures the diverse population of the city of
-Amsterdam by including the six main ethnic groups and provides
-comprehensive health and lifestyle data, including depressive symptoms
-as measured by the PHQ-9 (Galenkamp et al., 2017). To operationalize
-indicators of precariousness, we draw on the framework outlined in
-previous research (Elsenburg et al., 2025) and select a set of relevant
-variables. To ensure a robust representation of precariousness in our
-causal discovery models, we conducted exploratory analyses to identify
-consistent and meaningful data-driven factor structures. These analyses
-led to the identification of five distinct precariousness
-factors---three domain-based and two reflecting recent stressors---each
-composed of multiple observed variables, as outlined below. Additional
-details on the HELIUS study and the factor extraction process are
-provided in the \hyperref[sec-appendix]{Appendix}.
+We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study
+(Snijder et al., 2017; Stronks et al., 2013). It captures the diverse
+population of the city of Amsterdam by including the six main ethnic
+groups and provides comprehensive health and lifestyle data, including
+depressive symptoms as measured by the PHQ-9 (Galenkamp et al., 2017).
+To operationalize indicators of precariousness, we draw on the framework
+outlined in previous research (Elsenburg et al., 2025) and select a set
+of relevant variables. To ensure a robust representation of
+precariousness in our causal discovery models, we conducted exploratory
+analyses to identify consistent and meaningful data-driven factor
+structures. These analyses led to the identification of five distinct
+precariousness factors: three reflecting structural
+domains---employment, housing, and social relationships---and two
+capturing recent life stressors---financial and relational. Each factor
+is composed of multiple observed variables, as outlined below.
+Additional details on the HELIUS study and the factor extraction process
+are provided in the \hyperref[sec-appendix]{Appendix}.
 
@@
 
-After preprocessing, the HELIUS dataset includes 21,628 individuals. In
-addition to the five identified precariousness factors, we computed an
-overall precariousness score by by summing these five factor scores,
-thereby capturing a composite measure of accumulated precariousness
-across domains. Depression is represented using the PHQ-9 (Kroenke et
-al., 2001), both as a total sum score and as individual symptom scores.
-In the subsequent causal discovery analysis, we explore the relationship
-between depression and precariousness using both aggregated scores and
-their disaggregated components. See Figure~\ref{fig-dist} for the
-overall distributions of the variables used in the analysis.
+Following data cleaning and normalization procedures, the analytical
+sample comprises 21,628 individuals. In addition to the five identified
+precariousness factors, we computed an overall precariousness score by
+summing these five factor scores, thereby capturing a composite measure
+of accumulated precariousness across domains. Depression is represented
+using the PHQ-9 (Kroenke et al., 2001), both as a total sum score and as
+individual symptom scores. In the subsequent causal discovery analysis,
+we explore the relationship between depression and precariousness using
+both aggregated scores and their disaggregated components. See
+Figure~\ref{fig-dist} for the overall distributions of the variables
+used in the analysis.
 
@@
 restrictive assumptions, such as acyclicity and no unmeasured
-confounding. CCI, while capable of modeling cycles explicitly, comes
-with certain theoretical limitations. Results from both PC and CCI are
-provided in the \hyperref[sec-appendix]{Appendix}
+confounding. CCI, while capable of modeling cycles explicitly, has its
+own theoretical limitations; for example, its output may not fully
+preserve the d-separation properties of the true causal graph (Strobl,
+2019). Results from both PC and CCI are provided in the
+\hyperref[sec-appendix]{Appendix}
 (Section~\ref{sec-cci}, Section~\ref{sec-pc}) for completeness and
@@
 \item
-  \textbf{Within each condition} (e.g., fixed \(\alpha\), threshold, CI
-  test, and algorithm), we identify the most frequently occurring edge
-  type (e.g., arrowhead, circle, none) across bootstraps for each pair
-  of variables (Step 2 in Figure~\ref{fig-workflow}).
+  \textbf{Within each condition} (fixed \(\alpha\), threshold, CI test,
+  and algorithm), we identify the most frequently occurring edge type
+  (e.g., arrowhead, circle, none) across bootstraps for each pair of
+  variables (Step 2 in Figure~\ref{fig-workflow}).
 \item
   \textbf{Across conditions}, we then determine the most dominant edge
-  type based on majority voting. In cases where no edge type is clearly
+  type based on its frequency. In cases where no edge type is clearly
   dominant (i.e., a tie), we mark the edge as uncertain, represented
@@
 summarizing the relative frequency of each edge type for every variable
-pair---enabling readers to assess the degree of uncertainty or agreement
+pair, enabling readers to assess the degree of uncertainty or agreement
 underlying each connection.
@@
 As illustrated in Figure~\ref{fig-concept}, the model includes three
-components: depression (\(D\)), represented by the PHQ-9 sum score,
-social precariousness (\(P\)), and external financial stress (\(S\)).
-Depression and precariousness are modeled as coupled state variables
-that influence each other over time, with \(S\) acting as a shared
-input. These dynamics are formalized as a pair of linear stochastic
-differential equations (SDEs):
+components: social precariousness \(P_i(t)\), depressive symptoms
+\(D_i(t)\), and a static individual-specific external stressor \(S_i\),
+representing financial stress. Depression and precariousness are modeled
+as coupled state variables that influence each other over time, with
+\(S_i\) acting as a shared input. These dynamics are formalized as a
+pair of linear stochastic differential equations (SDEs):
 
@@
 
-Here, \(D\) denotes depression, \(P\) denotes social precariousness, and
-\(S\) is the external stressor associated with financial stress. The
-parameters \(\lambda_D\) and \(\lambda_P\) control the timescales over
-which each variable adjusts toward its input-driven target state. The
-coefficients \(\alpha_{PS}\) and \(\alpha_{DS}\) quantify the direct
-effects of the stressor on \(P\) and \(D\), while \(\alpha_{DP}\) and
-\(\alpha_{PD}\) represent the cross-influences between depression and
-precariousness. Stochastic fluctuations are introduced through the noise
-terms \(\sigma_P, dW_P\) and \(\sigma_D, dW_D\), with \(dW_P\) and
-\(dW_D\) representing independent Wiener processes. This linear
-structure offers a transparent and analytically tractable framework for
-examining how variations in external financial stress influence the
-population-level dynamics of depression and social precariousness.
-Crucially, the model's simplicity enables direct calibration from
-empirical data: all parameters can be derived analytically from observed
-variances and covariances in the HELIUS cohort. The following
-Section~\ref{sec-calibration} details the procedure used to perform this
-calibration.
+Here, \(D_i(t)\) denotes the depressive state and \(P_i(t)\) denotes the
+level of social precariousness for individual \(i\) at time \(t\), with
+\(S_i\) representing a static, individual-specific external stressor
+(financial stress). The parameters \(\lambda_D\), \(\lambda_P\), as well
+as the coefficients \(\alpha\) and noise terms \(\sigma\), are assumed
+to be constant across individuals and time. The parameters \(\lambda_D\)
+and \(\lambda_P\) govern the timescales over which \(D_i(t)\) and
+\(P_i(t)\) adjust toward their respective input-driven target states.
+The coefficients \(\alpha_{DS}\) and \(\alpha_{PS}\) quantify the direct
+effects of the external stressor \(S_i\) on depression and
+precariousness, respectively, while \(\alpha_{DP}\) and \(\alpha_{PD}\)
+capture the mutual influence between the two dynamic processes.
+Stochastic fluctuations are introduced via the independent Wiener
+processes \(dW_{D,i}(t)\) and \(dW_{P,i}(t)\), scaled by their
+respective volatility parameters \(\sigma_D\) and \(\sigma_P\). This
+linear structure offers a transparent and analytically tractable
+framework for modeling how exogenous variation in financial stress may
+propagate through individual dynamics of depression and precariousness.
+Crucially, all parameters can be calibrated directly from empirical
+covariances and variances observed in the HELIUS cohort. The resulting
+steady-state distribution of the model aligns with empirical measures:
+PHQ-9 total score (PHQsum) for depression, and the derived social
+precariousness factor (P.soc) for social precarity. The full calibration
+procedure is described in Section~\ref{sec-calibration}.
 
@@
 
-\includegraphics[width=0.4\textwidth,height=\textheight]{img/conceptmodel.png}
-
-}
-
-\caption{\label{fig-concept}High-level conceptual model of causal
-structure linking social precariousness (P.soc) and depression (PHQsum),
-influenced by external financial stress (S.fin).}
+\includegraphics[width=0.4\textwidth,height=\textheight]{img/conceptmodel.pdf}
+
+}
+
+\caption{\label{fig-concept}High-level schematic of the dynamic model
+structure used in simulation. Arrows represent influence coefficients
+\(\alpha\), matching the terms in the system of stochastic differential
+equations. \(D_i(t)\) and \(P_i(t)\) denote depressive symptoms and
+social precariousness for individual \(i\), with \(S_i\) representing
+static financial stress.}
 
@@
 normalizes the time scale given the cross-sectional nature of the HELIUS
-data. To resolve the remaining degrees of freedom, we treat the feedback
+data. To resolve the remaining degree of freedom, we treat the feedback
 from precariousness to depression (\(\alpha_{DP}\)) as a free parameter,
-and solve algebraically for all other coefficients --- including the
-direct effects of stress (\(\alpha_{DS}, \alpha_{PS}\)), the reverse
+and we solve algebraically for all other coefficients --- specifically
+the direct effects of stress (\(\alpha_{DS}, \alpha_{PS}\)), the reverse
 feedback from depression to precariousness (\(\alpha_{PD}\)), and the
@@
 consistent with the same observed covariances but differing in their
-internal configuration. However, we constrain the space of valid
-solutions by imposing two empirical and theoretical conditions: (1) all
-structural parameters must be positive, reflecting the assumption that
-stress and precariousness contribute adversely to depression; and (2)
-the system must be dynamically stable, ensuring convergence to a
-stationary distribution. These constraints restrict the allowable range
-of \(\alpha_{DP}\) to approximately \([0, 0.698]\) (see Appendix
+internal configuration. We constrain the space of valid solutions by
+imposing that all structural parameters must be positive, reflecting the
+assumption that stress and precariousness contribute adversely to
+depression. This constraint restricts the allowable range of
+\(\alpha_{DP}\) to approximately \([0, 0.698]\) (see Appendix
 Section~\ref{sec-analcalibration}).
@@
 empirical covariances. As \(\alpha_{DP}\) increases, \(\alpha_{PD}\)
-must decline to preserve stability, since the product
-\(\alpha_{DP} \cdot \alpha_{PD}\) determines the strength of the
-feedback loop and must remain below 1. This creates a compensatory
-trade-off: as influence from \(P\) to \(D\) strengthens, the influence
-from \(D\) to \(P\) weakens. Meanwhile, the external coupling terms ---
-\(\alpha_{DS}\) and \(\alpha_{PS}\) --- also shift to absorb variance in
-a way that maintains alignment with the empirical data. Notably, the
-noise amplitudes \(\sigma_D\) and \(\sigma_P\) decline slightly as
-\(\alpha_{DP}\) increases. This indicates that more of the observed
-variability is captured by deterministic interactions, with less left to
-be explained by stochastic fluctuations.
+tends to decrease, reflecting a compensatory trade-off between the
+mutual influences of depression and precariousness. This interplay helps
+preserve the model's match with observed covariances, without requiring
+major shifts in external input strength. Meanwhile, the external
+coupling terms --- \(\alpha_{DS}\) and \(\alpha_{PS}\) --- also shift to
+absorb variance in a way that maintains alignment with the empirical
+data. Notably, the noise amplitudes \(\sigma_D\) and \(\sigma_P\)
+decline slightly as \(\alpha_{DP}\) increases. This indicates that more
+of the observed variability is captured by deterministic interactions,
+with less left to be explained by stochastic fluctuations.
 
@@
 
-In each simulation, financial stress (\(S\)) is proportionally reduced
-toward the lowest observed value, modeling a population-wide
-intervention applied at varying intensities (0--100\%). For each model,
-we simulate 1000 virtual individuals whose baseline stress levels are
-sampled from the empirical distribution and scaled according to the
-intervention level.
+In each simulation, financial stress of individual \(i\) (\(S_i\)) is
+proportionally reduced toward the lowest observed value, modeling a
+population-wide intervention applied at varying intensities (0--100\%).
+For each model, we simulate 1000 virtual individuals whose baseline
+stress levels are sampled from the empirical distribution and scaled
+according to the intervention level.
 
@@
 emerges as a potential cause, as indicated by its circle edge endpoint.
-Supporting this, the proportion matrix plot in panel (b) shows that
-\emph{S.fin} has some probability of causally influencing either
-\emph{P.emp} or \emph{P.soc.}
+This pattern is further supported by the edge-type proportion matrix
+shown in panel (b), which aggregates edge inference outcomes across
+bootstrapped datasets.
 
@@
 connection to \emph{P.soc} is less definitive---the proportion matrix in
-Figure~\ref{fig-sum-2} indicates a considerable probability that this
-relationship is either absent, ambiguous, or potentially bidirectional.
-Meanwhile, housing precariousness (\emph{P.hou}) is not directly
-causally related to depression but is linked to \emph{P.emp} and
-\emph{S.fin}. While \emph{P.emp} and \emph{S.fin} are identified as
+Figure~\ref{fig-sum-2} suggests that this relationship is frequently
+absent, ambiguous, or potentially bidirectional across bootstrap
+samples. Meanwhile, housing precariousness (\emph{P.hou}) is not
+directly causally related to depression but is linked to \emph{P.emp}
+and \emph{S.fin}. While \emph{P.emp} and \emph{S.fin} are identified as
 non-causes of \emph{P.hou}, it remains unclear whether \emph{P.hou}
@@
 
+The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
+largely consistent with those from the FCI analysis, with one key
+distinction: CCI shows greater ambiguity in the directional relationship
+between \emph{S.fin} and \emph{PHQsum}: CCI assigns an arrowhead from
+\emph{S.fin} to \emph{PHQsum} in roughly 50\% of bootstraps, while the
+remaining cases display an undirected (circle) endpoint. Overall, CCI
+introduces greater uncertainty in edge orientations relative to FCI,
+yielding more undirected endpoints and more frequent bidirectional
+(confounded) edges, which in turn contribute to a slightly higher
+overall count of arrowheads when directions are inferred. Despite this,
+the skeleton structure and overall causal directions derived from CCI
+align well with the results of the FCI algorithm. For example, both
+algorithms identify directional pathways from the stressors
+(\emph{S.rel}, \emph{S.fin}) to \emph{P.soc}, and suggest that
+\emph{P.soc} is not a cause of \emph{PHQsum}, but more likely affected
+by it. \emph{P.emp} is not inferred to cause any of its neighboring
+variables (\emph{P.hou}, \emph{PHQsum}, \emph{S.fin}), and instead
+appears to be affected by them. \emph{P.hou}, in turn, is connected to
+\emph{P.emp} and \emph{S.fin} with circle endpoints, indicating that it
+may be a potential source of influence rather than an outcome. These
+consistent pathways underscore the robustness of the core relational
+architecture across algorithms amid variation in directional confidence.
+
 \begin{figure}
@@
 
-\caption{\label{fig-sum}Resulting graph of precariousness factors and
-depression sum score using FCI and proportion of edge endpoint types.}
+\caption{\label{fig-sum}Causal relationships between five precariousness
+factors and depression sum score (\emph{PHQsum}), as estimated using the
+FCI algorithm across bootstrapped samples. \textbf{(a)} Aggregated
+causal graph showing the most frequent edge types across bootstraps.
+Dashed edges indicate ties, with a bold horizontal bar marking where the
+tie occurs. \textbf{(b)} Proportion matrix summarizing edge types across
+all bootstraps: each matrix cell \([i, j]\) indicates how often a given
+edge from variable \(i\) to variable \(j\) exhibited a particular edge
+mark (e.g., arrowhead, tail, or circle) adjacent to \(j\). Green
+indicates arrowheads, coral arrowtails, blue circle marks, and gray
+absence of an edge. Colors are blended and shaded by frequency, with
+dominant symbols appearing more opaque. For example,
+matrix{[}\(\mathit{S.fin}\), \(\mathit{PHQsum}\){]} shows a green
+arrowhead (\(>\)) and matrix{[}\(\mathit{PHQsum}\),
+\(\mathit{S.fin}\){]} shows a coral arrowtail (\textemdash), together
+indicating that \(\mathit{S.fin} \rightarrow \mathit{PHQsum}\) in the
+majority of bootstrapped graphs.}
 
@@
 Unlike the sum score graph, which aggregates all symptoms into a single
-measure---potentially obscuring nuanced relationships---the
-symptom-level graph reveals the heterogeneity in how specific symptoms
-(\emph{con} (concentration), \emph{slp} (sleep), \emph{ene} (energy),
-\emph{app} (appetite), \emph{mot} (motor), \emph{sui} (suicidal),
-\emph{anh} (anhedonia), \emph{glt} (guilt), and \emph{dep} (depressed
-mood)) are influenced by, and in turn influence, different forms of
-precariousness.
+measure, the symptom-level graph reveals the heterogeneity in how
+specific symptoms (\emph{con} (concentration), \emph{slp} (sleep),
+\emph{ene} (energy), \emph{app} (appetite), \emph{mot} (motor),
+\emph{sui} (suicidal), \emph{anh} (anhedonia), \emph{glt} (guilt), and
+\emph{dep} (depressed mood)) are influenced by, and in turn influence,
+different forms of precariousness.
 
@@
 the proportion matrix, this appears as a 50--50 tie between absence and
-arrowhead symbols. This systematic divergence illustrates the broader
-challenge of causal discovery in high-dimensional data: different
-statistical assumptions can lead to different conclusions about
-whether---and in which direction---variables are causally related.
+arrowhead symbols.
 
@@
 social precariousness rather than the other way around. Additionally,
-\emph{anh} also shows some probability of causally influencing
+\emph{anh} also shows signs of a potential causal influence on
 \emph{P.soc}, reinforcing the idea that social precariousness is more
@@
 
-Other precariousness factors exhibit more uncertain but still notable
+Other precariousness factors exhibit more uncertainty but still notable
 relationships. \emph{S.rel} shows weak but existing connections with
@@
 To assess how internal structure influences the system's responsiveness
-to external support, we simulate a set of linear models that differ in
-the assumed strength of influence from precariousness to depression,
-\(\alpha_{DP}\). For each model, we generate a virtual population of
-1,000 individuals and examine system behavior in two complementary ways:
-by analyzing average outcomes at selected time points following
-intervention onset, and by simulating full temporal trajectories for two
-representative cases at the extremes of the \(\alpha_{DP}\) range.
-
-Figure~\ref{fig-intervention} summarizes how the internal structure of
-the model---specifically, variation in \(\alpha_{DP}\)---shapes the
-system's short-term and long-term responses to external support. Panel A
-shows the mean levels of depression (\(D\)) and precariousness (\(P\))
-at time steps 100, 300, and 500 after intervention onset, across a range
-of \(\alpha_{DP}\) values. Each line represents a different model,
-color-coded from low (purple) to high (yellow) values of
-\(\alpha_{DP}\).
+to external support, we simulate a set of linear models that are
+consistent with the data but vary in the assumed strength of influence
+from precariousness to depression (\(\alpha_{DP}\)). For each model, we
+generate a virtual population of 1,000 individuals and simulate an
+external intervention that reduces financial stress (\(S\)), which acts
+as a shared input across individuals. This intervention defines the
+control input, while different values of \(\alpha_{DP}\) represent
+alternative internal feedback scenarios. We examine system behavior in
+two complementary ways: (1) by analyzing average levels of depression
+and precariousness at selected time points following intervention onset,
+and (2) by simulating full temporal trajectories under sustained
+intervention followed by recovery, using two representative cases at the
+extremes of the \(\alpha_{DP}\) range.
+
+Figure~\ref{fig-intervention} summarizes how variation in
+\(\alpha_{DP}\) shapes system-level responses. Panel A shows the mean
+levels of depression (\(D\)) and precariousness (\(P\)) at time steps
+100, 300, and 500, with the x-axis indicating the strength of external
+support, defined as the proportional reduction of \(S\) toward its
+minimum observed value. Each line represents a different scenario with a
+different \(\alpha_{DP}\), color-coded from low (purple) to high
+(yellow).
 
@@
 precariousness exhibit contrasting patterns. Depression declines more
-rapidly in models with lower \(\alpha_{DP}\), where its dynamics are
+rapidly in scenarios with lower \(\alpha_{DP}\), where its dynamics are
 driven more directly by the external stressor \(S\). As \(\alpha_{DP}\)
@@
 Precariousness, by contrast, shows slightly faster early reductions in
-models with higher \(\alpha_{DP}\). This reflects a shift in influence:
-as \(\alpha_{DP}\) increases, \(P\) becomes more directly responsive to
-\(S\) (via increasing \(\alpha_{PS}\)) and less influenced by \(D\),
-allowing it to adjust more promptly to changes in the stressor.
-
-Despite these short-term differences, all models ultimately converge to
-similar steady-state levels under sustained intervention. By time step
-300, convergence is already substantial---though some variation remains,
-particularly in \(P\), where gradients between models are still visible.
-By step 500, trajectories have nearly fully aligned across all
+scenarios with higher \(\alpha_{DP}\). This reflects a shift in
+influence: as \(\alpha_{DP}\) increases, \(P\) becomes more directly
+responsive to \(S\) (via increasing \(\alpha_{PS}\)) and less influenced
+by \(D\), allowing it to adjust more promptly to changes in the
+stressor.
+
+Despite these short-term differences, all scenarios ultimately converge
+to similar steady-state levels under sustained intervention. By time
+step 300, convergence is already substantial---though some variation
+remains, particularly in \(P\), where gradients between models are still
+visible. By step 500, trajectories have nearly fully aligned across all
 configurations, consistent with the shared equilibrium structure of the
@@
 
-Panel B shows full time trajectories for two representative models: one
-with no coupling (\(\alpha_{DP} = 0\)) and one at the upper admissible
-limit (\(\alpha_{DP} = 0.65\)). In both, intervention begins at step 300
-and ends at step 1500. In the uncoupled model, depression responds
-rapidly and rebounds shortly after support is removed. In the
-high-\(\alpha_{DP}\) model, \(D\) responds more gradually and recovers
-more slowly---reflecting its greater dependence on the slower-changing
-\(P\). Precariousness declines in both models during intervention and
+Panel B shows full time trajectories for two representative cases: one
+with minimal internal feedback (\(\alpha_{DP} = 0\)) and one at the
+upper admissible limit (\(\alpha_{DP} = 0.65\)). In both, intervention
+begins at step 300 and ends at step 1500. When \(\alpha_{DP} = 0\),
+depression responds rapidly to the reduction in stress and rebounds
+shortly after support is withdrawn. In contrast, when \(\alpha_{DP}\) is
+high, \(D\) responds more gradually and recovers more
+slowly---reflecting its increased dependence on the slower-changing
+\(P\). Precariousness declines in both scenarios during intervention and
 returns to baseline afterward, but with a slightly sharper drop in the
-higher \(\alpha_{DP}\) condition. This is because \(P\) becomes more
-directly sensitive to \(S\) as \(\alpha_{PS}\) increases with
+higher \(\alpha_{DP}\) condition. This occurs because \(P\) becomes more
+directly responsive to \(S\), as \(\alpha_{PS}\) increases with
 \(\alpha_{DP}\).
@@
 directional relationships among multiple domains of precariousness
-factors, and depressive symptoms. By analyzing both aggregate constructs
+factors and depressive symptoms. By analyzing both aggregate constructs
 and individual variables, we aimed to capture patterns that may be
@@
 plausible driver of depression, appearing across both aggregated and
-disaggregated models. At the symptom level, we found that sleep
-disturbance, depressed mood, guilt, and anhedonia were especially
-sensitive to external stressors. Among these, the symptom depressed mood
-emerged as a central hub: not only was it heavily influenced by other
-symptoms, it also exerted directional influence on social
-precariousness, suggesting a feedback loop in which worsening mood
-contributes to social disconnection, which in turn may exacerbate
-depressive symptoms. The CCI algorithm even suggested a possible cyclic
-relationship between depressed mood and overall precariousness, hinting
-at a structurally self-reinforcing mechanism. Symptom sleep disturbance
-also played a key initiating role. It showed directed connections to
-both social and employment precariousness and exhibited many outgoing
-edges within the symptom network. This suggests the symptom sleep
-disturbance may function as an early warning signal or entry point into
-the depression--precariousness cycle, making it a promising target for
-preventive intervention. In contrast, Social precariousness more often
-appeared as an effect rather than a cause---receiving arrows from
-symptoms like depressed mood and guilt, but rarely sending them. This
-asymmetry implies that social precariousness may often reflect
-downstream consequences of worsening mental health rather than being its
-root cause. These patterns offer actionable insights for intervention.
-Addressing symptoms like sleep disturbance early may prevent escalation
-into broader depressive states or worsening precariousness. Intervening
-on depressed mood, as a central node with potential feedback effects,
-could also disrupt the self-reinforcing dynamics between mental health
-and social isolation.
+disaggregated models. At the symptom level, several nodes---sleep
+disturbance, depressed mood, guilt, and anhedonia---emerged as
+especially sensitive to external influences. Among these, depressed mood
+played a particularly central role: it was influenced by other symptoms
+and, in turn, showed directional influence on social precariousness,
+suggesting a feedback mechanism wherein worsening mood fosters social
+withdrawal, which may subsequently exacerbate depressive symptoms. The
+CCI algorithm further suggested a potential cyclic link between
+depressed mood and overall precariousness, hinting that symptom and
+precariousness factors may co-evolve in mutually reinforcing ways.
+Symptom sleep disturbance also played a key initiating role. It showed
+directed connections to both social and employment precariousness and
+exhibited many outgoing edges within the symptom network. This suggests
+the symptom sleep disturbance may function as an early warning signal or
+entry point into the depression--precariousness cycle, making it a
+promising target for preventive intervention. In contrast, social
+precariousness more often appeared as an effect rather than a
+cause---receiving arrows from symptoms like depressed mood and guilt,
+but rarely sending them. This asymmetry implies that social
+precariousness may often reflect downstream consequences of worsening
+mental health rather than being its root cause (Compton \& Shim, 2015;
+Lund et al., 2010). While such relationships are complex and likely
+reciprocal over time, the directional patterns observed provide a useful
+foundation for hypothesis generation and targeted intervention.
+
+Taken together, these patterns offer actionable insights for
+intervention. Symptoms such as sleep disturbance, which show early and
+wide-reaching influence within the symptom network, may serve as
+effective targets for prevention---limiting escalation into broader
+depressive states or worsening precariousness. Similarly, depressed
+mood, which functions as a central hub influencing both symptom and
+precariousness variables, may offer a leverage point for disrupting
+reinforcing cycles of mental health decline and social isolation. At the
+same time, the broader causal structure points to interconnected
+feedback loops that span both psychological symptoms and external
+conditions such as financial and social precariousness---factors that
+often reflect the influence of structural inequalities (e.g., income
+instability, housing insecurity, or social marginalization). These
+complex dynamics suggest that single-point interventions may be
+insufficient to produce lasting change. Supporting this, recent
+simulation work by Park, Li, et al. (2025) demonstrates that feedback
+loops---even when modest in strength---can substantially reduce the
+long-term impact of isolated interventions, as system dynamics tend to
+re-amplify vulnerability through alternate pathways. In line with
+emerging perspectives in systems-oriented mental health research
+(Borsboom, 2017), our findings underscore the need for multi-pronged
+strategies that address not only internal symptom dynamics but also the
+broader socioeconomic conditions that sustain vulnerability. Intervening
+simultaneously at multiple points of reinforcement may be essential to
+shift the system out of a maladaptive attractor and toward more stable
+trajectories of well-being.
 
@@
 \(\alpha_{DP}\)---the effect of precariousness on depression---as a free
-variable, and derived all other parameters accordingly, yielding a
-family of valid models consistent with the data. Simulation results
-showed that even with identical steady-state covariances, differences in
-internal structure produced systematic variation in the timing and
-profile of system responses to intervention. In models with lower
-\(\alpha_{DP}\), depression responded more quickly to reduced stress
-because its dynamics were more directly coupled to the external input.
-In contrast, as \(\alpha_{DP}\) increased and depression became more
-indirectly influenced---mediated by the slower-changing precariousness
-variable---its response was more gradual and persistent. Precariousness,
-while generally more directly responsive across all models, also showed
-subtle differences, with slightly faster early declines in
-configurations where it was more sensitive to external stress. These
-results demonstrate that intervention impact depends not only on the
-external conditions (e.g., stress reduction), but also on the internal
-structure of the system---how variables are coupled and through which
-pathways change flows. Tailoring interventions to system
-configuration---using more intensive or earlier support in systems with
-indirect or delayed pathways, and sustained support in systems with
-rapid but unstable responses---may help optimize mental health outcomes
-under conditions of social and economic adversity.
+variable and derived all other parameters accordingly, yielding a family
+of valid models consistent with the data. Simulation results showed that
+even with identical steady-state covariances, differences in internal
+structure produced systematic variation in the timing and profile of
+system responses to intervention. When the effect of precariousness on
+depression was weaker, depression responded more rapidly to reductions
+in financial stress, reflecting a direct coupling to the external input.
+In contrast, as the effect of precariousness on depression increased and
+depression became more indirectly influenced---mediated by the
+slower-changing precariousness variable---its response was more gradual
+and persistent. Precariousness, while generally more directly responsive
+across all models, also showed subtle differences, with slightly faster
+early declines in configurations where it was more sensitive to external
+stress. These results highlight that intervention effectiveness is
+shaped not just by external input (e.g., stress reduction) but also by
+internal system architecture---how tightly different elements are
+coupled and through what paths change propagates. One-size-fits-all
+approaches may therefore fall short. From a policy or clinical
+perspective, more tailored interventions---targeting early-acting
+symptoms in some cases, or sustaining support over time in others---may
+be needed to achieve durable change, particularly in populations facing
+layered psychosocial vulnerabilities.
 
@@
 Finally, the structure of our dynamical model introduces important
-limitations. While the linear formulation offered analytical
-tractability and transparent calibration, it likely oversimplifies the
-true dynamics of depression and socioeconomic precariousness. A growing
-body of evidence suggests that psychosocial systems are fundamentally
-nonlinear. Empirical and theoretical work points to phenomena such as
-threshold effects, where improvements only occur after a tipping point
-is crossed (Leemput et al., 2014; Park et al., 2025); saturation, where
-the impact of support diminishes as basic needs are met (Cramer et al.,
-2010); and path dependence, where past adversity shapes sensitivity to
-future stressors (Wichers et al., 2016). These behaviors are
-fundamentally incompatible with linear dynamics, yet they may be
-essential for understanding why some individuals remain trapped in
-high-risk states while others recover more easily. Future modeling
-efforts should therefore incorporate mechanisms such as saturating
-input-response functions (e.g., sigmoidal or tanh forms) and
-state-dependent feedback (where coupling strength varies by system
-state). Incorporating such features would allow models to capture not
-just average effects, but also the critical conditions under which
+limitations. The model focused on a restricted set of variables
+(depression, social precariousness, and financial stress). This design
+was intentionally simplified to isolate core mechanisms of interest, but
+it necessarily omits other relevant domains of precariousness,
+additional stressors, and potential unintended consequences of
+intervention. In addition, while the linear formulation offered
+analytical tractability and transparent calibration, it likely
+oversimplifies the true dynamics of depression and socioeconomic
+precariousness. A growing body of evidence suggests that psychosocial
+systems are fundamentally nonlinear. Empirical and theoretical work
+points to phenomena such as threshold effects, where improvements only
+occur after a tipping point is crossed (Leemput et al., 2014; Park,
+Waldorp, et al., 2025); saturation, where the impact of support
+diminishes as basic needs are met (Cramer et al., 2010); and path
+dependence, where past adversity shapes sensitivity to future stressors
+(Wichers et al., 2016). These behaviors are fundamentally incompatible
+with linear dynamics, yet they may be essential for understanding why
+some individuals remain trapped in high-risk states while others recover
+more easily. Future modeling efforts should therefore incorporate
+mechanisms such as saturating input-response functions (e.g., sigmoidal
+form) and state-dependent feedback (where coupling strength varies by
+system state). Incorporating such features would allow models to capture
+not just average effects but also the critical conditions under which
 meaningful change becomes possible---clarifying whether some subgroups
 require forceful, threshold-crossing interventions, while others may
-benefit from sustained, moderate support. Ultimately, embracing
-nonlinear dynamics may be key to understanding whether different
-populations face genuinely distinct stability regimes---and whether
-effective intervention requires shifting the system across a critical
-boundary, rather than simply nudging it along its current trajectory.
+benefit from sustained, moderate support.
 
 Our findings demonstrate that intervention effectiveness depends not
-only on external stressors, but also on the internal structure of the
+only on external stressors but also on the internal structure of the
 system. While all models converge to the same long-term outcome,
@@
 unfolds. This highlights the importance of designing interventions that
-are sensitive to the system's internal architecture.
-
-As public systems research has long shown, social and psychological
+are sensitive to the system's internal architecture, particularly when
+timely or rapid impact is a key concern.
+
+As public health research has long shown, social and psychological
 systems do not always respond passively to external inputs; internal
@@
 behave under external change. Together, these tools move us beyond
-static descriptions toward mechanistic insight---with direct relevance
-for real-world mental health policy and intervention design.
+static descriptions toward mechanistic insight with direct relevance for
+real-world mental health policy and intervention design.
 
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-borsboom2017network}
+Borsboom, D. (2017). A network theory of mental disorders. \emph{World
+Psychiatry}, \emph{16}(1), 5--13.
+
 \bibitem[\citeproctext]{ref-butterworth2012role}
@@
 \emph{46}(4), 364--373.
+
+\bibitem[\citeproctext]{ref-compton2015social}
+Compton, M. T., \& Shim, R. S. (2015). The social determinants of mental
+health. \emph{Focus}, \emph{13}(4), 419--425.
 
@@
 
+\bibitem[\citeproctext]{ref-lund2010poverty}
+Lund, C., Breen, A., Flisher, A. J., Kakuma, R., Corrigall, J., Joska,
+J. A., Swartz, L., \& Patel, V. (2010). Poverty and common mental
+disorders in low and middle income countries: A systematic review.
+\emph{Social Science \& Medicine}, \emph{71}(3), 517--528.
+
 \bibitem[\citeproctext]{ref-mckee2017living}
@@
 38). Pearson Prentice Hall Upper Saddle River.
+
+\bibitem[\citeproctext]{ref-park2025role}
+Park, K., Li, X., Waldorp, L., Lees, M., \& Vasconcelos, V. V. (2025).
+\emph{The role of feedback loops in dynamical symptom networks}.
 
@@
 \item
-  \(\alpha_{DS}\) and \(alpha_{PS}\) capture the direct effects of
+  \(\alpha_{DS}\) and \(\alpha_{PS}\) capture the direct effects of
   \(S\),

### 2025-08-12 draft_v1/draft_v2.qmd
draft update
vitor_feedback_2nd

@@
 # Load the dataset 
-dat <- read_sav("../../data/HELIUS_LEONIE.sav")
+dat <- read_sav("../data/helius_dat/HELIUS_LEONIE.sav")
 
@@
 ### Depression as sum score
-The sum score graphs provide a high-level summary of how precariousness factors collectively influence overall depression severity, focusing on aggregated relationships.
-@fig-sum illustrates the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
-As described in [@sec-analysis], dashed edges indicate ties, and the edge endpoint at which the tie occurs is marked by a bold horizontal bar.
-
-The causal graph in panel (a) highlights key pathways in the relationships between precariousness factors and depression (*PHQsum*).
-Employment precariousness (*P.emp*) and social precariousness (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role.
+@fig-sum provides a high-level summary of how precariousness factors collectively influence and are influenced by overall depression severity, focusing on the causal relationships between precariousness factors (*P.hou*, *P.emp*, *P.soc*, *S.rel*, *S.fin*) and the depression sum score (*PHQsum*), as identified by the FCI algorithm.
+<!-- As described in [@sec-analysis], dashed edges indicate ties, and the edge endpoint at which the tie occurs is marked by a bold horizontal bar. -->
+
+<!-- The causal graph in panel (a) highlights key pathways in the relationships between precariousness factors and depression (*PHQsum*).  -->
+In panel (a) we see that employment precariousness (*P.emp*) and social precariousness (*P.soc*) are not identified as causes of depression, whereas financial stress (*S.fin*) appears to play a causal role.
 While *P.emp* and *P.soc* are generally not recognized as causes of other precariousness factors, *S.fin* emerges as a potential cause, as indicated by its circle edge endpoint.
@@
 
-The matrix plot can be interpreted such that the symbol in matrix[$i$, $j$] represents the relationship $i -~[symbol] \  j$.
+<!-- The matrix plot can be interpreted such that the symbol in matrix[$i$, $j$] represents the relationship $i -~[symbol] \  j$.
 For example, if matrix[$i$, $j$] is > and matrix[j, i] is $\circ$, then the inferred relationship is $i \circarrow j$.
@@
 The colors are blended based on the proportion of each edge type, with higher proportions increasing the opacity of the corresponding color, making dominant symbols more visually prominent.
+-->
 
@@
 
-The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI algorithm, with one exception: CCI does not identify *S.fin* as a cause of depression.
-Additionally, CCI introduces greater uncertainty in edge directions, with more ties mainly between circle and arrowhead endpoints, and tends to favor arrowheads more frequently than FCI.
-Despite these differences, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm, supporting the key pathways.
-
 The findings from the CCI algorithm (see [@fig-sum-cci]) are largely consistent with those from the FCI analysis, with one key distinction: CCI shows greater ambiguity in the directional relationship between *S.fin* and *PHQsum*: CCI assigns an arrowhead from *S.fin* to *PHQsum* in roughly 50% of bootstraps, while the remaining cases display an undirected (circle) endpoint.
 Overall, CCI introduces greater uncertainty in edge orientations relative to FCI, yielding more undirected endpoints and more frequent bidirectional (confounded) edges, which in turn contribute to a slightly higher overall count of arrowheads when directions are inferred.
-Despite this, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm.
-For example, both algorithms identify directional pathways from the stressors (*S.rel*, *S.fin*) to *P.soc*, and suggest that *P.soc* is not a cause of *PHQsum*, but more likely affected by it.
-*P.emp* is not inferred to cause any of its neighboring variables (*P.hou*, *PHQsum*, *S.fin*), and instead appears to be affected by them.
+<!-- Despite this, the skeleton structure and overall causal directions derived from CCI align well with the results of the FCI algorithm. For example,  -->
+Both algorithms identify directional pathways from the stressors (*S.rel*, *S.fin*) to *P.soc* and suggest that *P.soc* is not a cause of *PHQsum*, but more likely affected by it.
+*P.emp* is not inferred to cause any of its neighboring variables (*P.hou*, *PHQsum*, *S.fin*) and instead appears to be affected by them.
 *P.hou*, in turn, is connected to *P.emp* and *S.fin* with circle endpoints, indicating that it may be a potential source of influence rather than an outcome.
-These consistent pathways underscore the robustness of the core relational architecture across algorithms amid variation in directional confidence.
+<!-- These consistent pathways underscore the robustness of the core relational architecture across algorithms amid variation in directional confidence. : " This is discussion point : Vitor-->
 
@@
 -->
+
+Causal relationships between five precariousness factors and depression sum score (PHQsum), as estimated using the FCI algorithm across bootstrapped samples.
+(a) Aggregated causal graph showing the most frequent edge types across bootstraps. Dashed edges indicate ties, with a bold horizontal bar marking where the tie occurs.
+(b) Proportion matrix summarizing edge types across all bootstraps.
+Each matrix cell $[i, j]$ represents how often an edge from variable $i$ to variable $j$ exhibited a particular edge mark—arrowhead, tail, circle, or no edge—adjacent to $j$.
+The edge mark in matrix$[i,j]$ reflects the orientation of the edge ending at $j$.
+For example, if matrix$[i, j]$ shows a green arrowhead ($>$) and matrix$[j, i]$ shows a blue circle ($\circ$), then the inferred relationship is $i \circarrow j$.
+Edge types are encoded by distinct colors: green for arrowheads, coral for arrowtails, blue for circles, and light gray for absence of an edge.
+Colors are blended and shaded by frequency, with greater opacity indicating higher confidence in the inferred edge type.
 
@@
 Moving from the sum score representation to the disaggregated symptom-level graph provides a more granular perspective on the causal relationships between precariousness factors and depression symptoms.
-Unlike the sum score graph, which aggregates all symptoms into a single measure, the symptom-level graph reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.
-
-The symptom-level graph in @fig-sym-1 reveals a more complex and interconnected structure than the sum score graph, reflecting the strong interdependence among symptoms and suggesting much presence of latent confounders, as indicated by numerous bidirectional edges.
-As before, dashed edges denote ties, with specific ties marked by bold horizontal bars.
-In @fig-sym-2, the proportion matrix provides more insight into these directionalities.
-Here, opaque green dominates most symptom-to-symptom connections, indicating that arrowheads are the most frequently inferred edge type.
-However, blue regions, which correspond to circle endpoints, are particularly common for *anh*, *slp*, *ene*, and *sui*, reflecting uncertainty in the causal direction among these symptoms.
+Unlike the sum score graph, which aggregates all symptoms into a single measure, the symptom-level graph in @fig-sym reveals the heterogeneity in how specific symptoms (*con* (concentration), *slp* (sleep), *ene* (energy), *app* (appetite), *mot* (motor), *sui* (suicidal), *anh* (anhedonia), *glt* (guilt), and *dep* (depressed mood)) are influenced by, and in turn influence, different forms of precariousness.
+
+@fig-sym-1 reveals a strong interdependence among symptoms and suggests much presence of latent confounders, as indicated by numerous bidirectional edges. 
+<!-- As before, dashed edges denote ties, with specific ties marked by bold horizontal bars.  -->
+In @fig-sym-2, the proportion matrix indicates that arrowheads are the most frequently inferred edge type.
+However, circle endpoints are particularly common for *anh*, *slp*, *ene*, and *sui*, reflecting uncertainty in the causal direction among these symptoms.
 Notably, the only arrowtail connection appears between *dep* and *anh*, suggesting that anhedonia is a potential cause of depressed mood.
@@
 
-This discrepancy likely arises from a combination of factors. 
-As the analysis moves from aggregated to disaggregated variables, the number of CI tests increases substantially, leading to a loss of statistical power.
+<!-- This discrepancy likely arises from a combination of factors.  -->
+<!-- As the analysis moves from aggregated to disaggregated variables, the number of CI tests increases substantially, leading to a loss of statistical power.
 With more variables being conditioned on, the effective sample size per test diminishes, making it harder to detect subtle dependencies.
 Over-conditioning—controlling for too many variables—can also obscure real relationships, resulting in false negatives.
-Beyond these statistical power issues, the underlying assumptions of the two CI tests contribute to their divergence.
+-->
+<!-- Beyond these statistical power issues, the underlying assumptions of the two CI tests contribute to their divergence.
 RCoT is a nonparametric test that captures both linear and nonlinear dependencies without assuming a specific distributional form.
@@
 In contrast, the Gaussian CI test is based on partial correlations and assumes linear Gaussian relationships, making it more permissive and more likely to detect weak or spurious associations.
-As a result, it is common to see edges where Gaussian CI suggests a directional relationship, while RCoT indicates no edge.
+-->
+<!-- As a result, it is common to see edges where Gaussian CI suggests a directional relationship, while RCoT indicates no edge.
 In the proportion matrix, this appears as a 50–50 tie between absence and arrowhead symbols.
+-->
 <!-- This systematic divergence illustrates the broader challenge of causal discovery in high-dimensional data: different statistical assumptions can lead to different conclusions about whether—and in which direction—variables are causally related.
@@
 Other precariousness factors exhibit more uncertainty but still notable relationships.
-*S.rel* shows weak but existing connections with *slp*, *glt*, and *app*, though directionality remains unclear in many cases.
+*S.rel* is connected to *slp*, *glt*, and *app*, though directionality remains unclear in many cases.
 *P.emp* also connects to symptoms like *slp*, *mot*, and *con*, but, like *S.rel*, these relationships exhibit a 50/50 split in directionality, reflecting uncertainty in the inferred causal paths.
@@
 Considering both symptom-precariousness connections and symptom-level dynamics, *slp* emerges as a central symptom, given its strong ties to precariousness factors and its influential role within the symptom network.
-This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precariousness.
-On the other hand, *P.soc* serves as a key bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precariousness, reinforcing a self-sustaining dynamic between depression and precarious conditions.
+<!-- This suggests that sleep issues could be an initiating symptom, particularly sensitive to relational stress and employment precariousness.  -->
+*P.soc* appears to function as a bridge between the depression and precariousness subsystems, as depressive symptoms appear to feed back into social precariousness, reinforcing a self-sustaining dynamic between depression and precarious conditions.
 
@@
 
-Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum-1) shows more certainty in causal directions, with only a few edges resulting in ties, all of which involve the overall precariousness factor.
+Unlike the disaggregated graph in @fig-sym, the aggregated symptom-precariousness graph (@fig-presum-1) shows greater certainty in causal directions, with ties occurring only in edges involving the overall precariousness factor.
 This is even more evident in the proportion matrix (@fig-presum-2), where symptom-to-symptom interactions almost fully converge to a single edge type.
@@
 Examining the symptom-precariousness connections, we observe a stronger overall trend of depressive symptoms influencing precariousness, rather than the other way around.
-Specifically, *anh*, *dep*, and *slp* show a high probability of causing precariousness, whereas *app*, *glt*, and *mot* exhibit more uncertainty in directionality, often resulting in circle endpoints.
+Specifically, *anh*, *dep*, and *slp* frequently emerge as sources of influence on precariousness, whereas *app*, *glt*, and *mot* exhibit more uncertainty in directionality, often resulting in circle endpoints.
 This pattern suggests that when precariousness factors are aggregated, the dominant causal flow is from depressive symptoms to precariousness, rather than precariousness driving depression.
@@
 
-Overall, the aggregated precariousness graph supports the key findings from both the disaggregated analysis and the depression sum score analysis, providing greater certainty in several important patterns.
+Across multiple levels of analysis, our results point to a consistent and directional link between precariousness and depression.
+Financial stress (*S.fin*) emerges as a particularly robust and modifiable upstream driver: in both aggregated and disaggregated models, it shows strong and stable causal influence on depressive symptoms.
+By contrast, other forms of precariousness—particularly social precarity—more often appear as outcomes rather than antecedents of depression, frequently receiving rather than sending causal arrows.
+Among symptoms, *dep* and *slp* stand out for their centrality and connections to precariousness factors, with *dep* also implicated in potential feedback loops with social precarity (*P.soc*).
+These findings motivate the use of a simplified three-node model in the simulation—capturing depression, financial stress, and social precarity—as it preserves the key dynamic motifs observed in the causal graphs while enabling tractable exploration of system-level responses to stress-reducing interventions.
+
+<!-- Overall, the aggregated precariousness graph supports the key findings from both the disaggregated analysis and the depression sum score analysis, providing greater certainty in several important patterns.
 The results highlight the significant role of latent confounders in symptom-to-symptom interactions.
@@
 Together, these findings suggest that depression—especially symptoms like sleep disturbances and guilt—may contribute to sustaining precarious conditions, possibly forming a reciprocal cycle between depression and precariousness.
-
-<!-- As before, black solid lines indicate edges agreed upon by both FCI and CCI, while gray dashed lines represent edges where the two algorithms disagree.
-Circles at the endpoints of gray dashed edges denote unresolved directionality.
-Additionally, navy dashed edges highlight relationships that appear exclusively in graphs generated using Gaussian CI tests.
--->
-<!-- Compared to @fig-sym, this analysis reveals greater disagreement between FCI and CCI regarding edge directionality, as evidenced by more dashed edges.
-FCI tends to estimate more undirected edges (circles), while CCI more frequently assigns directed arrows between symptom-symptom relationships.
-For a detailed breakdown, see [@sec-propmatrix], which shows the proportion of symbols produced by each algorithm.
--->
-<!-- When examining overall precarity, we observe that more symptoms are now connected to the precarity score (*Precarity*) compared to the individual precarity factor analysis in @fig-sym.
-In addition to *glt*, *slp*, *anh*, and *sui*, which are previously shown to be connected to individual precarity factors, *mot*, *app*, and *dep* are now identified as having causal relationships with overall precarity.
-Furthermore, when using the Gaussian CI test, con is also shown to be causally related to overall precarity.
-Notably, all these connections are directed towards precarity (o→), indicating that overall precarity is not the cause of symptoms.
-The exception is the bidirectional edge between *sui* <-> *Precarity*, which suggests a connection via latent confounders.
--->
-<!-- FCI predominantly estimates directed arrows (→) for symptom-precarity relationships, suggesting that symptoms are driving precarity.
-With fewer variables (reduced from 14 to 9), this approach appears to have slightly more statistical power, enabling algorithms to estimate relationships with greater confidence compared to the previous analysis involving individual symptoms and individual precarity factors.
 -->
@@
 #| label: fig-presum
-#| fig-cap: "Resulting graph of precariousness sum score and individual depression symptoms using FCI and proportion of edge endpoint types."
+#| fig-cap: "Resulting graph of precariousness sum score and individual depression symptoms using FCI and proportion of edge endpoint types.
+As before, dashed edges denote ties, with specific ties marked by bold horizontal bars."
 #| fig-subcap: 
@@
 To assess how internal structure influences the system’s responsiveness to external support, we simulate a set of linear models that are consistent with the data but vary in the assumed strength of influence from precariousness to depression ($\alpha_{DP}$).
-For each model, we generate a virtual population of 1,000 individuals and simulate an external intervention that reduces financial stress ($S$), which acts as a shared input across individuals.
-This intervention defines the control input, while different values of $\alpha_{DP}$ represent alternative internal feedback scenarios.
+For each model, we generate a virtual population of 1,000 individuals and simulate an external intervention that reduces each individual’s financial stress ($S$) proportionally toward the minimum observed value in the dataset.
+While individuals begin with different baseline stress levels, the intervention applies a shared proportional reduction across the population.
+This intervention defines the external input, while variation in $\alpha_{DP}$ reflects different assumptions about internal feedback from precariousness to depression.
 We examine system behavior in two complementary ways: (1) by analyzing average levels of depression and precariousness at selected time points following intervention onset, and (2) by simulating full temporal trajectories under sustained intervention followed by recovery, using two representative cases at the extremes of the $\alpha_{DP}$ range.
 
-@fig-intervention summarizes how variation in $\alpha_{DP}$ shapes system-level responses.
-Panel A shows the mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, with the x-axis indicating the strength of external support, defined as the proportional reduction of $S$ toward its minimum observed value.
-Each line represents a different scenario with a different $\alpha_{DP}$, color-coded from low (purple) to high (yellow).
-
-In the early phase of intervention ($t = 100$), depression and precariousness exhibit contrasting patterns.
+@fig-intervention summarizes how variation in $\alpha_{DP}$ shapes system-level responses to external support.
+<!-- Panel A shows the mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, with the x-axis indicating the strength of external support, defined as the proportional reduction of $S$ toward its minimum observed value.
+-->
+<!-- Each line represents a different scenario with a different $\alpha_{DP}$, color-coded from low (purple) to high (yellow). -->
+In Panel A, during the early phase of intervention, depression and precariousness exhibit contrasting patterns.
 Depression declines more rapidly in scenarios with lower $\alpha_{DP}$, where its dynamics are driven more directly by the external stressor $S$.
@@
 Despite these short-term differences, all scenarios ultimately converge to similar steady-state levels under sustained intervention.
-By time step 300, convergence is already substantial—though some variation remains, particularly in $P$, where gradients between models are still visible.
-By step 500, trajectories have nearly fully aligned across all configurations, consistent with the shared equilibrium structure of the linear system.
+AS the systm approaches convergence, some variation remains, particularly in $P$, where gradients between models are still visible.
+<!-- By step 500, trajectories have nearly fully aligned across all configurations, consistent with the shared equilibrium structure of the linear system.  -->
 This convergence highlights that the key distinctions lie not in whether intervention is effective, but in how quickly and through which pathways change unfolds.
@@
 Panel B shows full time trajectories for two representative cases: one with minimal internal feedback ($\alpha_{DP} = 0$) and one at the upper admissible limit ($\alpha_{DP} = 0.65$).
-In both, intervention begins at step 300 and ends at step 1500.
+In both, there is a temporary intervention period.
 When $\alpha_{DP} = 0$, depression responds rapidly to the reduction in stress and rebounds shortly after support is withdrawn.
@@
 Precariousness declines in both scenarios during intervention and returns to baseline afterward, but with a slightly sharper drop in the higher $\alpha_{DP}$ condition. 
-This occurs because $P$ becomes more directly responsive to $S$, as $\alpha_{PS}$ increases with $\alpha_{DP}$.
+This occurs because $P$ becomes more directly responsive to $S$, as consistency with the data requires $\alpha_{PS}$ to increase with $\alpha_{DP}$ (see @fig-tradeoff).
+These parameter shifts reflect trade-offs needed to preserve the empirical covariances, with further details provided in Appendix @sec-analcalibration.
 
@@
 #|   System response to external intervention across variations in internal structure.
-#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across different values of $\alpha_{DP}$ and intervention strengths.
-#|   Models with lower $\alpha_{DP}$ show faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
+#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across varying values of $\alpha_{DP}$ and intervention strengths.
+The x-axis represents the strength of external support, operationalized as the proportional reduction of financial stress ($S$) toward its minimum observed value.
+Each line corresponds to a different model defined by a specific $\alpha_{DP}$ value, color-coded from low (purple) to high (yellow).
+Models with lower $\alpha_{DP}$ exhibit faster early reductions in $D$, while higher $\alpha_{DP}$ leads to more immediate declines in $P$.
+#|   
 #|   **Panel B**: Full time trajectories for two models—one with no coupling ($\alpha_{DP} = 0$) and one with strong coupling ($\alpha_{DP} = 0.65$)—illustrating system response and recovery following intervention onset (step 300) and withdrawal (step 1500).
@@
 
-<!-- Summmary & Interpretation of Findings -->
-This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact—and how their internal structure modulates responsiveness to intervention.
-Using cross-sectional data from the HELIUS study, we applied cycle-capable discovery algorithms to estimate plausible directional relationships among multiple domains of precariousness factors and depressive symptoms.
-By analyzing both aggregate constructs and individual variables, we aimed to capture patterns that may be masked in either extreme of abstraction (i.e., purely aggregate or disaggregated analyses).
-
-The causal analysis consistently identified financial stress as a key plausible driver of depression, appearing across both aggregated and disaggregated models. 
-At the symptom level, several nodes—sleep disturbance, depressed mood, guilt, and anhedonia—emerged as especially sensitive to external influences.
-Among these, depressed mood played a particularly central role: it was influenced by other symptoms and, in turn, showed directional influence on social precariousness, suggesting a feedback mechanism wherein worsening mood fosters social withdrawal, which may subsequently exacerbate depressive symptoms.
-The CCI algorithm further suggested a potential cyclic link between depressed mood and overall precariousness, hinting that symptom and precariousness factors may co-evolve in mutually reinforcing ways.
-Symptom sleep disturbance also played a key initiating role.
-It showed directed connections to both social and employment precariousness and exhibited many outgoing edges within the symptom network.
-This suggests the symptom sleep disturbance may function as an early warning signal or entry point into the depression–precariousness cycle, making it a promising target for preventive intervention.
-In contrast, social precariousness more often appeared as an effect rather than a cause—receiving arrows from symptoms like depressed mood and guilt, but rarely sending them.
-This asymmetry implies that social precariousness may often reflect downstream consequences of worsening mental health rather than being its root cause [@lund2010poverty; @compton2015social].
-While such relationships are complex and likely reciprocal over time, the directional patterns observed provide a useful foundation for hypothesis generation and targeted intervention.
-
-Taken together, these patterns offer actionable insights for intervention.
-Symptoms such as sleep disturbance, which show early and wide-reaching influence within the symptom network, may serve as effective targets for prevention—limiting escalation into broader depressive states or worsening precariousness.
-Similarly, depressed mood, which functions as a central hub influencing both symptom and precariousness variables, may offer a leverage point for disrupting reinforcing cycles of mental health decline and social isolation.
-At the same time, the broader causal structure points to interconnected feedback loops that span both psychological symptoms and external conditions such as financial and social precariousness—factors that often reflect the influence of structural inequalities (e.g., income instability, housing insecurity, or social marginalization).
-These complex dynamics suggest that single-point interventions may be insufficient to produce lasting change.
-Supporting this, recent simulation work by @park2025role demonstrates that feedback loops—even when modest in strength—can substantially reduce the long-term impact of isolated interventions, as system dynamics tend to re-amplify vulnerability through alternate pathways.
-In line with emerging perspectives in systems-oriented mental health research [@borsboom2017network], our findings underscore the need for multi-pronged strategies that address not only internal symptom dynamics but also the broader socioeconomic conditions that sustain vulnerability.
-Intervening simultaneously at multiple points of reinforcement may be essential to shift the system out of a maladaptive attractor and toward more stable trajectories of well-being.
-
-To explore how internal structure shapes response to external support, we developed a linear stochastic model of depression and precariousness, influenced by financial stress.
-Crucially, we analytically calibrated this model to match empirical covariances from the HELIUS data, ensuring that all parameter sets were observationally equivalent but differed in their internal configuration.
-We treated the parameter $\alpha_{DP}$—the effect of precariousness on depression—as a free variable and derived all other parameters accordingly, yielding a family of valid models consistent with the data.
-Simulation results showed that even with identical steady-state covariances, differences in internal structure produced systematic variation in the timing and profile of system responses to intervention.
-When the effect of precariousness on depression was weaker, depression responded more rapidly to reductions in financial stress, reflecting a direct coupling to the external input.
-In contrast, as the effect of precariousness on depression increased and depression became more indirectly influenced—mediated by the slower-changing precariousness variable—its response was more gradual and persistent.
-Precariousness, while generally more directly responsive across all models, also showed subtle differences, with slightly faster early declines in configurations where it was more sensitive to external stress.
-These results highlight that intervention effectiveness is shaped not just by external input (e.g., stress reduction) but also by internal system architecture—how tightly different elements are coupled and through what paths change propagates.
-One-size-fits-all approaches may therefore fall short.
-From a policy or clinical perspective, more tailored interventions—targeting early-acting symptoms in some cases, or sustaining support over time in others—may be needed to achieve durable change, particularly in populations facing layered psychosocial vulnerabilities.
-
-<!-- To examine how such structures shape responsiveness to support, we developed a simplified two-variable nonlinear dynamical model capturing the feedback between depression and precariousness, modulated by financial stress.
-Calibrated against empirical summary statistics from the HELIUS dataset, the model allowed us to simulate how these dynamics evolve under varying levels of stress reduction.
--->
-<!-- The simulations revealed that intervention effectiveness depends not just on the magnitude of external stressors, but on the internal configuration of the system itself.
-In bistable regimes—characterized by strong feedback and low noise—the system could remain stuck in a high-risk state unless an intervention crosses a critical threshold, tipping it into a healthier equilibrium.
-By contrast, unistable systems—driven by weaker coupling or greater noise—responded more gradually and proportionally to increasing support.
--->
-<!-- These dynamics carry practical policy implications.
-In bistable systems, even moderate interventions may have little effect unless they are forceful enough to tip the system toward a lower-risk state.
-This underscores the importance of early or intensive support when targeting self-reinforcing cycles of depression and precariousness.
-Unistable systems, by contrast, are more responsive to incremental support, suggesting that sustained, moderate interventions may be sufficient in such contexts.
--->
+This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact and how their internal structure modulates responsiveness to intervention.
+Consistent with prior work linking financial strain to psychological distress [@lund2010poverty; @ezzy1993unemployment], our causal analysis identified financial stress as a key upstream driver of depression—emerging across both aggregated and disaggregated models.
+This provides empirical support for longstanding hypotheses that economic stressors can act as modifiable entry points into broader cycles of psychosocial risk.
+
+At the symptom level, our analysis aligns with established theories of psychopathology but offers a more fine-grained map of potential causal relationships, generating concrete, testable hypotheses.
+It also points to possible limitations of prevailing models, as some observed symptom–symptom links may be shaped by shared confounders rather than direct causal influence.
+This suggests that, in some contexts, broader mechanisms or clusters of symptoms may provide more effective units of analysis and intervention than individual symptoms alone.
+Within this framework, sleep disturbance emerged as a potentially initiating symptom, with directed connections to both symptom-level and precariousness variables—mirroring longitudinal findings that identify sleep disruption as an early warning signal for depressive episodes and social withdrawal [@zawadzki2013rumination; @alvaro2013systematic].
+Depressed mood, another core symptom, occupied a central position in the network.
+Not only was it influenced by upstream symptoms like anhedonia and guilt, but it also showed directional influence on social precariousness—suggesting that internal psychological states may actively shape external vulnerability.
+The CCI algorithm further identified a potential feedback loop between depressed mood and overall precariousness, resonating with social drift theories in depression, which posit that declining mental health can lead to social exclusion, job instability, and housing insecurity, reinforcing depressive cycles [@compton2015social; @boschloo2015network].
+
+In contrast, social precariousness itself appeared more often as an effect than a cause, receiving arrows from symptoms like depressed mood and guilt but rarely sending them.
+This asymmetry supports prior observations that indicators of social vulnerability often reflect downstream consequences of poor mental health, though the relationships are likely bidirectional over time [@lund2010poverty].
+These patterns, while derived from Amsterdam-specific data, align with mental health research in other high-income city settings, where structural inequalities such as poverty, unemployment, and housing insecurity are persistent drivers of poor psychological outcomes [@fryers2003social; @weich1998poverty].
+Although the specific dynamics may vary by context, the parallels with broader literature suggest that similar mechanisms could operate in comparable settings.
+Nonetheless, replication in other populations will be important for assessing robustness and external validity.
+
+Taken together, these findings yield actionable insights for mental health intervention.
+Symptoms like sleep disturbance, which demonstrate early and broad influence across both symptom and precariousness networks, may serve as effective targets for prevention—potentially halting the progression into more entrenched depressive states or worsening precariousness.
+Similarly, intervening on central nodes like depressed mood may offer strategic leverage for breaking mutually reinforcing cycles of mental decline and social isolation.
+Importantly, interventions targeting such nodes may have ripple effects across domains.
+For instance, evidence suggests that treating core depressive symptoms can improve social functioning, employment engagement, and housing stability—particularly when combined with social or financial supports [@mcdaid2019economic; @lund2018social].
+These dual benefits underscore the value of integrated, cross-domain strategies that simultaneously address internal symptoms and external stressors.
+
+From a systems perspective, our results also highlight why single-point interventions may fail to produce lasting change.
+The broader causal structure reveals multiple interlocking feedback loops spanning both psychological symptoms and structural risk factors, which may re-amplify vulnerability even after temporary relief.
+Supporting this, recent simulation work by @park2025role shows that feedback loops, even when modest in strength, can substantially diminish the long-term impact of isolated interventions unless multiple reinforcing links are simultaneously disrupted.
+Our own simulations support this view: while all models converged to similar long-term outcomes under sustained support, differences in internal structure—specifically, the strength of the link from precariousness to depression—produced systematic variation in how quickly and through which pathways systems responded to intervention.
+Models with stronger internal coupling exhibited slower recovery dynamics and greater persistence of depression following stress reduction, suggesting that internal system architecture, not just external input, shapes the effectiveness of support.
+This supports growing calls for system-aware intervention design [@borsboom2017network].
+Rather than assuming uniform responses, more durable outcomes may require (1) targeting early-acting symptoms, (2) sustaining support where recovery is slow, and (3) combining structural and psychological interventions.
+Tailoring strategies to match the system’s architecture can improve effectiveness, particularly in populations with layered vulnerabilities.
+In this light, understanding causal architecture is not merely theoretical—it is foundational to designing more effective and equitable interventions.
+By identifying key modifiable drivers like financial stress and influential symptoms like sleep disturbance, and by modeling how system dynamics unfold under intervention, our study offers a flexible framework for improving mental health outcomes in complex social environments.
 
 <!-- Limitations -->
-That said, our study has several limitations.
-While the FCI and CCI algorithms are designed to recover causal structure from cross-sectional data—including feedback and latent confounding—many edge directions in our results remain ambiguous, as indicated by the prevalence of circle endpoints and bidirectional arrows, especially in weakly connected regions of the graph.
-These uncertainties were most evident in the links between individual symptoms and precariousness factors, where edge estimates varied across algorithms and CI tests.
-This likely reflects limited statistical signal in high-dimensional settings, where large conditioning sets reduce power and small effect sizes are harder to detect.
-Although the CCI algorithm is explicitly designed to detect cycles, we found little clear evidence of feedback: most candidate loops appeared only as bidirectional edges, which may reflect latent confounding or statistical imprecision rather than true cyclic structure.
-This could point to either an absence of strong feedback at the symptom level or limitations in the sensitivity of full-graph discovery approaches applied to densely interconnected variables.
-More focused strategies, such as local structure tests or theory-guided subgraph modeling, may help resolve these ambiguities.
-In addition, while our analyses relied solely on observational data, some structural uncertainties—particularly those involving latent confounding or bidirectional influence—may ultimately require interventional or semi-interventional designs.
-Extensions such as LLC [@hyttinen2012learning], NODAGS-Flow [@sethuraman2023nodags], or Bicycle [@rohbeck2024], which integrate observational and interventional data, may prove valuable where such data are available.
-
-Second, our findings showed sensitivity to the choice of CI test.
-The Gaussian CI test, which assumes linear relationships and Gaussian noise, often produced denser graphs—likely due to its tendency to overestimate weak or noisy linear associations.
+While the FCI and CCI algorithms are designed to recover causal structure from cross-sectional data—including feedback and latent confounding—many edge directions in our results remain ambiguous, particularly in links between individual symptoms and precariousness factors.
+These inconsistencies likely reflect limited statistical signal in high-dimensional settings, where large conditioning sets reduce power and small effect sizes are harder to detect.
+Moreover, our findings showed sensitivity to the choice of conditional independence (CI) test.
+The Gaussian CI test, which assumes linearity and Gaussian noise, often produced denser graphs, likely due to overestimating weak or noisy linear associations.
 In contrast, RCoT, a nonparametric alternative, avoids these assumptions but relies on a Gaussian radial basis function (RBF) kernel to assess dependence.
 While flexible in capturing nonlinear relationships, this kernel is optimized for smooth, continuous variables and may underperform with discrete or ordinal data.
-Its similarity estimates can be unreliable when applied to variables with discontinuities, potentially leading to missed dependencies—particularly relevant in the HELIUS dataset, where some variables are ordinal or mixed-type [@howlett2001radial].
+Its similarity estimates can be unreliable when applied to variables with discontinuities, potentially leading to missed dependencies.
+This is particularly relevant in the HELIUS dataset, where some variables are ordinal or mixed-type [@howlett2001radial].
 To address this, future work should consider adapting kernel-based CI tests to better handle mixed data structures—for example, by employing hybrid kernels that explicitly combine continuous and discrete similarity measures.
 This could improve the reliability of nonparametric causal discovery in settings with heterogeneous variable types.
+
+Although the CCI algorithm is explicitly designed to detect cycles, aside from the connection between precarity and depression in the aggregate precarity setting, we found little clear evidence of feedback: most candidate loops appeared only as bidirectional edges, which may reflect latent confounding or statistical imprecision.
+This could point to either an absence of strong feedback at the symptom and the precariousness factor levels or limitations in the sensitivity of full-graph discovery approaches applied to densely interconnected variables.
+More focused strategies, such as local structure tests or theory-guided subgraph modeling, may help resolve these ambiguities.
+Moreover, our analyses relied solely on observational data from the Amsterdam population.
+Some structural uncertainties, particularly those involving latent confounding or bidirectional influence, may ultimately require interventional or semi-interventional designs.
+Extensions such as LLC [@hyttinen2012learning], NODAGS-Flow [@sethuraman2023nodags], or Bicycle [@rohbeck2024], which integrate observational and interventional data, could be especially valuable where such data are available.
 
@@
 <!-- Implications -->
-Our findings demonstrate that intervention effectiveness depends not only on external stressors but also on the internal structure of the system.
-While all models converge to the same long-term outcome, internal coupling shapes how quickly and through which pathways change unfolds.
-This highlights the importance of designing interventions that are sensitive to the system’s internal architecture, particularly when timely or rapid impact is a key concern.
 
@@
 By using causal discovery and computational modeling, researchers and policymakers can better identify these structural constraints and tailor interventions accordingly.
-
 More broadly, this study offers a generalizable framework for investigating complex mental health dynamics.
@@
 Together, these tools move us beyond static descriptions toward mechanistic insight with direct relevance for real-world mental health policy and intervention design.
-
 Future research should build on this foundation by developing richer, more nuanced dynamical models grounded in longitudinal and interventional data.
-Incorporating time, multiple domains of precarity, and adaptive feedback processes would help assess whether the mechanisms proposed here—such as delayed response, saturation, or multi-stability—manifest in actual mental health trajectories.
+Incorporating time, multiple domains of precarity, and adaptive feedback processes would help assess whether the mechanisms proposed manifest in actual mental health trajectories.
 Ultimately, we hope this study encourages further integration of causal structure learning and systems modeling in efforts to understand and  improve mental health outcomes under conditions of socioeconomic precariousness.

### 2025-08-12 draft_v1/draft_v2.tex
vitor_feedback_2nd

@@
 
-\date{2025-07-26}
+\date{2025-08-12}
 \begin{document}
@@
 
-The sum score graphs provide a high-level summary of how precariousness
-factors collectively influence overall depression severity, focusing on
-aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
-relationships between precariousness factors (\emph{P.hou},
-\emph{P.emp}, \emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the
-depression sum score (\emph{PHQsum}), as identified by the FCI
-algorithm. As described in Section~\ref{sec-analysis}, dashed edges
-indicate ties, and the edge endpoint at which the tie occurs is marked
-by a bold horizontal bar.
-
-The causal graph in panel (a) highlights key pathways in the
-relationships between precariousness factors and depression
-(\emph{PHQsum}). Employment precariousness (\emph{P.emp}) and social
-precariousness (\emph{P.soc}) are not identified as causes of
+Figure~\ref{fig-sum} provides a high-level summary of how precariousness
+factors collectively influence and are influenced by overall depression
+severity, focusing on the causal relationships between precariousness
+factors (\emph{P.hou}, \emph{P.emp}, \emph{P.soc}, \emph{S.rel},
+\emph{S.fin}) and the depression sum score (\emph{PHQsum}), as
+identified by the FCI algorithm.
+
+In panel (a) we see that employment precariousness (\emph{P.emp}) and
+social precariousness (\emph{P.soc}) are not identified as causes of
 depression, whereas financial stress (\emph{S.fin}) appears to play a
@@
 bootstrapped datasets.
-
-The matrix plot can be interpreted such that the symbol in
-matrix{[}\(i\), \(j\){]} represents the relationship
-\(i -~[symbol] \  j\). For example, if matrix{[}\(i\), \(j\){]} is
-\textgreater{} and matrix{[}j, i{]} is \(\circ\), then the inferred
-relationship is \(i \circarrow j\). In the table, different edge types
-are represented by distinct colors: light gray for the absence of an
-edge, blue for circles, green for arrowheads, and coral for arrowtails.
-The colors are blended based on the proportion of each edge type, with
-higher proportions increasing the opacity of the corresponding color,
-making dominant symbols more visually prominent.
 
@@
 The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
-largely consistent with those from the FCI algorithm, with one
-exception: CCI does not identify \emph{S.fin} as a cause of depression.
-Additionally, CCI introduces greater uncertainty in edge directions,
-with more ties mainly between circle and arrowhead endpoints, and tends
-to favor arrowheads more frequently than FCI. Despite these differences,
-the skeleton structure and overall causal directions derived from CCI
-align well with the results of the FCI algorithm, supporting the key
-pathways.
-
-The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
 largely consistent with those from the FCI analysis, with one key
@@
 (confounded) edges, which in turn contribute to a slightly higher
-overall count of arrowheads when directions are inferred. Despite this,
-the skeleton structure and overall causal directions derived from CCI
-align well with the results of the FCI algorithm. For example, both
+overall count of arrowheads when directions are inferred. Both
 algorithms identify directional pathways from the stressors
-(\emph{S.rel}, \emph{S.fin}) to \emph{P.soc}, and suggest that
+(\emph{S.rel}, \emph{S.fin}) to \emph{P.soc} and suggest that
 \emph{P.soc} is not a cause of \emph{PHQsum}, but more likely affected
 by it. \emph{P.emp} is not inferred to cause any of its neighboring
-variables (\emph{P.hou}, \emph{PHQsum}, \emph{S.fin}), and instead
+variables (\emph{P.hou}, \emph{PHQsum}, \emph{S.fin}) and instead
 appears to be affected by them. \emph{P.hou}, in turn, is connected to
 \emph{P.emp} and \emph{S.fin} with circle endpoints, indicating that it
-may be a potential source of influence rather than an outcome. These
-consistent pathways underscore the robustness of the core relational
-architecture across algorithms amid variation in directional confidence.
+may be a potential source of influence rather than an outcome.
+
+Causal relationships between five precariousness factors and depression
+sum score (PHQsum), as estimated using the FCI algorithm across
+bootstrapped samples. (a) Aggregated causal graph showing the most
+frequent edge types across bootstraps. Dashed edges indicate ties, with
+a bold horizontal bar marking where the tie occurs. (b) Proportion
+matrix summarizing edge types across all bootstraps. Each matrix cell
+\([i, j]\) represents how often an edge from variable \(i\) to variable
+\(j\) exhibited a particular edge mark---arrowhead, tail, circle, or no
+edge---adjacent to \(j\). The edge mark in matrix\([i,j]\) reflects the
+orientation of the edge ending at \(j\). For example, if
+matrix\([i, j]\) shows a green arrowhead (\(>\)) and matrix\([j, i]\)
+shows a blue circle (\(\circ\)), then the inferred relationship is
+\(i \circarrow j\). Edge types are encoded by distinct colors: green for
+arrowheads, coral for arrowtails, blue for circles, and light gray for
+absence of an edge. Colors are blended and shaded by frequency, with
+greater opacity indicating higher confidence in the inferred edge type.
 
@@
 Unlike the sum score graph, which aggregates all symptoms into a single
-measure, the symptom-level graph reveals the heterogeneity in how
-specific symptoms (\emph{con} (concentration), \emph{slp} (sleep),
-\emph{ene} (energy), \emph{app} (appetite), \emph{mot} (motor),
-\emph{sui} (suicidal), \emph{anh} (anhedonia), \emph{glt} (guilt), and
-\emph{dep} (depressed mood)) are influenced by, and in turn influence,
-different forms of precariousness.
-
-The symptom-level graph in Figure~\ref{fig-sym-1} reveals a more complex
-and interconnected structure than the sum score graph, reflecting the
-strong interdependence among symptoms and suggesting much presence of
-latent confounders, as indicated by numerous bidirectional edges. As
-before, dashed edges denote ties, with specific ties marked by bold
-horizontal bars. In Figure~\ref{fig-sym-2}, the proportion matrix
-provides more insight into these directionalities. Here, opaque green
-dominates most symptom-to-symptom connections, indicating that
-arrowheads are the most frequently inferred edge type. However, blue
-regions, which correspond to circle endpoints, are particularly common
-for \emph{anh}, \emph{slp}, \emph{ene}, and \emph{sui}, reflecting
-uncertainty in the causal direction among these symptoms. Notably, the
-only arrowtail connection appears between \emph{dep} and \emph{anh},
-suggesting that anhedonia is a potential cause of depressed mood.
+measure, the symptom-level graph in Figure~\ref{fig-sym} reveals the
+heterogeneity in how specific symptoms (\emph{con} (concentration),
+\emph{slp} (sleep), \emph{ene} (energy), \emph{app} (appetite),
+\emph{mot} (motor), \emph{sui} (suicidal), \emph{anh} (anhedonia),
+\emph{glt} (guilt), and \emph{dep} (depressed mood)) are influenced by,
+and in turn influence, different forms of precariousness.
+
+Figure~\ref{fig-sym-1} reveals a strong interdependence among symptoms
+and suggests much presence of latent confounders, as indicated by
+numerous bidirectional edges. In Figure~\ref{fig-sym-2}, the proportion
+matrix indicates that arrowheads are the most frequently inferred edge
+type. However, circle endpoints are particularly common for \emph{anh},
+\emph{slp}, \emph{ene}, and \emph{sui}, reflecting uncertainty in the
+causal direction among these symptoms. Notably, the only arrowtail
+connection appears between \emph{dep} and \emph{anh}, suggesting that
+anhedonia is a potential cause of depressed mood.
 
@@
 edge.
-
-This discrepancy likely arises from a combination of factors. As the
-analysis moves from aggregated to disaggregated variables, the number of
-CI tests increases substantially, leading to a loss of statistical
-power. With more variables being conditioned on, the effective sample
-size per test diminishes, making it harder to detect subtle
-dependencies. Over-conditioning---controlling for too many
-variables---can also obscure real relationships, resulting in false
-negatives. Beyond these statistical power issues, the underlying
-assumptions of the two CI tests contribute to their divergence. RCoT is
-a nonparametric test that captures both linear and nonlinear
-dependencies without assuming a specific distributional form. However,
-it is more conservative and requires larger sample sizes, especially in
-high-dimensional settings. In contrast, the Gaussian CI test is based on
-partial correlations and assumes linear Gaussian relationships, making
-it more permissive and more likely to detect weak or spurious
-associations. As a result, it is common to see edges where Gaussian CI
-suggests a directional relationship, while RCoT indicates no edge. In
-the proportion matrix, this appears as a 50--50 tie between absence and
-arrowhead symbols.
 
@@
 Other precariousness factors exhibit more uncertainty but still notable
-relationships. \emph{S.rel} shows weak but existing connections with
-\emph{slp}, \emph{glt}, and \emph{app}, though directionality remains
-unclear in many cases. \emph{P.emp} also connects to symptoms like
-\emph{slp}, \emph{mot}, and \emph{con}, but, like \emph{S.rel}, these
-relationships exhibit a 50/50 split in directionality, reflecting
-uncertainty in the inferred causal paths. In line with the aggregated
-graph in Figure~\ref{fig-sum}, \emph{P.hou} does not appear to have any
-direct relationship with depression symptoms but consistently shows
-associations with \emph{P.emp} and \emph{S.fin}. The causal
-relationships among precariousness factors remain largely unchanged from
-the aggregated analysis, with \emph{S.fin} exhibiting the strongest
-tendency to influence other precariousness factors.
+relationships. \emph{S.rel} is connected to \emph{slp}, \emph{glt}, and
+\emph{app}, though directionality remains unclear in many cases.
+\emph{P.emp} also connects to symptoms like \emph{slp}, \emph{mot}, and
+\emph{con}, but, like \emph{S.rel}, these relationships exhibit a 50/50
+split in directionality, reflecting uncertainty in the inferred causal
+paths. In line with the aggregated graph in Figure~\ref{fig-sum},
+\emph{P.hou} does not appear to have any direct relationship with
+depression symptoms but consistently shows associations with
+\emph{P.emp} and \emph{S.fin}. The causal relationships among
+precariousness factors remain largely unchanged from the aggregated
+analysis, with \emph{S.fin} exhibiting the strongest tendency to
+influence other precariousness factors.
 
@@
 its strong ties to precariousness factors and its influential role
-within the symptom network. This suggests that sleep issues could be an
-initiating symptom, particularly sensitive to relational stress and
-employment precariousness. On the other hand, \emph{P.soc} serves as a
-key bridge between the depression and precariousness subsystems, as
-depressive symptoms appear to feed back into social precariousness,
-reinforcing a self-sustaining dynamic between depression and precarious
-conditions.
+within the symptom network. \emph{P.soc} appears to function as a bridge
+between the depression and precariousness subsystems, as depressive
+symptoms appear to feed back into social precariousness, reinforcing a
+self-sustaining dynamic between depression and precarious conditions.
 
@@
 Unlike the disaggregated graph in Figure~\ref{fig-sym}, the aggregated
-symptom-precariousness graph (Figure~\ref{fig-presum-1}) shows more
-certainty in causal directions, with only a few edges resulting in ties,
-all of which involve the overall precariousness factor. This is even
-more evident in the proportion matrix (Figure~\ref{fig-presum-2}), where
+symptom-precariousness graph (Figure~\ref{fig-presum-1}) shows greater
+certainty in causal directions, with ties occurring only in edges
+involving the overall precariousness factor. This is even more evident
+in the proportion matrix (Figure~\ref{fig-presum-2}), where
 symptom-to-symptom interactions almost fully converge to a single edge
@@
 than the other way around. Specifically, \emph{anh}, \emph{dep}, and
-\emph{slp} show a high probability of causing precariousness, whereas
-\emph{app}, \emph{glt}, and \emph{mot} exhibit more uncertainty in
-directionality, often resulting in circle endpoints. This pattern
+\emph{slp} frequently emerge as sources of influence on precariousness,
+whereas \emph{app}, \emph{glt}, and \emph{mot} exhibit more uncertainty
+in directionality, often resulting in circle endpoints. This pattern
 suggests that when precariousness factors are aggregated, the dominant
@@
 
-Overall, the aggregated precariousness graph supports the key findings
-from both the disaggregated analysis and the depression sum score
-analysis, providing greater certainty in several important patterns. The
-results highlight the significant role of latent confounders in
-symptom-to-symptom interactions. While certain stressor-related
-precariousness factors, particularly \emph{S.fin} and \emph{S.rel},
-appear to causally influence depressive symptoms, other precariousness
-factors are more likely driven by depression rather than acting as
-primary causes, suggesting that precariousness may function more as a
-consequence of depression than its source. Some symptoms, particularly
-\emph{glt}, \emph{slp}, and \emph{dep}, appear to influence precarious
-conditions while also occupying central roles within the symptom
-network. Among them, \emph{slp} and \emph{glt} emerge as influential
-symptoms within the depression network, whereas \emph{con} and
-\emph{dep}, despite their high connectivity, behave more as outcomes
-rather than primary drivers. Together, these findings suggest that
-depression---especially symptoms like sleep disturbances and guilt---may
-contribute to sustaining precarious conditions, possibly forming a
-reciprocal cycle between depression and precariousness.
+Across multiple levels of analysis, our results point to a consistent
+and directional link between precariousness and depression. Financial
+stress (\emph{S.fin}) emerges as a particularly robust and modifiable
+upstream driver: in both aggregated and disaggregated models, it shows
+strong and stable causal influence on depressive symptoms. By contrast,
+other forms of precariousness---particularly social precarity---more
+often appear as outcomes rather than antecedents of depression,
+frequently receiving rather than sending causal arrows. Among symptoms,
+\emph{dep} and \emph{slp} stand out for their centrality and connections
+to precariousness factors, with \emph{dep} also implicated in potential
+feedback loops with social precarity (\emph{P.soc}). These findings
+motivate the use of a simplified three-node model in the
+simulation---capturing depression, financial stress, and social
+precarity---as it preserves the key dynamic motifs observed in the
+causal graphs while enabling tractable exploration of system-level
+responses to stress-reducing interventions.
 
@@
 and individual depression symptoms using FCI and proportion of edge
-endpoint types.}
+endpoint types. As before, dashed edges denote ties, with specific ties
+marked by bold horizontal bars.}
 
@@
 generate a virtual population of 1,000 individuals and simulate an
-external intervention that reduces financial stress (\(S\)), which acts
-as a shared input across individuals. This intervention defines the
-control input, while different values of \(\alpha_{DP}\) represent
-alternative internal feedback scenarios. We examine system behavior in
-two complementary ways: (1) by analyzing average levels of depression
-and precariousness at selected time points following intervention onset,
-and (2) by simulating full temporal trajectories under sustained
-intervention followed by recovery, using two representative cases at the
-extremes of the \(\alpha_{DP}\) range.
+external intervention that reduces each individual's financial stress
+(\(S\)) proportionally toward the minimum observed value in the dataset.
+While individuals begin with different baseline stress levels, the
+intervention applies a shared proportional reduction across the
+population. This intervention defines the external input, while
+variation in \(\alpha_{DP}\) reflects different assumptions about
+internal feedback from precariousness to depression. We examine system
+behavior in two complementary ways: (1) by analyzing average levels of
+depression and precariousness at selected time points following
+intervention onset, and (2) by simulating full temporal trajectories
+under sustained intervention followed by recovery, using two
+representative cases at the extremes of the \(\alpha_{DP}\) range.
 
 Figure~\ref{fig-intervention} summarizes how variation in
-\(\alpha_{DP}\) shapes system-level responses. Panel A shows the mean
-levels of depression (\(D\)) and precariousness (\(P\)) at time steps
-100, 300, and 500, with the x-axis indicating the strength of external
-support, defined as the proportional reduction of \(S\) toward its
-minimum observed value. Each line represents a different scenario with a
-different \(\alpha_{DP}\), color-coded from low (purple) to high
-(yellow).
-
-In the early phase of intervention (\(t = 100\)), depression and
+\(\alpha_{DP}\) shapes system-level responses to external support. In
+Panel A, during the early phase of intervention, depression and
 precariousness exhibit contrasting patterns. Depression declines more
@@
 Despite these short-term differences, all scenarios ultimately converge
-to similar steady-state levels under sustained intervention. By time
-step 300, convergence is already substantial---though some variation
-remains, particularly in \(P\), where gradients between models are still
-visible. By step 500, trajectories have nearly fully aligned across all
-configurations, consistent with the shared equilibrium structure of the
-linear system. This convergence highlights that the key distinctions lie
-not in whether intervention is effective, but in how quickly and through
-which pathways change unfolds.
+to similar steady-state levels under sustained intervention. AS the
+systm approaches convergence, some variation remains, particularly in
+\(P\), where gradients between models are still visible. This
+convergence highlights that the key distinctions lie not in whether
+intervention is effective, but in how quickly and through which pathways
+change unfolds.
 
@@
 with minimal internal feedback (\(\alpha_{DP} = 0\)) and one at the
-upper admissible limit (\(\alpha_{DP} = 0.65\)). In both, intervention
-begins at step 300 and ends at step 1500. When \(\alpha_{DP} = 0\),
-depression responds rapidly to the reduction in stress and rebounds
-shortly after support is withdrawn. In contrast, when \(\alpha_{DP}\) is
-high, \(D\) responds more gradually and recovers more
-slowly---reflecting its increased dependence on the slower-changing
-\(P\). Precariousness declines in both scenarios during intervention and
-returns to baseline afterward, but with a slightly sharper drop in the
-higher \(\alpha_{DP}\) condition. This occurs because \(P\) becomes more
-directly responsive to \(S\), as \(\alpha_{PS}\) increases with
-\(\alpha_{DP}\).
+upper admissible limit (\(\alpha_{DP} = 0.65\)). In both, there is a
+temporary intervention period. When \(\alpha_{DP} = 0\), depression
+responds rapidly to the reduction in stress and rebounds shortly after
+support is withdrawn. In contrast, when \(\alpha_{DP}\) is high, \(D\)
+responds more gradually and recovers more slowly---reflecting its
+increased dependence on the slower-changing \(P\). Precariousness
+declines in both scenarios during intervention and returns to baseline
+afterward, but with a slightly sharper drop in the higher
+\(\alpha_{DP}\) condition. This occurs because \(P\) becomes more
+directly responsive to \(S\), as consistency with the data requires
+\(\alpha_{PS}\) to increase with \(\alpha_{DP}\) (see
+Figure~\ref{fig-tradeoff}). These parameter shifts reflect trade-offs
+needed to preserve the empirical covariances, with further details
+provided in Appendix Section~\ref{sec-analcalibration}.
 
@@
 \emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
-precariousness (\(P\)) at time steps 100, 300, and 500, across different
-values of \(\alpha_{DP}\) and intervention strengths. Models with lower
-\(\alpha_{DP}\) show faster early reductions in \(D\), while higher
-\(\alpha_{DP}\) leads to more immediate declines in \(P\). \textbf{Panel
-B}: Full time trajectories for two models---one with no coupling
-(\(\alpha_{DP} = 0\)) and one with strong coupling
-(\(\alpha_{DP} = 0.65\))---illustrating system response and recovery
-following intervention onset (step 300) and withdrawal (step 1500).
-Depression responds more gradually and persistently in the high
-\(\alpha_{DP}\) model, whereas precariousness adjusts more quickly.}}
+precariousness (\(P\)) at time steps 100, 300, and 500, across varying
+values of \(\alpha_{DP}\) and intervention strengths. The x-axis
+represents the strength of external support, operationalized as the
+proportional reduction of financial stress (\(S\)) toward its minimum
+observed value. Each line corresponds to a different model defined by a
+specific \(\alpha_{DP}\) value, color-coded from low (purple) to high
+(yellow). Models with lower \(\alpha_{DP}\) exhibit faster early
+reductions in \(D\), while higher \(\alpha_{DP}\) leads to more
+immediate declines in \(P\). \textbf{Panel B}: Full time trajectories
+for two models---one with no coupling (\(\alpha_{DP} = 0\)) and one with
+strong coupling (\(\alpha_{DP} = 0.65\))---illustrating system response
+and recovery following intervention onset (step 300) and withdrawal
+(step 1500). Depression responds more gradually and persistently in the
+high \(\alpha_{DP}\) model, whereas precariousness adjusts more
+quickly.}}
 
@@
 modeling to investigate how depression and socioeconomic precariousness
-interact---and how their internal structure modulates responsiveness to
-intervention. Using cross-sectional data from the HELIUS study, we
-applied cycle-capable discovery algorithms to estimate plausible
-directional relationships among multiple domains of precariousness
-factors and depressive symptoms. By analyzing both aggregate constructs
-and individual variables, we aimed to capture patterns that may be
-masked in either extreme of abstraction (i.e., purely aggregate or
-disaggregated analyses).
-
-The causal analysis consistently identified financial stress as a key
-plausible driver of depression, appearing across both aggregated and
-disaggregated models. At the symptom level, several nodes---sleep
-disturbance, depressed mood, guilt, and anhedonia---emerged as
-especially sensitive to external influences. Among these, depressed mood
-played a particularly central role: it was influenced by other symptoms
-and, in turn, showed directional influence on social precariousness,
-suggesting a feedback mechanism wherein worsening mood fosters social
-withdrawal, which may subsequently exacerbate depressive symptoms. The
-CCI algorithm further suggested a potential cyclic link between
-depressed mood and overall precariousness, hinting that symptom and
-precariousness factors may co-evolve in mutually reinforcing ways.
-Symptom sleep disturbance also played a key initiating role. It showed
-directed connections to both social and employment precariousness and
-exhibited many outgoing edges within the symptom network. This suggests
-the symptom sleep disturbance may function as an early warning signal or
-entry point into the depression--precariousness cycle, making it a
-promising target for preventive intervention. In contrast, social
-precariousness more often appeared as an effect rather than a
-cause---receiving arrows from symptoms like depressed mood and guilt,
-but rarely sending them. This asymmetry implies that social
-precariousness may often reflect downstream consequences of worsening
-mental health rather than being its root cause (Compton \& Shim, 2015;
-Lund et al., 2010). While such relationships are complex and likely
-reciprocal over time, the directional patterns observed provide a useful
-foundation for hypothesis generation and targeted intervention.
-
-Taken together, these patterns offer actionable insights for
-intervention. Symptoms such as sleep disturbance, which show early and
-wide-reaching influence within the symptom network, may serve as
-effective targets for prevention---limiting escalation into broader
-depressive states or worsening precariousness. Similarly, depressed
-mood, which functions as a central hub influencing both symptom and
-precariousness variables, may offer a leverage point for disrupting
-reinforcing cycles of mental health decline and social isolation. At the
-same time, the broader causal structure points to interconnected
-feedback loops that span both psychological symptoms and external
-conditions such as financial and social precariousness---factors that
-often reflect the influence of structural inequalities (e.g., income
-instability, housing insecurity, or social marginalization). These
-complex dynamics suggest that single-point interventions may be
-insufficient to produce lasting change. Supporting this, recent
-simulation work by Park, Li, et al. (2025) demonstrates that feedback
-loops---even when modest in strength---can substantially reduce the
-long-term impact of isolated interventions, as system dynamics tend to
-re-amplify vulnerability through alternate pathways. In line with
-emerging perspectives in systems-oriented mental health research
-(Borsboom, 2017), our findings underscore the need for multi-pronged
-strategies that address not only internal symptom dynamics but also the
-broader socioeconomic conditions that sustain vulnerability. Intervening
-simultaneously at multiple points of reinforcement may be essential to
-shift the system out of a maladaptive attractor and toward more stable
-trajectories of well-being.
-
-To explore how internal structure shapes response to external support,
-we developed a linear stochastic model of depression and precariousness,
-influenced by financial stress. Crucially, we analytically calibrated
-this model to match empirical covariances from the HELIUS data, ensuring
-that all parameter sets were observationally equivalent but differed in
-their internal configuration. We treated the parameter
-\(\alpha_{DP}\)---the effect of precariousness on depression---as a free
-variable and derived all other parameters accordingly, yielding a family
-of valid models consistent with the data. Simulation results showed that
-even with identical steady-state covariances, differences in internal
-structure produced systematic variation in the timing and profile of
-system responses to intervention. When the effect of precariousness on
-depression was weaker, depression responded more rapidly to reductions
-in financial stress, reflecting a direct coupling to the external input.
-In contrast, as the effect of precariousness on depression increased and
-depression became more indirectly influenced---mediated by the
-slower-changing precariousness variable---its response was more gradual
-and persistent. Precariousness, while generally more directly responsive
-across all models, also showed subtle differences, with slightly faster
-early declines in configurations where it was more sensitive to external
-stress. These results highlight that intervention effectiveness is
-shaped not just by external input (e.g., stress reduction) but also by
-internal system architecture---how tightly different elements are
-coupled and through what paths change propagates. One-size-fits-all
-approaches may therefore fall short. From a policy or clinical
-perspective, more tailored interventions---targeting early-acting
-symptoms in some cases, or sustaining support over time in others---may
-be needed to achieve durable change, particularly in populations facing
-layered psychosocial vulnerabilities.
-
-That said, our study has several limitations. While the FCI and CCI
-algorithms are designed to recover causal structure from cross-sectional
-data---including feedback and latent confounding---many edge directions
-in our results remain ambiguous, as indicated by the prevalence of
-circle endpoints and bidirectional arrows, especially in weakly
-connected regions of the graph. These uncertainties were most evident in
-the links between individual symptoms and precariousness factors, where
-edge estimates varied across algorithms and CI tests. This likely
-reflects limited statistical signal in high-dimensional settings, where
-large conditioning sets reduce power and small effect sizes are harder
-to detect. Although the CCI algorithm is explicitly designed to detect
-cycles, we found little clear evidence of feedback: most candidate loops
-appeared only as bidirectional edges, which may reflect latent
-confounding or statistical imprecision rather than true cyclic
-structure. This could point to either an absence of strong feedback at
-the symptom level or limitations in the sensitivity of full-graph
-discovery approaches applied to densely interconnected variables. More
-focused strategies, such as local structure tests or theory-guided
-subgraph modeling, may help resolve these ambiguities. In addition,
-while our analyses relied solely on observational data, some structural
-uncertainties---particularly those involving latent confounding or
-bidirectional influence---may ultimately require interventional or
-semi-interventional designs. Extensions such as LLC (Hyttinen et al.,
-2012), NODAGS-Flow (Sethuraman et al., 2023), or Bicycle (Rohbeck et
-al., 2024), which integrate observational and interventional data, may
-prove valuable where such data are available.
-
-Second, our findings showed sensitivity to the choice of CI test. The
-Gaussian CI test, which assumes linear relationships and Gaussian noise,
-often produced denser graphs---likely due to its tendency to
-overestimate weak or noisy linear associations. In contrast, RCoT, a
-nonparametric alternative, avoids these assumptions but relies on a
-Gaussian radial basis function (RBF) kernel to assess dependence. While
-flexible in capturing nonlinear relationships, this kernel is optimized
-for smooth, continuous variables and may underperform with discrete or
-ordinal data. Its similarity estimates can be unreliable when applied to
-variables with discontinuities, potentially leading to missed
-dependencies---particularly relevant in the HELIUS dataset, where some
-variables are ordinal or mixed-type (Howlett, 2001). To address this,
-future work should consider adapting kernel-based CI tests to better
-handle mixed data structures---for example, by employing hybrid kernels
-that explicitly combine continuous and discrete similarity measures.
-This could improve the reliability of nonparametric causal discovery in
+interact and how their internal structure modulates responsiveness to
+intervention. Consistent with prior work linking financial strain to
+psychological distress (Ezzy, 1993; Lund et al., 2010), our causal
+analysis identified financial stress as a key upstream driver of
+depression---emerging across both aggregated and disaggregated models.
+This provides empirical support for longstanding hypotheses that
+economic stressors can act as modifiable entry points into broader
+cycles of psychosocial risk.
+
+At the symptom level, our analysis aligns with established theories of
+psychopathology but offers a more fine-grained map of potential causal
+relationships, generating concrete, testable hypotheses. It also points
+to possible limitations of prevailing models, as some observed
+symptom--symptom links may be shaped by shared confounders rather than
+direct causal influence. This suggests that, in some contexts, broader
+mechanisms or clusters of symptoms may provide more effective units of
+analysis and intervention than individual symptoms alone. Within this
+framework, sleep disturbance emerged as a potentially initiating
+symptom, with directed connections to both symptom-level and
+precariousness variables---mirroring longitudinal findings that identify
+sleep disruption as an early warning signal for depressive episodes and
+social withdrawal (Alvaro et al., 2013; Zawadzki et al., 2013).
+Depressed mood, another core symptom, occupied a central position in the
+network. Not only was it influenced by upstream symptoms like anhedonia
+and guilt, but it also showed directional influence on social
+precariousness---suggesting that internal psychological states may
+actively shape external vulnerability. The CCI algorithm further
+identified a potential feedback loop between depressed mood and overall
+precariousness, resonating with social drift theories in depression,
+which posit that declining mental health can lead to social exclusion,
+job instability, and housing insecurity, reinforcing depressive cycles
+(Boschloo et al., 2015; Compton \& Shim, 2015).
+
+In contrast, social precariousness itself appeared more often as an
+effect than a cause, receiving arrows from symptoms like depressed mood
+and guilt but rarely sending them. This asymmetry supports prior
+observations that indicators of social vulnerability often reflect
+downstream consequences of poor mental health, though the relationships
+are likely bidirectional over time (Lund et al., 2010). These patterns,
+while derived from Amsterdam-specific data, align with mental health
+research in other high-income city settings, where structural
+inequalities such as poverty, unemployment, and housing insecurity are
+persistent drivers of poor psychological outcomes (Fryers et al., 2003;
+Weich \& Lewis, 1998). Although the specific dynamics may vary by
+context, the parallels with broader literature suggest that similar
+mechanisms could operate in comparable settings. Nonetheless,
+replication in other populations will be important for assessing
+robustness and external validity.
+
+Taken together, these findings yield actionable insights for mental
+health intervention. Symptoms like sleep disturbance, which demonstrate
+early and broad influence across both symptom and precariousness
+networks, may serve as effective targets for prevention---potentially
+halting the progression into more entrenched depressive states or
+worsening precariousness. Similarly, intervening on central nodes like
+depressed mood may offer strategic leverage for breaking mutually
+reinforcing cycles of mental decline and social isolation. Importantly,
+interventions targeting such nodes may have ripple effects across
+domains. For instance, evidence suggests that treating core depressive
+symptoms can improve social functioning, employment engagement, and
+housing stability---particularly when combined with social or financial
+supports (Lund et al., 2018; McDaid et al., 2019). These dual benefits
+underscore the value of integrated, cross-domain strategies that
+simultaneously address internal symptoms and external stressors.
+
+From a systems perspective, our results also highlight why single-point
+interventions may fail to produce lasting change. The broader causal
+structure reveals multiple interlocking feedback loops spanning both
+psychological symptoms and structural risk factors, which may re-amplify
+vulnerability even after temporary relief. Supporting this, recent
+simulation work by Park, Li, et al. (2025) shows that feedback loops,
+even when modest in strength, can substantially diminish the long-term
+impact of isolated interventions unless multiple reinforcing links are
+simultaneously disrupted. Our own simulations support this view: while
+all models converged to similar long-term outcomes under sustained
+support, differences in internal structure---specifically, the strength
+of the link from precariousness to depression---produced systematic
+variation in how quickly and through which pathways systems responded to
+intervention. Models with stronger internal coupling exhibited slower
+recovery dynamics and greater persistence of depression following stress
+reduction, suggesting that internal system architecture, not just
+external input, shapes the effectiveness of support. This supports
+growing calls for system-aware intervention design (Borsboom, 2017).
+Rather than assuming uniform responses, more durable outcomes may
+require (1) targeting early-acting symptoms, (2) sustaining support
+where recovery is slow, and (3) combining structural and psychological
+interventions. Tailoring strategies to match the system's architecture
+can improve effectiveness, particularly in populations with layered
+vulnerabilities. In this light, understanding causal architecture is not
+merely theoretical---it is foundational to designing more effective and
+equitable interventions. By identifying key modifiable drivers like
+financial stress and influential symptoms like sleep disturbance, and by
+modeling how system dynamics unfold under intervention, our study offers
+a flexible framework for improving mental health outcomes in complex
+social environments.
+
+While the FCI and CCI algorithms are designed to recover causal
+structure from cross-sectional data---including feedback and latent
+confounding---many edge directions in our results remain ambiguous,
+particularly in links between individual symptoms and precariousness
+factors. These inconsistencies likely reflect limited statistical signal
+in high-dimensional settings, where large conditioning sets reduce power
+and small effect sizes are harder to detect. Moreover, our findings
+showed sensitivity to the choice of conditional independence (CI) test.
+The Gaussian CI test, which assumes linearity and Gaussian noise, often
+produced denser graphs, likely due to overestimating weak or noisy
+linear associations. In contrast, RCoT, a nonparametric alternative,
+avoids these assumptions but relies on a Gaussian radial basis function
+(RBF) kernel to assess dependence. While flexible in capturing nonlinear
+relationships, this kernel is optimized for smooth, continuous variables
+and may underperform with discrete or ordinal data. Its similarity
+estimates can be unreliable when applied to variables with
+discontinuities, potentially leading to missed dependencies. This is
+particularly relevant in the HELIUS dataset, where some variables are
+ordinal or mixed-type (Howlett, 2001). To address this, future work
+should consider adapting kernel-based CI tests to better handle mixed
+data structures---for example, by employing hybrid kernels that
+explicitly combine continuous and discrete similarity measures. This
+could improve the reliability of nonparametric causal discovery in
 settings with heterogeneous variable types.
+
+Although the CCI algorithm is explicitly designed to detect cycles,
+aside from the connection between precarity and depression in the
+aggregate precarity setting, we found little clear evidence of feedback:
+most candidate loops appeared only as bidirectional edges, which may
+reflect latent confounding or statistical imprecision. This could point
+to either an absence of strong feedback at the symptom and the
+precariousness factor levels or limitations in the sensitivity of
+full-graph discovery approaches applied to densely interconnected
+variables. More focused strategies, such as local structure tests or
+theory-guided subgraph modeling, may help resolve these ambiguities.
+Moreover, our analyses relied solely on observational data from the
+Amsterdam population. Some structural uncertainties, particularly those
+involving latent confounding or bidirectional influence, may ultimately
+require interventional or semi-interventional designs. Extensions such
+as LLC (Hyttinen et al., 2012), NODAGS-Flow (Sethuraman et al., 2023),
+or Bicycle (Rohbeck et al., 2024), which integrate observational and
+interventional data, could be especially valuable where such data are
+available.
 
@@
 
-Our findings demonstrate that intervention effectiveness depends not
-only on external stressors but also on the internal structure of the
-system. While all models converge to the same long-term outcome,
-internal coupling shapes how quickly and through which pathways change
-unfolds. This highlights the importance of designing interventions that
-are sensitive to the system's internal architecture, particularly when
-timely or rapid impact is a key concern.
-
 As public health research has long shown, social and psychological
@@
 better identify these structural constraints and tailor interventions
-accordingly.
-
-More broadly, this study offers a generalizable framework for
-investigating complex mental health dynamics. Causal discovery supports
-principled hypothesis generation from observational data, while
+accordingly. More broadly, this study offers a generalizable framework
+for investigating complex mental health dynamics. Causal discovery
+supports principled hypothesis generation from observational data, while
 dynamical modeling enables exploration of how hypothesized structures
@@
 static descriptions toward mechanistic insight with direct relevance for
-real-world mental health policy and intervention design.
-
-Future research should build on this foundation by developing richer,
-more nuanced dynamical models grounded in longitudinal and
-interventional data. Incorporating time, multiple domains of precarity,
-and adaptive feedback processes would help assess whether the mechanisms
-proposed here---such as delayed response, saturation, or
-multi-stability---manifest in actual mental health trajectories.
-Ultimately, we hope this study encourages further integration of causal
-structure learning and systems modeling in efforts to understand and
-improve mental health outcomes under conditions of socioeconomic
-precariousness.
+real-world mental health policy and intervention design. Future research
+should build on this foundation by developing richer, more nuanced
+dynamical models grounded in longitudinal and interventional data.
+Incorporating time, multiple domains of precarity, and adaptive feedback
+processes would help assess whether the mechanisms proposed manifest in
+actual mental health trajectories. Ultimately, we hope this study
+encourages further integration of causal structure learning and systems
+modeling in efforts to understand and improve mental health outcomes
+under conditions of socioeconomic precariousness.
 
@@
 \begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-alvaro2013systematic}
+Alvaro, P. K., Roberts, R. M., \& Harris, J. K. (2013). A systematic
+review assessing bidirectionality between sleep disturbances, anxiety,
+and depression. \emph{Sleep}, \emph{36}(7), 1059--1068.
+
 \bibitem[\citeproctext]{ref-borsboom2017network}
@@
 Psychiatry}, \emph{16}(1), 5--13.
+
+\bibitem[\citeproctext]{ref-boschloo2015network}
+Boschloo, L., Van Borkulo, C. D., Rhemtulla, M., Keyes, K. M., Borsboom,
+D., \& Schoevers, R. A. (2015). The network structure of symptoms of the
+diagnostic and statistical manual of mental disorders. \emph{PloS One},
+\emph{10}(9), e0137621.
 
@@
 
+\bibitem[\citeproctext]{ref-epskamp2018personalized}
+Epskamp, S., Borkulo, C. D. van, Veen, D. C. van der, Servaas, M. N.,
+Isvoranu, A.-M., Riese, H., \& Cramer, A. O. (2018). Personalized
+network modeling in psychopathology: The importance of contemporaneous
+and temporal connections. \emph{Clinical Psychological Science},
+\emph{6}(3), 416--427.
+
+\bibitem[\citeproctext]{ref-ezzy1993unemployment}
+Ezzy, D. (1993). Unemployment and mental health: A critical review.
+\emph{Social Science \& Medicine}, \emph{37}(1), 41--52.
+
 \bibitem[\citeproctext]{ref-fone2014effect}
@@
 
+\bibitem[\citeproctext]{ref-fryers2003social}
+Fryers, T., Melzer, D., \& Jenkins, R. (2003). Social inequalities and
+the common mental disorders: A systematic review of the evidence.
+\emph{Social Psychiatry and Psychiatric Epidemiology}, \emph{38}(5),
+229--237.
+
 \bibitem[\citeproctext]{ref-galenkamp2017measurement}
@@
 
+\bibitem[\citeproctext]{ref-lund2018social}
+Lund, C., Brooke-Sumner, C., Baingana, F., Baron, E. C., Breuer, E.,
+Chandra, P., Haushofer, J., Herrman, H., Jordans, M., Kieling, C., et
+al. (2018). Social determinants of mental disorders and the sustainable
+development goals: A systematic review of reviews. \emph{The Lancet
+Psychiatry}, \emph{5}(4), 357--369.
+
+\bibitem[\citeproctext]{ref-mcdaid2019economic}
+McDaid, D., Park, A.-L., \& Wahlbeck, K. (2019). The economic case for
+the prevention of mental illness. \emph{Annual Review of Public Health},
+\emph{40}(1), 373--389.
+
 \bibitem[\citeproctext]{ref-mckee2017living}
@@
 
+\bibitem[\citeproctext]{ref-weich1998poverty}
+Weich, S., \& Lewis, G. (1998). Poverty, unemployment, and common mental
+disorders: Population based cohort study. \emph{Bmj}, \emph{317}(7151),
+115--119.
+
 \bibitem[\citeproctext]{ref-wichers2016critical}
@@
 \url{https://www.who.int/news-room/fact-sheets/detail/mental-disorders}
+
+\bibitem[\citeproctext]{ref-zawadzki2013rumination}
+Zawadzki, M. J., Graham, J. E., \& Gerin, W. (2013). Rumination and
+anxiety mediate the effect of loneliness on depressed mood and sleep
+quality in college students. \emph{Health Psychology}, \emph{32}(2),
+212.
 

### 2025-08-12 draft/draft_v2.qmd
update_folder

--- 2025-08-12 draft_v1/draft_v2.qmd
@@
     keep-tex: true
+    geometry: 
+      #- showframe
+      - top=40mm
+      - left=45mm
+      - right=45mm
+      - bottom=40mm
     template-partials:
@@
 
+\newgeometry{top=20mm, bottom=20mm, left=40mm, right=40mm}
+
 ```{r}
 #| label: fig-intervention
+#| fig-align: center
+#| out-width: 80%
 #| fig-cap: >
@@
 
-#| fig-align: center
-#| out-width: 100%
-
 knitr::include_graphics("img/intervention_linear.pdf")
@@
 
+\restoregeometry
+
 # Discussion
@@
 This study combined constraint-based causal discovery and dynamical modeling to investigate how depression and socioeconomic precariousness interact and how their internal structure modulates responsiveness to intervention.
-Consistent with prior work linking financial strain to psychological distress [@lund2010poverty; @ezzy1993unemployment], our causal analysis identified financial stress as a key upstream driver of depression—emerging across both aggregated and disaggregated models.
-This provides empirical support for longstanding hypotheses that economic stressors can act as modifiable entry points into broader cycles of psychosocial risk.
-
-At the symptom level, our analysis aligns with established theories of psychopathology but offers a more fine-grained map of potential causal relationships, generating concrete, testable hypotheses.
+Consistent with prior work linking financial strain to psychological distress [@lund2010poverty; @ezzy1993unemployment], our analysis identified financial stress as a key upstream driver of depression in both aggregated and disaggregated models.
+This supports longstanding hypotheses that economic stressors can act as modifiable entry points into broader cycles of psychosocial risk.
+
+At the symptom level, our analysis aligns with established theories of psychopathology but provides a more fine-grained map of potential causal relationships, generating concrete, testable hypotheses.
 It also points to possible limitations of prevailing models, as some observed symptom–symptom links may be shaped by shared confounders rather than direct causal influence.
@@
 This asymmetry supports prior observations that indicators of social vulnerability often reflect downstream consequences of poor mental health, though the relationships are likely bidirectional over time [@lund2010poverty].
-These patterns, while derived from Amsterdam-specific data, align with mental health research in other high-income city settings, where structural inequalities such as poverty, unemployment, and housing insecurity are persistent drivers of poor psychological outcomes [@fryers2003social; @weich1998poverty].
-Although the specific dynamics may vary by context, the parallels with broader literature suggest that similar mechanisms could operate in comparable settings.
-Nonetheless, replication in other populations will be important for assessing robustness and external validity.
+Although our findings are based on Amsterdam-specific data, they align with research in other high-income urban settings, where structural inequalities such as poverty, unemployment, and housing insecurity consistently predict poor psychological outcomes [@fryers2003social; @weich1998poverty].
+While local contexts may alter specific pathways, these parallels suggest that similar mechanisms could operate elsewhere, though replication in other populations is needed to assess robustness.
 
@@
 Symptoms like sleep disturbance, which demonstrate early and broad influence across both symptom and precariousness networks, may serve as effective targets for prevention—potentially halting the progression into more entrenched depressive states or worsening precariousness.
-Similarly, intervening on central nodes like depressed mood may offer strategic leverage for breaking mutually reinforcing cycles of mental decline and social isolation.
+Similarly, targeting symptoms such as depressed mood, which appear embedded in potential feedback loops linking psychological and social domains, may offer strategic leverage for breaking mutually reinforcing cycles of mental decline and social isolation.
 Importantly, interventions targeting such nodes may have ripple effects across domains.
@@
 
-From a systems perspective, our results also highlight why single-point interventions may fail to produce lasting change.
+Our results also highlight why single-point interventions may fail to produce lasting change.
 The broader causal structure reveals multiple interlocking feedback loops spanning both psychological symptoms and structural risk factors, which may re-amplify vulnerability even after temporary relief.
@@
 Our own simulations support this view: while all models converged to similar long-term outcomes under sustained support, differences in internal structure—specifically, the strength of the link from precariousness to depression—produced systematic variation in how quickly and through which pathways systems responded to intervention.
-Models with stronger internal coupling exhibited slower recovery dynamics and greater persistence of depression following stress reduction, suggesting that internal system architecture, not just external input, shapes the effectiveness of support.
-This supports growing calls for system-aware intervention design [@borsboom2017network].
+Stronger coupling slowed recovery and prolonged depression, showing that internal architecture can be as important as external input.
+This supports growing calls for systems-oriented mental health perspectives [@borsboom2017network].
 Rather than assuming uniform responses, more durable outcomes may require (1) targeting early-acting symptoms, (2) sustaining support where recovery is slow, and (3) combining structural and psychological interventions.
@@
 In this light, understanding causal architecture is not merely theoretical—it is foundational to designing more effective and equitable interventions. 
-By identifying key modifiable drivers like financial stress and influential symptoms like sleep disturbance, and by modeling how system dynamics unfold under intervention, our study offers a flexible framework for improving mental health outcomes in complex social environments.
+<!-- By identifying key modifiable drivers like financial stress and influential symptoms like sleep disturbance, and by modeling how system dynamics unfold under intervention, our study offers a flexible framework for improving mental health outcomes in complex social environments.
+-->
 

### 2025-08-12 draft/draft_v2.tex
update_folder

--- 2025-08-12 draft_v1/draft_v2.tex
@@
 \usepackage{xcolor}
+\usepackage[top=40mm,left=45mm,right=45mm,bottom=40mm]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \usepackage{amsmath}
+\usepackage{geometry}
+
 \usetikzlibrary{arrows.meta, shapes.geometric}
@@
 
+\newgeometry{top=20mm, bottom=20mm, left=40mm, right=40mm}
+
 \begin{figure}
@@
 
-\includegraphics{img/intervention_linear.pdf}
+\includegraphics[width=0.8\textwidth,height=\textheight]{img/intervention_linear.pdf}
 
@@
 
+\restoregeometry
+
 \section{Discussion}\label{discussion}
@@
 intervention. Consistent with prior work linking financial strain to
-psychological distress (Ezzy, 1993; Lund et al., 2010), our causal
-analysis identified financial stress as a key upstream driver of
-depression---emerging across both aggregated and disaggregated models.
-This provides empirical support for longstanding hypotheses that
-economic stressors can act as modifiable entry points into broader
-cycles of psychosocial risk.
+psychological distress (Ezzy, 1993; Lund et al., 2010), our analysis
+identified financial stress as a key upstream driver of depression in
+both aggregated and disaggregated models. This supports longstanding
+hypotheses that economic stressors can act as modifiable entry points
+into broader cycles of psychosocial risk.
 
 At the symptom level, our analysis aligns with established theories of
-psychopathology but offers a more fine-grained map of potential causal
+psychopathology but provides a more fine-grained map of potential causal
 relationships, generating concrete, testable hypotheses. It also points
@@
 downstream consequences of poor mental health, though the relationships
-are likely bidirectional over time (Lund et al., 2010). These patterns,
-while derived from Amsterdam-specific data, align with mental health
-research in other high-income city settings, where structural
-inequalities such as poverty, unemployment, and housing insecurity are
-persistent drivers of poor psychological outcomes (Fryers et al., 2003;
-Weich \& Lewis, 1998). Although the specific dynamics may vary by
-context, the parallels with broader literature suggest that similar
-mechanisms could operate in comparable settings. Nonetheless,
-replication in other populations will be important for assessing
-robustness and external validity.
+are likely bidirectional over time (Lund et al., 2010). Although our
+findings are based on Amsterdam-specific data, they align with research
+in other high-income urban settings, where structural inequalities such
+as poverty, unemployment, and housing insecurity consistently predict
+poor psychological outcomes (Fryers et al., 2003; Weich \& Lewis, 1998).
+While local contexts may alter specific pathways, these parallels
+suggest that similar mechanisms could operate elsewhere, though
+replication in other populations is needed to assess robustness.
 
@@
 halting the progression into more entrenched depressive states or
-worsening precariousness. Similarly, intervening on central nodes like
-depressed mood may offer strategic leverage for breaking mutually
-reinforcing cycles of mental decline and social isolation. Importantly,
-interventions targeting such nodes may have ripple effects across
-domains. For instance, evidence suggests that treating core depressive
-symptoms can improve social functioning, employment engagement, and
-housing stability---particularly when combined with social or financial
-supports (Lund et al., 2018; McDaid et al., 2019). These dual benefits
-underscore the value of integrated, cross-domain strategies that
-simultaneously address internal symptoms and external stressors.
-
-From a systems perspective, our results also highlight why single-point
-interventions may fail to produce lasting change. The broader causal
-structure reveals multiple interlocking feedback loops spanning both
-psychological symptoms and structural risk factors, which may re-amplify
-vulnerability even after temporary relief. Supporting this, recent
-simulation work by Park, Li, et al. (2025) shows that feedback loops,
-even when modest in strength, can substantially diminish the long-term
-impact of isolated interventions unless multiple reinforcing links are
-simultaneously disrupted. Our own simulations support this view: while
-all models converged to similar long-term outcomes under sustained
-support, differences in internal structure---specifically, the strength
-of the link from precariousness to depression---produced systematic
-variation in how quickly and through which pathways systems responded to
-intervention. Models with stronger internal coupling exhibited slower
-recovery dynamics and greater persistence of depression following stress
-reduction, suggesting that internal system architecture, not just
-external input, shapes the effectiveness of support. This supports
-growing calls for system-aware intervention design (Borsboom, 2017).
-Rather than assuming uniform responses, more durable outcomes may
-require (1) targeting early-acting symptoms, (2) sustaining support
-where recovery is slow, and (3) combining structural and psychological
-interventions. Tailoring strategies to match the system's architecture
-can improve effectiveness, particularly in populations with layered
-vulnerabilities. In this light, understanding causal architecture is not
-merely theoretical---it is foundational to designing more effective and
-equitable interventions. By identifying key modifiable drivers like
-financial stress and influential symptoms like sleep disturbance, and by
-modeling how system dynamics unfold under intervention, our study offers
-a flexible framework for improving mental health outcomes in complex
-social environments.
+worsening precariousness. Similarly, targeting symptoms such as
+depressed mood, which appear embedded in potential feedback loops
+linking psychological and social domains, may offer strategic leverage
+for breaking mutually reinforcing cycles of mental decline and social
+isolation. Importantly, interventions targeting such nodes may have
+ripple effects across domains. For instance, evidence suggests that
+treating core depressive symptoms can improve social functioning,
+employment engagement, and housing stability---particularly when
+combined with social or financial supports (Lund et al., 2018; McDaid et
+al., 2019). These dual benefits underscore the value of integrated,
+cross-domain strategies that simultaneously address internal symptoms
+and external stressors.
+
+Our results also highlight why single-point interventions may fail to
+produce lasting change. The broader causal structure reveals multiple
+interlocking feedback loops spanning both psychological symptoms and
+structural risk factors, which may re-amplify vulnerability even after
+temporary relief. Supporting this, recent simulation work by Park, Li,
+et al. (2025) shows that feedback loops, even when modest in strength,
+can substantially diminish the long-term impact of isolated
+interventions unless multiple reinforcing links are simultaneously
+disrupted. Our own simulations support this view: while all models
+converged to similar long-term outcomes under sustained support,
+differences in internal structure---specifically, the strength of the
+link from precariousness to depression---produced systematic variation
+in how quickly and through which pathways systems responded to
+intervention. Stronger coupling slowed recovery and prolonged
+depression, showing that internal architecture can be as important as
+external input. This supports growing calls for systems-oriented mental
+health perspectives (Borsboom, 2017). Rather than assuming uniform
+responses, more durable outcomes may require (1) targeting early-acting
+symptoms, (2) sustaining support where recovery is slow, and (3)
+combining structural and psychological interventions. Tailoring
+strategies to match the system's architecture can improve effectiveness,
+particularly in populations with layered vulnerabilities. In this light,
+understanding causal architecture is not merely theoretical---it is
+foundational to designing more effective and equitable interventions.
 

### 2025-08-14 draft/draft_v2.qmd
added corresponding author

@@
 author: 
-  - name: Kyuri Park
+  - name: Kyuri Park \thanks{To whom correspondence should be addressed. \url{k.park@uva.nl}}
     affil-id: 1
+    corresponding: true
+    corresponding-note: "Corresponding author: kyuri.park@uva.nl"
   - name: Leonie K. Elsenburg
@@
     fontfamily: palatino
-    toc: true
+    #toc: true
     number-sections: true
@@
 This approach offers a scalable framework for diagnosing hidden system constraints and informing targeted public health strategies in complex psychosocial domains.
-
-  <!-- Understanding the causal mechanisms linking precariousness and depression is critical for developing effective interventions.
-This study utilizes data from the HELIUS cohort study to explore these relationships using advanced causal discovery methods.
-By applying algorithms such as FCI, and CCI, and combining traditional Gaussian CI tests with non-parametric approaches like RCoT, we investigate how different indicators of precariousness—including factors related to employment, social relations and relational stress, financial situation, and housing—affect depressed mood, both as a sum score and at the individual symptom level.
-Our findings reveal that relational stress consistently emerges as a potential causal factor for depressed mood, while symptoms such as sleep disturbances, guilt, and anhedonia are particularly sensitive to external stressors, acting as potential early warning signals or intervention points for prevention.
-Moreover, the results highlight complexities in the data, including the influence of latent confounders and the challenges of capturing cyclic relationships.
-Despite some limitations, such as unresolved ambiguities in causal directions and challenges with mixed data distributions, this study demonstrates the utility of causal discovery tools in disentangling the intricate interplay between social and mental health dynamics.
-By mapping these causal structures into computational models, future research can simulate intervention effects, providing actionable insights to mitigate the impact of precariousness on mental health.
-This study serves as a foundational effort, offering both methodological advancements and practical implications for addressing depression at a population level.
--->
 ---

### 2025-08-14 draft/draft_v2.tex
added corresponding author

@@
 \renewcommand\Affilfont{\small}
+
 \usepackage{amssymb}
@@
 \usepackage{geometry}
+\usepackage{hyperref}
 
@@
   pdftitle={From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study},
-  pdfauthor={Kyuri Park; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
+  pdfauthor={Kyuri Park ; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
   colorlinks=true,
@@
 
-\author[1]{Kyuri Park}
-\author[2]{Leonie K. Elsenburg}
-\author[2]{Mary Nicolao}
-\author[2]{Karien Stronks}
-\author[1, 3]{Vítor V. Vasconcelos}
+\author[1]{%
+  Kyuri Park
+\thanks{To whom correspondence should be addressed. \url{k.park@uva.nl}}%
+  %
+}
+\author[2]{%
+  Leonie K. Elsenburg%
+  %
+}
+\author[2]{%
+  Mary Nicolao%
+  %
+}
+\author[2]{%
+  Karien Stronks%
+  %
+}
+\author[1, 3]{%
+  Vítor V. Vasconcelos%
+  %
+}
 
@@
 
-\date{2025-08-12}
+\date{2025-08-14}
+
 \begin{document}
@@
 
-\renewcommand*\contentsname{Table of contents}
-{
-\hypersetup{linkcolor=}
-\setcounter{tocdepth}{3}
-\tableofcontents
-}
-
 \section{Introduction}\label{introduction}

### 2025-08-25 draft/draft_v3.qmd
add draft_v3 (nature format)

--- 2025-08-14 draft/draft_v2.qmd
@@
     affil-id: 2
-  - name: Mary Nicolao
+  - name: Mary Nicolaou
     affil-id: 2
@@
   @*
-csl: apa.csl
+csl: nature.csl
 abstract: |
@@
 Specifically, we ask: *How does the internal configuration of the precariousness–depression relationship affect the outcomes of stress-reducing interventions?*
-To answer this, we combine two approaches.
-First, we use cycle-capable causal discovery algorithms to infer directional relationships among domain-specific precariousness factors and depressive symptoms.
-By examining both aggregated measures (depression sum score and a composite precariousness index) and disaggregated variables (individual depressive symptoms and specific life-domain indicators), we capture broad patterns of relationship as well as fine-grained causal pathways.
+To address this question, we analyze cross-sectional data from more than 21,000 participants in the HELIUS (HEalthy LIfe in an Urban Setting) study [@stronks2013unravelling; @snijder2017cohort] using a two-step approach.
+First, we apply cycle-capable causal discovery algorithms to infer directional relationships among financial stress, domain-specific precariousness, and depressive symptoms.
+By examining both aggregated measures (a depression sum score and a composite precariousness index) and disaggregated indicators (individual symptoms and life-domain factors), we capture both broad patterns and fine-grained pathways.
 Second, we translate these insights into a computational dynamical model that simulates how depression and precariousness co-evolve under varying levels of external stress.
-This model enables us to probe how system properties—such as feedback strength and stochastic noise—govern responsiveness to intervention. 
-When the effects of depression symptoms onto the social precarity subsystem are weaker than the reverse path, there is higher sensitivity of depression to intervention, reflecting both a faster reduction post financial intervention but also quicker relapse once financial stresses are reestablished.
-By integrating causal discovery with dynamic simulation, we demonstrate that intervention strategies can be tailored to the underlying architecture of vulnerability.
-
-<!-- Mental health problems are rising globally, and populations facing chronic stress and socioeconomic instability bear a disproportionate share of this burden.
-Despite growing recognition of these challenges, designing effective policies remains difficult—largely because mental health outcomes emerge from complex, interacting factors that do not operate in isolation.
-To inform better interventions, we must move beyond identifying which factors are associated with mental health problems and instead uncover how those factors influence one another, the underlying causal mechanisms.
+This framework enables us to examine how system properties—such as feedback strength and stochastic noise—shape responsiveness to intervention.
+
+@fig-concept presents this high-level model, derived from the causal discovery analysis and serving as the foundation for our simulation experiments.
+It highlights financial stress ($S_i$) as an upstream driver influencing both depressive symptoms ($D_i$) and social precariousness ($P_i$).
+Depression contributes to broader social vulnerability by feeding into precariousness, while precariousness feeds back into depression, creating reinforcing cycles.
+This framework illustrates that intervention effects depend not only on external inputs but also on the internal feedback architecture of the system.
+Full methodological details, including data processing, algorithmic specifications, and model calibration procedures, are provided in the [Methods](#sec-methods) section.
+
+<!-- When the effects of depression symptoms onto the social precarity subsystem are weaker than the reverse path, there is higher sensitivity of depression to intervention, reflecting both a faster reduction post financial intervention but also quicker relapse once financial stresses are reestablished.
 -->
-
-<!-- Precarious life conditions—such as unstable housing, insecure employment, and financial hardship—are now well established as key correlates of depression and other mental health outcomes.
-However, the causal dynamics linking these domains remain poorly understood.
-In particular, it is unclear how different forms of precariousness and depressive symptoms influence one another, and whether their interplay creates self-reinforcing loops that stabilize poor mental health and social adversity over time.
--->
-
-<!-- This study aims to investigate the causal mechanisms linking precariousness and depression, and to examine how those mechanisms shape the system’s response to external intervention.
-Specifically, we ask: How do different configurations of the precariousness–depression relationship affect the outcomes of interventions aimed at reducing external stress?
--->
-
-<!-- To address this, we combine causal discovery and mechanistic modeling.
-We apply cycle-capable causal discovery algorithms to infer directional relationships among financial stress, precariousness indicators, and depressive symptoms.
-Our analysis includes both aggregate scores (e.g., PHQ-9 total, precariousness composite index) and disaggregated variables (individual symptoms and domain-specific precariousness), allowing us to capture both broad associations and fine-grained pathways.
--->
-
-<!-- Building on the discovered causal patterns, we simulate a set of scenarios using a computational model that captures the co-evolution of depression and precariousness under varying levels of financial stress.
-This framework allows us to examine how internal system properties—particularly feedback strength and stochastic noise—influence responsiveness to intervention.
-In some regimes, the system exhibits bistability, resisting change unless pushed past a critical threshold.
-In others, it transitions gradually, reflecting unistable dynamics.
-Through this integrative approach, we aim to show how the internal architecture of the depression–precariousness system fundamentally shapes its susceptibility to change—highlighting the importance of structural understanding in designing effective, equitable mental health interventions.
--->
-
-# Methods
-
-## Data
-
-We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study [@stronks2013unravelling; @snijder2017cohort].
-It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement].
-To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
-To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures.
-These analyses led to the identification of five distinct precariousness factors: three reflecting structural domains—employment, housing, and social relationships—and two capturing recent life stressors—financial and relational.
-Each factor is composed of multiple observed variables, as outlined below.
-Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).
-
--   Employment precariousness: `emp_stat` (current employment status), `work_sit` (nature of work situation).
--   Social precariousness: `soc_freq` (frequency of social contact), `soc_adq` (perceived adequacy of social support).
--   Housing precariousness: `nb_safe` (neighborhood safety), `nb_res` (access to neighborhood resources), `nb_rent` ( proportion of rental housing), `cul_rec` (availability of cultural resources).
--   Recent relational stressors: `frd_brk12` (experience of a friendship breakup), `conf12` (recent interpersonal or relational conflict).
--   Recent financial stressors: `fincri12` (experience of a financial crisis), `inc_diff` ( difficulty managing household income).
-
-Following data cleaning and normalization procedures, the analytical sample comprises 21,628 individuals.
-In addition to the five identified precariousness factors, we computed an overall precariousness score by summing these five factor scores, thereby capturing a composite measure of accumulated precariousness across domains.
-Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores.
-In the subsequent causal discovery analysis, we explore the relationship between depression and precariousness using both aggregated scores and their disaggregated components.
-See @fig-dist for the overall distributions of the variables used in the analysis.
-
-```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%", cache=TRUE}
-#| fig-cap: >
-#|   Distributions of variables with density overlay.  
-#|   *This shows the distributions of the variables used in the causal analysis, plotted as histograms with overlaid density estimates.
-The x-axis represents the value of each variable after preprocessing.
-The y-axis shows the corresponding estimated probability density.*
-#|   ***P.emp*** = *employment precariousness*;  ***P.hou*** = *housing precariousness*;  ***P.soc*** = *social precariousness*;  ***S.fin*** = *recent financial stressors*;  ***S.rel*** = *recent relational stressors*;  ***precariousness*** = *overall precariousness*;  ***PHQsum*** = *PHQ-9 sum score*;  ***anh*** = *anhedonia*;  ***app*** = *appetite*;  ***con*** = *concentration*;  ***dep*** = *depressed mood*;  ***ene*** = *energy*;  ***glt*** = *guilty*;  ***mot*** = *motor*;  ***slp*** = *sleep*;  ***sui*** = *suicidal*.
-#|   
-#| label: fig-dist
-
-clust_data <- scaled_data |> 
-  mutate(
-    P.emp = rowMeans(cbind(emp_stat, work_sit)),
-    P.soc = rowMeans(cbind(soc_freq, soc_adq)),
-    P.hou = rowMeans(cbind(nb_safe, nb_res, nb_rent, cul_rec)),
-    S.rel = rowMeans(cbind(frd_brk12, conf12)),
-    S.fin = rowMeans(cbind(fincri12, inc_dif)),
-    precariousness = P.emp + P.soc + P.hou + S.rel + S.fin
-  ) |>
-  dplyr::select(17:34) |>
-  as.data.frame()
-
-# Reshape the data to long format
-clust_data_long <- clust_data |>
-  dplyr::select(-age, -ethn) |>
-  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")
-
-# Plot distributions with density lines
-ggplot(clust_data_long, aes(x = value)) +
-  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "white", bins = 15) +  # Use density for histogram
-  geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
-  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "precariousness", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui", "PHQsum")), scales = "free") +
-  labs(x = "", y = "") +
-  theme_minimal()
-
-```
-
-## Causal Discovery{#sec-analysis}
-
-To uncover directional relationships between precariousness and depression, we use constraint-based causal discovery algorithms that allow for feedback loops and latent confounding.
-Specifically, we apply the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019].
-For reference, we also include the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].
-While our full analysis includes results from three causal discovery algorithms—FCI, CCI, and PC—we focus on FCI in the main text, as it offers a strong balance between flexibility and theoretical soundness.
-FCI accommodates latent confounding and potential feedback, making it well-suited for the complexity of our data [@mooij2020constraint].
-The PC algorithm is included for reference but relies on more restrictive assumptions, such as acyclicity and no unmeasured confounding.
-CCI, while capable of modeling cycles explicitly, has its own theoretical limitations; for example, its output may not fully preserve the d-separation properties of the true causal graph [@strobl2019].
-Results from both PC and CCI are provided in the [Appendix](#sec-appendix) ([@sec-cci; @sec-pc]) for completeness and comparison.
-For readers interested in detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, please refer to the [Appendix](#sec-appendix) ([@sec-causalprimer]).
-
-A key challenge in applying these algorithms to the HELIUS dataset is that many variables are non-Gaussian and likely exhibit nonlinear relationships.
-While kernel-based conditional independence tests (such as KCIT) are well-suited for capturing such complex dependencies, they are computationally intensive—scaling quadratically with sample size due to the need to invert large kernel matrices—making them impractical for large datasets like HELIUS [@rahimi2007random].
-To address this, we use the Randomized Conditional Correlation Test (RCoT), a nonparametric alternative that approximates kernel methods using random Fourier features.
-This reduces computational complexity from quadratic to linear in sample size, substantially lowering runtime while preserving sensitivity to nonlinear relationships [@strobl2019approximate; @zhang2012kernel].
-For further details, see @sec-rcot.
-
-We examine the causal structure using three complementary approaches:
-
-1. Aggregate-level analysis: relationships between five domain-specific precariousness factors and the PHQ-9 sum score.
-2. Fully disaggregated analysis: relationships between individual precariousness items and individual depressive symptoms.
-3. Mixed analysis: relationships between individual symptoms and a composite index of overall precariousness.
-
-The aggregate analysis reduces dimensionality and yields interpretable summaries of how domains of precariousness relate to overall depression severity.
-However, it may obscure finer-grained effects.
-The disaggregated analysis allows for precise mapping of which symptoms are influenced by (or influence) specific precariousness conditions, but it introduces complexity due to the high dimensionality and distributional properties of the data.
-The mixed approach, in turn, integrates these perspectives by examining how individual symptoms respond to cumulative precariousness across domains.
-
-To robustly estimate the causal structure between depression and precariousness variables, we implement a comprehensive bootstrapped causal discovery procedure.
-This procedure systematically varies key analysis parameters across a grid of configurations, enabling us to assess the stability of inferred relationships under different settings.
-Specifically, we vary four main components:
-
-* Significance levels for conditional independence tests: $\alpha$ = 0.01 and 0.05.
-* Stability thresholds: 0.5, 0.6, 0.7, and 0.8, which determine how frequently an edge must appear across bootstraps to be retained.
-* Conditional independence (CI) tests: Gaussian CI test (partial correlation) and RCoT (a nonparametric test).
-* Causal discovery algorithms: FCI, CCI, and PC.
-
-For each unique parameter combination (2 significance levels × 4 thresholds × 2 CI tests = 16 settings), we run 100 bootstrap samples per algorithm for the aggregated analyses (see Step 1 in @fig-workflow).
-For symptom-level analyses, which are more computationally intensive, we use 30 bootstrap samples per setting.[^5]
-
-[^5]: For symptom-level analyses, we further improve computational efficiency by fixing the skeleton structure—the undirected network of potential connections—using a consensus graph derived from all three algorithms (with $\alpha = 0.01$ and the RCoT test).
-This allows subsequent analyses to focus exclusively on estimating edge directions, avoiding the repeated and costly computation of skeletons, and thus substantially reducing runtime.
-
-To summarize these results, we apply a two-level aggregation strategy:
-
-1.
-**Within each condition** (fixed $\alpha$, threshold, CI test, and algorithm), we identify the most frequently occurring edge type (e.g., arrowhead, circle, none) across bootstraps for each pair of variables (Step 2 in @fig-workflow).
-2.
-**Across conditions**, we then determine the most dominant edge type based on its frequency.
-In cases where no edge type is clearly dominant (i.e., a tie), we mark the edge as uncertain, represented with a dashed line in the final graph (Step 3 in @fig-workflow).
-
-This summarization process is repeated separately for each causal discovery algorithm (FCI, CCI, PC), yielding a single robust graph per algorithm.
-These final graphs represent only the most stable and consistent relationships across a wide range of plausible parameterizations, enhancing both reliability and interpretability.
-To support transparency, each final graph is accompanied by a matrix summarizing the relative frequency of each edge type for every variable pair, enabling readers to assess the degree of uncertainty or agreement underlying each connection.
-
-<!-- To ensure robust estimation and guard against sensitivity to parameter settings, we implemented a systematic grid of analysis conditions: -->
-
-<!-- * Significance levels: $\alpha$ = 0.01 and 0.05 -->
-<!-- * Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8 -->
-<!-- * CI tests: Gaussian and RCoT -->
-<!-- * Algorithms: FCI, CCI, and PC -->
-
-<!-- Each combination was bootstrapped 100 times for sum-score level analyses and 30 times for symptom-level analyses, yielding a total of 1,600 graphs per algorithm for the full analysis.
-For symptom networks, we fixed the skeleton structure using a consensus graph estimated across PC, FCI, and CCI ($\alpha$ = 0.01, RCoT).
-This constraint improves computational efficiency without limiting the estimation of directional edges between symptoms and precariousness variables.
--->
-
-<!-- To summarize the results, we adopted a two-level aggregation strategy.
-First, within each analysis condition, we recorded the most frequently observed edge-endpoint type (e.g., directed, undirected, uncertain) from the bootstraps.
-Then, across all conditions, we tallied the most frequent endpoint type per edge.
-In cases of ties, we assigned dashed edges to reflect uncertainty.
-For transparency, we accompany each graph with a table showing the relative frequencies of each edge symbol.
--->
-
-<!-- This procedure yields a final graph for each algorithm and analysis type—offering both robustness and interpretability. See @fig-workflow for an overview of the causal discovery pipeline. -->
-<!-- For detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, see the [Appendix](#sec-appendix).
--->
-
-<!-- To uncover directional relationships between precariousness and depression, we used constraint-based causal discovery algorithms capable of detecting feedback loops and latent confounding.
-Specifically, we applied the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which support cyclic and non-acyclic structures [@mooij2020constraint; @strobl2019].
-Additionally, we include the *PC* algorithm as a reference, given its simplicity and prominence as one of the most widely known causal discovery methods [@spirtes2001causation].
--->
-
-<!-- One practical challenge in applying these algorithms to the HELIUS dataset is that most variables do not follow a Gaussian distribution, and the relationships among variables are unlikely to be strictly linear.
-To address this, we supplement the commonly used Gaussian conditional independence (CI) test, which relies on partial correlations, with a non-parametric CI test based on reproducing kernel methods [@zhang2012kernel].
--->
-<!-- While kernel-based conditional independence tests (KCIT) are theoretically effective in capturing non-linear dependencies and subtle interactions in data with non-Gaussian noise, they are computationally demanding for large datasets like HELIUS.
-Their quadratic scaling with sample size is primarily due to the need to invert large kernel matrices, making them impractical for large-scale analysis [@rahimi2007random].
--->
-<!-- To mitigate this issue, we use the Randomized Conditional Correlation Test (RCoT), which approximates kernel methods using random Fourier features.
-This approach reduces computational complexity from quadratic to linear scaling with sample size, significantly lowering computational costs while maintaining robustness [@strobl2019approximate].
-For a more detailed explanation of RCoT, refer to @sec-rcot.
--->
-
-<!-- We analyze the causal structure using three complementary approaches: (1) Examining the relationship between the five individual precariousness factors and the PHQ sum score, which represents overall depression severity.
--->
-<!-- (2) Investigating the relationship between individual depressive symptoms and individual precariousness factors.
-(3) Assessing the relationship between individual depressive symptoms and overall precariousness (the sum of all precariousness factors).
--->
-<!-- Using the PHQ sum score as a measure of depression simplifies the analysis by reducing dimensionality, improving computational efficiency, and providing a broad, interpretable perspective on how precariousness factors relate to overall depression severity.
-However, this aggregated approach may overlook granular symptom-level relationships.
--->
-
-<!-- By zooming in on individual symptoms, we capture a more detailed picture of how specific depressive symptoms interact with different precariousness factors.
-This approach allows us to map individual causal pathways, identifying which symptoms are particularly sensitive to certain precarious conditions and vice versa.
-However, this level of detail introduces methodological challenges, including non-standard distributions of symptom variables and the high-dimensional nature of the analysis, which can reduce statistical power and significantly increase computational demands—especially when using non-parametric conditional independence (CI) tests.
--->
-
-<!-- Finally, analyzing individual symptoms in relation to overall precariousness provides an integrative perspective, revealing how depression symptoms collectively relate to precarious conditions.
-This approach complements the fully disaggregated analysis by aggregating precariousness factors, potentially uncovering relationships that may be too weak to detect when examined separately.
-This is particularly relevant when a symptom is associated with multiple precariousness factors, but these associations are not strong enough to emerge in an individual-level analysis.
--->
-<!-- By integrating these three approaches — ranging from partially aggregated to fully disaggregated — we aim to achieve a balance between capturing detailed granularity and gaining a broader understanding of the relationship between precariousness and depression.
--->
-
-<!-- Furthermore, to address sensitivity to parameter values and enhance the robustness of our results, we implement multiple settings combined with bootstrapping.
-For each condition, we generate 100 bootstrap samples, estimate causal graphs, and retain only the edge-endpoints that exceed predefined thresholds.
-If no edge-endpoint type surpasses the threshold, we assign a circle ($\circ$) to indicate uncertainty in the edge direction.
--->
-
-<!-- The analyses are conducted under the following conditions: -->
-
-<!-- -   Significance levels ($\alpha$): 0.01 and 0.05 -->
-<!-- -   Thresholds: 0.5, 0.6, 0.7, and 0.8 -->
-<!-- -   CI test: Gaussian CI test, RCoT -->
-<!-- -   Algorithms: FCI, CCI, and PC -->
-
-<!-- This setup yields 16 combinations (2 significance levels × 4 thresholds × 2 CI tests), applied across three algorithms.
-Each combination is repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per algorithm.
--->
-<!-- For analyses involving individual symptom variables, we simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing the number of bootstrap samples to 30, easing computational demands.
--->
-<!-- To further enhance efficiency, we fix the skeleton of the symptom network using a common structure derived from PC, FCI, and CCI algorithms ($\alpha$ = 0.01, RCoT CI test).
-This step constrains which edges can be present among symptoms but does not restrict the estimation of causal directions within the symptom network.
-Additionally, the entire causal structure between symptoms and precariousness factors remains fully estimable, ensuring that key relationships are still inferred.
--->
-<!-- Beyond its computational benefits — particularly in reducing the time required for skeleton estimation — this fixed structure also aligns well with commonly reported symptom network structures in the literature [@park2025individual], ensuring consistency with existing findings.
--->
-
-<!-- To obtain a final, stable summary of causal relationships, we identify the most frequently occurring edge-endpoints across different experimental setups.
-First, within each condition, we determine the most frequent edge-endpoint from all bootstrap samples.
-Then, across different conditions, we aggregate these selected symbols and determine the overall most common edge-endpoint for each graph entry.
--->
-
-<!-- Since we have an even number of experimental setups, ties may occasionally occur.
-In such cases, we explicitly denote them in the resulting graph using dashed lines.
-To further help with the interpretation and understanding of the resulting graph, we provide a corresponding table for each graph, showing the proportion of each edge-endpoint type to clarify how dominant a given symbol is in each case.
--->
-
-<!-- By double summarizing the results in this way, we generate one final graph per algorithm for each analysis approach: the partially aggregated analysis, which uses either the sum of precariousness factors or the sum of depression factors to capture overall trends, and the fully disaggregated analysis, which examines individual precariousness and depression factors separately.
--->
-<!-- This approach ensures that only stable and consistent edges — those that persist across different parameter settings — are retained, leading to a robust and reliable understanding of the relationships between precariousness factors and depression.
-See @fig-workflow for an overview of the analysis workflow.
--->
-
-```{r}
-#| label: fig-workflow
-#| fig-cap: "Causal discovery workflow applied across all three algorithms (FCI, CCI, and PC).
-*The process involves three main steps: (1) bootstrapping the dataset to create multiple resampled datasets; (2) estimating causal graphs under varying parameter configurations; and (3) summarizing results by identifying the most frequent edge type across bootstraps and settings to construct a final stable graph.*"
-#| fig-align: center
-#| out-width: 90%
-
-knitr::include_graphics("img/simsetup.pdf")
-
-```
-
-## Modeling Dynamics between precariousness and Depression
-
-To explore how internal system structure shapes responses to external support, we implement a high-level computational model that formalizes key mechanisms suggested by the causal discovery analysis.
-Rather than attempting to replicate the full estimated causal graph, this model abstracts the essential structure into a compact, tractable system focused on dynamic interactions between depression and social precariousness, both influenced by financial stress.
-
-As illustrated in [@fig-concept], the model includes three components: social precariousness $P_i(t)$, depressive symptoms $D_i(t)$, and a static individual-specific external stressor $S_i$, representing financial stress.
-Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S_i$ acting as a shared input.
-These dynamics are formalized as a pair of linear stochastic differential equations (SDEs):
-
-$$
-\begin{aligned}
-dP_i(t) &= \lambda_P \big( \alpha_{PS} S_i + \alpha_{PD} D_i(t) - P_i(t) \big) \, dt + \sigma_P \, dW_{P,i}(t) \\
-dD_i(t) &= \lambda_D \big( \alpha_{DS} S_i + \alpha_{DP} P_i(t) - D_i(t) \big) \, dt + \sigma_D \, dW_{D,i}(t)
-\end{aligned}
-$$
-
-Here, $D_i(t)$ denotes the depressive state and $P_i(t)$ denotes the level of social precariousness for individual $i$ at time $t$, with $S_i$ representing a static, individual-specific external stressor (financial stress).
-The parameters $\lambda_D$, $\lambda_P$, as well as the coefficients $\alpha$ and noise terms $\sigma$, are assumed to be constant across individuals and time.
-The parameters $\lambda_D$ and $\lambda_P$ govern the timescales over which $D_i(t)$ and $P_i(t)$ adjust toward their respective input-driven target states.
-The coefficients $\alpha_{DS}$ and $\alpha_{PS}$ quantify the direct effects of the external stressor $S_i$ on depression and precariousness, respectively, while $\alpha_{DP}$ and $\alpha_{PD}$ capture the mutual influence between the two dynamic processes.
-Stochastic fluctuations are introduced via the independent Wiener processes $dW_{D,i}(t)$ and $dW_{P,i}(t)$, scaled by their respective volatility parameters $\sigma_D$ and $\sigma_P$.
-This linear structure offers a transparent and analytically tractable framework for modeling how exogenous variation in financial stress may propagate through individual dynamics of depression and precariousness.
-Crucially, all parameters can be calibrated directly from empirical covariances and variances observed in the HELIUS cohort.
-The resulting steady-state distribution of the model aligns with empirical measures: PHQ-9 total score (PHQsum) for depression, and the derived social precariousness factor (P.soc) for social precarity.
-The full calibration procedure is described in [@sec-calibration].
-
-<!-- Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress.
-The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which each variable adjusts toward its input-driven target state.
-The coefficients $\alpha_{PS}$ and $\alpha_{DS}$ quantify the direct effects of the stressor on $P$ and $D$, while $\alpha_{DP}$ and $\alpha_{PD}$ represent the cross-influences between depression and precariousness.
-Stochastic fluctuations are introduced through the noise terms $\sigma_P, dW_P$ and $\sigma_D, dW_D$, with $dW_P$ and $dW_D$ representing independent Wiener processes.
--->
-<!-- This linear structure offers a transparent and analytically tractable framework for examining how variations in external financial stress influence the population-level dynamics of depression and social precariousness.
-Crucially, the model's simplicity enables direct calibration from empirical data: all parameters can be derived analytically from observed variances and covariances in the HELIUS cohort.
-The following  [@sec-calibration] details the procedure used to perform this calibration.
--->
+<!-- By integrating causal discovery with dynamic simulation, we demonstrate that intervention strategies can be tailored to the underlying architecture of vulnerability. -->
 
@@
 #| label: fig-concept
-#| fig-cap: "High-level schematic of the dynamic model structure used in simulation.
-Arrows represent influence coefficients $\\alpha$, matching the terms in the system of stochastic differential equations.
-$D_i(t)$ and $P_i(t)$ denote depressive symptoms and social precariousness for individual $i$, with $S_i$ representing static financial stress."
+#| fig-cap: "The model abstracts the main dependencies revealed by causal discovery into a compact system of dynamic interactions.
+Financial stress ($S_i$) is treated as an upstream driver that influences both depressive symptoms ($D_i(t)$) and social precariousness ($P_i(t)$).
+Depression and precariousness also influence each other through feedback loops, creating the potential for reinforcing cycles of vulnerability.
+Arrows represent influence coefficients ($\\alpha$), which correspond to terms in the system of stochastic differential equations used for simulation.
+This high-level structure served as the foundation for our simulation experiments.
+Full methodological details, refer to the [Methods](#sec-methods) section."
 #| fig-align: center
@@
 
-### Model Calibration {#sec-calibration}
-
-In our linear model, all parameters are derived analytically to reproduce the empirical second-order moments observed in the HELIUS dataset.
-The system treats depression ($D$) and social precariousness ($P$) as jointly evolving under the influence of a shared external stressor ($S$), forming a coupled system of Ornstein–Uhlenbeck processes.
-Under the assumption of stationarity, the steady-state covariances between these variables admit closed-form solutions for the system’s structural parameters.
-
-We fix the adjustment rates $\lambda_D = \lambda_P = 1$, which normalizes the time scale given the cross-sectional nature of the HELIUS data.
-To resolve the remaining degree of freedom, we treat the feedback from precariousness to depression ($\alpha_{DP}$) as a free parameter, and we solve algebraically for all other coefficients — specifically the direct effects of stress ($\alpha_{DS}, \alpha_{PS}$), the reverse feedback from depression to precariousness ($\alpha_{PD}$), and the noise terms ($\sigma_D, \sigma_P$).
-These solutions are expressed as closed-form functions of the empirical covariances: $\mathrm{Var}(D)$, $\mathrm{Var}(P)$, $\mathrm{Var}(S)$, $\mathrm{Cov}(D, P)$, $\mathrm{Cov}(D, S)$, and $\mathrm{Cov}(P, S)$.
-
-This structure creates a family of analytically calibrated models, each consistent with the same observed covariances but differing in their internal configuration.
-We constrain the space of valid solutions by imposing that all structural parameters must be positive, reflecting the assumption that stress and precariousness contribute adversely to depression.
-This constraint restricts the allowable range of $\alpha_{DP}$ to approximately $[0, 0.698]$ (see Appendix [@sec-analcalibration]).
-
-@fig-tradeoff illustrates how the remaining parameters adjust as $\alpha_{DP}$ increases to maintain consistency with the empirical covariances.
-As $\alpha_{DP}$ increases, $\alpha_{PD}$ tends to decrease, reflecting a compensatory trade-off between the mutual influences of depression and precariousness.
-This interplay helps preserve the model’s match with observed covariances, without requiring major shifts in external input strength.
-Meanwhile, the external coupling terms — $\alpha_{DS}$ and $\alpha_{PS}$ — also shift to absorb variance in a way that maintains alignment with the empirical data.
-Notably, the noise amplitudes $\sigma_D$ and $\sigma_P$ decline slightly as $\alpha_{DP}$ increases.
-This indicates that more of the observed variability is captured by deterministic interactions, with less left to be explained by stochastic fluctuations.
-
-Rather than being a limitation, this flexibility enables exploration of a spectrum of equally data-consistent feedback regimes.
-Each represents a different plausible causal architecture, offering insight into how internal dynamics modulate the system's responsiveness to external stress — even when the observed covariances remain unchanged.
-
-```{r}
-#| label: fig-tradeoff
-#| fig-cap: >
-#|   Trade-off curves showing how calibrated parameters vary with $\alpha_{DP}$.
-#|   *Each line shows how a different parameter adjusts in response to changes in $\alpha_{DP}$ to maintain consistency with empirical covariances from the HELIUS data. 
-#|   All configurations shown are equally consistent with the data, illustrating how internal dynamics adjust as feedback strength changes.*
-#| fig-align: center
-#| out-width: 100%
-knitr::include_graphics("img/param_tradeoff.pdf")
-```
-
-### Intervention Simulation
-
-To evaluate how internal structure affects system responsiveness, we simulate a family of linear models subjected to external intervention.
-Each model corresponds to a different value of $\alpha_{DP}$, representing the strength of influence from social precariousness to depression.
-The range spans from 0 (no coupling) to 0.65 (maximal admissible value), with all other parameters derived analytically to preserve the empirical covariances observed in the HELIUS data (see @fig-tradeoff).
-
-In each simulation, financial stress of individual $i$ ($S_i$) is proportionally reduced toward the lowest observed value, modeling a population-wide intervention applied at varying intensities (0–100%).
-For each model, we simulate 1000 virtual individuals whose baseline stress levels are sampled from the empirical distribution and scaled according to the intervention level.
-
-We assess the system’s behavior using two complementary approaches:
-
-- Snapshot analysis captures the average levels of depression ($D$) and precariousness ($P$) at selected time points ($t = 100$, $300$, $500$) following intervention onset.
-Since all linear models converge to the same equilibrium, this highlights how internal feedback affects the speed and pathway of adjustment.
-
-- Full trajectory simulation tracks system dynamics over time for two representative models: one with minimal ($\alpha_{DP} = 0$) and one with maximal ($\alpha_{DP} = 0.65$) feedback.
-Each simulation includes an intervention period followed by its withdrawal, revealing differences in reactivity, delay, and recovery stability.
-
-Together, these simulations demonstrate how internal structure governs not just whether change occurs, but how it unfolds — shaping the pace, persistence, and variability of response to external support.
-
 # Results
+
+We present the results in two parts.
+First, we report the causal structures inferred from the HELIUS data at the levels of aggregate depression, individual symptoms, and aggregated precariousness.
+Second, we embed these findings in a simplified dynamical model to simulate how alternative feedback structures shape the system’s response to stress-reducing interventions.
 
@@
 <!-- These consistent pathways underscore the robustness of the core relational architecture across algorithms amid variation in directional confidence. : " This is discussion point : Vitor-->
-
-<!-- In  graph  (a),  the  key  pathways  suggest  that  employment  precariousness  (P.emp) and social precariousness (P.soc) do not cause depression (PHQsum).P.soc does not cause recent relational stress (S.rel) or financial stress (S.fin), and these stressors are likely related by a latent confounder.
-Additionally, P.emp and S.fin are identified as non-causes of housing precariousness (P.hou).
-Both FCI and CCI detect a dependency between P.emp and S.fin, but they disagree on the direction of the relationship, leaving it unclear whether S.fin causes P.emp or whether a latent confounder mediates their relationship.
-Similar ambiguities are found in the connections between S.rel and PHQsum and between S.fin and PHQsum -->
-
-<!-- Both FCI and CCI detect a dependency between *P.emp* and *S.fin*, but they disagree on the direction of the relationship, leaving it unclear whether *S.fin* causes *P.emp* or whether a latent confounder mediates their relationship.
-Similar ambiguities are found in the connections between *S.rel* and *PHQsum* and between *S.fin* and *PHQsum*.
--->
-
-<!-- Graph (b), derived solely from the nonparametric RCoT method.
-Both graphs consistently identify that *P.emp* is not a cause of *P.hou* or *PHQsum*, and that *P.soc* does not cause *PHQsum* or *S.fin*.
--->
-<!-- The relationship between stressors and depression remains unresolved between the two algorithms.
-However, the graph derived from RCoT provides greater confidence that depression is not the cause of *S.fin* and suggests a stronger likelihood that the financial stress may contribute to depression or that their relationship is mediated by a latent confounder.
--->
-<!-- The role of relational stress has become less pronounced, as the edge between *P.soc* and *S.rel* is omitted, and the relationship between *S.rel* and *S.fin* is now inconsistent between the algorithms.
--->
-
-<!-- Overall, both graphs together highlight the potential roles of stressors like *S.fin* and *S.rel*, which are closely tied to depression, either causally or via latent confounders.
-Employment and social precariousness may also be influenced by depression, either directly or through latent variables.
-Lastly, housing precariousness does not directly impact depression but remains connected to employment precariousness and financial stress, likely mediated by direct or latent pathways.
--->
 
@@
 #|   System response to external intervention across variations in internal structure.
-#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across varying values of $\alpha_{DP}$ and intervention strengths.
-The x-axis represents the strength of external support, operationalized as the proportional reduction of financial stress ($S$) toward its minimum observed value.
+#|   ***Panel A**: Mean levels of depression ($D$) and precariousness ($P$) at time steps 100, 300, and 500, across models that differ in the coupling coefficient $\alpha_{DP}$, which represents the strength of depression’s influence on precariousness in the dynamical model.
+The x‑axis shows intervention strength, defined as the proportional reduction of financial stress ($S$) relative to its baseline (with higher values indicating more external financial support).
 Each line corresponds to a different model defined by a specific $\alpha_{DP}$ value, color-coded from low (purple) to high (yellow).
@@
 Ultimately, we hope this study encourages further integration of causal structure learning and systems modeling in efforts to understand and  improve mental health outcomes under conditions of socioeconomic precariousness.
+
+# Methods {#sec-methods}
+
+## Data
+
+We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study [@stronks2013unravelling; @snijder2017cohort].
+It  captures the diverse population of the city of Amsterdam by including the six main ethnic groups and provides comprehensive health and lifestyle data, including depressive symptoms as measured by the PHQ-9 [@galenkamp2017measurement].
+To operationalize indicators of precariousness, we draw on the framework outlined in previous research [@elsenburg2025clustering] and select a set of relevant variables.
+To ensure a robust representation of precariousness in our causal discovery models, we conducted exploratory analyses to identify consistent and meaningful data-driven factor structures.
+These analyses led to the identification of five distinct precariousness factors: three reflecting structural domains—employment, housing, and social relationships—and two capturing recent life stressors—financial and relational.
+Each factor is composed of multiple observed variables, as outlined below.
+Additional details on the HELIUS study and the factor extraction process are provided in the [Appendix](#sec-appendix).
+
+-   Employment precariousness: `emp_stat` (current employment status), `work_sit` (nature of work situation).
+-   Social precariousness: `soc_freq` (frequency of social contact), `soc_adq` (perceived adequacy of social support).
+-   Housing precariousness: `nb_safe` (neighborhood safety), `nb_res` (access to neighborhood resources), `nb_rent` ( proportion of rental housing), `cul_rec` (availability of cultural resources).
+-   Recent relational stressors: `frd_brk12` (experience of a friendship breakup), `conf12` (recent interpersonal or relational conflict).
+-   Recent financial stressors: `fincri12` (experience of a financial crisis), `inc_diff` ( difficulty managing household income).
+
+Following data cleaning and normalization procedures, the analytical sample comprises 21,628 individuals.
+In addition to the five identified precariousness factors, we computed an overall precariousness score by summing these five factor scores, thereby capturing a composite measure of accumulated precariousness across domains.
+Depression is represented using the PHQ-9 [@kroenke2001phq], both as a total sum score and as individual symptom scores.
+In the subsequent causal discovery analysis, we explore the relationship between depression and precariousness using both aggregated scores and their disaggregated components.
+See @fig-dist for the overall distributions of the variables used in the analysis.
+
+```{r echo=FALSE, fig.align='center', fig.height=5, fig.width=8, out.width="90%", cache=TRUE}
+#| fig-cap: >
+#|   Distributions of variables with density overlay.  
+#|   *This shows the distributions of the variables used in the causal analysis, plotted as histograms with overlaid density estimates.
+The x-axis represents the value of each variable after preprocessing.
+The y-axis shows the corresponding estimated probability density.*
+#|   ***P.emp*** = *employment precariousness*;  ***P.hou*** = *housing precariousness*;  ***P.soc*** = *social precariousness*;  ***S.fin*** = *recent financial stressors*;  ***S.rel*** = *recent relational stressors*;  ***precariousness*** = *overall precariousness*;  ***PHQsum*** = *PHQ-9 sum score*;  ***anh*** = *anhedonia*;  ***app*** = *appetite*;  ***con*** = *concentration*;  ***dep*** = *depressed mood*;  ***ene*** = *energy*;  ***glt*** = *guilty*;  ***mot*** = *motor*;  ***slp*** = *sleep*;  ***sui*** = *suicidal*.
+#|   
+#| label: fig-dist
+
+clust_data <- scaled_data |> 
+  mutate(
+    P.emp = rowMeans(cbind(emp_stat, work_sit)),
+    P.soc = rowMeans(cbind(soc_freq, soc_adq)),
+    P.hou = rowMeans(cbind(nb_safe, nb_res, nb_rent, cul_rec)),
+    S.rel = rowMeans(cbind(frd_brk12, conf12)),
+    S.fin = rowMeans(cbind(fincri12, inc_dif)),
+    precariousness = P.emp + P.soc + P.hou + S.rel + S.fin
+  ) |>
+  dplyr::select(17:34) |>
+  as.data.frame()
+
+# Reshape the data to long format
+clust_data_long <- clust_data |>
+  dplyr::select(-age, -ethn) |>
+  pivot_longer(cols = everything(), names_to = "variable", values_to = "value")
+
+# Plot distributions with density lines
+ggplot(clust_data_long, aes(x = value)) +
+  geom_histogram(aes(y = after_stat(density)), fill = "lightblue", color = "white", bins = 15) +  # Use density for histogram
+  geom_density(adjust=2, linewidth=0.3, color = "gray", alpha = 0.5) +  # Add density line in red
+  facet_wrap(~ factor(variable, levels =c("P.emp", "P.hou", "P.soc", "S.fin", "S.rel", "precariousness", "anh", "app", "con", "dep", "ene", "glt", "mot", "slp", "sui", "PHQsum")), scales = "free") +
+  labs(x = "", y = "") +
+  theme_minimal()
+
+```
+
+## Causal Discovery{#sec-analysis}
+
+To uncover directional relationships between precariousness and depression, we use constraint-based causal discovery algorithms that allow for feedback loops and latent confounding.
+Specifically, we apply the *Fast Causal Inference* (FCI) and *Cyclic Causal Inference* (CCI) algorithms, both of which can detect cycles and non-acyclic structures [@mooij2020constraint; @strobl2019].
+For reference, we also include the *PC* algorithm, one of the most widely used methods for acyclic causal discovery [@spirtes2001causation].
+While our full analysis includes results from three causal discovery algorithms—FCI, CCI, and PC—we focus on FCI in the main text, as it offers a strong balance between flexibility and theoretical soundness.
+FCI accommodates latent confounding and potential feedback, making it well-suited for the complexity of our data [@mooij2020constraint].
+The PC algorithm is included for reference but relies on more restrictive assumptions, such as acyclicity and no unmeasured confounding.
+CCI, while capable of modeling cycles explicitly, has its own theoretical limitations; for example, its output may not fully preserve the d-separation properties of the true causal graph [@strobl2019].
+Results from both PC and CCI are provided in the [Appendix](#sec-appendix) ([@sec-cci; @sec-pc]) for completeness and comparison.
+For readers interested in detailed explanations of the algorithms, edge interpretations, and graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions and limitations, please refer to the [Appendix](#sec-appendix) ([@sec-causalprimer]).
+
+A key challenge in applying these algorithms to the HELIUS dataset is that many variables are non-Gaussian and likely exhibit nonlinear relationships.
+While kernel-based conditional independence tests (such as KCIT) are well-suited for capturing such complex dependencies, they are computationally intensive—scaling quadratically with sample size due to the need to invert large kernel matrices—making them impractical for large datasets like HELIUS [@rahimi2007random].
+To address this, we use the Randomized Conditional Correlation Test (RCoT), a nonparametric alternative that approximates kernel methods using random Fourier features.
+This reduces computational complexity from quadratic to linear in sample size, substantially lowering runtime while preserving sensitivity to nonlinear relationships [@strobl2019approximate; @zhang2012kernel].
+For further details, see @sec-rcot.
+
+We examine the causal structure using three complementary approaches:
+
+1. Aggregate-level analysis: relationships between five domain-specific precariousness factors and the PHQ-9 sum score.
+2. Fully disaggregated analysis: relationships between individual precariousness items and individual depressive symptoms.
+3. Mixed analysis: relationships between individual symptoms and a composite index of overall precariousness.
+
+The aggregate analysis reduces dimensionality and yields interpretable summaries of how domains of precariousness relate to overall depression severity.
+However, it may obscure finer-grained effects.
+The disaggregated analysis allows for precise mapping of which symptoms are influenced by (or influence) specific precariousness conditions, but it introduces complexity due to the high dimensionality and distributional properties of the data.
+The mixed approach, in turn, integrates these perspectives by examining how individual symptoms respond to cumulative precariousness across domains.
+
+To robustly estimate the causal structure between depression and precariousness variables, we implement a comprehensive bootstrapped causal discovery procedure.
+This procedure systematically varies key analysis parameters across a grid of configurations, enabling us to assess the stability of inferred relationships under different settings.
+Specifically, we vary four main components:
+
+* Significance levels for conditional independence tests: $\alpha$ = 0.01 and 0.05.
+* Stability thresholds: 0.5, 0.6, 0.7, and 0.8, which determine how frequently an edge must appear across bootstraps to be retained.
+* Conditional independence (CI) tests: Gaussian CI test (partial correlation) and RCoT (a nonparametric test).
+* Causal discovery algorithms: FCI, CCI, and PC.
+
+For each unique parameter combination (2 significance levels × 4 thresholds × 2 CI tests = 16 settings), we run 100 bootstrap samples per algorithm for the aggregated analyses (see Step 1 in @fig-workflow).
+For symptom-level analyses, which are more computationally intensive, we use 30 bootstrap samples per setting.[^5]
+
+[^5]: For symptom-level analyses, we further improve computational efficiency by fixing the skeleton structure—the undirected network of potential connections—using a consensus graph derived from all three algorithms (with $\alpha = 0.01$ and the RCoT test).
+This allows subsequent analyses to focus exclusively on estimating edge directions, avoiding the repeated and costly computation of skeletons, and thus substantially reducing runtime.
+
+To summarize these results, we apply a two-level aggregation strategy:
+
+1.
+**Within each condition** (fixed $\alpha$, threshold, CI test, and algorithm), we identify the most frequently occurring edge type (e.g., arrowhead, circle, none) across bootstraps for each pair of variables (Step 2 in @fig-workflow).
+2.
+**Across conditions**, we then determine the most dominant edge type based on its frequency.
+In cases where no edge type is clearly dominant (i.e., a tie), we mark the edge as uncertain, represented with a dashed line in the final graph (Step 3 in @fig-workflow).
+
+This summarization process is repeated separately for each causal discovery algorithm (FCI, CCI, PC), yielding a single robust graph per algorithm.
+These final graphs represent only the most stable and consistent relationships across a wide range of plausible parameterizations, enhancing both reliability and interpretability.
+To support transparency, each final graph is accompanied by a matrix summarizing the relative frequency of each edge type for every variable pair, enabling readers to assess the degree of uncertainty or agreement underlying each connection.
+
+```{r}
+#| label: fig-workflow
+#| fig-cap: "Causal discovery workflow applied across all three algorithms (FCI, CCI, and PC).
+*The process involves three main steps: (1) bootstrapping the dataset to create multiple resampled datasets; (2) estimating causal graphs under varying parameter configurations; and (3) summarizing results by identifying the most frequent edge type across bootstraps and settings to construct a final stable graph.*"
+#| fig-align: center
+#| out-width: 90%
+
+knitr::include_graphics("img/simsetup.pdf")
+
+```
+
+## Modeling Dynamics between precariousness and Depression
+
+To explore how internal system structure shapes responses to external support, we implement a high-level computational model that formalizes key mechanisms suggested by the causal discovery analysis.
+Rather than attempting to replicate the full estimated causal graph, this model abstracts the essential structure into a compact, tractable system focused on dynamic interactions between depression and social precariousness, both influenced by financial stress.
+
+As illustrated in [@fig-concept], the model includes three components: social precariousness $P_i(t)$, depressive symptoms $D_i(t)$, and a static individual-specific external stressor $S_i$, representing financial stress.
+Depression and precariousness are modeled as coupled state variables that influence each other over time, with $S_i$ acting as a shared input.
+These dynamics are formalized as a pair of linear stochastic differential equations (SDEs):
+
+$$
+\begin{aligned}
+dP_i(t) &= \lambda_P \big( \alpha_{PS} S_i + \alpha_{PD} D_i(t) - P_i(t) \big) \, dt + \sigma_P \, dW_{P,i}(t) \\
+dD_i(t) &= \lambda_D \big( \alpha_{DS} S_i + \alpha_{DP} P_i(t) - D_i(t) \big) \, dt + \sigma_D \, dW_{D,i}(t)
+\end{aligned}
+$$
+
+Here, $D_i(t)$ denotes the depressive state and $P_i(t)$ denotes the level of social precariousness for individual $i$ at time $t$, with $S_i$ representing a static, individual-specific external stressor (financial stress).
+The parameters $\lambda_D$, $\lambda_P$, as well as the coefficients $\alpha$ and noise terms $\sigma$, are assumed to be constant across individuals and time.
+The parameters $\lambda_D$ and $\lambda_P$ govern the timescales over which $D_i(t)$ and $P_i(t)$ adjust toward their respective input-driven target states.
+The coefficients $\alpha_{DS}$ and $\alpha_{PS}$ quantify the direct effects of the external stressor $S_i$ on depression and precariousness, respectively, while $\alpha_{DP}$ and $\alpha_{PD}$ capture the mutual influence between the two dynamic processes.
+Stochastic fluctuations are introduced via the independent Wiener processes $dW_{D,i}(t)$ and $dW_{P,i}(t)$, scaled by their respective volatility parameters $\sigma_D$ and $\sigma_P$.
+This linear structure offers a transparent and analytically tractable framework for modeling how exogenous variation in financial stress may propagate through individual dynamics of depression and precariousness.
+Crucially, all parameters can be calibrated directly from empirical covariances and variances observed in the HELIUS cohort.
+The resulting steady-state distribution of the model aligns with empirical measures: PHQ-9 total score (PHQsum) for depression, and the derived social precariousness factor (P.soc) for social precarity.
+The full calibration procedure is described in [@sec-calibration].
+
+<!-- Here, $D$ denotes depression, $P$ denotes social precariousness, and $S$ is the external stressor associated with financial stress.
+The parameters $\lambda_D$ and $\lambda_P$ control the timescales over which each variable adjusts toward its input-driven target state.
+The coefficients $\alpha_{PS}$ and $\alpha_{DS}$ quantify the direct effects of the stressor on $P$ and $D$, while $\alpha_{DP}$ and $\alpha_{PD}$ represent the cross-influences between depression and precariousness.
+Stochastic fluctuations are introduced through the noise terms $\sigma_P, dW_P$ and $\sigma_D, dW_D$, with $dW_P$ and $dW_D$ representing independent Wiener processes.
+-->
+<!-- This linear structure offers a transparent and analytically tractable framework for examining how variations in external financial stress influence the population-level dynamics of depression and social precariousness.
+Crucially, the model's simplicity enables direct calibration from empirical data: all parameters can be derived analytically from observed variances and covariances in the HELIUS cohort.
+The following  [@sec-calibration] details the procedure used to perform this calibration.
+-->
+
+### Model Calibration {#sec-calibration}
+
+In our linear model, all parameters are derived analytically to reproduce the empirical second-order moments observed in the HELIUS dataset.
+The system treats depression ($D$) and social precariousness ($P$) as jointly evolving under the influence of a shared external stressor ($S$), forming a coupled system of Ornstein–Uhlenbeck processes.
+Under the assumption of stationarity, the steady-state covariances between these variables admit closed-form solutions for the system’s structural parameters.
+
+We fix the adjustment rates $\lambda_D = \lambda_P = 1$, which normalizes the time scale given the cross-sectional nature of the HELIUS data.
+To resolve the remaining degree of freedom, we treat the feedback from precariousness to depression ($\alpha_{DP}$) as a free parameter, and we solve algebraically for all other coefficients — specifically the direct effects of stress ($\alpha_{DS}, \alpha_{PS}$), the reverse feedback from depression to precariousness ($\alpha_{PD}$), and the noise terms ($\sigma_D, \sigma_P$).
+These solutions are expressed as closed-form functions of the empirical covariances: $\mathrm{Var}(D)$, $\mathrm{Var}(P)$, $\mathrm{Var}(S)$, $\mathrm{Cov}(D, P)$, $\mathrm{Cov}(D, S)$, and $\mathrm{Cov}(P, S)$.
+
+This structure creates a family of analytically calibrated models, each consistent with the same observed covariances but differing in their internal configuration.
+We constrain the space of valid solutions by imposing that all structural parameters must be positive, reflecting the assumption that stress and precariousness contribute adversely to depression.
+This constraint restricts the allowable range of $\alpha_{DP}$ to approximately $[0, 0.698]$ (see Appendix [@sec-analcalibration]).
+
+@fig-tradeoff illustrates how the remaining parameters adjust as $\alpha_{DP}$ increases to maintain consistency with the empirical covariances.
+As $\alpha_{DP}$ increases, $\alpha_{PD}$ tends to decrease, reflecting a compensatory trade-off between the mutual influences of depression and precariousness.
+This interplay helps preserve the model’s match with observed covariances, without requiring major shifts in external input strength.
+Meanwhile, the external coupling terms — $\alpha_{DS}$ and $\alpha_{PS}$ — also shift to absorb variance in a way that maintains alignment with the empirical data.
+Notably, the noise amplitudes $\sigma_D$ and $\sigma_P$ decline slightly as $\alpha_{DP}$ increases.
+This indicates that more of the observed variability is captured by deterministic interactions, with less left to be explained by stochastic fluctuations.
+
+Rather than being a limitation, this flexibility enables exploration of a spectrum of equally data-consistent feedback regimes.
+Each represents a different plausible causal architecture, offering insight into how internal dynamics modulate the system's responsiveness to external stress — even when the observed covariances remain unchanged.
+
+```{r}
+#| label: fig-tradeoff
+#| fig-cap: >
+#|   Trade-off curves showing how calibrated parameters vary with $\alpha_{DP}$.
+#|   *Each line shows how a different parameter adjusts in response to changes in $\alpha_{DP}$ to maintain consistency with empirical covariances from the HELIUS data. 
+#|   All configurations shown are equally consistent with the data, illustrating how internal dynamics adjust as feedback strength changes.*
+#| fig-align: center
+#| out-width: 100%
+knitr::include_graphics("img/param_tradeoff.pdf")
+```
+
+### Intervention Simulation
+
+To evaluate how internal structure affects system responsiveness, we simulate a family of linear models subjected to external intervention.
+Each model corresponds to a different value of $\alpha_{DP}$, representing the strength of influence from social precariousness to depression.
+The range spans from 0 (no coupling) to 0.65 (maximal admissible value), with all other parameters derived analytically to preserve the empirical covariances observed in the HELIUS data (see @fig-tradeoff).
+
+In each simulation, financial stress of individual $i$ ($S_i$) is proportionally reduced toward the lowest observed value, modeling a population-wide intervention applied at varying intensities (0–100%).
+For each model, we simulate 1000 virtual individuals whose baseline stress levels are sampled from the empirical distribution and scaled according to the intervention level.
+
+We assess the system’s behavior using two complementary approaches:
+
+- Snapshot analysis captures the average levels of depression ($D$) and precariousness ($P$) at selected time points ($t = 100$, $300$, $500$) following intervention onset.
+Since all linear models converge to the same equilibrium, this highlights how internal feedback affects the speed and pathway of adjustment.
+
+- Full trajectory simulation tracks system dynamics over time for two representative models: one with minimal ($\alpha_{DP} = 0$) and one with maximal ($\alpha_{DP} = 0.65$) feedback.
+Each simulation includes an intervention period followed by its withdrawal, revealing differences in reactivity, delay, and recovery stability.
+
+Together, these simulations demonstrate how internal structure governs not just whether change occurs, but how it unfolds — shaping the pace, persistence, and variability of response to external support.
+
+# Data Availibility
+The data used in this study come from the HELIUS (HEalthy LIfe in an Urban Setting) cohort.
+Due to privacy and ethical restrictions, the raw HELIUS data are not publicly available.
+Researchers who wish to access the data can submit a request to the HELIUS study team via the study website (<https://heliusstudy.nl>), following the governance procedures of the HELIUS Executive Board.
+
+# Code Availibility
+All analysis scripts used for causal discovery, model calibration, and simulation are available at <https://github.com/KyuriP/Precarity_HELIUS>.
+The repository contains documented R code and instructions for reproducing the analyses reported in this manuscript.
 

### 2025-08-25 draft/draft_v3.tex
add draft_v3 (nature format)

--- 2025-08-14 draft/draft_v2.tex
@@
 \renewcommand\Affilfont{\small}
-
 \usepackage{amssymb}
@@
 \usepackage{geometry}
-\usepackage{hyperref}
 
@@
   pdftitle={From Causal Discovery to Intervention Simulation: Modeling Precariousness and Depression in the HELIUS Study},
-  pdfauthor={Kyuri Park ; Leonie K. Elsenburg; Mary Nicolao; Karien Stronks; Vítor V. Vasconcelos},
+  pdfauthor={Kyuri Park ; Leonie K. Elsenburg; Mary Nicolaou; Karien Stronks; Vítor V. Vasconcelos},
   colorlinks=true,
@@
 \author[2]{%
-  Mary Nicolao%
+  Mary Nicolaou%
   %
@@
 
-\date{2025-08-14}
+\date{2025-08-25}
 
@@
 Mental health disorders represent a growing global health concern,
-especially in urbanized regions (Gruebner et al., 2017; World Health
-Organization, 2022). In 2019, one in eight people worldwide were living
-with a mental health condition, with the highest disability-adjusted
-life years (DALYs) due to mental and addictive disorders present in
-high-income countries such as those in Northern Europe, North America,
-and Australia (Rehm \& Shield, 2019; World Health Organization, 2022).
+especially in urbanized regions\textsuperscript{1,2}. In 2019, one in
+eight people worldwide were living with a mental health condition, with
+the highest disability-adjusted life years (DALYs) due to mental and
+addictive disorders present in high-income countries such as those in
+Northern Europe, North America, and Australia\textsuperscript{1,3}.
 Despite growing awareness, governments have struggled to design
@@
 structures of urban environments---making both diagnosis and
-intervention design deeply challenging (Van Der Wal et al., 2021).
+intervention design deeply challenging\textsuperscript{4}.
 
@@
 show elevated overall DALYs, within-country disparities reveal that
-income inequality is a significant predictor of mental health burden
-(Rehm \& Shield, 2019). Moreover, precarious working conditions, housing
-instability, and neighborhood disadvantage have all been linked to
-increased risk of depression and anxiety (Fone et al., 2014; Pevalin et
-al., 2017; Rönnblad et al., 2019; Rugulies et al., 2023). This growing
-literature highlights that poor mental health is not merely a personal
-issue but one rooted in sustained socioeconomic stress and structural
-uncertainty.
+income inequality is a significant predictor of mental health
+burden\textsuperscript{3}. Moreover, precarious working conditions,
+housing instability, and neighborhood disadvantage have all been linked
+to increased risk of depression and anxiety\textsuperscript{5--8}. This
+growing literature highlights that poor mental health is not merely a
+personal issue but one rooted in sustained socioeconomic stress and
+structural uncertainty.
 
@@
 instability and lack of control across several life domains, including
-employment, finances, housing, social relations, and cultural belonging
-(Elsenburg et al., 2025; McKee et al., 2017). This broader view reveals
-how different forms of insecurity can co-occur and compound, forming an
-ecosystem of risk. However, despite growing evidence of association
-between precariousness and mental health, a fundamental question remains
+employment, finances, housing, social relations, and cultural
+belonging\textsuperscript{9,10}. This broader view reveals how different
+forms of insecurity can co-occur and compound, forming an ecosystem of
+risk. However, despite growing evidence of association between
+precariousness and mental health, a fundamental question remains
 unresolved: \emph{How do these factors causally influence one another?}
@@
 relationship affect the outcomes of stress-reducing interventions?} To
-answer this, we combine two approaches. First, we use cycle-capable
-causal discovery algorithms to infer directional relationships among
-domain-specific precariousness factors and depressive symptoms. By
-examining both aggregated measures (depression sum score and a composite
-precariousness index) and disaggregated variables (individual depressive
-symptoms and specific life-domain indicators), we capture broad patterns
-of relationship as well as fine-grained causal pathways. Second, we
+address this question, we analyze cross-sectional data from more than
+21,000 participants in the HELIUS (HEalthy LIfe in an Urban Setting)
+study\textsuperscript{11,12} using a two-step approach. First, we apply
+cycle-capable causal discovery algorithms to infer directional
+relationships among financial stress, domain-specific precariousness,
+and depressive symptoms. By examining both aggregated measures (a
+depression sum score and a composite precariousness index) and
+disaggregated indicators (individual symptoms and life-domain factors),
+we capture both broad patterns and fine-grained pathways. Second, we
 translate these insights into a computational dynamical model that
 simulates how depression and precariousness co-evolve under varying
-levels of external stress. This model enables us to probe how system
-properties---such as feedback strength and stochastic noise---govern
-responsiveness to intervention. When the effects of depression symptoms
-onto the social precarity subsystem are weaker than the reverse path,
-there is higher sensitivity of depression to intervention, reflecting
-both a faster reduction post financial intervention but also quicker
-relapse once financial stresses are reestablished. By integrating causal
-discovery with dynamic simulation, we demonstrate that intervention
-strategies can be tailored to the underlying architecture of
-vulnerability.
-
-\section{Methods}\label{methods}
+levels of external stress. This framework enables us to examine how
+system properties---such as feedback strength and stochastic
+noise---shape responsiveness to intervention.
+
+Figure~\ref{fig-concept} presents this high-level model, derived from
+the causal discovery analysis and serving as the foundation for our
+simulation experiments. It highlights financial stress (\(S_i\)) as an
+upstream driver influencing both depressive symptoms (\(D_i\)) and
+social precariousness (\(P_i\)). Depression contributes to broader
+social vulnerability by feeding into precariousness, while
+precariousness feeds back into depression, creating reinforcing cycles.
+This framework illustrates that intervention effects depend not only on
+external inputs but also on the internal feedback architecture of the
+system. Full methodological details, including data processing,
+algorithmic specifications, and model calibration procedures, are
+provided in the \hyperref[sec-methods]{Methods} section.
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.4\textwidth,height=\textheight]{img/conceptmodel.pdf}
+
+}
+
+\caption{\label{fig-concept}The model abstracts the main dependencies
+revealed by causal discovery into a compact system of dynamic
+interactions. Financial stress (\(S_i\)) is treated as an upstream
+driver that influences both depressive symptoms (\(D_i(t)\)) and social
+precariousness (\(P_i(t)\)). Depression and precariousness also
+influence each other through feedback loops, creating the potential for
+reinforcing cycles of vulnerability. Arrows represent influence
+coefficients (\(\alpha\)), which correspond to terms in the system of
+stochastic differential equations used for simulation. This high-level
+structure served as the foundation for our simulation experiments. Full
+methodological details, refer to the \hyperref[sec-methods]{Methods}
+section.}
+
+\end{figure}%
+
+\section{Results}\label{results}
+
+We present the results in two parts. First, we report the causal
+structures inferred from the HELIUS data at the levels of aggregate
+depression, individual symptoms, and aggregated precariousness. Second,
+we embed these findings in a simplified dynamical model to simulate how
+alternative feedback structures shape the system's response to
+stress-reducing interventions.
+
+\subsection{Causal Structure Linking Precariousness and
+Depression}\label{causal-structure-linking-precariousness-and-depression}
+
+\subsubsection{Depression as sum score}\label{depression-as-sum-score}
+
+Figure~\ref{fig-sum} provides a high-level summary of how precariousness
+factors collectively influence and are influenced by overall depression
+severity, focusing on the causal relationships between precariousness
+factors (\emph{P.hou}, \emph{P.emp}, \emph{P.soc}, \emph{S.rel},
+\emph{S.fin}) and the depression sum score (\emph{PHQsum}), as
+identified by the FCI algorithm.
+
+In panel (a) we see that employment precariousness (\emph{P.emp}) and
+social precariousness (\emph{P.soc}) are not identified as causes of
+depression, whereas financial stress (\emph{S.fin}) appears to play a
+causal role. While \emph{P.emp} and \emph{P.soc} are generally not
+recognized as causes of other precariousness factors, \emph{S.fin}
+emerges as a potential cause, as indicated by its circle edge endpoint.
+This pattern is further supported by the edge-type proportion matrix
+shown in panel (b), which aggregates edge inference outcomes across
+bootstrapped datasets.
+
+Relational stress (\emph{S.rel}) plays a more nuanced role, interacting
+with depression and \emph{S.fin} through a latent confounder. Its
+connection to \emph{P.soc} is less definitive---the proportion matrix in
+Figure~\ref{fig-sum-2} suggests that this relationship is frequently
+absent, ambiguous, or potentially bidirectional across bootstrap
+samples. Meanwhile, housing precariousness (\emph{P.hou}) is not
+directly causally related to depression but is linked to \emph{P.emp}
+and \emph{S.fin}. While \emph{P.emp} and \emph{S.fin} are identified as
+non-causes of \emph{P.hou}, it remains unclear whether \emph{P.hou}
+causally influences \emph{P.emp} or if their relationship is mediated by
+an unobserved confounder.
+
+The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
+largely consistent with those from the FCI analysis, with one key
+distinction: CCI shows greater ambiguity in the directional relationship
+between \emph{S.fin} and \emph{PHQsum}: CCI assigns an arrowhead from
+\emph{S.fin} to \emph{PHQsum} in roughly 50\% of bootstraps, while the
+remaining cases display an undirected (circle) endpoint. Overall, CCI
+introduces greater uncertainty in edge orientations relative to FCI,
+yielding more undirected endpoints and more frequent bidirectional
+(confounded) edges, which in turn contribute to a slightly higher
+overall count of arrowheads when directions are inferred. Both
+algorithms identify directional pathways from the stressors
+(\emph{S.rel}, \emph{S.fin}) to \emph{P.soc} and suggest that
+\emph{P.soc} is not a cause of \emph{PHQsum}, but more likely affected
+by it. \emph{P.emp} is not inferred to cause any of its neighboring
+variables (\emph{P.hou}, \emph{PHQsum}, \emph{S.fin}) and instead
+appears to be affected by them. \emph{P.hou}, in turn, is connected to
+\emph{P.emp} and \emph{S.fin} with circle endpoints, indicating that it
+may be a potential source of influence rather than an outcome.
+
+Causal relationships between five precariousness factors and depression
+sum score (PHQsum), as estimated using the FCI algorithm across
+bootstrapped samples. (a) Aggregated causal graph showing the most
+frequent edge types across bootstraps. Dashed edges indicate ties, with
+a bold horizontal bar marking where the tie occurs. (b) Proportion
+matrix summarizing edge types across all bootstraps. Each matrix cell
+\([i, j]\) represents how often an edge from variable \(i\) to variable
+\(j\) exhibited a particular edge mark---arrowhead, tail, circle, or no
+edge---adjacent to \(j\). The edge mark in matrix\([i,j]\) reflects the
+orientation of the edge ending at \(j\). For example, if
+matrix\([i, j]\) shows a green arrowhead (\(>\)) and matrix\([j, i]\)
+shows a blue circle (\(\circ\)), then the inferred relationship is
+\(i \circarrow j\). Edge types are encoded by distinct colors: green for
+arrowheads, coral for arrowtails, blue for circles, and light gray for
+absence of an edge. Colors are blended and shaded by frequency, with
+greater opacity indicating higher confidence in the inferred edge type.
+
+\begin{figure}
+
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{img/FCI_depsum.png}
+
+}
+
+\subcaption{\label{fig-sum-1}FCI PAG}
+
+\end{minipage}%
+%
+\begin{minipage}{0.50\linewidth}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{img/depsum_mat_fci.pdf}
+
+}
+
+\subcaption{\label{fig-sum-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-sum}Causal relationships between five precariousness
+factors and depression sum score (\emph{PHQsum}), as estimated using the
+FCI algorithm across bootstrapped samples. \textbf{(a)} Aggregated
+causal graph showing the most frequent edge types across bootstraps.
+Dashed edges indicate ties, with a bold horizontal bar marking where the
+tie occurs. \textbf{(b)} Proportion matrix summarizing edge types across
+all bootstraps: each matrix cell \([i, j]\) indicates how often a given
+edge from variable \(i\) to variable \(j\) exhibited a particular edge
+mark (e.g., arrowhead, tail, or circle) adjacent to \(j\). Green
+indicates arrowheads, coral arrowtails, blue circle marks, and gray
+absence of an edge. Colors are blended and shaded by frequency, with
+dominant symbols appearing more opaque. For example,
+matrix{[}\(\mathit{S.fin}\), \(\mathit{PHQsum}\){]} shows a green
+arrowhead (\(>\)) and matrix{[}\(\mathit{PHQsum}\),
+\(\mathit{S.fin}\){]} shows a coral arrowtail (\textemdash), together
+indicating that \(\mathit{S.fin} \rightarrow \mathit{PHQsum}\) in the
+majority of bootstrapped graphs.}
+
+\end{figure}%
+
+\subsubsection{Individual depression
+symptom}\label{individual-depression-symptom}
+
+Moving from the sum score representation to the disaggregated
+symptom-level graph provides a more granular perspective on the causal
+relationships between precariousness factors and depression symptoms.
+Unlike the sum score graph, which aggregates all symptoms into a single
+measure, the symptom-level graph in Figure~\ref{fig-sym} reveals the
+heterogeneity in how specific symptoms (\emph{con} (concentration),
+\emph{slp} (sleep), \emph{ene} (energy), \emph{app} (appetite),
+\emph{mot} (motor), \emph{sui} (suicidal), \emph{anh} (anhedonia),
+\emph{glt} (guilt), and \emph{dep} (depressed mood)) are influenced by,
+and in turn influence, different forms of precariousness.
+
+Figure~\ref{fig-sym-1} reveals a strong interdependence among symptoms
+and suggests much presence of latent confounders, as indicated by
+numerous bidirectional edges. In Figure~\ref{fig-sym-2}, the proportion
+matrix indicates that arrowheads are the most frequently inferred edge
+type. However, circle endpoints are particularly common for \emph{anh},
+\emph{slp}, \emph{ene}, and \emph{sui}, reflecting uncertainty in the
+causal direction among these symptoms. Notably, the only arrowtail
+connection appears between \emph{dep} and \emph{anh}, suggesting that
+anhedonia is a potential cause of depressed mood.
+
+Looking at the symptom-precariousness connections, one of the key
+patterns is that ties are most frequently found in the relationships
+between individual symptoms and precariousness factors. This suggests
+that these causal links may be less stable across different conditions.
+Closer examination reveals that most ties arise due to discrepancies
+between the Gaussian CI test and the RCoT test---with Gaussian CI
+favoring arrowheads and RCoT more often predicting the absence of an
+edge.
+
+Despite these differences, some consistent patterns emerge across both
+CI tests, particularly in the case of \emph{S.fin}, which appears to be
+connected to nearly all other variables---though many of these
+connections are marked by ties, reflecting uncertainty in
+directionality. One exception is the stronger evidence suggesting that
+\emph{S.fin} may cause changes in \emph{app}, while its relationships
+with other symptoms, such as \emph{ene}, \emph{mot}, and \emph{dep},
+remain ambiguous. Similarly, \emph{P.soc} exhibits numerous connections
+with depressive symptoms, with most edges pointing toward \emph{P.soc}
+rather than outward from it. This suggests that depressive symptoms,
+particularly \emph{dep} and \emph{glt}, may contribute to worsening
+social precariousness rather than the other way around. Additionally,
+\emph{anh} also shows signs of a potential causal influence on
+\emph{P.soc}, reinforcing the idea that social precariousness is more
+often a consequence rather than a driver of depressive symptoms.
+
+Other precariousness factors exhibit more uncertainty but still notable
+relationships. \emph{S.rel} is connected to \emph{slp}, \emph{glt}, and
+\emph{app}, though directionality remains unclear in many cases.
+\emph{P.emp} also connects to symptoms like \emph{slp}, \emph{mot}, and
+\emph{con}, but, like \emph{S.rel}, these relationships exhibit a 50/50
+split in directionality, reflecting uncertainty in the inferred causal
+paths. In line with the aggregated graph in Figure~\ref{fig-sum},
+\emph{P.hou} does not appear to have any direct relationship with
+depression symptoms but consistently shows associations with
+\emph{P.emp} and \emph{S.fin}. The causal relationships among
+precariousness factors remain largely unchanged from the aggregated
+analysis, with \emph{S.fin} exhibiting the strongest tendency to
+influence other precariousness factors.
+
+Among depressive symptoms, \emph{dep}, \emph{glt}, and \emph{slp} appear
+to be the most connected to precariousness factors, while \emph{P.soc}
+has the most connections with symptoms, predominantly as a recipient
+rather than a driver of influence. Within the symptom network,
+\emph{slp}, \emph{sui}, and \emph{anh} emerge as causally influential
+symptoms, as they exhibit more outgoing arrows compared to other
+symptoms. On the other hand, \emph{con}, \emph{mot}, \emph{dep}, and
+\emph{glt}, despite having high connectivity, predominantly receive
+incoming arrows, indicating they are more likely effects rather than
+causes. Considering both symptom-precariousness connections and
+symptom-level dynamics, \emph{slp} emerges as a central symptom, given
+its strong ties to precariousness factors and its influential role
+within the symptom network. \emph{P.soc} appears to function as a bridge
+between the depression and precariousness subsystems, as depressive
+symptoms appear to feed back into social precariousness, reinforcing a
+self-sustaining dynamic between depression and precarious conditions.
+
+Finally, the results from CCI are largely consistent with those from
+FCI, but with some key differences. CCI tends to favor arrowheads more
+frequently, resulting in a greater number of bidirectional edges, which
+suggests a higher involvement of latent confounders. Additionally, CCI
+exhibits more tie situations, though, similar to FCI, most ties occur
+between absence and arrowhead edges, reflecting the discrepancy between
+the Gaussian CI test and RCoT---where the Gaussian test favors
+arrowheads, while RCoT more often suggests the absence of an edge. For a
+detailed visualization of the CCI-derived graph and proportion matrix,
+see Figure~\ref{fig-sym-cci}.
+
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/symptom_graph_FCI.png}
+
+}
+
+\subcaption{\label{fig-sym-1}FCI PAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.7\textwidth,height=\textheight]{img/symptom_mat_fci.pdf}
+
+}
+
+\subcaption{\label{fig-sym-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-sym}Resulting graph of precariousness factors and
+individual depression symptoms using FCI and proportion of edge endpoint
+types.}
+
+\end{figure}%
+
+\subsubsection{Precariousness as sum
+score}\label{precariousness-as-sum-score}
+
+Lastly, we examine the relationships between individual symptoms and
+overall precariousness, an aggregated measure represented by the sum of
+five precariousness factors. This analysis provides a complementary
+perspective, capturing broader patterns that may not be evident in the
+disaggregated symptom-precariousness analysis. By aggregating
+precariousness factors into a single score, this approach may capture
+distributed relationships across different precariousness factors that
+were previously overlooked in the disaggregated analysis.
+
+Unlike the disaggregated graph in Figure~\ref{fig-sym}, the aggregated
+symptom-precariousness graph (Figure~\ref{fig-presum-1}) shows greater
+certainty in causal directions, with ties occurring only in edges
+involving the overall precariousness factor. This is even more evident
+in the proportion matrix (Figure~\ref{fig-presum-2}), where
+symptom-to-symptom interactions almost fully converge to a single edge
+type. Additionally, all symptom interactions are connected either
+through bidirectional edges or a combination of an arrowhead and a
+circle, suggesting a strong presence of latent confounders.
+
+Examining the symptom-precariousness connections, we observe a stronger
+overall trend of depressive symptoms influencing precariousness, rather
+than the other way around. Specifically, \emph{anh}, \emph{dep}, and
+\emph{slp} frequently emerge as sources of influence on precariousness,
+whereas \emph{app}, \emph{glt}, and \emph{mot} exhibit more uncertainty
+in directionality, often resulting in circle endpoints. This pattern
+suggests that when precariousness factors are aggregated, the dominant
+causal flow is from depressive symptoms to precariousness, rather than
+precariousness driving depression. Particularly, \emph{dep} emerges as
+the strongest predictor of precariousness, exhibiting the highest
+proportion of arrowtails, consistent with findings from the
+disaggregated analysis.
+
+Comparing this with the CCI-derived graph (Figure~\ref{fig-presum-cci}),
+several key differences stand out. The CCI results reveal that most
+symptoms are interconnected through bidirectional edges, suggesting a
+strong presence of latent confounders. However, \emph{dep} appears to be
+a distinct exception, as it is predominantly caused by nearly all other
+symptoms, with one exception \emph{sui}, which does not contribute to
+\emph{dep}. Similarly, \emph{con} is influenced by multiple symptoms,
+albeit with weaker support compared to \emph{dep}. Most interestingly,
+the CCI results highlight a distinct feedback loop between \emph{dep}
+and overall precariousness, represented by a tail-tail edge
+(\textemdash ). This suggests a possible reinforcing cycle between
+depression and overall precariousness, where depression not only arises
+from precarious conditions but also contributes to their persistence,
+creating a self-sustaining dynamic.
+
+Across multiple levels of analysis, our results point to a consistent
+and directional link between precariousness and depression. Financial
+stress (\emph{S.fin}) emerges as a particularly robust and modifiable
+upstream driver: in both aggregated and disaggregated models, it shows
+strong and stable causal influence on depressive symptoms. By contrast,
+other forms of precariousness---particularly social precarity---more
+often appear as outcomes rather than antecedents of depression,
+frequently receiving rather than sending causal arrows. Among symptoms,
+\emph{dep} and \emph{slp} stand out for their centrality and connections
+to precariousness factors, with \emph{dep} also implicated in potential
+feedback loops with social precarity (\emph{P.soc}). These findings
+motivate the use of a simplified three-node model in the
+simulation---capturing depression, financial stress, and social
+precarity---as it preserves the key dynamic motifs observed in the
+causal graphs while enabling tractable exploration of system-level
+responses to stress-reducing interventions.
+
+\begin{figure}
+
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_graph_FCI.png}
+
+}
+
+\subcaption{\label{fig-presum-1}FCI PAG}
+
+\end{minipage}%
+\newline
+\begin{minipage}{\linewidth}
+
+\centering{
+
+\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_mat_fci.pdf}
+
+}
+
+\subcaption{\label{fig-presum-2}Proportion matrix}
+
+\end{minipage}%
+
+\caption{\label{fig-presum}Resulting graph of precariousness sum score
+and individual depression symptoms using FCI and proportion of edge
+endpoint types. As before, dashed edges denote ties, with specific ties
+marked by bold horizontal bars.}
+
+\end{figure}%
+
+\subsection{Simulating the Dynamics Between Precariousness and
+Depression Under
+Intervention}\label{simulating-the-dynamics-between-precariousness-and-depression-under-intervention}
+
+To assess how internal structure influences the system's responsiveness
+to external support, we simulate a set of linear models that are
+consistent with the data but vary in the assumed strength of influence
+from precariousness to depression (\(\alpha_{DP}\)). For each model, we
+generate a virtual population of 1,000 individuals and simulate an
+external intervention that reduces each individual's financial stress
+(\(S\)) proportionally toward the minimum observed value in the dataset.
+While individuals begin with different baseline stress levels, the
+intervention applies a shared proportional reduction across the
+population. This intervention defines the external input, while
+variation in \(\alpha_{DP}\) reflects different assumptions about
+internal feedback from precariousness to depression. We examine system
+behavior in two complementary ways: (1) by analyzing average levels of
+depression and precariousness at selected time points following
+intervention onset, and (2) by simulating full temporal trajectories
+under sustained intervention followed by recovery, using two
+representative cases at the extremes of the \(\alpha_{DP}\) range.
+
+Figure~\ref{fig-intervention} summarizes how variation in
+\(\alpha_{DP}\) shapes system-level responses to external support. In
+Panel A, during the early phase of intervention, depression and
+precariousness exhibit contrasting patterns. Depression declines more
+rapidly in scenarios with lower \(\alpha_{DP}\), where its dynamics are
+driven more directly by the external stressor \(S\). As \(\alpha_{DP}\)
+increases, \(D\) becomes more dependent on precariousness (\(P\)), which
+itself adjusts more slowly---delaying the depressive response.
+Precariousness, by contrast, shows slightly faster early reductions in
+scenarios with higher \(\alpha_{DP}\). This reflects a shift in
+influence: as \(\alpha_{DP}\) increases, \(P\) becomes more directly
+responsive to \(S\) (via increasing \(\alpha_{PS}\)) and less influenced
+by \(D\), allowing it to adjust more promptly to changes in the
+stressor.
+
+Despite these short-term differences, all scenarios ultimately converge
+to similar steady-state levels under sustained intervention. AS the
+systm approaches convergence, some variation remains, particularly in
+\(P\), where gradients between models are still visible. This
+convergence highlights that the key distinctions lie not in whether
+intervention is effective, but in how quickly and through which pathways
+change unfolds.
+
+Panel B shows full time trajectories for two representative cases: one
+with minimal internal feedback (\(\alpha_{DP} = 0\)) and one at the
+upper admissible limit (\(\alpha_{DP} = 0.65\)). In both, there is a
+temporary intervention period. When \(\alpha_{DP} = 0\), depression
+responds rapidly to the reduction in stress and rebounds shortly after
+support is withdrawn. In contrast, when \(\alpha_{DP}\) is high, \(D\)
+responds more gradually and recovers more slowly---reflecting its
+increased dependence on the slower-changing \(P\). Precariousness
+declines in both scenarios during intervention and returns to baseline
+afterward, but with a slightly sharper drop in the higher
+\(\alpha_{DP}\) condition. This occurs because \(P\) becomes more
+directly responsive to \(S\), as consistency with the data requires
+\(\alpha_{PS}\) to increase with \(\alpha_{DP}\) (see
+Figure~\ref{fig-tradeoff}). These parameter shifts reflect trade-offs
+needed to preserve the empirical covariances, with further details
+provided in Appendix Section~\ref{sec-analcalibration}.
+
+These asymmetries further highlight how internal structure---not just
+external inputs---shapes the system's trajectory. Specifically, changes
+in coupling parameters reconfigure how stress reduction propagates
+through the system, altering the speed and pathway of response even when
+long-run outcomes remain unchanged.
+
+\newgeometry{top=20mm, bottom=20mm, left=40mm, right=40mm}
+
+\begin{figure}
+
+\centering{
+
+\includegraphics[width=0.8\textwidth,height=\textheight]{img/intervention_linear.pdf}
+
+}
+
+\caption{\label{fig-intervention}System response to external
+intervention across variations in internal structure.
+\emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
+precariousness (\(P\)) at time steps 100, 300, and 500, across models
+that differ in the coupling coefficient \(\alpha_{DP}\), which
+represents the strength of depression's influence on precariousness in
+the dynamical model. The x‑axis shows intervention strength, defined as
+the proportional reduction of financial stress (\(S\)) relative to its
+baseline (with higher values indicating more external financial
+support). Each line corresponds to a different model defined by a
+specific \(\alpha_{DP}\) value, color-coded from low (purple) to high
+(yellow). Models with lower \(\alpha_{DP}\) exhibit faster early
+reductions in \(D\), while higher \(\alpha_{DP}\) leads to more
+immediate declines in \(P\). \textbf{Panel B}: Full time trajectories
+for two models---one with no coupling (\(\alpha_{DP} = 0\)) and one with
+strong coupling (\(\alpha_{DP} = 0.65\))---illustrating system response
+and recovery following intervention onset (step 300) and withdrawal
+(step 1500). Depression responds more gradually and persistently in the
+high \(\alpha_{DP}\) model, whereas precariousness adjusts more
+quickly.}}
+
+\end{figure}%
+
+\restoregeometry
+
+\section{Discussion}\label{discussion}
+
+This study combined constraint-based causal discovery and dynamical
+modeling to investigate how depression and socioeconomic precariousness
+interact and how their internal structure modulates responsiveness to
+intervention. Consistent with prior work linking financial strain to
+psychological distress\textsuperscript{13,14}, our analysis identified
+financial stress as a key upstream driver of depression in both
+aggregated and disaggregated models. This supports longstanding
+hypotheses that economic stressors can act as modifiable entry points
+into broader cycles of psychosocial risk.
+
+At the symptom level, our analysis aligns with established theories of
+psychopathology but provides a more fine-grained map of potential causal
+relationships, generating concrete, testable hypotheses. It also points
+to possible limitations of prevailing models, as some observed
+symptom--symptom links may be shaped by shared confounders rather than
+direct causal influence. This suggests that, in some contexts, broader
+mechanisms or clusters of symptoms may provide more effective units of
+analysis and intervention than individual symptoms alone. Within this
+framework, sleep disturbance emerged as a potentially initiating
+symptom, with directed connections to both symptom-level and
+precariousness variables---mirroring longitudinal findings that identify
+sleep disruption as an early warning signal for depressive episodes and
+social withdrawal\textsuperscript{15,16}. Depressed mood, another core
+symptom, occupied a central position in the network. Not only was it
+influenced by upstream symptoms like anhedonia and guilt, but it also
+showed directional influence on social precariousness---suggesting that
+internal psychological states may actively shape external vulnerability.
+The CCI algorithm further identified a potential feedback loop between
+depressed mood and overall precariousness, resonating with social drift
+theories in depression, which posit that declining mental health can
+lead to social exclusion, job instability, and housing insecurity,
+reinforcing depressive cycles\textsuperscript{17,18}.
+
+In contrast, social precariousness itself appeared more often as an
+effect than a cause, receiving arrows from symptoms like depressed mood
+and guilt but rarely sending them. This asymmetry supports prior
+observations that indicators of social vulnerability often reflect
+downstream consequences of poor mental health, though the relationships
+are likely bidirectional over time\textsuperscript{13}. Although our
+findings are based on Amsterdam-specific data, they align with research
+in other high-income urban settings, where structural inequalities such
+as poverty, unemployment, and housing insecurity consistently predict
+poor psychological outcomes\textsuperscript{19,20}. While local contexts
+may alter specific pathways, these parallels suggest that similar
+mechanisms could operate elsewhere, though replication in other
+populations is needed to assess robustness.
+
+Taken together, these findings yield actionable insights for mental
+health intervention. Symptoms like sleep disturbance, which demonstrate
+early and broad influence across both symptom and precariousness
+networks, may serve as effective targets for prevention---potentially
+halting the progression into more entrenched depressive states or
+worsening precariousness. Similarly, targeting symptoms such as
+depressed mood, which appear embedded in potential feedback loops
+linking psychological and social domains, may offer strategic leverage
+for breaking mutually reinforcing cycles of mental decline and social
+isolation. Importantly, interventions targeting such nodes may have
+ripple effects across domains. For instance, evidence suggests that
+treating core depressive symptoms can improve social functioning,
+employment engagement, and housing stability---particularly when
+combined with social or financial supports\textsuperscript{21,22}. These
+dual benefits underscore the value of integrated, cross-domain
+strategies that simultaneously address internal symptoms and external
+stressors.
+
+Our results also highlight why single-point interventions may fail to
+produce lasting change. The broader causal structure reveals multiple
+interlocking feedback loops spanning both psychological symptoms and
+structural risk factors, which may re-amplify vulnerability even after
+temporary relief. Supporting this, recent simulation work
+by\textsuperscript{23} shows that feedback loops, even when modest in
+strength, can substantially diminish the long-term impact of isolated
+interventions unless multiple reinforcing links are simultaneously
+disrupted. Our own simulations support this view: while all models
+converged to similar long-term outcomes under sustained support,
+differences in internal structure---specifically, the strength of the
+link from precariousness to depression---produced systematic variation
+in how quickly and through which pathways systems responded to
+intervention. Stronger coupling slowed recovery and prolonged
+depression, showing that internal architecture can be as important as
+external input. This supports growing calls for systems-oriented mental
+health perspectives\textsuperscript{24}. Rather than assuming uniform
+responses, more durable outcomes may require (1) targeting early-acting
+symptoms, (2) sustaining support where recovery is slow, and (3)
+combining structural and psychological interventions. Tailoring
+strategies to match the system's architecture can improve effectiveness,
+particularly in populations with layered vulnerabilities. In this light,
+understanding causal architecture is not merely theoretical---it is
+foundational to designing more effective and equitable interventions.
+
+While the FCI and CCI algorithms are designed to recover causal
+structure from cross-sectional data---including feedback and latent
+confounding---many edge directions in our results remain ambiguous,
+particularly in links between individual symptoms and precariousness
+factors. These inconsistencies likely reflect limited statistical signal
+in high-dimensional settings, where large conditioning sets reduce power
+and small effect sizes are harder to detect. Moreover, our findings
+showed sensitivity to the choice of conditional independence (CI) test.
+The Gaussian CI test, which assumes linearity and Gaussian noise, often
+produced denser graphs, likely due to overestimating weak or noisy
+linear associations. In contrast, RCoT, a nonparametric alternative,
+avoids these assumptions but relies on a Gaussian radial basis function
+(RBF) kernel to assess dependence. While flexible in capturing nonlinear
+relationships, this kernel is optimized for smooth, continuous variables
+and may underperform with discrete or ordinal data. Its similarity
+estimates can be unreliable when applied to variables with
+discontinuities, potentially leading to missed dependencies. This is
+particularly relevant in the HELIUS dataset, where some variables are
+ordinal or mixed-type\textsuperscript{25}. To address this, future work
+should consider adapting kernel-based CI tests to better handle mixed
+data structures---for example, by employing hybrid kernels that
+explicitly combine continuous and discrete similarity measures. This
+could improve the reliability of nonparametric causal discovery in
+settings with heterogeneous variable types.
+
+Although the CCI algorithm is explicitly designed to detect cycles,
+aside from the connection between precarity and depression in the
+aggregate precarity setting, we found little clear evidence of feedback:
+most candidate loops appeared only as bidirectional edges, which may
+reflect latent confounding or statistical imprecision. This could point
+to either an absence of strong feedback at the symptom and the
+precariousness factor levels or limitations in the sensitivity of
+full-graph discovery approaches applied to densely interconnected
+variables. More focused strategies, such as local structure tests or
+theory-guided subgraph modeling, may help resolve these ambiguities.
+Moreover, our analyses relied solely on observational data from the
+Amsterdam population. Some structural uncertainties, particularly those
+involving latent confounding or bidirectional influence, may ultimately
+require interventional or semi-interventional designs. Extensions such
+as LLC\textsuperscript{26}, NODAGS-Flow\textsuperscript{27}, or
+Bicycle\textsuperscript{28}, which integrate observational and
+interventional data, could be especially valuable where such data are
+available.
+
+Finally, the structure of our dynamical model introduces important
+limitations. The model focused on a restricted set of variables
+(depression, social precariousness, and financial stress). This design
+was intentionally simplified to isolate core mechanisms of interest, but
+it necessarily omits other relevant domains of precariousness,
+additional stressors, and potential unintended consequences of
+intervention. In addition, while the linear formulation offered
+analytical tractability and transparent calibration, it likely
+oversimplifies the true dynamics of depression and socioeconomic
+precariousness. A growing body of evidence suggests that psychosocial
+systems are fundamentally nonlinear. Empirical and theoretical work
+points to phenomena such as threshold effects, where improvements only
+occur after a tipping point is crossed\textsuperscript{29,30};
+saturation, where the impact of support diminishes as basic needs are
+met\textsuperscript{31}; and path dependence, where past adversity
+shapes sensitivity to future stressors\textsuperscript{32}. These
+behaviors are fundamentally incompatible with linear dynamics, yet they
+may be essential for understanding why some individuals remain trapped
+in high-risk states while others recover more easily. Future modeling
+efforts should therefore incorporate mechanisms such as saturating
+input-response functions (e.g., sigmoidal form) and state-dependent
+feedback (where coupling strength varies by system state). Incorporating
+such features would allow models to capture not just average effects but
+also the critical conditions under which meaningful change becomes
+possible---clarifying whether some subgroups require forceful,
+threshold-crossing interventions, while others may benefit from
+sustained, moderate support.
+
+As public health research has long shown, social and psychological
+systems do not always respond passively to external inputs; internal
+constraints can limit the impact of even well-intentioned
+policies\textsuperscript{33,34}. By using causal discovery and
+computational modeling, researchers and policymakers can better identify
+these structural constraints and tailor interventions accordingly. More
+broadly, this study offers a generalizable framework for investigating
+complex mental health dynamics. Causal discovery supports principled
+hypothesis generation from observational data, while dynamical modeling
+enables exploration of how hypothesized structures behave under external
+change. Together, these tools move us beyond static descriptions toward
+mechanistic insight with direct relevance for real-world mental health
+policy and intervention design. Future research should build on this
+foundation by developing richer, more nuanced dynamical models grounded
+in longitudinal and interventional data. Incorporating time, multiple
+domains of precarity, and adaptive feedback processes would help assess
+whether the mechanisms proposed manifest in actual mental health
+trajectories. Ultimately, we hope this study encourages further
+integration of causal structure learning and systems modeling in efforts
+to understand and improve mental health outcomes under conditions of
+socioeconomic precariousness.
+
+\section{Methods}\label{sec-methods}
 
@@
 
-We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study
-(Snijder et al., 2017; Stronks et al., 2013). It captures the diverse
-population of the city of Amsterdam by including the six main ethnic
-groups and provides comprehensive health and lifestyle data, including
-depressive symptoms as measured by the PHQ-9 (Galenkamp et al., 2017).
-To operationalize indicators of precariousness, we draw on the framework
-outlined in previous research (Elsenburg et al., 2025) and select a set
-of relevant variables. To ensure a robust representation of
-precariousness in our causal discovery models, we conducted exploratory
-analyses to identify consistent and meaningful data-driven factor
-structures. These analyses led to the identification of five distinct
-precariousness factors: three reflecting structural
-domains---employment, housing, and social relationships---and two
-capturing recent life stressors---financial and relational. Each factor
-is composed of multiple observed variables, as outlined below.
-Additional details on the HELIUS study and the factor extraction process
-are provided in the \hyperref[sec-appendix]{Appendix}.
+We use data from the HELIUS (HEalthy LIfe in an Urban Setting)
+study\textsuperscript{11,12}. It captures the diverse population of the
+city of Amsterdam by including the six main ethnic groups and provides
+comprehensive health and lifestyle data, including depressive symptoms
+as measured by the PHQ-9\textsuperscript{35}. To operationalize
+indicators of precariousness, we draw on the framework outlined in
+previous research\textsuperscript{9} and select a set of relevant
+variables. To ensure a robust representation of precariousness in our
+causal discovery models, we conducted exploratory analyses to identify
+consistent and meaningful data-driven factor structures. These analyses
+led to the identification of five distinct precariousness factors: three
+reflecting structural domains---employment, housing, and social
+relationships---and two capturing recent life stressors---financial and
+relational. Each factor is composed of multiple observed variables, as
+outlined below. Additional details on the HELIUS study and the factor
+extraction process are provided in the
+\hyperref[sec-appendix]{Appendix}.
 
@@
 of accumulated precariousness across domains. Depression is represented
-using the PHQ-9 (Kroenke et al., 2001), both as a total sum score and as
+using the PHQ-9\textsuperscript{36}, both as a total sum score and as
 individual symptom scores. In the subsequent causal discovery analysis,
@@
 
-\includegraphics[width=0.9\textwidth,height=\textheight]{draft_v2_files/figure-pdf/fig-dist-1.pdf}
+\includegraphics[width=0.9\textwidth,height=\textheight]{draft_v3_files/figure-pdf/fig-dist-1.pdf}
 
@@
 Inference} (CCI) algorithms, both of which can detect cycles and
-non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
-reference, we also include the \emph{PC} algorithm, one of the most
-widely used methods for acyclic causal discovery (Spirtes et al., 2001).
-While our full analysis includes results from three causal discovery
-algorithms---FCI, CCI, and PC---we focus on FCI in the main text, as it
-offers a strong balance between flexibility and theoretical soundness.
-FCI accommodates latent confounding and potential feedback, making it
-well-suited for the complexity of our data (Mooij \& Claassen, 2020).
-The PC algorithm is included for reference but relies on more
-restrictive assumptions, such as acyclicity and no unmeasured
-confounding. CCI, while capable of modeling cycles explicitly, has its
-own theoretical limitations; for example, its output may not fully
-preserve the d-separation properties of the true causal graph (Strobl,
-2019). Results from both PC and CCI are provided in the
-\hyperref[sec-appendix]{Appendix}
+non-acyclic structures\textsuperscript{37,38}. For reference, we also
+include the \emph{PC} algorithm, one of the most widely used methods for
+acyclic causal discovery\textsuperscript{39}. While our full analysis
+includes results from three causal discovery algorithms---FCI, CCI, and
+PC---we focus on FCI in the main text, as it offers a strong balance
+between flexibility and theoretical soundness. FCI accommodates latent
+confounding and potential feedback, making it well-suited for the
+complexity of our data\textsuperscript{37}. The PC algorithm is included
+for reference but relies on more restrictive assumptions, such as
+acyclicity and no unmeasured confounding. CCI, while capable of modeling
+cycles explicitly, has its own theoretical limitations; for example, its
+output may not fully preserve the d-separation properties of the true
+causal graph\textsuperscript{38}. Results from both PC and CCI are
+provided in the \hyperref[sec-appendix]{Appendix}
 (Section~\ref{sec-cci}, Section~\ref{sec-pc}) for completeness and
@@
 due to the need to invert large kernel matrices---making them
-impractical for large datasets like HELIUS (Rahimi \& Recht, 2007). To
+impractical for large datasets like HELIUS\textsuperscript{40}. To
 address this, we use the Randomized Conditional Correlation Test (RCoT),
@@
 quadratic to linear in sample size, substantially lowering runtime while
-preserving sensitivity to nonlinear relationships (Strobl et al., 2019;
-Zhang et al., 2012). For further details, see Section~\ref{sec-rcot}.
+preserving sensitivity to nonlinear
+relationships\textsuperscript{41,42}. For further details, see
+Section~\ref{sec-rcot}.
 
@@
 
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.4\textwidth,height=\textheight]{img/conceptmodel.pdf}
-
-}
-
-\caption{\label{fig-concept}High-level schematic of the dynamic model
-structure used in simulation. Arrows represent influence coefficients
-\(\alpha\), matching the terms in the system of stochastic differential
-equations. \(D_i(t)\) and \(P_i(t)\) denote depressive symptoms and
-social precariousness for individual \(i\), with \(S_i\) representing
-static financial stress.}
-
-\end{figure}%
-
 \subsubsection{Model Calibration}\label{sec-calibration}
@@
 
-\section{Results}\label{results}
-
-\subsection{Causal Structure Linking Precariousness and
-Depression}\label{causal-structure-linking-precariousness-and-depression}
-
-\subsubsection{Depression as sum score}\label{depression-as-sum-score}
-
-Figure~\ref{fig-sum} provides a high-level summary of how precariousness
-factors collectively influence and are influenced by overall depression
-severity, focusing on the causal relationships between precariousness
-factors (\emph{P.hou}, \emph{P.emp}, \emph{P.soc}, \emph{S.rel},
-\emph{S.fin}) and the depression sum score (\emph{PHQsum}), as
-identified by the FCI algorithm.
-
-In panel (a) we see that employment precariousness (\emph{P.emp}) and
-social precariousness (\emph{P.soc}) are not identified as causes of
-depression, whereas financial stress (\emph{S.fin}) appears to play a
-causal role. While \emph{P.emp} and \emph{P.soc} are generally not
-recognized as causes of other precariousness factors, \emph{S.fin}
-emerges as a potential cause, as indicated by its circle edge endpoint.
-This pattern is further supported by the edge-type proportion matrix
-shown in panel (b), which aggregates edge inference outcomes across
-bootstrapped datasets.
-
-Relational stress (\emph{S.rel}) plays a more nuanced role, interacting
-with depression and \emph{S.fin} through a latent confounder. Its
-connection to \emph{P.soc} is less definitive---the proportion matrix in
-Figure~\ref{fig-sum-2} suggests that this relationship is frequently
-absent, ambiguous, or potentially bidirectional across bootstrap
-samples. Meanwhile, housing precariousness (\emph{P.hou}) is not
-directly causally related to depression but is linked to \emph{P.emp}
-and \emph{S.fin}. While \emph{P.emp} and \emph{S.fin} are identified as
-non-causes of \emph{P.hou}, it remains unclear whether \emph{P.hou}
-causally influences \emph{P.emp} or if their relationship is mediated by
-an unobserved confounder.
-
-The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
-largely consistent with those from the FCI analysis, with one key
-distinction: CCI shows greater ambiguity in the directional relationship
-between \emph{S.fin} and \emph{PHQsum}: CCI assigns an arrowhead from
-\emph{S.fin} to \emph{PHQsum} in roughly 50\% of bootstraps, while the
-remaining cases display an undirected (circle) endpoint. Overall, CCI
-introduces greater uncertainty in edge orientations relative to FCI,
-yielding more undirected endpoints and more frequent bidirectional
-(confounded) edges, which in turn contribute to a slightly higher
-overall count of arrowheads when directions are inferred. Both
-algorithms identify directional pathways from the stressors
-(\emph{S.rel}, \emph{S.fin}) to \emph{P.soc} and suggest that
-\emph{P.soc} is not a cause of \emph{PHQsum}, but more likely affected
-by it. \emph{P.emp} is not inferred to cause any of its neighboring
-variables (\emph{P.hou}, \emph{PHQsum}, \emph{S.fin}) and instead
-appears to be affected by them. \emph{P.hou}, in turn, is connected to
-\emph{P.emp} and \emph{S.fin} with circle endpoints, indicating that it
-may be a potential source of influence rather than an outcome.
-
-Causal relationships between five precariousness factors and depression
-sum score (PHQsum), as estimated using the FCI algorithm across
-bootstrapped samples. (a) Aggregated causal graph showing the most
-frequent edge types across bootstraps. Dashed edges indicate ties, with
-a bold horizontal bar marking where the tie occurs. (b) Proportion
-matrix summarizing edge types across all bootstraps. Each matrix cell
-\([i, j]\) represents how often an edge from variable \(i\) to variable
-\(j\) exhibited a particular edge mark---arrowhead, tail, circle, or no
-edge---adjacent to \(j\). The edge mark in matrix\([i,j]\) reflects the
-orientation of the edge ending at \(j\). For example, if
-matrix\([i, j]\) shows a green arrowhead (\(>\)) and matrix\([j, i]\)
-shows a blue circle (\(\circ\)), then the inferred relationship is
-\(i \circarrow j\). Edge types are encoded by distinct colors: green for
-arrowheads, coral for arrowtails, blue for circles, and light gray for
-absence of an edge. Colors are blended and shaded by frequency, with
-greater opacity indicating higher confidence in the inferred edge type.
-
-\begin{figure}
-
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{img/FCI_depsum.png}
-
-}
-
-\subcaption{\label{fig-sum-1}FCI PAG}
-
-\end{minipage}%
-%
-\begin{minipage}{0.50\linewidth}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{img/depsum_mat_fci.pdf}
-
-}
-
-\subcaption{\label{fig-sum-2}Proportion matrix}
-
-\end{minipage}%
-
-\caption{\label{fig-sum}Causal relationships between five precariousness
-factors and depression sum score (\emph{PHQsum}), as estimated using the
-FCI algorithm across bootstrapped samples. \textbf{(a)} Aggregated
-causal graph showing the most frequent edge types across bootstraps.
-Dashed edges indicate ties, with a bold horizontal bar marking where the
-tie occurs. \textbf{(b)} Proportion matrix summarizing edge types across
-all bootstraps: each matrix cell \([i, j]\) indicates how often a given
-edge from variable \(i\) to variable \(j\) exhibited a particular edge
-mark (e.g., arrowhead, tail, or circle) adjacent to \(j\). Green
-indicates arrowheads, coral arrowtails, blue circle marks, and gray
-absence of an edge. Colors are blended and shaded by frequency, with
-dominant symbols appearing more opaque. For example,
-matrix{[}\(\mathit{S.fin}\), \(\mathit{PHQsum}\){]} shows a green
-arrowhead (\(>\)) and matrix{[}\(\mathit{PHQsum}\),
-\(\mathit{S.fin}\){]} shows a coral arrowtail (\textemdash), together
-indicating that \(\mathit{S.fin} \rightarrow \mathit{PHQsum}\) in the
-majority of bootstrapped graphs.}
-
-\end{figure}%
-
-\subsubsection{Individual depression
-symptom}\label{individual-depression-symptom}
-
-Moving from the sum score representation to the disaggregated
-symptom-level graph provides a more granular perspective on the causal
-relationships between precariousness factors and depression symptoms.
-Unlike the sum score graph, which aggregates all symptoms into a single
-measure, the symptom-level graph in Figure~\ref{fig-sym} reveals the
-heterogeneity in how specific symptoms (\emph{con} (concentration),
-\emph{slp} (sleep), \emph{ene} (energy), \emph{app} (appetite),
-\emph{mot} (motor), \emph{sui} (suicidal), \emph{anh} (anhedonia),
-\emph{glt} (guilt), and \emph{dep} (depressed mood)) are influenced by,
-and in turn influence, different forms of precariousness.
-
-Figure~\ref{fig-sym-1} reveals a strong interdependence among symptoms
-and suggests much presence of latent confounders, as indicated by
-numerous bidirectional edges. In Figure~\ref{fig-sym-2}, the proportion
-matrix indicates that arrowheads are the most frequently inferred edge
-type. However, circle endpoints are particularly common for \emph{anh},
-\emph{slp}, \emph{ene}, and \emph{sui}, reflecting uncertainty in the
-causal direction among these symptoms. Notably, the only arrowtail
-connection appears between \emph{dep} and \emph{anh}, suggesting that
-anhedonia is a potential cause of depressed mood.
-
-Looking at the symptom-precariousness connections, one of the key
-patterns is that ties are most frequently found in the relationships
-between individual symptoms and precariousness factors. This suggests
-that these causal links may be less stable across different conditions.
-Closer examination reveals that most ties arise due to discrepancies
-between the Gaussian CI test and the RCoT test---with Gaussian CI
-favoring arrowheads and RCoT more often predicting the absence of an
-edge.
-
-Despite these differences, some consistent patterns emerge across both
-CI tests, particularly in the case of \emph{S.fin}, which appears to be
-connected to nearly all other variables---though many of these
-connections are marked by ties, reflecting uncertainty in
-directionality. One exception is the stronger evidence suggesting that
-\emph{S.fin} may cause changes in \emph{app}, while its relationships
-with other symptoms, such as \emph{ene}, \emph{mot}, and \emph{dep},
-remain ambiguous. Similarly, \emph{P.soc} exhibits numerous connections
-with depressive symptoms, with most edges pointing toward \emph{P.soc}
-rather than outward from it. This suggests that depressive symptoms,
-particularly \emph{dep} and \emph{glt}, may contribute to worsening
-social precariousness rather than the other way around. Additionally,
-\emph{anh} also shows signs of a potential causal influence on
-\emph{P.soc}, reinforcing the idea that social precariousness is more
-often a consequence rather than a driver of depressive symptoms.
-
-Other precariousness factors exhibit more uncertainty but still notable
-relationships. \emph{S.rel} is connected to \emph{slp}, \emph{glt}, and
-\emph{app}, though directionality remains unclear in many cases.
-\emph{P.emp} also connects to symptoms like \emph{slp}, \emph{mot}, and
-\emph{con}, but, like \emph{S.rel}, these relationships exhibit a 50/50
-split in directionality, reflecting uncertainty in the inferred causal
-paths. In line with the aggregated graph in Figure~\ref{fig-sum},
-\emph{P.hou} does not appear to have any direct relationship with
-depression symptoms but consistently shows associations with
-\emph{P.emp} and \emph{S.fin}. The causal relationships among
-precariousness factors remain largely unchanged from the aggregated
-analysis, with \emph{S.fin} exhibiting the strongest tendency to
-influence other precariousness factors.
-
-Among depressive symptoms, \emph{dep}, \emph{glt}, and \emph{slp} appear
-to be the most connected to precariousness factors, while \emph{P.soc}
-has the most connections with symptoms, predominantly as a recipient
-rather than a driver of influence. Within the symptom network,
-\emph{slp}, \emph{sui}, and \emph{anh} emerge as causally influential
-symptoms, as they exhibit more outgoing arrows compared to other
-symptoms. On the other hand, \emph{con}, \emph{mot}, \emph{dep}, and
-\emph{glt}, despite having high connectivity, predominantly receive
-incoming arrows, indicating they are more likely effects rather than
-causes. Considering both symptom-precariousness connections and
-symptom-level dynamics, \emph{slp} emerges as a central symptom, given
-its strong ties to precariousness factors and its influential role
-within the symptom network. \emph{P.soc} appears to function as a bridge
-between the depression and precariousness subsystems, as depressive
-symptoms appear to feed back into social precariousness, reinforcing a
-self-sustaining dynamic between depression and precarious conditions.
-
-Finally, the results from CCI are largely consistent with those from
-FCI, but with some key differences. CCI tends to favor arrowheads more
-frequently, resulting in a greater number of bidirectional edges, which
-suggests a higher involvement of latent confounders. Additionally, CCI
-exhibits more tie situations, though, similar to FCI, most ties occur
-between absence and arrowhead edges, reflecting the discrepancy between
-the Gaussian CI test and RCoT---where the Gaussian test favors
-arrowheads, while RCoT more often suggests the absence of an edge. For a
-detailed visualization of the CCI-derived graph and proportion matrix,
-see Figure~\ref{fig-sym-cci}.
-
-\begin{figure}
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/symptom_graph_FCI.png}
-
-}
-
-\subcaption{\label{fig-sym-1}FCI PAG}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.7\textwidth,height=\textheight]{img/symptom_mat_fci.pdf}
-
-}
-
-\subcaption{\label{fig-sym-2}Proportion matrix}
-
-\end{minipage}%
-
-\caption{\label{fig-sym}Resulting graph of precariousness factors and
-individual depression symptoms using FCI and proportion of edge endpoint
-types.}
-
-\end{figure}%
-
-\subsubsection{Precariousness as sum
-score}\label{precariousness-as-sum-score}
-
-Lastly, we examine the relationships between individual symptoms and
-overall precariousness, an aggregated measure represented by the sum of
-five precariousness factors. This analysis provides a complementary
-perspective, capturing broader patterns that may not be evident in the
-disaggregated symptom-precariousness analysis. By aggregating
-precariousness factors into a single score, this approach may capture
-distributed relationships across different precariousness factors that
-were previously overlooked in the disaggregated analysis.
-
-Unlike the disaggregated graph in Figure~\ref{fig-sym}, the aggregated
-symptom-precariousness graph (Figure~\ref{fig-presum-1}) shows greater
-certainty in causal directions, with ties occurring only in edges
-involving the overall precariousness factor. This is even more evident
-in the proportion matrix (Figure~\ref{fig-presum-2}), where
-symptom-to-symptom interactions almost fully converge to a single edge
-type. Additionally, all symptom interactions are connected either
-through bidirectional edges or a combination of an arrowhead and a
-circle, suggesting a strong presence of latent confounders.
-
-Examining the symptom-precariousness connections, we observe a stronger
-overall trend of depressive symptoms influencing precariousness, rather
-than the other way around. Specifically, \emph{anh}, \emph{dep}, and
-\emph{slp} frequently emerge as sources of influence on precariousness,
-whereas \emph{app}, \emph{glt}, and \emph{mot} exhibit more uncertainty
-in directionality, often resulting in circle endpoints. This pattern
-suggests that when precariousness factors are aggregated, the dominant
-causal flow is from depressive symptoms to precariousness, rather than
-precariousness driving depression. Particularly, \emph{dep} emerges as
-the strongest predictor of precariousness, exhibiting the highest
-proportion of arrowtails, consistent with findings from the
-disaggregated analysis.
-
-Comparing this with the CCI-derived graph (Figure~\ref{fig-presum-cci}),
-several key differences stand out. The CCI results reveal that most
-symptoms are interconnected through bidirectional edges, suggesting a
-strong presence of latent confounders. However, \emph{dep} appears to be
-a distinct exception, as it is predominantly caused by nearly all other
-symptoms, with one exception \emph{sui}, which does not contribute to
-\emph{dep}. Similarly, \emph{con} is influenced by multiple symptoms,
-albeit with weaker support compared to \emph{dep}. Most interestingly,
-the CCI results highlight a distinct feedback loop between \emph{dep}
-and overall precariousness, represented by a tail-tail edge
-(\textemdash ). This suggests a possible reinforcing cycle between
-depression and overall precariousness, where depression not only arises
-from precarious conditions but also contributes to their persistence,
-creating a self-sustaining dynamic.
-
-Across multiple levels of analysis, our results point to a consistent
-and directional link between precariousness and depression. Financial
-stress (\emph{S.fin}) emerges as a particularly robust and modifiable
-upstream driver: in both aggregated and disaggregated models, it shows
-strong and stable causal influence on depressive symptoms. By contrast,
-other forms of precariousness---particularly social precarity---more
-often appear as outcomes rather than antecedents of depression,
-frequently receiving rather than sending causal arrows. Among symptoms,
-\emph{dep} and \emph{slp} stand out for their centrality and connections
-to precariousness factors, with \emph{dep} also implicated in potential
-feedback loops with social precarity (\emph{P.soc}). These findings
-motivate the use of a simplified three-node model in the
-simulation---capturing depression, financial stress, and social
-precarity---as it preserves the key dynamic motifs observed in the
-causal graphs while enabling tractable exploration of system-level
-responses to stress-reducing interventions.
-
-\begin{figure}
-
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_graph_FCI.png}
-
-}
-
-\subcaption{\label{fig-presum-1}FCI PAG}
-
-\end{minipage}%
-\newline
-\begin{minipage}{\linewidth}
-
-\centering{
-
-\includegraphics[width=0.6\textwidth,height=\textheight]{img/presum_mat_fci.pdf}
-
-}
-
-\subcaption{\label{fig-presum-2}Proportion matrix}
-
-\end{minipage}%
-
-\caption{\label{fig-presum}Resulting graph of precariousness sum score
-and individual depression symptoms using FCI and proportion of edge
-endpoint types. As before, dashed edges denote ties, with specific ties
-marked by bold horizontal bars.}
-
-\end{figure}%
-
-\subsection{Simulating the Dynamics Between Precariousness and
-Depression Under
-Intervention}\label{simulating-the-dynamics-between-precariousness-and-depression-under-intervention}
-
-To assess how internal structure influences the system's responsiveness
-to external support, we simulate a set of linear models that are
-consistent with the data but vary in the assumed strength of influence
-from precariousness to depression (\(\alpha_{DP}\)). For each model, we
-generate a virtual population of 1,000 individuals and simulate an
-external intervention that reduces each individual's financial stress
-(\(S\)) proportionally toward the minimum observed value in the dataset.
-While individuals begin with different baseline stress levels, the
-intervention applies a shared proportional reduction across the
-population. This intervention defines the external input, while
-variation in \(\alpha_{DP}\) reflects different assumptions about
-internal feedback from precariousness to depression. We examine system
-behavior in two complementary ways: (1) by analyzing average levels of
-depression and precariousness at selected time points following
-intervention onset, and (2) by simulating full temporal trajectories
-under sustained intervention followed by recovery, using two
-representative cases at the extremes of the \(\alpha_{DP}\) range.
-
-Figure~\ref{fig-intervention} summarizes how variation in
-\(\alpha_{DP}\) shapes system-level responses to external support. In
-Panel A, during the early phase of intervention, depression and
-precariousness exhibit contrasting patterns. Depression declines more
-rapidly in scenarios with lower \(\alpha_{DP}\), where its dynamics are
-driven more directly by the external stressor \(S\). As \(\alpha_{DP}\)
-increases, \(D\) becomes more dependent on precariousness (\(P\)), which
-itself adjusts more slowly---delaying the depressive response.
-Precariousness, by contrast, shows slightly faster early reductions in
-scenarios with higher \(\alpha_{DP}\). This reflects a shift in
-influence: as \(\alpha_{DP}\) increases, \(P\) becomes more directly
-responsive to \(S\) (via increasing \(\alpha_{PS}\)) and less influenced
-by \(D\), allowing it to adjust more promptly to changes in the
-stressor.
-
-Despite these short-term differences, all scenarios ultimately converge
-to similar steady-state levels under sustained intervention. AS the
-systm approaches convergence, some variation remains, particularly in
-\(P\), where gradients between models are still visible. This
-convergence highlights that the key distinctions lie not in whether
-intervention is effective, but in how quickly and through which pathways
-change unfolds.
-
-Panel B shows full time trajectories for two representative cases: one
-with minimal internal feedback (\(\alpha_{DP} = 0\)) and one at the
-upper admissible limit (\(\alpha_{DP} = 0.65\)). In both, there is a
-temporary intervention period. When \(\alpha_{DP} = 0\), depression
-responds rapidly to the reduction in stress and rebounds shortly after
-support is withdrawn. In contrast, when \(\alpha_{DP}\) is high, \(D\)
-responds more gradually and recovers more slowly---reflecting its
-increased dependence on the slower-changing \(P\). Precariousness
-declines in both scenarios during intervention and returns to baseline
-afterward, but with a slightly sharper drop in the higher
-\(\alpha_{DP}\) condition. This occurs because \(P\) becomes more
-directly responsive to \(S\), as consistency with the data requires
-\(\alpha_{PS}\) to increase with \(\alpha_{DP}\) (see
-Figure~\ref{fig-tradeoff}). These parameter shifts reflect trade-offs
-needed to preserve the empirical covariances, with further details
-provided in Appendix Section~\ref{sec-analcalibration}.
-
-These asymmetries further highlight how internal structure---not just
-external inputs---shapes the system's trajectory. Specifically, changes
-in coupling parameters reconfigure how stress reduction propagates
-through the system, altering the speed and pathway of response even when
-long-run outcomes remain unchanged.
-
-\newgeometry{top=20mm, bottom=20mm, left=40mm, right=40mm}
-
-\begin{figure}
-
-\centering{
-
-\includegraphics[width=0.8\textwidth,height=\textheight]{img/intervention_linear.pdf}
-
-}
-
-\caption{\label{fig-intervention}System response to external
-intervention across variations in internal structure.
-\emph{\textbf{Panel A}: Mean levels of depression (\(D\)) and
-precariousness (\(P\)) at time steps 100, 300, and 500, across varying
-values of \(\alpha_{DP}\) and intervention strengths. The x-axis
-represents the strength of external support, operationalized as the
-proportional reduction of financial stress (\(S\)) toward its minimum
-observed value. Each line corresponds to a different model defined by a
-specific \(\alpha_{DP}\) value, color-coded from low (purple) to high
-(yellow). Models with lower \(\alpha_{DP}\) exhibit faster early
-reductions in \(D\), while higher \(\alpha_{DP}\) leads to more
-immediate declines in \(P\). \textbf{Panel B}: Full time trajectories
-for two models---one with no coupling (\(\alpha_{DP} = 0\)) and one with
-strong coupling (\(\alpha_{DP} = 0.65\))---illustrating system response
-and recovery following intervention onset (step 300) and withdrawal
-(step 1500). Depression responds more gradually and persistently in the
-high \(\alpha_{DP}\) model, whereas precariousness adjusts more
-quickly.}}
-
-\end{figure}%
-
-\restoregeometry
-
-\section{Discussion}\label{discussion}
-
-This study combined constraint-based causal discovery and dynamical
-modeling to investigate how depression and socioeconomic precariousness
-interact and how their internal structure modulates responsiveness to
-intervention. Consistent with prior work linking financial strain to
-psychological distress (Ezzy, 1993; Lund et al., 2010), our analysis
-identified financial stress as a key upstream driver of depression in
-both aggregated and disaggregated models. This supports longstanding
-hypotheses that economic stressors can act as modifiable entry points
-into broader cycles of psychosocial risk.
-
-At the symptom level, our analysis aligns with established theories of
-psychopathology but provides a more fine-grained map of potential causal
-relationships, generating concrete, testable hypotheses. It also points
-to possible limitations of prevailing models, as some observed
-symptom--symptom links may be shaped by shared confounders rather than
-direct causal influence. This suggests that, in some contexts, broader
-mechanisms or clusters of symptoms may provide more effective units of
-analysis and intervention than individual symptoms alone. Within this
-framework, sleep disturbance emerged as a potentially initiating
-symptom, with directed connections to both symptom-level and
-precariousness variables---mirroring longitudinal findings that identify
-sleep disruption as an early warning signal for depressive episodes and
-social withdrawal (Alvaro et al., 2013; Zawadzki et al., 2013).
-Depressed mood, another core symptom, occupied a central position in the
-network. Not only was it influenced by upstream symptoms like anhedonia
-and guilt, but it also showed directional influence on social
-precariousness---suggesting that internal psychological states may
-actively shape external vulnerability. The CCI algorithm further
-identified a potential feedback loop between depressed mood and overall
-precariousness, resonating with social drift theories in depression,
-which posit that declining mental health can lead to social exclusion,
-job instability, and housing insecurity, reinforcing depressive cycles
-(Boschloo et al., 2015; Compton \& Shim, 2015).
-
-In contrast, social precariousness itself appeared more often as an
-effect than a cause, receiving arrows from symptoms like depressed mood
-and guilt but rarely sending them. This asymmetry supports prior
-observations that indicators of social vulnerability often reflect
-downstream consequences of poor mental health, though the relationships
-are likely bidirectional over time (Lund et al., 2010). Although our
-findings are based on Amsterdam-specific data, they align with research
-in other high-income urban settings, where structural inequalities such
-as poverty, unemployment, and housing insecurity consistently predict
-poor psychological outcomes (Fryers et al., 2003; Weich \& Lewis, 1998).
-While local contexts may alter specific pathways, these parallels
-suggest that similar mechanisms could operate elsewhere, though
-replication in other populations is needed to assess robustness.
-
-Taken together, these findings yield actionable insights for mental
-health intervention. Symptoms like sleep disturbance, which demonstrate
-early and broad influence across both symptom and precariousness
-networks, may serve as effective targets for prevention---potentially
-halting the progression into more entrenched depressive states or
-worsening precariousness. Similarly, targeting symptoms such as
-depressed mood, which appear embedded in potential feedback loops
-linking psychological and social domains, may offer strategic leverage
-for breaking mutually reinforcing cycles of mental decline and social
-isolation. Importantly, interventions targeting such nodes may have
-ripple effects across domains. For instance, evidence suggests that
-treating core depressive symptoms can improve social functioning,
-employment engagement, and housing stability---particularly when
-combined with social or financial supports (Lund et al., 2018; McDaid et
-al., 2019). These dual benefits underscore the value of integrated,
-cross-domain strategies that simultaneously address internal symptoms
-and external stressors.
-
-Our results also highlight why single-point interventions may fail to
-produce lasting change. The broader causal structure reveals multiple
-interlocking feedback loops spanning both psychological symptoms and
-structural risk factors, which may re-amplify vulnerability even after
-temporary relief. Supporting this, recent simulation work by Park, Li,
-et al. (2025) shows that feedback loops, even when modest in strength,
-can substantially diminish the long-term impact of isolated
-interventions unless multiple reinforcing links are simultaneously
-disrupted. Our own simulations support this view: while all models
-converged to similar long-term outcomes under sustained support,
-differences in internal structure---specifically, the strength of the
-link from precariousness to depression---produced systematic variation
-in how quickly and through which pathways systems responded to
-intervention. Stronger coupling slowed recovery and prolonged
-depression, showing that internal architecture can be as important as
-external input. This supports growing calls for systems-oriented mental
-health perspectives (Borsboom, 2017). Rather than assuming uniform
-responses, more durable outcomes may require (1) targeting early-acting
-symptoms, (2) sustaining support where recovery is slow, and (3)
-combining structural and psychological interventions. Tailoring
-strategies to match the system's architecture can improve effectiveness,
-particularly in populations with layered vulnerabilities. In this light,
-understanding causal architecture is not merely theoretical---it is
-foundational to designing more effective and equitable interventions.
-
-While the FCI and CCI algorithms are designed to recover causal
-structure from cross-sectional data---including feedback and latent
-confounding---many edge directions in our results remain ambiguous,
-particularly in links between individual symptoms and precariousness
-factors. These inconsistencies likely reflect limited statistical signal
-in high-dimensional settings, where large conditioning sets reduce power
-and small effect sizes are harder to detect. Moreover, our findings
-showed sensitivity to the choice of conditional independence (CI) test.
-The Gaussian CI test, which assumes linearity and Gaussian noise, often
-produced denser graphs, likely due to overestimating weak or noisy
-linear associations. In contrast, RCoT, a nonparametric alternative,
-avoids these assumptions but relies on a Gaussian radial basis function
-(RBF) kernel to assess dependence. While flexible in capturing nonlinear
-relationships, this kernel is optimized for smooth, continuous variables
-and may underperform with discrete or ordinal data. Its similarity
-estimates can be unreliable when applied to variables with
-discontinuities, potentially leading to missed dependencies. This is
-particularly relevant in the HELIUS dataset, where some variables are
-ordinal or mixed-type (Howlett, 2001). To address this, future work
-should consider adapting kernel-based CI tests to better handle mixed
-data structures---for example, by employing hybrid kernels that
-explicitly combine continuous and discrete similarity measures. This
-could improve the reliability of nonparametric causal discovery in
-settings with heterogeneous variable types.
-
-Although the CCI algorithm is explicitly designed to detect cycles,
-aside from the connection between precarity and depression in the
-aggregate precarity setting, we found little clear evidence of feedback:
-most candidate loops appeared only as bidirectional edges, which may
-reflect latent confounding or statistical imprecision. This could point
-to either an absence of strong feedback at the symptom and the
-precariousness factor levels or limitations in the sensitivity of
-full-graph discovery approaches applied to densely interconnected
-variables. More focused strategies, such as local structure tests or
-theory-guided subgraph modeling, may help resolve these ambiguities.
-Moreover, our analyses relied solely on observational data from the
-Amsterdam population. Some structural uncertainties, particularly those
-involving latent confounding or bidirectional influence, may ultimately
-require interventional or semi-interventional designs. Extensions such
-as LLC (Hyttinen et al., 2012), NODAGS-Flow (Sethuraman et al., 2023),
-or Bicycle (Rohbeck et al., 2024), which integrate observational and
-interventional data, could be especially valuable where such data are
-available.
-
-Finally, the structure of our dynamical model introduces important
-limitations. The model focused on a restricted set of variables
-(depression, social precariousness, and financial stress). This design
-was intentionally simplified to isolate core mechanisms of interest, but
-it necessarily omits other relevant domains of precariousness,
-additional stressors, and potential unintended consequences of
-intervention. In addition, while the linear formulation offered
-analytical tractability and transparent calibration, it likely
-oversimplifies the true dynamics of depression and socioeconomic
-precariousness. A growing body of evidence suggests that psychosocial
-systems are fundamentally nonlinear. Empirical and theoretical work
-points to phenomena such as threshold effects, where improvements only
-occur after a tipping point is crossed (Leemput et al., 2014; Park,
-Waldorp, et al., 2025); saturation, where the impact of support
-diminishes as basic needs are met (Cramer et al., 2010); and path
-dependence, where past adversity shapes sensitivity to future stressors
-(Wichers et al., 2016). These behaviors are fundamentally incompatible
-with linear dynamics, yet they may be essential for understanding why
-some individuals remain trapped in high-risk states while others recover
-more easily. Future modeling efforts should therefore incorporate
-mechanisms such as saturating input-response functions (e.g., sigmoidal
-form) and state-dependent feedback (where coupling strength varies by
-system state). Incorporating such features would allow models to capture
-not just average effects but also the critical conditions under which
-meaningful change becomes possible---clarifying whether some subgroups
-require forceful, threshold-crossing interventions, while others may
-benefit from sustained, moderate support.
-
-As public health research has long shown, social and psychological
-systems do not always respond passively to external inputs; internal
-constraints can limit the impact of even well-intentioned policies
-(Savigny \& Adam, 2009; Wagenaar \& Burris, 2013). By using causal
-discovery and computational modeling, researchers and policymakers can
-better identify these structural constraints and tailor interventions
-accordingly. More broadly, this study offers a generalizable framework
-for investigating complex mental health dynamics. Causal discovery
-supports principled hypothesis generation from observational data, while
-dynamical modeling enables exploration of how hypothesized structures
-behave under external change. Together, these tools move us beyond
-static descriptions toward mechanistic insight with direct relevance for
-real-world mental health policy and intervention design. Future research
-should build on this foundation by developing richer, more nuanced
-dynamical models grounded in longitudinal and interventional data.
-Incorporating time, multiple domains of precarity, and adaptive feedback
-processes would help assess whether the mechanisms proposed manifest in
-actual mental health trajectories. Ultimately, we hope this study
-encourages further integration of causal structure learning and systems
-modeling in efforts to understand and improve mental health outcomes
-under conditions of socioeconomic precariousness.
+\section{Data Availibility}\label{data-availibility}
+
+The data used in this study come from the HELIUS (HEalthy LIfe in an
+Urban Setting) cohort. Due to privacy and ethical restrictions, the raw
+HELIUS data are not publicly available. Researchers who wish to access
+the data can submit a request to the HELIUS study team via the study
+website (\url{https://heliusstudy.nl}), following the governance
+procedures of the HELIUS Executive Board.
+
+\section{Code Availibility}\label{code-availibility}
+
+All analysis scripts used for causal discovery, model calibration, and
+simulation are available at
+\url{https://github.com/KyuriP/Precarity_HELIUS}. The repository
+contains documented R code and instructions for reproducing the analyses
+reported in this manuscript.
 
@@
 \phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
+\begin{CSLReferences}{0}{0}
+\bibitem[\citeproctext]{ref-WHO_2022}
+\CSLLeftMargin{1. }%
+\CSLRightInline{World Health Organization.
+\href{https://www.who.int/news-room/fact-sheets/detail/mental-disorders}{Mental
+disorders}. (2022).}
+
+\bibitem[\citeproctext]{ref-gruebner2017cities}
+\CSLLeftMargin{2. }%
+\CSLRightInline{Gruebner, O. \emph{et al.} Cities and mental health.
+\emph{Deutsches {Ä}rzteblatt International} \textbf{114}, 121 (2017).}
+
+\bibitem[\citeproctext]{ref-rehm2019global}
+\CSLLeftMargin{3. }%
+\CSLRightInline{Rehm, J. \& Shield, K. D. Global burden of disease and
+the impact of mental and addictive disorders. \emph{Current psychiatry
+reports} \textbf{21}, 1--7 (2019).}
+
+\bibitem[\citeproctext]{ref-van2021advancing}
+\CSLLeftMargin{4. }%
+\CSLRightInline{Van Der Wal, J. M. \emph{et al.} Advancing urban mental
+health research: From complexity science to actionable targets for
+intervention. \emph{The Lancet Psychiatry} \textbf{8}, 991--1000
+(2021).}
+
+\bibitem[\citeproctext]{ref-pevalin2017impact}
+\CSLLeftMargin{5. }%
+\CSLRightInline{Pevalin, D. J., Reeves, A., Baker, E. \& Bentley, R. The
+impact of persistent poor housing conditions on mental health: A
+longitudinal population-based study. \emph{Preventive medicine}
+\textbf{105}, 304--310 (2017).}
+
+\bibitem[\citeproctext]{ref-fone2014effect}
+\CSLLeftMargin{6. }%
+\CSLRightInline{Fone, D. \emph{et al.} Effect of neighbourhood
+deprivation and social cohesion on mental health inequality: A
+multilevel population-based longitudinal study. \emph{Psychological
+medicine} \textbf{44}, 2449--2460 (2014).}
+
+\bibitem[\citeproctext]{ref-rugulies2023work}
+\CSLLeftMargin{7. }%
+\CSLRightInline{Rugulies, R. \emph{et al.} Work-related causes of mental
+health conditions and interventions for their improvement in workplaces.
+\emph{The Lancet} \textbf{402}, 1368--1381 (2023).}
+
+\bibitem[\citeproctext]{ref-ronnblad2019precarious}
+\CSLLeftMargin{8. }%
+\CSLRightInline{Rönnblad, T. \emph{et al.} Precarious employment and
+mental health. \emph{Scandinavian journal of work, environment \&
+health} \textbf{45}, 429--443 (2019).}
+
+\bibitem[\citeproctext]{ref-elsenburg2025clustering}
+\CSLLeftMargin{9. }%
+\CSLRightInline{Elsenburg, L. K., Nicolaou, M., Galenkamp, H.,
+Lakerveld, J. \& Stronks, K. The clustering of disadvantage in different
+life dimensions across ethnic groups: A network analysis of indicators
+of precariousness in the HELIUS study. \emph{Social Science \& Medicine}
+117970 (2025).}
+
+\bibitem[\citeproctext]{ref-mckee2017living}
+\CSLLeftMargin{10. }%
+\CSLRightInline{McKee, M., Reeves, A., Clair, A. \& Stuckler, D. Living
+on the edge: Precariousness and why it matters for health.
+\emph{Archives of Public Health} \textbf{75}, 1--10 (2017).}
+
+\bibitem[\citeproctext]{ref-stronks2013unravelling}
+\CSLLeftMargin{11. }%
+\CSLRightInline{Stronks, K. \emph{et al.} Unravelling the impact of
+ethnicity on health in europe: The HELIUS study. \emph{BMC public
+health} \textbf{13}, 1--10 (2013).}
+
+\bibitem[\citeproctext]{ref-snijder2017cohort}
+\CSLLeftMargin{12. }%
+\CSLRightInline{Snijder, M. B. \emph{et al.} Cohort profile: The healthy
+life in an urban setting (HELIUS) study in amsterdam, the netherlands.
+\emph{BMJ open} \textbf{7}, e017873 (2017).}
+
+\bibitem[\citeproctext]{ref-lund2010poverty}
+\CSLLeftMargin{13. }%
+\CSLRightInline{Lund, C. \emph{et al.} Poverty and common mental
+disorders in low and middle income countries: A systematic review.
+\emph{Social science \& medicine} \textbf{71}, 517--528 (2010).}
+
+\bibitem[\citeproctext]{ref-ezzy1993unemployment}
+\CSLLeftMargin{14. }%
+\CSLRightInline{Ezzy, D. Unemployment and mental health: A critical
+review. \emph{Social science \& medicine} \textbf{37}, 41--52 (1993).}
+
+\bibitem[\citeproctext]{ref-zawadzki2013rumination}
+\CSLLeftMargin{15. }%
+\CSLRightInline{Zawadzki, M. J., Graham, J. E. \& Gerin, W. Rumination
+and anxiety mediate the effect of loneliness on depressed mood and sleep
+quality in college students. \emph{Health Psychology} \textbf{32}, 212
+(2013).}
+
 \bibitem[\citeproctext]{ref-alvaro2013systematic}
-Alvaro, P. K., Roberts, R. M., \& Harris, J. K. (2013). A systematic
-review assessing bidirectionality between sleep disturbances, anxiety,
-and depression. \emph{Sleep}, \emph{36}(7), 1059--1068.
+\CSLLeftMargin{16. }%
+\CSLRightInline{Alvaro, P. K., Roberts, R. M. \& Harris, J. K. A
+systematic review assessing bidirectionality between sleep disturbances,
+anxiety, and depression. \emph{Sleep} \textbf{36}, 1059--1068 (2013).}
+
+\bibitem[\citeproctext]{ref-compton2015social}
+\CSLLeftMargin{17. }%
+\CSLRightInline{Compton, M. T. \& Shim, R. S. The social determinants of
+mental health. \emph{Focus} \textbf{13}, 419--425 (2015).}
+
+\bibitem[\citeproctext]{ref-boschloo2015network}
+\CSLLeftMargin{18. }%
+\CSLRightInline{Boschloo, L. \emph{et al.} The network structure of
+symptoms of the diagnostic and statistical manual of mental disorders.
+\emph{PloS one} \textbf{10}, e0137621 (2015).}
+
+\bibitem[\citeproctext]{ref-fryers2003social}
+\CSLLeftMargin{19. }%
+\CSLRightInline{Fryers, T., Melzer, D. \& Jenkins, R. Social
+inequalities and the common mental disorders: A systematic review of the
+evidence. \emph{Social psychiatry and psychiatric epidemiology}
+\textbf{38}, 229--237 (2003).}
+
+\bibitem[\citeproctext]{ref-weich1998poverty}
+\CSLLeftMargin{20. }%
+\CSLRightInline{Weich, S. \& Lewis, G. Poverty, unemployment, and common
+mental disorders: Population based cohort study. \emph{Bmj}
+\textbf{317}, 115--119 (1998).}
+
+\bibitem[\citeproctext]{ref-mcdaid2019economic}
+\CSLLeftMargin{21. }%
+\CSLRightInline{McDaid, D., Park, A.-L. \& Wahlbeck, K. The economic
+case for the prevention of mental illness. \emph{Annual review of public
+health} \textbf{40}, 373--389 (2019).}
+
+\bibitem[\citeproctext]{ref-lund2018social}
+\CSLLeftMargin{22. }%
+\CSLRightInline{Lund, C. \emph{et al.} Social determinants of mental
+disorders and the sustainable development goals: A systematic review of
+reviews. \emph{The lancet psychiatry} \textbf{5}, 357--369 (2018).}
+
+\bibitem[\citeproctext]{ref-park2025role}
+\CSLLeftMargin{23. }%
+\CSLRightInline{Park, K., Li, X., Waldorp, L., Lees, M. \& Vasconcelos,
+V. V. The role of feedback loops in dynamical symptom networks. (2025).}
 
 \bibitem[\citeproctext]{ref-borsboom2017network}
-Borsboom, D. (2017). A network theory of mental disorders. \emph{World
-Psychiatry}, \emph{16}(1), 5--13.
-
-\bibitem[\citeproctext]{ref-boschloo2015network}
-Boschloo, L., Van Borkulo, C. D., Rhemtulla, M., Keyes, K. M., Borsboom,
-D., \& Schoevers, R. A. (2015). The network structure of symptoms of the
-diagnostic and statistical manual of mental disorders. \emph{PloS One},
-\emph{10}(9), e0137621.
+\CSLLeftMargin{24. }%
+\CSLRightInline{Borsboom, D. A network theory of mental disorders.
+\emph{World psychiatry} \textbf{16}, 5--13 (2017).}
+
+\bibitem[\citeproctext]{ref-howlett2001radial}
+\CSLLeftMargin{25. }%
+\CSLRightInline{Howlett, R. J. Radial basis function networks 1: Recent
+developments in theory and applications. (2001).}
+
+\bibitem[\citeproctext]{ref-hyttinen2012learning}
+\CSLLeftMargin{26. }%
+\CSLRightInline{Hyttinen, A., Eberhardt, F. \& Hoyer, P. O. Learning
+linear cyclic causal models with latent variables. \emph{The Journal of
+Machine Learning Research} \textbf{13}, 3387--3439 (2012).}
+
+\bibitem[\citeproctext]{ref-sethuraman2023nodags}
+\CSLLeftMargin{27. }%
+\CSLRightInline{Sethuraman, M. G. \emph{et al.} NODAGS-flow: Nonlinear
+cyclic causal structure learning. in \emph{International conference on
+artificial intelligence and statistics} 6371--6387 (PMLR, 2023).}
+
+\bibitem[\citeproctext]{ref-rohbeck2024}
+\CSLLeftMargin{28. }%
+\CSLRightInline{Rohbeck, M. \emph{et al.} Bicycle: Intervention-based
+causal discovery with cycles. in \emph{Proceedings of the third
+conference on causal learning and reasoning} (eds. Locatello, F. \&
+Didelez, V.) vol. 236 209--242 (PMLR, 2024).}
+
+\bibitem[\citeproctext]{ref-van2014critical}
+\CSLLeftMargin{29. }%
+\CSLRightInline{Leemput, I. A. van de \emph{et al.} Critical slowing
+down as early warning for the onset and termination of depression.
+\emph{Proceedings of the National Academy of Sciences} \textbf{111},
+87--92 (2014).}
+
+\bibitem[\citeproctext]{ref-park2025individual}
+\CSLLeftMargin{30. }%
+\CSLRightInline{Park, K., Waldorp, L. \& Vasconcelos, V. V. The
+individual-and population-level mechanistic implications of statistical
+networks of symptoms. (2025).}
+
+\bibitem[\citeproctext]{ref-cramer2010comorbidity}
+\CSLLeftMargin{31. }%
+\CSLRightInline{Cramer, A. O., Waldorp, L. J., Van Der Maas, H. L. \&
+Borsboom, D. Comorbidity: A network perspective. \emph{Behavioral and
+brain sciences} \textbf{33}, 137--150 (2010).}
+
+\bibitem[\citeproctext]{ref-wichers2016critical}
+\CSLLeftMargin{32. }%
+\CSLRightInline{Wichers, M., Groot, P. C., Psychosystems, E., Group, E.,
+\emph{et al.} Critical slowing down as a personalized early warning
+signal for depression. \emph{Psychotherapy and psychosomatics}
+\textbf{85}, 114--116 (2016).}
+
+\bibitem[\citeproctext]{ref-wagenaar2013public}
+\CSLLeftMargin{33. }%
+\CSLRightInline{Wagenaar, A. C. \& Burris, S. \emph{Public Health Law
+Research: Theory and Methods}. (John Wiley \& Sons, 2013).}
+
+\bibitem[\citeproctext]{ref-desavigny2009systems}
+\CSLLeftMargin{34. }%
+\CSLRightInline{\emph{Systems Thinking for Health Systems
+Strengthening}. (Alliance for Health Policy; Systems Research, World
+Health Organization, Geneva, 2009).}
+
+\bibitem[\citeproctext]{ref-galenkamp2017measurement}
+\CSLLeftMargin{35. }%
+\CSLRightInline{Galenkamp, H., Stronks, K., Snijder, M. B. \& Derks, E.
+M. Measurement invariance testing of the PHQ-9 in a multi-ethnic
+population in europe: The HELIUS study. \emph{BMC psychiatry}
+\textbf{17}, 1--14 (2017).}
+
+\bibitem[\citeproctext]{ref-kroenke2001phq}
+\CSLLeftMargin{36. }%
+\CSLRightInline{Kroenke, K., Spitzer, R. L. \& Williams, J. B. The
+PHQ-9: Validity of a brief depression severity measure. \emph{Journal of
+general internal medicine} \textbf{16}, 606--613 (2001).}
+
+\bibitem[\citeproctext]{ref-mooij2020constraint}
+\CSLLeftMargin{37. }%
+\CSLRightInline{Mooij, J. M. \& Claassen, T. Constraint-based causal
+discovery using partial ancestral graphs in the presence of cycles. in
+\emph{Conference on uncertainty in artificial intelligence} 1159--1168
+(Pmlr, 2020).}
+
+\bibitem[\citeproctext]{ref-strobl2019}
+\CSLLeftMargin{38. }%
+\CSLRightInline{Strobl, E. V.
+\href{https://doi.org/10.1007/s41060-018-0158-2}{A constraint-based
+algorithm for causal discovery with cycles, latent variables and
+selection bias}. \emph{International Journal of Data Science and
+Analytics} \textbf{8}, 33--56 (2019).}
+
+\bibitem[\citeproctext]{ref-spirtes2001causation}
+\CSLLeftMargin{39. }%
+\CSLRightInline{Spirtes, P., Glymour, C. \& Scheines, R.
+\emph{Causation, Prediction, and Search}. (MIT press, 2001).}
+
+\bibitem[\citeproctext]{ref-rahimi2007random}
+\CSLLeftMargin{40. }%
+\CSLRightInline{Rahimi, A. \& Recht, B. Random features for large-scale
+kernel machines. \emph{Advances in neural information processing
+systems} \textbf{20}, (2007).}
+
+\bibitem[\citeproctext]{ref-strobl2019approximate}
+\CSLLeftMargin{41. }%
+\CSLRightInline{Strobl, E. V., Zhang, K. \& Visweswaran, S. Approximate
+kernel-based conditional independence tests for fast non-parametric
+causal discovery. \emph{Journal of Causal Inference} \textbf{7},
+20180017 (2019).}
+
+\bibitem[\citeproctext]{ref-zhang2012kernel}
+\CSLLeftMargin{42. }%
+\CSLRightInline{Zhang, K., Peters, J., Janzing, D. \& Schölkopf, B.
+Kernel-based conditional independence test and application in causal
+discovery. \emph{arXiv preprint arXiv:1202.3775} (2012).}
+
+\bibitem[\citeproctext]{ref-park2024discovering}
+\CSLLeftMargin{43. }%
+\CSLRightInline{Park, K., Waldorp, L. J. \& Ryan, O. Discovering cyclic
+causal models in psychological research. \emph{advances. in/psychology}
+\textbf{2}, e72425 (2024).}
+
+\bibitem[\citeproctext]{ref-forre2018constraint}
+\CSLLeftMargin{44. }%
+\CSLRightInline{Forré, P. \& Mooij, J. M. Constraint-based causal
+discovery for non-linear structural causal models with cycles and latent
+confounders. \emph{arXiv preprint arXiv:1807.03024} (2018).}
+
+\bibitem[\citeproctext]{ref-lindsay2000moment}
+\CSLLeftMargin{45. }%
+\CSLRightInline{Lindsay, B. G., Pilla, R. S. \& Basak, P. Moment-based
+approximations of distributions using mixtures: Theory and applications.
+\emph{Annals of the Institute of Statistical Mathematics} \textbf{52},
+215--230 (2000).}
 
 \bibitem[\citeproctext]{ref-butterworth2012role}
-Butterworth, P., Olesen, S. C., \& Leach, L. S. (2012). The role of
-hardship in the association between socio-economic position and
-depression. \emph{Australian \& New Zealand Journal of Psychiatry},
-\emph{46}(4), 364--373.
-
-\bibitem[\citeproctext]{ref-compton2015social}
-Compton, M. T., \& Shim, R. S. (2015). The social determinants of mental
-health. \emph{Focus}, \emph{13}(4), 419--425.
-
-\bibitem[\citeproctext]{ref-cramer2010comorbidity}
-Cramer, A. O., Waldorp, L. J., Van Der Maas, H. L., \& Borsboom, D.
-(2010). Comorbidity: A network perspective. \emph{Behavioral and Brain
-Sciences}, \emph{33}(2-3), 137--150.
+\CSLLeftMargin{46. }%
+\CSLRightInline{Butterworth, P., Olesen, S. C. \& Leach, L. S. The role
+of hardship in the association between socio-economic position and
+depression. \emph{Australian \& New Zealand Journal of Psychiatry}
+\textbf{46}, 364--373 (2012).}
 
 \bibitem[\citeproctext]{ref-deb2002fast}
-Deb, K., Pratap, A., Agarwal, S., \& Meyarivan, T. (2002). A fast and
-elitist multiobjective genetic algorithm: NSGA-II. \emph{IEEE
-Transactions on Evolutionary Computation}, \emph{6}(2), 182--197.
+\CSLLeftMargin{47. }%
+\CSLRightInline{Deb, K., Pratap, A., Agarwal, S. \& Meyarivan, T. A fast
+and elitist multiobjective genetic algorithm: NSGA-II. \emph{IEEE
+transactions on evolutionary computation} \textbf{6}, 182--197 (2002).}
 
 \bibitem[\citeproctext]{ref-dojer2016learning}
-Dojer, N. (2016). Learning bayesian networks from datasets joining
-continuous and discrete variables. \emph{International Journal of
-Approximate Reasoning}, \emph{78}, 116--124.
-
-\bibitem[\citeproctext]{ref-elsenburg2025clustering}
-Elsenburg, L. K., Nicolaou, M., Galenkamp, H., Lakerveld, J., \&
-Stronks, K. (2025). The clustering of disadvantage in different life
-dimensions across ethnic groups: A network analysis of indicators of
-precariousness in the HELIUS study. \emph{Social Science \& Medicine},
-117970.
+\CSLLeftMargin{48. }%
+\CSLRightInline{Dojer, N. Learning bayesian networks from datasets
+joining continuous and discrete variables. \emph{International Journal
+of Approximate Reasoning} \textbf{78}, 116--124 (2016).}
+
+\bibitem[\citeproctext]{ref-epskamp2018personalized}
+\CSLLeftMargin{49. }%
+\CSLRightInline{Epskamp, S. \emph{et al.} Personalized network modeling
+in psychopathology: The importance of contemporaneous and temporal
+connections. \emph{Clinical Psychological Science} \textbf{6}, 416--427
+(2018).}
 
 \bibitem[\citeproctext]{ref-entner2010causal}
-Entner, D., \& Hoyer, P. O. (2010). On causal discovery from time series
-data using FCI. \emph{Probabilistic Graphical Models}, \emph{16}.
-
-\bibitem[\citeproctext]{ref-epskamp2018personalized}
-Epskamp, S., Borkulo, C. D. van, Veen, D. C. van der, Servaas, M. N.,
-Isvoranu, A.-M., Riese, H., \& Cramer, A. O. (2018). Personalized
-network modeling in psychopathology: The importance of contemporaneous
-and temporal connections. \emph{Clinical Psychological Science},
-\emph{6}(3), 416--427.
-
-\bibitem[\citeproctext]{ref-ezzy1993unemployment}
-Ezzy, D. (1993). Unemployment and mental health: A critical review.
-\emph{Social Science \& Medicine}, \emph{37}(1), 41--52.
-
-\bibitem[\citeproctext]{ref-fone2014effect}
-Fone, D., White, J., Farewell, D., Kelly, M., John, G., Lloyd, K.,
-Williams, G., \& Dunstan, F. (2014). Effect of neighbourhood deprivation
-and social cohesion on mental health inequality: A multilevel
-population-based longitudinal study. \emph{Psychological Medicine},
-\emph{44}(11), 2449--2460.
-
-\bibitem[\citeproctext]{ref-forre2018constraint}
-Forré, P., \& Mooij, J. M. (2018). Constraint-based causal discovery for
-non-linear structural causal models with cycles and latent confounders.
-\emph{arXiv Preprint arXiv:1807.03024}.
+\CSLLeftMargin{50. }%
+\CSLRightInline{Entner, D. \& Hoyer, P. O. On causal discovery from time
+series data using FCI. \emph{Probabilistic graphical models}
+\textbf{16}, (2010).}
+
+\bibitem[\citeproctext]{ref-lorant2003socioeconomic}
+\CSLLeftMargin{51. }%
+\CSLRightInline{Lorant, V. \emph{et al.} Socioeconomic inequalities in
+depression: A meta-analysis. \emph{American journal of epidemiology}
+\textbf{157}, 98--112 (2003).}
+
+\bibitem[\citeproctext]{ref-neapolitan2004learning}
+\CSLLeftMargin{52. }%
+\CSLRightInline{Neapolitan, R. E. \emph{et al.} \emph{Learning Bayesian
+Networks}. vol. 38 (Pearson Prentice Hall Upper Saddle River, 2004).}
 
 \bibitem[\citeproctext]{ref-nsga2R}
-Franz, R., \& Nakamura, R. (2015). \emph{nsga2R: Elitist non-dominated
-sorting genetic algorithm for multi-objective optimization}.
-\url{https://CRAN.R-project.org/package=nsga2R}
-
-\bibitem[\citeproctext]{ref-fryers2003social}
-Fryers, T., Melzer, D., \& Jenkins, R. (2003). Social inequalities and
-the common mental disorders: A systematic review of the evidence.
-\emph{Social Psychiatry and Psychiatric Epidemiology}, \emph{38}(5),
-229--237.
-
-\bibitem[\citeproctext]{ref-galenkamp2017measurement}
-Galenkamp, H., Stronks, K., Snijder, M. B., \& Derks, E. M. (2017).
-Measurement invariance testing of the PHQ-9 in a multi-ethnic population
-in europe: The HELIUS study. \emph{BMC Psychiatry}, \emph{17}, 1--14.
-
-\bibitem[\citeproctext]{ref-gruebner2017cities}
-Gruebner, O., Rapp, M. A., Adli, M., Kluge, U., Galea, S., \& Heinz, A.
-(2017). Cities and mental health. \emph{Deutsches {Ä}rzteblatt
-International}, \emph{114}(8), 121.
-
-\bibitem[\citeproctext]{ref-howlett2001radial}
-Howlett, R. J. (2001). \emph{Radial basis function networks 1: Recent
-developments in theory and applications}.
-
-\bibitem[\citeproctext]{ref-hyttinen2012learning}
-Hyttinen, A., Eberhardt, F., \& Hoyer, P. O. (2012). Learning linear
-cyclic causal models with latent variables. \emph{The Journal of Machine
-Learning Research}, \emph{13}(1), 3387--3439.
-
-\bibitem[\citeproctext]{ref-kroenke2001phq}
-Kroenke, K., Spitzer, R. L., \& Williams, J. B. (2001). The PHQ-9:
-Validity of a brief depression severity measure. \emph{Journal of
-General Internal Medicine}, \emph{16}(9), 606--613.
-
-\bibitem[\citeproctext]{ref-van2014critical}
-Leemput, I. A. van de, Wichers, M., Cramer, A. O., Borsboom, D.,
-Tuerlinckx, F., Kuppens, P., Van Nes, E. H., Viechtbauer, W., Giltay, E.
-J., Aggen, S. H., et al. (2014). Critical slowing down as early warning
-for the onset and termination of depression. \emph{Proceedings of the
-National Academy of Sciences}, \emph{111}(1), 87--92.
-
-\bibitem[\citeproctext]{ref-lindsay2000moment}
-Lindsay, B. G., Pilla, R. S., \& Basak, P. (2000). Moment-based
-approximations of distributions using mixtures: Theory and applications.
-\emph{Annals of the Institute of Statistical Mathematics}, \emph{52},
-215--230.
-
-\bibitem[\citeproctext]{ref-lorant2003socioeconomic}
-Lorant, V., Deliège, D., Eaton, W., Robert, A., Philippot, P., \&
-Ansseau, M. (2003). Socioeconomic inequalities in depression: A
-meta-analysis. \emph{American Journal of Epidemiology}, \emph{157}(2),
-98--112.
-
-\bibitem[\citeproctext]{ref-lund2010poverty}
-Lund, C., Breen, A., Flisher, A. J., Kakuma, R., Corrigall, J., Joska,
-J. A., Swartz, L., \& Patel, V. (2010). Poverty and common mental
-disorders in low and middle income countries: A systematic review.
-\emph{Social Science \& Medicine}, \emph{71}(3), 517--528.
-
-\bibitem[\citeproctext]{ref-lund2018social}
-Lund, C., Brooke-Sumner, C., Baingana, F., Baron, E. C., Breuer, E.,
-Chandra, P., Haushofer, J., Herrman, H., Jordans, M., Kieling, C., et
-al. (2018). Social determinants of mental disorders and the sustainable
-development goals: A systematic review of reviews. \emph{The Lancet
-Psychiatry}, \emph{5}(4), 357--369.
-
-\bibitem[\citeproctext]{ref-mcdaid2019economic}
-McDaid, D., Park, A.-L., \& Wahlbeck, K. (2019). The economic case for
-the prevention of mental illness. \emph{Annual Review of Public Health},
-\emph{40}(1), 373--389.
-
-\bibitem[\citeproctext]{ref-mckee2017living}
-McKee, M., Reeves, A., Clair, A., \& Stuckler, D. (2017). Living on the
-edge: Precariousness and why it matters for health. \emph{Archives of
-Public Health}, \emph{75}, 1--10.
-
-\bibitem[\citeproctext]{ref-mooij2020constraint}
-Mooij, J. M., \& Claassen, T. (2020). Constraint-based causal discovery
-using partial ancestral graphs in the presence of cycles.
-\emph{Conference on Uncertainty in Artificial Intelligence}, 1159--1168.
-
-\bibitem[\citeproctext]{ref-neapolitan2004learning}
-Neapolitan, R. E. et al. (2004). \emph{Learning bayesian networks} (Vol.
-38). Pearson Prentice Hall Upper Saddle River.
-
-\bibitem[\citeproctext]{ref-park2025role}
-Park, K., Li, X., Waldorp, L., Lees, M., \& Vasconcelos, V. V. (2025).
-\emph{The role of feedback loops in dynamical symptom networks}.
-
-\bibitem[\citeproctext]{ref-park2024discovering}
-Park, K., Waldorp, L. J., \& Ryan, O. (2024). Discovering cyclic causal
-models in psychological research. \emph{Advances. In/Psychology},
-\emph{2}, e72425.
-
-\bibitem[\citeproctext]{ref-park2025individual}
-Park, K., Waldorp, L., \& Vasconcelos, V. V. (2025). \emph{The
-individual-and population-level mechanistic implications of statistical
-networks of symptoms}.
-
-\bibitem[\citeproctext]{ref-pevalin2017impact}
-Pevalin, D. J., Reeves, A., Baker, E., \& Bentley, R. (2017). The impact
-of persistent poor housing conditions on mental health: A longitudinal
-population-based study. \emph{Preventive Medicine}, \emph{105},
-304--310.
-
-\bibitem[\citeproctext]{ref-rahimi2007random}
-Rahimi, A., \& Recht, B. (2007). Random features for large-scale kernel
-machines. \emph{Advances in Neural Information Processing Systems},
-\emph{20}.
-
-\bibitem[\citeproctext]{ref-rehm2019global}
-Rehm, J., \& Shield, K. D. (2019). Global burden of disease and the
-impact of mental and addictive disorders. \emph{Current Psychiatry
-Reports}, \emph{21}, 1--7.
+\CSLLeftMargin{53. }%
+\CSLRightInline{Franz, R. \& Nakamura, R.
+\emph{\href{https://CRAN.R-project.org/package=nsga2R}{nsga2R: Elitist
+Non-Dominated Sorting Genetic Algorithm for Multi-Objective
+Optimization}}. (2015).}
 
 \bibitem[\citeproctext]{ref-ridley2020poverty}
-Ridley, M., Rao, G., Schilbach, F., \& Patel, V. (2020). Poverty,
-depression, and anxiety: Causal evidence and mechanisms. \emph{Science},
-\emph{370}(6522), eaay0214.
-
-\bibitem[\citeproctext]{ref-rohbeck2024}
-Rohbeck, M., Clarke, B., Mikulik, K., Pettet, A., Stegle, O., \&
-Ueltzhöffer, K. (2024). Bicycle: Intervention-based causal discovery
-with cycles. In F. Locatello \& V. Didelez (Eds.), \emph{Proceedings of
-the third conference on causal learning and reasoning} (Vol. 236, pp.
-209--242). PMLR.
-
-\bibitem[\citeproctext]{ref-ronnblad2019precarious}
-Rönnblad, T., Grönholm, E., Jonsson, J., Koranyi, I., Orellana, C.,
-Kreshpaj, B., Chen, L., Stockfelt, L., \& Bodin, T. (2019). Precarious
-employment and mental health. \emph{Scandinavian Journal of Work,
-Environment \& Health}, \emph{45}(5), 429--443.
-
-\bibitem[\citeproctext]{ref-rugulies2023work}
-Rugulies, R., Aust, B., Greiner, B. A., Arensman, E., Kawakami, N.,
-LaMontagne, A. D., \& Madsen, I. E. (2023). Work-related causes of
-mental health conditions and interventions for their improvement in
-workplaces. \emph{The Lancet}, \emph{402}(10410), 1368--1381.
+\CSLLeftMargin{54. }%
+\CSLRightInline{Ridley, M., Rao, G., Schilbach, F. \& Patel, V. Poverty,
+depression, and anxiety: Causal evidence and mechanisms. \emph{Science}
+\textbf{370}, eaay0214 (2020).}
 
 \bibitem[\citeproctext]{ref-runge2019detecting}
-Runge, J., Nowack, P., Kretschmer, M., Flaxman, S., \& Sejdinovic, D.
-(2019). Detecting and quantifying causal associations in large nonlinear
-time series datasets. \emph{Science Advances}, \emph{5}(11), eaau4996.
-
-\bibitem[\citeproctext]{ref-desavigny2009systems}
-Savigny, D. de, \& Adam, T. (Eds.). (2009). \emph{Systems thinking for
-health systems strengthening}. Alliance for Health Policy; Systems
-Research, World Health Organization.
-
-\bibitem[\citeproctext]{ref-sethuraman2023nodags}
-Sethuraman, M. G., Lopez, R., Mohan, R., Fekri, F., Biancalani, T., \&
-Hütter, J.-C. (2023). NODAGS-flow: Nonlinear cyclic causal structure
-learning. \emph{International Conference on Artificial Intelligence and
-Statistics}, 6371--6387.
-
-\bibitem[\citeproctext]{ref-snijder2017cohort}
-Snijder, M. B., Galenkamp, H., Prins, M., Derks, E. M., Peters, R. J.,
-Zwinderman, A. H., \& Stronks, K. (2017). Cohort profile: The healthy
-life in an urban setting (HELIUS) study in amsterdam, the netherlands.
-\emph{BMJ Open}, \emph{7}(12), e017873.
-
-\bibitem[\citeproctext]{ref-spirtes2001causation}
-Spirtes, P., Glymour, C., \& Scheines, R. (2001). \emph{Causation,
-prediction, and search}. MIT press.
+\CSLLeftMargin{55. }%
+\CSLRightInline{Runge, J., Nowack, P., Kretschmer, M., Flaxman, S. \&
+Sejdinovic, D. Detecting and quantifying causal associations in large
+nonlinear time series datasets. \emph{Science advances} \textbf{5},
+eaau4996 (2019).}
 
 \bibitem[\citeproctext]{ref-spirtes_causal_1995}
-Spirtes, P., Meek, C., \& Richardson, T. (1995). Causal inference in the
-presence of latent variables and selection bias. \emph{Proceedings of
-the {Eleventh} Conference on {Uncertainty} in Artificial Intelligence},
-499--506.
+\CSLLeftMargin{56. }%
+\CSLRightInline{Spirtes, P., Meek, C. \& Richardson, T. Causal inference
+in the presence of latent variables and selection bias. in
+\emph{Proceedings of the {Eleventh} conference on {Uncertainty} in
+artificial intelligence} 499--506 (Morgan Kaufmann, 1995).}
 
 \bibitem[\citeproctext]{ref-storn1997differential}
-Storn, R., \& Price, K. (1997). Differential evolution--a simple and
-efficient heuristic for global optimization over continuous spaces.
-\emph{Journal of Global Optimization}, \emph{11}, 341--359.
-
-\bibitem[\citeproctext]{ref-strobl2019}
-Strobl, E. V. (2019). A constraint-based algorithm for causal discovery
-with cycles, latent variables and selection bias. \emph{International
-Journal of Data Science and Analytics}, \emph{8}(1), 33--56.
-\url{https://doi.org/10.1007/s41060-018-0158-2}
-
-\bibitem[\citeproctext]{ref-strobl2019approximate}
-Strobl, E. V., Zhang, K., \& Visweswaran, S. (2019). Approximate
-kernel-based conditional independence tests for fast non-parametric
-causal discovery. \emph{Journal of Causal Inference}, \emph{7}(1),
-20180017.
-
-\bibitem[\citeproctext]{ref-stronks2013unravelling}
-Stronks, K., Snijder, M. B., Peters, R. J., Prins, M., Schene, A. H., \&
-Zwinderman, A. H. (2013). Unravelling the impact of ethnicity on health
-in europe: The HELIUS study. \emph{BMC Public Health}, \emph{13}, 1--10.
-
-\bibitem[\citeproctext]{ref-van2021advancing}
-Van Der Wal, J. M., Van Borkulo, C. D., Deserno, M. K., Breedvelt, J.
-J., Lees, M., Lokman, J. C., Borsboom, D., Denys, D., Holst, R. J. van,
-Smidt, M. P., et al. (2021). Advancing urban mental health research:
-From complexity science to actionable targets for intervention.
-\emph{The Lancet Psychiatry}, \emph{8}(11), 991--1000.
-
-\bibitem[\citeproctext]{ref-wagenaar2013public}
-Wagenaar, A. C., \& Burris, S. (2013). \emph{Public health law research:
-Theory and methods}. John Wiley \& Sons.
-
-\bibitem[\citeproctext]{ref-weich1998poverty}
-Weich, S., \& Lewis, G. (1998). Poverty, unemployment, and common mental
-disorders: Population based cohort study. \emph{Bmj}, \emph{317}(7151),
-115--119.
-
-\bibitem[\citeproctext]{ref-wichers2016critical}
-Wichers, M., Groot, P. C., Psychosystems, E., Group, E., et al. (2016).
-Critical slowing down as a personalized early warning signal for
-depression. \emph{Psychotherapy and Psychosomatics}, \emph{85}(2),
-114--116.
-
-\bibitem[\citeproctext]{ref-WHO_2022}
-World Health Organization. (2022). \emph{Mental disorders}. World Health
-Organization.
-\url{https://www.who.int/news-room/fact-sheets/detail/mental-disorders}
-
-\bibitem[\citeproctext]{ref-zawadzki2013rumination}
-Zawadzki, M. J., Graham, J. E., \& Gerin, W. (2013). Rumination and
-anxiety mediate the effect of loneliness on depressed mood and sleep
-quality in college students. \emph{Health Psychology}, \emph{32}(2),
-212.
-
-\bibitem[\citeproctext]{ref-zhang2012kernel}
-Zhang, K., Peters, J., Janzing, D., \& Schölkopf, B. (2012).
-Kernel-based conditional independence test and application in causal
-discovery. \emph{arXiv Preprint arXiv:1202.3775}.
+\CSLLeftMargin{57. }%
+\CSLRightInline{Storn, R. \& Price, K. Differential evolution--a simple
+and efficient heuristic for global optimization over continuous spaces.
+\emph{Journal of global optimization} \textbf{11}, 341--359 (1997).}
 
@@
 conditional independence relationships, ensuring that the same
-\emph{d-separation} conditions hold across all graphs in the class
-(Spirtes et al., 2001). Conversely, an edge marked as \(A \startail B\)
-indicates that \(B\) is an ancestor of \(A\) across all graphs in
-\(Equiv(G)\). Circle endpoints, \(A \starcirc B\), represent ambiguity
-in the ancestral relationship, meaning \(B\)'s ancestral status relative
-to \(A\) varies across graphs in \(Equiv(G)\).\footnote{\(*\) serves as
-  a \textit{meta-symbol}, representing one of the three possible
-  edge-endpoints. For instance, \(A \tailstar B\) can indicate any of
-  the following edges: \(A\) \textemdash~\(B\), \(A\tailarrow B\), or
-  \(A \tailcirc B\) (Park et al., 2024).} Finally, when an edge is
+\emph{d-separation} conditions hold across all graphs in the
+class\textsuperscript{39}. Conversely, an edge marked as
+\(A \startail B\) indicates that \(B\) is an ancestor of \(A\) across
+all graphs in \(Equiv(G)\). Circle endpoints, \(A \starcirc B\),
+represent ambiguity in the ancestral relationship, meaning \(B\)'s
+ancestral status relative to \(A\) varies across graphs in
+\(Equiv(G)\).\footnote{\(*\) serves as a \textit{meta-symbol},
+  representing one of the three possible edge-endpoints. For instance,
+  \(A \tailstar B\) can indicate any of the following edges: \(A\)
+  \textemdash~\(B\), \(A\tailarrow B\), or
+  \(A \tailcirc B\)\textsuperscript{43}.} Finally, when an edge is
 represented as \(A \arrowarrow B\), it implies that neither \(A\) nor
@@
 distinguish variables that are not part of a cycle, offering a more
-nuanced approach to handling cyclic relationships (Mooij \& Claassen,
-2020).
+nuanced approach to handling cyclic relationships\textsuperscript{37}.
 
@@
 general faithfulness and Markov conditions (i.e.,
-\emph{\(\sigma\)-separation} and \emph{\(\sigma\)-faithfulness} setting)
-(Forré \& Mooij, 2018). When the respective assumptions of each
+\emph{\(\sigma\)-separation} and \emph{\(\sigma\)-faithfulness}
+setting)\textsuperscript{44}. When the respective assumptions of each
 algorithm are met, their inferred cyclic relationships align with those
@@
 not always fully preserve d-separation relations from the original
-graph, \(\mathcal{G}\) (Strobl, 2019). Given these considerations, we
-examine and compare results from both algorithms, placing more weight on
-the FCI results, which are presented in the main results section, while
-discussing CCI results where relevant, with detailed findings provided
-in Section~\ref{sec-cci}. A full discussion of causal discovery concepts
-and algorithmic details is beyond the scope of this paper; however,
-readers seeking a more in-depth understanding can refer to Park et al.
-(2024) for a comprehensive exploration of these methods and their
-applications.
+graph, \(\mathcal{G}\)\textsuperscript{38}. Given these considerations,
+we examine and compare results from both algorithms, placing more weight
+on the FCI results, which are presented in the main results section,
+while discussing CCI results where relevant, with detailed findings
+provided in Section~\ref{sec-cci}. A full discussion of causal discovery
+concepts and algorithmic details is beyond the scope of this paper;
+however, readers seeking a more in-depth understanding can refer
+to\textsuperscript{43} for a comprehensive exploration of these methods
+and their applications.
 
@@
 \begin{center}
-\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-12-1.pdf}
+\includegraphics{draft_v3_files/figure-pdf/unnamed-chunk-12-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-13-1.pdf}
+\includegraphics[width=0.55\textwidth,height=\textheight]{draft_v3_files/figure-pdf/unnamed-chunk-13-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-14-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v3_files/figure-pdf/unnamed-chunk-14-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-15-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v3_files/figure-pdf/unnamed-chunk-15-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-16-1.pdf}
+\includegraphics{draft_v3_files/figure-pdf/unnamed-chunk-16-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-17-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v3_files/figure-pdf/unnamed-chunk-17-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-18-1.pdf}
+\includegraphics[width=0.7\textwidth,height=\textheight]{draft_v3_files/figure-pdf/unnamed-chunk-18-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-19-1.pdf}
+\includegraphics{draft_v3_files/figure-pdf/unnamed-chunk-19-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-19-2.pdf}
+\includegraphics{draft_v3_files/figure-pdf/unnamed-chunk-19-2.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-19-3.pdf}
+\includegraphics{draft_v3_files/figure-pdf/unnamed-chunk-19-3.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-19-4.pdf}
+\includegraphics{draft_v3_files/figure-pdf/unnamed-chunk-19-4.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v2_files/figure-pdf/unnamed-chunk-20-1.pdf}
+\includegraphics[width=0.6\textwidth,height=\textheight]{draft_v3_files/figure-pdf/unnamed-chunk-20-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-21-1.pdf}
+\includegraphics{draft_v3_files/figure-pdf/unnamed-chunk-21-1.pdf}
 \end{center}
@@
 \begin{center}
-\includegraphics{draft_v2_files/figure-pdf/unnamed-chunk-22-1.pdf}
+\includegraphics{draft_v3_files/figure-pdf/unnamed-chunk-22-1.pdf}
 \end{center}
@@
 inherent in kernel-based CI tests. In this section, we provide a
-high-level overview of RCIT and RCoT based on (Strobl et al., 2019).
+high-level overview of RCIT and RCoT based on\textsuperscript{41}.
 
@@
 Independence Criterion (HSIC) in reproducing kernel Hilbert spaces
-(RKHS) (Zhang et al., 2012). KCIT uses the following hypothesis
+(RKHS)\textsuperscript{42}. KCIT uses the following hypothesis
 framework: \[
@@
 threshold determined by permutation or moment-matching-based null
-distribution (Lindsay et al., 2000).
+distribution\textsuperscript{45}.
 
